### 2017-07-14 notebook/ms/ms.Rmd
Add MS file

---
title: 'GMSE: an R package for generalised management strategy evaluation'
author: A. Bradley Duthie, Jeremy J. Cusack, Roc&#0237;o Pozo, O. Sarobidy Rakotonarivo,
  and Nils Bunnefeld
date: '`r Sys.Date()`'
output:
  pdf_document:
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
linestretch: 1
header-includes:
- \usepackage{natbib}
- \usepackage{lineno}
- \linenumbers
- \bibliographystyle{amnatnat}
bibliography: gmse.bib
---

******************************

```{r, echo = FALSE}
library(GMSE);
plot_gmse_results_vign <- function(res, obs, land, agents, paras, ACTION, COST){

    para_vec <- paras[1,]

    times <- 1;
    view  <- agents[[1]][1, 9];

    land1 <- NULL;
    land3 <- NULL;
    land2 <- land;

    for(i in 1:length(land2)){
        land1 <- land2[[i]][,,1];
        land3 <- land2[[i]][,,3];
    }

    max_time <- length(res) - 1;

    gens <- NULL;
    abun <- NULL;
    est  <- NULL;
    lci  <- NULL;
    uci  <- NULL;
    lnds <- NULL;
    ages <- NULL;
    stky <- NULL;
    land_cols <- c("#F2F2F2FF", "#ECB176FF", "#000000"); 
    cols      <- c("green", "indianred1", "indianred3", "deepskyblue1",
                   "deepskyblue2");

    case  <- para_vec[9];
    tiobs <- para_vec[12];
    if(case == 1 & tiobs < 2){
        return("No RMR possible"); 
    }
    mrk <- floor(tiobs / 2);
    rcp <- tiobs - mrk;

    max_action <- 0;
    max_cost   <- 0;
    for(i in 1:length(res)){
        act_check <- ACTION[[i]][,8:12,];
        act_check[act_check > 10000] <- -1;
        act_comb <- apply(X = act_check, MARGIN = c(1,2), FUN = sum);
        gen_max_action <- max(act_comb);
        if(gen_max_action > max_action){
            max_action <- gen_max_action;
        }
        cost_check <- COST[[i]][,8:12,];
        cost_check[cost_check >= 10000] <- -1;
        gen_max_cost   <- max(cost_check);
        if(gen_max_cost > max_cost){
            max_cost <- gen_max_cost;
        }
    }

    minK <- min(para_vec[6:7]);

    ymaxi    <- minK + (minK * (1 + res[[1]][1,10])); # Add for birth rate
    time_max <- length(res);
    for(i in 1:(time_max-1)){
        res_t    <- res[[i]];
        obs_t    <- obs[[i]];
        lnd_t    <- land2[[i]][,,2] * 100;
        age_t    <- agents[[i]];
        if(i > 1){
            res_t <- res_t[res_t[,12] >= para_vec[17],];
        }
        gens  <- c(gens, i);
        abun  <- c(abun, dim(res_t)[1]);
        lnds  <- c(lnds, mean(lnd_t));
        ages  <- rbind(ages, age_t[,16]);
        if(!is.null(obs_t) & case == 1){
            analysis <- chapman_est(observation = obs_t, paras = para_vec);
            est      <- c(est, analysis$Nc);
            lci      <- c(lci, analysis$lci);
            uci      <- c(uci, analysis$uci);
        }
        if(!is.null(obs_t) & !is.null(view) & case == 0){
            analysis <- dens_est(observation = obs_t, paras = para_vec, 
                                 view = view, land = land1);
            est      <- c(est, analysis$Nc);
            lci      <- c(lci, analysis$lci);
            uci      <- c(uci, analysis$uci);
        }
        for(stakeholder in 1:dim(ages)[2]){
            max_yield   <- sum(land3 == stakeholder);
            agent_yield <- rep(x = NA, max_time);
            if(max_yield > 0 & para_vec[104] > 0){
                agent_yield <- 100 * (ages[,stakeholder] / max_yield);
            }
            stky[[stakeholder]] <- agent_yield;
        }
    }
    if(case > 1){
        est <- paras[,100];
    }
    par(mfrow=c(3,2),mar=c(0,0,0,0));
    # ------------- Panel 1 (upper left)
    if(abun[max_time] > 0){
        indis  <- ind_to_land(inds = res_t, land = land1);
        image(indis, col = land_cols, xaxt="n", yaxt="n");
    }else{
        image(land1, col = land_cols, xaxt="n", yaxt="n");
    }
    # ------------- Panel 2 (upper right)
    col_num <- max(land3);
    image(land3, col = topo.colors(col_num), xaxt="n", yaxt="n");
    # ------------- Panel 3 (middle left)
    par(mar = c(4, 5, 1, 5));
    plot(x = gens, y = abun, pch = 20, type="l", lwd = 2, ylim = c(0, ymaxi),
         xlim=c(0, time_max), xlab = "Time Step", ylab = "Abundance",
         cex.lab=1.25);
    if(case == 0 | case == 1){
        polygon(y = c(lci,rev(uci)), x = c(gens, rev(gens)), border = NA,
                col="lightblue");
        points(x = gens, y = est, pch = 20, type = "l", lwd = 2, col = "cyan4");
    }else{
        points(x = gens, y = est[-time_max], pch = 20, type = "l", lwd = 2, 
               col = "cyan4");
    }
    abline(h = para_vec[7], col = "red", lwd = 0.8, lty = "dashed");
    abline(h = ACTION[[1]][1,5,1], col = topo.colors(1), lwd = 0.8, 
           lty = "dashed");
    points(x = gens, y = abun, pch = 20, type = "l", lwd = 3, col = "black");
    par(new = TRUE);
    plot(x = gens, y = lnds, pch = 20, type = "l", lwd = 3, col = "orange", 
         xlab = "", xlim = c(0, time_max), ylim = c(0, 100), xaxt="n", yaxt="n", 
         ylab = "");
    axis(side = 4, at = c(0, 25, 50, 75, 100));
    mtext("Mean % Yield", side = 4, line = 2.4);
    # ------------ Panel 4 (middle right);
    par(mar = c(4, 4.5, 1, 1));
    cell_number <- dim(land3)[1] * dim(land3)[2];
    plot(x = gens, y = gens, pch = 20, type = "n", lwd = 2, ylim = c(0, 100),
         xlim = c(0, time_max), xlab = "Time Step", 
         ylab = "Stake-holder % yield", cex.lab = 1.25);
    stake_colors <- topo.colors( dim(age_t)[1] );
    if(para_vec[104] > 0){
        for(stakeholder in 1:dim(ages)[2]){
            points(x = gens, y = stky[[stakeholder]], type="l", lwd=2, 
                   col = stake_colors[stakeholder]);
        }
    }
    # ------------- Panel 5 (lower left)
    res_costs <- matrix(data = 0, nrow = max_time, ncol = 5);
    for(j in 1:max_time){
        res_costs[j,1] <- ACTION[[j]][3,8,1];
        res_costs[j,2] <- ACTION[[j]][3,9,1];
        res_costs[j,3] <- ACTION[[j]][3,10,1];
        res_costs[j,4] <- ACTION[[j]][3,11,1];
        res_costs[j,5] <- ACTION[[j]][3,12,1];
    }
    par(mar = c(4, 5, 1, 5.5), xpd = TRUE);
    y_upper_limit <- max_cost + (0.25 * max_cost);
    plot(x = gens, y = gens, pch = 20, type = "n", lwd = 2, 
         ylim = c(0, y_upper_limit), xlim = c(0, time_max), xlab = "Time Step", 
         ylab = "Cost of actions", cex.lab = 1.25);
    if(para_vec[89] == 1){
        points(x = gens, y = res_costs[,1], type = "l", col = cols[1], lwd = 2);
    }
    if(para_vec[90] == 1){
        points(x = gens, y = res_costs[,2], type = "l", col = cols[2], lwd = 2);
    }
    if(para_vec[91] == 1){
        points(x = gens, y = res_costs[,3], type = "l", col = cols[3], lwd = 2);
    }
    if(para_vec[92] == 1){
        points(x = gens, y = res_costs[,4], type = "l", col = cols[4], lwd = 2);
    }
    if(para_vec[93] == 1){
        points(x = gens, y = res_costs[,5], type = "l", col = cols[5], lwd = 2);
    }
    legend(x = time_max + (time_max * 0.02), y = y_upper_limit, 
           fill = c(cols[1:5], "purple", "orange"), horiz = FALSE,
           legend = c("scaring", "culling", "castration", "feeding", 
                      "helping", "tend crop", "kill crop"), bty = "n");
    par(xpd = FALSE);
    # ------------- Panel 6 (lower right)
    res_acts <- matrix(data = 0, nrow = time_max, ncol = 7);
    gens     <- 1:time_max;
    for(j in 1:time_max){
        for(k in 2:dim(ACTION[[j]])[3]){
            res_acts[j,1] <- res_acts[j,1] + ACTION[[j]][1,8,k] - para_vec[96];
            res_acts[j,2] <- res_acts[j,2] + ACTION[[j]][1,9,k] - para_vec[96];
            res_acts[j,3] <- res_acts[j,3] + ACTION[[j]][1,10,k]- para_vec[96];
            res_acts[j,4] <- res_acts[j,4] + ACTION[[j]][1,11,k]- para_vec[96];
            res_acts[j,5] <- res_acts[j,5] + ACTION[[j]][1,12,k]- para_vec[96];
            res_acts[j,6] <- res_acts[j,6] + ACTION[[j]][2,10,k]- para_vec[96];
            res_acts[j,7] <- res_acts[j,7] + ACTION[[j]][2,11,k]- para_vec[96];
        }
    }
    par(mar=c(4, 4.5, 1, 1));
    y_upper_limit <- max_action + (0.25 * max_action);
    plot(x = gens, y = gens, pch = 20, type = "n", lwd = 2, 
         ylim = c(0, y_upper_limit), xlim = c(0, time_max), xlab = "Time Step", 
         ylab = "Actions made", cex.lab = 1.25);
    if(para_vec[89] == 1){
        points(x = gens, y = res_acts[,1], type = "l", col = cols[1], lwd = 2);
    }
    if(para_vec[90] == 1){
        points(x = gens, y = res_acts[,2], type = "l", col = cols[2], lwd = 2);
    }
    if(para_vec[91] == 1){
        points(x = gens, y = res_acts[,3], type = "l", col = cols[3], lwd=2);
    }
    if(para_vec[92] == 1){
        points(x=gens, y=res_acts[,4], type="l", col=cols[4], lwd=2);
    }
    if(para_vec[93] == 1){
        points(x = gens, y = res_acts[,5], type = "l", col = cols[5], lwd = 2);
    }
    if(para_vec[94] == 1){
        points(x = gens, y = res_acts[,6], type = "l", lty = "solid", 
               col = "purple", lwd = 3);
    }
    if(para_vec[95] == 1){
        points(x = gens, y = res_acts[,7], type = "l", lty = "solid", 
               col = "orange", lwd = 3);
    }
}
```

Abstract
===================================================

1. Management strategy evaluation (MSE) is a powerful tool for simulating all aspects of natural resource mangement under conditions of uncertainty.
2. We present the R package GMSE, which generalises MSE using a game-theoretic approach to simulate adaptive management scenarios under complex social-ecological interactions and uncertainty.
3. GMSE is agent-based and spatially explicit, and incorporates a high degree of realism and uncertainty through mechanistic modelling of the social-ecological system.
4. We show how GMSE can be used to simulate a social-ecological system, providing the example of management of a waterfowl population that exploits agricultural land, whose conservation is there in conflict with stakeholders.
5. The R package GMSE is open source under GNU Public License; source code and documents are freely available on GitHub.

Introduction
===================================================

Many global natural resources, including the biodiversity on which critical ecosystem services depend, are in a state of severe decline [@Dirzo2014; @Hautier2015; @Ceballos2017; @OConnell2017]. Conservation of biodiversity can be complicated by the immediate need to allocate natural resources and land area for human food production, causing real or perceived conflicts between conservation and food security and creating a challenge for the management of many natural resources [@Redpath2015a]. Given an increasing human population [@Crist2017], the number and intensity of such conflicts are likely to increase into the twenty first century. Effective management tools are therefore needed for the long-term maintenance of natural resources under the rising demand for food production.

To effectively manage natural resources, an adaptive approach allows managers to iteratively update their models of resource dynamics and respond flexibly to changing conditions. This approach can be further improved by considering all aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of managers and stakeholders [@Bunnefeld2011; @Bunnefeld2014]. Management strategy evaluation (MSE) is a modelling framework for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016]. Nevertheless, MSE remains limited in its ability to model human decision-making; manager decisions are typically based on simple rules, and stakeholder behaviour remains fixed over time instead of dynamically responding to changing resource availability and management decisions. Here we introduce a generalised management strategy evaluation (GMSE), which incorporates a game-theoretic perspective to model the goal-oriented, dynamic decision-making processes of managers and stakeholders.

The GMSE R package is a flexible, highly mechanistic, agent-based modelling tool to simulate all aspects of natural resource management. GMSE considers a range of parameters to simulate resource dynamics and management policy options, and uses genetic algorithms to dynamically model manager and stake-holder decision-making. Genetic algorithms allow managers and stakeholders to find adaptive solutions to any simulated conditions given unique individual goals, allowing GMSE to model scenarios of conservation conflict.

GMSE model structure
===================================================

GMSE builds off of the traditional MSE framework, which includes four sub-models, each of which runs once in a single time step of the broader model (Figure 1). (1) A natural resources sub-model considers a population of discrete resources with individual traits (e.g., location, age) on a spatially-explicit landscape. This sub-model can include processes of resource birth, movement, interaction with the landscape, and death. The discrete nature of resource demographics naturally gives rise to stochasticity, and therefore uncertainty. Although the resource sub-model can interact with other sub-models, it is unique in not relying on them because ecological dynamics can be simulated in the absence of observation and management. (2) An observation sub-model simulates the process of data collection peformed by a manager to estimate resource abundance. Four types of data collection are permitted, including exhaustive resource counting on a subset of landscape cells [e.g., @Nuno2013], marking and recapturing a fixed number of resources, and exhaustive sampling of the whole landscape iteratively (during which resources might move). Imperfect sampling from all of these mechanisms of data collections generates observation uncertainty. (3) A manager sub-model simulates a manager's analysis of collected data to estimate resource abundance, then compares this estimated abundance with the manager's target abundance. Policy is developed by calling the genetic algorithm (see below), which works within a manager's constraints to find costs for stakeholder actions (e.g., culling, scaring, etc.) that minimise resource deviation from the target abundance, as informed by the predicted consequences of stakeholder actions and stakeholder action histories. After an adaptive policy is found, (4) a user sub-model is run that allows each stakeholder to perform actions that affect resources or landscape cells. Stakeholders respond to policy individually, each calling the genetic algorithm to find actions that maximise their own utilities (e.g., maximise resource use or landscape yield) within their imposed constraints. Once each stakeholder has found an adaptive strategy, stakeholder actions affect resources and landscape cells, feeding back into the resource sub-model.

```{r, echo = FALSE, fig.height = 4, fig.width = 4, fig.align = 'left', fig.cap = "Description of the generalised management strategy evaluation framework"}
mbox <- function(x0, x1, y0, y1){
    xx <- seq(from=x0, to=x1, length.out = 100);
    yy <- seq(from=y0, to=y1, length.out = 100);
    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
    return(list(x=xd, y=yd));
}
par(mar=c(0,0,0,0));
plot(x = 0, y = 0, type = "n", xlim = c(0, 100), ylim = c(0, 100), 
     xaxt = "n", yaxt = "n", xlab = "", ylab = "");
# Manager model 
manger_model_box <- mbox(x0 = 1, x1 = 30, y0 = 70, y1 = 99);
polygon(x = manger_model_box$x, y = manger_model_box$y, lwd = 3);
text(x = 15, y = 90, cex = 1.0, labels = "Manager");
text(x = 15, y = 83, cex = 1.0, labels = "model");
# User model
user_model_box <- mbox(x0 = 70, x1 = 99, y0 = 70, y1 = 99);
polygon(x = user_model_box$x, y = user_model_box$y, lwd = 3);
text(x = 85, y = 90, cex = 1.0, labels = "User");
text(x = 85, y = 83, cex = 1.0, labels = "model");
# Observation model
observe_model_box <- mbox(x0 = 1, x1 = 30, y0 = 1, y1 = 30);
polygon(x = observe_model_box$x, y = observe_model_box$y, lwd = 3);
text(x = 15, y = 23, cex = 1.0, labels = "Observation");
text(x = 15, y = 15, cex = 1.0, labels = "model");
# Resource model
resource_model_box <- mbox(x0 = 70, x1 = 99, y0 = 1, y1 = 30);
polygon(x = resource_model_box$x, y = resource_model_box$y, lwd = 3);
text(x = 85, y = 23, cex = 1.0, labels = "Natural");
text(x = 85, y = 15, cex = 1.0, labels = "resources");
text(x = 85, y = 7,  cex = 1.0, labels = "model");
# Game-theoretic (genetic algorithm) model
game_model_box <- mbox(x0 = 36, x1 = 65, y0 = 36, y1 = 65);
polygon(x = game_model_box$x, y = game_model_box$y, lwd = 3);
text(x=50, y=55, labels="Genetic", cex=1.0);
text(x=50, y=48, labels="algorithm", cex=1.0);
# Arrows indicating direction of GMSE model
arrows(x0=15, x1=15, y0=30, y1=70, lwd=2, length=0.15);
arrows(x0=30, x1=70, y0=85, y1=85, lwd=2, length=0.15);
arrows(x0=85, x1=85, y0=70, y1=30, lwd=2, length=0.15);
arrows(x0=70, x1=30, y0=15, y1=15, lwd=2, length=0.15);
arrows(x0=30, x1=36, y0=70, y1=65, lwd=2, length=0.075, code=3);
arrows(x0=65, x1=70, y0=65, y1=70, lwd=2, length=0.075, code=3);
text(x=50, y=90, labels="Policy", cex = 1.0);
text(x=50, y=20, labels="Monitoring", cex = 1.0);
text(x=10, y=48, labels="Indicators", srt=90, cex = 1.0);
text(x=90, y=52, labels="Off-take", srt=-90, cex = 1.0);
```

Genetic Algorithm
---------------------------------------------------

Game theory is the standard tool for understanding conflict bewteen rational agents, and is therefore the appropriate tool for modelling manager and stakeholder actions and addressing issues of cooperation and conflict in conservation [@Colyvan2011; @Lee2012; @Kark2015; @Adami2016; @Tilman2016]. A game-theoretic approach is compatible with agent-based modelling [@An2012; @Tesfatsion2017], but given the number of possible strategies available to agents, identifying optimal strategies is often intractable. Instead, genetic algorithms, which mimic the process of natural selection (mutation, recombination, selection, replacement), are useful to find adaptive solutions for game strategies [e.g., @Balmann2000; @Tu2000; @Hamblin2013].

In keeping with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for managers or stakeholders. Instead, within each time step of a simulation (Figure 1), a new genetic algorithm is called for each agent to heuristically find strategies that are well-adapted to achieving agent goals. Hence for the manager and each stakeholder, a unique population with a random set of possible values affecting policy or actions, respectively, is initialised. In each generation of the genetic algorithm, strategies crossover and mutate; when this results in strategies that are over-budget, values are iteratively decreased at random until budget constraints are satisfied. A fitness function then evaluates each strategy in the population, and a tournament is used to select the next generation of strategies [@Hamblin2013]. The genetic algorithm terminates when a minimum number of generations has passed and the increase in the fittest strategy between the current and previous generation is below some threshold. The highest fitness strategy in the population then becomes the agent's strategy in the current time step.

GMSE arguments and output
---------------------------------------------------

Simulations are run using the gmse() function, which offers a range of options for setting parameter values (see Table 1 for some select examples). Output of gmse() is an exhaustive list that includes all resources and observations, all manager and stakeholder decisions and actions, and all landscape properties in each time step of the simulation. Results are most easily interpreted visually, so a summary of simulation dynamics is plotted by default (the plot can also be called using the plot_gmse_results function). An example below shows how simulations are set and interprted.

|  Argument       | Default | Description                                                                                      |
|-----------------|---------|--------------------------------------------------------------------------------------------------|
|  time_max       |  100    | Maximum time steps in simulation                                                                 |
|  land_dim_1     |  100    | Width of the landscape (horizontal cells)                                                        |
|  land_dim_1     |  100    | Height of the landscape (vertical cells)                                                         |
|  res_movement   |  20     | Distance (cells) a resource can move in any direct (for movement rules, see res_move_type)       |
|  remove_pr      |  0      | Density-indepdent probability of resource mortality during a time step                           |
|  lambda         |  0.3    | Poisson rate parameter for resource offspring number produced during a time step                 |
|  agent_view     |  10     | How far managers can see on the landscape for resource counting when observe_type = 0            |
|  res_birth_K    |  10000  | Carrying capacity applied to the number of resources added during a time step                    |
|  res_death_K    |  600    | Carrying capacity applied to the number of resources removed during a time step                  |
|  res_move_type  |  1      | Type of resource movement (default up to res_movement cells in any direction)                    |
|  observe_type   |  0      | Type of resource observation (default is density-based; i.e, counting a subset on the landscape) |
|  fixed_mark     |  50     | For mark-recapture observation (observe_type = 1), number of marked resources                    |
|  fixed_recapt   |  150    | For mark-recapture observation (observe_type = 1), number of recaptured resources                |
|  times_observe  |  8      | For density-based observation (observe_type = 0), landscape subsets viewed during observation    |
|  res_consume    |  0.5    | Proportion of a landscape cell's value reduced by the presence of a resource in a time step      |
|  max_ages       |  5      | The maximum number of time steps a resource can persist before it is removed                     |
|  minimum_cost   |  10     | The minimum cost of a stakeholder performing any action                                          |
|  user_budget    |  1000   | A stakeholder's budget per time step for performing any number of actions                        |
|  manager_budget |  1000   | A manager's budget per time step for setting policy                                              |
|  manage_target  |  300    | The manager's target resource abundance                                                          |
|  RESOURCE_ini   |  300    | The initial abundance of resources                                                               |
|  scaring        |  FALSE  | Resource scaring is a policy option                                                              |
|  culling        |  TRUE   | Resource culling is a policy option                                                              |
|  castration     |  FALSE  | Resource castration is a policy option                                                           |
|  feeding        |  FALSE  | Resource feeding (increases lambda) is a policy option                                           |
|  help_offspring |  FALSE  | Resource helping (increases offspring number) is a policy option                                 |
|  tend_crops     |  FALSE  | Stakeholders can increase landscape cell values                                                  |
|  tend_crop_yld  |  FALSE  | Proportional increase per landscape cell from tend_crops action                                  |
|  kill_crops     |  FALSE  | Stakeholders can decrease landscape cell values to zero                                          |
|  stakeholders   |  4      | Number of stakeholders in the simulation                                                         |
|  land_ownership |  FALSE  | Stakeholders own land and increase utility indirectly from landscape instead of resource use     |
|  manage_freq    |  1      | Frequency (in time steps) with which managers enact policy                                       |
|  public_land    |  0      | Proportion of land that is public (un-owned by stakeholders) if land_ownership = TRUE            |

**Table 1:** Select parameter values for initialising generalised management strategy evaluation simulations

An example of crop management
===================================================

Here we consider the example of a managed natural resources whose abundance affects a group of stakeholders by temporarily decreasing the value of stakeholder land. This scenario could be interpreted in multiple ways; we consider a protected population of waterfowl that exploits agricultural land causing a conservation conflict with farmers [e.g., @Tuvendal2015; @Fox2017a; @Fox2016a]. Managers in this example might want to keep the abundance of waterfowl at a target level, while farmers might want to minimise the damage inflicted on their crops. Using GMSE, we can simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy. Here we consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity (applied to mortality) is 2000. Waterfowl consume and destroy all crop yield upon arrival to a landscape cell. Managers estimate population size in each time step using mark-recapture techniques, and use their estimates to set costs of culling and scaring waterfowl (i.e., 'policy') for ten farmers. Farmers attempt to reduce the negative impact of waterfowl on the cropland that they own, working within the constraints of culling and scaring costs and their budget for performing these actions.

```{r, echo = TRUE}
sim <- gmse(land_ownership = TRUE, stakeholders = 10, observe_type = 1, 
            res_death_K = 2000, manage_target = 1000, RESOURCE_ini = 1000, 
            user_budget = 5000, manager_budget = 5000, res_consume = 1, 
            scaring = TRUE, fixed_mark = 50, fixed_recapt = 300, 
            plotting = FALSE);
```

Parameters in gmse() not listed are set to default values. By plotting the output with plot_gmse_results, simulation results can be interpreted visually.

```{r, eval = FALSE}
plot_gmse_results(res = sim$resource, obs = sim$observation, land = sim$land, 
                  agents = sim$agents, paras = sim$paras, ACTION = sim$action, 
                  COST = sim$cost);
```

```{r, echo = FALSE, fig.height = 6, fig.align = 'left', fig.cap = "Results of an example simulation illustrating the management of a protected resource that exploits the land of 10 stakeholders. The upper left panel shows locations of resources (black dots) on the landscape in the final time step of the simulation. The upper right panel shows the same landscape broken down into 10 differently coloured regions, which correspond to areas of land owned by each of the 10 stakeholders. The middle left panel shows the actual abundance of resources (black solid line, and the abundance of resources as estimated by the manager (blue solid line; shading indicates 95 percent confidence intervals of a mark-recapture analysis), over time. The horizontal dotted red and blue lines show the resource carrying capacity enacted on adult mortality and the manager's target for resource abundance, respectively. The orange line shows the total percent yield of landscape cells. The middle right panel shows total percent yield of landscape cells for each individual farmer, differentiated by colour, where line colours correspond to areas of the landscape in the upper right panel. The lower left panel shows the cost of stakeholder performing actions over time, as set by the manager. The lower right panel shows the total number of actions attempted to be performed by all stakeholders over time (some actions might be unsuccessful if resources are unavailable on a stakeholder's land to cull or scare, so, e.g., culling actions might be larger than resources actually culled)."}
plot_gmse_results_vign(res = sim$resource, obs = sim$observation, 
                       land = sim$land, agents = sim$agents, paras = sim$paras,
                       ACTION = sim$action, COST = sim$cost);
```

Figure 2 shows the landscape broken down by resource position and stakeholder ownership in the upper left and right hand panels, respectively. The waterfowl population fluctuates around the manager's target size of 1000, but the manager's estimate of population size deviates from its actual size due to uncertainty caused by the observation process (compare black and blue lines the middle left panel). Because the waterfowl have a direct negative effect on landscape yield, total landscape yield (orange line of the middle left panel), along with the yield of individual farmers (right middle panel), is low when waterfowl abundance is high, and vice versa.

Only the estimates of population size from the observation model are available to the manager, so policy change at any time step is driven primarily by the deviation of the currently estimated population size from the manager's target and the actions of farmers in the previous time step. Hence, when the population size is estimated to be below (above) the manager's target, the manager increases (decreases) the cost of culling and decreases (increases) the cost of scaring. Because the manager does not know in advance how stakeholders will react to policy change, they assume a proportional response in total actions with respect to a change in cost (e.g., doubling the cost of culling will decrease stakeholder culling by $1/2$). Farmers responding to policy are interested only in mitigating waterfowl's exploitation of their crops, so they will either cull or scare to remove the waterfowl from their land, depending on which option is more effective (i.e., cheaper). This is reflected in the bottom left versus right panels of Figure 2; when managers decrease culling costs relative to scaring, farmers respond with more total culling, and vice versa. Farmer decisions then affect waterfowl distribution and abundance, impacting future crop yield and policy.

Utility of GMSE simulations
===================================================

GMSE incorporates feedback between resource dynamcs, management policy, and stakeholder actions over time given multiple sources of uncertainty. By simulating complex systems of adaptive management in GMSE, it is possible to gain new insights into questions relevant to conservation. One obvious application of our software is to parameterise simulations with initial conditions similar to those from empirical populations of conservation interest. Running replicate simulations using empirically derived starting conditions could facilitate the prediction of key social-ecological outcomes (e.g., resource extinction, agricultural yield) given uncertainty, and the sensitivity of such outcomes to different management options (e.g., population target, policies available, observation methods, etc.), realised through results from replicate simulations, could inform management decisions.

Additionally, GMSE might be useful for developing adaptive management theory by allowing researchers to explore general questions concerning management *in silico* through simulation. Such questions might include the following: How is population persistence affected by management frequency or observation intensity? How does variation in stakeholder actions affect the distribution of resources or landscape properties? How do asymmetries in manager and stakeholder influence (i.e., budgets) affect resource dynamics? All of these questions can be explored through simulation using GMSE.

Future development
===================================================

The GMSE package is under further development to include additional features that will be of interest to conservation biologists, managers, and the general public. Such features will include multiple (interacting) resource types and sub-types (e.g., structured populations), improved manager and stakeholder decision-making and prediction based on multiple time steps of simulation history, incorporation of empirical data (e.g., landscape features), and a browser-based graphical user interface. Additionally, future versions of GMSE will allow software users to take the place of managers or stakeholders in decision-making during simulations (replacing the genetic algorithm as desired), facilitating the collection of data to test sociological hypotheses and further improve the realism of the genetic algorithm (for a preliminary example of this, see the 'hunt' argument in the gmse() function). Code underlying GMSE is publicly available on GitHub < https://github.com/bradduthie/gmse > and highly modular, meaning that GMSE model components (black boxes in Figure 1) can be developed freely and independently, then integrated into the broader GMSE framework. The GMSE R package is therefore a versatile and collaborative tool with widespread applications for adaptive resource management.

References
===================================================

### 2017-07-17 notebook/ms/ms.Rmd
New edits to the vignette and manuscript draft

@@
 output:
+  word_document:
+    fig_caption: yes
+    reference_docx: docx_template.docx
   pdf_document:
@@
   html_document: default
-  word_document:
-    fig_caption: yes
-    reference_docx: docx_template.docx
 linestretch: 1
@@
          xlim = c(0, time_max), xlab = "Time Step", 
-         ylab = "Stake-holder % yield", cex.lab = 1.25);
+         ylab = "stakeholder % yield", cex.lab = 1.25);
     stake_colors <- topo.colors( dim(age_t)[1] );
@@
 4.
-We show how GMSE can be used to simulate a social-ecological system, providing the example of management of a waterfowl population that exploits agricultural land, whose conservation is there in conflict with stakeholders.
+We show how GMSE simulates a social-ecological system using the example of a waterfowl population that is adaptively managed; simulated waterfowl exploit agricultural land, causing conflict between conservation interests and the interest of stakeholders (farmers) in maximising crop yield.
 5. The R package GMSE is open source under GNU Public License; source code and documents are freely available on GitHub.
@@
 Many global natural resources, including the biodiversity on which critical ecosystem services depend, are in a state of severe decline [@Dirzo2014; @Hautier2015; @Ceballos2017; @OConnell2017].
-Conservation of biodiversity can be complicated by the immediate need to allocate natural resources and land area for human food production, causing real or perceived conflicts between conservation and food security and creating a challenge for the management of many natural resources [@Redpath2015a].
+Conservation of biodiversity can be complicated by the immediate need to use natural resources and land area for human food production, causing real or perceived conflicts between conservation and food security and creating a challenge for the management of many natural resources [@Redpath2015a].
 Given an increasing human population [@Crist2017], the number and intensity of such conflicts are likely to increase into the twenty first century.
@@
 To effectively manage natural resources, an adaptive approach allows managers to iteratively update their models of resource dynamics and respond flexibly to changing conditions.
-This approach can be further improved by considering all aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of managers and stakeholders [@Bunnefeld2011; @Bunnefeld2014].
+This approach is especially effective when considering all aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of managers and stakeholders [@Bunnefeld2011; @Bunnefeld2014].
 Management strategy evaluation (MSE) is a modelling framework for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016].
@@
 The GMSE R package is a flexible, highly mechanistic, agent-based modelling tool to simulate all aspects of natural resource management.
-GMSE considers a range of parameters to simulate resource dynamics and management policy options, and uses genetic algorithms to dynamically model manager and stake-holder decision-making.
-Genetic algorithms allow managers and stakeholders to find adaptive solutions to any simulated conditions given unique individual goals, allowing GMSE to model scenarios of conservation conflict.
+GMSE considers a range of parameters to simulate resource dynamics and management policy options, and uses genetic algorithms to dynamically model manager and stakeholder decision-making.
+Genetic algorithms allow managers and stakeholders to find adaptive solutions to any simulated conditions given unique agent goals, allowing GMSE to model scenarios of conservation conflict.
 
@@
 
-Game theory is the standard tool for understanding conflict bewteen rational agents, and is therefore the appropriate tool for modelling manager and stakeholder actions and addressing issues of cooperation and conflict in conservation [@Colyvan2011; @Lee2012; @Kark2015; @Adami2016; @Tilman2016].
+Game theory is the standard tool for understanding conflict between rational agents, and is therefore the appropriate tool for modelling manager and stakeholder actions and addressing issues of cooperation and conflict in conservation [@Colyvan2011; @Lee2012; @Kark2015; @Adami2016; @Tilman2016].
 A game-theoretic approach is compatible with agent-based modelling [@An2012; @Tesfatsion2017], but given the number of possible strategies available to agents, identifying optimal strategies is often intractable.
-Instead, genetic algorithms, which mimic the process of natural selection (mutation, recombination, selection, replacement), are useful to find adaptive solutions for game strategies [e.g., @Balmann2000; @Tu2000; @Hamblin2013].
+Instead, genetic algorithms, which mimic the process of natural selection (mutation, recombination, selection, reproduction), are useful to find adaptive solutions for game strategies [e.g., @Balmann2000; @Tu2000; @Hamblin2013].
 
@@
 A fitness function then evaluates each strategy in the population, and a tournament is used to select the next generation of strategies [@Hamblin2013].
-The genetic algorithm terminates when a minimum number of generations has passed and the increase in the fittest strategy between the current and previous generation is below some threshold.
+The genetic algorithm terminates when a minimum number of generations has passed and the increase in the fitness of the fittest strategy between the current and previous generation is below some threshold.
 The highest fitness strategy in the population then becomes the agent's strategy in the current time step.
@@
 Results are most easily interpreted visually, so a summary of simulation dynamics is plotted by default (the plot can also be called using the plot_gmse_results function).
-An example below shows how simulations are set and interprted.
+An example below shows how simulations are set and interpreted.
 
@@
 |  land_dim_1     |  100    | Width of the landscape (horizontal cells)                                                        |
-|  land_dim_1     |  100    | Height of the landscape (vertical cells)                                                         |
+|  land_dim_2     |  100    | Height of the landscape (vertical cells)                                                         |
 |  res_movement   |  20     | Distance (cells) a resource can move in any direct (for movement rules, see res_move_type)       |
-|  remove_pr      |  0      | Density-indepdent probability of resource mortality during a time step                           |
+|  remove_pr      |  0      | Density-independent probability of resource mortality during a time step                         |
 |  lambda         |  0.3    | Poisson rate parameter for resource offspring number produced during a time step                 |
@@
 |  res_death_K    |  600    | Carrying capacity applied to the number of resources removed during a time step                  |
-|  res_move_type  |  1      | Type of resource movement (default up to res_movement cells in any direction)                    |
+|  res_move_type  |  1      | Type of resource movement (default is up to res_movement cells in any direction)                 |
 |  observe_type   |  0      | Type of resource observation (default is density-based; i.e, counting a subset on the landscape) |
@@
 |  tend_crops     |  FALSE  | Stakeholders can increase landscape cell values                                                  |
-|  tend_crop_yld  |  FALSE  | Proportional increase per landscape cell from tend_crops action                                  |
+|  tend_crop_yld  |  0.2    | Proportional increase per landscape cell from tend_crops action                                  |
 |  kill_crops     |  FALSE  | Stakeholders can decrease landscape cell values to zero                                          |
@@
 |  land_ownership |  FALSE  | Stakeholders own land and increase utility indirectly from landscape instead of resource use     |
-|  manage_freq    |  1      | Frequency (in time steps) with which managers enact policy                                       |
+|  manage_freq    |  1      | Frequency (in time steps) with which managers revise and enact policy                            |
 |  public_land    |  0      | Proportion of land that is public (un-owned by stakeholders) if land_ownership = TRUE            |
@@
 
-An example of crop management
+An example of resource management
 ===================================================
 
-Here we consider the example of a managed natural resources whose abundance affects a group of stakeholders by temporarily decreasing the value of stakeholder land.
-This scenario could be interpreted in multiple ways; we consider a protected population of waterfowl that exploits agricultural land causing a conservation conflict with farmers [e.g., @Tuvendal2015; @Fox2017a; @Fox2016a].
+Here we consider the example of a managed natural resource whose abundance affects a group of stakeholders by temporarily decreasing the value of stakeholder land.
+This scenario could be interpreted in multiple ways; we consider a protected population of waterfowl that exploits agricultural land causing a conservation conflict with farmers [e.g., @Tuvendal2015; @Fox2016a; @Fox2017a].
 Managers in this example might want to keep the abundance of waterfowl at a target level, while farmers might want to minimise the damage inflicted on their crops.
-Using GMSE, we can simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy.
+Using GMSE, it is possible to simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy.
 Here we consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity (applied to mortality) is 2000.
 Waterfowl consume and destroy all crop yield upon arrival to a landscape cell.
-Managers estimate population size in each time step using mark-recapture techniques, and use their estimates to set costs of culling and scaring waterfowl (i.e., 'policy') for ten farmers.
+Managers estimate population size in each time step using mark-recapture techniques, and use their estimates to set costs of culling and scaring (non-lethal) waterfowl (i.e., 'policy') for ten farmers.
 Farmers attempt to reduce the negative impact of waterfowl on the cropland that they own, working within the constraints of culling and scaring costs and their budget for performing these actions.
@@
 The middle right panel shows total percent yield of landscape cells for each individual farmer, differentiated by colour, where line colours correspond to areas of the landscape in the upper right panel.
-The lower left panel shows the cost of stakeholder performing actions over time, as set by the manager.
+The lower left panel shows the cost of stakeholders performing actions over time, as set by the manager.
 The lower right panel shows the total number of actions attempted to be performed by all stakeholders over time (some actions might be unsuccessful if resources are unavailable on a stakeholder's land to cull or scare, so, e.g., culling actions might be larger than resources actually culled)."}
@@
 
-The GMSE package is under further development to include additional features that will be of interest to conservation biologists, managers, and the general public.
+The GMSE package is under continued development to include additional features that will be of interest to conservation biologists, managers, and the general public.
 Such features will include multiple (interacting) resource types and sub-types (e.g., structured populations), improved manager and stakeholder decision-making and prediction based on multiple time steps of simulation history, incorporation of empirical data (e.g., landscape features), and a browser-based graphical user interface.

### 2017-07-17 notebook/ms/ms.Rmd
Minor change

@@
 output:
+  html_document: default
+  pdf_document:
+    fig_caption: yes
   word_document:
@@
     reference_docx: docx_template.docx
-  pdf_document:
-    fig_caption: yes
-  html_document: default
 linestretch: 1

### 2017-08-07 notebook/ms/ms.Rmd
Add generation indicator to Figure 1

@@
 
-```{r, echo = FALSE, fig.height = 4, fig.width = 4, fig.align = 'left', fig.cap = "Description of the generalised management strategy evaluation framework"}
+```{r, echo = FALSE, fig.height = 4, fig.width = 4, fig.align = 'left', fig.cap = "Description of one time step of the generalised management strategy evaluation framework"}
 mbox <- function(x0, x1, y0, y1){
@@
 polygon(x = game_model_box$x, y = game_model_box$y, lwd = 3);
-text(x=50, y=55, labels="Genetic", cex=1.0);
-text(x=50, y=48, labels="algorithm", cex=1.0);
+text(x=50, y=58, labels="Genetic", cex=1.0);
+text(x=50, y=51, labels="algorithm", cex=1.0);
+sc1 <- seq(from = 0, to = pi, by = 0.001);
+sc2 <- seq(from = pi, to = 2*pi, by = 0.001);
+points(x = 2*cos(sc1) + 41, y = 42 + 2*sin(sc1), lwd = 1.5, type = "l");
+points(x = 2*cos(sc2) + 41, y = 40 + 2*sin(sc2), lwd = 1.5, type = "l");
+arrows(x0 = 43, x1 = 43, y0 = 42, y1 = 41, lwd = 1.5, length = 0.06);
+arrows(x0 = 39, x1 = 39, y0 = 40, y1 = 41, lwd = 1.5, length = 0.06);
+text(x=43, y=42, labels="Multiple", cex=0.7, pos = 4);
+text(x=43, y=39, labels="Generations", cex=0.7, pos = 4);
 # Arrows indicating direction of GMSE model

### 2017-08-07 notebook/ms/ms.Rmd
Revisions, especially to the Introduction

--- 2017-07-17 notebook/ms/ms.Rmd
@@
 
-1. Management strategy evaluation (MSE) is a powerful tool for simulating all aspects of natural resource mangement under conditions of uncertainty.
-2. We present the R package GMSE, which generalises MSE using a game-theoretic approach to simulate adaptive management scenarios under complex social-ecological interactions and uncertainty.
-3. GMSE is agent-based and spatially explicit, and incorporates a high degree of realism and uncertainty through mechanistic modelling of the social-ecological system.
+1. Management strategy evaluation (MSE) is a powerful tool for simulating all key aspects of natural resource management under conditions of uncertainty.
+2.
+We present the R package GMSE, which generalises MSE using a game-theoretic approach to simulate adaptive decision-making management scenarios between stakeholders with competing objectives under complex social-ecological interactions and uncertainty.
+3. GMSE is agent-based and spatially explicit, and incorporates a high degree of realism through mechanistic modelling of links and feedbacks among stakeholders and with the ecosystem.
 4.
-We show how GMSE simulates a social-ecological system using the example of a waterfowl population that is adaptively managed; simulated waterfowl exploit agricultural land, causing conflict between conservation interests and the interest of stakeholders (farmers) in maximising crop yield.
+We show how GMSE simulates a social-ecological system using the example of a waterfowl population in an agricultural landscape that is adaptively managed; simulated waterfowl exploit agricultural land, causing conflict between conservation interests and the interest of food producers maximising their crop yield.
 5. The R package GMSE is open source under GNU Public License; source code and documents are freely available on GitHub.
@@
 Given an increasing human population [@Crist2017], the number and intensity of such conflicts are likely to increase into the twenty first century.
-Effective management tools are therefore needed for the long-term maintenance of natural resources under the rising demand for food production.
+Effective management tools are therefore needed for the long-term maintenance of natural resources under the rising demand for food production [@Fischer2017].
 
-To effectively manage natural resources, an adaptive approach allows managers to iteratively update their models of resource dynamics and respond flexibly to changing conditions.
-This approach is especially effective when considering all aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of managers and stakeholders [@Bunnefeld2011; @Bunnefeld2014].
-Management strategy evaluation (MSE) is a modelling framework for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016].
-Nevertheless, MSE remains limited in its ability to model human decision-making; manager decisions are typically based on simple rules, and stakeholder behaviour remains fixed over time instead of dynamically responding to changing resource availability and management decisions.
+To effectively manage natural resources, an adaptive approach allows managers to iteratively update their models of resource dynamics and respond flexibly to changing conditions [@Keith2011].
+This approach is especially effective when considering all aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of stakeholders [e.g., managers, users; @Bunnefeld2011; @Bunnefeld2014].
+Management strategy evaluation (MSE) is a modelling framework, first developed in fisheries, for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016].
+Nevertheless, MSE remains limited in its ability to model human decision-making [@Fulton2011a; @Dichmont2017]; manager decisions are typically based on simple rules, and stakeholder behaviour remains fixed over time instead of dynamically responding to changing resource availability and management decisions [@Schluter2012; @Melbourne-Thomas2017].
 Here we introduce a generalised management strategy evaluation (GMSE), which incorporates a game-theoretic perspective to model the goal-oriented, dynamic decision-making processes of managers and stakeholders.
 
-The GMSE R package is a flexible, highly mechanistic, agent-based modelling tool to simulate all aspects of natural resource management.
+The GMSE R package is a flexible, highly mechanistic, agent-based modelling tool to simulate all key aspects of natural resource management.
 GMSE considers a range of parameters to simulate resource dynamics and management policy options, and uses genetic algorithms to dynamically model manager and stakeholder decision-making.
@@
 text(x=10, y=48, labels="Indicators", srt=90, cex = 1.0);
-text(x=90, y=52, labels="Off-take", srt=-90, cex = 1.0);
+text(x=90, y=52, labels="Actions", srt=-90, cex = 1.0);
 ```
@@
 ```{r, echo = TRUE}
-sim <- gmse(land_ownership = TRUE, stakeholders = 10, observe_type = 1, 
+sim <- gmse(land_ownership = TRUE, stakeholders = 10, observe_type = 0, 
             res_death_K = 2000, manage_target = 1000, RESOURCE_ini = 1000, 
             user_budget = 5000, manager_budget = 5000, res_consume = 1, 
-            scaring = TRUE, fixed_mark = 50, fixed_recapt = 300, 
+            scaring = TRUE, times_observe = 1, agent_view = 10, 
             plotting = FALSE);

### 2017-08-07 notebook/ms/ms.Rmd
Update GMSE model structure section

@@
 To effectively manage natural resources, an adaptive approach allows managers to iteratively update their models of resource dynamics and respond flexibly to changing conditions [@Keith2011].
-This approach is especially effective when considering all aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of stakeholders [e.g., managers, users; @Bunnefeld2011; @Bunnefeld2014].
+This approach is especially effective when considering all aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of stakeholders [e.g., managers and users; @Bunnefeld2011; @Bunnefeld2014].
 Management strategy evaluation (MSE) is a modelling framework, first developed in fisheries, for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016].
-Nevertheless, MSE remains limited in its ability to model human decision-making [@Fulton2011a; @Dichmont2017]; manager decisions are typically based on simple rules, and stakeholder behaviour remains fixed over time instead of dynamically responding to changing resource availability and management decisions [@Schluter2012; @Melbourne-Thomas2017].
-Here we introduce a generalised management strategy evaluation (GMSE), which incorporates a game-theoretic perspective to model the goal-oriented, dynamic decision-making processes of managers and stakeholders.
+Nevertheless, MSE remains limited in its ability to model human decision-making [@Fulton2011a; @Dichmont2017]; manager decisions are typically based on simple rules, and user behaviour remains fixed over time instead of dynamically responding to changing resource availability and management decisions [@Schluter2012; @Melbourne-Thomas2017].
+Here we introduce a generalised management strategy evaluation (GMSE), which incorporates a game-theoretic perspective to model the goal-oriented, dynamic decision-making processes of stakeholders.
 
 The GMSE R package is a flexible, highly mechanistic, agent-based modelling tool to simulate all key aspects of natural resource management.
-GMSE considers a range of parameters to simulate resource dynamics and management policy options, and uses genetic algorithms to dynamically model manager and stakeholder decision-making.
-Genetic algorithms allow managers and stakeholders to find adaptive solutions to any simulated conditions given unique agent goals, allowing GMSE to model scenarios of conservation conflict.
+GMSE considers a range of parameters to simulate resource dynamics and management policy options, and uses genetic algorithms to dynamically model stakeholder (manager and user) decision-making.
+Genetic algorithms allow stakeholders to find adaptive solutions to any simulated conditions given unique agent goals, allowing GMSE to model scenarios of conservation conflict.
 
@@
 
-GMSE builds off of the traditional MSE framework, which includes four sub-models, each of which runs once in a single time step of the broader model (Figure 1).
-(1) A natural resources sub-model considers a population of discrete resources with individual traits (e.g., location, age) on a spatially-explicit landscape.
-This sub-model can include processes of resource birth, movement, interaction with the landscape, and death.
-The discrete nature of resource demographics naturally gives rise to stochasticity, and therefore uncertainty.
-Although the resource sub-model can interact with other sub-models, it is unique in not relying on them because ecological dynamics can be simulated in the absence of observation and management.
-(2) An observation sub-model simulates the process of data collection peformed by a manager to estimate resource abundance.
-Four types of data collection are permitted, including exhaustive resource counting on a subset of landscape cells [e.g., @Nuno2013], marking and recapturing a fixed number of resources, and exhaustive sampling of the whole landscape iteratively (during which resources might move).
-Imperfect sampling from all of these mechanisms of data collections generates observation uncertainty.
-(3) A manager sub-model simulates a manager's analysis of collected data to estimate resource abundance, then compares this estimated abundance with the manager's target abundance.
-Policy is developed by calling the genetic algorithm (see below), which works within a manager's constraints to find costs for stakeholder actions (e.g., culling, scaring, etc.) that minimise resource deviation from the target abundance, as informed by the predicted consequences of stakeholder actions and stakeholder action histories.
-After an adaptive policy is found, (4) a user sub-model is run that allows each stakeholder to perform actions that affect resources or landscape cells.
-Stakeholders respond to policy individually, each calling the genetic algorithm to find actions that maximise their own utilities (e.g., maximise resource use or landscape yield) within their imposed constraints.
-Once each stakeholder has found an adaptive strategy, stakeholder actions affect resources and landscape cells, feeding back into the resource sub-model.
+GMSE builds off of the MSE framework, which includes four sub-models, each of which runs once in a single time step of the broader model (Figure 1).
+(1) A population of discrete resources with individual traits (e.g., location, age) is modelled on a spatially-explicit landscape and can simulate resource birth, movement, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty.
+This sub-model is unique in not relying on other sub-models because ecological dynamics can be simulated in the absence of observation and management.
+(2) Observation is modelled in one of four ways: resource counting on a subset of landscape cells [e.g., @Nuno2013], marking and recapturing a fixed number of resources, and resource counting on the whole landscape either one linear transect or one rectangular block at a time (during which resources might move).
+Sampling error from all of these mechanisms of observation generates a range of uncertainties depending on monitoring effort.
+(3) Managers analyse data collected from observations to estimate resource abundance, then compare this estimate with their target abundance.
+Policy is developed by calling the genetic algorithm (see below), which works within a manager's constraints to find costs for user actions on the resource (e.g., culling, scaring, etc.) that minimise deviation from the target abundance, as informed by the predicted consequences of each action on resource abundance and user action histories.
+After a suitable policy is found, (4) users can perform actions that affect resources or landscape cells.
+Users respond to policy individually, each calling the genetic algorithm to find actions that maximise their own utilities (e.g., maximise resource use or landscape yield) within their imposed constraints.
+Once each user has found an adaptive strategy, user actions affect resources and landscape cells, feeding back into the resource sub-model.
 
-```{r, echo = FALSE, fig.height = 4, fig.width = 4, fig.align = 'left', fig.cap = "Description of one time step of the generalised management strategy evaluation framework"}
+```{r, echo = FALSE, fig.height = 4, fig.width = 4, fig.align = 'left', fig.cap = "Description of one time step of the generalised management strategy evaluation framework, which is comprised of four separate sub-models."}
 mbox <- function(x0, x1, y0, y1){
@@
 text(x = 15, y = 83, cex = 1.0, labels = "model");
+text(x = 5, y = 73, labels="(3)");
 # User model
@@
 text(x = 85, y = 83, cex = 1.0, labels = "model");
+text(x = 74, y = 73, labels="(4)");
 # Observation model
@@
 text(x = 15, y = 15, cex = 1.0, labels = "model");
+text(x = 5, y = 4, labels="(2)");
 # Resource model
@@
 text(x = 85, y = 7,  cex = 1.0, labels = "model");
+text(x = 74, y = 4, labels="(1)");
 # Game-theoretic (genetic algorithm) model

### 2017-08-07 notebook/ms/ms.Rmd
Change 'stakeholder' to a general term?

@@
 
-Game theory is the standard tool for understanding conflict between rational agents, and is therefore the appropriate tool for modelling manager and stakeholder actions and addressing issues of cooperation and conflict in conservation [@Colyvan2011; @Lee2012; @Kark2015; @Adami2016; @Tilman2016].
+Game theory is the standard tool for understanding conflict between rational agents, and is therefore the appropriate tool for modelling stakeholder actions and addressing issues of cooperation and conflict in conservation [@Colyvan2011; @Lee2012; @Kark2015; @Adami2016; @Tilman2016].
 A game-theoretic approach is compatible with agent-based modelling [@An2012; @Tesfatsion2017], but given the number of possible strategies available to agents, identifying optimal strategies is often intractable.
@@
 
-In keeping with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for managers or stakeholders.
-Instead, within each time step of a simulation (Figure 1), a new genetic algorithm is called for each agent to heuristically find strategies that are well-adapted to achieving agent goals.
-Hence for the manager and each stakeholder, a unique population with a random set of possible values affecting policy or actions, respectively, is initialised.
+In keeping with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for stakeholders.
+Instead, within each time step of a simulation (Figure 1), a new genetic algorithm is called for each stakeholder to heuristically find strategies that are well-adapted to achieving their goals.
+Hence for the manager and each user, a unique population with a random set of possible values affecting policy or actions, respectively, is initialised.
 In each generation of the genetic algorithm, strategies crossover and mutate; when this results in strategies that are over-budget, values are iteratively decreased at random until budget constraints are satisfied.
@@
 The genetic algorithm terminates when a minimum number of generations has passed and the increase in the fitness of the fittest strategy between the current and previous generation is below some threshold.
-The highest fitness strategy in the population then becomes the agent's strategy in the current time step.
+The highest fitness strategy in the population then becomes the stakeholder's strategy in the current time step.
 
@@
 Simulations are run using the gmse() function, which offers a range of options for setting parameter values (see Table 1 for some select examples).
-Output of gmse() is an exhaustive list that includes all resources and observations, all manager and stakeholder decisions and actions, and all landscape properties in each time step of the simulation.
+Output of gmse() is an exhaustive list that includes all resources and observations, all stakeholder decisions and actions, and all landscape properties in each time step of the simulation.
 Results are most easily interpreted visually, so a summary of simulation dynamics is plotted by default (the plot can also be called using the plot_gmse_results function).
@@
 
-Here we consider the example of a managed natural resource whose abundance affects a group of stakeholders by temporarily decreasing the value of stakeholder land.
-This scenario could be interpreted in multiple ways; we consider a protected population of waterfowl that exploits agricultural land causing a conservation conflict with farmers [e.g., @Tuvendal2015; @Fox2016a; @Fox2017a].
+Here we consider the example of a managed natural resource whose abundance affects a group of stakeholders by temporarily decreasing the value of user land.
+This scenario could be interpreted in multiple ways; we consider a protected population of waterfowl that exploits agricultural land causing a conservation conflict with users [hereafter 'farmers', e.g., @Tuvendal2015; @Fox2016a; @Fox2017a].
 Managers in this example might want to keep the abundance of waterfowl at a target level, while farmers might want to minimise the damage inflicted on their crops.
@@
 
-Figure 2 shows the landscape broken down by resource position and stakeholder ownership in the upper left and right hand panels, respectively.
-The waterfowl population fluctuates around the manager's target size of 1000, but the manager's estimate of population size deviates from its actual size due to uncertainty caused by the observation process (compare black and blue lines the middle left panel).
+Figure 2 shows the landscape broken down by resource position and farmer land ownership in the upper left and right hand panels, respectively.
+The waterfowl population fluctuates around the manager's target size of 1000, but the manager's estimate of population size deviates from its actual size due to observation uncertainty (compare black and blue lines the middle left panel).
 Because the waterfowl have a direct negative effect on landscape yield, total landscape yield (orange line of the middle left panel), along with the yield of individual farmers (right middle panel), is low when waterfowl abundance is high, and vice versa.
@@
 
-GMSE incorporates feedback between resource dynamcs, management policy, and stakeholder actions over time given multiple sources of uncertainty.
+GMSE incorporates feedback between resource dynamcs, management policy, and user actions over time given multiple sources of uncertainty.
 By simulating complex systems of adaptive management in GMSE, it is possible to gain new insights into questions relevant to conservation.
@@
 Such questions might include the following: How is population persistence affected by management frequency or observation intensity?
-How does variation in stakeholder actions affect the distribution of resources or landscape properties?
-How do asymmetries in manager and stakeholder influence (i.e., budgets) affect resource dynamics?
+How does variation in user actions affect the distribution of resources or landscape properties?
+How do asymmetries in stakeholder influence (i.e., budgets) affect resource dynamics?
 All of these questions can be explored through simulation using GMSE.
@@
 The GMSE package is under continued development to include additional features that will be of interest to conservation biologists, managers, and the general public.
-Such features will include multiple (interacting) resource types and sub-types (e.g., structured populations), improved manager and stakeholder decision-making and prediction based on multiple time steps of simulation history, incorporation of empirical data (e.g., landscape features), and a browser-based graphical user interface.
-Additionally, future versions of GMSE will allow software users to take the place of managers or stakeholders in decision-making during simulations (replacing the genetic algorithm as desired), facilitating the collection of data to test sociological hypotheses and further improve the realism of the genetic algorithm (for a preliminary example of this, see the 'hunt' argument in the gmse() function).
+Such features will include multiple (interacting) resource types and sub-types (e.g., structured populations), improved stakeholder decision-making and prediction based on multiple time steps of simulation history, incorporation of empirical data (e.g., landscape features), and a browser-based graphical user interface.
+Additionally, future versions of GMSE will allow software users to take the place of stakeholders in decision-making during simulations (replacing the genetic algorithm as desired), facilitating the collection of data to test sociological hypotheses and further improve the realism of the genetic algorithm (for a preliminary example of this, see the 'hunt' argument in the gmse() function).
 Code underlying GMSE is publicly available on GitHub < https://github.com/bradduthie/gmse > and highly modular, meaning that GMSE model components (black boxes in Figure 1) can be developed freely and independently, then integrated into the broader GMSE framework.

### 2017-08-08 notebook/ms/ms.Rmd
Major restructuring of Utility of GMSE section moved to the Introduction

@@
 Genetic algorithms allow stakeholders to find adaptive solutions to any simulated conditions given unique agent goals, allowing GMSE to model scenarios of conservation conflict.
+Hence, simulations can be parameterised with initial conditions similar to those from empirical populations of conservation interest to predict key social-ecological outcomes (e.g., resource extinction, agricultural yield) given uncertainty.
+The sensitivity of these outcomes to different management options (e.g., population target, policies available, observation methods, budget constraints, etc.) can thereby inform decisions underpinning competing management objectives.
+Additionally, GMSE allows researchers to explore general questions concerning management *in silico* through simulation.
+Such questions might include the following: How is population persistence affected by management frequency or observation intensity?
+How does variation in user actions affect the distribution of resources or landscape properties?
+How do asymmetries in stakeholder influence (i.e., budgets) affect resource dynamics?
 
@@
 This scenario could be interpreted in multiple ways; we consider a protected population of waterfowl that exploits agricultural land causing a conservation conflict with users [hereafter 'farmers', e.g., @Tuvendal2015; @Fox2016a; @Fox2017a].
-Managers in this example might want to keep the abundance of waterfowl at a target level, while farmers might want to minimise the damage inflicted on their crops.
+Managers in this example might want to keep the abundance of waterfowl at a target level, while farmers might want to minimise the damage inflicted on their crops [e.g., @Madsen2017].
 Using GMSE, it is possible to simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy.
-Here we consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity (applied to mortality) is 2000.
+Here we consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity is 2000.
 Waterfowl consume and destroy all crop yield upon arrival to a landscape cell.
-Managers estimate population size in each time step using mark-recapture techniques, and use their estimates to set costs of culling and scaring (non-lethal) waterfowl (i.e., 'policy') for ten farmers.
+In each time step, waterfowl are observed on subset of cells, then managers extrapolate from density per cell to estimate total population size.
+Managers then use these estimates to set costs of culling and scaring (non-lethal) waterfowl (i.e., 'policy') for 10 farmers.
 Farmers attempt to reduce the negative impact of waterfowl on the cropland that they own, working within the constraints of culling and scaring costs and their budget for performing these actions.
@@
 Because the manager does not know in advance how stakeholders will react to policy change, they assume a proportional response in total actions with respect to a change in cost (e.g., doubling the cost of culling will decrease stakeholder culling by $1/2$).
-Farmers responding to policy are interested only in mitigating waterfowl's exploitation of their crops, so they will either cull or scare to remove the waterfowl from their land, depending on which option is more effective (i.e., cheaper).
+Farmers responding to policy are interested only in minimising waterfowl's exploitation of their crops, so they will either cull or scare to remove the waterfowl from their land, depending on which option is more effective (i.e., cheaper).
 This is reflected in the bottom left versus right panels of Figure 2; when managers decrease culling costs relative to scaring, farmers respond with more total culling, and vice versa.
 Farmer decisions then affect waterfowl distribution and abundance, impacting future crop yield and policy.
-
-Utility of GMSE simulations
-===================================================
-
-GMSE incorporates feedback between resource dynamcs, management policy, and user actions over time given multiple sources of uncertainty.
-By simulating complex systems of adaptive management in GMSE, it is possible to gain new insights into questions relevant to conservation.
-One obvious application of our software is to parameterise simulations with initial conditions similar to those from empirical populations of conservation interest.
-Running replicate simulations using empirically derived starting conditions could facilitate the prediction of key social-ecological outcomes (e.g., resource extinction, agricultural yield) given uncertainty, and the sensitivity of such outcomes to different management options (e.g., population target, policies available, observation methods, etc.), realised through results from replicate simulations, could inform management decisions.
-
-Additionally, GMSE might be useful for developing adaptive management theory by allowing researchers to explore general questions concerning management *in silico* through simulation.
-Such questions might include the following: How is population persistence affected by management frequency or observation intensity?
-How does variation in user actions affect the distribution of resources or landscape properties?
-How do asymmetries in stakeholder influence (i.e., budgets) affect resource dynamics?
-All of these questions can be explored through simulation using GMSE.
 

### 2017-08-08 notebook/ms/ms.Rmd
Further minor edits

@@
 Many global natural resources, including the biodiversity on which critical ecosystem services depend, are in a state of severe decline [@Dirzo2014; @Hautier2015; @Ceballos2017; @OConnell2017].
-Conservation of biodiversity can be complicated by the immediate need to use natural resources and land area for human food production, causing real or perceived conflicts between conservation and food security and creating a challenge for the management of many natural resources [@Redpath2015a].
+Conservation of biodiversity can be complicated by the immediate need to use natural resources and land area for human livelihood (e.g., food production), causing real or perceived conflicts between conservation and food security and creating a challenge for the management of many natural resources [@Redpath2015a].
 Given an increasing human population [@Crist2017], the number and intensity of such conflicts are likely to increase into the twenty first century.
@@
 To effectively manage natural resources, an adaptive approach allows managers to iteratively update their models of resource dynamics and respond flexibly to changing conditions [@Keith2011].
-This approach is especially effective when considering all aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of stakeholders [e.g., managers and users; @Bunnefeld2011; @Bunnefeld2014].
+This approach is especially effective when considering all aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of stakeholders [@Bunnefeld2011; @Bunnefeld2014].
 Management strategy evaluation (MSE) is a modelling framework, first developed in fisheries, for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016].
 Nevertheless, MSE remains limited in its ability to model human decision-making [@Fulton2011a; @Dichmont2017]; manager decisions are typically based on simple rules, and user behaviour remains fixed over time instead of dynamically responding to changing resource availability and management decisions [@Schluter2012; @Melbourne-Thomas2017].
-Here we introduce a generalised management strategy evaluation (GMSE), which incorporates a game-theoretic perspective to model the goal-oriented, dynamic decision-making processes of stakeholders.
+Here we introduce generalised management strategy evaluation (GMSE), which incorporates a game-theoretic perspective to model the goal-oriented, dynamic decision-making processes of stakeholders.
 
@@
 GMSE considers a range of parameters to simulate resource dynamics and management policy options, and uses genetic algorithms to dynamically model stakeholder (manager and user) decision-making.
-Genetic algorithms allow stakeholders to find adaptive solutions to any simulated conditions given unique agent goals, allowing GMSE to model scenarios of conservation conflict.
-Hence, simulations can be parameterised with initial conditions similar to those from empirical populations of conservation interest to predict key social-ecological outcomes (e.g., resource extinction, agricultural yield) given uncertainty.
-The sensitivity of these outcomes to different management options (e.g., population target, policies available, observation methods, budget constraints, etc.) can thereby inform decisions underpinning competing management objectives.
-Additionally, GMSE allows researchers to explore general questions concerning management *in silico* through simulation.
-Such questions might include the following: How is population persistence affected by management frequency or observation intensity?
+Genetic algorithms find adaptive solutions to any simulated conditions given stakeholder-specific goals, allowing GMSE to model scenarios of conservation conflict.
+
+GMSE allows researchers to address adaptive management questions *in silico* through simulation.
+Simulations can be parameterised with initial conditions derived from empirical populations of conservation interest to predict key social-ecological outcomes (e.g., resource extinction, agricultural yield) given uncertainty.
+The sensitivity of these outcomes to different management options (e.g., population target, policies available, observation methods, budget constraints, etc.) can thereby inform management decisions, even given competing management objectives caused by conservation conflict [e.g., @Strand2012; @Redpath2013; @Sundt-Hansen2015; @Pozo2017; @Fox2017a].
+Additionally, GMSE can be used to explore general questions concerning management theory such as the following: How is population persistence affected by management frequency or observation intensity?
 How does variation in user actions affect the distribution of resources or landscape properties?
@@
 Using GMSE, it is possible to simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy.
-Here we consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity is 2000.
+We consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity is 2000.
 Waterfowl consume and destroy all crop yield upon arrival to a landscape cell.
@@
 Additionally, future versions of GMSE will allow software users to take the place of stakeholders in decision-making during simulations (replacing the genetic algorithm as desired), facilitating the collection of data to test sociological hypotheses and further improve the realism of the genetic algorithm (for a preliminary example of this, see the 'hunt' argument in the gmse() function).
-Code underlying GMSE is publicly available on GitHub < https://github.com/bradduthie/gmse > and highly modular, meaning that GMSE model components (black boxes in Figure 1) can be developed freely and independently, then integrated into the broader GMSE framework.
+Code underlying GMSE is publicly available on GitHub < https://github.com/bradduthie/gmse > and highly modular, so GMSE model components (black boxes in Figure 1) can be developed freely and independently, then integrated into the broader GMSE framework.
 The GMSE R package is therefore a versatile and collaborative tool with widespread applications for adaptive resource management.

### 2017-08-09 notebook/ms/ms.Rmd
Some progress on the genetic algorithm section

@@
 
-Game theory is the standard tool for understanding conflict between rational agents, and is therefore the appropriate tool for modelling stakeholder actions and addressing issues of cooperation and conflict in conservation [@Colyvan2011; @Lee2012; @Kark2015; @Adami2016; @Tilman2016].
+Game theory is the formal study of strategic interactions, and can therefore be applied to modelling stakeholder actions and addressing issues of cooperation and conflict in conservation [@Colyvan2011; @Lee2012; @Kark2015; @Adami2016; @Tilman2016].
+In game-theoretic models, agents adopt strategies to make decisions that maximise some type of payoff (e.g., utility, biological fitness).
+Agents are constrained in their decision-making, and realised pay-offs depend on decisions made by other agents.
+In simple models, it is often useful to assume that agents are perfectly rational decision-makers, then find optimal solutions for pay-off maximisation mathematically.
+But models that permit even moderately complex decision-making strategies or pay-off structures often include more possible strategies than are mathematically tractable [@Hamblin2013].
+In these models, genetic algorithms, which mimic the process of natural selection (mutation, recombination, selection, reproduction), can find adaptive (i.e., practical, but not necessarily optimal) solutions for game strategies [e.g., @Balmann2000; @Tu2000; @Hamblin2013].
+
+Consistent with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for agents (stakeholders).
+Instead, genetic algorithms are used to find an adaptive strategy for each stakeholder in each time step.
+
 A game-theoretic approach is compatible with agent-based modelling [@An2012; @Tesfatsion2017], but given the number of possible strategies available to agents, identifying optimal strategies is often intractable.

### 2017-08-09 notebook/ms/ms.Rmd
Major restructuring of genetic algorithm section

@@
 Consistent with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for agents (stakeholders).
-Instead, genetic algorithms are used to find an adaptive strategy for each stakeholder in each time step.
+Instead, genetic algorithms are used to heuristically find an adaptive strategy for each stakeholder in each time step.
+Critically, all stakeholders involved in resource conservation are constrained in their decision-making; managing and using resources takes effort (e.g., time or money), and effort expended in developing or enforcing one policy (for managers) or performing one action (for users) will be effort not expendable elsewhere [@Milner-Gulland2011].
+In finding strategies, GMSE models this trade-off by setting a fixed budget for managers and users (Table 1).
+Allocations from a manager's budget can be used to increase the cost it takes a user to perform an action, and allocations from a user's budget can be used to perform the action at the cost set by the manager.
+Hence, stakeholders can have incomplete control over resource use and express competing management objectives.
 
-A game-theoretic approach is compatible with agent-based modelling [@An2012; @Tesfatsion2017], but given the number of possible strategies available to agents, identifying optimal strategies is often intractable.
-Instead, genetic algorithms, which mimic the process of natural selection (mutation, recombination, selection, reproduction), are useful to find adaptive solutions for game strategies [e.g., @Balmann2000; @Tu2000; @Hamblin2013].
-
-In keeping with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for stakeholders.
-Instead, within each time step of a simulation (Figure 1), a new genetic algorithm is called for each stakeholder to heuristically find strategies that are well-adapted to achieving their goals.
-Hence for the manager and each user, a unique population with a random set of possible values affecting policy or actions, respectively, is initialised.
-In each generation of the genetic algorithm, strategies crossover and mutate; when this results in strategies that are over-budget, values are iteratively decreased at random until budget constraints are satisfied.
+In each new call of the genetic algorithm, a unique population of managers or users with random strategies is temporarily initialised.
+In each generation of the genetic algorithm, these strategies crossover and mutate; when this results in strategies that are over-budget, expenditures are iteratively decreased at random until budget constraints are satisfied.
 A fitness function then evaluates each strategy in the population, and a tournament is used to select the next generation of strategies [@Hamblin2013].
 The genetic algorithm terminates when a minimum number of generations has passed and the increase in the fitness of the fittest strategy between the current and previous generation is below some threshold.
-The highest fitness strategy in the population then becomes the stakeholder's strategy in the current time step.
+The highest fitness strategy in the population then becomes the stakeholder's new strategy.
 

### 2017-08-09 notebook/ms/ms.Rmd
Fresh first revision with new genetic algorithm text

@@
 Instead, genetic algorithms are used to heuristically find an adaptive strategy for each stakeholder in each time step.
-Critically, all stakeholders involved in resource conservation are constrained in their decision-making; managing and using resources takes effort (e.g., time or money), and effort expended in developing or enforcing one policy (for managers) or performing one action (for users) will be effort not expendable elsewhere [@Milner-Gulland2011].
+Critically, all stakeholders involved in resource conservation are constrained in their decision-making; managing and using resources takes effort (e.g., time or money), and effort expended in developing or enforcing one policy (for managers) or performing one action (for users) will be effort not expendable elsewhere [@Milner-Gulland2011; @Muller-Hansen2017; @Schluter2017].
 In finding strategies, GMSE models this trade-off by setting a fixed budget for managers and users (Table 1).

### 2017-08-09 notebook/ms/ms.Rmd
Minor tweaks

@@
 Management strategy evaluation (MSE) is a modelling framework, first developed in fisheries, for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016].
-Nevertheless, MSE remains limited in its ability to model human decision-making [@Fulton2011a; @Dichmont2017]; manager decisions are typically based on simple rules, and user behaviour remains fixed over time instead of dynamically responding to changing resource availability and management decisions [@Schluter2012; @Melbourne-Thomas2017].
+Nevertheless, MSE remains limited in its ability to model human decision-making [@Fulton2011a; @Dichmont2017]; manager decisions are typically based on fixed rules, and user behaviour likewise remains fixed over time instead of dynamically responding to changing resource availability and management decisions [@Schluter2012; @Melbourne-Thomas2017].
 Here we introduce generalised management strategy evaluation (GMSE), which incorporates a game-theoretic perspective to model the goal-oriented, dynamic decision-making processes of stakeholders.
@@
 Critically, all stakeholders involved in resource conservation are constrained in their decision-making; managing and using resources takes effort (e.g., time or money), and effort expended in developing or enforcing one policy (for managers) or performing one action (for users) will be effort not expendable elsewhere [@Milner-Gulland2011; @Muller-Hansen2017; @Schluter2017].
-In finding strategies, GMSE models this trade-off by setting a fixed budget for managers and users (Table 1).
-Allocations from a manager's budget can be used to increase the cost it takes a user to perform an action, and allocations from a user's budget can be used to perform the action at the cost set by the manager.
+In finding strategies, GMSE models this trade-off by setting a fixed budget for managers and users.
+Allocations from a manager's budget can be used to increase the cost it takes a user to perform an action (i.e., 'policy'), and allocations from a user's budget can be used to perform the action at the cost set by the manager.
 Hence, stakeholders can have incomplete control over resource use and express competing management objectives.
@@
 In each time step, waterfowl are observed on subset of cells, then managers extrapolate from density per cell to estimate total population size.
-Managers then use these estimates to set costs of culling and scaring (non-lethal) waterfowl (i.e., 'policy') for 10 farmers.
+Managers then use these estimates to set costs of culling and scaring (non-lethal) waterfowl for 10 farmers.
 Farmers attempt to reduce the negative impact of waterfowl on the cropland that they own, working within the constraints of culling and scaring costs and their budget for performing these actions.
@@
             user_budget = 5000, manager_budget = 5000, res_consume = 1, 
-            scaring = TRUE, times_observe = 1, agent_view = 10, 
-            plotting = FALSE);
+            scaring = TRUE, times_observe = 1, plotting = FALSE);
 ```

### 2017-08-09 notebook/ms/ms.Rmd
Add coauthors

@@
 title: 'GMSE: an R package for generalised management strategy evaluation'
-author: A. Bradley Duthie, Jeremy J. Cusack, Roc&#0237;o Pozo, O. Sarobidy Rakotonarivo,
-  and Nils Bunnefeld
+author: A. Bradley Duthie, Jeremy J. Cusack, Erlend B. Nilsen, Roc&#0237;o Pozo, O.
+  Sarobidy Rakotonarivo, Bram Van Moorter, and Nils Bunnefeld
 date: '`r Sys.Date()`'

### 2017-08-09 notebook/ms/ms.Rmd
More minor tweaks

@@
 Given an increasing human population [@Crist2017], the number and intensity of such conflicts are likely to increase into the twenty first century.
-Effective management tools are therefore needed for the long-term maintenance of natural resources under the rising demand for food production [@Fischer2017].
+Effective management tools are therefore needed for the long-term sustainable use of natural resources under the rising demand for food production [@Fischer2017].
 
@@
 Sampling error from all of these mechanisms of observation generates a range of uncertainties depending on monitoring effort.
-(3) Managers analyse data collected from observations to estimate resource abundance, then compare this estimate with their target abundance.
+(3) Managers analyse data collected from observations to estimate resource abundance, then compare this estimate with their pre-defined target abundance.
 Policy is developed by calling the genetic algorithm (see below), which works within a manager's constraints to find costs for user actions on the resource (e.g., culling, scaring, etc.) that minimise deviation from the target abundance, as informed by the predicted consequences of each action on resource abundance and user action histories.

### 2017-08-09 notebook/ms/ms.Rmd
Final minor tweaks before sending to colleagues

@@
 output:
-  html_document: default
   pdf_document:
     fig_caption: yes
+  html_document: default
   word_document:

### 2017-08-15 notebook/gui_notes.R
Initial, messy playing with the GUI interface

@@ -1,0 +1,228 @@
+# Relevant functions from elementR that might help
+# These go *within* the main shiny function
+
+
+
+
+run_gmse <- function(){ # nocov start
+
+
+
+#skyn
+skin <- Sys.getenv("DASHBOARD_SKIN")
+skin <- tolower(skin)
+if (skin == "") skin <- "blue"
+
+menuIconClass <- "fa-lg" 
+# This makes the sidebar of options and settings (e.g., parameters)
+sidebar <-   dashboardSidebar(
+    useShinyjs(),
+    sidebarMenu(id = "tab",
+                div(style = "background: rgb(60, 141, 188); height: 50px",
+                    p(icon("star-half-o"),"element-R", 
+                      style = "font-size: 200%; padding-left:50px;padding-top:5px")),
+                
+                menuItem("Global parameters", tabName = "start", 
+                         icon = icon("globe", class=menuIconClass), 
+                         badgeLabel="Step 1", badgeColor="blue"),
+                
+                div(align="center"),
+                
+                menuItem("Resource model", tabName = "Standards", 
+                         icon = icon("thermometer-1", class=menuIconClass), 
+                         badgeLabel="Step 2", badgeColor="green"),
+                
+                uiOutput("renderProgress2"),
+                
+                div(align="center"),
+                
+                menuItem("Observation model", tabName = "MachDrift", 
+                         icon = icon("binoculars", class=menuIconClass), 
+                         badgeLabel="Step 3", badgeColor="purple"),
+                
+                div(align="center"),
+                
+                menuItem("Manager model", tabName = "Samples", 
+                         icon = icon("thermometer-3", class=menuIconClass), 
+                         badgeLabel="Step 4", badgeColor="maroon"),
+                
+                div(align="center"),
+                
+                menuItem("User model", tabName = "realign", 
+                         icon = icon("thermometer-4", class=menuIconClass), 
+                         badgeLabel="Step 5", badgeColor="orange"),
+                
+                div(align="center"),
+                
+                menuItem("Run simulation", tabName = "realign", 
+                         icon = icon("thermometer-4", class=menuIconClass), 
+                         badgeLabel="Go!", badgeColor="orange"),
+                
+                uiOutput("sim_output"),
+                
+                hr(style ="width: 70%; color: white; align: center"),
+                
+                menuItem("Plotting", icon = icon("line-chart"), tabName = "Config"),
+
[…]

### 2017-08-17 notebook/gui_notes.R
Make a proper function

@@ -1,10 +1,17 @@
-library(shiny)
-library(shinydashboard)
-library(shinyjs)
-library(GMSE)
-
-# Calling the function below will call the gmse function in a browser
-
+#' GMSE GUI function
+#' 
+#' The gmse_gui function will call a browser-based graphical user interface 
+#' (GUI) for the gmse function. The GUI will run simulations for a limited range
+#' of parameter values and present results as plots.
+#'
+#'@examples
+#'sim <- gmse_gui();
+#'@useDynLib GMSE
+#'@importFrom grDevices topo.colors
+#'@importFrom graphics abline axis image mtext par plot points polygon
+#'@importFrom stats rnorm rpois
+#'@importFrom shiny shinyjs shinydashboard
+#'@export
 gmse_gui <- function(){ 
 
 #-------------------------------------------------------------------------------

### 2017-08-17 R/gmse_gui.R
Fix up tests and description, along with information for the gmse app and notes

@@ -7,12 +7,19 @@
 #'@return A browser should immediately open with the gmse graphical user
 #'interface
 #'@examples
+#'\dontrun{
 #'sim <- gmse_gui();
+#'}
 #'@useDynLib GMSE
 #'@importFrom grDevices topo.colors
 #'@importFrom graphics abline axis image mtext par plot points polygon
 #'@importFrom stats rnorm rpois
-#'@importFrom shiny shinyjs shinydashboard
+#'@importFrom shiny div p icon headerPanel column hr sliderInput selectInput
+#'@importFrom shiny radioButtons actionButton plotOutput eventReactive runApp
+#'@importFrom shiny shinyApp renderPlot uiOutput
+#'@importFrom shinydashboard dashboardSidebar sidebarMenu menuItem dashboardBody
+#'@importFrom shinydashboard tabItems tabItem dashboardHeader dashboardPage
+#'@importFrom shinyjs useShinyjs
 #'@export
 gmse_gui <- function(){ 
 
@@ -68,7 +75,7 @@ sidebar <-   dashboardSidebar(
                 
                 hr(style ="width: 70%; color: white; align: center"),
                 
-                menuItem("Results", icon = icon("line-chart"), 
+                menuItem("General results", icon = icon("line-chart"), 
                          tabName = "plotting"),
 
                 menuItem("Source code for app", icon = icon("code"),
@@ -300,7 +307,7 @@ body <- dashboardBody(
                                    value = 300),
                        
                        sliderInput("manage_freq",
-                                   "Frequency management decisions are enacted",
+                                   "Manage frequency (once per N time steps)",
                                    min   = 1,
                                    max   = 5,
                                    step  = 1,
@@ -573,7 +580,5 @@ app <- shinyApp(ui, server)
 
 #-------------------------------------------------------------------------------
 }
-
-
 
 

### 2017-08-17 R/gmse_gui/app.R

@@ -3,6 +3,11 @@ library(shinydashboard)
 library(shinyjs)
 library(GMSE)
 
+################################################################################
+# Note: This code is for uploading to the shiny server, the gui called from the
+# GMSE is located in the gmse_gui.R file.
+################################################################################
+
 skin <- Sys.getenv("DASHBOARD_SKIN")
 skin <- tolower(skin)
 if (skin == "") skin <- "green"
@@ -53,7 +58,7 @@ sidebar <-   dashboardSidebar(
                 
                 hr(style ="width: 70%; color: white; align: center"),
                 
-                menuItem("Results", icon = icon("line-chart"), 
+                menuItem("General results", icon = icon("line-chart"), 
                          tabName = "plotting"),
                 
                 menuItem("Source code for app", icon = icon("code"),
@@ -285,7 +290,7 @@ body <- dashboardBody(
                                    value = 300),
                        
                        sliderInput("manage_freq",
-                                   "Management frequency (every N years)",
+                                   "Manage frequency (once per N time steps)",
                                    min   = 1,
                                    max   = 5,
                                    step  = 1,
@@ -477,6 +482,8 @@ header <- dashboardHeader(
 #The below calls the user environment (I assume in the normal shiny way)
 ui <- dashboardPage(header, sidebar, body, skin = skin)
 
+#The server function itself contains everything else to get the program to run
+# server <- function(input, output, session){}
 server <- function(input, output){
     
     run_gmse <- eventReactive(input$go, {
@@ -547,7 +554,7 @@ server <- function(input, output){
                           COST = sim$cost);
     })
     
-}
+}    
 
 # Run the application 
 shinyApp(ui = ui, server = server)

### 2017-08-17 tests/testthat/test-make_gmse.R

@@ -10,25 +10,19 @@ test_that("Dimensions of simulation list output are correct", {
 
 test_that("Dimensions of simulation resource output are correct", {
     expect_equal(length(sim[[1]]), 2);
-    expect_equal(dim(sim[[1]][[1]]), c(208, 20));
-    expect_equal(dim(sim[[1]][[2]]), c(219, 20));
 })
 
 test_that("Dimensions of simulation observation output are correct", {
     expect_equal(length(sim[[2]]), 2);
-    expect_equal(dim(sim[[2]][[1]]), c(151, 61));
-    expect_equal(dim(sim[[2]][[2]]), c(169, 61));
 })
 
 test_that("Dimensions of simulation paras output are correct", {
-    expect_equal(length(sim[[3]]), 204);
-    expect_equal(dim(sim[[3]]), c(2, 102));
+    expect_equal(dim(sim[[3]])[2], 105);
 })
 
 test_that("Dimensions of simulation landscape output are correct", {
     expect_equal(length(sim[[4]]), 2);
-    expect_equal(dim(sim[[4]][[1]]), c(100, 100));
-    expect_equal(dim(sim[[4]][[2]]), c(100, 100));
+    expect_equal(dim(sim[[4]][[1]]), c(100, 100, 3));
 })
 
 test_that("Dimensions of time elapsed are correct", {
@@ -36,19 +30,16 @@ test_that("Dimensions of time elapsed are correct", {
 })
 
 test_that("Dimensions of simulation agent output are correct", {
-     expect_equal(dim(sim[[6]]), c(5, 17));
+     expect_equal(dim(sim[[6]][[1]]), c(5, 17));
 })
 
 test_that("Dimensions of simulation cost array output are correct", {
-    expect_equal(length(sim[[7]]), 2);
     expect_equal(dim(sim[[7]][[1]]), c(7, 13, 5));
-    expect_equal(dim(sim[[7]][[2]]), c(7, 13, 5));
 })
 
 test_that("Dimensions of simulation cost array output are correct", {
     expect_equal(length(sim[[8]]), 2);
     expect_equal(dim(sim[[8]][[1]]), c(7, 13, 5));
-    expect_equal(dim(sim[[8]][[2]]), c(7, 13, 5));
 })
 
 set.seed(Sys.time())

### 2017-09-06 R/gmse_apply.R
A bit more to the apply function

@@ -19,3 +19,5 @@ gmse_apply <- function(){
 }
 
 
+
+

### 2017-09-13 notebook/presentation/GMSE_presentation.Rmd
Initial commit of presentation

---
title: "GMSE: a general tool for management strategy evaluation"
subtitle: "https://bradduthie.shinyapps.io/GMSE_presentation/#1"
author: "Brad Duthie"
date: "19 September 2017"
output: 
  ioslides_presentation:
    logo: Confoobio-stacked-logo.png
    css: slide.css
    widescreen: no
    fig_width: 5
    fig_height: 5
runtime: shiny
---

## Introduction

## Background on ConFooBio

Need some background on management

## Management Strategy Evaluation

Show the general structure, point out these four submodels being quite novel
Lead into next slide by pointing out that users and managers are agents

## Conservation conflict

## Point about conflict and games being complex

## Genetic algorithms

Show the basic idea
Use the gamesGA package to illustrate it in a game-theoretic context

## Introduce GMSE

## Geese example

## Future software development

### 2017-09-14 notebook/presentation/GMSE_presentation.Rmd
New scattered slides

@@
 
+##
+
+![](wildlife_conflict_sm.jpg)
+
+<center>
+**Image**: [Public domain](https://commons.wikimedia.org/wiki/File:Human-wildlife_conflict.jpg)
+</center>
+
+## Conservation conflict
+
+- People with strongly held views clash over objectives
+- Negative effect on sustainability, biodiversity, and human livelihood
+
+## Game theory: conflict and cooperation
+
+<div class="columns-2">
+
+- Formalisation and study of conflict and cooperation between rational decision-makers
+- Broadly encompasses combinatorial games (e.g., tic-tac-toe, chess) and evolutionary invasion analysis (adaptive dynamics)
+- Compatibile with agent-based modelling
+
+![](payoff_matrix_sm.png)
+
+</div>
+<hr>
+
+[Adami et al. (2016) Phys. Life. Rev. 19:1-26](http://dx.doi.org/10.1016/j.plrev.2016.08.015)
+
+## {.smaller}
+
+<div class="columns-2">
+
+![Public license](US_USSR_nukes_sm.png).
+
+![Wilkinson 1990](Wilkinson1990_sm.png).
+
+![Isaacdm](penalty_sm.png).
+
+![Dennig](brexit_sm.png).
+
+</div>
+
+**Images:** (1) [Public license](https://commons.wikimedia.org/wiki/File:US_and_USSR_nuclear_stockpiles.svg), (2) [Wilkinson 1990](https://projects.ncsu.edu/project/evoresources/Evolutionary%20medicine/Unitsofselection%20DM%20copy/wilkinson-foodsharing.pdf), (3) [Isaacdm](https://commons.wikimedia.org/wiki/File:PhysicsofPenalty.JPG?uselang=en-gb), (4) [Public license](http://www.publicdomainpictures.net/view-image.php?image=180560&picture=brexit-referendum-uk)
+
+## Conflict in conservation
+
+Game theory provides a natural framework for addressing issues of cooperation and conflict in conservation.
+
+- Game theory can be used to understand stake-holder decision making ([Kark et al. 2015 *Curr. Opin. Environ. Sustainability* **12**:12-24]()).
+
+- Game-theoretic models can suggest novel management strategies ([Colvan et al.
+2011 *Biol.
+Conserv.* **144**:1246-1253](http://dx.doi.org/10.1016/j.biocon.2010.10.028); [Tilman et al.
+2016 *Theor.
+Ecol.*](http://link.springer.com/article/10.1007%2Fs12080-016-0318-8)).
+
+- Game theory can consider multiple management objectives simultaneously ([Lee 2012 *Chemosphere* **87**:608-613](http://dx.doi.org/10.1016/j.chemosphere.2012.01.014)).
+
+## Point about conflict and games being complex
+
+Nils points out that typical game theory assumes that
+
+- pay-offs are known and fixed
+- Actions are clear and always available
+- Players behave rationally
+- Game rules are known to all
+- There is a clear rank order of outcomes
+
 ## Background on ConFooBio
 
-Need some background on management
+Resolving conflicts between food security and biodiversity conservation under uncertainty
+
+<div class = "columns-2">
+
+![](confoobio_sans.png)
+
+- How to solve conflict
+- What leads to conflict
+- Impact on sustainability
+
+All of the above are characterised by uncertainty
+
+</div>
+
+<!--- Get the ERC logo and EU flag in here? --->
 
@@
 
-## Conservation conflict
-
-## Point about conflict and games being complex
+- Confoobio brings in a game-theoretic perspective to this
 
@@
 
+## One run
+
 ## Introduce GMSE
+
+<div class = "columns-2">
+
+![](GMSE_Logo_Goose_300.png)
+
+</div>
 
@@
 
+- Use some data from Jeremy's paper?
+
 ## Future software development

### 2017-09-14 notebook/presentation/GMSE_presentation.Rmd
Restructuring and new mse figure
More details and structure

@@
 
+```{r, echo=FALSE}
+library(gamesGA);
+library(GMSE);
+library(knitr);
+```
+
 ## Introduction
+
+## Process of adaptive management
+
+## Management Strategy Evaluation
+
+![](mse_overview_sm.png)
+
+## Uncertainty
 
@@
 - Negative effect on sustainability, biodiversity, and human livelihood
+
+## Background on ConFooBio
+
+Resolving conflicts between food security and biodiversity conservation under uncertainty
+
+<div class = "columns-2">
+
+![](confoobio_sans.png)
+
+- How to solve conflict
+- What leads to conflict
+- Impact on sustainability
+
+All of the above are characterised by uncertainty
+
+</div>
+
+<!--- Get the ERC logo and EU flag in here? --->
 
@@
 
-## Background on ConFooBio
-
-Resolving conflicts between food security and biodiversity conservation under uncertainty
-
-<div class = "columns-2">
-
-![](confoobio_sans.png)
-
-- How to solve conflict
-- What leads to conflict
-- Impact on sustainability
-
-All of the above are characterised by uncertainty
-
-</div>
-
-<!--- Get the ERC logo and EU flag in here? --->
-
-## Management Strategy Evaluation
-
-Show the general structure, point out these four submodels being quite novel
-Lead into next slide by pointing out that users and managers are agents
-
-- Confoobio brings in a game-theoretic perspective to this
-
 ## Genetic algorithms
@@
 
-## One run
+<!--- 
+    START OF THE GENETIC ALGORITHM SET OF SLIDES  ------------------------------
+--->
+
+## Structure of a genetic algorithm
+
+<div class="columns-2">
+
+```{r, echo=FALSE}
+mbox <- function(x0, x1, y0, y1){
+    xx <- seq(from=x0, to=x1, length.out = 100);
+    yy <- seq(from=y0, to=y1, length.out = 100);
+    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
+    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
+    return(list(x=xd, y=yd));
+}
+
+par(mar=c(0,0,0,0));
+# ===============================================================
+plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
+     xlab="",ylab="", bty="n");
+abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
+polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
+text(x=22, y=85, labels="Random", col="black");
+text(x=22, y=80, labels="seed of", col="black");
+text(x=22, y=75, labels="strategies", col="black");
+# --------------------------------------------------------------
+cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
+polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
+arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
+dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
+polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=84, labels="Begin new", col="black");
+text(x=105, y=79, labels="generation", col="black");
+# ===============================================================
+```
+
+<br>
+
+**Table of strategies**
+
+```{r, echo=FALSE}
+randvals <- sample(x=1:5, size=32, replace=TRUE);
+agents   <- matrix(data=randvals, nrow=8);
+rownames(agents) <- c("Strategy 1", "Strategy 2", "Strategy 3", "Strategy 4",
+                      "Strategy 5", "Strategy 6", "Strategy 7", "Strategy 8");
+colnames(agents) <- c("L1", "L2", "L3", "L4");
+kable(agents, format="html");
+```
+
+</div>
+
+## Structure of a genetic algorithm
+
+<div class="columns-2">
+
+```{r, echo=FALSE}
+mbox <- function(x0, x1, y0, y1){
+    xx <- seq(from=x0, to=x1, length.out = 100);
+    yy <- seq(from=y0, to=y1, length.out = 100);
+    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
+    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
+    return(list(x=xd, y=yd));
+}
+
+par(mar=c(0,0,0,0));
+# ===============================================================
+plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
+     xlab="",ylab="", bty="n");
+abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
+polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
+text(x=22, y=85, labels="Random", col="black");
+text(x=22, y=80, labels="seed of", col="black");
+text(x=22, y=75, labels="strategies", col="black");
+# --------------------------------------------------------------
+cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
+polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
+arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
+dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
+polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=84, labels="Begin new", col="black");
+text(x=105, y=79, labels="generation", col="black");
+# --------------------------------------------------------------
+ebox <- mbox(x0 = 75, x1 = 135, y0 = 63, y1 = 73);
+polygon(x=ebox$x, y=ebox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=68, labels="recombination", col="black");
+# ===============================================================
+```
+
+<br>
+
+**Table of strategies**
+
+```{r, echo=FALSE}
+for(i in 1:dim(agents)[1]){
+    for(j in 1:dim(agents)[2]){
+        check <- runif(n=1);
+        if(check < 0.2){
+            other            <- sample(x=1:dim(agents)[1], size=1);
+            temp             <- agents[i,j];
+            agents[i,j]      <- agents[other, j];
+            agents[other, j] <- temp;
+        }
+    }
+}
+kable(agents, format="html");
+```
+
+</div>
+
+## Structure of a genetic algorithm
+
+<div class="columns-2">
+
+```{r, echo=FALSE}
+mbox <- function(x0, x1, y0, y1){
+    xx <- seq(from=x0, to=x1, length.out = 100);
+    yy <- seq(from=y0, to=y1, length.out = 100);
+    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
+    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
+    return(list(x=xd, y=yd));
+}
+
+par(mar=c(0,0,0,0));
+# ===============================================================
+plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
+     xlab="",ylab="", bty="n");
+abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
+polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
+text(x=22, y=85, labels="Random", col="black");
+text(x=22, y=80, labels="seed of", col="black");
+text(x=22, y=75, labels="strategies", col="black");
+# --------------------------------------------------------------
+cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
+polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
+arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
+dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
+polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=84, labels="Begin new", col="black");
+text(x=105, y=79, labels="generation", col="black");
+# --------------------------------------------------------------
+ebox <- mbox(x0 = 75, x1 = 135, y0 = 63, y1 = 73);
+polygon(x=ebox$x, y=ebox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=68, labels="recombination", col="black");
+# --------------------------------------------------------------
+fbox <- mbox(x0 = 75, x1 = 135, y0 = 51, y1 = 61);
+polygon(x=fbox$x, y=fbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=56, labels="mutation", col="black");
+# ===============================================================
+```
+
+<br>
+
+**Table of strategies**
+
+```{r, echo=FALSE}
+for(i in 1:dim(agents)[1]){
+    for(j in 1:dim(agents)[2]){
+        check <- runif(n=1);
+        if(check < 0.2){
+            agents[i,j]  <- sample(x=1:5, size=1);
+        }
+    }
+}
+kable(agents, format="html");
+```
+
+</div>
+
+## Structure of a genetic algorithm
+
+<div class="columns-2">
+
+```{r, echo=FALSE}
+mbox <- function(x0, x1, y0, y1){
+    xx <- seq(from=x0, to=x1, length.out = 100);
+    yy <- seq(from=y0, to=y1, length.out = 100);
+    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
+    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
+    return(list(x=xd, y=yd));
+}
+
+par(mar=c(0,0,0,0));
+# ===============================================================
+plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
+     xlab="",ylab="", bty="n");
+abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
+polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
+text(x=22, y=85, labels="Random", col="black");
+text(x=22, y=80, labels="seed of", col="black");
+text(x=22, y=75, labels="strategies", col="black");
+# --------------------------------------------------------------
+cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
+polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
+arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
+dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
+polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=84, labels="Begin new", col="black");
+text(x=105, y=79, labels="generation", col="black");
+# --------------------------------------------------------------
+ebox <- mbox(x0 = 75, x1 = 135, y0 = 63, y1 = 73);
+polygon(x=ebox$x, y=ebox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=68, labels="recombination", col="black");
+# --------------------------------------------------------------
+fbox <- mbox(x0 = 75, x1 = 135, y0 = 51, y1 = 61);
+polygon(x=fbox$x, y=fbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=56, labels="mutation", col="black");
+# --------------------------------------------------------------
+gbox <- mbox(x0 = 75, x1 = 135, y0 = 39, y1 = 49);
+polygon(x=gbox$x, y=gbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=44, labels="check fitness", col="black");
+# ===============================================================
+```
+
+<br>
+
+**Table of strategies**
+
+```{r, echo=FALSE}
+fit <- apply(agents,1,sum);
+kable(cbind(agents,fit), format="html");
+```
+
+</div>
+
+## Structure of a genetic algorithm
+
+<div class="columns-2">
+
+```{r, echo=FALSE}
+mbox <- function(x0, x1, y0, y1){
+    xx <- seq(from=x0, to=x1, length.out = 100);
+    yy <- seq(from=y0, to=y1, length.out = 100);
+    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
+    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
+    return(list(x=xd, y=yd));
+}
+
+par(mar=c(0,0,0,0));
+# ===============================================================
+plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
+     xlab="",ylab="", bty="n");
+abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
+polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
+text(x=22, y=85, labels="Random", col="black");
+text(x=22, y=80, labels="seed of", col="black");
+text(x=22, y=75, labels="strategies", col="black");
+# --------------------------------------------------------------
+cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
+polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
+arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
+dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
+polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=84, labels="Begin new", col="black");
+text(x=105, y=79, labels="generation", col="black");
+# --------------------------------------------------------------
+ebox <- mbox(x0 = 75, x1 = 135, y0 = 63, y1 = 73);
+polygon(x=ebox$x, y=ebox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=68, labels="recombination", col="black");
+# --------------------------------------------------------------
+fbox <- mbox(x0 = 75, x1 = 135, y0 = 51, y1 = 61);
+polygon(x=fbox$x, y=fbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=56, labels="mutation", col="black");
+# --------------------------------------------------------------
+gbox <- mbox(x0 = 75, x1 = 135, y0 = 39, y1 = 49);
+polygon(x=gbox$x, y=gbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=44, labels="check fitness", col="black");
+# --------------------------------------------------------------
+hbox <- mbox(x0 = 75, x1 = 135, y0 = 27, y1 = 37);
+polygon(x=hbox$x, y=hbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=32, labels="selection", col="black");
+# ===============================================================
+```
+
+<br>
+
+**Table of strategies**
+
+```{r, echo=FALSE}
+ord <- order(fit, decreasing=TRUE);
+agents <- agents[ord,];
+fit    <- fit[ord];
+kable(cbind(agents,fit), format="html");
+```
+
+</div>
+
+## Structure of a genetic algorithm
+
+<div class="columns-2">
+
+```{r, echo=FALSE}
+mbox <- function(x0, x1, y0, y1){
+    xx <- seq(from=x0, to=x1, length.out = 100);
+    yy <- seq(from=y0, to=y1, length.out = 100);
+    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
+    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
+    return(list(x=xd, y=yd));
+}
+
+par(mar=c(0,0,0,0));
+# ===============================================================
+plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
+     xlab="",ylab="", bty="n");
+abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
+polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
+text(x=22, y=85, labels="Random", col="black");
+text(x=22, y=80, labels="seed of", col="black");
+text(x=22, y=75, labels="strategies", col="black");
+# --------------------------------------------------------------
+cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
+polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
+arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
+dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
+polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=84, labels="Begin new", col="black");
+text(x=105, y=79, labels="generation", col="black");
+# --------------------------------------------------------------
+ebox <- mbox(x0 = 75, x1 = 135, y0 = 63, y1 = 73);
+polygon(x=ebox$x, y=ebox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=68, labels="recombination", col="black");
+# --------------------------------------------------------------
+fbox <- mbox(x0 = 75, x1 = 135, y0 = 51, y1 = 61);
+polygon(x=fbox$x, y=fbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=56, labels="mutation", col="black");
+# --------------------------------------------------------------
+gbox <- mbox(x0 = 75, x1 = 135, y0 = 39, y1 = 49);
+polygon(x=gbox$x, y=gbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=44, labels="check fitness", col="black");
+# --------------------------------------------------------------
+hbox <- mbox(x0 = 75, x1 = 135, y0 = 27, y1 = 37);
+polygon(x=hbox$x, y=hbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=32, labels="selection", col="black");
+# --------------------------------------------------------------
+ibox <- mbox(x0 = 75, x1 = 135, y0 = 15, y1 = 25);
+polygon(x=ibox$x, y=ibox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=20, labels="replacement", col="black");
+# ===============================================================
+```
+
+<br>
+
+**Table of strategies**
+
+```{r, echo=FALSE}
+agents <- rbind(agents[1,],
+                agents[1,],
+                agents[2,],
+                agents[2,],
+                agents[3,],
+                agents[3,],
+                agents[4,],
+                agents[4,]
+                );
+fit    <- c(fit[1], fit[1], fit[2], fit[2], fit[3], fit[3], fit[4], fit[4]);
+kable(cbind(agents,fit), format="html");
+```
+
+</div>
+
+## Structure of a genetic algorithm
+
+<div class="columns-2">
+
+```{r, echo=FALSE}
+mbox <- function(x0, x1, y0, y1){
+    xx <- seq(from=x0, to=x1, length.out = 100);
+    yy <- seq(from=y0, to=y1, length.out = 100);
+    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
+    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
+    return(list(x=xd, y=yd));
+}
+
+par(mar=c(0,0,0,0));
+# ===============================================================
+plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
+     xlab="",ylab="", bty="n");
+abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
+polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
+text(x=22, y=85, labels="Random", col="black");
+text(x=22, y=80, labels="seed of", col="black");
+text(x=22, y=75, labels="strategies", col="black");
+# --------------------------------------------------------------
+cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
+polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
+arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
+dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
+polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=84, labels="Begin new", col="black");
+text(x=105, y=79, labels="generation", col="black");
+# --------------------------------------------------------------
+ebox <- mbox(x0 = 75, x1 = 135, y0 = 63, y1 = 73);
+polygon(x=ebox$x, y=ebox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=68, labels="recombination", col="black");
+# --------------------------------------------------------------
+fbox <- mbox(x0 = 75, x1 = 135, y0 = 51, y1 = 61);
+polygon(x=fbox$x, y=fbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=56, labels="mutation", col="black");
+# --------------------------------------------------------------
+gbox <- mbox(x0 = 75, x1 = 135, y0 = 39, y1 = 49);
+polygon(x=gbox$x, y=gbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=44, labels="check fitness", col="black");
+# --------------------------------------------------------------
+hbox <- mbox(x0 = 75, x1 = 135, y0 = 27, y1 = 37);
+polygon(x=hbox$x, y=hbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=32, labels="selection", col="black");
+# --------------------------------------------------------------
+ibox <- mbox(x0 = 75, x1 = 135, y0 = 15, y1 = 25);
+polygon(x=ibox$x, y=ibox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=20, labels="replacement", col="black");
+# --------------------------------------------------------------
+jbox <- mbox(x0 = 75, x1 = 135, y0 = 3, y1 = 13);
+polygon(x=jbox$x, y=jbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=8, labels="converge?", col="black");
+# --------------------------------------------------------------
+bbox <- mbox(x0 = 0, x1 = 45, y0 = 20, y1 = 40);
+polygon(x=bbox$x, y=bbox$y, lwd=3, border="blue", col="lightsteelblue1");
+arrows(x0=70, x1=22.5, y0=5, y1=5, lwd=2, length=0.0);
+arrows(x0=22, x1=22, y0=5, y1=20, lwd=2, length=0.15);
+arrows(x0=55, x1=55, y0=5, y1=82, lwd=2, length=0.15);
+text(x=15, y=10, labels="Yes", col="black", srt=90);
+text(x=50, y=10, labels="No", col="black", srt=90);
+text(x=22, y=33, labels="Adaptive", col="black");
+text(x=22, y=28, labels="strategies", col="black");
+# ===============================================================
+```
+
+<br>
+
+**Table of strategies**
+
+```{r, echo=FALSE}
+kable(cbind(agents,fit), format="html");
+```
+
+</div>
+
+<!--- 
+    END OF THE GENETIC ALGORITHM SET OF SLIDES  --------------------------------
+--->
+
+##
+
+![](mse_overview_sm.png)
+
+##
+
+![](gmse_overview_sm.png)
 
@@
 ![](GMSE_Logo_Goose_300.png)
+
+- Objective of GMSE 1
+- Objective of GMSE 2
+- Objective of GMSE 3
+
+</div>
+
+## Introduce GMSE
+
+<div class = "columns-2">
+
+![](GMSE_Logo_Goose_300.png)
+
+- Technical detail of GMSE 1
+- Technical detail of GMSE 2
+- Technical detail of GMSE 3
 

### 2017-09-14 notebook/presentation/GMSE_presentation.Rmd
More progress

@@
 
-## Introduction
+<!---
+https://welikia.org/
+http://ngm.nationalgeographic.com/2009/09/manhattan/miller-text
+https://welikia.org/explore/mannahatta-map/
+
+--->
+
+##
+
+<center>
+
+![](mannahatta.jpg)
+
+</center>
+
+##
+
+![](food_biodiversity.png)
+
+##
+
+![](wicked_modelling.png)
 
@@
 
-## Management Strategy Evaluation
+##
+
+<center>
 
@@
 
-## Uncertainty
+</center>
+
+## Uncertainty in natural resource management
+
+Uncertainty is fundamental in ecological and economic systems, and their measurement
+
+**Natural resources**: Structural and demographic uncertainty
+
+**Resource observation**: Observation uncertainty
+
+**Management**: Implementation uncertainty
+
+**Stake-holder actions**: Pay-off uncertainty
 
 ##
+
+<!--- Transition from global to local scale --->
 
@@
 - Negative effect on sustainability, biodiversity, and human livelihood
+- MSE does not integrate the livelihood decisions of people
 
@@
 
+## Iterated Prisoner's dilemma rules {.smaller}
+
+<div class="columns-2">
+
+![](payoff_matrix_sm.png)
+**General structure of Prisoner's dilemma scenario**
+
+![](pd_pays.png)
+**Specific payoffs satisfying a Prisoner's dilemma scenario**
+
+</div>
+
+## Use of genetic algorithms {.smaller}
+
+**Each locus is an focal player's choice given an opponent's past three choices (plus one choice of first move)**
+
+<hr>
+
+|                  | **CCC** | **CCD** | **CDC** | **CDD** | **DCC** | **DCD** | **DDC** | **DDD** | **1st** |
+|------------------|---------|---------|---------|---------|---------|---------|---------|---------|---------|
+| **Strategy 1**   | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  |
+| **Strategy 2**   | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  |
+| **...**          |   ...   |   ...   |   ...   |   ...   |   ...   |   ...   |   ...   |   ...   |   ...   |
+| **Strategy N-1** | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  |
+| **Strategy N**   | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  |
+
+<hr>
+
+**Total of 512 possible strategies**
+
+## Use of genetic algorithms {.smaller}
+
+```{r, echo=FALSE}
+inputPanel(
+
+  numericInput("CC", label = "CC", 0, width="30%"),
+  numericInput("DC", label = "DC", 0, width="30%"),
+  numericInput("CD", label = "CD", 0, width="30%"),
+  numericInput("DD", label = "DD", 0, width="30%"),
+
+  style='width: 2000px; height: 60px'
+)
+```
+
+<br>
+
+```{r, echo=FALSE}
+gres <- reactive({ 
+     cc <- input$CC;
+     dc <- input$DC;
+     cd <- input$CD;
+     dd <- input$DD;
+     rd <- 100;
+     gn <- 100;
+     co <- 0.05;
+     mu <- 0.05;
+
+     game_res <- games_ga(CC = cc, DC = dc, CD = cd, DD = dd, rounds = rd,
+                          generations = gn, cross_prob = co, 
+                          mutation_prob = mu, num_opponents = 100);     
+     all <- c(game_res$genos, "break_1",
+              rownames(game_res$genos), "break_2",
+              game_res$fitness);
+    })
+```
+
+```{r, echo = FALSE}
+renderPlot({
+    break2    <- which(gres()=="break_2")[1]+1;
+    end       <- length(gres());
+    fitnesses <- gres()[break2:end];
+    fitnesses <- as.numeric(fitnesses);
+    mean_fit  <- fitnesses / (100 * 100);
+    maxpt     <- max(c(input$CC, input$CD, input$DC, input$DD));
+    par(mar=c(5,5,1,1));
+    plot(x = 1:length(mean_fit), y=mean_fit, type="l", ylim=c(0,maxpt), lwd = 3,
+        xlab="Generation", ylab = "Mean strategy fitness per round",
+         cex.axis=1.5, cex.lab=1.5);
+})
+```
+
+## Adaptive strategies simulated {.smaller}
+
+```{r, echo=FALSE}
+renderTable({
+      break1   <- which(gres()=="break_1")[1] - 1;
+      break2   <- which(gres()=="break_2")[1] - 1;
+      geno_vec <- gres()[1:break1];
+      geno_row <- gres()[(break1+2):break2];
+      genos    <- matrix(data =  geno_vec, ncol = 10, byrow = FALSE);
+      rownames(genos) <- geno_row;
+      colnames(genos) <- c("CCC", "CCD", "CDC", "CDD", "DCC", "DCD", "DDC",
+                           "DDD", "1st", "Final %");
+      final_table <- genos[1:10,];
+
+}, include.rownames=TRUE)
+```
+
 ##
+
+<center>
 
@@
 
+</center>
+
 ##
 
+<center>
+
 ![](gmse_overview_sm.png)
+
+</center>
 

### 2017-09-14 notebook/presentation/GMSE_presentation.Rmd
New structures and before-after image

@@
 --->
+
+##
+
+<center>
+
+![](mannahatta_before.jpg)
+
+</center>
 
@@
 
+**Conservation conflicts are complex, and characterised by incomplete information that is rapidly changing.**
+
 - People with strongly held views clash over objectives
@@
 
+- What leads to conflict
 - How to solve conflict
-- What leads to conflict
 - Impact on sustainability
@@
 
-## Point about conflict and games being complex
+## Simple game-theoretic models insufficient
 
-Nils points out that typical game theory assumes that
+![](payoff_matrix_sm.png)
 
@@
 
-- Objective of GMSE 1
-- Objective of GMSE 2
-- Objective of GMSE 3
+- Integrates game theory and ecological theory to construct social-ecological models for adaptive management
+- Simulates all aspects of management: population dynamics, manager observation of populations, manager decision making, and stakeholder responses to management decisions
+- Mimic the decision-making process of managers and stakeholders under conditions of change, uncertainty, and conflict
+- Provide a flexible modelling framework for future development of complex adaptive management models to be applied to conservation theory and specific case studies
 
@@
 
-- Technical detail of GMSE 1
-- Technical detail of GMSE 2
-- Technical detail of GMSE 3
+- Written in R and C
+- Requires R (>= 3.3.3)
+- License GPL (>=2)
 

### 2017-09-14 notebook/presentation/GMSE_presentation.Rmd
A bit more on the geese example

@@
 
-- Use some data from Jeremy's paper?
+- Wild geese declined starting in the 1930s, likely caused by habitat loss and increased hunting
+- Refuges were established and hunting restricted in the 1950s, leading to recovery
+
+[Fox & Madsen 2017 *Ambio* **46(s2)**:170-187](http://doi.org/10.1007/s13280-016-0878-2)
 

### 2017-09-14 notebook/presentation/GMSE_presentation.Rmd
More on geese

--- 2017-09-14 notebook/presentation/GMSE_presentation.Rmd
@@
 
-- Use some data from Jeremy's paper?
+- Wild geese declined starting in the 1930s, likely caused by habitat loss and increased hunting
+- Refuges were established and hunting restricted in the 1950s, leading to recovery
+- Since the 1960s, many populations have exponentially increased
+- Population growth coincides with a switch from natural wetlands to agricultural landscapes
+- Switch to agricultural exploitation causes conflict with farmers
+
+[Fox & Madsen 2017 *Ambio* **46(s2)**:170-187](http://doi.org/10.1007/s13280-016-0878-2)
+[Fox & Abraham 2017 *Ambio* **46(s2)**:188-197](http://doi.org/10.1007/s13280-016-0879-1)
 

### 2017-09-15 notebook/presentation/GMSE_presentation.Rmd
Update slides, mostly to the introduction

@@
 
-##
+## {.smaller}
 
@@
 
-![](mannahatta_before.jpg)
+![](mannahatta_before.png)
 
@@
 
-![](mannahatta.jpg)
+![](mannahatta.png)
 
 </center>
+
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Computer Image: [Markley Boyer](https://www.facebook.com/markley.boyer), Photo Image: [Robert Clark](http://www.nationalgeographic.com/contributors/c/photographer-robert-clark/)
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [National Geographic](http://ngm.nationalgeographic.com/2009/09/manhattan/miller-text), [The Welikia Project](https://welikia.org/)
+
+## {.smaller}
+
+<center>
+
+![](potato_fields_Janet_Tench.jpg)
+
+<br>
+
+Image: [Janet Tench](https://commons.wikimedia.org/wiki/File:Norfolk_potato_field_-_geograph.org.uk_-_180521.jpg)
+</center>
+
+## {.smaller}
+
+<center>
+
+![](KehoeFig1.png)
+
+</center>
+
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Kehoe et al.
+(2017) *Nat.
+Ecol.
+Evol.* **1**:1129–1135](https://www.nature.com/articles/s41559-017-0234-3)
 
@@
 
-## Uncertainty in natural resource management
+## {.smaller}
 
-Uncertainty is fundamental in ecological and economic systems, and their measurement
+<center>
 
-**Natural resources**: Structural and demographic uncertainty
+![](mse_overview_sm_pt.png)
 
-**Resource observation**: Observation uncertainty
+</center>
 
-**Management**: Implementation uncertainty
+<br>
 
-**Stake-holder actions**: Pay-off uncertainty
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Bunnefeld et al.
+(2011) *Trends.
+Ecol.
+Evol.* **26**:441-447](https://doi.org/10.1016/j.tree.2011.05.003)
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Chadès et al. (2017) *Theor. Ecol.* **10**:1-10](https://doi.org/10.1007/s12080-016-0313-0)
+
+## {.smaller}
+
+<center>
+
+![](mse_overview_sm_un.png)
+
+</center>
+
+<br>
+
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Bunnefeld et al.
+(2011) *Trends.
+Ecol.
+Evol.* **26**:441-447](https://doi.org/10.1016/j.tree.2011.05.003)
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Chadès et al. (2017) *Theor. Ecol.* **10**:1-10](https://doi.org/10.1007/s12080-016-0313-0)
 
@@
 
-[Adami et al. (2016) Phys. Life. Rev. 19:1-26](http://dx.doi.org/10.1016/j.plrev.2016.08.015)
+[Adami et al. (2016) *Phys. Life. Rev.* **19**:1-26](http://dx.doi.org/10.1016/j.plrev.2016.08.015)
 

### 2017-09-15 notebook/presentation/GMSE_presentation.Rmd
Minor adjustments to the introduction

@@
 
-<br>
+<br><br>
 
-Image: [Janet Tench](https://commons.wikimedia.org/wiki/File:Norfolk_potato_field_-_geograph.org.uk_-_180521.jpg)
 </center>
+
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Image: [Janet Tench](https://commons.wikimedia.org/wiki/File:Norfolk_potato_field_-_geograph.org.uk_-_180521.jpg)
 
@@
 </center>
+
+<br><br><br>
 

### 2017-09-15 notebook/presentation/GMSE_presentation.Rmd
Nearing the endish now... finally

@@
 subtitle: "https://bradduthie.shinyapps.io/GMSE_presentation/#1"
-author: "Brad Duthie"
+author: "Brad Duthie (@bradduthie)"
 date: "19 September 2017"
@@
 
-## Conservation conflict
+##
+
+<div class = "columns-2">
+
+<!--- ![](African_Bush_Elephant.jpg) --->
+
+![](grouse_shooting.jpg)
+
+<br>
 
 **Conservation conflicts are complex, and characterised by incomplete information that is rapidly changing.**
+
+<br>
 
@@
 
-## Background on ConFooBio
+</div>
 
-Resolving conflicts between food security and biodiversity conservation under uncertainty
+<center>
+
+<br>
+
+**Image:** [Public](https://commons.wikimedia.org/wiki/File:Hardy_Heywood_-_The_Grouse_Shoot_1898.jpg) [Domain](https://commons.wikimedia.org/wiki/File:Henharrier.jpg)
+<!--- **Image:** [Muhammad Mahdi Karim](https://en.wikipedia.org/wiki/User:Muhammad_Mahdi_Karim) --->
+
+</center>
+
+##
+
+<div class = "columns-2">
+
+<!--- ![](African_Bush_Elephant.jpg) --->
+
+![](grouse_shooting2.jpg)
+
+<br>
+
+**Conservation conflicts are complex, and characterised by incomplete information that is rapidly changing.**
+
+<br>
+
+- People with strongly held views clash over objectives
+- Negative effect on sustainability, biodiversity, and human livelihood
+- MSE does not integrate the livelihood decisions of people
+
+</div>
+
+<center>
+
+<br>
+
+**Image:** [Public](https://commons.wikimedia.org/wiki/File:Hardy_Heywood_-_The_Grouse_Shoot_1898.jpg) [Domain](https://commons.wikimedia.org/wiki/File:Henharrier.jpg)
+<!--- **Image:** [Muhammad Mahdi Karim](https://en.wikipedia.org/wiki/User:Muhammad_Mahdi_Karim) --->
+
+</center>
+
+## The ConFooBio Project
+
+**Resolving conflicts between food security and biodiversity conservation under uncertainty**
 
@@
 
-- What leads to conflict
-- How to solve conflict
-- Impact on sustainability
+![](group.png)
 
-All of the above are characterised by uncertainty
+</div>
+
+## The ConFooBio Project
+
+**Resolving conflicts between food security and biodiversity conservation under uncertainty**
+
+<div class = "columns-2">
+
+![](confoobio_sans.png)
+
+<br>
+
+- What leads to conservation conflict
+- How to resolve conservation conflict
+- Impact on sustainability of social-ecological systems
 
@@
 
-[Adami et al. (2016) *Phys. Life. Rev.* **19**:1-26](http://dx.doi.org/10.1016/j.plrev.2016.08.015)
+<br><br>
+
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Adami et al. (2016) *Phys. Life. Rev.* **19**:1-26](http://dx.doi.org/10.1016/j.plrev.2016.08.015)
 
@@
 
-**Images:** (1) [Public license](https://commons.wikimedia.org/wiki/File:US_and_USSR_nuclear_stockpiles.svg), (2) [Wilkinson 1990](https://projects.ncsu.edu/project/evoresources/Evolutionary%20medicine/Unitsofselection%20DM%20copy/wilkinson-foodsharing.pdf), (3) [Isaacdm](https://commons.wikimedia.org/wiki/File:PhysicsofPenalty.JPG?uselang=en-gb), (4) [Public license](http://www.publicdomainpictures.net/view-image.php?image=180560&picture=brexit-referendum-uk)
+<br><br>
+
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; **Images:** (1) [Public license](https://commons.wikimedia.org/wiki/File:US_and_USSR_nuclear_stockpiles.svg), (2) [Wilkinson 1990](https://projects.ncsu.edu/project/evoresources/Evolutionary%20medicine/Unitsofselection%20DM%20copy/wilkinson-foodsharing.pdf), (3) [Isaacdm](https://commons.wikimedia.org/wiki/File:PhysicsofPenalty.JPG?uselang=en-gb), (4) [Public license](http://www.publicdomainpictures.net/view-image.php?image=180560&picture=brexit-referendum-uk)
 
@@
 
-## Simple game-theoretic models insufficient
+## Stake-holders are complex
+
+<div class="columns-2">
+
+**Simple models assume that stake-holders:**
+
+- Know what pay-offs will be
+- Have clearly defined choices
+- Behave rationally
+- Understand the games
+- Prioritise outcomes clearly
 
@@
 
-- pay-offs are known and fixed
-- Actions are clear and always available
-- Players behave rationally
-- Game rules are known to all
-- There is a clear rank order of outcomes
+</div>
 
-## Genetic algorithms
+## Conservation conflicts are complex {.smaller}
 
-Show the basic idea
-Use the gamesGA package to illustrate it in a game-theoretic context
+<div class="columns-2">
+
+ - **Asymmetric payoffs**: Payoffs of choices may depend instead on stakeholder identity.
+
+![](payoff_matrix_asym.png)
+
+</div>
+
+## But most games are more complex {.smaller}
+
+<div class="columns-2">
+
+ - **Asymmetric payoffs**: Payoffs of choices may depend instead on stakeholder identity.
+  - **Multiple choices**: Players might have three or more choices.
+
+![](payoff_matrix_cho.png)
+
+</div>
+
+## But most games are more complex {.smaller}
+
+<div class="columns-2">
+
+ - **Asymmetric payoffs**: Payoffs of choices may depend instead on stakeholder identity.
+ - **Multiple choices**: Players might have three or more choices.
+ - **Multiple players**: More than two stakeholders might interact at one time.
+
+![](payoff_matrix_3pl.png)
+
+</div>
+
+## But most games are more complex {.smaller}
+
+<div class="columns-2">
+
+ - **Asymmetric payoffs**: Payoffs of choices may depend instead on stakeholder identity.
+ - **Multiple choices**: Players might have three or more choices.
+ - **Multiple players**: More than two stakeholders might interact at one time.
+ - **Iterated interactions**: Strategies maybe affected by interaction history.
+
+```{r, echo=FALSE}
+rounds  <- as.character(1:9);
+players <- c("Player 1","Player 2");
+plays   <- sample(x=c(0,1), size=18, replace=TRUE);
+plays[plays == 0] <- "A";
+plays[plays == 1] <- "B";
+history <- matrix(data=plays, nrow=2);
+rownames(history) <- players;
+colnames(history) <- rounds;
+kable(noquote(history));
+```
+
+![](payoff_matrix_rd10_sm.png)
+
+</div>
+
+##{.smaller}
+
+<center>
+
+![](mse_overview_sm.png)
+
+</center>
+
+## {.smaller}
+
+<center>
+
+![](gmse_overview_sm_ga.png)
+
+</center>
+
+<br><br><br>
+
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Bunnefeld et al.
+(2011) *Trends.
+Ecol.
+Evol.* **26**:441-447](https://doi.org/10.1016/j.tree.2011.05.003)
+
+## {.smaller}
+
+<center>
+
+![](gmse_overview_sm_ga2.png)
+
+</center>
 
@@
 
-![](mse_overview_sm.png)
+![](gmse_overview_sm_ga2.png)
 
@@
 
-## Introduce GMSE
+## Genearlised MSE {.smaller}
 
 <div class = "columns-2">
-
-![](GMSE_Logo_Goose_300.png)
 
@@
 
+<br><br>
+
+<center>
+
+![](GMSE_Logo_Goose_375.png)
+
+</center>
+
 </div>
 
-## Introduce GMSE
+## Genearlised MSE ('GMSE' R package)
 
@@
 
-![](GMSE_Logo_Goose_300.png)
-
+- Individual-based
+- Spatially explicit
 - Written in R and C
@@
 - License GPL (>=2)
+- *v0.2.2.7* [available](https://cran.r-project.org/package=GMSE) on CRAN
+- Browser-based GUI [available](https://bradduthie.shinyapps.io/gmse_gui/) via shiny
+- Source code [available](https://github.com/bradduthie/gmse) on GitHub
+- Intended for a wide range of software users
+
+<br><br>
+
+<center>
+
+![](GMSE_Logo_Goose_375.png)
+
+</center>
 
 </div>
+
+##
+
+<center>
+
+![](gmse_overview_sm.png)
+
+</center>
+
+## GMSE: Natural resources
+
+<div class = "columns-2">
+
+- Spatially explicit landscape
+- Discrete individual resources
+- Individual birth, movement, death
+
+<br>
+
+<center>
+
+![](sep_land.gif)
+
+</center>
+
+## GMSE: Resource observation
+
+<div class = "columns-2">
+
+- Four types of observation options
+    1. Counting on a subset of landscape
+    2. Mark and recapture
+    3. Linear transect sampling
+    4. Block transect sampling
+- Mechanistic sampling process
+- Generates simulated data frame used by the manager
+
+<br>
+
+<center>
+
+![](binos.png)
+
+</center>
+
+</div>
+
+<center>
+
+**Image:** [Public Domain](https://commons.wikimedia.org/wiki/File:Fernglas(alt).JPG)
+
+</center>
+
+## GMSE: Management
+
+<div class = "columns-2">
+
+- One manager analyses observations and compares with a target abundance
+- Five possible policy options (scaring, culling, castration, feeding, helping)
+- Fixed budget and a costs for setting policy
+- Genetic algorithm anticipates policy effects from previous actions to set costs of actions for stakeholders
+
+<br>
+
+<center>
+
+![](balance.png)
+
+</center>
+
+## GMSE: Stake-holder actions {.smaller}
+
+<div class = "columns-2">
+
+- Any number of stake-holders
+- Stake-holders want to maximise harvest of resources or minimise resource effects on their land
+- Seven possible actions (scaring, culling, castration, feeding, helping, tend crops, kill crops)
+- Fixed budget for performing actions of potentially differing costs
+- Genetic algorithm anticipates effects of actions on harvest or land yield
+
+<br>
+
+<center>
+
+![](harvesting.jpg)
+
+</center>
+
+</div>
+
+<center>
+
+<br>
+
+**Image:** [Public Domain](https://commons.wikimedia.org/wiki/File:Libiaz-zniwa1938.a.jpg)
+
+</center>
+
+## Running the gmse() function {.smaller}
+
+```{r, echo = TRUE}
+sim <- gmse(land_ownership = TRUE, stakeholders = 10, observe_type = 1,
+            res_death_K = 2000, manage_target = 1000, RESOURCE_ini = 1000,
+            user_budget = 5000, manager_budget = 5000, res_consume = 1,
+            scaring = TRUE, fixed_mark = 50, fixed_recapt = 300,
+            plotting = FALSE);
+```
+
+## Explain two types of plots
+
+##
+
+<center>
+
+```{r, echo = FALSE, fig.height = 6, fig.width = 7}
+plot_gmse_results(res = sim$resource, obs = sim$observation, land = sim$land,
+                  agents = sim$agents, paras = sim$paras, ACTION = sim$action,
+                  COST = sim$cost);
+```
+
+</center>
+
+## Other plot type
 

### 2017-09-18 notebook/presentation/GMSE_presentation.Rmd
Add new plotting function, tweak slides

@@
 <!---
+1. Know what to say about the Welikia project
+
+--->
+
+```{r, echo = FALSE}
+plot_gmse_effort <- function(agents, paras, ACTION, COST){
+
+    cols      <- c("green", "indianred1", "indianred3", "deepskyblue1",
+                   "deepskyblue2");
+
+    users    <- dim(agents[[1]])[1];
+    max_time <- dim(paras)[1];
+
+    para_vec <- paras[1,];
+
+    allowed  <- sum(para_vec[89:93]);
+
+    scar_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
+    cull_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
+    cast_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
+    feed_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
+    help_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
+
+    scar_act  <- matrix(data = 0, nrow = max_time, ncol = users);
+    cull_act  <- matrix(data = 0, nrow = max_time, ncol = users);
+    cast_act  <- matrix(data = 0, nrow = max_time, ncol = users);
+    feed_act  <- matrix(data = 0, nrow = max_time, ncol = users);
+    help_act  <- matrix(data = 0, nrow = max_time, ncol = users);
+
+    scar_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
+    cull_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
+    cast_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
+    feed_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
+    help_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
+
+    act_costs  <- matrix(data = 0, nrow = max_time, ncol = 5);
+    pol_effort <- matrix(data = 0, nrow = max_time, ncol = 5);
+    min_cost   <- para_vec[97];
+
+    for(time in 1:max_time){
+        #-- Scaring cost and actions
+        scar_cst[time, 1] <- COST[[time]][3, 8, 1];
+        for(user in 2:users){
+            scar_cst[time, user] <- COST[[time]][1, 8, user];
+        }
+        scar_act[time, 1] <- ACTION[[time]][3, 8, 1] - min_cost;
+        for(user in 2:users){
+            scar_act[time, user] <- ACTION[[time]][1, 8, user];
+        }
+        scar_eff[time,] <- scar_act[time,]*scar_cst[time,]/agents[[time]][,17];
+        #-- Culling cost and actions
+        cull_cst[time, 1] <- COST[[time]][3, 9, 1];
+        for(user in 2:users){
+            cull_cst[time, user] <- COST[[time]][1, 9, user];
+        }
+        cull_act[time, 1] <- ACTION[[time]][3, 9, 1] - min_cost;
+        for(user in 2:users){
+            cull_act[time, user] <- ACTION[[time]][1, 9, user];
+        }
+        cull_eff[time,] <- cull_act[time,]*cull_cst[time,]/agents[[time]][,17];
+        #-- Castration cost and actions
+        cast_cst[time, 1] <- COST[[time]][3, 10, 1];
+        for(user in 2:users){
+            cast_cst[time, user] <- COST[[time]][1, 10, user];
+        }
+        cast_act[time, 1] <- ACTION[[time]][3, 10, 1] - min_cost;
+        for(user in 2:users){
+            cast_act[time, user] <- ACTION[[time]][1, 10, user];
+        }
+        cast_eff[time,] <- cast_act[time,]*cast_cst[time,]/agents[[time]][,17];
+        #-- Feeding cost and actions
+        feed_cst[time, 1] <- COST[[time]][3, 11, 1];
+        for(user in 2:users){
+            feed_cst[time, user] <- COST[[time]][1, 11, user];
+        }
+        feed_act[time, 1] <- ACTION[[time]][3, 11, 1] - min_cost;
+        for(user in 2:users){
+            feed_act[time, user] <- ACTION[[time]][1, 11, user];
+        }
+        feed_eff[time,] <- feed_act[time,]*feed_cst[time,]/agents[[time]][,17];
+        #-- Helping cost and actions
+        help_cst[time, 1] <- COST[[time]][3, 12, 1];
+        for(user in 2:users){
+            help_cst[time, user] <- COST[[time]][1, 12, user];
+        }
+        help_act[time, 1] <- ACTION[[time]][3, 12, 1] - min_cost;
+        for(user in 2:users){
+            help_act[time, user] <- ACTION[[time]][1, 12, user];
+        }
+        help_eff[time,] <- help_act[time,]*help_cst[time,]/agents[[time]][,17];
+    }
+
+    # -- Turn these into percentages:
+    scar_eff <- scar_eff * 100;
+    cull_eff <- cull_eff * 100;
+    cast_eff <- cast_eff * 100;
+    feed_eff <- feed_eff * 100;
+    help_eff <- help_eff * 100;
+
+    par(mfrow = c(5, 1), mar = c(0, 0, 0, 0), oma = c(6, 6, 4, 6));
+    #---- Scaring
+    if(para_vec[89] == 1){
+        y1 <- 100;
+        y2 <- y1 * 1.2
+        plot(x = 1:max_time, y = 100 - scar_eff[,1], type = "l", lwd = 2, 
+             cex.axis = 1.5, xaxt = "n", ylim = c(0, y2), yaxt = "n");
+        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+        par(new = TRUE);
+        plot(x = 1:max_time, y = scar_eff[,2], type = "n", lwd = 2, 
+             ylim = c(0, y2), xaxt="n", yaxt="n", 
+             cex.axis = 1.5);
+        for(stakeholder in 2:users){
+            points(x = 1:max_time, y = scar_eff[,stakeholder], type = "l", 
+                   lwd = 1, col = cols[1]);
+        }
+        axis(side = 4, at = c(50, 100),  labels = c(50, 100), cex.axis = 1.5);
+        legend(x = 1, y = y2 + y2*0.1, fill = cols[1:5], horiz = TRUE,
+               legend = c("scaring", "culling", "castration", "feeding", 
+                          "helping"), bty = "n", cex = 1.5);
+        abline(h = y1 + y1*0.1, lwd = 2);
+    }else{
+        plot(x = 1:max_time, y = scar_eff[,1], type = "n", xaxt= "n", 
+             yaxt = "n", ylim = c(0, 110));
+        text(x = 0, y = 90, cex = 2, labels = "No scaring allowed",
+             pos = 4);
+        abline(h = 100, lwd = 2);
+        legend(x = 1, y = 119, fill = cols[1:5], horiz = TRUE,
+               legend = c("scaring", "culling", "castration", "feeding", 
+                          "helping"), bty = "n", cex = 1.5);
+    }
+    #---- Culling
+    if(para_vec[90] == 1){
+        plot(x = 1:max_time, y = 100 - cull_eff[,1], type = "l", lwd = 2, 
+             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
+        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+        par(new = TRUE);
+        plot(x = 1:max_time, y = cull_eff[,2], type = "n", lwd = 2, 
+             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
+        for(stakeholder in 2:users){
+            points(x = 1:max_time, y = cull_eff[,stakeholder], type = "l", 
+                   lwd = 1, col = cols[2]);
+        }
+        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+    }else{
+        plot(x = 1:max_time, y = cull_act[,1], type = "n", xaxt= "n", 
+             yaxt = "n", ylim = c(0, 100));
+        text(x = 0, y = 90, cex = 2, labels = "No culling allowed",
+             pos = 4);
+    }
+    #---- Castrating
+    if(para_vec[91] == 1){
+        plot(x = 1:max_time, y = 100 - cast_eff[,1], type = "l", lwd = 2, 
+             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
+        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+        par(new = TRUE);
+        plot(x = 1:max_time, y = cast_eff[,2], type = "n", lwd = 2, 
+             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
+        for(stakeholder in 2:users){
+            points(x = 1:max_time, y = cast_eff[,stakeholder], type = "l", 
+                   lwd = 1, col = cols[3]);
+        }
+        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+    }else{
+        plot(x = 1:max_time, y = cast_act[,1], type = "n", xaxt= "n", 
+             yaxt = "n", ylim = c(0, 100));
+        text(x = 0, y = 90, cex = 2, labels = "No castration allowed",
+             pos = 4);
+    }
+    #---- Axes labels
+    mtext("Manager's permissiveness of user action", 
+          side = 2, line = 3.5, cex = 1.5, col = "black");
+    mtext("Percentage of user action effort expended", 
+          side = 4, line = 3.5, cex = 1.5, col = "black");
+    #---- Feeding
+    if(para_vec[92] == 1){
+        plot(x = 1:max_time, y = 100 - feed_eff[,1], type = "l", lwd = 2, 
+             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
+        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+        par(new = TRUE);
+        plot(x = 1:max_time, y = feed_eff[,2], type = "n", lwd = 2, 
+             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
+        for(stakeholder in 2:users){
+            points(x = 1:max_time, y = feed_eff[,stakeholder], type = "l", 
+                   lwd = 1, col = cols[4]);
+        }
+        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+    }else{
+        plot(x = 1:max_time, y = feed_act[,1], type = "n", xaxt= "n", 
+             yaxt = "n", ylim = c(0, 100));
+        text(x = 0, y = 90, cex = 2, labels = "No feeding allowed",
+             pos = 4);
+    }
+    #---- Helping
+    if(para_vec[93] == 1){
+        plot(x = 1:max_time, y = 100 - help_eff[,1], type = "l", lwd = 2, 
+             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
+        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+        par(new = TRUE);
+        plot(x = 1:max_time, y = help_eff[,2], type = "n", lwd = 2, 
+             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
+        for(stakeholder in 2:users){
+            points(x = 1:max_time, y = help_eff[,stakeholder], type = "l", 
+                   lwd = 1, col = cols[5]);
+        }
+        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+    }else{
+        plot(x = 1:max_time, y = help_act[,1], type = "n", cex.axis = 1.5,
+             yaxt = "n", ylim = c(0, 100));
+        text(x = 0, y = 90, cex = 2, labels = "No helping offspring allowed",
+             pos = 4);
+    }    
+    mtext("Time step", side = 1, line = 3.5, cex = 1.5, col = "black");
+}
+```
+
+<!---
 https://welikia.org/
@@
 
-<br>
-
 **Image:** [Public](https://commons.wikimedia.org/wiki/File:Hardy_Heywood_-_The_Grouse_Shoot_1898.jpg) [Domain](https://commons.wikimedia.org/wiki/File:Henharrier.jpg)
@@
 
-<br>
+<br><br><br>
 
@@
 
-<br><br>
+<br>
 
@@
 
-## GMSE: Resource observation
+## GMSE: Resource observation {.smaller}
 
@@
 
-## GMSE: Management
+## GMSE: Management {.smaller}
 
@@
 
-## Explain two types of plots
+## Inferring GMSE from plots {.smaller}
+
+**General plot of simulation results.**
+
+```{r, echo = TRUE, eval = FALSE}
+plot_gmse_results(res = sim$resource, obs = sim$observation, land = sim$land,
+                  agents = sim$agents, paras = sim$paras, ACTION = sim$action,
+                  COST = sim$cost);
+```
+
+<br>
+
+**Plot of conflict among stakeholders**
+
+```{r, echo = TRUE, eval = FALSE}
+plot_gmse_effort(agents = sim$agents, paras = sim$paras, ACTION = sim$action, 
+                 COST   = sim$cost);
+```
 
@@
 
-## Other plot type
+##
+
+<center>
+
+```{r, echo = FALSE, fig.height = 6, fig.width = 6}
+plot_gmse_effort(agents = sim$agents, paras = sim$paras, ACTION = sim$action, 
+                 COST   = sim$cost);
+```
+
+</center>
 

### 2017-09-18 notebook/presentation/GMSE_presentation.Rmd
Update with title slides

@@
 
-    par(mfrow = c(5, 1), mar = c(0, 0, 0, 0), oma = c(6, 6, 4, 6));
+    par(mfrow = c(5, 1), mar = c(0, 0, 0, 0), oma = c(6, 6, 0, 6));
     #---- Scaring
@@
                legend = c("scaring", "culling", "castration", "feeding", 
-                          "helping"), bty = "n", cex = 1.5);
+                          "helping"), bty = "n", cex = 1);
         abline(h = y1 + y1*0.1, lwd = 2);
@@
                legend = c("scaring", "culling", "castration", "feeding", 
-                          "helping"), bty = "n", cex = 1.5);
+                          "helping"), bty = "n", cex = 1);
     }
@@
 
+# Biodiversity and food security
+
 ## {.smaller}
@@
 
+# Adaptive management and conservation conflicts
+
 ## Process of adaptive management
+
+<div class = "columns-2">
+
+- Uses an iterative process of decision-making
+- Sets explicit management goals
+- Develops alternative management strategies
+- Deals with uncertainty by updating knowledge of the system
+
+<br>
+
+![](adaptive_management.png)
+
+</div>
+
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Keith et al. (2011) *Biol. Conserv.* **144**: 1175-1778](https://doi.org/10.1016/j.biocon.2010.11.022)
 
@@
 
+# Game theory in conservation conflict
+
 <!--- Get the ERC logo and EU flag in here? --->
@@
 
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; **Images:** (1) [Public license](https://commons.wikimedia.org/wiki/File:US_and_USSR_nuclear_stockpiles.svg), (2) [Wilkinson 1990](https://projects.ncsu.edu/project/evoresources/Evolutionary%20medicine/Unitsofselection%20DM%20copy/wilkinson-foodsharing.pdf), (3) [Isaacdm](https://commons.wikimedia.org/wiki/File:PhysicsofPenalty.JPG?uselang=en-gb), (4) [Public license](http://www.publicdomainpictures.net/view-image.php?image=180560&picture=brexit-referendum-uk)
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; **Images:** (1) [Public license](https://commons.wikimedia.org/wiki/File:US_and_USSR_nuclear_stockpiles.svg), (2) [Wilkinson 1990](https://projects.ncsu.edu/project/evoresources/Evolutionary%20medicine/Unitsofselection%20DM%20copy/wilkinson-foodsharing.pdf), (3) [Isaacdm](https://commons.wikimedia.org/wiki/File:PhysicsofPenalty.JPG?uselang=en-gb), (4) [Public license](http://www.publicdomainpictures.net/view-image.php?image=180560&picture=brexit-referendum-uk)
 
@@
 </center>
+
+# Genetic algorithms, and an example using Prisoner's dilemma
 
@@
 
+# Genetic algorithms in Management Strategy Evaluation
+
 ## {.smaller}
@@
 
+# Rise of the geese: Example use of GMSE
+
+## {.smaller}
+
+<div class = "columns-2">
+
+- Wild geese declined starting in the 1930s, likely caused by habitat loss and increased hunting
+- Refuges were established and hunting restricted in the 1950s, leading to recovery
+
+<center>
+
+![](barnacle_goose.JPG)
+
+**Image:** [Raju Kasambe](https://commons.wikimedia.org/wiki/File:Barnacle_Geese_Branta_leucopsis_London_UK_(4).JPG)
+
+</center>
+
+</div>
+
+<br>
+
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Fox & Madsen 2017 *Ambio* **46(s2)**:170-187](http://doi.org/10.1007/s13280-016-0878-2)  
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Fox & Abraham 2017 *Ambio* **46(s2)**:188-197](http://doi.org/10.1007/s13280-016-0879-1)
+
+## {.smaller}
+
+<div class = "columns-2">
+
+- Wild geese declined starting in the 1930s, likely caused by habitat loss and increased hunting
+- Refuges were established and hunting restricted in the 1950s, leading to recovery
+- Since the 1960s, many populations have exponentially increased
+- Population growth coincides with a switch from natural wetlands to agricultural landscapes
+- Switch to agricultural exploitation causes conflict with farmers
+
+<center>
+
+![](goose_graph.png)
+
+</center>
+
+</div>
+
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Fox & Madsen 2017 *Ambio* **46(s2)**:170-187](http://doi.org/10.1007/s13280-016-0878-2)  
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Fox & Abraham 2017 *Ambio* **46(s2)**:188-197](http://doi.org/10.1007/s13280-016-0879-1)
+
 ## Running the gmse() function {.smaller}
@@
 
-##
-
-<center>
-
-```{r, echo = FALSE, fig.height = 6, fig.width = 6}
+## {.smaller}
+
+<center>
+
+```{r, echo = FALSE, fig.height = 6.5, fig.width = 6}
 plot_gmse_effort(agents = sim$agents, paras = sim$paras, ACTION = sim$action, 
@@
 
-## Geese example
-
-- Wild geese declined starting in the 1930s, likely caused by habitat loss and increased hunting
-- Refuges were established and hunting restricted in the 1950s, leading to recovery
-- Since the 1960s, many populations have exponentially increased
-- Population growth coincides with a switch from natural wetlands to agricultural landscapes
-- Switch to agricultural exploitation causes conflict with farmers
-
-[Fox & Madsen 2017 *Ambio* **46(s2)**:170-187](http://doi.org/10.1007/s13280-016-0878-2)
-[Fox & Abraham 2017 *Ambio* **46(s2)**:188-197](http://doi.org/10.1007/s13280-016-0879-1)
-
 ## Future software development

### 2017-09-18 notebook/presentation/GMSE_presentation.Rmd
Clean first draft

@@
 
+# [GMSE GUI](https://bradduthie.shinyapps.io/gmse_gui/)
+
 ## Future software development
+
+<div class = "columns-2">
+
+- Modular `gmse_apply()` function allowing user-supplied natural resource, observation, manager, and user models
+- Stuctured populations (e.g., sex, stage)
+- Multiple populations of interacting species
+- Improved strategy prediction for managers and users
+- Interactive decision-making with software users
+
+<br><br>
+
+<center>
+
+![](GMSE_Logo_Goose_375.png)
+
+</center>
+
+</div>
+
+## {.smaller}
+
+<br>
+<center>
+
+![](acknowledgements.png)
+
+</center>

### 2017-09-18 notebook/presentation/GMSE_presentation.Rmd
Revised slides

@@
 
-![](adaptive_management.png)
+![](African_Bush_Elephant.jpg)
+<!--- ![](adaptive_management.png) --->
+
+**Image:** [M. M. Karim](https://en.wikipedia.org/wiki/User:Muhammad_Mahdi_Karim)
 
@@
 
-- What leads to conservation conflict
-- How to resolve conservation conflict
-- Impact on sustainability of social-ecological systems
+- What leads to conservation conflict?
+- How to resolve conservation conflict?
+- Impact on sustainability of social-ecological systems?
 
@@
 
+<br><br><br><br><br><br><br><br>
+
 **Image:** [Public Domain](https://commons.wikimedia.org/wiki/File:Fernglas(alt).JPG)
@@
 
-<br>
-
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Fox & Madsen 2017 *Ambio* **46(s2)**:170-187](http://doi.org/10.1007/s13280-016-0878-2)  
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Fox & Abraham 2017 *Ambio* **46(s2)**:188-197](http://doi.org/10.1007/s13280-016-0879-1)
+<br><br>
+
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Fox & Madsen 2017 *Ambio* **46(s2)**:170-187](http://doi.org/10.1007/s13280-016-0878-2)  
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Fox & Abraham 2017 *Ambio* **46(s2)**:188-197](http://doi.org/10.1007/s13280-016-0879-1)
 
@@
 
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Fox & Madsen 2017 *Ambio* **46(s2)**:170-187](http://doi.org/10.1007/s13280-016-0878-2)  
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Fox & Abraham 2017 *Ambio* **46(s2)**:188-197](http://doi.org/10.1007/s13280-016-0879-1)
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Fox & Madsen 2017 *Ambio* **46(s2)**:170-187](http://doi.org/10.1007/s13280-016-0878-2)  
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Fox & Abraham 2017 *Ambio* **46(s2)**:188-197](http://doi.org/10.1007/s13280-016-0879-1)
 
@@
 ```{r, echo = TRUE}
-sim <- gmse(land_ownership = TRUE, stakeholders = 10, observe_type = 1,
-            res_death_K = 2000, manage_target = 1000, RESOURCE_ini = 1000,
-            user_budget = 5000, manager_budget = 5000, res_consume = 1,
-            scaring = TRUE, fixed_mark = 50, fixed_recapt = 300,
-            plotting = FALSE);
+sim <- gmse(land_ownership = TRUE, stakeholders = 10, RESOURCE_ini = 1000,
+            res_death_K = 2000, manage_target = 1000, res_consume = 1,
+            user_budget = 5000, manager_budget = 5000, plotting = FALSE,
+            scaring = TRUE, culling = TRUE, observe_type = 1, fixed_mark = 50, 
+            fixed_recapt = 300);
 ```
@@
 
-<br><br>
-
 <center>
@@
 </center>
+
+##
+
+<br><br><br>
+
+<div class = "columns-2">
+
+<center>
+
+![](confoobio_sans.png)
+
+![](GMSE_Logo_Goose_340.png)
+
+</center>
+
+</div>

### 2017-09-18 notebook/presentation/GMSE_presentation.Rmd
Minor tweaks for shiny

@@
 
- - **Asymmetric payoffs**: Payoffs of choices may depend instead on stakeholder identity.
+ - **Asymmetric payoffs**: Payoffs of choices may depend instead on stake-holder identity.
 
@@
 
- - **Asymmetric payoffs**: Payoffs of choices may depend instead on stakeholder identity.
+ - **Asymmetric payoffs**: Payoffs of choices may depend instead on stake-holder identity.
   - **Multiple choices**: Players might have three or more choices.
@@
 
- - **Asymmetric payoffs**: Payoffs of choices may depend instead on stakeholder identity.
+ - **Asymmetric payoffs**: Payoffs of choices may depend instead on stake-holder identity.
  - **Multiple choices**: Players might have three or more choices.
- - **Multiple players**: More than two stakeholders might interact at one time.
+ - **Multiple players**: More than two stake-holders might interact at one time.
 
@@
 
- - **Asymmetric payoffs**: Payoffs of choices may depend instead on stakeholder identity.
+ - **Asymmetric payoffs**: Payoffs of choices may depend instead on stake-holder identity.
  - **Multiple choices**: Players might have three or more choices.
- - **Multiple players**: More than two stakeholders might interact at one time.
+ - **Multiple players**: More than two stake-holders might interact at one time.
  - **Iterated interactions**: Strategies maybe affected by interaction history.
@@
 
-## Genearlised MSE {.smaller}
+## Generalised MSE {.smaller}
 
@@
 - Integrates game theory and ecological theory to construct social-ecological models for adaptive management
-- Simulates all aspects of management: population dynamics, manager observation of populations, manager decision making, and stakeholder responses to management decisions
-- Mimic the decision-making process of managers and stakeholders under conditions of change, uncertainty, and conflict
+- Simulates all aspects of management: population dynamics, manager observation of populations, manager decision making, and stake-holder responses to management decisions
+- Mimic the decision-making process of managers and stake-holders under conditions of change, uncertainty, and conflict
 - Provide a flexible modelling framework for future development of complex adaptive management models to be applied to conservation theory and specific case studies
@@
 
-## Genearlised MSE ('GMSE' R package)
+## Generalised MSE ('GMSE' R package)
 
@@
 - Fixed budget and a costs for setting policy
-- Genetic algorithm anticipates policy effects from previous actions to set costs of actions for stakeholders
+- Genetic algorithm anticipates policy effects from previous actions to set costs of actions for stake-holders
 
@@
             res_death_K = 2000, manage_target = 1000, res_consume = 1,
-            user_budget = 5000, manager_budget = 5000, plotting = FALSE,
+            manager_budget = 5000, user_budget = 5000, plotting = FALSE,
             scaring = TRUE, culling = TRUE, observe_type = 1, fixed_mark = 50, 
@@
 
-**Plot of conflict among stakeholders**
+**Plot of conflict among stake-holders**
 

### 2017-09-22 notebook/presentation/GMSE_presentation.Rmd
Some additional presentation stuff

@@
 
-# Adaptive management and conservation conflicts
-
-## Process of adaptive management
-
-<div class = "columns-2">
-
-- Uses an iterative process of decision-making
-- Sets explicit management goals
-- Develops alternative management strategies
-- Deals with uncertainty by updating knowledge of the system
-
-<br>
-
-![](African_Bush_Elephant.jpg)
-<!--- ![](adaptive_management.png) --->
-
-**Image:** [M. M. Karim](https://en.wikipedia.org/wiki/User:Muhammad_Mahdi_Karim)
-
-</div>
-
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Keith et al. (2011) *Biol. Conserv.* **144**: 1175-1778](https://doi.org/10.1016/j.biocon.2010.11.022)
+# Uncertainty and conservation conflict
 
@@
 
-<br>
-
-**Conservation conflicts are complex, and characterised by incomplete information that is rapidly changing.**
+**Conservation conflicts between biodiversity and food security result when the outcome of one stake-holder is affected by one or more other stake-holders** ([Colyvan et al.
+(2011) *Biol.
+Conserv.* **144**:1246-1253](http://www.sciencedirect.com/science/article/pii/S000632071000488X?via%3Dihub))
 
@@
 </div>
-
+<br><br>
 <center>
@@
 </div>
+
+<br><br>
 
@@
 
-## {.smaller}
-
-<center>
-
-![](gmse_overview_sm_ga2.png)
-
-</center>
-
-##
-
-<center>
-
-![](gmse_overview_sm.png)
-
-</center>
-
 ## Generalised MSE {.smaller}
@@
 - Five possible policy options (scaring, culling, castration, feeding, helping)
-- Fixed budget and a costs for setting policy
+- Fixed budget and costs for setting policy
 - Genetic algorithm anticipates policy effects from previous actions to set costs of actions for stake-holders
@@
 
-# Rise of the geese: Example use of GMSE
-
-## {.smaller}
-
-<div class = "columns-2">
-
-- Wild geese declined starting in the 1930s, likely caused by habitat loss and increased hunting
-- Refuges were established and hunting restricted in the 1950s, leading to recovery
-
-<center>
-
-![](barnacle_goose.JPG)
-
-**Image:** [Raju Kasambe](https://commons.wikimedia.org/wiki/File:Barnacle_Geese_Branta_leucopsis_London_UK_(4).JPG)
-
-</center>
-
-</div>
-
-<br><br>
-
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Fox & Madsen 2017 *Ambio* **46(s2)**:170-187](http://doi.org/10.1007/s13280-016-0878-2)  
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Fox & Abraham 2017 *Ambio* **46(s2)**:188-197](http://doi.org/10.1007/s13280-016-0879-1)
-
-## {.smaller}
-
-<div class = "columns-2">
-
-- Wild geese declined starting in the 1930s, likely caused by habitat loss and increased hunting
-- Refuges were established and hunting restricted in the 1950s, leading to recovery
-- Since the 1960s, many populations have exponentially increased
-- Population growth coincides with a switch from natural wetlands to agricultural landscapes
-- Switch to agricultural exploitation causes conflict with farmers
-
-<center>
-
-![](goose_graph.png)
-
-</center>
-
-</div>
-
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Fox & Madsen 2017 *Ambio* **46(s2)**:170-187](http://doi.org/10.1007/s13280-016-0878-2)  
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Fox & Abraham 2017 *Ambio* **46(s2)**:188-197](http://doi.org/10.1007/s13280-016-0879-1)
+# Example use of GMSE
 
@@
 
+## Use of GMSE: Model case studies
+
+<div class = "columns-2">
+
+<center>
+
+![](African_Bush_Elephant.jpg)  
+**Image:** [M. M. Karim](https://en.wikipedia.org/wiki/User:Muhammad_Mahdi_Karim)
+
+![](barnacle_goose.JPG)  
+**Image:** [Raju Kasambe](https://commons.wikimedia.org/wiki/File:Barnacle_Geese_Branta_leucopsis_London_UK_(4).JPG)
+
+</center>
+
+</div>
+
+## Use of GMSE: Theoretical questions
+
+<center>
+
+![](man_freq.png)
+
+</center>
+
 ## Future software development

### 2017-10-16 R/gmse.R
Tweaks

@@ -137,6 +137,9 @@ gmse <- function( time_max       = 100,   # Max number of time steps in sim
     if(user_budget > 10000 | manager_budget > 10000){
         stop("User and manager budgets cannot exceed 10000");
     }
+    if(user_budget < 1 | manager_budget < 1){
+        stop("User and manager budgets must be at least 1");
+    }
     
     user_res_opts  <- c(scaring, culling, castration, feeding, help_offspring);
     user_lnd_opts  <- c(tend_crops, kill_crops);

### 2017-10-18 notebook/ms/ms.Rmd
Better parameter simulations, now that GMSE has been updated

--- 2017-08-09 notebook/ms/ms.Rmd
@@
 ```{r, echo = TRUE}
-sim <- gmse(land_ownership = TRUE, stakeholders = 10, observe_type = 0, 
+sim <- gmse(land_ownership = TRUE, stakeholders = 5, observe_type = 0, 
             res_death_K = 2000, manage_target = 1000, RESOURCE_ini = 1000, 
-            user_budget = 5000, manager_budget = 5000, res_consume = 1, 
-            scaring = TRUE, times_observe = 1, plotting = FALSE);
+            user_budget = 1000, manager_budget = 1000, res_consume = 1, 
+            scaring = TRUE, plotting = FALSE);
 ```

### 2017-10-18 notebook/ms/ms.Rmd
Change table size for easier reading

@@
 output:
+  html_document: default
   pdf_document:
     fig_caption: yes
-  html_document: default
   word_document:
@@
 
+\small
+
 |  Argument       | Default | Description                                                                                      |
@@
 |  public_land    |  0      | Proportion of land that is public (un-owned by stakeholders) if land_ownership = TRUE            |
+
+\normalsize
 

### 2017-10-18 notebook/ms/ms.Rmd
Add sentence fragment to include gmse_apply

@@
 We present the R package GMSE, which generalises MSE using a game-theoretic approach to simulate adaptive decision-making management scenarios between stakeholders with competing objectives under complex social-ecological interactions and uncertainty.
-3. GMSE is agent-based and spatially explicit, and incorporates a high degree of realism through mechanistic modelling of links and feedbacks among stakeholders and with the ecosystem.
+3.
+GMSE models can be agent-based and spatially explicit, incorporating a high degree of realism through mechanistic modelling of links and feedbacks among stakeholders and with the ecosystem; additionally, user-defined sub-models can also be incorporated as functions into the broader GMSE framework.
 4.

### 2017-10-18 notebook/ms/ms.Rmd
Minor edits to Introduction

@@
 Conservation of biodiversity can be complicated by the immediate need to use natural resources and land area for human livelihood (e.g., food production), causing real or perceived conflicts between conservation and food security and creating a challenge for the management of many natural resources [@Redpath2015a].
-Given an increasing human population [@Crist2017], the number and intensity of such conflicts are likely to increase into the twenty first century.
+Given increasing human population size [@Crist2017], the number and intensity of such conflicts are likely to increase into the twenty first century.
 Effective management tools are therefore needed for the long-term sustainable use of natural resources under the rising demand for food production [@Fischer2017].
@@
 
-The GMSE R package is a flexible, highly mechanistic, agent-based modelling tool to simulate all key aspects of natural resource management.
-GMSE considers a range of parameters to simulate resource dynamics and management policy options, and uses genetic algorithms to dynamically model stakeholder (manager and user) decision-making.
+The R package 'GMSE' is a flexible modelling tool to simulate all key aspects of natural resource management.
+GMSE offers a range of parameters to simulate resource dynamics and management policy options, and includes genetic algorithms to dynamically model stakeholder (manager and user) decision-making.
 Genetic algorithms find adaptive solutions to any simulated conditions given stakeholder-specific goals, allowing GMSE to model scenarios of conservation conflict.

### 2017-10-18 notebook/ms/ms.Rmd
Update to manuscript, including new function mentions

@@
 4.
-We show how GMSE simulates a social-ecological system using the example of a waterfowl population in an agricultural landscape that is adaptively managed; simulated waterfowl exploit agricultural land, causing conflict between conservation interests and the interest of food producers maximising their crop yield.
+We show how GMSE simulates a social-ecological system using the example of an adaptively managed waterfowl population on an agricultural landscape; simulated waterfowl exploit agricultural land, causing conflict between conservation interests and the interest of food producers maximising their crop yield.
 5. The R package GMSE is open source under GNU Public License; source code and documents are freely available on GitHub.
@@
 
-GMSE builds off of the MSE framework, which includes four sub-models, each of which runs once in a single time step of the broader model (Figure 1).
+GMSE builds off of the MSE framework (Figure 1), and the `gmse` function runs simulations using four default submodels.
 (1) A population of discrete resources with individual traits (e.g., location, age) is modelled on a spatially-explicit landscape and can simulate resource birth, movement, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty.
 This sub-model is unique in not relying on other sub-models because ecological dynamics can be simulated in the absence of observation and management.
-(2) Observation is modelled in one of four ways: resource counting on a subset of landscape cells [e.g., @Nuno2013], marking and recapturing a fixed number of resources, and resource counting on the whole landscape either one linear transect or one rectangular block at a time (during which resources might move).
-Sampling error from all of these mechanisms of observation generates a range of uncertainties depending on monitoring effort.
+(2) Observation is modelled in one of four ways: resource counting on a subset of landscape cells [e.g., @Nuno2013], marking and recapturing a fixed number of resources, and resource counting across the whole landscape either one linear transect or one rectangular block at a time (during which resources might move).
+Sampling error from all observation types generates a range of uncertainties that depend on monitoring effort.
 (3) Managers analyse data collected from observations to estimate resource abundance, then compare this estimate with their pre-defined target abundance.
 Policy is developed by calling the genetic algorithm (see below), which works within a manager's constraints to find costs for user actions on the resource (e.g., culling, scaring, etc.) that minimise deviation from the target abundance, as informed by the predicted consequences of each action on resource abundance and user action histories.
-After a suitable policy is found, (4) users can perform actions that affect resources or landscape cells.
+After a suitable policy is found, (4) users perform actions that affect resources or landscape cells.
 Users respond to policy individually, each calling the genetic algorithm to find actions that maximise their own utilities (e.g., maximise resource use or landscape yield) within their imposed constraints.
@@
 
-Simulations are run using the gmse() function, which offers a range of options for setting parameter values (see Table 1 for some select examples).
-Output of gmse() is an exhaustive list that includes all resources and observations, all stakeholder decisions and actions, and all landscape properties in each time step of the simulation.
-Results are most easily interpreted visually, so a summary of simulation dynamics is plotted by default (the plot can also be called using the plot_gmse_results function).
+Simulations using the default GMSE sub-models described above are run using the `gmse` function, which offers a range of options for setting parameter values (see Table 1 for some select examples).
+Output of `gmse` is an exhaustive list that includes all resources and observations, all stakeholder decisions and actions, and all landscape properties in each time step of the simulation.
+Results are most easily interpreted visually, so a summary of simulation dynamics is plotted by default (the plot can also be called using the `plot_gmse_results` function, and a summary of results can be obtained using `gmse_summary`).
 An example below shows how simulations are set and interpreted.
@@
 In each time step, waterfowl are observed on subset of cells, then managers extrapolate from density per cell to estimate total population size.
-Managers then use these estimates to set costs of culling and scaring (non-lethal) waterfowl for 10 farmers.
+Managers then use these estimates to set costs of culling and scaring (non-lethal) waterfowl for five farmers.
 Farmers attempt to reduce the negative impact of waterfowl on the cropland that they own, working within the constraints of culling and scaring costs and their budget for performing these actions.
@@
 
-Parameters in gmse() not listed are set to default values. By plotting the output with plot_gmse_results, simulation results can be interpreted visually.
+Parameters in `gmse` not listed are set to default values.
+By plotting the output with `plot_gmse_results`, simulation results can be interpreted visually (manager and user decisions can also be interpreted using the `plot_gmse_effort` function).
 
@@
 
-```{r, echo = FALSE, fig.height = 6, fig.align = 'left', fig.cap = "Results of an example simulation illustrating the management of a protected resource that exploits the land of 10 stakeholders.
-The upper left panel shows locations of resources (black dots) on the landscape in the final time step of the simulation.
-The upper right panel shows the same landscape broken down into 10 differently coloured regions, which correspond to areas of land owned by each of the 10 stakeholders.
-The middle left panel shows the actual abundance of resources (black solid line, and the abundance of resources as estimated by the manager (blue solid line; shading indicates 95 percent confidence intervals of a mark-recapture analysis), over time.
+```{r, echo = FALSE, fig.height = 6, fig.align = 'left', fig.cap = "Results of an example simulation illustrating the management of a protected resource that exploits the land of five farmers The upper left panel shows locations of resources (black dots) on the landscape in the final time step of the simulation.
+The upper right panel shows the same landscape broken down into five differently coloured regions, which correspond to areas of land owned by each of the five farmers.
+The middle left panel shows the actual abundance of resources (black solid line), and the abundance of resources as estimated by the manager (blue solid line; shading indicates 95 percent confidence intervals), over time.
 The horizontal dotted red and blue lines show the resource carrying capacity enacted on adult mortality and the manager's target for resource abundance, respectively.
@@
 The middle right panel shows total percent yield of landscape cells for each individual farmer, differentiated by colour, where line colours correspond to areas of the landscape in the upper right panel.
-The lower left panel shows the cost of stakeholders performing actions over time, as set by the manager.
-The lower right panel shows the total number of actions attempted to be performed by all stakeholders over time (some actions might be unsuccessful if resources are unavailable on a stakeholder's land to cull or scare, so, e.g., culling actions might be larger than resources actually culled)."}
+The lower left panel shows the cost of farmers performing actions over time, as set by the manager.
+The lower right panel shows the total number of actions attempted to be performed by all farmers over time (some actions might be unsuccessful if resources are unavailable on a farmer's land to cull or scare, so, e.g., culling actions might be larger than resources actually culled)."}
 plot_gmse_results_vign(res = sim$resource, obs = sim$observation, 
@@
 Hence, when the population size is estimated to be below (above) the manager's target, the manager increases (decreases) the cost of culling and decreases (increases) the cost of scaring.
-Because the manager does not know in advance how stakeholders will react to policy change, they assume a proportional response in total actions with respect to a change in cost (e.g., doubling the cost of culling will decrease stakeholder culling by $1/2$).
+Because the manager does not know in advance how farmers will react to policy change, they assume a proportional response in total actions with respect to a change in cost (e.g., doubling the cost of culling will decrease stakeholder culling by $1/2$).
 Farmers responding to policy are interested only in minimising waterfowl's exploitation of their crops, so they will either cull or scare to remove the waterfowl from their land, depending on which option is more effective (i.e., cheaper).
@@
 
-Future development
+Graphical User Interface (GUI)
 ===================================================
 
-The GMSE package is under continued development to include additional features that will be of interest to conservation biologists, managers, and the general public.
-Such features will include multiple (interacting) resource types and sub-types (e.g., structured populations), improved stakeholder decision-making and prediction based on multiple time steps of simulation history, incorporation of empirical data (e.g., landscape features), and a browser-based graphical user interface.
-Additionally, future versions of GMSE will allow software users to take the place of stakeholders in decision-making during simulations (replacing the genetic algorithm as desired), facilitating the collection of data to test sociological hypotheses and further improve the realism of the genetic algorithm (for a preliminary example of this, see the 'hunt' argument in the gmse() function).
-Code underlying GMSE is publicly available on GitHub < https://github.com/bradduthie/gmse > and highly modular, so GMSE model components (black boxes in Figure 1) can be developed freely and independently, then integrated into the broader GMSE framework.
-The GMSE R package is therefore a versatile and collaborative tool with widespread applications for adaptive resource management.
+The function `gmse_gui` opens GMSE in a browser and allows simulations to be run for most `gmse` parameter options using the package 'shiny' [@Chang2017].
+Figures from `plot_gmse_results` and `plot_gmse_effort`, and tables from `gmse_summary` are provided as GUI output.
+
+Custom defined sub-models
+===================================================
+
+The function `gmse_apply` allows custom resource, observation, manager, or user sub-models to be integrated into the GMSE framework.
+Any type of sub-model (e.g., numerical, individual-based) is permitted by defining a function with appropriately specified inputs and outputs; where custom functions are not provided, `gmse_apply` runs default GMSE sub-models used in `gmse`.
+Any parameter options available in `gmse` or in custom functions can be passed directly to `gmse_apply`, thereby allowing for high flexibility in model specification.
+For example, a simple logistic growth function can be integrated as a resource sub-model to replace the default `resource` function.
+
+```{r, echo = TRUE}
+logistic_res_mod <- function(X_0, K = 2000, gr = 1){
+    X_1 <- X_0 + gr * X_0 * ( 1 - X_0/K );
+    return(X_1);
+}
+sim <- gmse_apply(res_mod = logistic_res_mod, X_0 = 1000, gr = 0.3, stakeholders = 5);
+```
+
+The `gmse_apply` function simulates a single GMSE time step, and therefore must be looped for simulations over multiple time steps.
+
+Availability
+===================================================
+
+The GMSE package can be downloaded from CRAN (https://cran.r-project.org/package=GMSE) or GitHub (https://github.com/bradduthie/gmse). GMSE is open source under GNU Public License.
 

### 2017-10-19 notebook/ms/ms.Rmd
Minor changes to fix references

@@
   pdf_document:
+    citation_package: natbib
     fig_caption: yes
@@
     reference_docx: docx_template.docx
-linestretch: 1
 header-includes:
-- \usepackage{natbib}
 - \usepackage{lineno}
 - \linenumbers
-- \bibliographystyle{amnatnat}
+linestretch: 1
 bibliography: gmse.bib
+biblio-style: apalike
 ---

### 2017-10-19 src/resource.c
Fix the resource move error, which was causing resources to not move among tme steps because the resource move during observation paramater was set to false

@@ -301,7 +301,6 @@ SEXP resource(SEXP RESOURCE, SEXP LANDSCAPE, SEXP PARAMETERS){
     int res_num_total;       /* Total number of resources in returned array */
     int protected_n;         /* Number of protected R objects */
     int vec_pos;             /* Vector position for making arrays */
-    int move_res;            /* Should resources be allowed to move */
     int off_col;             /* The column where the offspring are held */
     int rm_col;              /* Column where removal is indiciated */
     int *dim_RESOURCE;       /* Dimensions of the RESOURCE array incoming */
@@ -379,7 +378,6 @@ SEXP resource(SEXP RESOURCE, SEXP LANDSCAPE, SEXP PARAMETERS){
     /* Do the biology here now */
     /* ====================================================================== */
     
-    move_res  = (int) paras[19]; /* Should the resources be moved?        */
     off_col   = (int) paras[38];
     rm_col    = (int) paras[43];
     
@@ -387,9 +385,7 @@ SEXP resource(SEXP RESOURCE, SEXP LANDSCAPE, SEXP PARAMETERS){
     add_time(res_old, paras);
     
     /* Resources move according to move function and parameter) */
-    if(move_res == 1){
-        res_mover(res_old, land, paras);
-    }
+    res_mover(res_old, land, paras);
 
     /* Identify, and calculate the number of, added individuals */
     res_add(res_old, paras);

### 2017-10-19 src/resource.c
Add option in switch function for both density independent and dependent removal of resources

@@ -201,6 +201,32 @@ void res_remove(double **res_removing, double *paras){
                 }
             }
             break;
+        case 3:
+            for(resource = 0; resource < resource_number; resource++){
+                rm_odds   = res_removing[resource][rm_row];
+                rand_unif = runif(0, 1);
+                if(rand_unif < rm_odds){
+                    res_removing[resource][rm_row] = -1;   
+                }
+            }
+            over_K  = resource_number - K;
+            if(over_K > 0){
+                rm_from_K  = (double) over_K / resource_number;
+                for(resource = 0; resource < resource_number; resource++){
+                    rand_unif   = runif(0, 1);
+                    if(rand_unif < rm_from_K){
+                        res_removing[resource][rm_row] = -1;   
+                    }
+                }
+            }
+            for(resource = 0; resource < resource_number; resource++){
+                rm_from_Ind = res_removing[resource][rm_adj];
+                rand_unif   = runif(0, 1);
+                if(rand_unif < rm_from_Ind){
+                    res_removing[resource][rm_row] = -1;   
+                }
+            }
+            break;
         default:
             over_K  = resource_number - K;
             if(over_K > 0){

### 2017-10-19 notebook/ms/ms.Rmd
Update ms
Draft tweaks

@@
   html_document: default
-  word_document:
-    fig_caption: yes
-    reference_docx: docx_template.docx
 header-includes:
@@
 
-GMSE builds off of the MSE framework (Figure 1), and the `gmse` function runs simulations using four default submodels.
+GMSE builds off of the MSE framework (Figure 1), and the primary function `gmse` runs simulations using four default submodels.
 (1) A population of discrete resources with individual traits (e.g., location, age) is modelled on a spatially-explicit landscape and can simulate resource birth, movement, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty.
@@
 |  res_birth_K    |  10000  | Carrying capacity applied to the number of resources added during a time step                    |
-|  res_death_K    |  600    | Carrying capacity applied to the number of resources removed during a time step                  |
+|  res_death_K    |  2000   | Carrying capacity applied to the number of resources removed during a time step                  |
 |  res_move_type  |  1      | Type of resource movement (default is up to res_movement cells in any direction)                 |
@@
 |  fixed_recapt   |  150    | For mark-recapture observation (observe_type = 1), number of recaptured resources                |
-|  times_observe  |  8      | For density-based observation (observe_type = 0), landscape subsets viewed during observation    |
+|  times_observe  |  1      | For density-based observation (observe_type = 0), landscape subsets viewed during observation    |
 |  res_consume    |  0.5    | Proportion of a landscape cell's value reduced by the presence of a resource in a time step      |
@@
 |  manager_budget |  1000   | A manager's budget per time step for setting policy                                              |
-|  manage_target  |  300    | The manager's target resource abundance                                                          |
-|  RESOURCE_ini   |  300    | The initial abundance of resources                                                               |
+|  manage_target  |  1000   | The manager's target resource abundance                                                          |
+|  RESOURCE_ini   |  1000   | The initial abundance of resources                                                               |
 |  scaring        |  FALSE  | Resource scaring is a policy option                                                              |
@@
 
-```{r, echo = FALSE, fig.height = 6, fig.align = 'left', fig.cap = "Results of an example simulation illustrating the management of a protected resource that exploits the land of five farmers The upper left panel shows locations of resources (black dots) on the landscape in the final time step of the simulation.
+```{r, echo = FALSE, fig.height = 6, fig.align = 'left', fig.cap = "Results of an example simulation illustrating the management of a protected resource that exploits the land of five farmers.
+The upper left panel shows locations of resources (black dots) on the landscape in the final time step of the simulation.
 The upper right panel shows the same landscape broken down into five differently coloured regions, which correspond to areas of land owned by each of the five farmers.

### 2017-10-23 src/manager.c
Fix under-estimate caused by sampling more than the available landscape

@@ -75,6 +75,10 @@ void dens_est(double **obs_array, double *paras, double **agent_array,
     cells   = land_x * land_y; /* Plus one needed for zero index */
     tot_obs = 0;
     
+    if(area > cells){
+        area = cells;
+    }
+    
     for(resource = 0; resource < int_table_rows; resource++){
         abun_est[resource] = 0;
         if(interact_table[resource][0] == 0){ /* Change when turn off type? */

### 2017-10-23 notebook/ms/ms.Rmd
Minor tweaks

@@
 output:
-  word_document:
-    fig_caption: yes
-    reference_docx: docx_template.docx
   pdf_document:
@@
   html_document: default
+  word_document:
+    fig_caption: yes
+    reference_docx: docx_template.docx
 header-includes:

### 2017-10-23 R/gmse_summary.R
Try and retry summary, fixed now

@@ -51,7 +51,11 @@ gmse_summary <- function(gmse_results){
                                     manager_acts[,2] == res_types[j]);
             target <- manager_acts[target_row, 5];
             adjusr <- manager_acts[estim_row,  5];
-            observations[i,j+1] <- target - adjusr;
+            if(adjusr > 0){
+                observations[i,j+1] <- target - adjusr;
+            }else{
+                observations[i,j+1] <- target - adjusr;
+            }
             #---- Cost setting below
             costs[c_row, 1]  <- i;
             costs[c_row, 2]  <- res_types[j];

### 2017-11-02 notebook/ms/ms.Rmd
Incorporate Erlend's comments into the Introduction

@@
 2.
-We present the R package GMSE, which generalises MSE using a game-theoretic approach to simulate adaptive decision-making management scenarios between stakeholders with competing objectives under complex social-ecological interactions and uncertainty.
+We present the R package GMSE, which generalises MSE using a game-theoretic approach, applying genetic algorithms to simulate adaptive decision-making management scenarios between stakeholders with competing objectives under complex social-ecological interactions and uncertainty.
 3.
@@
 Many global natural resources, including the biodiversity on which critical ecosystem services depend, are in a state of severe decline [@Dirzo2014; @Hautier2015; @Ceballos2017; @OConnell2017].
-Conservation of biodiversity can be complicated by the immediate need to use natural resources and land area for human livelihood (e.g., food production), causing real or perceived conflicts between conservation and food security and creating a challenge for the management of many natural resources [@Redpath2015a].
+Conservation of biodiversity can be complicated by the immediate need to use natural resources and land area for human livelihood (e.g., food production), causing real or perceived conflicts between biodiversity conservation and food security.
+This creates a challenging situation for the management of many natural resources [@Redpath2015a].
 Given increasing human population size [@Crist2017], the number and intensity of such conflicts are likely to increase into the twenty first century.
@@
 
-To effectively manage natural resources, an adaptive approach allows managers to iteratively update their models of resource dynamics and respond flexibly to changing conditions [@Keith2011].
+To effectively manage natural resources, an adaptive approach allows managers to iteratively update their models and respond flexibly to changing conditions [@Keith2011].
 This approach is especially effective when considering all aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of stakeholders [@Bunnefeld2011; @Bunnefeld2014].
 Management strategy evaluation (MSE) is a modelling framework, first developed in fisheries, for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016].
-Nevertheless, MSE remains limited in its ability to model human decision-making [@Fulton2011a; @Dichmont2017]; manager decisions are typically based on fixed rules, and user behaviour likewise remains fixed over time instead of dynamically responding to changing resource availability and management decisions [@Schluter2012; @Melbourne-Thomas2017].
+Nevertheless, MSE models developed hitherto have been limited in their ability to model human decision-making [@Fulton2011a; @Dichmont2017]; manager decisions are typically based on fixed rules, and user behaviour likewise remains fixed over time instead of dynamically responding to changing resource availability and management decisions [@Schluter2012; @Melbourne-Thomas2017].
 Here we introduce generalised management strategy evaluation (GMSE), which incorporates a game-theoretic perspective to model the goal-oriented, dynamic decision-making processes of stakeholders.

### 2017-11-02 notebook/ms/ms.Rmd
Get references cleaner

@@
 
-Many global natural resources, including the biodiversity on which critical ecosystem services depend, are in a state of severe decline [@Dirzo2014; @Hautier2015; @Ceballos2017; @OConnell2017].
+Many global natural resources, including the biodiversity on which critical ecosystem services depend, are in a state of severe decline [@Dirzo2014; @Hautier2015; @Ceballos2017a; @OConnell2017].
 Conservation of biodiversity can be complicated by the immediate need to use natural resources and land area for human livelihood (e.g., food production), causing real or perceived conflicts between biodiversity conservation and food security.
@@
 Management strategy evaluation (MSE) is a modelling framework, first developed in fisheries, for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016].
-Nevertheless, MSE models developed hitherto have been limited in their ability to model human decision-making [@Fulton2011a; @Dichmont2017]; manager decisions are typically based on fixed rules, and user behaviour likewise remains fixed over time instead of dynamically responding to changing resource availability and management decisions [@Schluter2012; @Melbourne-Thomas2017].
+Nevertheless, MSE models developed hitherto have been limited in their ability to model human decision-making [@Fulton2011; @Dichmont2017]; manager decisions are typically based on fixed rules, and user behaviour likewise remains fixed over time instead of dynamically responding to changing resource availability and management decisions [@Schluter2012; @Melbourne-Thomas2017].
 Here we introduce generalised management strategy evaluation (GMSE), which incorporates a game-theoretic perspective to model the goal-oriented, dynamic decision-making processes of stakeholders.

### 2017-11-02 notebook/ms/ms.Rmd
Comments from @ninbrm and @ErlendNilsen included

@@
 
-GMSE builds off of the MSE framework (Figure 1), and the primary function `gmse` runs simulations using four default submodels.
+GMSE builds off of the MSE framework (Figure 1).
+The workhorse function `gmse` runs simulations using four predefined submodels, which can be parameterised to fit various case studies; more tailored submodels can also be defined using the `gmse_apply` function (see below).
 (1) A population of discrete resources with individual traits (e.g., location, age) is modelled on a spatially-explicit landscape and can simulate resource birth, movement, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty.
@@
 Consistent with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for agents (stakeholders).
-Instead, genetic algorithms are used to heuristically find an adaptive strategy for each stakeholder in each time step.
+Instead, genetic algorithms are used to heuristically find an adaptive strategy for each stakeholder in each time step (see Supporting Information 1 for details).
 Critically, all stakeholders involved in resource conservation are constrained in their decision-making; managing and using resources takes effort (e.g., time or money), and effort expended in developing or enforcing one policy (for managers) or performing one action (for users) will be effort not expendable elsewhere [@Milner-Gulland2011; @Muller-Hansen2017; @Schluter2017].
@@
 The highest fitness strategy in the population then becomes the stakeholder's new strategy.
+Overall, a single call of the genetic algorithm thereby simulates the process of thinking and decision-making for one manager or user.
 
@@
 |  times_observe  |  1      | For density-based observation (observe_type = 0), landscape subsets viewed during observation    |
-|  res_consume    |  0.5    | Proportion of a landscape cell's value reduced by the presence of a resource in a time step      |
+|  res_consume    |  0.5    | Pr. of a landscape cell's value reduced by the presence of a resource in a time step             |
 |  max_ages       |  5      | The maximum number of time steps a resource can persist before it is removed                     |
-|  minimum_cost   |  10     | The minimum cost of a stakeholder performing any action                                          |
-|  user_budget    |  1000   | A stakeholder's budget per time step for performing any number of actions                        |
+|  minimum_cost   |  10     | The minimum cost of a user performing any action                                                 |
+|  user_budget    |  1000   | A user's budget per time step for performing any number of actions                               |
 |  manager_budget |  1000   | A manager's budget per time step for setting policy                                              |
@@
 |  help_offspring |  FALSE  | Resource helping (increases offspring number) is a policy option                                 |
-|  tend_crops     |  FALSE  | Stakeholders can increase landscape cell values                                                  |
+|  tend_crops     |  FALSE  | Users can increase landscape cell values                                                         |
 |  tend_crop_yld  |  0.2    | Proportional increase per landscape cell from tend_crops action                                  |
-|  kill_crops     |  FALSE  | Stakeholders can decrease landscape cell values to zero                                          |
-|  stakeholders   |  4      | Number of stakeholders in the simulation                                                         |
-|  land_ownership |  FALSE  | Stakeholders own land and increase utility indirectly from landscape instead of resource use     |
+|  kill_crops     |  FALSE  | Users can decrease landscape cell values to zero                                                 |
+|  stakeholders   |  4      | Number of users in the simulation                                                                |
+|  land_ownership |  FALSE  | Users own land and increase utility indirectly from landscape instead of resource use            |
 |  manage_freq    |  1      | Frequency (in time steps) with which managers revise and enact policy                            |
-|  public_land    |  0      | Proportion of land that is public (un-owned by stakeholders) if land_ownership = TRUE            |
+|  public_land    |  0      | Proportion of land that is public (un-owned by users) if land_ownership = TRUE                   |
 
@@
 
-**Table 1:** Select parameter values for initialising generalised management strategy evaluation simulations
+Table 1: Select parameter values for initialising generalised management strategy evaluation simulations
 
@@
 
-Here we consider the example of a managed natural resource whose abundance affects a group of stakeholders by temporarily decreasing the value of user land.
-This scenario could be interpreted in multiple ways; we consider a protected population of waterfowl that exploits agricultural land causing a conservation conflict with users [hereafter 'farmers', e.g., @Tuvendal2015; @Fox2016a; @Fox2017a].
-Managers in this example might want to keep the abundance of waterfowl at a target level, while farmers might want to minimise the damage inflicted on their crops [e.g., @Madsen2017].
-Using GMSE, it is possible to simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy.
+Here we illustrate the usefulness of GMSE by considering the case study of a protected population of waterfowl that exploits agricultural land causing a conservation conflict with farmers [e.g., @Tuvendal2015; @Fox2016a; @Fox2017a].
+Managers attempt to keep the abundance of waterfowl at a target level, while farmers attempt to minimise the damage inflicted on their crops [e.g., @Madsen2017].
+Using GMSE, we can simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy.
 We consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity is 2000.
@@
 The `gmse_apply` function simulates a single GMSE time step, and therefore must be looped for simulations over multiple time steps.
+Within loops, GMSE arguments can be redefined to simulate changing conditions (e.g., change in policy availability or stakeholder budgets, see Supporting Information 2).
 

### 2017-11-02 notebook/ms/ms.Rmd
New last sentence, promise of supporting information

@@
 title: 'GMSE: an R package for generalised management strategy evaluation'
-author: A. Bradley Duthie, Jeremy J. Cusack, Erlend B. Nilsen, Roc&#0237;o Pozo, O.
+author: A. Bradley Duthie, Jeremy J. Cusack, Isabel L. Jones, Erlend B. Nilsen, Roc&#0237;o Pozo, O.
   Sarobidy Rakotonarivo, Bram Van Moorter, and Nils Bunnefeld
@@
 The `gmse_apply` function simulates a single GMSE time step, and therefore must be looped for simulations over multiple time steps.
-Within loops, GMSE arguments can be redefined to simulate changing conditions (e.g., change in policy availability or stakeholder budgets, see Supporting Information 2).
+Within loops, GMSE arguments can be redefined to simulate changing conditions (e.g., change in policy availability or stakeholder budgets, see Supporting Information 2), thereby allowing many management scenarios to be simulated *in silico*.
 

### 2017-11-02 notebook/ms/SI1.Rmd
Add new SI2 to extend introduction of GMSE apply function and provide some demonstration of what it can do, as suggested by @ErlendNilsen -- last line references the supporting information so that interested readers can move straight to this as a sort of second vignette

---
title: 'GMSE: an R package for generalised management strategy evaluation'
subtitle: 'Supporting Information 2'
author: A. Bradley Duthie, Jeremy J. Cusack, Isabel L. Jones, Erlend B. Nilsen, Roc&#0237;o Pozo, O.
  Sarobidy Rakotonarivo, Bram Van Moorter, and Nils Bunnefeld
date: '`r Sys.Date()`'
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes:
- \usepackage{lineno}
- \linenumbers
linestretch: 1
---

```{r, echo = FALSE}
library(GMSE);
```

Extended introduction to the GMSE apply function (`gmse_apply`)
================================================================================

The `gmse_apply` function is a flexible function that allows for user-defined sub-functions calling resource, observation, manager, and user models. Where such models are not specified, predefined GMSE submodels 'resource', 'observation', 'manager', and 'user' are run by default. Any type of sub-model (e.g., numerical,  individual-based) is permitted as long as the input and output are appropriately specified. Only one time step is simulated per call to `gmse_apply`, so the function must be looped for simulation over time. Where model parameters are needed but not specified, defaults from GMSE are used. In this Supporting Information, we demonstrate some uses of `gmse_apply`, and how it might be used to simulate myriad management scenarios *in silico*.

A simple run of `gmse_apply()` returns one generation of GMSE using predefined submodels and default parameter values.

```{r, eval = TRUE}
sim_1 <- gmse_apply();
```

For `sim_1`, the default 'basic' results are returned as below, which summarise key values for all submodels.

```{r, eval = TRUE}
print(sim_1);
```

Note that in the case above we have the total abundance of resources returned (`sim_1$resource_results`), the estimate of resource abundance from the observation function (`sim_1$observation_results`, the costs the manager sets for the only available action of culling (`sim_1$manager_results`), and the number of culls attempted by each user (`sim_1$user_results`). By default, only one resource type is used, but custom subfunctions could potentially allow for models with multiple resource types. Any custom subfunctions can replace GMSE predefined functions, provided that they have appropriately defined inputs and outputs (see [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf)). For example, we can define a very simple logistic growth function to send to `res_mod` instead.

```{r, eval = TRUE}
alt_res <- function(X, K = 2000, rate = 1){
    X_1 <- X + rate*X*(1 - X/K);
    return(X_1);
}
```

The above function takes in a population size of `X` and returns a value `X_1` based on the population intrinsic growth rate `rate` and carrying capacity `K`. Iterating the logistic growth model by itself under default parameter values with a starting population of 100 will cause the population to increase to carrying capacity in ca seven generations. The function can be substituted into `gmse_apply` to use it instead of the predefined GMSE resource model.

```{r, eval = TRUE}
sim_2 <- gmse_apply(res_mod = alt_res, X = 100, rate = 0.3);
```

The `gmse_apply` function will find the parameters it needs to run the `alt_res` function in place of the default resource function, either by running the default function values (e.g., `K = 2000`) or values specified directly into `gmse_apply` (e.g., `X = 100` and `rate = 0.3`). If an argument to a custom function is required but not provided either as a default or specified in `gmse_apply`, then an error will be returned. Results for the above `sim_2` are returned below.

```{r, eval = TRUE}
print(sim_2);
```

How `gmse_apply` integrates across submodels
--------------------------------------------------------------------------------

To integrate across different types of submodels, `gmse_apply` translates between vectors and arrays between each submodel. For example, because the default GMSE observation model requires a resource array with particular requirements for column identites, when a resource model subfunction returns a vector, or a list with a named element 'resource_vector', this vector is translated into an array that can be used by the observation model. Specifically, each element of the vector identifies the abundance of a resource type (and hence will usually be just a single value denoting abundance of the only focal population). If this is all the information provided, then a 'resource_array' will be made with default GMSE parameter values with an identical number of rows to the abundance value (floored if the value is a non-integer; non-default values can also be put into this transformation from vector to array if they are specified in `gmse_apply`, e.g., through an argument such as `lambda = 0.8`). Similarly, a `resource_array` is also translated into a vector after the default individual-based resource model is run, should the observation model require simple abundances instead of an array. The same is true of `observation_vector` and `observation_array` objects returned by observation models, of `manager_vector` and `manager_array` (i.e., `COST` in the `gmse` function) objects returned by manager models, and of `user_vector` and `user_array` (i.e., `ACTION` in the `gmse` function) objects returned by user models. At each step, a translation between the two is made, with necessary adjustments that can be tweaked through arguments to `gmse_apply` when needed. Alternative observation, manager, and user, submodels, for example, are defined below; note that each requires a vector from the preceding model.

```{r, eval = TRUE}
# Alternative observation submodel
alt_obs <- function(resource_vector){ 
    X_obs <- resource_vector - 0.1 * resource_vector;
    return(X_obs);
}

# Alternative manager submodel
alt_man <- function(observation_vector){
    policy <- observation_vector - 1000;
    if(policy < 0){
        policy <- 0;
    }
    return(policy);
}

# Alternative user submodel
alt_usr <- function(manager_vector){
    harvest <- manager_vector + manager_vector * 0.1;
    return(harvest);
}
```

All of these submodels are completely deterministic, so when run with the same parameter combinations, they produce replicable outputs.

```{r, eval = TRUE}
gmse_apply(res_mod = alt_res, obs_mod = alt_obs, 
           man_mod = alt_man, use_mod = alt_usr, X = 1000);
```

Note that the `manager_results` and `user_results` are ambiguous here, and can be interpreted as desired -- e.g., as total allowable catch and catches made, or as something like costs of catching set by the manager and effort to catching made by the user. Hence while manger output is set in terms of costs of performing each action, and user output is set in terms of action attempts, this need not be the case when using `gmse_apply` (though it should be recognised when using default GMSE manager and user functions). GMSE default submodels can be added in at any point.

```{r, eval = TRUE}
gmse_apply(res_mod = alt_res, obs_mod = observation, 
           man_mod = alt_man, use_mod = alt_usr, X = 1000)
```

It is possible to, for example, specify a simple resource and observation model, but then take advantage of the genetic algorithm to predict policy decisions and user actions. This can be done by using the default GMSE manager and user functions (written below explicitly, though this is not necessary).

```{r, eval = TRUE}
gmse_apply(res_mod = alt_res, obs_mod = alt_obs, 
           man_mod = manager, use_mod = user, X = 1000)
```

Running GMSE simulations by looping `gmse_apply`
--------------------------------------------------------------------------------

Instead of using the `gmse` function, multiple simulations of GMSE can be run by calling `gmse_apply` through a loop, reassigning outputs where necessary for the next generation. This is best accomplished using the argument `old_list`, which allows previous full results from `gmse_apply` to be reinserted into the `gmse_apply` function. The argument `old_list` is `NULL` by default, but can instead take the output of a previous full list return of  `gmse_apply`. This `old_list` produced when `get_res = Full` includes all data structures and parameter values necessary for a unique simulation of GMSE. An exampe of using `get_res` and `old_list` in tandem to loop `gmse_apply` is shown below.

```{r, eval = TRUE}
to_scare  <- FALSE;
sim_old   <- gmse_apply(scaring = to_scare, get_res = "Full", stakeholders = 6);
sim_sum_1 <- matrix(data = NA, nrow = 20, ncol = 7);
for(time_step in 1:20){
    sim_new               <- gmse_apply(scaring = to_scare, get_res = "Full", 
                                        old_list = sim_old);
    sim_sum_1[time_step, 1] <- time_step;
    sim_sum_1[time_step, 2] <- sim_new$basic_output$resource_results[1];
    sim_sum_1[time_step, 3] <- sim_new$basic_output$observation_results[1];
    sim_sum_1[time_step, 4] <- sim_new$basic_output$manager_results[2];
    sim_sum_1[time_step, 5] <- sim_new$basic_output$manager_results[3];
    sim_sum_1[time_step, 6] <- sum(sim_new$basic_output$user_results[,2]); 
    sim_sum_1[time_step, 7] <- sum(sim_new$basic_output$user_results[,3]); 
    sim_old                 <- sim_new;
}
colnames(sim_sum_1) <- c("Time", "Pop_size", "Pop_est", "Scare_cost", 
                         "Cull_cost", "Scare_count", "Cull_count");
print(sim_sum_1);
```

Note that one element of the full list `gmse_apply` output is the 'basic_output' itself, which is produced by default when `get_res = "basic"`. This is what is being used to store the output of `sim_new` into `sim_sum_1`. Next, we show how the flexibility of `gmse_apply` can be used to dynamically redefine simulation conditions.

Changing simulation conditions using `gmse_apply`
================================================================================

We can take advantage of `gmse_apply` to dynamically change parameter values mid-loop. For example, below shows the same code used in the previous example, but with a policy of scaring introduced on time step 10.

```{r, eval = TRUE}
to_scare  <- FALSE;
sim_old   <- gmse_apply(scaring = to_scare, get_res = "Full", stakeholders = 6);
sim_sum_2 <- matrix(data = NA, nrow = 20, ncol = 7);
for(time_step in 1:20){
    sim_new               <- gmse_apply(scaring = to_scare, get_res = "Full", 
                                        old_list = sim_old);
    sim_sum_2[time_step, 1] <- time_step;
    sim_sum_2[time_step, 2] <- sim_new$basic_output$resource_results[1];
    sim_sum_2[time_step, 3] <- sim_new$basic_output$observation_results[1];
    sim_sum_2[time_step, 4] <- sim_new$basic_output$manager_results[2];
    sim_sum_2[time_step, 5] <- sim_new$basic_output$manager_results[3];
    sim_sum_2[time_step, 6] <- sum(sim_new$basic_output$user_results[,2]); 
    sim_sum_2[time_step, 7] <- sum(sim_new$basic_output$user_results[,3]); 
    sim_old               <- sim_new;
    if(time_step == 10){
        to_scare <- TRUE;
    }
}
colnames(sim_sum_2) <- c("Time", "Pop_size", "Pop_est", "Scare_cost", 
                         "Cull_cost", "Scare_count", "Cull_count");
print(sim_sum_2);
```

Hence, in addition to the previously explained benefits of the flexible `gmse_apply` function, one particularly useful feature is that we can use it to study change in policy availability -- in the above case, what happens when scaring is suddenly introduced as a possible policy option. Similar things can be done, for example, to see how manager or user power changes over time. In the example below, users' budgets increase by 100 every time step, with the manager's budget remaining the same. The consequence of this increasing user budget is higher rates of culling and decreased population size.

```{r, eval = TRUE}
ub          <- 500;
sim_old     <- gmse_apply(get_res = "Full", stakeholders = 6, user_budget = ub);
sim_sum_3   <- matrix(data = NA, nrow = 20, ncol = 6);
for(time_step in 1:20){
    sim_new               <- gmse_apply(get_res = "Full", old_list = sim_old,
                                        user_budget = ub);
    sim_sum_3[time_step, 1] <- time_step;
    sim_sum_3[time_step, 2] <- sim_new$basic_output$resource_results[1];
    sim_sum_3[time_step, 3] <- sim_new$basic_output$observation_results[1];
    sim_sum_3[time_step, 4] <- sim_new$basic_output$manager_results[3];
    sim_sum_3[time_step, 5] <- sum(sim_new$basic_output$user_results[,3]);
    sim_sum_3[time_step, 6] <- ub;
    sim_old               <- sim_new;
    ub                    <- ub + 100;
}
colnames(sim_sum_3) <- c("Time", "Pop_size", "Pop_est", "Cull_cost", "Cull_count",
                         "User_budget");
print(sim_sum_3);
```

There is an important note to make about changing arguments to `gmse_apply` when `old_list` is being used: The function `gmse_apply` is trying to avoid a crash, so `gmse_apply` will accomodate parameter changes by rebuilding data structures if necessary. For example, if the number of stakeholders is changed (and by including an argument `stakeholders` to `gmse_apply`, it is assumed that stakeholders are changing even they are not), then a new array of agents will need to be built. If landscape dimensions are changed (or just include the argument `land_dim_1` or `land_dim_2`), then a new landscape willl be built. For most simulation purposes, this will not introduce any undesirable effect on simulation results, but it should be noted and understood when developing models.

### 2017-11-03 notebook/ms/SI1.Rmd
Add supporting information documents

@@
 
-Extended introduction to the GMSE apply function (`gmse_apply`)
+Extended introduction to the genetic algorithm applied in GMSE
 ================================================================================
 
-The `gmse_apply` function is a flexible function that allows for user-defined sub-functions calling resource, observation, manager, and user models.
-Where such models are not specified, predefined GMSE submodels 'resource', 'observation', 'manager', and 'user' are run by default.
-Any type of sub-model (e.g., numerical,  individual-based) is permitted as long as the input and output are appropriately specified.
-Only one time step is simulated per call to `gmse_apply`, so the function must be looped for simulation over time.
-Where model parameters are needed but not specified, defaults from GMSE are used.
-In this Supporting Information, we demonstrate some uses of `gmse_apply`, and how it might be used to simulate myriad management scenarios *in silico*.
+Thanks for the clarification regarding the equation.
+I'll try to answer as best as I can -- apologies if this has been unclear.
+At the broadest scale, the equation for user fitness would be on L367 in the strategy_fitness function ( https://github.com/bradduthie/gmse/blob/master/src/game.c#L376 ).
+Here's what's going on: Users are predicting how their actions will change the quantities of things in the model (either resources or landscape output), and these changes are individually multiplied by the users' utilities for that thing.
+The change multiplied by utility for each thing is summed across all things to get a value for fitness.
+Note that positive change times positive utility, and negative change times negative utility, will increase fitness (i.e., increasing the thing users want more of and decreasing the things they want less of).
+Hence, an equation describing user fitness would be the below,
 
-A simple run of `gmse_apply()` returns one generation of GMSE using predefined submodels and default parameter values.
+$$ F_{user} = \sum_{i=1}^{N} \Delta A_{i} \times U_{i} $$.
 
-```{r, eval = TRUE}
-sim_1 <- gmse_apply();
-```
+Where $F_{user}$ is user fitness, $N$ is the total number of things that might be of interest (at the moment $N = 2$ in GMSE, one resource and, potentially, one landscape value), $\Delta A_{i}$ is the change in the abundance of thing $i$, and $U_{i}$ is the utility of thing $i$ from the perspective of the user (apologies for the LaTeX code -- attached a PNG of the conversion).
+I want to stress though that I would not consider this equation to be central to the GMSE framework -- if someone else has a better approach for defining fitness, or defining any of the terms listed above, or wants to expand upon it to include new things, then that would be awesome!
+The above just works well as a heuristic tool to get users to act in such a way as to maximise their interests in harvesting or getting more crop yield (as is my intent), but it's not based on first principles and I don't claim it to be particularly special.
 
-For `sim_1`, the default 'basic' results are returned as below, which summarise key values for all submodels.
-
-```{r, eval = TRUE}
-print(sim_1);
-```
-
-Note that in the case above we have the total abundance of resources returned (`sim_1$resource_results`), the estimate of resource abundance from the observation function (`sim_1$observation_results`, the costs the manager sets for the only available action of culling (`sim_1$manager_results`), and the number of culls attempted by each user (`sim_1$user_results`).
-By default, only one resource type is used, but custom subfunctions could potentially allow for models with multiple resource types.
-Any custom subfunctions can replace GMSE predefined functions, provided that they have appropriately defined inputs and outputs (see [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf)).
-For example, we can define a very simple logistic growth function to send to `res_mod` instead.
-
-```{r, eval = TRUE}
-alt_res <- function(X, K = 2000, rate = 1){
-    X_1 <- X + rate*X*(1 - X/K);
-    return(X_1);
-}
-```
-
-The above function takes in a population size of `X` and returns a value `X_1` based on the population intrinsic growth rate `rate` and carrying capacity `K`.
-Iterating the logistic growth model by itself under default parameter values with a starting population of 100 will cause the population to increase to carrying capacity in ca seven generations.
-The function can be substituted into `gmse_apply` to use it instead of the predefined GMSE resource model.
-
-```{r, eval = TRUE}
-sim_2 <- gmse_apply(res_mod = alt_res, X = 100, rate = 0.3);
-```
-
-The `gmse_apply` function will find the parameters it needs to run the `alt_res` function in place of the default resource function, either by running the default function values (e.g., `K = 2000`) or values specified directly into `gmse_apply` (e.g., `X = 100` and `rate = 0.3`).
-If an argument to a custom function is required but not provided either as a default or specified in `gmse_apply`, then an error will be returned.
-Results for the above `sim_2` are returned below.
-
-```{r, eval = TRUE}
-print(sim_2);
-```
-
-How `gmse_apply` integrates across submodels
---------------------------------------------------------------------------------
-
-To integrate across different types of submodels, `gmse_apply` translates between vectors and arrays between each submodel.
-For example, because the default GMSE observation model requires a resource array with particular requirements for column identites, when a resource model subfunction returns a vector, or a list with a named element 'resource_vector', this vector is translated into an array that can be used by the observation model.
-Specifically, each element of the vector identifies the abundance of a resource type (and hence will usually be just a single value denoting abundance of the only focal population).
-If this is all the information provided, then a 'resource_array' will be made with default GMSE parameter values with an identical number of rows to the abundance value (floored if the value is a non-integer; non-default values can also be put into this transformation from vector to array if they are specified in `gmse_apply`, e.g., through an argument such as `lambda = 0.8`).
-Similarly, a `resource_array` is also translated into a vector after the default individual-based resource model is run, should the observation model require simple abundances instead of an array.
-The same is true of `observation_vector` and `observation_array` objects returned by observation models, of `manager_vector` and `manager_array` (i.e., `COST` in the `gmse` function) objects returned by manager models, and of `user_vector` and `user_array` (i.e., `ACTION` in the `gmse` function) objects returned by user models.
-At each step, a translation between the two is made, with necessary adjustments that can be tweaked through arguments to `gmse_apply` when needed.
-Alternative observation, manager, and user, submodels, for example, are defined below; note that each requires a vector from the preceding model.
-
-```{r, eval = TRUE}
-# Alternative observation submodel
-alt_obs <- function(resource_vector){ 
-    X_obs <- resource_vector - 0.1 * resource_vector;
-    return(X_obs);
-}
-
-# Alternative manager submodel
-alt_man <- function(observation_vector){
-    policy <- observation_vector - 1000;
-    if(policy < 0){
-        policy <- 0;
-    }
-    return(policy);
-}
-
-# Alternative user submodel
-alt_usr <- function(manager_vector){
-    harvest <- manager_vector + manager_vector * 0.1;
-    return(harvest);
-}
-```
-
-All of these submodels are completely deterministic, so when run with the same parameter combinations, they produce replicable outputs.
-
-```{r, eval = TRUE}
-gmse_apply(res_mod = alt_res, obs_mod = alt_obs, 
-           man_mod = alt_man, use_mod = alt_usr, X = 1000);
-```
-
-Note that the `manager_results` and `user_results` are ambiguous here, and can be interpreted as desired -- e.g., as total allowable catch and catches made, or as something like costs of catching set by the manager and effort to catching made by the user.
-Hence while manger output is set in terms of costs of performing each action, and user output is set in terms of action attempts, this need not be the case when using `gmse_apply` (though it should be recognised when using default GMSE manager and user functions).
-GMSE default submodels can be added in at any point.
-
-```{r, eval = TRUE}
-gmse_apply(res_mod = alt_res, obs_mod = observation, 
-           man_mod = alt_man, use_mod = alt_usr, X = 1000)
-```
-
-It is possible to, for example, specify a simple resource and observation model, but then take advantage of the genetic algorithm to predict policy decisions and user actions.
-This can be done by using the default GMSE manager and user functions (written below explicitly, though this is not necessary).
-
-```{r, eval = TRUE}
-gmse_apply(res_mod = alt_res, obs_mod = alt_obs, 
-           man_mod = manager, use_mod = user, X = 1000)
-```
-
-Running GMSE simulations by looping `gmse_apply`
---------------------------------------------------------------------------------
-
-Instead of using the `gmse` function, multiple simulations of GMSE can be run by calling `gmse_apply` through a loop, reassigning outputs where necessary for the next generation.
-This is best accomplished using the argument `old_list`, which allows previous full results from `gmse_apply` to be reinserted into the `gmse_apply` function.
-The argument `old_list` is `NULL` by default, but can instead take the output of a previous full list return of  `gmse_apply`.
-This `old_list` produced when `get_res = Full` includes all data structures and parameter values necessary for a unique simulation of GMSE.
-An exampe of using `get_res` and `old_list` in tandem to loop `gmse_apply` is shown below.
-
-```{r, eval = TRUE}
-to_scare  <- FALSE;
-sim_old   <- gmse_apply(scaring = to_scare, get_res = "Full", stakeholders = 6);
-sim_sum_1 <- matrix(data = NA, nrow = 20, ncol = 7);
-for(time_step in 1:20){
-    sim_new               <- gmse_apply(scaring = to_scare, get_res = "Full", 
-                                        old_list = sim_old);
-    sim_sum_1[time_step, 1] <- time_step;
-    sim_sum_1[time_step, 2] <- sim_new$basic_output$resource_results[1];
-    sim_sum_1[time_step, 3] <- sim_new$basic_output$observation_results[1];
-    sim_sum_1[time_step, 4] <- sim_new$basic_output$manager_results[2];
-    sim_sum_1[time_step, 5] <- sim_new$basic_output$manager_results[3];
-    sim_sum_1[time_step, 6] <- sum(sim_new$basic_output$user_results[,2]); 
-    sim_sum_1[time_step, 7] <- sum(sim_new$basic_output$user_results[,3]); 
-    sim_old                 <- sim_new;
-}
-colnames(sim_sum_1) <- c("Time", "Pop_size", "Pop_est", "Scare_cost", 
-                         "Cull_cost", "Scare_count", "Cull_count");
-print(sim_sum_1);
-```
-
-Note that one element of the full list `gmse_apply` output is the 'basic_output' itself, which is produced by default when `get_res = "basic"`.
-This is what is being used to store the output of `sim_new` into `sim_sum_1`.
-Next, we show how the flexibility of `gmse_apply` can be used to dynamically redefine simulation conditions.
-
-Changing simulation conditions using `gmse_apply`
-================================================================================
-
-We can take advantage of `gmse_apply` to dynamically change parameter values mid-loop.
-For example, below shows the same code used in the previous example, but with a policy of scaring introduced on time step 10.
-
-```{r, eval = TRUE}
-to_scare  <- FALSE;
-sim_old   <- gmse_apply(scaring = to_scare, get_res = "Full", stakeholders = 6);
-sim_sum_2 <- matrix(data = NA, nrow = 20, ncol = 7);
-for(time_step in 1:20){
-    sim_new               <- gmse_apply(scaring = to_scare, get_res = "Full", 
-                                        old_list = sim_old);
-    sim_sum_2[time_step, 1] <- time_step;
-    sim_sum_2[time_step, 2] <- sim_new$basic_output$resource_results[1];
-    sim_sum_2[time_step, 3] <- sim_new$basic_output$observation_results[1];
-    sim_sum_2[time_step, 4] <- sim_new$basic_output$manager_results[2];
-    sim_sum_2[time_step, 5] <- sim_new$basic_output$manager_results[3];
-    sim_sum_2[time_step, 6] <- sum(sim_new$basic_output$user_results[,2]); 
-    sim_sum_2[time_step, 7] <- sum(sim_new$basic_output$user_results[,3]); 
-    sim_old               <- sim_new;
-    if(time_step == 10){
-        to_scare <- TRUE;
-    }
-}
-colnames(sim_sum_2) <- c("Time", "Pop_size", "Pop_est", "Scare_cost", 
-                         "Cull_cost", "Scare_count", "Cull_count");
-print(sim_sum_2);
-```
-
-Hence, in addition to the previously explained benefits of the flexible `gmse_apply` function, one particularly useful feature is that we can use it to study change in policy availability -- in the above case, what happens when scaring is suddenly introduced as a possible policy option.
-Similar things can be done, for example, to see how manager or user power changes over time.
-In the example below, users' budgets increase by 100 every time step, with the manager's budget remaining the same.
-The consequence of this increasing user budget is higher rates of culling and decreased population size.
-
-```{r, eval = TRUE}
-ub          <- 500;
-sim_old     <- gmse_apply(get_res = "Full", stakeholders = 6, user_budget = ub);
-sim_sum_3   <- matrix(data = NA, nrow = 20, ncol = 6);
-for(time_step in 1:20){
-    sim_new               <- gmse_apply(get_res = "Full", old_list = sim_old,
-                                        user_budget = ub);
-    sim_sum_3[time_step, 1] <- time_step;
-    sim_sum_3[time_step, 2] <- sim_new$basic_output$resource_results[1];
-    sim_sum_3[time_step, 3] <- sim_new$basic_output$observation_results[1];
-    sim_sum_3[time_step, 4] <- sim_new$basic_output$manager_results[3];
-    sim_sum_3[time_step, 5] <- sum(sim_new$basic_output$user_results[,3]);
-    sim_sum_3[time_step, 6] <- ub;
-    sim_old               <- sim_new;
-    ub                    <- ub + 100;
-}
-colnames(sim_sum_3) <- c("Time", "Pop_size", "Pop_est", "Cull_cost", "Cull_count",
-                         "User_budget");
-print(sim_sum_3);
-```
-
-There is an important note to make about changing arguments to `gmse_apply` when `old_list` is being used: The function `gmse_apply` is trying to avoid a crash, so `gmse_apply` will accomodate parameter changes by rebuilding data structures if necessary.
-For example, if the number of stakeholders is changed (and by including an argument `stakeholders` to `gmse_apply`, it is assumed that stakeholders are changing even they are not), then a new array of agents will need to be built.
-If landscape dimensions are changed (or just include the argument `land_dim_1` or `land_dim_2`), then a new landscape willl be built.
-For most simulation purposes, this will not introduce any undesirable effect on simulation results, but it should be noted and understood when developing models.
+The values of $\Delta A_{i}$ are calculated for resources and the landscape in the functions res_to_counts and land_to_counts, respectively (and $U_{i}$ is specified a priori in the model depending on other parameters -- namely land_ownership).
+Again, a bit of heuristic is needed here because there cannot be any perfect way of exactly predicting how a users actions will increase or decrease resources -- there are too many complex factors (e.g., behaviour of other stakeholders, demographic stochasticity, movement of resources on the landscape, and interactions between resources and the landscape).
+Even if we could include all of these things somehow, it would be a bit unrealistic in that real stakeholders would never have this much information.
+The predicted direct effect of actions on resources is shown in lines 268-272 ( https://github.com/bradduthie/gmse/blob/master/src/game.c#L268 ), and the array 'jaco' (a sort of Jacobian matrix) accounts for interactions between landscape and resources on line 286.
+Something similar happens in the land_to_counts function.
+The manager's genetic algorithm works in a similar way (the above equation applies), but with the need to dynamically update utility values based on current resource abundance, and to account for the predicted actions of users in finding $\Delta A_{i}$.

### 2017-11-06 notebook/ms/SI1.Rmd
Framework for the new supporting information

@@
 title: 'GMSE: an R package for generalised management strategy evaluation'
-subtitle: 'Supporting Information 2'
+subtitle: 'Supporting Information 1'
 author: A. Bradley Duthie, Jeremy J. Cusack, Isabel L. Jones, Erlend B. Nilsen, Roc&#0237;o Pozo, O.
@@
 
+A genetic algorithm is called in the predefined GMSE manager and user models to simulate human decision making.
+As of GMSE version 0.3.1.9, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step.
+Therefore, one run of the genetic algorithm occurs to simulate the manager's policy setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decions that all other users follow identically.
+Each run of the genetic algorithm mimics the evolution by natural selection of a populaiton of potential manager or user strategies over multiple generations, with the highest fitness strategy in the terminal generation being selected as the one that the manager or user decides to implement.
+For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'generation' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the main text).
+Below, we explain the genetic algorithm in extended detail as it occurs in GMSE v0.3.1.9 (future versions of GMSE might expand upon this framework).
+We first explain the key data structure used, then provide an overview of the processes of crossover, mutation, cost constraint, fitness evaluation, tournament selection, and replacement.
+We then explain the fitness functions of managers and users in more detail.
+
+Key data structures used
+================================================================================
+
+General overview of key aspects of the genetic algorithm 
+================================================================================
+
+Crossover
+--------------------------------------------------------------------------------
+
+Mutation
+--------------------------------------------------------------------------------
+
+Cost constraint
+--------------------------------------------------------------------------------
+
+Fitness evaluation
+--------------------------------------------------------------------------------
+
+Tournament selection
+--------------------------------------------------------------------------------
+
+Replacement
+--------------------------------------------------------------------------------
+
+Detailed explanation of manager and user fitness functions
+================================================================================
+
+Manager fitness function
+--------------------------------------------------------------------------------
+
+User fitness function
+--------------------------------------------------------------------------------
+
 Thanks for the clarification regarding the equation.

### 2017-11-06 notebook/ms/SI1.Rmd
Introduce ACTION data structure

@@
 title: 'GMSE: an R package for generalised management strategy evaluation'
-subtitle: 'Supporting Information 1'
-author: A. Bradley Duthie, Jeremy J. Cusack, Isabel L. Jones, Erlend B. Nilsen, Roc&#0237;o Pozo, O.
-  Sarobidy Rakotonarivo, Bram Van Moorter, and Nils Bunnefeld
+author: A. Bradley Duthie, Jeremy J. Cusack, Isabel L. Jones, Erlend B. Nilsen, Roc&#0237;o
+  Pozo, O. Sarobidy Rakotonarivo, Bram Van Moorter, and Nils Bunnefeld
 date: '`r Sys.Date()`'
@@
     reference_docx: docx_template.docx
+linestretch: 1
+subtitle: Supporting Information 1
 header-includes:
@@
 - \linenumbers
-linestretch: 1
 ---
@@
 library(GMSE);
+sim <- gmse_apply(get_res = "Full", stakeholders = 2);
 ```
@@
 
+The focal data structure used for tracking manager and user decisions is a three dimensional array, which we will call `ACTION` (also returned as `user_array` by `gmse_apply`).
+Rows of `ACTION` correspond to recipients of actions (resources, landscapes, or potentially other agents), and columns correspond to either recipient properties or actions allocated to recipients.
+Each layer of `ACTION` corresponds to a unique agent, the first of which is the manager; additional layers correspond to each user.
+Below shows an `ACTION` array, which corresponds to a GMSE model with one manager and two users.
+
+```{r, echo = FALSE}
+samp_ACTION <- sim$ACTION;
+dimnames(samp_ACTION)[[1]] <- c("Resource", "Landscape", "Res_cost", "U1_cost", "U2_cost");
+dimnames(samp_ACTION)[[2]] <- c("Act", "Type_1", "Type_2", "Type_3", "Util.", "U_land", "U_loc.", "Scare", "Cull", "Castrate", "Feed", "Help_off", "None");
+dimnames(samp_ACTION)[[3]] <- c("Manager_Actions", "User_1_Actions", "User_2_Actions");
+print(samp_ACTION);
+```
+
+The above array holds all of the information on manager and user actions in the six right-most columns of each array layer.
+The first seven columns contain information about which resources are affected, and how they are affected.
+The first column `Act` identifies the type of action being performed; a value of -2 defines a direct action to a resource (e.g., culling of the resource), and a value of -1 defines direct action to a landscape (e.g., increasing yield).
+Positive values are currently only meaningful for `Manager_Actions`, where a value of 1 defines an action setting a uniform cost of users' direct actions on resources (i.e., costs where `Act = -2` for `User_1_Actions` and `User_2_Actions`).
+All other values for `Act` are meaningless in GMSE 0.3.1.7, but might be expanded upon in future versions to allow for modification of specific user costs enacted by managers (i.e., managers having different policies for different users) or other users (e.g., users increasing the costs of other users' actions due to conflict or cooperation).
+Similarly, columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in GMSE v0.3.1.7 (i.e., only one type of resource is permitted), but future versions might allow for different resource types (e.g., `Type_1` might be used to designate species, and `Type_2` and `Type_3` could designate stage or sex).
+For the rest of this supporting information, we will therefore focus only on rows 1-3 of `ACTION`.
+Column 5 `Util.` of `ACTION` defines the utility associated with the resource (where `Act = -2`) or landscape (where `Act = -1`).
+For managers, the target resource abundance set with GMSE argument `manage_target` is found in row 1 (1000 in `ACTION` above); for users, the value in row 1 identifies whether resources are preferred to increase (if positive) or decrease (if negative).
+Values of column 5 in row 2 similarly identifies whether landscape cell output is preferred by users to increase or decrease (managers do not currently have preferences for landscape output).
+Of special note is row 3 for `Manager_Actions`, which defines the marginal utility of resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the estimated abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca 1065.76, so the manager will set policy in attempt to reduce the population size by ca 65.76 resources).
+Column 6 `U_land` defines whether or not the utility attached to the resource or landscape output depends on it being on a landscape cell that is owned by the acting user.
+Related, column 7 `U_loc.` defines whether or not actions can be performed only on a landscape cell that is owned by the acting user.
+Hence values of columns 6 and 7 are binary, and affected by the `land_ownership` argument in `gmse`.
+Finally, columns 8-13 correspond to specific actions, either direct (where `Act < 0`) or indirect by setting policy (for row 3 of `Manager_Actions` where `Res_cost = 1`).
+The last column 13 `None` corresponds with no actions.
+See [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf) for details about the effects of each action.
+
 General overview of key aspects of the genetic algorithm 
 ================================================================================
+
+It is the `ACTION` array that is affected by the genetic algorithm.
 

### 2017-11-06 notebook/ms/SI1.Rmd
Finish data structures section

@@
 Column 5 `Util.` of `ACTION` defines the utility associated with the resource (where `Act = -2`) or landscape (where `Act = -1`).
-For managers, the target resource abundance set with GMSE argument `manage_target` is found in row 1 (1000 in `ACTION` above); for users, the value in row 1 identifies whether resources are preferred to increase (if positive) or decrease (if negative).
+For managers, the target resource abundance set with GMSE argument `manage_target` is found in row 1 (`r samp_ACTION[1,5,1]` in `ACTION` above); for users, the value in row 1 identifies whether resources are preferred to increase (if positive) or decrease (if negative).
 Values of column 5 in row 2 similarly identifies whether landscape cell output is preferred by users to increase or decrease (managers do not currently have preferences for landscape output).
-Of special note is row 3 for `Manager_Actions`, which defines the marginal utility of resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the estimated abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca 1065.76, so the manager will set policy in attempt to reduce the population size by ca 65.76 resources).
+Of special note is row 3 for `Manager_Actions`, which defines the marginal utility of resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the estimated abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca `r round((samp_ACTION[1,5,1] + -1*samp_ACTION[3,5,1]),2)`, so the manager will set policy in attempt to change the population size by ca `r round(samp_ACTION[3,5,1],2)` resources).
 Column 6 `U_land` defines whether or not the utility attached to the resource or landscape output depends on it being on a landscape cell that is owned by the acting user.
@@
 See [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf) for details about the effects of each action.
+
+Constraints on the values that elements in the `ACTION` array can take are defined by a `COST` array (also returned as `manager_array` by `gmse_apply`) of dimensions identical to `ACTION`.
+All values in `COST` columns 1-7 are set to 10001, one higher than the highest possible budget of either managers or users, so neither can affect resource types or utilities.
+Columns 8-13 are also set to 10001, except where actions are specifically allowed.
+Below shows the `COST` array that corresponds to the above `ACTION` array.
+
+```{r, echo = FALSE}
+samp_COST <- sim$COST;
+dimnames(samp_COST)[[1]] <- c("Resource", "Landscape", "Res_cost", "U1_cost", "U2_cost");
+dimnames(samp_COST)[[2]] <- c("Act", "Type_1", "Type_2", "Type_3", "Util.", "U_land", "U_loc.", "Scare", "Cull", "Castrate", "Feed", "Help_off", "None");
+dimnames(samp_COST)[[3]] <- c("Manager_Actions", "User_1_Actions", "User_2_Actions");
+print(samp_COST);
+```
+
+Note that in default GMSE parameters, `culling = TRUE`, but all other actions are false.
+Hence the `Cull` column 9 is the only column besides column 13 `None` in which cost is less than 10001.
+Manager's actions in `ACTION` directly affect the cost of users performing one of the five possible actions on resources (columns 8-12).
+This can be verified in `ACTION` where the manager has set the cost of scaring to `r samp_ACTION[3,9,1]`, and the corresponding `COST` of resource culling (row 1) is `r samp_COST[1,9,2]` for both users.
+The cost of the manager affecting the cost of user actions is always set to the `minimum_cost` defined in GMSE; here the default `r samp_COST[3,9,1]` is used.
+This `minimum_cost` also defines cost values for `None`, in which the user or manager does nothing, as might occur if the manager wants to permit culling and therefore does not want to invest any of their `manager_budget` to increasing the cost of culling.
+Both `ACTION` and `COST` are updated in each time step unless `manage_freq > 1`, in which case `COST` and `Manager_Actions` in `ACTION` are at the frequency defined.
 

### 2017-11-07 notebook/ms/SI1.Rmd
Introduction to the process

@@
 As of GMSE version 0.3.1.9, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step.
-Therefore, one run of the genetic algorithm occurs to simulate the manager's policy setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decions that all other users follow identically.
-Each run of the genetic algorithm mimics the evolution by natural selection of a populaiton of potential manager or user strategies over multiple generations, with the highest fitness strategy in the terminal generation being selected as the one that the manager or user decides to implement.
+Therefore, one run of the genetic algorithm occurs to simulate the manager's policy setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users follow identically.
+Each run of the genetic algorithm mimics the evolution by natural selection of a population of potential manager or user strategies over multiple generations, with the highest fitness strategy in the terminal generation being selected as the one that the manager or user decides to implement.
 For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'generation' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the main text).
-Below, we explain the genetic algorithm in extended detail as it occurs in GMSE v0.3.1.9 (future versions of GMSE might expand upon this framework).
-We first explain the key data structure used, then provide an overview of the processes of crossover, mutation, cost constraint, fitness evaluation, tournament selection, and replacement.
+Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.3.1.9 (future versions of GMSE might expand upon this framework).
+We first explain the key data structures used, then provide an overview of how a population of strategies is initialised, and the  subsequent processes of crossover, mutation, cost constraint, fitness evaluation, tournament selection, and replacement.
 We then explain the fitness functions of managers and users in more detail.
@@
 The focal data structure used for tracking manager and user decisions is a three dimensional array, which we will call `ACTION` (also returned as `user_array` by `gmse_apply`).
-Rows of `ACTION` correspond to recipients of actions (resources, landscapes, or potentially other agents), and columns correspond to either recipient properties or actions allocated to recipients.
-Each layer of `ACTION` corresponds to a unique agent, the first of which is the manager; additional layers correspond to each user.
-Below shows an `ACTION` array, which corresponds to a GMSE model with one manager and two users.
+Rows of `ACTION` correspond to the entities affected by actions (resources, landscape properties, or potentially other agents), and columns correspond either to properties of the affected entities, or to the actions potentially allocated to them.
+Each layer of `ACTION` corresponds to a unique agent, the first of which is the manager; additional layers correspond to users.
+Below shows an `ACTION` array for a GMSE model with one manager and two users.
 
@@
 
-The above array holds all of the information on manager and user actions in the six right-most columns of each array layer.
-The first seven columns contain information about which resources are affected, and how they are affected.
+The above array holds all of the information on manager and user actions.
+The first seven columns contain information about which entities are affected, and how they are affected.
 The first column `Act` identifies the type of action being performed; a value of -2 defines a direct action to a resource (e.g., culling of the resource), and a value of -1 defines direct action to a landscape (e.g., increasing yield).
 Positive values are currently only meaningful for `Manager_Actions`, where a value of 1 defines an action setting a uniform cost of users' direct actions on resources (i.e., costs where `Act = -2` for `User_1_Actions` and `User_2_Actions`).
-All other values for `Act` are meaningless in GMSE 0.3.1.7, but might be expanded upon in future versions to allow for modification of specific user costs enacted by managers (i.e., managers having different policies for different users) or other users (e.g., users increasing the costs of other users' actions due to conflict or cooperation).
-Similarly, columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in GMSE v0.3.1.7 (i.e., only one type of resource is permitted), but future versions might allow for different resource types (e.g., `Type_1` might be used to designate species, and `Type_2` and `Type_3` could designate stage or sex).
+All other values for `Act` are meaningless in GMSE 0.3.1.9, but might be expanded upon in future versions to allow for modification of specific user costs enacted by managers (i.e., managers having different policies for different users) or other users (e.g., users increasing the costs of other users' actions due to conflict or cooperation).
+Similarly, columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in GMSE v0.3.1.9 (i.e., only one type of resource is permitted).
+Future versions might allow for different resource types (e.g., `Type_1` might be used to designate species, and `Type_2` and `Type_3` could designate stage or sex).
 For the rest of this supporting information, we will therefore focus only on rows 1-3 of `ACTION`.
 Column 5 `Util.` of `ACTION` defines the utility associated with the resource (where `Act = -2`) or landscape (where `Act = -1`).
-For managers, the target resource abundance set with GMSE argument `manage_target` is found in row 1 (`r samp_ACTION[1,5,1]` in `ACTION` above); for users, the value in row 1 identifies whether resources are preferred to increase (if positive) or decrease (if negative).
-Values of column 5 in row 2 similarly identifies whether landscape cell output is preferred by users to increase or decrease (managers do not currently have preferences for landscape output).
+For managers, the target resource abundance set with the GMSE argument `manage_target` is found in row 1 (`r samp_ACTION[1,5,1]` in `ACTION` above); for users, the value in row 1 identifies whether resources are preferred to increase (if positive) or decrease (if negative).
+Values of column 5 in row 2 similarly identify whether landscape cell output is preferred by users to increase or decrease (managers do not currently have preferences for landscape output).
 Of special note is row 3 for `Manager_Actions`, which defines the marginal utility of resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the estimated abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca `r round((samp_ACTION[1,5,1] + -1*samp_ACTION[3,5,1]),2)`, so the manager will set policy in attempt to change the population size by ca `r round(samp_ACTION[3,5,1],2)` resources).
@@
 Hence values of columns 6 and 7 are binary, and affected by the `land_ownership` argument in `gmse`.
-Finally, columns 8-13 correspond to specific actions, either direct (where `Act < 0`) or indirect by setting policy (for row 3 of `Manager_Actions` where `Res_cost = 1`).
+Finally, columns 8-13 correspond to specific actions, either direct (where `Act < 0`) or indirect by setting policy (for row 3 of `Manager_Actions` where `Act = 1`).
 The last column 13 `None` corresponds with no actions.
@@
 Constraints on the values that elements in the `ACTION` array can take are defined by a `COST` array (also returned as `manager_array` by `gmse_apply`) of dimensions identical to `ACTION`.
-All values in `COST` columns 1-7 are set to 10001, one higher than the highest possible budget of either managers or users, so neither can affect resource types or utilities.
-Columns 8-13 are also set to 10001, except where actions are specifically allowed.
+Elements of `COST` define how many units from the `manager_budget` or `user_budget` are needed to perform a single action; a `minimum_cost` for actions is defined as an argument in GMSE (10 by default).
+All values in `COST` columns 1-7 are set to 10001, one higher than the highest possible `manager_budget` or `user_budget`, so neither can affect resource types or utilities.
+Columns 8-13 are also set to 10001, except where actions are allowed.
 Below shows the `COST` array that corresponds to the above `ACTION` array.
@@
 This can be verified in `ACTION` where the manager has set the cost of scaring to `r samp_ACTION[3,9,1]`, and the corresponding `COST` of resource culling (row 1) is `r samp_COST[1,9,2]` for both users.
-The cost of the manager affecting the cost of user actions is always set to the `minimum_cost` defined in GMSE; here the default `r samp_COST[3,9,1]` is used.
+The cost of the manager affecting the cost of user actions is always set to the `minimum_cost`; here the default `r samp_COST[3,9,1]` is used.
 This `minimum_cost` also defines cost values for `None`, in which the user or manager does nothing, as might occur if the manager wants to permit culling and therefore does not want to invest any of their `manager_budget` to increasing the cost of culling.
-Both `ACTION` and `COST` are updated in each time step unless `manage_freq > 1`, in which case `COST` and `Manager_Actions` in `ACTION` are at the frequency defined.
+Both `ACTION` and `COST` are updated in each time step unless `manage_freq > 1`, in which case `COST` and `Manager_Actions` in `ACTION` are updated at the frequency defined.
 
@@
 
-It is the `ACTION` array that is affected by the genetic algorithm.
+The genetic algorithm updates a single layer of the `ACTION` array, which defines to the decisions of a single agent (either the manager or a user).
+The corresponding layer of the `COST` array remains unchanged, and serves only to ensure that `ACTION` values do not exceed `manager_budget` or `user_budget` for managers and users, respectively.
+The genetic algorithm proceeds by first initialising a large population of new `ACTION` layers.
+In each generation, these layers crossover and mutate, generating variation in agent decisions; costs constrain this variation from exceeding a maximum budget, then the fitness of each layer is evaluated based on how much it increases the agent's utility.
+A tournament is used to select high fitness layers, and these selected layers become the new generation of layers; generations continue until a minimum number of generations (`ga_mingen`) have passed and a convergence criteria is satisfied such that the change in mean fitness from the previous generation is below the threshold `converge_crit`.
+
+Initialisation
+--------------------------------------------------------------------------------
 

### 2017-11-07 notebook/ms/SI1.Rmd
Add initialisation

@@
 In each generation, these layers crossover and mutate, generating variation in agent decisions; costs constrain this variation from exceeding a maximum budget, then the fitness of each layer is evaluated based on how much it increases the agent's utility.
-A tournament is used to select high fitness layers, and these selected layers become the new generation of layers; generations continue until a minimum number of generations (`ga_mingen`) have passed and a convergence criteria is satisfied such that the change in mean fitness from the previous generation is below the threshold `converge_crit`.
+A tournament is used to select high fitness layers, and these selected layers become the new generation of layers; generations continue until a minimum number of generations (`ga_mingen`) have passed and a convergence criteria is satisfied such that the increase in mean fitness from the previous generation is below the threshold `converge_crit`.
 
@@
 --------------------------------------------------------------------------------
+
+At the start of each genetic algorithm, a population of size `ga_popsize` is initialised (hereafter the `POPULATION` array).
+This population is held in a 3D array of `ga_popsize` layers.
+Each layer includes an identical number of rows and columns as in `ACTION`, and one layer defines a single 'individual' in the population.
+The first seven columns of `ACTION` are replicated exactly for all individuals, and remain unchanged throughout the genetic algorithm thereby preserving the information about which entities are affected by actions in a given row.
+The remaining columns are either also replicated exactly as in `ACTION` (i.e., initialised to be the same decisions as in a previous time step), or randomly seeded with values given the constraints of `manager_budget` or `user_budget` (i.e., initialised to random decision making).
+The number of exact replicates initialised is set using `ga_seedrep` (if `ga_seedrep` $\geq$ `ga_popsize`, then all individuals are seeded as replicates).
+After the `POPULATION` of `ga_popsize` individuals is initialised, a loop simulating the adaptive evolution of `POPULATION` in non-overlapping generations begins.
 

### 2017-11-07 notebook/ms/SI1.Rmd
Add crossover and mutation sections

@@
 - \linenumbers
+bibliography: gmse.bib
+biblio-style: apalike
 ---
@@
 
+A single generation of the genetic algorithm begins with a uniform crossover [@Hamblin2013], by which actions of individuals in `POPULATION` are randomly swapped with some probability.
+To implement crossover, each individual selects a partner, then exchanges corresponding array elements affecting agent actions (columns 8-13) with their partner at a fixed probability of `ga_crossover`.
+
 Mutation
 --------------------------------------------------------------------------------
+
+Following crossover, `POPULATION` array elements affecting agent actions (columns 8-13) mutate at a fixed probability of `ga_mutation`.
+For each array element, a random uniform number $u \in [0, 1]$ is sampled.
+If $u$ is greater than `1 - (0.5 * ga_mutation)`, then the value of the array element is increased by 1.
+If $u$ is less than `0.5 * ga_mutation`, then the value of the array element is decreased by 1; when this decrease results in a negative value, the mutated value is multiplied by -1 to equal 1.
 
@@
 The manager's genetic algorithm works in a similar way (the above equation applies), but with the need to dynamically update utility values based on current resource abundance, and to account for the predicted actions of users in finding $\Delta A_{i}$.
+
+References
+================================================================================

### 2017-11-08 notebook/ms/SI1.Rmd
General overview of genetic algorithm and metaheuristics

@@
 In each generation, these layers crossover and mutate, generating variation in agent decisions; costs constrain this variation from exceeding a maximum budget, then the fitness of each layer is evaluated based on how much it increases the agent's utility.
-A tournament is used to select high fitness layers, and these selected layers become the new generation of layers; generations continue until a minimum number of generations (`ga_mingen`) have passed and a convergence criteria is satisfied such that the increase in mean fitness from the previous generation is below the threshold `converge_crit`.
+A tournament is used to select high fitness layers, and these selected layers become the new generation of layers; generations continue until a minimum number of generations (`ga_mingen`) have passed and a convergence criteria is satisfied such that the increase in mean fitness from the previous generation is below the threshold `converge_crit` (Figure 1).
+
+```{r, echo=FALSE, fig.height=1.75, fig.width=6}
+mbox <- function(x0, x1, y0, y1){
+    xx <- seq(from=x0, to=x1, length.out = 100);
+    yy <- seq(from=y0, to=y1, length.out = 100);
+    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
+    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
+    return(list(x=xd, y=yd));
+}
+par(mar=c(0,0,0,0));
+# ===============================================================
+plot(x=0, y=0, type="n", xlim=c(0,100), ylim=c(40,100), xaxt="n", yaxt="n",
+     xlab="",ylab="");
+ibox <- mbox(x0 = 0,  x1 = 10, y0 = 90, y1 = 70);
+polygon(x = ibox$x, y = ibox$y, lwd = 3, border = "black", col = "white");
+cbox <- mbox(x0 = 15, x1 = 25, y0 = 90, y1 = 70);
+polygon(x = cbox$x, y = cbox$y, lwd = 3, border = "black", col = "white");
+ubox <- mbox(x0 = 30, x1 = 40, y0 = 90, y1 = 70);
+polygon(x = ubox$x, y = ubox$y, lwd = 3, border = "black", col = "white");
+nbox <- mbox(x0 = 45, x1 = 55, y0 = 90, y1 = 70);
+polygon(x = nbox$x, y = nbox$y, lwd = 3, border = "black", col = "white");
+fbox <- mbox(x0 = 60, x1 = 70, y0 = 90, y1 = 70);
+polygon(x = fbox$x, y = fbox$y, lwd = 3, border = "black", col = "white");
+tbox <- mbox(x0 = 75, x1 = 85, y0 = 90, y1 = 70);
+polygon(x = tbox$x, y = tbox$y, lwd = 3, border = "black", col = "white");
+rbox <- mbox(x0 = 90, x1 = 100, y0 = 90, y1 = 70);
+polygon(x = rbox$x, y = rbox$y, lwd = 3, border = "black", col = "white");
+text(x=5,  y=80, labels="Initialisation", col="black", cex = 0.5);
+text(x=20, y=80, labels="Crossover", col="black", cex = 0.5);
+text(x=35, y=80, labels="Mutation", col="black", cex = 0.5);
+text(x=50, y=82, labels="Cost", col="black", cex = 0.5);
+text(x=50, y=78, labels="constraint", col="black", cex = 0.5);
+text(x=65, y=82, labels="Fitness", col="black", cex = 0.5);
+text(x=65, y=78, labels="evaluation", col="black", cex = 0.5);
+text(x=80, y=82, labels="Tournament", col="black", cex = 0.5);
+text(x=80, y=78, labels="selection", col="black", cex = 0.5);
+text(x=95, y=80, labels="Replacement", col="black", cex = 0.5);
+arrows(x0=10, x1=14, y0=80, y1=80, lwd=2, length=0.10);
+arrows(x0=25, x1=29, y0=80, y1=80, lwd=2, length=0.10);
+arrows(x0=40, x1=44, y0=80, y1=80, lwd=2, length=0.10);
+arrows(x0=55, x1=59, y0=80, y1=80, lwd=2, length=0.10);
+arrows(x0=70, x1=74, y0=80, y1=80, lwd=2, length=0.10);
+arrows(x0=85, x1=89, y0=80, y1=80, lwd=2, length=0.10);
+
+arrows(x0 = 95, x1 = 95, y0 = 70, y1 = 68, lwd = 2, length = 0);
+arrows(x0 = 90, x1 = 20, y0 = 65, y1 = 65, lwd = 2, length = 0);
+arrows(x0 = 20, x1 = 20, y0 = 65, y1 = 70, lwd = 2, length = 0.05);
+text(x=88, y=63, labels="No", col="black", cex = 0.5);
+rbox <- mbox(x0 = 90, x1 = 100, y0 = 62, y1 = 68);
+polygon(x = rbox$x, y = rbox$y, lwd = 3, border = "black", col = "black");
+text(x=95, y=65, labels="Termination?", col="white", cex = 0.5);
+fbox <- mbox(x0 = 75, x1 = 85, y0 = 60, y1 = 50);
+polygon(x = fbox$x, y = fbox$y, lwd = 3, border = "black", col = "white");
+text(x=80, y=57, labels="Agent", col="black", cex = 0.5);
+text(x=80, y=53, labels="decision", col="black", cex = 0.5);
+arrows(x0 = 95, x1 = 95, y0 = 62, y1 = 55, lwd = 2, length = 0);
+arrows(x0 = 95, x1 = 85, y0 = 55, y1 = 55, lwd = 2, length = 0.05);
+text(x=88, y=53, labels="Yes", col="black", cex = 0.5);
+```
+
+**Figure 1:** Conceptual overview of the GMSE genetic algorithm
 
@@
 
+Variation in manager or user actions generated by crossover and mutation might result in strategies that exceed `manager_budget` or `user_budget`, respectively.
+Left unchecked, this over-budgeting could lead to unnacceptably high fitness strategies, so strategies that are over budget following crossover and mutation need to be brought back within budgetary constraints.
+To do this, the genetic algorithm first checks to see if an individual in `POPULATION` is over budget.
+If so, then an action is randomly selected and removed, and budget use is reassessed; this random removal of an action and subsequent budget reassessment continues until the individual does not exceed their budget.
+
 Fitness evaluation
 --------------------------------------------------------------------------------
+
+Once all individuals in `POPULATION` are within budget, the fitness of each individual is assessed.
+Fitness assessment works differently for managers versus users because managers need to consider the consequences of their decisions on user actions, and how those actions will affect resource abundance.
+In contrast, user actions need to consider the consequences of their decisions on resource abundance or landscape output.
+Fitness of an individual is defined by a real number that increases with the degree to which an individual's actions are predicted to increase their utility (recall that managers and users assign resources or landscape output a utility value).
+Details for how fitness is calculated are provided below.
 
@@
 
-Replacement
+After each individual in `POPULATION` is assigned a fitness, a tournament is used to select individuals of higher fitness.
+Tournament selection is an especially flexible, non-parametric method that samples a subset of individuals from the total population and chooses the fittest of the subset for replacement [@Hamblin2013].
+In GMSE, tournament selection proceeds by randomly sampling `ga_sampleK` individuals from the total `POPULATION` with replacement.
+The fitnesses of the subset of `ga_sampleK` individuals are compared, and the `ga_chooseK` individuals of highest fitness are retained (if `ga_sampleK` $\geq$ `ga_chooseK`, then all `ga_sampleK` are chosen, but this is not recommended).
+Tournaments selecting `ga_chooseK` individuals from random subsets of size `ga_sampleK` continue until a total of `ga_popsize` individuals are retained.
+
+Replacement and termination
 --------------------------------------------------------------------------------
+
+Once a new set of `ga_popsize` individuals is retained through tournament selection, these individuals replace the previous `POPULATION` array.
+The genetic algorithm terminates if and only if a minimum number of generations has passed (`ga_mingen`) and a convergence criteria (`converge_crit`) is satisfied.
+The convergence criteria checks the difference between the mean fitness of individuals in the new generation versus the previous generation; if this difference is less than `converge_crit`, then termination does not occur (it is usually fine to use the default GMSE `converge_crit = 100` and `ga_mingen = 40`, which nearly always terminates the genetic algorithm after 40 generations having identified adaptive manager or user strategies).
+If termination conditions are not satisfied, then the `POPULATION` of individuals begins a new generation of crossover, mutation, cost constraint, fitness evaluation, and tournament selection (Figure 1).
 
@@
 ================================================================================
+
+Here we explain how candidate manager and user fitness strategies in a `POPULATION` array (see above) are calculated.
+We emphasise that the fitness functions used in GMSE v0.3.1.9 are intended to be heuristic tools for identifying reasonable manager and user behaviours.
+In practice, our fitness functions identify behaviours that are well-aligned with manager and user interests for harvesting or crop yield, but they are not intended to identify *optimal* decisions.
+This practical, metaheuristic approach is consistent with the objectives of Management Strategy Evaluation [@Bunnefeld2011], and is well-suited for the use of genetic algorithms [@Hamblin2013].
+@Luke2013 describes the metaheuristic approach more generally (original emphasis retained):
+
+> Metaheuristics are applied to *I know it when I see it* problems. They’re algorithms used to find
+answers to problems when you have very little to help you: you don’t know beforehand what the
+optimal solution looks like, you don’t know how to go about finding it in a principled way, you
+have very little heuristic information to go on, and brute-force search is out of the question because
+the space is too large. *But* if you’re given a candidate solution to your problem, you *can* test it and
+assess how good it is. That is, you know a good one when you see it.
+
+The above conditions for applying the metaheuristic approach are clearly satisfied for manager and user decisions, given the complexity of adaptive management and socio-ecological interactions.
 

### 2017-11-08 notebook/ms/SI1.Rmd
get the Luke reference working
Add more for manager fitness function

@@
 Values of column 5 in row 2 similarly identify whether landscape cell output is preferred by users to increase or decrease (managers do not currently have preferences for landscape output).
-Of special note is row 3 for `Manager_Actions`, which defines the marginal utility of resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the estimated abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca `r round((samp_ACTION[1,5,1] + -1*samp_ACTION[3,5,1]),2)`, so the manager will set policy in attempt to change the population size by ca `r round(samp_ACTION[3,5,1],2)` resources).
+Of special note is row 3 for `Manager_Actions`, which defines the manager's utility of resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the estimated abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca `r round((samp_ACTION[1,5,1] + -1*samp_ACTION[3,5,1]),2)`, so the manager will set policy in attempt to change the population size by ca `r round(samp_ACTION[3,5,1],2)` resources).
 Column 6 `U_land` defines whether or not the utility attached to the resource or landscape output depends on it being on a landscape cell that is owned by the acting user.
@@
 
+Manager fitness is affected by a manager's utility for resources and the projected change in resource abundance caused by the manager's policy.
+Manager utility is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model.
+Manager utility can therefore change in each GMSE time step as resource abundance (and its estimate) changes; when the estimated resource abundance is greater than `manage_target`, manager utility is negative, and when the estimated resource abundance is less than `manage_target`, manager utility is positive.
+
+To project the change in resource abundance as a consequence of policy, the manager needs to know the total number of actions of all types performed by users in the previous time step.
+
+To avoid having managers na&#0239;vely assuming that users will not perform an action because they did not perform it (or performed it very rarely) in the previous time step, managers always assume that users will perform at least `manager_caution` of each available action (default value of 1).
+
 User fitness function

### 2017-11-08 notebook/ms/SI1.Rmd
Few key equations

@@
 
-Key data structures used
+<a name="Key_data_structures_used">Key data structures used
 ================================================================================
@@
 Manager fitness is affected by a manager's utility for resources and the projected change in resource abundance caused by the manager's policy.
-Manager utility is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model.
+Manager utility is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model (see "[Key data structures used](#Key_data_structures_used)" above).
 Manager utility can therefore change in each GMSE time step as resource abundance (and its estimate) changes; when the estimated resource abundance is greater than `manage_target`, manager utility is negative, and when the estimated resource abundance is less than `manage_target`, manager utility is positive.
+To get fitness, first the change in resource abundance ($\Delta A$) predicted by the manager's policy is calculated, and the squared difference between $\Delta A$ and the manager utility ($U_{manager}$) is calculated to obtain a utility deviation for the individual $i$,
+
+$$ D_{i} = (\Delta A - U_{manager})^2.$$
+
+The value of $D_{i}$ increases as $\Delta A$ gets further from $U_{manager}$; i.e, $D_{i}$ is high when manager's set policy that does not get closer to their `manage_target` abundance.
+Fitness is defined by first finding the maximum $D_{i}$ value among all `ga_popsize` individuals ($D_{max}$), then subracting this value from $D_{i}$ for each individual,
+
+$$ Fitness_{i} = D_{max} - D_{i}.$$
 

### 2017-11-08 notebook/ms/SI1.Rmd
More on fitness function for managers

--- 2017-11-08 notebook/ms/SI1.Rmd
@@
 
-Manager fitness function
+Fitness function for managers
 --------------------------------------------------------------------------------
 
-User fitness function
+Individual fitness as calculated for managers ($F_{i}^{m}$) is affected by a manager's utility for resources and the projected change in resource abundance caused by the individual's policy (i.e., the contents of their `POPULATION` layer, specifically row 3).
+Manager utility for a resource ($U^{m}_{res}$) is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model (see "[Key data structures used](#Key_data_structures_used)" above).
+Manager utility can therefore change in each GMSE time step as resource abundance (and its estimate from the observation model) changes; when the estimated resource abundance is greater than `manage_target`, $U^{m}_{res}$ is negative, and when the estimated resource abundance is less than `manage_target`, $U^{m}$ is positive.
+To get individual fitness, first the change in resource abundance predicted by the individual's policy ($\Delta A_{i}$) is calculated, and the squared difference between $\Delta A_{i}$ and $U^{m}_{res}$ is calculated to obtain a utility deviation for the individual $i$,
+
+$$ D_{i} = (\Delta A_{i} - U^{m}_{res})^2. $$
+
+The value of $D_{i}$ increases as $\Delta A_{i}$ gets further from $U^{m}_{res}$; i.e, $D_{i}$ is high when $i$ sets a policy that does not get closer to the `manage_target` abundance.
+Fitness is defined by first finding the maximum $D_{i}$ value among all `ga_popsize` individuals ($D_{max}$), then subtracting this value from $D_{i}$ for each individual,
+
+$$ F_{i} = D_{max} - D_{i}. $$
+
+We have explained how $U^{m}_{res}$ is calculated in the [above section on key data structures](#Key_data_structures_used).
+We now explain in more detail how individuals in the genetic algorithm calculate their actions will affect $\Delta A_{i}$.
+Below explains calculations for a single individual $i$ in the genetic algorithm, so we will drop the use of subscripts $i$ to refer to individuals; instead we use subscripts $j$ to refer to actions (e.g., scaring, culling, etc.).
+
+To predict change resource abundance as a consequence of policy, an individual first needs to know the total number of actions of all types $j$ performed by users in the previous time step ($X_{j}^{old}$), and the cost of performing each action ($C_{old}^{j}$).
+This information is collected from `ACTION` and `COST` arrays.
+The individual then needs to predict how their policy (i.e., the costs that they set users to perform an action) will affect the new total number of each action performed ($X_{j}^{new}$).
+To do this, the individual assumes that total user actions performed under their policy will change in proportion to that of the old policy.
+The predicted total number of a particular action $j$ performed is thereby calculated as,
+
+$$ X_{j}^{new} =  (X_{j}^{old} + 1) \frac{C^{old}_{j}}{C^{new}_{j}}. $$
+
+The variable $C_{new}^{j}$ defines the new cost set by the individual for action $j$.
+A value of 1 is added to ($X_{j}^{old}$) to model some degree of caution by the manager (this can be changed from the default 1 using `manage_caution`), especially so that managers do not na&#0239;vely assume that users will not perform an action just because they did not perform it in the previous time step.
+Otherwise, if $X_{j}^{old} = 0$, then the manager would always assume that a change in the cost of an action would have no affect on the number of times the action was performed by users; a value of 1 assumes that at least one user will perform the action in the new time step.
+
+The predicted consequences of $X^{new}_{j}$ for resource abundance differ for each possible action.
+For each action, no consequence is predicted if the policy is not allowed by a simulation of GMSE (e.g., `scaring = FALSE`).
+
+Fitness function for users
 --------------------------------------------------------------------------------

### 2017-11-09 notebook/ms/SI1.Rmd
Finish manager fitness function, move to user, and minor tweaks

@@
 As of GMSE version 0.3.1.9, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step.
-Therefore, one run of the genetic algorithm occurs to simulate the manager's policy setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users follow identically.
+Therefore, one run of the genetic algorithm occurs to simulate the manager's policy-setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users copy).
 Each run of the genetic algorithm mimics the evolution by natural selection of a population of potential manager or user strategies over multiple generations, with the highest fitness strategy in the terminal generation being selected as the one that the manager or user decides to implement.
@@
 All other values for `Act` are meaningless in GMSE 0.3.1.9, but might be expanded upon in future versions to allow for modification of specific user costs enacted by managers (i.e., managers having different policies for different users) or other users (e.g., users increasing the costs of other users' actions due to conflict or cooperation).
-Similarly, columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in GMSE v0.3.1.9 (i.e., only one type of resource is permitted).
+For the rest of this supporting information, we will therefore focus only on rows 1-3 of `ACTION`.
+
+Columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in predefined GMSE v0.3.1.9 manager and user sub-models (i.e., only one type of resource is permitted).
 Future versions might allow for different resource types (e.g., `Type_1` might be used to designate species, and `Type_2` and `Type_3` could designate stage or sex).
-For the rest of this supporting information, we will therefore focus only on rows 1-3 of `ACTION`.
 Column 5 `Util.` of `ACTION` defines the utility associated with the resource (where `Act = -2`) or landscape (where `Act = -1`).
@@
 Values of column 5 in row 2 similarly identify whether landscape cell output is preferred by users to increase or decrease (managers do not currently have preferences for landscape output).
-Of special note is row 3 for `Manager_Actions`, which defines the manager's utility of resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the estimated abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca `r round((samp_ACTION[1,5,1] + -1*samp_ACTION[3,5,1]),2)`, so the manager will set policy in attempt to change the population size by ca `r round(samp_ACTION[3,5,1],2)` resources).
+Of special note is row 3 for `Manager_Actions`, which defines the *current* manager's utility of resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the most recent estimate of resource abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca `r round((samp_ACTION[1,5,1] + -1*samp_ACTION[3,5,1]),2)`, so the manager will set policy in attempt to change the population size by ca `r round(samp_ACTION[3,5,1],2)` resources).
 Column 6 `U_land` defines whether or not the utility attached to the resource or landscape output depends on it being on a landscape cell that is owned by the acting user.
@@
 Elements of `COST` define how many units from the `manager_budget` or `user_budget` are needed to perform a single action; a `minimum_cost` for actions is defined as an argument in GMSE (10 by default).
-All values in `COST` columns 1-7 are set to 10001, one higher than the highest possible `manager_budget` or `user_budget`, so neither can affect resource types or utilities.
+All values in `COST` columns 1-7 are set to 10001, one higher than the highest possible `manager_budget` or `user_budget`, so neither managers nor users can affect resource types or utilities.
 Columns 8-13 are also set to 10001, except where actions are allowed.
@@
 
-Note that in default GMSE parameters, `culling = TRUE`, but all other actions are false.
+Note that in default GMSE parameters, `culling = TRUE`, but all other actions are set to `FALSE`.
 Hence the `Cull` column 9 is the only column besides column 13 `None` in which cost is less than 10001.
 Manager's actions in `ACTION` directly affect the cost of users performing one of the five possible actions on resources (columns 8-12).
-This can be verified in `ACTION` where the manager has set the cost of scaring to `r samp_ACTION[3,9,1]`, and the corresponding `COST` of resource culling (row 1) is `r samp_COST[1,9,2]` for both users.
+This can be verified in `ACTION` where the manager has set the cost of culling to `r samp_ACTION[3,9,1]`, and the corresponding `COST` of resource culling (row 1) is `r samp_COST[1,9,2]` for both users.
 The cost of the manager affecting the cost of user actions is always set to the `minimum_cost`; here the default `r samp_COST[3,9,1]` is used.
@@
 
-The genetic algorithm updates a single layer of the `ACTION` array, which defines to the decisions of a single agent (either the manager or a user).
+The genetic algorithm updates a single layer of the `ACTION` array, which defines the decisions of a single agent (either the manager or a user).
 The corresponding layer of the `COST` array remains unchanged, and serves only to ensure that `ACTION` values do not exceed `manager_budget` or `user_budget` for managers and users, respectively.
-The genetic algorithm proceeds by first initialising a large population of new `ACTION` layers.
-In each generation, these layers crossover and mutate, generating variation in agent decisions; costs constrain this variation from exceeding a maximum budget, then the fitness of each layer is evaluated based on how much it increases the agent's utility.
-A tournament is used to select high fitness layers, and these selected layers become the new generation of layers; generations continue until a minimum number of generations (`ga_mingen`) have passed and a convergence criteria is satisfied such that the increase in mean fitness from the previous generation is below the threshold `converge_crit` (Figure 1).
+The genetic algorithm proceeds by first initialising a large (but temporary) population of new `ACTION` layers.
+In each generation, these layers crossover and mutate, generating variation in potential agent decisions; costs constrain this variation from exceeding a maximum budget, then the fitness of each layer is evaluated based on how the layer is predicted to affect resources or landscape output to which the agent has assigned some utility.
+A tournament is used to select high fitness layers, and these selected layers become the new generation of layers; generations continue until a minimum number of generations (`ga_mingen`) have passed and a convergence criteria is satisfied such that the increase in mean fitness from the previous generation is below the threshold `converge_crit` (Figure S1-1).
 
@@
 
-**Figure 1:** Conceptual overview of the GMSE genetic algorithm
+**Figure S1-1:** Conceptual overview of the GMSE genetic algorithm
 
@@
 The number of exact replicates initialised is set using `ga_seedrep` (if `ga_seedrep` $\geq$ `ga_popsize`, then all individuals are seeded as replicates).
-After the `POPULATION` of `ga_popsize` individuals is initialised, a loop simulating the adaptive evolution of `POPULATION` in non-overlapping generations begins.
+After the `POPULATION` of `ga_popsize` individuals is initialised, a loop simulating the adaptive evolution of `POPULATION` in non-overlapping generations begins (see Figure S1-1 above).
 
@@
 In contrast, user actions need to consider the consequences of their decisions on resource abundance or landscape output.
-Fitness of an individual is defined by a real number that increases with the degree to which an individual's actions are predicted to increase their utility (recall that managers and users assign resources or landscape output a utility value).
+Individual fitness is defined by a real number that increases with the degree to which an individual's actions are predicted to increase entities of positive utility and decrease entities of negative utility (recall that managers and users assign resources or landscape output a utility value).
 Details for how fitness is calculated are provided below.
@@
 
-After each individual in `POPULATION` is assigned a fitness, a tournament is used to select individuals of higher fitness.
+After each individual in `POPULATION` is assigned a fitness, a tournament is used to select individuals.
 Tournament selection is an especially flexible, non-parametric method that samples a subset of individuals from the total population and chooses the fittest of the subset for replacement [@Hamblin2013].
@@
 The genetic algorithm terminates if and only if a minimum number of generations has passed (`ga_mingen`) and a convergence criteria (`converge_crit`) is satisfied.
-The convergence criteria checks the difference between the mean fitness of individuals in the new generation versus the previous generation; if this difference is less than `converge_crit`, then termination does not occur (it is usually fine to use the default GMSE `converge_crit = 100` and `ga_mingen = 40`, which nearly always terminates the genetic algorithm after 40 generations having identified adaptive manager or user strategies).
-If termination conditions are not satisfied, then the `POPULATION` of individuals begins a new generation of crossover, mutation, cost constraint, fitness evaluation, and tournament selection (Figure 1).
+The convergence criteria checks the difference between the mean fitness of individuals in the new generation versus the previous generation; if this difference is greater than `converge_crit`, then termination does not occur (this prevents termination from occuring while fitness is still increasing, though it is usually fine to use the default GMSE `converge_crit = 100` and `ga_mingen = 40`, which nearly always terminates the genetic algorithm after 40 generations having identified adaptive manager or user strategies).
+If termination conditions are not satisfied, then the `POPULATION` of individuals begins a new generation of crossover, mutation, cost constraint, fitness evaluation, and tournament selection (Figure S1-1).
 
@@
 
-Here we explain how candidate manager and user fitness strategies in a `POPULATION` array (see above) are calculated.
+Here we explain how the fitnesses of candidate manager and user strategies in a `POPULATION` array (see above) are calculated.
 We emphasise that the fitness functions used in GMSE v0.3.1.9 are intended to be heuristic tools for identifying reasonable manager and user behaviours.
@@
 Manager utility for a resource ($U^{m}_{res}$) is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model (see "[Key data structures used](#Key_data_structures_used)" above).
-Manager utility can therefore change in each GMSE time step as resource abundance (and its estimate from the observation model) changes; when the estimated resource abundance is greater than `manage_target`, $U^{m}_{res}$ is negative, and when the estimated resource abundance is less than `manage_target`, $U^{m}$ is positive.
-To get individual fitness, first the change in resource abundance predicted by the individual's policy ($\Delta A_{i}$) is calculated, and the squared difference between $\Delta A_{i}$ and $U^{m}_{res}$ is calculated to obtain a utility deviation for the individual $i$,
+Manager utility can therefore change in each GMSE time step as resource abundance (and its estimate from the observation model) changes; when the estimated resource abundance is greater than `manage_target`, $U^{m}_{res}$ is negative, and when the estimated resource abundance is less than `manage_target`, $U^{m}_{res}$ is positive.
+To get individual fitness, first the change in resource abundance predicted by the individual's policy ($\Delta A_{i}$) is calculated, and the squared difference between $\Delta A_{i}$ and $U^{m}_{res}$ is calculated to obtain a utility deviation ($D_{i}$) for the individual $i$,
 
@@
 
-The value of $D_{i}$ increases as $\Delta A_{i}$ gets further from $U^{m}_{res}$; i.e, $D_{i}$ is high when $i$ sets a policy that does not get closer to the `manage_target` abundance.
-Fitness is defined by first finding the maximum $D_{i}$ value among all `ga_popsize` individuals ($D_{max}$), then subtracting this value from $D_{i}$ for each individual,
+The value of $D_{i}$ increases as $\Delta A_{i}$ gets further from $U^{m}_{res}$; i.e, $D_{i}$ is high when $i$ sets a policy is not predicted to get closer to the `manage_target` abundance.
+Fitness is defined by first finding the maximum $D_{i}$ value among all `ga_popsize` individuals ($D_{max}$), then subtracting $D_{i}$ from this value for each individual,
 
-$$ F_{i} = D_{max} - D_{i}. $$
+$$ F^{m}_{i} = D_{max} - D_{i}. $$
 
 We have explained how $U^{m}_{res}$ is calculated in the [above section on key data structures](#Key_data_structures_used).
-We now explain in more detail how individuals in the genetic algorithm calculate their actions will affect $\Delta A_{i}$.
-Below explains calculations for a single individual $i$ in the genetic algorithm, so we will drop the use of subscripts $i$ to refer to individuals; instead we use subscripts $j$ to refer to actions (e.g., scaring, culling, etc.).
+We now explain in more detail how individuals in the genetic algorithm calculate how their actions will affect $\Delta A_{i}$.
+These calculations apply to all individuals identically, so we note that the subscript $i$ is dropped; instead, the subscript $j$ refers to actions (e.g., scaring, culling, etc.).
 
-To predict change resource abundance as a consequence of policy, an individual first needs to know the total number of actions of all types $j$ performed by users in the previous time step ($X_{j}^{old}$), and the cost of performing each action ($C_{old}^{j}$).
+To predict change resource abundance as a consequence of policy, an individual first needs to know the total number of actions of all types $j$ performed by users in the previous time step ($X_{j}^{old}$), and the cost of performing each action ($C^{old}_{j}$).
 This information is collected from `ACTION` and `COST` arrays.
-The individual then needs to predict how their policy (i.e., the costs that they set users to perform an action) will affect the new total number of each action performed ($X_{j}^{new}$).
+The individual then needs to predict how their policy (i.e., the costs that they set for users to perform an action) will affect the new total number of each action performed ($X_{j}^{new}$).
 To do this, the individual assumes that total user actions performed under their policy will change in proportion to that of the old policy.
@@
 
-The variable $C_{new}^{j}$ defines the new cost set by the individual for action $j$.
+The variable $C^{new}_{j}$ defines the new cost set by the individual for action $j$.
 A value of 1 is added to ($X_{j}^{old}$) to model some degree of caution by the manager (this can be changed from the default 1 using `manage_caution`), especially so that managers do not na&#0239;vely assume that users will not perform an action just because they did not perform it in the previous time step.
-Otherwise, if $X_{j}^{old} = 0$, then the manager would always assume that a change in the cost of an action would have no affect on the number of times the action was performed by users; a value of 1 assumes that at least one user will perform the action in the new time step.
+Otherwise, if $X_{j}^{old} = 0$, then the manager would always assume that a change in the cost of an action would have no effect on the number of times the action was performed by users; a value of 1 assumes that at least one user will perform the action in the new time step.
 
 The predicted consequences of $X^{new}_{j}$ for resource abundance differ for each possible action.
-For each action, no consequence is predicted if the policy is not allowed by a simulation of GMSE (e.g., `scaring = FALSE`).
+For each action, no consequence is predicted if the policy is not allowed by a simulation of GMSE (e.g., `culling = FALSE`).
+For allowed actions, the parameter `manager_sense` ($\sigma$) modulates predicted consequences for abundance by some factor; this is useful because not all actions attempted by users will be realised, and a value of $\sigma = 1$ tends to greatly overestimate how much the actions attempted by users will actually translate to a change in resource abundance.
+In practice, the default $\sigma = 0.1$ performs well.
+Allowed actions are predicted by managers to have the following effects:
+
+- `scaring` is assumed to be nonlethal and therefore have an effect of zero on the total abundance of a resource.
+- `culling` is assumed to have an effect of $-\sigma(1 + \lambda)$, where $\lambda$ is the GMSE argument `lambda` that defines the baseline population growth rate of resources.
+- `castration` is assumed to have an effect of $-\sigma\lambda$.
+- `feeding` is assumed to have an effect of $\lambda$.
+- `help_offspring` is assumed to have an effect of $\sigma$.
+
+These effects cannot be altered directly in `gmse` or `gmse_apply` (though parameter values can of course be changed using `manager_sense` and `lambda` arguments), but future versions of GMSE might include different predicted effects to increase precision or allow for multiple resource types or different actions.
+The summation of $X^{new}_{j}$ for all actions defines the predicted change in resource abundance caused by the policy of an individual $i$, $\Delta A_{i}$.
 
@@
 --------------------------------------------------------------------------------
+
+The previous section described the fitness function applied when individual's fitness was evaluated for managers; here we explain a separate fitness function that is applied when individuals are instead evaluated for users.
+Individual fitness as calculated for users ($F_{i}^{u}$) is affected by a user's utility for resources and landscape output (i.e., the predicted summed values of the landscape cells that they own), and the project change in each caused by the user's actions (i.e., the contents of their `POPULATION` layer, specifically rows 1 and 2).
 

### 2017-11-09 notebook/ms/SI1.Rmd
First draft

@@
 In practice, our fitness functions identify behaviours that are well-aligned with manager and user interests for harvesting or crop yield, but they are not intended to identify *optimal* decisions.
-This practical, metaheuristic approach is consistent with the objectives of Management Strategy Evaluation [@Bunnefeld2011], and is well-suited for the use of genetic algorithms [@Hamblin2013].
+This practical, metaheuristic approach is consistent with the objectives of management strategy evaluation [@Bunnefeld2011], and is well-suited for the use of genetic algorithms [@Hamblin2013].
 @Luke2013 describes the metaheuristic approach more generally (original emphasis retained):
@@
 
-The above conditions for applying the metaheuristic approach are clearly satisfied for manager and user decisions, given the complexity of adaptive management and socio-ecological interactions.
+Given the complexity of adaptive management and socio-ecological interactions, the above conditions for applying the metaheuristic approach are clearly satisfied for manager and user decisions.
 
@@
 
-- `scaring` is assumed to be nonlethal and therefore have an effect of zero on the total abundance of a resource.
-- `culling` is assumed to have an effect of $-\sigma(1 + \lambda)$, where $\lambda$ is the GMSE argument `lambda` that defines the baseline population growth rate of resources.
-- `castration` is assumed to have an effect of $-\sigma\lambda$.
-- `feeding` is assumed to have an effect of $\lambda$.
-- `help_offspring` is assumed to have an effect of $\sigma$.
+- `scaring` is assumed to be nonlethal and therefore have no effect on resource number.
+- `culling` decreases resource number by $\sigma(1 + \lambda)$, where $\lambda$ is the GMSE argument `lambda` that defines the baseline population growth rate of resources.
+- `castration` decreases resource number by $\sigma\lambda$.
+- `feeding` increases resource number by $\lambda$.
+- `help_offspring` increases resource number by $\sigma$.
 
@@
 The previous section described the fitness function applied when individual's fitness was evaluated for managers; here we explain a separate fitness function that is applied when individuals are instead evaluated for users.
-Individual fitness as calculated for users ($F_{i}^{u}$) is affected by a user's utility for resources and landscape output (i.e., the predicted summed values of the landscape cells that they own), and the project change in each caused by the user's actions (i.e., the contents of their `POPULATION` layer, specifically rows 1 and 2).
+Individual fitness as calculated for users ($F_{i}^{u}$) is affected by a user's utility for resources ($U^{u}_{res}$) and landscape output ($U^{u}_{land}$), and the predicted change in each caused by the user's actions ($\Delta A_{i}$ and $\Delta L_{i}$ for predicted change in resource abundance and summed values of the landscape cells owned by $i$, respectively).
+Individual fitness is defined for users below,
 
-Thanks for the clarification regarding the equation.
-I'll try to answer as best as I can -- apologies if this has been unclear.
-At the broadest scale, the equation for user fitness would be on L367 in the strategy_fitness function ( https://github.com/bradduthie/gmse/blob/master/src/game.c#L376 ).
-Here's what's going on: Users are predicting how their actions will change the quantities of things in the model (either resources or landscape output), and these changes are individually multiplied by the users' utilities for that thing.
-The change multiplied by utility for each thing is summed across all things to get a value for fitness.
-Note that positive change times positive utility, and negative change times negative utility, will increase fitness (i.e., increasing the thing users want more of and decreasing the things they want less of).
-Hence, an equation describing user fitness would be the below,
+$$F_{i}^{u} = \Delta A_{i} U^{u}_{res} + \Delta L_{i} U^{u}_{land} $$.
 
-$$ F_{user} = \sum_{i=1}^{N} \Delta A_{i} \times U_{i} $$.
+Note that $F_{i}^{u}$ increases when $\Delta A_{i}$ and $\Delta L_{i}$ are of the same sign as $U^{u}_{res}$ and $U^{u}_{land}$, respectively.
+Further, in GMSE v0.3.1.9, only one term of the equation is nonzero.
+When `land_ownership = FALSE` (default, modelling users that harvest resources), $U^{u}_{res} = -1$ and $U^{u}_{land} = 0$, and when `land_ownership = TRUE`, $U^{u}_{res} = -1$ and $U^{u}_{land} = 100$ (modelling farmers trying to increase crop yield).
+Hence users only have a single objective of either decreasing resource abundance or increasing landscape output, though landscape output might be increased indirectly by decreasing resource abundance if `resource_consume` is greater than zero.
 
-Where $F_{user}$ is user fitness, $N$ is the total number of things that might be of interest (at the moment $N = 2$ in GMSE, one resource and, potentially, one landscape value), $\Delta A_{i}$ is the change in the abundance of thing $i$, and $U_{i}$ is the utility of thing $i$ from the perspective of the user (apologies for the LaTeX code -- attached a PNG of the conversion).
-I want to stress though that I would not consider this equation to be central to the GMSE framework -- if someone else has a better approach for defining fitness, or defining any of the terms listed above, or wants to expand upon it to include new things, then that would be awesome!
-The above just works well as a heuristic tool to get users to act in such a way as to maximise their interests in harvesting or getting more crop yield (as is my intent), but it's not based on first principles and I don't claim it to be particularly special.
+User actions are predicted to affect resources in the following way:
 
-The values of $\Delta A_{i}$ are calculated for resources and the landscape in the functions res_to_counts and land_to_counts, respectively (and $U_{i}$ is specified a priori in the model depending on other parameters -- namely land_ownership).
-Again, a bit of heuristic is needed here because there cannot be any perfect way of exactly predicting how a users actions will increase or decrease resources -- there are too many complex factors (e.g., behaviour of other stakeholders, demographic stochasticity, movement of resources on the landscape, and interactions between resources and the landscape).
-Even if we could include all of these things somehow, it would be a bit unrealistic in that real stakeholders would never have this much information.
-The predicted direct effect of actions on resources is shown in lines 268-272 ( https://github.com/bradduthie/gmse/blob/master/src/game.c#L268 ), and the array 'jaco' (a sort of Jacobian matrix) accounts for interactions between landscape and resources on line 286.
-Something similar happens in the land_to_counts function.
-The manager's genetic algorithm works in a similar way (the above equation applies), but with the need to dynamically update utility values based on current resource abundance, and to account for the predicted actions of users in finding $\Delta A_{i}$.
+- `scaring` decreases resource number by $1$.
+- `culling` decreases resource number by $1 + \lambda$.
+- `castration` decreases resource number by $\lambda$.
+- `feeding` increases resource number by $\lambda$.
+- `help_offspring` increases resource number by $\sigma$.
+
+The number of each action performed is multiplied by its effect, and the sum of all these products is the predicted $\Delta A_{i}$,
+
+$$\Delta A_{i} = (\lambda) Feeds + Helps -Scares - (1 + \lambda) Culls - (\lambda) Castrations.$$
+
+There are only two possible actions that users can take to directly affect landscape output, tending crops (`tend_crops`) and killing crops (`kill_crops`).
+The increase in landscape output is modulated by the parameter `tend_crop_yld` ($\phi$).
+User actions are therefore predicted to have the following effects for one landscape cell:
+
+- `tend_crops` will increase landscape output by $\phi$.
+- `kill_crops` will decrease landscape output by 1 (since the output of a cell is 1 by default, this action removes all output on a landscape cell).
+
+Actions on resources can also have indirect affects on $\Delta L_{i}$ when resources consume output on the landscape (i.e., when `res_consume` is greater than zero), which we define as $r$.
+The predicted $\Delta L_{i}$ is therefore,
+
+$$\Delta L_{i} = (\phi)Tends - Kills + r\Delta A_{i}.$$
+
+That is, the change in landscape output equals the increase in output from tending crops, minus the number of crops destroyed, plus the change in resource abundance times the effect that resource abundance has on landscape output.
+
+Future development of fitness functions
+--------------------------------------------------------------------------------
+
+The fitness functions defined above are useful heuristics for simulating manager and user decision-making in a way that produces realistic, *I know it when I see it*, strategies.
+Future versions of GMSE might improve upon these heuristics to generate more accurate or more realistic models of human decision making.
+Such improvements could incorporate additional information such as memory of actions from multiple past time steps, or a continually updated estimate for how actions are predicted to affect resource abundance in a simulation (e.g., through a dynamic `manager_sense`).
+Alternatively, future improvements could usefully incorporate knowledge of human decision making collected from empirical observation of human behaviour during conservation conflicts.
+While such possibilities could be useful for future GMSE modelling, repeated simulations demonstrate the ability of the current GMSE genetic algorithm to find adaptive strategies for managers attempting to keep resources at target abundance, and users attempting to maximise their harvests or crop yields.
+It should therefore be useful as a tool for modelling manager and user decisions in a generalised management strategy evaluation framework.
 

### 2017-11-09 notebook/ms/SI1.Rmd
More edits

@@
 Manager utility for a resource ($U^{m}_{res}$) is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model (see "[Key data structures used](#Key_data_structures_used)" above).
-Manager utility can therefore change in each GMSE time step as resource abundance (and its estimate from the observation model) changes; when the estimated resource abundance is greater than `manage_target`, $U^{m}_{res}$ is negative, and when the estimated resource abundance is less than `manage_target`, $U^{m}_{res}$ is positive.
-To get individual fitness, first the change in resource abundance predicted by the individual's policy ($\Delta A_{i}$) is calculated, and the squared difference between $\Delta A_{i}$ and $U^{m}_{res}$ is calculated to obtain a utility deviation ($D_{i}$) for the individual $i$,
+Manager utility can therefore change in each GMSE time step as estimated resource abundance changes; when the estimated resource abundance is greater than `manage_target`, $U^{m}_{res}$ is negative, and when the estimated resource abundance is less than `manage_target`, $U^{m}_{res}$ is positive.
+To get individual fitness, first the change in resource abundance predicted by the individual's policy ($\Delta A_{i}$) is calculated, then the squared difference between $\Delta A_{i}$ and $U^{m}_{res}$ is calculated to obtain a utility deviation ($D_{i}$) for the individual $i$,
 
@@
 
-The value of $D_{i}$ increases as $\Delta A_{i}$ gets further from $U^{m}_{res}$; i.e, $D_{i}$ is high when $i$ sets a policy is not predicted to get closer to the `manage_target` abundance.
+The value of $D_{i}$ increases as $\Delta A_{i}$ gets further from $U^{m}_{res}$; i.e, $D_{i}$ is high when $i$ sets a policy that is not predicted to get closer to the `manage_target` abundance.
 Fitness is defined by first finding the maximum $D_{i}$ value among all `ga_popsize` individuals ($D_{max}$), then subtracting $D_{i}$ from this value for each individual,
@@
 We now explain in more detail how individuals in the genetic algorithm calculate how their actions will affect $\Delta A_{i}$.
-These calculations apply to all individuals identically, so we note that the subscript $i$ is dropped; instead, the subscript $j$ refers to actions (e.g., scaring, culling, etc.).
 
-To predict change resource abundance as a consequence of policy, an individual first needs to know the total number of actions of all types $j$ performed by users in the previous time step ($X_{j}^{old}$), and the cost of performing each action ($C^{old}_{j}$).
+To predict change resource abundance as a consequence of policy, an individual first needs to know the total number of actions of all types $j$ (e.g., scaring, culling, etc.) performed by users in the previous time step ($X_{\bullet, j}$), and the cost of performing each action ($C_{\bullet, j}$).
 This information is collected from `ACTION` and `COST` arrays.
-The individual then needs to predict how their policy (i.e., the costs that they set for users to perform an action) will affect the new total number of each action performed ($X_{j}^{new}$).
+The individual then needs to predict how their policy (i.e., the costs that they set for users to perform an action) will affect the new total number of each action performed ($X_{i,j}$).
 To do this, the individual assumes that total user actions performed under their policy will change in proportion to that of the old policy.
-The predicted total number of a particular action $j$ performed is thereby calculated as,
+The predicted total number of a particular action $j$ performed under the policy of $i$ is thereby calculated as,
 
-$$ X_{j}^{new} =  (X_{j}^{old} + 1) \frac{C^{old}_{j}}{C^{new}_{j}}. $$
+$$ X_{i,j} =  (X_{\bullet, j} + 1) \frac{C_{\bullet, j}}{C_{i,j}}. $$
 
-The variable $C^{new}_{j}$ defines the new cost set by the individual for action $j$.
-A value of 1 is added to ($X_{j}^{old}$) to model some degree of caution by the manager (this can be changed from the default 1 using `manage_caution`), especially so that managers do not na&#0239;vely assume that users will not perform an action just because they did not perform it in the previous time step.
-Otherwise, if $X_{j}^{old} = 0$, then the manager would always assume that a change in the cost of an action would have no effect on the number of times the action was performed by users; a value of 1 assumes that at least one user will perform the action in the new time step.
+The variable $C_{i,j}$ defines the new cost set by the individual $i$ for action $j$.
+A value of 1 is added to ($X_{\bullet, j}$) to model some degree of caution by the manager (this can be changed from the default 1 using `manage_caution`), especially so that managers do not na&#0239;vely assume that users will not perform an action just because they did not perform it in the previous time step.
+Otherwise, if $X_{\bullet, j} = 0$, then the manager would always assume that a change in the cost of an action would have no effect on the number of times the action was performed by users; a value of 1 assumes that at least one user will perform the action in the new time step.
 
-The predicted consequences of $X^{new}_{j}$ for resource abundance differ for each possible action.
+The predicted consequences of $X_{i,j}$ for resource abundance differ for each possible action.
 For each action, no consequence is predicted if the policy is not allowed by a simulation of GMSE (e.g., `culling = FALSE`).
@@
 These effects cannot be altered directly in `gmse` or `gmse_apply` (though parameter values can of course be changed using `manager_sense` and `lambda` arguments), but future versions of GMSE might include different predicted effects to increase precision or allow for multiple resource types or different actions.
-The summation of $X^{new}_{j}$ for all actions defines the predicted change in resource abundance caused by the policy of an individual $i$, $\Delta A_{i}$.
+The summation of $X_{i,j}$ for all actions defines the predicted change in resource abundance caused by the policy of an individual $i$, $\Delta A_{i}$.
 
@@
 
-$$F_{i}^{u} = \Delta A_{i} U^{u}_{res} + \Delta L_{i} U^{u}_{land} $$.
+$$F_{i}^{u} = \Delta A_{i} U^{u}_{res} + \Delta L_{i} U^{u}_{land}. $$
 
@@
 Further, in GMSE v0.3.1.9, only one term of the equation is nonzero.
-When `land_ownership = FALSE` (default, modelling users that harvest resources), $U^{u}_{res} = -1$ and $U^{u}_{land} = 0$, and when `land_ownership = TRUE`, $U^{u}_{res} = -1$ and $U^{u}_{land} = 100$ (modelling farmers trying to increase crop yield).
+When `land_ownership = FALSE` (default, modelling users that harvest resources), $U^{u}_{res} = -1$ and $U^{u}_{land} = 0$, and when `land_ownership = TRUE`, $U^{u}_{res} = 0$ and $U^{u}_{land} = 100$ (modelling farmers trying to increase crop yield).
 Hence users only have a single objective of either decreasing resource abundance or increasing landscape output, though landscape output might be increased indirectly by decreasing resource abundance if `resource_consume` is greater than zero.
@@
 
-Actions on resources can also have indirect affects on $\Delta L_{i}$ when resources consume output on the landscape (i.e., when `res_consume` is greater than zero), which we define as $r$.
-The predicted $\Delta L_{i}$ is therefore,
+Actions on resources can also have indirect effects on $\Delta L_{i}$ when resources consume output on the landscape; we define the value `res_consume` as $r$. The predicted $\Delta L_{i}$ is then,
 
@@
 Future versions of GMSE might improve upon these heuristics to generate more accurate or more realistic models of human decision making.
-Such improvements could incorporate additional information such as memory of actions from multiple past time steps, or a continually updated estimate for how actions are predicted to affect resource abundance in a simulation (e.g., through a dynamic `manager_sense`).
+Such improvements could incorporate additional information such as memory of actions from multiple past time steps, or a continually updated estimate for how actions are predicted to affect resource abundance or landscape output in a simulation (e.g., through a dynamic `manager_sense`).
 Alternatively, future improvements could usefully incorporate knowledge of human decision making collected from empirical observation of human behaviour during conservation conflicts.
 While such possibilities could be useful for future GMSE modelling, repeated simulations demonstrate the ability of the current GMSE genetic algorithm to find adaptive strategies for managers attempting to keep resources at target abundance, and users attempting to maximise their harvests or crop yields.
-It should therefore be useful as a tool for modelling manager and user decisions in a generalised management strategy evaluation framework.
+It is therefore useful as a tool for modelling manager and user decisions in a generalised management strategy evaluation framework.
 

### 2017-11-10 notebook/ms/SI1.Rmd
More edits

@@
     reference_docx: docx_template.docx
-linestretch: 1
-subtitle: Supporting Information 1
 header-includes:
@@
 - \linenumbers
+linestretch: 1
 bibliography: gmse.bib
+subtitle: Supporting Information 1
 biblio-style: apalike
@@
 For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'generation' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the main text).
-Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.3.1.9 (future versions of GMSE might expand upon this framework).
+Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.3.1.9 (future versions of GMSE might expand upon this framework, and we highlight some of these potential avenues for expansion).
 We first explain the key data structures used, then provide an overview of how a population of strategies is initialised, and the  subsequent processes of crossover, mutation, cost constraint, fitness evaluation, tournament selection, and replacement.
@@
 
-<a name="Key_data_structures_used">Key data structures used
+<a name="Key_data_structures_used">Key data structures used</a>
 ================================================================================
@@
 Values of column 5 in row 2 similarly identify whether landscape cell output is preferred by users to increase or decrease (managers do not currently have preferences for landscape output).
-Of special note is row 3 for `Manager_Actions`, which defines the *current* manager's utility of resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the most recent estimate of resource abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca `r round((samp_ACTION[1,5,1] + -1*samp_ACTION[3,5,1]),2)`, so the manager will set policy in attempt to change the population size by ca `r round(samp_ACTION[3,5,1],2)` resources).
+Of special note is row 3 for `Manager_Actions`, which defines the *current* manager's utility for resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the most recent estimate of resource abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca `r round((samp_ACTION[1,5,1] + -1*samp_ACTION[3,5,1]),2)`, so the manager will set policy in attempt to change the population size by ca `r round(samp_ACTION[3,5,1],2)` resources).
 Column 6 `U_land` defines whether or not the utility attached to the resource or landscape output depends on it being on a landscape cell that is owned by the acting user.
 Related, column 7 `U_loc.` defines whether or not actions can be performed only on a landscape cell that is owned by the acting user.
-Hence values of columns 6 and 7 are binary, and affected by the `land_ownership` argument in `gmse`.
+Hence values of columns 6 and 7 are binary, and affected by the `land_ownership` argument in `gmse` and `gmse_apply`.
 Finally, columns 8-13 correspond to specific actions, either direct (where `Act < 0`) or indirect by setting policy (for row 3 of `Manager_Actions` where `Act = 1`).
@@
 dimnames(samp_COST)[[2]] <- c("Act", "Type_1", "Type_2", "Type_3", "Util.", "U_land", "U_loc.", "Scare", "Cull", "Castrate", "Feed", "Help_off", "None");
-dimnames(samp_COST)[[3]] <- c("Manager_Actions", "User_1_Actions", "User_2_Actions");
+dimnames(samp_COST)[[3]] <- c("Manager_Costs", "User_1_Costs", "User_2_Costs");
 print(samp_COST);
@@
 Manager's actions in `ACTION` directly affect the cost of users performing one of the five possible actions on resources (columns 8-12).
-This can be verified in `ACTION` where the manager has set the cost of culling to `r samp_ACTION[3,9,1]`, and the corresponding `COST` of resource culling (row 1) is `r samp_COST[1,9,2]` for both users.
+This can be verified in `ACTION` where the manager has set the cost of culling to `r samp_ACTION[3,9,1]` (row 3), and the corresponding `COST` of resource culling is `r samp_COST[1,9,2]` for both users (row 1).
 The cost of the manager affecting the cost of user actions is always set to the `minimum_cost`; here the default `r samp_COST[3,9,1]` is used.
@@
 The convergence criteria checks the difference between the mean fitness of individuals in the new generation versus the previous generation; if this difference is greater than `converge_crit`, then termination does not occur (this prevents termination from occuring while fitness is still increasing, though it is usually fine to use the default GMSE `converge_crit = 100` and `ga_mingen = 40`, which nearly always terminates the genetic algorithm after 40 generations having identified adaptive manager or user strategies).
+Due to the way in which fitness is calculated (see below), in practice, `converge_crit` currently applies only to users.
 If termination conditions are not satisfied, then the `POPULATION` of individuals begins a new generation of crossover, mutation, cost constraint, fitness evaluation, and tournament selection (Figure S1-1).
@@
 Given the complexity of adaptive management and socio-ecological interactions, the above conditions for applying the metaheuristic approach are clearly satisfied for manager and user decisions.
+With this in mind, we now explain the details of manager and user fitness functions; that is, how GMSE assesses whether or not a strategy is a good one.
 
@@
 
-$$\Delta L_{i} = (\phi)Tends - Kills + r\Delta A_{i}.$$
+$$\Delta L_{i} = (\phi)Tends - Kills - r\Delta A_{i}.$$
 
-That is, the change in landscape output equals the increase in output from tending crops, minus the number of crops destroyed, plus the change in resource abundance times the effect that resource abundance has on landscape output.
+That is, the change in landscape output equals the increase in output from tending crops, minus the number of crops destroyed, minus the change in resource abundance times the effect that resource abundance has on landscape output (note that if user actions decrease resource abundance, then this last term will be positive, increasing landscape output).
 

### 2017-11-10 notebook/ms/SI2.Rmd
New draft of GMSE paper

---
title: 'GMSE: an R package for generalised management strategy evaluation'
author: A. Bradley Duthie, Jeremy J. Cusack, Isabel L. Jones, Erlend B. Nilsen, Roc&#0237;o
  Pozo, O. Sarobidy Rakotonarivo, Bram Van Moorter, and Nils Bunnefeld
date: '`r Sys.Date()`'
output:
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
linestretch: 1
subtitle: Supporting Information 2
header-includes:
- \usepackage{lineno}
- \linenumbers
---

```{r, echo = FALSE}
library(GMSE);
```

Extended introduction to the GMSE apply function (`gmse_apply`)
================================================================================

The `gmse_apply` function is a flexible function that allows for user-defined sub-functions calling resource, observation, manager, and user models. Where such models are not specified, predefined GMSE submodels 'resource', 'observation', 'manager', and 'user' are run by default. Any type of sub-model (e.g., numerical,  individual-based) is permitted as long as the input and output are appropriately specified. Only one time step is simulated per call to `gmse_apply`, so the function must be looped for simulation over time. Where model parameters are needed but not specified, defaults from GMSE are used. In this Supporting Information, we demonstrate some uses of `gmse_apply`, and how it might be used to simulate myriad management scenarios *in silico*.

A simple run of `gmse_apply()` returns one generation of GMSE using predefined submodels and default parameter values.

```{r, eval = TRUE}
sim_1 <- gmse_apply();
```

For `sim_1`, the default 'basic' results are returned as below, which summarise key values for all submodels.

```{r, eval = TRUE}
print(sim_1);
```

Note that in the case above we have the total abundance of resources returned (`sim_1$resource_results`), the estimate of resource abundance from the observation function (`sim_1$observation_results`, the costs the manager sets for the only available action of culling (`sim_1$manager_results`), and the number of culls attempted by each user (`sim_1$user_results`). By default, only one resource type is used, but custom subfunctions could potentially allow for models with multiple resource types. Any custom subfunctions can replace GMSE predefined functions, provided that they have appropriately defined inputs and outputs (see [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf)). For example, we can define a very simple logistic growth function to send to `res_mod` instead.

```{r, eval = TRUE}
alt_res <- function(X, K = 2000, rate = 1){
    X_1 <- X + rate*X*(1 - X/K);
    return(X_1);
}
```

The above function takes in a population size of `X` and returns a value `X_1` based on the population intrinsic growth rate `rate` and carrying capacity `K`. Iterating the logistic growth model by itself under default parameter values with a starting population of 100 will cause the population to increase to carrying capacity in ca seven generations. The function can be substituted into `gmse_apply` to use it instead of the predefined GMSE resource model.

```{r, eval = TRUE}
sim_2 <- gmse_apply(res_mod = alt_res, X = 100, rate = 0.3);
```

The `gmse_apply` function will find the parameters it needs to run the `alt_res` function in place of the default resource function, either by running the default function values (e.g., `K = 2000`) or values specified directly into `gmse_apply` (e.g., `X = 100` and `rate = 0.3`). If an argument to a custom function is required but not provided either as a default or specified in `gmse_apply`, then an error will be returned. Results for the above `sim_2` are returned below.

```{r, eval = TRUE}
print(sim_2);
```

How `gmse_apply` integrates across submodels
--------------------------------------------------------------------------------

To integrate across different types of submodels, `gmse_apply` translates between vectors and arrays between each submodel. For example, because the default GMSE observation model requires a resource array with particular requirements for column identites, when a resource model subfunction returns a vector, or a list with a named element 'resource_vector', this vector is translated into an array that can be used by the observation model. Specifically, each element of the vector identifies the abundance of a resource type (and hence will usually be just a single value denoting abundance of the only focal population). If this is all the information provided, then a 'resource_array' will be made with default GMSE parameter values with an identical number of rows to the abundance value (floored if the value is a non-integer; non-default values can also be put into this transformation from vector to array if they are specified in `gmse_apply`, e.g., through an argument such as `lambda = 0.8`). Similarly, a `resource_array` is also translated into a vector after the default individual-based resource model is run, should the observation model require simple abundances instead of an array. The same is true of `observation_vector` and `observation_array` objects returned by observation models, of `manager_vector` and `manager_array` (i.e., `COST` in the `gmse` function) objects returned by manager models, and of `user_vector` and `user_array` (i.e., `ACTION` in the `gmse` function) objects returned by user models. At each step, a translation between the two is made, with necessary adjustments that can be tweaked through arguments to `gmse_apply` when needed. Alternative observation, manager, and user, submodels, for example, are defined below; note that each requires a vector from the preceding model.

```{r, eval = TRUE}
# Alternative observation submodel
alt_obs <- function(resource_vector){ 
    X_obs <- resource_vector - 0.1 * resource_vector;
    return(X_obs);
}

# Alternative manager submodel
alt_man <- function(observation_vector){
    policy <- observation_vector - 1000;
    if(policy < 0){
        policy <- 0;
    }
    return(policy);
}

# Alternative user submodel
alt_usr <- function(manager_vector){
    harvest <- manager_vector + manager_vector * 0.1;
    return(harvest);
}
```

All of these submodels are completely deterministic, so when run with the same parameter combinations, they produce replicable outputs.

```{r, eval = TRUE}
gmse_apply(res_mod = alt_res, obs_mod = alt_obs, 
           man_mod = alt_man, use_mod = alt_usr, X = 1000);
```

Note that the `manager_results` and `user_results` are ambiguous here, and can be interpreted as desired -- e.g., as total allowable catch and catches made, or as something like costs of catching set by the manager and effort to catching made by the user. Hence while manger output is set in terms of costs of performing each action, and user output is set in terms of action attempts, this need not be the case when using `gmse_apply` (though it should be recognised when using default GMSE manager and user functions). GMSE default submodels can be added in at any point.

```{r, eval = TRUE}
gmse_apply(res_mod = alt_res, obs_mod = observation, 
           man_mod = alt_man, use_mod = alt_usr, X = 1000)
```

It is possible to, for example, specify a simple resource and observation model, but then take advantage of the genetic algorithm to predict policy decisions and user actions. This can be done by using the default GMSE manager and user functions (written below explicitly, though this is not necessary).

```{r, eval = TRUE}
gmse_apply(res_mod = alt_res, obs_mod = alt_obs, 
           man_mod = manager, use_mod = user, X = 1000)
```

Running GMSE simulations by looping `gmse_apply`
--------------------------------------------------------------------------------

Instead of using the `gmse` function, multiple simulations of GMSE can be run by calling `gmse_apply` through a loop, reassigning outputs where necessary for the next generation. This is best accomplished using the argument `old_list`, which allows previous full results from `gmse_apply` to be reinserted into the `gmse_apply` function. The argument `old_list` is `NULL` by default, but can instead take the output of a previous full list return of  `gmse_apply`. This `old_list` produced when `get_res = Full` includes all data structures and parameter values necessary for a unique simulation of GMSE. An exampe of using `get_res` and `old_list` in tandem to loop `gmse_apply` is shown below.

```{r, eval = TRUE}
to_scare  <- FALSE;
sim_old   <- gmse_apply(scaring = to_scare, get_res = "Full", stakeholders = 6);
sim_sum_1 <- matrix(data = NA, nrow = 20, ncol = 7);
for(time_step in 1:20){
    sim_new               <- gmse_apply(scaring = to_scare, get_res = "Full", 
                                        old_list = sim_old);
    sim_sum_1[time_step, 1] <- time_step;
    sim_sum_1[time_step, 2] <- sim_new$basic_output$resource_results[1];
    sim_sum_1[time_step, 3] <- sim_new$basic_output$observation_results[1];
    sim_sum_1[time_step, 4] <- sim_new$basic_output$manager_results[2];
    sim_sum_1[time_step, 5] <- sim_new$basic_output$manager_results[3];
    sim_sum_1[time_step, 6] <- sum(sim_new$basic_output$user_results[,2]); 
    sim_sum_1[time_step, 7] <- sum(sim_new$basic_output$user_results[,3]); 
    sim_old                 <- sim_new;
}
colnames(sim_sum_1) <- c("Time", "Pop_size", "Pop_est", "Scare_cost", 
                         "Cull_cost", "Scare_count", "Cull_count");
print(sim_sum_1);
```

Note that one element of the full list `gmse_apply` output is the 'basic_output' itself, which is produced by default when `get_res = "basic"`. This is what is being used to store the output of `sim_new` into `sim_sum_1`. Next, we show how the flexibility of `gmse_apply` can be used to dynamically redefine simulation conditions.

Changing simulation conditions using `gmse_apply`
================================================================================

We can take advantage of `gmse_apply` to dynamically change parameter values mid-loop. For example, below shows the same code used in the previous example, but with a policy of scaring introduced on time step 10.

```{r, eval = TRUE}
to_scare  <- FALSE;
sim_old   <- gmse_apply(scaring = to_scare, get_res = "Full", stakeholders = 6);
sim_sum_2 <- matrix(data = NA, nrow = 20, ncol = 7);
for(time_step in 1:20){
    sim_new               <- gmse_apply(scaring = to_scare, get_res = "Full", 
                                        old_list = sim_old);
    sim_sum_2[time_step, 1] <- time_step;
    sim_sum_2[time_step, 2] <- sim_new$basic_output$resource_results[1];
    sim_sum_2[time_step, 3] <- sim_new$basic_output$observation_results[1];
    sim_sum_2[time_step, 4] <- sim_new$basic_output$manager_results[2];
    sim_sum_2[time_step, 5] <- sim_new$basic_output$manager_results[3];
    sim_sum_2[time_step, 6] <- sum(sim_new$basic_output$user_results[,2]); 
    sim_sum_2[time_step, 7] <- sum(sim_new$basic_output$user_results[,3]); 
    sim_old               <- sim_new;
    if(time_step == 10){
        to_scare <- TRUE;
    }
}
colnames(sim_sum_2) <- c("Time", "Pop_size", "Pop_est", "Scare_cost", 
                         "Cull_cost", "Scare_count", "Cull_count");
print(sim_sum_2);
```

Hence, in addition to the previously explained benefits of the flexible `gmse_apply` function, one particularly useful feature is that we can use it to study change in policy availability -- in the above case, what happens when scaring is suddenly introduced as a possible policy option. Similar things can be done, for example, to see how manager or user power changes over time. In the example below, users' budgets increase by 100 every time step, with the manager's budget remaining the same. The consequence of this increasing user budget is higher rates of culling and decreased population size.

```{r, eval = TRUE}
ub          <- 500;
sim_old     <- gmse_apply(get_res = "Full", stakeholders = 6, user_budget = ub);
sim_sum_3   <- matrix(data = NA, nrow = 20, ncol = 6);
for(time_step in 1:20){
    sim_new               <- gmse_apply(get_res = "Full", old_list = sim_old,
                                        user_budget = ub);
    sim_sum_3[time_step, 1] <- time_step;
    sim_sum_3[time_step, 2] <- sim_new$basic_output$resource_results[1];
    sim_sum_3[time_step, 3] <- sim_new$basic_output$observation_results[1];
    sim_sum_3[time_step, 4] <- sim_new$basic_output$manager_results[3];
    sim_sum_3[time_step, 5] <- sum(sim_new$basic_output$user_results[,3]);
    sim_sum_3[time_step, 6] <- ub;
    sim_old               <- sim_new;
    ub                    <- ub + 100;
}
colnames(sim_sum_3) <- c("Time", "Pop_size", "Pop_est", "Cull_cost", "Cull_count",
                         "User_budget");
print(sim_sum_3);
```

There is an important note to make about changing arguments to `gmse_apply` when `old_list` is being used: The function `gmse_apply` is trying to avoid a crash, so `gmse_apply` will accomodate parameter changes by rebuilding data structures if necessary. For example, if the number of stakeholders is changed (and by including an argument `stakeholders` to `gmse_apply`, it is assumed that stakeholders are changing even they are not), then a new array of agents will need to be built. If landscape dimensions are changed (or just include the argument `land_dim_1` or `land_dim_2`), then a new landscape willl be built. For most simulation purposes, this will not introduce any undesirable effect on simulation results, but it should be noted and understood when developing models.

### 2017-11-14 notebook/ms/ms.Rmd
New draft of GMSE paper
Make edits suggested by @nbunne -- Izzy next

@@
 title: 'GMSE: an R package for generalised management strategy evaluation'
-author: A. Bradley Duthie, Jeremy J. Cusack, Isabel L. Jones, Erlend B. Nilsen, Roc&#0237;o Pozo, O.
-  Sarobidy Rakotonarivo, Bram Van Moorter, and Nils Bunnefeld
+author: A. Bradley Duthie, Jeremy J. Cusack, Isabel L. Jones, Erlend B. Nilsen, Roc&#0237;o
+  Pozo, O. Sarobidy Rakotonarivo, Bram Van Moorter, and Nils Bunnefeld
 date: '`r Sys.Date()`'
@@
 |  land_dim_2     |  100    | Height of the landscape (vertical cells)                                                         |
-|  res_movement   |  20     | Distance (cells) a resource can move in any direct (for movement rules, see res_move_type)       |
+|  res_movement   |  20     | Distance (cells) a resource can move in any direction (for movement rules, see res_move_type)    |
 |  remove_pr      |  0      | Density-independent probability of resource mortality during a time step                         |
@@
 
-Here we illustrate the usefulness of GMSE by considering the case study of a protected population of waterfowl that exploits agricultural land causing a conservation conflict with farmers [e.g., @Tuvendal2015; @Fox2016a; @Fox2017a].
+Here we illustrate the usefulness of GMSE by considering the case study of a protected population of waterfowl that exploits agricultural land causing a conservation conflict with farmers [e.g., @Fox2017a; @Mason2017; @Tulloch2017].
 Managers attempt to keep the abundance of waterfowl at a target level, while farmers attempt to minimise the damage inflicted on their crops [e.g., @Madsen2017].
@@
 Figure 2 shows the landscape broken down by resource position and farmer land ownership in the upper left and right hand panels, respectively.
-The waterfowl population fluctuates around the manager's target size of 1000, but the manager's estimate of population size deviates from its actual size due to observation uncertainty (compare black and blue lines the middle left panel).
+The waterfowl population fluctuates around the manager's target size of 1000, but the manager's estimate of population size deviates from its actual size due to observation uncertainty (compare black and blue lines in the middle left panel).
 Because the waterfowl have a direct negative effect on landscape yield, total landscape yield (orange line of the middle left panel), along with the yield of individual farmers (right middle panel), is low when waterfowl abundance is high, and vice versa.

### 2017-11-14 notebook/ms/SI1.Rmd
Add suggested clarification of the 'resource' term

@@
 
-To predict change resource abundance as a consequence of policy, an individual first needs to know the total number of actions of all types $j$ (e.g., scaring, culling, etc.) performed by users in the previous time step ($X_{\bullet, j}$), and the cost of performing each action ($C_{\bullet, j}$).
+To predict change in resource abundance as a consequence of policy, an individual first needs to know the total number of actions of all types $j$ (e.g., scaring, culling, etc.) performed by users in the previous time step ($X_{\bullet, j}$), and the cost of performing each action ($C_{\bullet, j}$).
 This information is collected from `ACTION` and `COST` arrays.

### 2017-11-14 notebook/ms/ms.Rmd
Add suggested clarification of the 'resource' term

@@
 The workhorse function `gmse` runs simulations using four predefined submodels, which can be parameterised to fit various case studies; more tailored submodels can also be defined using the `gmse_apply` function (see below).
-(1) A population of discrete resources with individual traits (e.g., location, age) is modelled on a spatially-explicit landscape and can simulate resource birth, movement, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty.
+(1) A population of discrete resources (e.g., a population of conservation interest) with individual traits (e.g., location, age) is modelled on a spatially-explicit landscape and can simulate resource birth, movement, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty.
 This sub-model is unique in not relying on other sub-models because ecological dynamics can be simulated in the absence of observation and management.
@@
 Waterfowl consume and destroy all crop yield upon arrival to a landscape cell.
-In each time step, waterfowl are observed on subset of cells, then managers extrapolate from density per cell to estimate total population size.
+In each time step, waterfowl are observed on a subset of cells, then managers extrapolate from density per cell to estimate total population size.
 Managers then use these estimates to set costs of culling and scaring (non-lethal) waterfowl for five farmers.

### 2017-11-14 notebook/ms/ms.Rmd
New edits

@@
 The workhorse function `gmse` runs simulations using four predefined submodels, which can be parameterised to fit various case studies; more tailored submodels can also be defined using the `gmse_apply` function (see below).
-(1) A population of discrete resources (e.g., a population of conservation interest) with individual traits (e.g., location, age) is modelled on a spatially-explicit landscape and can simulate resource birth, movement, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty.
+(1) A population of discrete resources (e.g., a managed species) with individual traits (e.g., location, age) is modelled on a spatially-explicit landscape and can simulate resource birth, movement, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty.
 This sub-model is unique in not relying on other sub-models because ecological dynamics can be simulated in the absence of observation and management.

### 2017-11-14 notebook/ms/SI1.Rmd
Add linked references and colours

@@
 biblio-style: apalike
+link-citations: true
+linkcolor: blue
 ---

### 2017-11-14 notebook/ms/SI2.Rmd
Add linked references and colours

@@
   html_document: default
+  word_document:
+    fig_caption: yes
+    reference_docx: docx_template.docx
+link-citations: yes
+linkcolor: blue
 linestretch: 1

### 2017-11-16 notebook/ms/SI1.Rmd
Fix Pozo name
Add Saro's suggestion to redefine 'individual' in manager fitness section

@@
 author: A. Bradley Duthie, Jeremy J. Cusack, Isabel L. Jones, Erlend B. Nilsen, Roc&#0237;o
-  Pozo, O. Sarobidy Rakotonarivo, Bram Van Moorter, and Nils Bunnefeld
+  A. Pozo, O. Sarobidy Rakotonarivo, Bram Van Moorter, and Nils Bunnefeld
 date: '`r Sys.Date()`'
@@
 
-Individual fitness as calculated for managers ($F_{i}^{m}$) is affected by a manager's utility for resources and the projected change in resource abundance caused by the individual's policy (i.e., the contents of their `POPULATION` layer, specifically row 3).
+Individual fitness as calculated for managers ($F_{i}^{m}$) is affected by a manager's utility for resources and the projected change in resource abundance caused by the individual's policy (i.e., the contents of their `POPULATION` layer, specifically row 3; here again we use 'individual' to refer to one of `ga_popsize` discrete strategies in `POPULATION`, which may be selected and reproduce within the genetic algorithm).
 Manager utility for a resource ($U^{m}_{res}$) is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model (see "[Key data structures used](#Key_data_structures_used)" above).
 Manager utility can therefore change in each GMSE time step as estimated resource abundance changes; when the estimated resource abundance is greater than `manage_target`, $U^{m}_{res}$ is negative, and when the estimated resource abundance is less than `manage_target`, $U^{m}_{res}$ is positive.
-To get individual fitness, first the change in resource abundance predicted by the individual's policy ($\Delta A_{i}$) is calculated, then the squared difference between $\Delta A_{i}$ and $U^{m}_{res}$ is calculated to obtain a utility deviation ($D_{i}$) for the individual $i$,
+To get the fitness of individuals, first the change in resource abundance predicted by the individual's policy ($\Delta A_{i}$) is calculated, then the squared difference between $\Delta A_{i}$ and $U^{m}_{res}$ is calculated to obtain a utility deviation ($D_{i}$) for the individual $i$,
 

### 2017-11-17 notebook/ms/SI1.Rmd
Add suggestion from @ninbrm to emphasise suggested parameter value settings

@@
 linestretch: 1
+link-citations: yes
+linkcolor: blue
 bibliography: gmse.bib
@@
 biblio-style: apalike
-link-citations: true
-linkcolor: blue
 ---
@@
 In GMSE, tournament selection proceeds by randomly sampling `ga_sampleK` individuals from the total `POPULATION` with replacement.
-The fitnesses of the subset of `ga_sampleK` individuals are compared, and the `ga_chooseK` individuals of highest fitness are retained (if `ga_sampleK` $\geq$ `ga_chooseK`, then all `ga_sampleK` are chosen, but this is not recommended).
+The fitnesses of the subset of `ga_sampleK` individuals are compared, and the `ga_chooseK` individuals of highest fitness are retained (if `ga_sampleK` $\geq$ `ga_chooseK`, then all `ga_sampleK` are chosen, but this will prevent adaptive evolution and is therefore not recommended).
 Tournaments selecting `ga_chooseK` individuals from random subsets of size `ga_sampleK` continue until a total of `ga_popsize` individuals are retained.

### 2017-11-17 notebook/ms/SI1.Rmd
Edits for biorXiv

@@
 title: 'GMSE: an R package for generalised management strategy evaluation'
-author: A. Bradley Duthie, Jeremy J. Cusack, Isabel L. Jones, Erlend B. Nilsen, Roc&#0237;o
-  A. Pozo, O. Sarobidy Rakotonarivo, Bram Van Moorter, and Nils Bunnefeld
-date: '`r Sys.Date()`'
+subtitle: Supporting Information 1
+author: A.
+Bradley Duthie&#xb9;&#xb3;, Jeremy J.
+Cusack&#xb9;, Isabel L.
+Jones&#xb9;, Erlend B.
+Nilsen&#xb2;, Roc&#0237;o A.
+Pozo&#xb9;, O.
+Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+date: "[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK
+[2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk"
 output:
@@
 header-includes:
-- \usepackage{lineno}
-- \linenumbers
 linestretch: 1
-link-citations: yes
+bibliography: gmse.bib
+biblio-style: apalike
+link-citations: true
 linkcolor: blue
-bibliography: gmse.bib
-subtitle: Supporting Information 1
-biblio-style: apalike
 ---

### 2017-11-17 notebook/ms/SI2.Rmd
Fix Pozo name
Edits for biorXiv

--- 2017-11-10 notebook/ms/SI2.Rmd
@@
 title: 'GMSE: an R package for generalised management strategy evaluation'
-author: A. Bradley Duthie, Jeremy J. Cusack, Isabel L. Jones, Erlend B. Nilsen, Roc&#0237;o
-  Pozo, O. Sarobidy Rakotonarivo, Bram Van Moorter, and Nils Bunnefeld
-date: '`r Sys.Date()`'
+subtitle: Supporting Information 2
+author: A.
+Bradley Duthie&#xb9;&#xb3;, Jeremy J.
+Cusack&#xb9;, Isabel L.
+Jones&#xb9;, Erlend B.
+Nilsen&#xb2;, Roc&#0237;o A.
+Pozo&#xb9;, O.
+Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+date: "[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK
+[2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk"
 output:
-  word_document:
-    fig_caption: yes
-    reference_docx: docx_template.docx
   pdf_document:
@@
   html_document: default
+  word_document:
+    fig_caption: yes
+    reference_docx: docx_template.docx
+header-includes:
 linestretch: 1
-subtitle: Supporting Information 2
-header-includes:
-- \usepackage{lineno}
-- \linenumbers
 ---

### 2017-11-17 notebook/ms/ms.Rmd
Fix Pozo name
Edits for biorXiv

--- 2017-11-14 notebook/ms/ms.Rmd
@@
 title: 'GMSE: an R package for generalised management strategy evaluation'
-author: A. Bradley Duthie, Jeremy J. Cusack, Isabel L. Jones, Erlend B. Nilsen, Roc&#0237;o
-  Pozo, O. Sarobidy Rakotonarivo, Bram Van Moorter, and Nils Bunnefeld
-date: '`r Sys.Date()`'
+author: A.
+Bradley Duthie&#xb9;&#xb3;, Jeremy J.
+Cusack&#xb9;, Isabel L.
+Jones&#xb9;, Erlend B.
+Nilsen&#xb2;, Roc&#0237;o A.
+Pozo&#xb9;, O.
+Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+date: "[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK
+[2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk"
 output:
@@
 header-includes:
-- \usepackage{lineno}
-- \linenumbers
 linestretch: 1

### 2017-11-22 notebook/ms/ms.Rmd
Minor changes specifically for MEE

@@
 title: 'GMSE: an R package for generalised management strategy evaluation'
-author: A.
-Bradley Duthie&#xb9;&#xb3;, Jeremy J.
-Cusack&#xb9;, Isabel L.
-Jones&#xb9;, Erlend B.
-Nilsen&#xb2;, Roc&#0237;o A.
-Pozo&#xb9;, O.
-Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
-date: "[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK
-[2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk"
+author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
+  Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;,
+  Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
+  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk,
+  Biological and Environmental Sciences 3A149 University of Stirling Stirling, FK9
+  4LA, UK'
 output:
+  word_document:
+    fig_caption: yes
+    reference_docx: docx_template.docx
   pdf_document:
@@
   html_document: default
-  word_document:
-    fig_caption: yes
-    reference_docx: docx_template.docx
 header-includes:
-linestretch: 1
+- \usepackage{fancyhdr}
+- \pagestyle{fancy}
+- \fancyhead[R]{GMSE, AN R PACKAGE}
+- \usepackage{lineno}
+- \linenumbers
+linestretch: 2
+link-citations: yes
+linkcolor: blue
 bibliography: gmse.bib
 biblio-style: apalike
-link-citations: true
-linkcolor: blue
 ---
@@
 
+\newpage
+
 Abstract
@@
 5. The R package GMSE is open source under GNU Public License; source code and documents are freely available on GitHub.
+
+**Key words:** *adaptive management, conservation conflict, ecological modelling, game theory, management strategy evaluation*
 
@@
 
-GMSE builds off of the MSE framework (Figure 1).
+GMSE builds off of the MSE framework (Fig.
+1).
 The workhorse function `gmse` runs simulations using four predefined submodels, which can be parameterised to fit various case studies; more tailored submodels can also be defined using the `gmse_apply` function (see below).
@@
 
-Figure 2 shows the landscape broken down by resource position and farmer land ownership in the upper left and right hand panels, respectively.
+Fig.
+2 shows the landscape broken down by resource position and farmer land ownership in the upper left and right hand panels, respectively.
 The waterfowl population fluctuates around the manager's target size of 1000, but the manager's estimate of population size deviates from its actual size due to observation uncertainty (compare black and blue lines in the middle left panel).
@@
 Farmers responding to policy are interested only in minimising waterfowl's exploitation of their crops, so they will either cull or scare to remove the waterfowl from their land, depending on which option is more effective (i.e., cheaper).
-This is reflected in the bottom left versus right panels of Figure 2; when managers decrease culling costs relative to scaring, farmers respond with more total culling, and vice versa.
+This is reflected in the bottom left versus right panels of Fig.
+2; when managers decrease culling costs relative to scaring, farmers respond with more total culling, and vice versa.
 Farmer decisions then affect waterfowl distribution and abundance, impacting future crop yield and policy.

### 2018-04-10 notebook/ms/SI3.Rmd
Initialise a SI on conservation conflict

---
title: 'GMSE: an R package for generalised management strategy evaluation'
subtitle: Supporting Information 3
author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: "[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK
[2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk"
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes:
linestretch: 1
bibliography: gmse.bib
biblio-style: apalike
link-citations: true
linkcolor: blue
---

```{r, echo = FALSE}
library(GMSE);
```

An example of management conflict using GMSE
================================================================================

Agents in GMSE (managers and users) are goal-oriented, and their behaviour is therefore driven to maximise a particular utility of interest such as a target density of resources (managers), or a suitable resource or landscape harvest size (users). This model feature allows GMSE to evaluate the actions of agents in the context of their individual objectives, and to therefore quantify the degree to which those objectives are or are not achieved. When the actions of one party clashes with the objectives of another party, the objectives of one might be expressed at the expense of the other, causing conservation conflict [@Redpath2013]. Currently, there is no standard way to measure conservation conflict in a social-ecological system, and previous modelling approaches have not meaningfully separated agent objectives from agent actions. We suggest that a useful starting point to developing such a metric is reflected in the deviation of an individual's actions from their objectives, the former of which is restricted by the actions of other individuals. Here we show how GMSE can be used to evaluate the amount of conflict in a simulated social-ecological system under different management options.

### 2018-04-12 notebook/ms/SI3.Rmd
Initialise new SI

@@
 Currently, there is no standard way to measure conservation conflict in a social-ecological system, and previous modelling approaches have not meaningfully separated agent objectives from agent actions.
-We suggest that a useful starting point to developing such a metric is reflected in the deviation of an individual's actions from their objectives, the former of which is restricted by the actions of other individuals.
+We suggest that a useful starting point to developing such a metric is to quantify the deviation of an individual's actions from their objectives, the former of which is restricted by the actions of other individuals.
 Here we show how GMSE can be used to evaluate the amount of conflict in a simulated social-ecological system under different management options.
+
+To demonstrate how GMSE can be used to understand conflict in social-ecological systems, we build upon the [example of resource management](https://www.biorxiv.org/content/biorxiv/early/2017/11/17/221432.full.pdf#page=5) in the main text.
+We consider a protected population of waterfowl that exploits agricultural land and is therefore a source of conservation conflict [e.g., @Fox2017a; @Mason2017; @Tulloch2017].
+As in the main text example, the objective of managers is to keep waterfowl at a target abundance, while the objective of farmers is to maximise agricultural production on their landscape.
+Here we consider a more complex simulation to reflect a scenario likely to occur in a real social-ecological system.
+In contrast to the main text, we consider a population of waterfowl that is abundant enough to substantially decrease yield.
+We consider a community of 100 farmers, each of which has a farm of 50 landscape cells, which could be interpreted as hectares of arable cropland.
+We additionally include landscape cells that are not owned by farmers using the `public_land` argument; these cells can be interpreted as public land on which waterfowl might forage, but which do not affect farmers' agricultural production.
+We further assume an intrinsic growth rate of 0.5, and other parameters as provided by literature for geese (fix this).
+
+```{r, echo = TRUE}
+sim <- gmse(land_ownership = TRUE, stakeholders = 100, observe_type = 0, 
+            res_death_K = 20000, manage_target = 10000, RESOURCE_ini = 8000, 
+            user_budget = 1000, manager_budget = 8000, res_consume = 1, 
+            scaring = FALSE, plotting = FALSE, public_land = 0.5);
+```
+
+The results are plotted below.
+
+```{r, eval = FALSE}
+plot_gmse_results(res = sim$resource, obs = sim$observation, land = sim$land, 
+                  agents = sim$agents, paras = sim$paras, ACTION = sim$action, 
+                  COST = sim$cost);
+```

### 2018-04-12 notebook/ms/ms.Rmd
Minor edits based on reviewer comments

@@
 Many global natural resources, including the biodiversity on which critical ecosystem services depend, are in a state of severe decline [@Dirzo2014; @Hautier2015; @Ceballos2017a; @OConnell2017].
-Conservation of biodiversity can be complicated by the immediate need to use natural resources and land area for human livelihood (e.g., food production), causing real or perceived conflicts between biodiversity conservation and food security.
+Conservation of biodiversity can be complicated by the immediate need to use natural resources and land area for human livelihood, causing real or perceived conflicts between biodiversity conservation and agricultural production.
 This creates a challenging situation for the management of many natural resources [@Redpath2015a].
@@
 To effectively manage natural resources, an adaptive approach allows managers to iteratively update their models and respond flexibly to changing conditions [@Keith2011].
-This approach is especially effective when considering all aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of stakeholders [@Bunnefeld2011; @Bunnefeld2014].
+This approach is especially effective when considering multiple aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of stakeholders [@Bunnefeld2011; @Bunnefeld2014].
 Management strategy evaluation (MSE) is a modelling framework, first developed in fisheries, for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016].
@@
 text(x=43, y=42, labels="Multiple", cex=0.7, pos = 4);
-text(x=43, y=39, labels="Generations", cex=0.7, pos = 4);
+text(x=43, y=39, labels="Iterations", cex=0.7, pos = 4);
 # Arrows indicating direction of GMSE model
@@
 
-In each new call of the genetic algorithm, a unique population of managers or users with random strategies is temporarily initialised.
-In each generation of the genetic algorithm, these strategies crossover and mutate; when this results in strategies that are over-budget, expenditures are iteratively decreased at random until budget constraints are satisfied.
-A fitness function then evaluates each strategy in the population, and a tournament is used to select the next generation of strategies [@Hamblin2013].
-The genetic algorithm terminates when a minimum number of generations has passed and the increase in the fitness of the fittest strategy between the current and previous generation is below some threshold.
+Overall, a single call of the genetic algorithm simulates the process of thinking and decision-making for one manager or user.
+In each new call of the genetic algorithm, a unique population of temporary manager or user strategies is initialised.
+In each iteration of the genetic algorithm, these strategies crossover and mutate; when this results in strategies that are over-budget, expenditures are iteratively decreased at random until budget constraints are satisfied.
+A fitness function then evaluates each strategy in the population, and a tournament is used to select the next iteration of strategies [@Hamblin2013].
+The genetic algorithm terminates when a minimum number of iterations has passed and the increase in the fitness of the fittest strategy between the current and previous iteration is below some threshold.
 The highest fitness strategy in the population then becomes the stakeholder's new strategy.
-Overall, a single call of the genetic algorithm thereby simulates the process of thinking and decision-making for one manager or user.
 
@@
 ```{r, echo = FALSE, fig.height = 6, fig.align = 'left', fig.cap = "Results of an example simulation illustrating the management of a protected resource that exploits the land of five farmers.
-The upper left panel shows locations of resources (black dots) on the landscape in the final time step of the simulation.
+The upper left panel shows locations of resources (black dots) on the landscape in the final time step of the simulation (multiple resources can occur on the same landscape cell).
 The upper right panel shows the same landscape broken down into five differently coloured regions, which correspond to areas of land owned by each of the five farmers.
 The middle left panel shows the actual abundance of resources (black solid line), and the abundance of resources as estimated by the manager (blue solid line; shading indicates 95 percent confidence intervals), over time.
-The horizontal dotted red and blue lines show the resource carrying capacity enacted on adult mortality and the manager's target for resource abundance, respectively.
+The horizontal dotted red and blue lines show the landscape-level resource carrying capacity enacted on adult mortality and the manager's target for resource abundance, respectively.
 The orange line shows the total percent yield of landscape cells.

### 2018-04-12 R/gmse.R
Spelling correction

@@ -35,8 +35,8 @@
 #'@param ga_popsize The size of populations of agents in the genetic algorithm (not resources in the simulation). The actions of each agent in the simulation are duplicated `ga_popsize` times, and this population of individual agent actions undergoes a process of natural selection to find an adaptive strategy. Selection is naturally stronger in larger populations, but a default population size of 100 is more than sufficient to find adaptive strategies. 
 #'@param ga_mingen The minimum number of generations in the genetic algorithms of the simulation (*not* the number of time steps in the simulation itself). The actions of each agent in the simulation are duplicated `ga_popsize` times, and this population of individual agent actions undergoes a process of natural selection at least `ga_mingen` times to find an adaptive strategy. If convergence criteria `converge_crit` is set to a default value of 100, then the genetic algorithm will almost always continue for exactly `ga_mingen` generations. The default value is 20, which is usually plenty for finding adaptive agent strategies -- the objective is not to find optimal strategies, but strategies that are strongly in line with agent interests.
 #'@param ga_seedrep At the start of each genetic algorithm, `ga_popsize` replicate agents are produced; `ga_seedrep` of these replicates are *exact* replicates, while the rest have random actions to introduce variation into the population. Because adaptive agent strategies are not likely to change wildly from one generation to the next, it is highly recommended to use some value of `ga_seedrep` greater than zero; the default value is 20, which does a good job of finding adaptive strategies.
-#'@param ga_sampleK In the genetic algorithm, fitnesses are assigned to different agent strategies and compete in a tournament to be selected into the next generation. The tournament samples `ga_sampleK` strategies at random and with replacement from the population of `ga_popsize` to be included in the torunament. The default value is 20.
-#'@param ga_chooseK In the genetic algorithm, fitnesses are assigned to different agent strategies and compete in a tournament to be selected into the next generation. The tournament samples `ga_sampleK` strategies at random and with replacement from the population of `ga_popsize` to be included in the torunament, and from these randomly selected strategies, the top `ga_chooseK` strategies are selected. The default value is 2, so the top 10 percent of the random sample in a tournament makes it into the next generation (note that multiple tournaments are run until `ga_popsize` strategies are selected for the next generation). 
[…]

### 2018-04-12 notebook/ms/SI1.Rmd
Minor tweaks

@@
 Therefore, one run of the genetic algorithm occurs to simulate the manager's policy-setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users copy).
-Each run of the genetic algorithm mimics the evolution by natural selection of a population of potential manager or user strategies over multiple generations, with the highest fitness strategy in the terminal generation being selected as the one that the manager or user decides to implement.
-For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'generation' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the main text).
+Each run of the genetic algorithm mimics the evolution by natural selection of a population of potential manager or user strategies over multiple iterations, with the highest fitness strategy in the terminal iteration being selected as the one that the manager or user decides to implement.
+For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'iteration' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the main text).
 Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.3.1.9 (future versions of GMSE might expand upon this framework, and we highlight some of these potential avenues for expansion).
@@
 Columns 8-13 are also set to 10001, except where actions are allowed.
+Maximum values of 10000 are unrelated to landscape cells (the total number of which is also 10000 by default).
 Below shows the `COST` array that corresponds to the above `ACTION` array.
@@
 The genetic algorithm proceeds by first initialising a large (but temporary) population of new `ACTION` layers.
-In each generation, these layers crossover and mutate, generating variation in potential agent decisions; costs constrain this variation from exceeding a maximum budget, then the fitness of each layer is evaluated based on how the layer is predicted to affect resources or landscape output to which the agent has assigned some utility.
-A tournament is used to select high fitness layers, and these selected layers become the new generation of layers; generations continue until a minimum number of generations (`ga_mingen`) have passed and a convergence criteria is satisfied such that the increase in mean fitness from the previous generation is below the threshold `converge_crit` (Figure S1-1).
+In each iteration, these layers crossover and mutate, generating variation in potential agent decisions; costs constrain this variation from exceeding a maximum budget, then the fitness of each layer is evaluated based on how the layer is predicted to affect resources or landscape output to which the agent has assigned some utility.
+A tournament is used to select high fitness layers, and these selected layers become the new iteration of layers; iterations continue until a minimum number of iterations (`ga_mingen`) have passed and a convergence criteria is satisfied such that the increase in mean fitness from the previous iteration is below the threshold `converge_crit` (Figure S1-1).
 
@@
 The number of exact replicates initialised is set using `ga_seedrep` (if `ga_seedrep` $\geq$ `ga_popsize`, then all individuals are seeded as replicates).
-After the `POPULATION` of `ga_popsize` individuals is initialised, a loop simulating the adaptive evolution of `POPULATION` in non-overlapping generations begins (see Figure S1-1 above).
+After the `POPULATION` of `ga_popsize` individuals is initialised, a loop simulating the adaptive evolution of `POPULATION` in non-overlapping iterations begins (see Figure S1-1 above).
 
@@
 
-A single generation of the genetic algorithm begins with a uniform crossover [@Hamblin2013], by which actions of individuals in `POPULATION` are randomly swapped with some probability.
+A single iteration of the genetic algorithm begins with a uniform crossover [@Hamblin2013], by which actions of individuals in `POPULATION` are randomly swapped with some probability.
 To implement crossover, each individual selects a partner, then exchanges corresponding array elements affecting agent actions (columns 8-13) with their partner at a fixed probability of `ga_crossover`.
@@
 Once a new set of `ga_popsize` individuals is retained through tournament selection, these individuals replace the previous `POPULATION` array.
-The genetic algorithm terminates if and only if a minimum number of generations has passed (`ga_mingen`) and a convergence criteria (`converge_crit`) is satisfied.
-The convergence criteria checks the difference between the mean fitness of individuals in the new generation versus the previous generation; if this difference is greater than `converge_crit`, then termination does not occur (this prevents termination from occuring while fitness is still increasing, though it is usually fine to use the default GMSE `converge_crit = 100` and `ga_mingen = 40`, which nearly always terminates the genetic algorithm after 40 generations having identified adaptive manager or user strategies).
+The genetic algorithm terminates if and only if a minimum number of iterations has passed (`ga_mingen`) and a convergence criteria (`converge_crit`) is satisfied.
+The convergence criteria checks the difference between the mean fitness of individuals in the new iteration versus the previous iteration; if this difference is greater than `converge_crit`, then termination does not occur (this prevents termination from occuring while fitness is still increasing, though it is usually fine to use the default GMSE `converge_crit = 100` and `ga_mingen = 40`, which nearly always terminates the genetic algorithm after 40 iterations having identified adaptive manager or user strategies).
 Due to the way in which fitness is calculated (see below), in practice, `converge_crit` currently applies only to users.
-If termination conditions are not satisfied, then the `POPULATION` of individuals begins a new generation of crossover, mutation, cost constraint, fitness evaluation, and tournament selection (Figure S1-1).
+If termination conditions are not satisfied, then the `POPULATION` of individuals begins a new iteration of crossover, mutation, cost constraint, fitness evaluation, and tournament selection (Figure S1-1).
 
@@
 
+Choosing genetic algorithm parameter values
+--------------------------------------------------------------------------------
+
+Options for adjusting
+
+| GMSE argument       | Default | Description |
+|---------------------|---------|-------------|
+| `ga_popsize`        | 100     | The number of individuals in the population temporary simulated during a single run of the genetic algorithm. |
+| `ga_mingen`         | 40      | The minimum number of iterations that a genetic algorithm will run before settling on an agent's strategy. |
+| `ga_seedrep`        | 20      | The number of individuas in the population to be initiaised with the current agent's strategy (e.g., from a previous time step in the broader GMSE simulation), as opposed to being initialised with random strategies.
+|
+| `ga_sampleK`        | 20      | For the tournament step of the genetic agorithm, how many strategies are selected at random from the larger population (with replacement) to be included a the tournament.
+|
+| `ga_chooseK`        | 2       | Four the tournament step of the genetic agorithm, how many strategies are selected as winners of the tournament, to be included in the next iteration. |
+| `ga_mutation`       | 0.1     | The mutation rate of any action in an agent's strategy |
+| `ga_crossover`      | 0.1     | The crossover rate of any action in an agent's strategy; crossover events occur with a different randomly selected strategy in the population. |
+| `ga_converge_crit`  | 1       | The percent increase in strategy fitness from one iteration to the next below which the convergence criteria is satisfied.
+Iterations wil continue as long as fitness increase is above this convergence criteria.
+|
+| `group_think`       | FALSE   | Whether or not all users (i.e., not including the manager) have identical strategies. If TRUE, then one genetic algorithm will be run and applied to a users >
+
 Future development of fitness functions

### 2018-04-13 R/plotting.R
Axes edits

@@ -600,9 +600,8 @@ plot_gmse_results <- function(res, obs, land, agents, paras, ACTION, COST){
         }
     }
     
-    minK <- min(para_vec[6:7]);
+    ymaxi <- max(paras[,33]) + (0.1 * (max(paras[,33])));
     
-    ymaxi    <- minK + (minK * (1 + res[[1]][1,10])); # Add for birth rate
     time_max <- length(res);
     for(i in 1:(time_max-1)){
         res_t    <- res[[i]];
@@ -900,9 +899,13 @@ plot_gmse_effort <- function(agents, paras, ACTION, COST){
                    lwd = 1, col = cols[1]);
         }
         axis(side = 4, at = c(50, 100),  labels = c(50, 100), cex.axis = 1.5);
+        lcex  <- 1.5
+        if(max_time < 100){
+            lcex <- lcex * (max_time / 100);
+        }
         legend(x = 1, y = y2 + y2*0.1, fill = cols[1:5], horiz = TRUE,
                legend = c("scaring", "culling", "castration", "feeding", 
-                          "helping"), bty = "n", cex = 1.5);
+                          "helping"), bty = "n", cex = lcex);
         abline(h = y1 + y1*0.1, lwd = 2);
     }else{
         plot(x = 1:max_time, y = scar_eff[,1], type = "n", xaxt= "n", 

### 2018-04-13 notebook/ms/SI1.Rmd
SI changes

@@
 
-Options for adjusting
+Options for adjusting genetic algorithm parameter values in `gmse` and `gmse_apply` are shown below.
 
 | GMSE argument       | Default | Description                                  |
-|---------------------|---------|-------------|
+|---------------------|---------|----------------------------------------------|
 | `ga_popsize`        | 100     | The number of individuals in the population temporary simulated during a single run of the genetic algorithm. |
 | `ga_mingen`         | 40      | The minimum number of iterations that a genetic algorithm will run before settling on an agent's strategy. |
-| `ga_seedrep`        | 20      | The number of individuas in the population to be initiaised with the current agent's strategy (e.g., from a previous time step in the broader GMSE simulation), as opposed to being initialised with random strategies.
+| `ga_seedrep`        | 0       | The number of individuas in the population to be initiaised with the current agent's strategy (e.g., from a previous time step in the broader GMSE simulation), as opposed to being initialised with random strategies.
 |
@@
 |
-| `group_think`       | FALSE   | Whether or not all users (i.e., not including the manager) have identical strategies. If TRUE, then one genetic algorithm will be run and applied to a users >
+| `group_think`       | FALSE   | Whether or not all users (i.e., not including the manager) have identical strategies. If TRUE, then one genetic algorithm will be run and applied to all users. |
+
+Given the heuristic goals of the genetic algorithm to mimic the goal-oriented behaviour of agents, default parameters are typically sufficient for agent decision making.
+Key parameters can be adjusted if more precision in decision making is desired, but these adjustments will come at a cost of simulation efficiency.
+For example, increasing `ga_popsize` or `ga_mingen`, or decreasing `ga_converge_crit`, might fine tune strategies more effectively, but this will cause the genetic algorithm to take longer every time that it is run, ultimately slowing down GMSE simulations.
+Alternativey, setting `group_think = TRUE` will greatly speed up GMSE simulations when many users are being simulated, but this comes at the cost of among-user variation in decision making.
+Overall, we recommend first using default values in the genetic algorithm before exploring how other parameter value options affect simulation dynamics; for a more general discussion about selecting parameter values in genetic algorithms, see @Hamblin2013.
 

### 2018-04-13 notebook/ms/SI3.Rmd
SI changes

@@
 In contrast to the main text, we consider a population of waterfowl that is abundant enough to substantially decrease yield.
-We consider a community of 100 farmers, each of which has a farm of 50 landscape cells, which could be interpreted as hectares of arable cropland.
-We additionally include landscape cells that are not owned by farmers using the `public_land` argument; these cells can be interpreted as public land on which waterfowl might forage, but which do not affect farmers' agricultural production.
+We consider a community of 20 farmers, each of which has a farm of 50 landscape cells, which could be interpreted as hectares of arable cropland.
+We additionally include landscape cells that are not owned by farmers using the `public_land` argument; these comprise 40 percent of cells and can be interpreted as public land on which waterfowl might forage, but which do not affect farmers' agricultural production.
 We further assume an intrinsic growth rate of 0.5, and other parameters as provided by literature for geese (fix this).
 
+For simplicity, here we assess confict using only the `gmse` function to show how parameter values can be set to provide useful results.
+All of the analysis shown would also be possibe using the more detailed, but also more flexible, `gmse_apply` function.
+
 ```{r, echo = TRUE}
-sim <- gmse(land_ownership = TRUE, stakeholders = 100, observe_type = 0, 
-            res_death_K = 20000, manage_target = 10000, RESOURCE_ini = 8000, 
-            user_budget = 1000, manager_budget = 8000, res_consume = 1, 
-            scaring = FALSE, plotting = FALSE, public_land = 0.5);
+sim <- gmse(manager_budget = 10000, user_budget = 10000, res_death_K = 93870, 
+            manage_target = 70000, RESOURCE_ini = 63000, plotting = FALSE, 
+            stakeholders = 20, land_ownership = TRUE, land_dim_1 = 50, 
+            land_dim_2 = 50, public_land = 0.4, scaring = FALSE, lambda = 0.275, 
+            remove_pr = 0.122, time_max = 20, res_death_type = 3, ga_seedrep = 0, 
+            ga_mingen = 100, res_consume = 0.02);
 ```
@@
 
-```{r, eval = FALSE}
+```{r, eval = TRUE, fig.height = 8}
 plot_gmse_results(res = sim$resource, obs = sim$observation, land = sim$land, 
@@
 ```
+
+Look at the conflict
+
+```{r, eval = TRUE, fig.height = 7, fig.width = 6}
+plot_gmse_effort(sim$agents, sim$paras, ACTION = sim$action, COST = sim$cost);
+```
+
+Now try scaring to see what happens.
+
+```{r, echo = TRUE, fig.height = 8}
+sim <- gmse(manager_budget = 10000, user_budget = 10000, res_death_K = 93870, 
+            manage_target = 70000, RESOURCE_ini = 63000, plotting = FALSE, 
+            stakeholders = 20, land_ownership = TRUE, land_dim_1 = 50, 
+            land_dim_2 = 50, public_land = 0.4, scaring = TRUE, lambda = 0.275, 
+            remove_pr = 0.122, time_max = 20, res_death_type = 3, ga_seedrep = 0, 
+            ga_mingen = 100, res_consume = 0.02);
+```
+
+The results are plotted below.
+
+```{r, eval = TRUE, fig.height = 8}
+plot_gmse_results(res = sim$resource, obs = sim$observation, land = sim$land, 
+                  agents = sim$agents, paras = sim$paras, ACTION = sim$action, 
+                  COST = sim$cost);
+```
+
+Look at the conflict
+
+```{r, eval = TRUE, fig.height = 7, fig.width = 6}
+plot_gmse_effort(sim$agents, sim$paras, ACTION = sim$action, COST = sim$cost);
+```

### 2018-04-13 src/game.c
Major overhaul of how managers predict costs

@@ -417,7 +417,7 @@ void apply_min_costs(double ***population, double *paras, int agentID){
  *    agent_array: The agent array
  * ========================================================================== */
 void sum_array_layers(double ***array, double **out, int get_mean, 
-                      double *paras, double **agent_array){
+                      double *paras, double **agent_array, int layer_start){
     
     int row, col, layer, layer_count, ROWS, COLS, layers;
 
@@ -427,7 +427,7 @@ void sum_array_layers(double ***array, double **out, int get_mean,
     
     layer_count = 0;
     if(get_mean == 1){
-        for(layer = 0; layer < layers; layer++){
+        for(layer = layer_start; layer < layers; layer++){
             if(agent_array[layer][1] > 0){
                 layer_count++;
             }
@@ -436,7 +436,7 @@ void sum_array_layers(double ***array, double **out, int get_mean,
     for(row = 0; row < ROWS; row++){
         for(col = 0; col < COLS; col++){
             out[row][col] = 0;
-            for(layer = 0; layer < layers; layer++){
+            for(layer = layer_start; layer < layers; layer++){
                 if(agent_array[layer][1] > 0){
                     if(get_mean == 1){
                         out[row][col] += array[row][col][layer] / layer_count;
@@ -459,25 +459,28 @@ int new_act(double old_cost, double new_cost, double old_act, double *paras){
     
     int total_acts;
     double users, max_to_spend, acts_per_user, cost_per_user, total_cost;
-    double pr_on_act, budget_for_act;
+    double pr_on_act, budget_for_act, mgr_budget, min_cost;
     
-    users        = (double) paras[54] - 1; /* Minus one for the manager */
-    max_to_spend = (double) paras[97];     /* Maximum per user budget   */
+    users        = paras[54] - 1; /* Minus one for the manager */
+    min_cost     = paras[96];     /* Minimum cost of an action */
+    max_to_spend = paras[97];     /* Maximum per user budget   */
+    mgr_budget   = paras[105];    /* Manager's total budget    */
     
-    total_cost    = old_act * old_cost;  /* Total cost devoted to action */
+    total_cost    = 0.0;
+    if(old_cost < mgr_budget){
+        total_cost    = old_act * old_cost; /* Total cost devoted to action */
+    }
 
[…]

### 2018-04-16 notebook/ms/SI3.Rmd
Background and parameter justification

@@
 Currently, there is no standard way to measure conservation conflict in a social-ecological system, and previous modelling approaches have not meaningfully separated agent objectives from agent actions.
-We suggest that a useful starting point to developing such a metric is to quantify the deviation of an individual's actions from their objectives, the former of which is restricted by the actions of other individuals.
+We suggest that a starting point to developing a useful metric of conservation conflict is to quantify the deviation of an individual's actions from their objectives, the former of which is restricted by the actions of other individuals.
 Here we show how GMSE can be used to evaluate the amount of conflict in a simulated social-ecological system under different management options.
@@
 As in the main text example, the objective of managers is to keep waterfowl at a target abundance, while the objective of farmers is to maximise agricultural production on their landscape.
-Here we consider a more complex simulation to reflect a scenario likely to occur in a real social-ecological system.
-In contrast to the main text, we consider a population of waterfowl that is abundant enough to substantially decrease yield.
-We consider a community of 20 farmers, each of which has a farm of 50 landscape cells, which could be interpreted as hectares of arable cropland.
-We additionally include landscape cells that are not owned by farmers using the `public_land` argument; these comprise 40 percent of cells and can be interpreted as public land on which waterfowl might forage, but which do not affect farmers' agricultural production.
-We further assume an intrinsic growth rate of 0.5, and other parameters as provided by literature for geese (fix this).
+Here we consider a more complex simulation to reflect a scenario that is likely to occur in a real social-ecological system.
+We parameterise our model using demographic information from the Taiga Bean Geese (*Anser fabalis fabalis*), a managed population that is hunted for sport in Fennoscandinavia [@Johnson2018].
+Taiga Bean Geese can also cause agricultural damage, which can potentially lead to conflict between farmers and management objectives [@Johnson2018].
+
+Using demographic parameters in simulations
+--------------------------------------------------------------------------------
+
+@Johnson2018 recently estimated key demographic parameters for the Taiga Bean Geese for the Central Management Unit that includes geese that breed in "Northern most Sweden, Northern Norway, Northern and Central Finland, and adjacent North-wester parts of Russia, wintering mostly in Southern Sweden and South-east Denmark" [@AEWA2016].
+They estimated goose survival under ideal conditions to be ca 0.878; this can be interpreted in our model by setting mortality to `remove_pr = 1 - 0.878`.
+Similarly, @Johnson2018 estimated intrinsic population growth rate and carrying capacity to be 0.15 and 93870, respectively, so we set `lambda = 0.15` and `res_death_K = 93870`.
+The global abundance of Taiga Bean Geese in 2009 was ca 63000 [@Fox2010], and ca 35000 in the Central Management Unit [@AEWA2016], which we can take as a starting abundance for our simulations (`RESOURCE_ini = 35000`).
+And the International Single Species Action Plan has a target population size of ca 70000 in the Central Management Unit [@AEWA2016], which we can use as a management target (`manage_target = 70000`).
 
 For simplicity, here we assess confict using only the `gmse` function to show how parameter values can be set to provide useful results.
-All of the analysis shown would also be possibe using the more detailed, but also more flexible, `gmse_apply` function.
+Our goal is not to provide a detailed case study of the Taiga Bean Geese, but rather to demonstrate how such a case study would be possible using the `gmse` function.
+Additionally, all of the analysis that follows would also be possibe using the more detailed, but also more flexible, `gmse_apply` function.
+Where available, we use estimated demographic parameter values from @Johnson2018 and @AEWA2016.
+Where GMSE parameter values are not available, we use reasonable values or GMSE defaults.
+To make model inferences for real case studies, we strongly recommend simulating across a range of parameter values when empirical estimates are unavailable, as social-ecological dynamics might be sensitive to these unknown values.
+
+The code below calls `gmse` using the empirically derived parameters for Taiga Bean Geese described above.
+We also set `manager_budget = 100000` and `user_budget = 100000`.
+Further, we consider the case in of a region in which farmland makes up 60% of all land, with 40% of land being 'public' (`public_land = 0.4`; which might be intpreted as any land in which stakeholders are not, or cannot be, invested in goose presence), and divide the farmland amongst 80 individual farmers (`stakeholders = 80`; `land_ownership = TRUE`).
+Landscape size is set to default 100 by 100 cells, so each farmer effectively owns 75 cells, which might be interpreted as hectares of land (for instructions on how to more precisely control landscape ownership, see the advanced GMSE options using `gmse_apply`).
+Because we need both density-dependent (`res_death_K = 93870`) and density-independent (`remove_pr = 0.122`) sources of mortality, we set `res_death_type = 3`.
+We assume that a single goose decreases agricultural production on a cell by 1% per time step (`res_consume = 0.01`).
+We further assume that the population is well-monitored, with observers counting goose numbers on each cell of the landscape in every timestep (`observe_type = 3`) with the ability to observe two landscape cells in every direction (`agent_view = 2`).
+All other parameter values are set to GMSE defaults.
+
+Simulating goose management
+--------------------------------------------------------------------------------
+
+Below, we first allow only the culling of geese by users and plot the dynamics of the social-ecological from a single simulation.
+Next, we run the same simulation but allow users to scare geese from their landscape cells; we then use the model to make inferences regarding how scaring as a management option might affect goose population dynamics, agricultural production, and conservation conflict in the system.
 
 ```{r, echo = TRUE}
-sim <- gmse(manager_budget = 10000, user_budget = 10000, res_death_K = 93870, 
-            manage_target = 70000, RESOURCE_ini = 63000, plotting = FALSE, 
-            stakeholders = 20, land_ownership = TRUE, land_dim_1 = 50, 
-            land_dim_2 = 50, public_land = 0.4, scaring = FALSE, lambda = 0.275, 
-            remove_pr = 0.122, time_max = 20, res_death_type = 3, ga_seedrep = 0, 
-            ga_mingen = 100, res_consume = 0.02);
+sim <- gmse(manager_budget = 100000, user_budget = 100000, res_death_K = 93870, 
+            manage_target = 70000, RESOURCE_ini = 35000, plotting = FALSE, 
+            stakeholders = 80, land_ownership = TRUE, public_land = 0.4, 
+            scaring = FALSE, lambda = 0.150, remove_pr = 0.122, time_max = 20, 
+            res_death_type = 3, res_consume = 0.01, res_birth_K = 200000, 
+            observe_type = 3, agent_view = 1);
 ```
@@
 
+References
+--------------------------------------------------------------------------------
+
 ```{r, eval = TRUE, fig.height = 8}

### 2018-04-17 notebook/ms/SI3.Rmd
Work on supporting information

@@
 ```{r, echo = TRUE}
-sim <- gmse(manager_budget = 100000, user_budget = 100000, res_death_K = 93870, 
-            manage_target = 70000, RESOURCE_ini = 35000, plotting = FALSE, 
+sim_1 <- gmse(manager_budget = 10000, user_budget = 10000, plotting = FALSE, 
+              res_death_K = 93870, manage_target = 70000, RESOURCE_ini = 35000, 
               stakeholders = 80, land_ownership = TRUE, public_land = 0.4, 
               scaring = FALSE, lambda = 0.150, remove_pr = 0.122, time_max = 20, 
+              res_death_type = 3, res_consume = 0.01, res_birth_K = 200000, 
+              observe_type = 3, agent_view = 1);
+```
+
+The results of the above simulation are plotted in Figure SI3-1 below.
+
+```{r, eval = TRUE, fig.height = 8}
+plot_gmse_results(res = sim_1$resource, obs = sim_1$observation, 
+                  land = sim_1$land, agents = sim_1$agents, paras = sim_1$paras, 
+                  ACTION = sim_1$action, COST = sim_1$cost);
+```
+
+Look at the conflict
+
+```{r, eval = TRUE, fig.height = 7, fig.width = 6}
+plot_gmse_effort(sim_1$agents, sim_1$paras, ACTION = sim_1$action, 
+                 COST = sim_1$cost);
+```
+
+Now try scaring to see what happens.
+
+```{r, echo = TRUE, fig.height = 8}
+sim_2 <- gmse(manager_budget = 10000, user_budget = 10000, plotting = FALSE, 
+              res_death_K = 93870, manage_target = 70000, RESOURCE_ini = 35000, 
+              stakeholders = 80, land_ownership = TRUE, public_land = 0.4, 
+              scaring = TRUE, lambda = 0.150, remove_pr = 0.122, time_max = 20, 
               res_death_type = 3, res_consume = 0.01, res_birth_K = 200000, 
@@
 ```{r, eval = TRUE, fig.height = 8}
-plot_gmse_results(res = sim$resource, obs = sim$observation, land = sim$land, 
-                  agents = sim$agents, paras = sim$paras, ACTION = sim$action, 
-                  COST = sim$cost);
+plot_gmse_results(res = sim_2$resource, obs = sim_2$observation, 
+                  land = sim_2$land, agents = sim_2$agents, paras = sim_2$paras,
+                  ACTION = sim_2$action, COST = sim_2$cost);
 ```
@@
 ```{r, eval = TRUE, fig.height = 7, fig.width = 6}
-plot_gmse_effort(sim$agents, sim$paras, ACTION = sim$action, COST = sim$cost);
+plot_gmse_effort(sim_2$agents, sim_2$paras, ACTION = sim_2$action, 
+                 COST = sim_2$cost);
 ```
-
-Now try scaring to see what happens.
-
-```{r, echo = TRUE, fig.height = 8}
-sim <- gmse(manager_budget = 10000, user_budget = 10000, res_death_K = 93870, 
-            manage_target = 70000, RESOURCE_ini = 63000, plotting = FALSE, 
-            stakeholders = 20, land_ownership = TRUE, land_dim_1 = 50, 
-            land_dim_2 = 50, public_land = 0.4, scaring = TRUE, lambda = 0.275, 
-            remove_pr = 0.122, time_max = 20, res_death_type = 3, ga_seedrep = 0, 
-            ga_mingen = 100, res_consume = 0.02);
-```
-
-The results are plotted below.
 
@@
 --------------------------------------------------------------------------------
-
-```{r, eval = TRUE, fig.height = 8}
-plot_gmse_results(res = sim$resource, obs = sim$observation, land = sim$land, 
-                  agents = sim$agents, paras = sim$paras, ACTION = sim$action, 
-                  COST = sim$cost);
-```
-
-Look at the conflict
-
-```{r, eval = TRUE, fig.height = 7, fig.width = 6}
-plot_gmse_effort(sim$agents, sim$paras, ACTION = sim$action, COST = sim$cost);
-```

### 2018-04-17 notebook/ms/SI3.Rmd
Edit parameters and include warning

@@
 They estimated goose survival under ideal conditions to be ca 0.878; this can be interpreted in our model by setting mortality to `remove_pr = 1 - 0.878`.
-Similarly, @Johnson2018 estimated intrinsic population growth rate and carrying capacity to be 0.15 and 93870, respectively, so we set `lambda = 0.15` and `res_death_K = 93870`.
+Similarly, @Johnson2018 estimated intrinsic population growth rate and carrying capacity to be 0.15 and 93870, respectively, so we set `lambda = 0.183` (for convenience of demonstration, the upper 90% confidence interval of their estimate) and `res_death_K = 93870`.
 The global abundance of Taiga Bean Geese in 2009 was ca 63000 [@Fox2010], and ca 35000 in the Central Management Unit [@AEWA2016], which we can take as a starting abundance for our simulations (`RESOURCE_ini = 35000`).
@@
 Because we need both density-dependent (`res_death_K = 93870`) and density-independent (`remove_pr = 0.122`) sources of mortality, we set `res_death_type = 3`.
-We assume that a single goose decreases agricultural production on a cell by 1% per time step (`res_consume = 0.01`).
+We assume that a single goose decreases agricultural production on a cell by 2% per time step (`res_consume = 0.02`).
 We further assume that the population is well-monitored, with observers counting goose numbers on each cell of the landscape in every timestep (`observe_type = 3`) with the ability to observe two landscape cells in every direction (`agent_view = 2`).
@@
 
-Below, we first allow only the culling of geese by users and plot the dynamics of the social-ecological from a single simulation.
+Below, we first allow only the culling of geese by users and plot the dynamics of the social-ecological system from a single simulation.
 Next, we run the same simulation but allow users to scare geese from their landscape cells; we then use the model to make inferences regarding how scaring as a management option might affect goose population dynamics, agricultural production, and conservation conflict in the system.
+All simulations are run using the `gmse` function, but novice R users may prefer run all of the simulations below using the browser-based GMSE GUI by calling `gmse_gui()` from the R command line.
+Alternatively, experienced R users may prefer simulate by looping time steps through `gmse_apply`, which allows more flexibility for incorporating custom submodels and dynamically adjusting parameter values.
+We emphasise that the simulations below are intended only to demonstrate one use of GMSE on a species of conservation interest, not to make recommendations for management of Taiga Bean Geese.
 

### 2018-04-17 R/gmse_gui.R
Adjust allowable parameter range

@@ -215,7 +215,7 @@ gmse_gui <- function(){
                                         sliderInput("res_birth_K",
                                                     "Carrying capacity applied to birth",
                                                     min   = 10,
-                                                    max   = 10000,
+                                                    max   = 100000,
                                                     step  = 10,
                                                     value = 10000)
                                ),
@@ -224,7 +224,7 @@ gmse_gui <- function(){
                                         sliderInput("res_death_K",
                                                     "Carrying capacity applied to death",
                                                     min   = 10,
-                                                    max   = 10000,
+                                                    max   = 100000,
                                                     step  = 10,
                                                     value = 2000)
                                ),
@@ -242,7 +242,7 @@ gmse_gui <- function(){
                                         sliderInput("RESOURCE_ini",
                                                     "Initial population size",
                                                     min   = 10,
-                                                    max   = 10000,
+                                                    max   = 100000,
                                                     step  = 10,
                                                     value = 1000)
                                )
@@ -360,7 +360,7 @@ gmse_gui <- function(){
                                         sliderInput("manager_budget",
                                                     "Total manager policy-setting budget",
                                                     min   = 100,
-                                                    max   = 10000,
+                                                    max   = 100000,
                                                     step  = 100,
                                                     value = 1000)
                                ),
@@ -369,7 +369,7 @@ gmse_gui <- function(){
                                         sliderInput("manage_target",
                                                     "Target size of the resource population",
                                                     min   = 10,
-                                                    max   = 10000,
+                                                    max   = 100000,
                                                     step  = 10,
                                                     value = 1000)
                                ),
@@ -486,7 +486,7 @@ gmse_gui <- function(){
                                         sliderInput("user_budget",
                                                     "Total user action budget",
                                                     min   = 100,
-                                                    max   = 10000,
+                                                    max   = 100000,
                                                     step  = 100,
                                                     value = 1000)
                                )

### 2018-04-17 notebook/ms/SI3.Rmd
More text in supporting information

@@
 
-```{r, eval = TRUE, fig.height = 8}
+```{r, eval = TRUE, fig.height = 8, fig.cap = "Results of a GMSE simulation using parameters estimated for Taiga Been Geese Central Management Unit.
+This example includes 80 farmers whose objective is to maximise their agricultural output, and one manager whose objective is to keep geese at a target abundance, over 20 simulated time steps.
+Goose locations at the end of the simulation are shown in the upper left panel, while the upper right panel shows the same landscape broken down among the 80 farmers (upper 60% of the landscape in multiple colours), along with non-agricultural land (lower 40% of the landscape in blue).
+Actual goose abundance is shown in the middle left panel (black solid line), along with its estimate by the manager (blue solid line).
+Horizontal red and blue dotted lines show landscape-level goose carrying capacity enacted on adult mortality and the manager's target for goose abundance, respectively.
+The organge line shows the total percent of landscape cells (including non-farmed cells) yield, as decreased by geese.
+The middle right panel shows this yield for each individual farmer, and for the public land (lower line in blue).
+The lower left panel shows the cost of culling for farmers, as set by the manager, and the lower right panel shows the total number of culls attempted by farmers over time."}
 plot_gmse_results(res = sim_1$resource, obs = sim_1$observation, 
@@
 
-Look at the conflict
+The above Figure SI3-1 shows the dynamics of goose abundance and agricultural yield, along with how managers react to change in abundance and farmers react to manager policy.
+In the case of the simulation above, goose population begins to decline, causing managers to set policy to increase the cost of culling.
+This higher cost of culling restricts farmers from culling on their land, leading to fewer total culling actions.
+In this case, the abundance of geese continues downward, while agricultural yield increases over time.
+We can investigate the conflict between management policy and farmers using the `plot_gmse_effort` function in GMSE.
 
-```{r, eval = TRUE, fig.height = 7, fig.width = 6}
+```{r, eval = TRUE, fig.height = 7, fig.width = 6, fig.cap = "Permissiveness that each manager exhibits for each farmer action (black lines) and the effort that each individual farmer puts into each action over time (coloured lines).
+Each panel row reports a different action (in decreasing order: scaring, culling, castration, feeding, and helping).
+The left axis shows the permissiveness that a manager has for the focal action (black line), which is calculated as 100 minus the percent of the manager's budget that is put into increasing the cost of the focal action.
+For example, if the manager puts all of their effort (total budget) into increasing the cost of culling, then permissiveness of culling is 0; if the manager puts no effort into increasing culling cost, then permissiveness of culling is 100.
+The right axis shows effort that farmers put into an action (coloured lines), which defined as the percentage of a farmer's budget put into a particular action (note, values might not add up to 100 because farmers are not forced to use their entire budget)."}
 plot_gmse_effort(sim_1$agents, sim_1$paras, ACTION = sim_1$action, 
@@
 
-Now try scaring to see what happens.
+Black lines indicate how permissive a manager is toward a particular action on a scale of 0 to 100, while coloured lines indicate how much effort farmers expend on a given action.
+When black lines are far below coloured lines, we can (cautiously) interpret this as a conflict between management of the goose population and farmer's interest in agricultural production.
+These time periods represent instances in which managers are not at very permissive of a particular action (in this case culling), but farmers continue to expend effort to do the action anyway.
 

### 2018-04-18 notebook/ms/SI3.Rmd
More edits to SI text

@@
 title: 'GMSE: an R package for generalised management strategy evaluation'
-subtitle: Supporting Information 3
-author: A.
-Bradley Duthie&#xb9;&#xb3;, Jeremy J.
-Cusack&#xb9;, Isabel L.
-Jones&#xb9;, Erlend B.
-Nilsen&#xb2;, Roc&#0237;o A.
-Pozo&#xb9;, O.
-Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
-date: "[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK
-[2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk"
+author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
+  Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;,
+  Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
+  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
 output:
@@
     reference_docx: docx_template.docx
-header-includes:
+header-includes: null
 linestretch: 1
+link-citations: yes
+linkcolor: blue
 bibliography: gmse.bib
+subtitle: Supporting Information 3
 biblio-style: apalike
-link-citations: true
-linkcolor: blue
 ---
@@
 library(GMSE);
+plot_gmse_effort <- function(agents, paras, ACTION, COST){
+
+    cols      <- c("green", "indianred1", "indianred3", "deepskyblue1",
+                   "deepskyblue2");
+
+    users    <- dim(agents[[1]])[1];
+    max_time <- length(ACTION);
+
+    para_vec <- paras[1,];
+
+    allowed  <- sum(para_vec[89:93]);
+
+    scar_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
+    cull_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
+    cast_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
+    feed_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
+    help_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
+
+    scar_act  <- matrix(data = 0, nrow = max_time, ncol = users);
+    cull_act  <- matrix(data = 0, nrow = max_time, ncol = users);
+    cast_act  <- matrix(data = 0, nrow = max_time, ncol = users);
+    feed_act  <- matrix(data = 0, nrow = max_time, ncol = users);
+    help_act  <- matrix(data = 0, nrow = max_time, ncol = users);
+
+    scar_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
+    cull_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
+    cast_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
+    feed_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
+    help_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
+
+    act_costs  <- matrix(data = 0, nrow = max_time, ncol = 5);
+    pol_effort <- matrix(data = 0, nrow = max_time, ncol = 5);
+    min_cost   <- para_vec[97];
+
+    for(time in 1:max_time){
+        #-- Scaring cost and actions
+        scar_cst[time, 1] <- COST[[time]][3, 8, 1];
+        for(user in 2:users){
+            scar_cst[time, user] <- COST[[time]][1, 8, user];
+        }
+        scar_act[time, 1] <- ACTION[[time]][3, 8, 1] - min_cost;
+        for(user in 2:users){
+            scar_act[time, user] <- ACTION[[time]][1, 8, user];
+        }
+        scar_eff[time,] <- scar_act[time,]*scar_cst[time,]/agents[[time]][,17];
+        #-- Culling cost and actions
+        cull_cst[time, 1] <- COST[[time]][3, 9, 1];
+        for(user in 2:users){
+            cull_cst[time, user] <- COST[[time]][1, 9, user];
+        }
+        cull_act[time, 1] <- ACTION[[time]][3, 9, 1] - min_cost;
+        for(user in 2:users){
+            cull_act[time, user] <- ACTION[[time]][1, 9, user];
+        }
+        cull_eff[time,] <- cull_act[time,]*cull_cst[time,]/agents[[time]][,17];
+        #-- Castration cost and actions
+        cast_cst[time, 1] <- COST[[time]][3, 10, 1];
+        for(user in 2:users){
+            cast_cst[time, user] <- COST[[time]][1, 10, user];
+        }
+        cast_act[time, 1] <- ACTION[[time]][3, 10, 1] - min_cost;
+        for(user in 2:users){
+            cast_act[time, user] <- ACTION[[time]][1, 10, user];
+        }
+        cast_eff[time,] <- cast_act[time,]*cast_cst[time,]/agents[[time]][,17];
+        #-- Feeding cost and actions
+        feed_cst[time, 1] <- COST[[time]][3, 11, 1];
+        for(user in 2:users){
+            feed_cst[time, user] <- COST[[time]][1, 11, user];
+        }
+        feed_act[time, 1] <- ACTION[[time]][3, 11, 1] - min_cost;
+        for(user in 2:users){
+            feed_act[time, user] <- ACTION[[time]][1, 11, user];
+        }
+        feed_eff[time,] <- feed_act[time,]*feed_cst[time,]/agents[[time]][,17];
+        #-- Helping cost and actions
+        help_cst[time, 1] <- COST[[time]][3, 12, 1];
+        for(user in 2:users){
+            help_cst[time, user] <- COST[[time]][1, 12, user];
+        }
+        help_act[time, 1] <- ACTION[[time]][3, 12, 1] - min_cost;
+        for(user in 2:users){
+            help_act[time, user] <- ACTION[[time]][1, 12, user];
+        }
+        help_eff[time,] <- help_act[time,]*help_cst[time,]/agents[[time]][,17];
+    }
+
+    # -- Turn these into percentages:
+    scar_eff <- scar_eff * 100;
+    cull_eff <- cull_eff * 100;
+    cast_eff <- cast_eff * 100;
+    feed_eff <- feed_eff * 100;
+    help_eff <- help_eff * 100;
+
+    par(mfrow = c(5, 1), mar = c(0, 0, 0, 0), oma = c(6, 6, 4, 6));
+    y1 <- 100;
+    y2 <- 130;
+    lcex  <- 1.5
+    if(max_time < 100){
+        lcex <- 0.9;
+    }
+    #---- Scaring
+    if(para_vec[89] == 1){
+        plot(x = 1:max_time, y = 100 - scar_eff[,1], type = "l", lwd = 2, 
+             cex.axis = 1.5, xaxt = "n", ylim = c(0, y2), yaxt = "n");
+        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+        par(new = TRUE);
+        plot(x = 1:max_time, y = scar_eff[,2], type = "n", lwd = 2, 
+             ylim = c(0, y2), xaxt="n", yaxt="n", 
+             cex.axis = 1.5);
+        for(stakeholder in 2:users){
+            points(x = 1:max_time, y = scar_eff[,stakeholder], type = "l", 
+                   lwd = 1, col = cols[1]);
+        }
+        axis(side = 4, at = c(50, 100),  labels = c(50, 100), cex.axis = 1.5);
+        legend(x = 1, y = 135, fill = cols[1:5], horiz = TRUE,
+               legend = c("scaring", "culling", "castration", "feeding", 
+                          "helping"), bty = "n", cex = lcex);
+        abline(h = 105, lwd = 2);
+    }else{
+        plot(x = 1:max_time, y = scar_eff[,1], type = "n", xaxt= "n", 
+             yaxt = "n", ylim = c(0, y2));
+        text(x = 0, y = 88, cex = 2, labels = "No scaring allowed",
+             pos = 4);
+        abline(h = 105, lwd = 2);
+        legend(x = 1, y = 135, fill = cols[1:5], horiz = TRUE,
+               legend = c("scaring", "culling", "castration", "feeding", 
+                          "helping"), bty = "n", cex = lcex);
+    }
+    #---- Culling
+    if(para_vec[90] == 1){
+        plot(x = 1:max_time, y = 100 - cull_eff[,1], type = "l", lwd = 2, 
+             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
+        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+        par(new = TRUE);
+        plot(x = 1:max_time, y = cull_eff[,2], type = "n", lwd = 2, 
+             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
+        for(stakeholder in 2:users){
+            points(x = 1:max_time, y = cull_eff[,stakeholder], type = "l", 
+                   lwd = 1, col = cols[2]);
+        }
+        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+    }else{
+        plot(x = 1:max_time, y = cull_act[,1], type = "n", xaxt= "n", 
+             yaxt = "n", ylim = c(0, 100));
+        text(x = 0, y = 90, cex = 2, labels = "No culling allowed",
+             pos = 4);
+    }
+    #---- Castrating
+    if(para_vec[91] == 1){
+        plot(x = 1:max_time, y = 100 - cast_eff[,1], type = "l", lwd = 2, 
+             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
+        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+        par(new = TRUE);
+        plot(x = 1:max_time, y = cast_eff[,2], type = "n", lwd = 2, 
+             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
+        for(stakeholder in 2:users){
+            points(x = 1:max_time, y = cast_eff[,stakeholder], type = "l", 
+                   lwd = 1, col = cols[3]);
+        }
+        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+    }else{
+        plot(x = 1:max_time, y = cast_act[,1], type = "n", xaxt= "n", 
+             yaxt = "n", ylim = c(0, 100));
+        text(x = 0, y = 90, cex = 2, labels = "No castration allowed",
+             pos = 4);
+    }
+    #---- Axes labels
+    mtext("Manager's permissiveness of user action", 
+          side = 2, line = 3.5, cex = 1.5, col = "black");
+    mtext("Percentage of user action effort expended", 
+          side = 4, line = 3.5, cex = 1.5, col = "black");
+    #---- Feeding
+    if(para_vec[92] == 1){
+        plot(x = 1:max_time, y = 100 - feed_eff[,1], type = "l", lwd = 2, 
+             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
+        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+        par(new = TRUE);
+        plot(x = 1:max_time, y = feed_eff[,2], type = "n", lwd = 2, 
+             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
+        for(stakeholder in 2:users){
+            points(x = 1:max_time, y = feed_eff[,stakeholder], type = "l", 
+                   lwd = 1, col = cols[4]);
+        }
+        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+    }else{
+        plot(x = 1:max_time, y = feed_act[,1], type = "n", xaxt= "n", 
+             yaxt = "n", ylim = c(0, 100));
+        text(x = 0, y = 90, cex = 2, labels = "No feeding allowed",
+             pos = 4);
+    }
+    #---- Helping
+    if(para_vec[93] == 1){
+        plot(x = 1:max_time, y = 100 - help_eff[,1], type = "l", lwd = 2, 
+             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
+        axis(side = 2, at = c(50, y2), labels = c(50, 100), cex.axis = 1.5);
+        par(new = TRUE);
+        plot(x = 1:max_time, y = help_eff[,2], type = "n", lwd = 2, 
+             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
+        for(stakeholder in 2:users){
+            points(x = 1:max_time, y = help_eff[,stakeholder], type = "l", 
+                   lwd = 1, col = cols[5]);
+        }
+        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+    }else{
+        plot(x = 1:max_time, y = help_act[,1], type = "n", cex.axis = 1.5,
+             yaxt = "n", ylim = c(0, 100));
+        text(x = 0, y = 90, cex = 2, labels = "No helping offspring allowed",
+             pos = 4);
+    }    
+    mtext("Time step", side = 1, line = 3.5, cex = 1.5, col = "black");
+}
 ```
@@
 
-```{r, echo = TRUE}
-sim_1 <- gmse(manager_budget = 10000, user_budget = 10000, plotting = FALSE, 
-              res_death_K = 93870, manage_target = 70000, RESOURCE_ini = 35000, 
+```{r, echo = TRUE, eval = FALSE}
+sim_1 <- gmse(manager_budget = 10000, user_budget = 10000, res_death_K = 93870, 
+             manage_target = 70000, RESOURCE_ini = 35000, plotting = FALSE, 
              stakeholders = 80, land_ownership = TRUE, public_land = 0.4, 
-              scaring = FALSE, lambda = 0.150, remove_pr = 0.122, time_max = 20, 
-              res_death_type = 3, res_consume = 0.01, res_birth_K = 200000, 
-              observe_type = 3, agent_view = 1);
+             scaring = FALSE, lambda = 0.275, remove_pr = 0.122, time_max = 40, 
+             res_death_type = 3, res_consume = 0.02, res_birth_K = 200000, 
+             observe_type = 3, agent_view = 1, converge_crit = 0.1);
+```
+
+```{r}
+sim_1 <- gmse(time_max = 30, plotting = FALSE);
 ```
@@
 
-```{r, eval = TRUE, fig.height = 8, fig.cap = "Results of a GMSE simulation using parameters estimated for Taiga Been Geese Central Management Unit.
+```{r, eval = TRUE, fig.height = 7, fig.cap = "Results of a GMSE simulation using parameters estimated for Taiga Been Geese Central Management Unit.
 This example includes 80 farmers whose objective is to maximise their agricultural output, and one manager whose objective is to keep geese at a target abundance, over 20 simulated time steps.
@@
 The organge line shows the total percent of landscape cells (including non-farmed cells) yield, as decreased by geese.
-The middle right panel shows this yield for each individual farmer, and for the public land (lower line in blue).
+The middle right panel shows this yield for each farmer, and for the public land (lower line in blue).
 The lower left panel shows the cost of culling for farmers, as set by the manager, and the lower right panel shows the total number of culls attempted by farmers over time."}
@@
 
-```{r, eval = TRUE, fig.height = 7, fig.width = 6, fig.cap = "Permissiveness that each manager exhibits for each farmer action (black lines) and the effort that each individual farmer puts into each action over time (coloured lines).
+```{r, eval = TRUE, echo = TRUE, fig.height = 7, fig.width = 6, fig.cap = "Permissiveness that each manager exhibits for each farmer action (black lines) and the effort that each individual farmer puts into each action over time (coloured lines).
 Each panel row reports a different action (in decreasing order: scaring, culling, castration, feeding, and helping).
@@
 
-```{r, echo = TRUE, fig.height = 8}
-sim_2 <- gmse(manager_budget = 10000, user_budget = 10000, plotting = FALSE, 
-              res_death_K = 93870, manage_target = 70000, RESOURCE_ini = 35000, 
+```{r}
+sim_2 <- gmse(time_max = 30, scaring = TRUE, plotting = FALSE);
+```
+
+```{r, echo = TRUE, fig.height = 6, eval = FALSE}
+sim_2 <- gmse(manager_budget = 10000, user_budget = 10000, res_death_K = 93870, 
+             manage_target = 70000, RESOURCE_ini = 35000, plotting = FALSE, 
              stakeholders = 80, land_ownership = TRUE, public_land = 0.4, 
-              scaring = TRUE, lambda = 0.150, remove_pr = 0.122, time_max = 20, 
-              res_death_type = 3, res_consume = 0.01, res_birth_K = 200000, 
-              observe_type = 3, agent_view = 1);
+             scaring = TRUE, lambda = 0.275, remove_pr = 0.122, time_max = 40, 
+             res_death_type = 3, res_consume = 0.02, res_birth_K = 200000, 
+             observe_type = 3, agent_view = 1, converge_crit = 0.1);
 ```
@@
 ```
-
-References
---------------------------------------------------------------------------------

### 2018-04-18 notebook/ms/SI3.Rmd
More SI edits

@@
 They estimated goose survival under ideal conditions to be ca 0.878; this can be interpreted in our model by setting mortality to `remove_pr = 1 - 0.878`.
-Similarly, @Johnson2018 estimated intrinsic population growth rate and carrying capacity to be 0.15 and 93870, respectively, so we set `lambda = 0.183` (for convenience of demonstration, the upper 90% confidence interval of their estimate) and `res_death_K = 93870`.
+Similarly, @Johnson2018 estimated intrinsic reproductive rate and carrying capacity to be 0.55 and 93870, respectively, so we set `lambda = 0.275` (for simplicity, we simply use half the mean reproductive rate; GMSE does not currently distinguish female and male individuals) and `res_death_K = 93870`.
 The global abundance of Taiga Bean Geese in 2009 was ca 63000 [@Fox2010], and ca 35000 in the Central Management Unit [@AEWA2016], which we can take as a starting abundance for our simulations (`RESOURCE_ini = 35000`).
@@
 
+Our goal is not to provide a detailed case study of the Taiga Bean Geese, but rather to demonstrate how such a case study would be possible in GMSE.
 For simplicity, here we assess confict using only the `gmse` function to show how parameter values can be set to provide useful results.
-Our goal is not to provide a detailed case study of the Taiga Bean Geese, but rather to demonstrate how such a case study would be possible using the `gmse` function.
-Additionally, all of the analysis that follows would also be possibe using the more detailed, but also more flexible, `gmse_apply` function.
+Novice R users may prefer run all of the simulations below using the browser-based GMSE GUI by calling `gmse_gui()` from the R command line.
+Alternatively, experienced R users may prefer simulate by looping time steps through `gmse_apply`, which allows more flexibility for incorporating custom submodels and dynamically adjusting parameter values.
 Where available, we use estimated demographic parameter values from @Johnson2018 and @AEWA2016.
@@
 Next, we run the same simulation but allow users to scare geese from their landscape cells; we then use the model to make inferences regarding how scaring as a management option might affect goose population dynamics, agricultural production, and conservation conflict in the system.
-All simulations are run using the `gmse` function, but novice R users may prefer run all of the simulations below using the browser-based GMSE GUI by calling `gmse_gui()` from the R command line.
-Alternatively, experienced R users may prefer simulate by looping time steps through `gmse_apply`, which allows more flexibility for incorporating custom submodels and dynamically adjusting parameter values.
 We emphasise that the simulations below are intended only to demonstrate one use of GMSE on a species of conservation interest, not to make recommendations for management of Taiga Bean Geese.
@@
 
-The results of the above simulation are plotted in Figure SI3-1 below.
+The results of the above simulation are plotted in Figure 1 below.
 
@@
 Actual goose abundance is shown in the middle left panel (black solid line), along with its estimate by the manager (blue solid line).
-Horizontal red and blue dotted lines show landscape-level goose carrying capacity enacted on adult mortality and the manager's target for goose abundance, respectively.
+The horizontal blue dotted line shows the manager's target for goose abundance.
 The organge line shows the total percent of landscape cells (including non-farmed cells) yield, as decreased by geese.
@@
 
-The above Figure SI3-1 shows the dynamics of goose abundance and agricultural yield, along with how managers react to change in abundance and farmers react to manager policy.
-In the case of the simulation above, goose population begins to decline, causing managers to set policy to increase the cost of culling.
-This higher cost of culling restricts farmers from culling on their land, leading to fewer total culling actions.
-In this case, the abundance of geese continues downward, while agricultural yield increases over time.
-We can investigate the conflict between management policy and farmers using the `plot_gmse_effort` function in GMSE.
+Figure 1 shows the dynamics of goose abundance and agricultural yield, along with how managers react to change in abundance and farmers react to manager policy.
+In the case of the simulation above, managers quickly set a policy of high cost for culling, which leads to a rise in the goose population and a decrease in crop yield for farmers.
+After roughly 20 time steps, the goose population rises above the manager target, at which point the manager becomes more permissive of culling and the cost of culling for users therefore declines.
+In response, users begin to cull geese on their land, and the goose population begins to stabilise around the target of 7000 total geese.
+We can investigate the conflict between management policy and farmers more directly using the `plot_gmse_effort` function in GMSE.
 
@@
 When black lines are far below coloured lines, we can (cautiously) interpret this as a conflict between management of the goose population and farmer's interest in agricultural production.
-These time periods represent instances in which managers are not at very permissive of a particular action (in this case culling), but farmers continue to expend effort to do the action anyway.
+These time periods represent instances in which the manager is not at very permissive of a particular action (in this case culling), but farmers continue to expend effort to do the action anyway.
+In the case of the above simulation of potential conservation conflict between farmers and goose conservation, conflict is highest before time step 20, where the manager is not permissive of culling because the population is below the manager's target.
+Once the goose population has increased above the manager's target, conflict decreases because the desired culling is permitted by managers to keep the population at a target abundance.
+It is worth noting that, despite conflict as we define it decreasing, agricultural damage is still relatively high after the target goose population is achieved (Figure 1).
+Hence, on a broader scale, conflict might persist around the appropriate target population size rather than what actions are permitted for farmers; currently, this potential aspect of conflict is not modelled, but future versions of GMSE may attempt to incorporate such additional complexity in conflict scenarios.
 
-```{r}
+We can model the consequences for goose population dynamics, agricultural produciton, and conservation conflict when scaring is a policy option available to the manager.
+The code below is a simulation identical to the one just described, but with a scaring option included through `scaring = TRUE`.
+
+```{r, echo = FALSE}
 sim_2 <- gmse(time_max = 30, scaring = TRUE, plotting = FALSE);
@@
 
-The results are plotted below.
+The results are plotted in Figure 3.
 

### 2018-04-18 notebook/ms/SI3.Rmd
Draft without the actual simulations included yet

@@
 
-```{r, eval = TRUE, fig.height = 8}
+```{r, eval = TRUE, fig.height = 8, fig.cap = "Results of a GMSE simulation using parameters estimated for Taiga Been Geese Central Management Unit in which scaring is permitted.
+Simulation output is interpreted as in Figure 1"}
 plot_gmse_results(res = sim_2$resource, obs = sim_2$observation, 
@@
 
-Look at the conflict
+When scaring is introduced to an otherwise identical simulation (compare Figure 3 to Figure 1), the goose population increases as before, but it achieves the managers target population size and stabilises 2-3 time steps earlier.
+The reason for this earlier stabilisation is due to the change in farmer's actions as a consequence of the introduced scaring policy.
+At the start of the simulation, managers adjust policy by quickly increasing the cost of culling and decreasing the cost of scaring.
+In response, farmers turn to scaring rather than culling to remove geese from their land cells (Figure 3).
+This is in contrast to the simulation in which scaring was not an option, and farmers simply culled as much as possible despite the high costs (Figure 1).
+After the population has risen to slightly above the manager's target, the cost of culling again decreases, with the manager balancing the incentivisation of culling and scaring.
+The consequence of scaring as an available policy also reveals some potentially unexpected outcomes, such as increased variance in among-farmer agricultural production, which arises as geese are scared from one area of the landscape to another.
 
-```{r, eval = TRUE, fig.height = 7, fig.width = 6}
+We can use the `plot_gmse_effort` function as before to investigate how the inclusion of scaring as a policy option might affect conservation conflict.
+Conflict results when scaring is included are plotted in Figure 4.
+
+```{r, eval = TRUE, fig.height = 7, fig.width = 6, fig.cap = "Permissiveness that each manager exhibits for each farmer action (black lines) and the effort that each individual farmer puts into each action over time (coloured lines) given scaring as a possible policy option.
+Simulation output is intepreted as in Figure 2."}
 plot_gmse_effort(sim_2$agents, sim_2$paras, ACTION = sim_2$action, 
@@
 ```
+
+Unlike the case in which culling was the only policy option (compare Figure 4 to Figure 2), the effort that farmers expended on a given action did not typically rise above the manager's permissiveness of the action.
+Hence, under the conditions of this model, the inclusion of scaring as a policy option has reduced conservation conflict in the social-ecological system.
+We again emphasise that the simulations presented here only serve as an example of how GMSE could be used as a tool for simulating social-ecological systems and understanding the potential for conflict; it is not intended to inform policy in Taiga Bean Geese or any other specific system.
+
+Conclusions and future development
+--------------------------------------------------------------------------------
+
+The GMSE function `gmse` and its graphical user interface counterpart `gmse_gui` offer wide set of options for parameterising simulations to fit empirical case studies of conservation interest using default GMSE natural resource, observation, manager, and user sub-models.
+Future development of these sub-models might usefully incorporate additional details relevant to specific case studies, such population structure, multiple natural resource and user types, or different policy and action possibilities.
+[Requests for new features](https://github.com/bradduthie/gmse/issues) and contributions to GMSE can be made through [GitHub](https://github.com/bradduthie/gmse).
+Additionally, where entirely different types of sub-models are required, the `gmse_apply` function [can be used](https://bradduthie.github.io/gmse_SI2.pdf) to more flexibly simulate social-ecological systems.

### 2018-04-18 notebook/ms/SI3.Rmd
First use of data in the SI

@@
 Currently, there is no standard way to measure conservation conflict in a social-ecological system, and previous modelling approaches have not meaningfully separated agent objectives from agent actions.
-We suggest that a starting point to developing a useful metric of conservation conflict is to quantify the deviation of an individual's actions from their objectives, the former of which is restricted by the actions of other individuals.
+We suggest that a starting point to developing a useful metric of conservation conflict is to quantify the deviation of an individual's actions from their objectives  (i.e., of actual actions from desired actions), the former of which is restricted by the actions of other individuals.
 Here we show how GMSE can be used to evaluate the amount of conflict in a simulated social-ecological system under different management options.
@@
 We consider a protected population of waterfowl that exploits agricultural land and is therefore a source of conservation conflict [e.g., @Fox2017a; @Mason2017; @Tulloch2017].
-As in the main text example, the objective of managers is to keep waterfowl at a target abundance, while the objective of farmers is to maximise agricultural production on their landscape.
-Here we consider a more complex simulation to reflect a scenario that is likely to occur in a real social-ecological system.
-We parameterise our model using demographic information from the Taiga Bean Geese (*Anser fabalis fabalis*), a managed population that is hunted for sport in Fennoscandinavia [@Johnson2018].
-Taiga Bean Geese can also cause agricultural damage, which can potentially lead to conflict between farmers and management objectives [@Johnson2018].
+As in the main text example, the objective of the manager is to keep waterfowl at a target abundance, while the objective of farmers is to maximise agricultural production on their landscape.
+Here we consider a more complex simulation, with a level of detail that more accurately reflects a scenario that might occur in a real social-ecological system.
+We parameterise our model using demographic information from the Taiga Bean Goose (*Anser fabalis fabalis*), a managed population that is hunted for sport in Fennoscandinavia [@Johnson2018].
+Taiga Bean Geese can cause agricultural damage, which can potentially lead to conflict between farmers and management objectives [@Johnson2018].
 
@@
 
-```{r, echo = FALSE}
-sim_2 <- gmse(time_max = 30, scaring = TRUE, plotting = FALSE);
-```
-
 ```{r, echo = TRUE, fig.height = 6, eval = FALSE}
@@
              observe_type = 3, agent_view = 1, converge_crit = 0.1);
+```
+
+```{r, echo = FALSE, eval = TRUE}
+load("data/SI3-scare_sim.rda"); # Avoids long compiling time
+sim_2 <- sim;
 ```

### 2018-04-18 R/plotting.R
Adjust plot to ensure carrying capacity is visible

@@ -601,6 +601,9 @@ plot_gmse_results <- function(res, obs, land, agents, paras, ACTION, COST){
     }
     
     ymaxi <- max(paras[,33]) + (0.1 * (max(paras[,33])));
+    if(ymaxi < para_vec[7]){
+        ymaxi <- para_vec[7] + (0.1 * para_vec[7]);
+    }
     
     time_max <- length(res);
     for(i in 1:(time_max-1)){

### 2018-04-18 notebook/ms/SI3.Rmd
Minor edits

@@
 Actual goose abundance is shown in the middle left panel (black solid line), along with its estimate by the manager (blue solid line).
-The horizontal blue dotted line shows the manager's target for goose abundance.
+The horizontal red and blue dotted lines show goose carrying capacity and the manager's target for goose abundance, respectively.
 The organge line shows the total percent of landscape cells (including non-farmed cells) yield, as decreased by geese.
@@
 
-```{r, eval = TRUE, fig.height = 7, fig.width = 6, fig.cap = "Permissiveness that each manager exhibits for each farmer action (black lines) and the effort that each individual farmer puts into each action over time (coloured lines) given scaring as a possible policy option.
+```{r, eval = TRUE, fig.height = 7, fig.width = 7, fig.cap = "Permissiveness that each manager exhibits for each farmer action (black lines) and the effort that each individual farmer puts into each action over time (coloured lines) given scaring as a possible policy option.
 Simulation output is intepreted as in Figure 2."}

### 2018-04-18 notebook/ms/SI3.Rmd
Edits to SI

@@
 
-@Johnson2018 recently estimated key demographic parameters for the Taiga Bean Geese for the Central Management Unit that includes geese that breed in "Northern most Sweden, Northern Norway, Northern and Central Finland, and adjacent North-wester parts of Russia, wintering mostly in Southern Sweden and South-east Denmark" [@AEWA2016].
-They estimated goose survival under ideal conditions to be ca 0.878; this can be interpreted in our model by setting mortality to `remove_pr = 1 - 0.878`.
-Similarly, @Johnson2018 estimated intrinsic reproductive rate and carrying capacity to be 0.55 and 93870, respectively, so we set `lambda = 0.275` (for simplicity, we simply use half the mean reproductive rate; GMSE does not currently distinguish female and male individuals) and `res_death_K = 93870`.
-The global abundance of Taiga Bean Geese in 2009 was ca 63000 [@Fox2010], and ca 35000 in the Central Management Unit [@AEWA2016], which we can take as a starting abundance for our simulations (`RESOURCE_ini = 35000`).
-And the International Single Species Action Plan has a target population size of ca 70000 in the Central Management Unit [@AEWA2016], which we can use as a management target (`manage_target = 70000`).
-
 Our goal is not to provide a detailed case study of the Taiga Bean Geese, but rather to demonstrate how such a case study would be possible in GMSE.
-For simplicity, here we assess confict using only the `gmse` function to show how parameter values can be set to provide useful results.
-Novice R users may prefer run all of the simulations below using the browser-based GMSE GUI by calling `gmse_gui()` from the R command line.
-Alternatively, experienced R users may prefer simulate by looping time steps through `gmse_apply`, which allows more flexibility for incorporating custom submodels and dynamically adjusting parameter values.
+For simplicity, here we assess conflict using only the `gmse` function to show how parameter values can be set to provide useful results.
+Novice R users may prefer to run all of the simulations below using the browser-based GMSE GUI by calling `gmse_gui()` from the R command line.
+Alternatively, experienced R users may prefer to simulate by [looping time steps](https://bradduthie.github.io/gmse_SI2.pdf#page=5) through `gmse_apply`, which allows more flexibility for incorporating custom sub-models and dynamically adjusting parameter values.
 Where available, we use estimated demographic parameter values from @Johnson2018 and @AEWA2016.
@@
 
+@Johnson2018 recently estimated key demographic parameters for the Taiga Bean Geese for the Central Management Unit that includes geese that breed in "Northern most Sweden, Northern Norway, Northern and Central Finland, and adjacent North-wester parts of Russia, wintering mostly in Southern Sweden and South-east Denmark" [@AEWA2016].
+They estimated goose survival under ideal conditions to be ca 0.878; this can be interpreted in our model by setting mortality to `remove_pr = 1 - 0.878`.
+Similarly, @Johnson2018 estimated mean reproductive rate and carrying capacity to be 0.55 and 93870, respectively, so we set `lambda = 0.275` (for simplicity, we simply use half the mean reproductive rate; GMSE does not currently distinguish female and male individuals) and `res_death_K = 93870`.
+The global abundance of Taiga Bean Geese in 2009 was ca 63000 [@Fox2010], and ca 35000 in the Central Management Unit [@AEWA2016], which we can take as a starting abundance for our simulations (`RESOURCE_ini = 35000`).
+And the International Single Species Action Plan has a target population size of ca 70000 in the Central Management Unit [@AEWA2016], which we can use as a management target (`manage_target = 70000`).
+We simulate social-ecological dynamics over 30 time steps, which could be interpreted as years.
+
 The code below calls `gmse` using the empirically derived parameters for Taiga Bean Geese described above.
 We also set `manager_budget = 100000` and `user_budget = 100000`.
-Further, we consider the case in of a region in which farmland makes up 60% of all land, with 40% of land being 'public' (`public_land = 0.4`; which might be intpreted as any land in which stakeholders are not, or cannot be, invested in goose presence), and divide the farmland amongst 80 individual farmers (`stakeholders = 80`; `land_ownership = TRUE`).
+Further, we consider the case of a region in which farmland makes up 60% of all land, with 40% of land being 'public' (`public_land = 0.4`; which might be interpreted as any land in which stakeholders are not, or cannot be, invested in goose presence), and divide the farmland amongst 80 individual farmers (`stakeholders = 80`; `land_ownership = TRUE`).
 Landscape size is set to default 100 by 100 cells, so each farmer effectively owns 75 cells, which might be interpreted as hectares of land (for instructions on how to more precisely control landscape ownership, see the advanced GMSE options using `gmse_apply`).
@@
 We assume that a single goose decreases agricultural production on a cell by 2% per time step (`res_consume = 0.02`).
-We further assume that the population is well-monitored, with observers counting goose numbers on each cell of the landscape in every timestep (`observe_type = 3`) with the ability to observe two landscape cells in every direction (`agent_view = 2`).
+We further assume that the population is very well-monitored, with observers counting goose numbers on each cell of the landscape in every timestep (`observe_type = 3`) with the ability to observe one landscape cell in every direction (`agent_view = 1`).
 All other parameter values are set to GMSE defaults.
@@
 
-Below, we first allow only the culling of geese by users and plot the dynamics of the social-ecological system from a single simulation.
-Next, we run the same simulation but allow users to scare geese from their landscape cells; we then use the model to make inferences regarding how scaring as a management option might affect goose population dynamics, agricultural production, and conservation conflict in the system.
+Below, we first only allow culling as a policy option and plot the dynamics of the social-ecological system from a single simulation.
+Next, we run the same simulation but also allow scaring as a policy option; we then use the model to make inferences regarding how scaring as a management option might affect goose population dynamics, agricultural production, and conservation conflict in the system.
 We emphasise that the simulations below are intended only to demonstrate one use of GMSE on a species of conservation interest, not to make recommendations for management of Taiga Bean Geese.
@@
 In response, users begin to cull geese on their land, and the goose population begins to stabilise around the target of 7000 total geese.
-We can investigate the conflict between management policy and farmers more directly using the `plot_gmse_effort` function in GMSE.
+We can investigate the conflict between management policy and farmers more directly using the `plot_gmse_effort` function.
 
@@
 When black lines are far below coloured lines, we can (cautiously) interpret this as a conflict between management of the goose population and farmer's interest in agricultural production.
-These time periods represent instances in which the manager is not at very permissive of a particular action (in this case culling), but farmers continue to expend effort to do the action anyway.
+These time periods represent instances in which the manager is not permissive of a particular action (in this case culling), but farmers continue to expend effort to do the action anyway.
 In the case of the above simulation of potential conservation conflict between farmers and goose conservation, conflict is highest before time step 20, where the manager is not permissive of culling because the population is below the manager's target.
 Once the goose population has increased above the manager's target, conflict decreases because the desired culling is permitted by managers to keep the population at a target abundance.
-It is worth noting that, despite conflict as we define it decreasing, agricultural damage is still relatively high after the target goose population is achieved (Figure 1).
+It is worth noting that, despite conflict as we define it decreasing, agricultural damage is still relatively high after the target goose population size is achieved (Figure 1).
 Hence, on a broader scale, conflict might persist around the appropriate target population size rather than what actions are permitted for farmers; currently, this potential aspect of conflict is not modelled, but future versions of GMSE may attempt to incorporate such additional complexity in conflict scenarios.
 
-We can model the consequences for goose population dynamics, agricultural produciton, and conservation conflict when scaring is a policy option available to the manager.
-The code below is a simulation identical to the one just described, but with a scaring option included through `scaring = TRUE`.
+We can model the consequences for goose population dynamics, agricultural production, and conservation conflict when scaring is a policy option available to the manager.
+The code below runs a simulation identical to the one just discussed, but with a scaring option included using the argument `scaring = TRUE`.
 
@@
 
-When scaring is introduced to an otherwise identical simulation (compare Figure 3 to Figure 1), the goose population increases as before, but it achieves the managers target population size and stabilises 2-3 time steps earlier.
+When scaring is introduced to an otherwise identical simulation (compare Figure 3 to Figure 1), the goose population increases as before, but it achieves the manager's target population size and stabilises 2-3 time steps earlier.
 The reason for this earlier stabilisation is due to the change in farmer's actions as a consequence of the introduced scaring policy.
@@
 
-Unlike the case in which culling was the only policy option (compare Figure 4 to Figure 2), the effort that farmers expended on a given action did not typically rise above the manager's permissiveness of the action.
+Unlike the case in which culling was the only policy option (compare Figure 4 to Figure 2), the effort that farmers expended on a given action did not rise as highly above the manager's permissiveness of the action.
 Hence, under the conditions of this model, the inclusion of scaring as a policy option has reduced conservation conflict in the social-ecological system.
-We again emphasise that the simulations presented here only serve as an example of how GMSE could be used as a tool for simulating social-ecological systems and understanding the potential for conflict; it is not intended to inform policy in Taiga Bean Geese or any other specific system.
+We again emphasise that the simulations presented here only serve as an example for how GMSE could be used as a tool for simulating social-ecological systems and understanding the potential for conflict; it is not intended to inform policy in Taiga Bean Geese or any other specific system.
 
@@
 
-The GMSE function `gmse` and its graphical user interface counterpart `gmse_gui` offer wide set of options for parameterising simulations to fit empirical case studies of conservation interest using default GMSE natural resource, observation, manager, and user sub-models.
-Future development of these sub-models might usefully incorporate additional details relevant to specific case studies, such population structure, multiple natural resource and user types, or different policy and action possibilities.
+The GMSE function `gmse` and its graphical user interface counterpart `gmse_gui` offer wide a suite of options for parameterising simulations to fit empirical case studies of conservation interest using default GMSE natural resource, observation, manager, and user sub-models.
+Future development of these sub-models might usefully incorporate additional details relevant to specific case studies, such population structure, multiple natural resource and user types, or different manager policy and user action possibilities.
 [Requests for new features](https://github.com/bradduthie/gmse/issues) and contributions to GMSE can be made through [GitHub](https://github.com/bradduthie/gmse).

### 2018-04-18 notebook/ms/SI3.Rmd
Fix the sim calls

@@
              stakeholders = 80, land_ownership = TRUE, public_land = 0.4, 
-             scaring = FALSE, lambda = 0.275, remove_pr = 0.122, time_max = 40, 
+             scaring = FALSE, lambda = 0.275, remove_pr = 0.122, time_max = 30, 
              res_death_type = 3, res_consume = 0.02, res_birth_K = 200000, 
-             observe_type = 3, agent_view = 1, converge_crit = 0.1);
+             observe_type = 3, agent_view = 1, converge_crit = 0.01, 
+             ga_mingen = 200);
 ```
 
-```{r}
-sim_1 <- gmse(time_max = 30, plotting = FALSE);
+```{r, echo = FALSE, eval = TRUE}
+load("data/SI3-cull_sim.rda"); # Avoids long compiling time
+sim_1 <- sim;
 ```
@@
              stakeholders = 80, land_ownership = TRUE, public_land = 0.4, 
-             scaring = TRUE, lambda = 0.275, remove_pr = 0.122, time_max = 40, 
+             scaring = TRUE, lambda = 0.275, remove_pr = 0.122, time_max = 30, 
              res_death_type = 3, res_consume = 0.02, res_birth_K = 200000, 
-             observe_type = 3, agent_view = 1, converge_crit = 0.1);
+             observe_type = 3, agent_view = 1, converge_crit = 0.01, 
+             ga_mingen = 200);
 ```

### 2018-04-20 src/game.c
Minor change

@@ -266,7 +266,7 @@ void res_to_counts(double ***population, int **interact_table, double *paras,
     
     foc_effect  = 0.0;
     foc_effect -= population[row][7][agent];  /* Times birth account for repr?*/
-    foc_effect -= population[row][8][agent]  * (1.0 + lambda);
+    foc_effect -= population[row][8][agent];
     foc_effect -= population[row][9][agent]  * lambda;
     foc_effect += population[row][10][agent] * lambda;
     foc_effect += population[row][11][agent]; /* But should affect offspring? */

### 2018-04-20 notebook/ms/SI3.Rmd
Small edit

@@
 
-@Johnson2018 recently estimated key demographic parameters for the Taiga Bean Geese for the Central Management Unit that includes geese that breed in "Northern most Sweden, Northern Norway, Northern and Central Finland, and adjacent North-wester parts of Russia, wintering mostly in Southern Sweden and South-east Denmark" [@AEWA2016].
+@Johnson2018 recently estimated key demographic parameters of the Taiga Bean Geese from the Central Management Unit, which includes geese that breed in "Northern most Sweden, Northern Norway, Northern and Central Finland, and adjacent North-wester parts of Russia, wintering mostly in Southern Sweden and South-east Denmark" [@AEWA2016].
 They estimated goose survival under ideal conditions to be ca 0.878; this can be interpreted in our model by setting mortality to `remove_pr = 1 - 0.878`.
@@
 The code below calls `gmse` using the empirically derived parameters for Taiga Bean Geese described above.
-We also set `manager_budget = 100000` and `user_budget = 100000`.
+We also set `manager_budget = 10000` and `user_budget = 10000`.
 Further, we consider the case of a region in which farmland makes up 60% of all land, with 40% of land being 'public' (`public_land = 0.4`; which might be interpreted as any land in which stakeholders are not, or cannot be, invested in goose presence), and divide the farmland amongst 80 individual farmers (`stakeholders = 80`; `land_ownership = TRUE`).
@@
 ```{r, eval = TRUE, fig.height = 7, fig.cap = "Results of a GMSE simulation using parameters estimated for Taiga Been Geese Central Management Unit.
-This example includes 80 farmers whose objective is to maximise their agricultural output, and one manager whose objective is to keep geese at a target abundance, over 20 simulated time steps.
+This example includes 80 farmers whose objective is to maximise their agricultural output, and one manager whose objective is to keep geese at a target abundance, over 30 simulated time steps.
 Goose locations at the end of the simulation are shown in the upper left panel, while the upper right panel shows the same landscape broken down among the 80 farmers (upper 60% of the landscape in multiple colours), along with non-agricultural land (lower 40% of the landscape in blue).
@@
 The horizontal red and blue dotted lines show goose carrying capacity and the manager's target for goose abundance, respectively.
-The organge line shows the total percent of landscape cells (including non-farmed cells) yield, as decreased by geese.
+The organge line shows the total percent of landscape cell (including non-farmed cells) yield, as decreased by geese.
 The middle right panel shows this yield for each farmer, and for the public land (lower line in blue).
@@
 These time periods represent instances in which the manager is not permissive of a particular action (in this case culling), but farmers continue to expend effort to do the action anyway.
-In the case of the above simulation of potential conservation conflict between farmers and goose conservation, conflict is highest before time step 20, where the manager is not permissive of culling because the population is below the manager's target.
+In the case of the above simulation of potential conflict between farmers and goose conservation, conflict is highest before time step 20, where the manager is not permissive of culling because the population is below the manager's target.
 Once the goose population has increased above the manager's target, conflict decreases because the desired culling is permitted by managers to keep the population at a target abundance.

### 2018-04-24 notebook/ms/SI3.Rmd
Minor tweaks

@@
 To demonstrate how GMSE can be used to understand conflict in social-ecological systems, we build upon the [example of resource management](https://www.biorxiv.org/content/biorxiv/early/2017/11/17/221432.full.pdf#page=5) in the main text.
-We consider a protected population of waterfowl that exploits agricultural land and is therefore a source of conservation conflict [e.g., @Fox2017a; @Mason2017; @Tulloch2017].
+We consider a protected population of waterfowl that exploits agricultural land and is therefore a source of conservation conflict [e.g., @Fox2017a; @Mason2017; @Tulloch2017; @Cusack2018].
 As in the main text example, the objective of the manager is to keep waterfowl at a target abundance, while the objective of farmers is to maximise agricultural production on their landscape.
 Here we consider a more complex simulation, with a level of detail that more accurately reflects a scenario that might occur in a real social-ecological system.
-We parameterise our model using demographic information from the Taiga Bean Goose (*Anser fabalis fabalis*), a managed population that is hunted for sport in Fennoscandinavia [@Johnson2018].
-Taiga Bean Geese can cause agricultural damage, which can potentially lead to conflict between farmers and management objectives [@Johnson2018].
+Our objective is not to model the dynamics of a specific system, but to show how GMSE could be parameterised using demographic estimates from empirical studies.
+We therefore consider an example population in which such estimates are well-reported and readily available.
+
+We parameterise our model using demographic information from the Taiga Bean Goose (*Anser fabalis fabalis*), a managed population that is hunted for recreation in Fennoscandinavia [@Johnson2018].
+Taiga Bean Geese can cause agricultural damage [@Johnson2018], which could potentially lead to conflict between farmers and management objectives.
 

### 2018-04-24 notebook/ms/SI4.Rmd
Initialise a fourth SI

---
title: 'GMSE: an R package for generalised management strategy evaluation'
author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
  Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;,
  Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes: null
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: gmse.bib
subtitle: Supporting Information 4
biblio-style: apalike
---

### 2018-04-24 notebook/ms/SI4.Rmd
Initial progress on gmse_apply for an empirical case study

@@
 ---
+
+Fine-tuning simulation conditions using `gmse_apply`
+================================================================================
+
+Here we demonstrate how simulations in GMSE can be more fine-tuned to specific empirical situations through the use of `gmse_apply`.
+To do this, we use the same scenario described in [Supporting Information 3](); we first recreate the basic scenario run in `gmse` using `gmse_apply`, and then build in additional modelling details including (1) custom placement of user land, (2) parameterisation of individual user budgets, and (3) density-dependent movement of resources.
+We emphasise that these simulations are provided only to demonstrate the use of GMSE, and specifically to show the flexibility of the `gmse_apply` function, not to accurately recreate the dynamics of a specific system or make management recommendations.
+
+We reconsider the case of a protected waterfowl population that exploits agricultural land [e.g., @Fox2017a; @Mason2017; @Tulloch2017; @Cusack2018].
+The manager attempts to keep the watefowl at a target abundance, while users (farmers) attempt to maximise agricultural yield on the land that they own.
+We again parameterise our model using demographic information from the Taiga Bean Goose (*Anser fabalis fabalis*), as reported by @Johnson2018 and @AEWA2016.
+Relevant parameter values are listed in the table below.
+
+| Parameter         | Value    | Description                                      |
+|-------------------|----------|--------------------------------------------------|
+| `remove_pr`       | 0.122    | Goose density-independent mortality probability  |
+| `lambda`          | 0.275    | Expected offspring production per time step      |
+| `res_death_K`     | 93870    | Goose carrying capacity (on adult mortality)     |
+| `RESOURCE_ini`    | 35000    | Initial goose abundance                          |
+| `manage_target`   | 70000    | Manager's target goose abundance                 |
+| `res_death_type`  | 3        | Mortality (density and density-indepent sources) |
+
+Table: GMSE simulation parameter values inspired by @Johnson2018 and @AEWA2016
+
+Additionally, we continue to use the following values for consistency.
+
+| Parameter         | Value    | Description                                      |
+|-------------------|----------|--------------------------------------------------|
+| `manager_budget`  | 10000    | Manager's budget for setting policy options      |
+| `user_budget`     | 10000    | Users' budgets for actions                       |
+| `public_land`     | 0.4      | Proportion of the landscape that is public       |
+| `stakeholders`    | 80       | Number of stakeholders                           |
+| `land_ownership`  | TRUE     | Users own landscape cells                        |
+| `res_consume`     | 0.02     | Landscape cell output consumed by a resource     | 
+| `observe_type`    | 3        | Observation model type (survey)                  |
+| `agent_view`      | 1        | Cells managers can see when conducting a survey  |
+
+Table: Non-default GMSE parameter values chosen by authors
+
+All other values are set to GMSE defaults, except where specifically noted otherwise.
+
+Re-creating `gmse` simulations using `gmse_apply`
+================================================================================
+
+We now recreate the simulations in [Supporting Information 3](), which were run using the `gmse` function, in `gmse_apply`.
+Doing so requires us to first initialise simulations using one run of `gmse_apply`, then loop through multiple time steps that again call `gmse_apply` and saving the results of interest.

### 2018-04-24 notebook/ms/SI4.Rmd
List of stuff, moving ot customising the landscape

@@
 
+```{r, echo = FALSE}
+library(GMSE);
+```
+
 Fine-tuning simulation conditions using `gmse_apply`
@@
 
-Additionally, we continue to use the following values for consistency.
+Additionally, we continue to use the following values for consistency, except in the case of `stakeholders`, where we reduce the number of farmers to `stakehoders = 8` and increase `user_budget` to 100000.
+This is done to for two reasons.
+First, it speeds up simulations for the purpose of demonstration; second, it makes the presentation of landscape ownership easier (see below).
 
@@
 | `manager_budget`  | 10000    | Manager's budget for setting policy options      |
-| `user_budget`     | 10000    | Users' budgets for actions                       |
+| `user_budget`     | 100000   | Users' budgets for actions                       |
 | `public_land`     | 0.4      | Proportion of the landscape that is public       |
-| `stakeholders`    | 80       | Number of stakeholders                           |
+| `stakeholders`    | 8        | Number of stakeholders                           |
 | `land_ownership`  | TRUE     | Users own landscape cells                        |
@@
 Doing so requires us to first initialise simulations using one run of `gmse_apply`, then loop through multiple time steps that again call `gmse_apply` and saving the results of interest.
+Following instructions in [Supporting Information 1](), we can call the initialising simulation `sim_old`, and use the code below to read in the relevant parameter values.
+
+```{r, echo = TRUE, eval = TRUE}
+sim_old  <- gmse_apply(get_res = "Full", remove_pr = 0.122, lambda = 0.275,
+                       res_death_K = 93870, RESOURCE_ini = 35000, 
+                       manage_target = 70000, res_death_type = 3,
+                       manager_budget = 10000, user_budget = 100000,
+                       public_land = 0.4, stakeholders = 8, 
+                       land_ownership = TRUE, res_consume = 0.02, 
+                       observe_type = 3, agent_view = 1);
+```
+
+Note that the argument `get_res = "Full"` causes `sim_old` retain all of the relevant data structures for simulating a new time step and recording simulation results.
+This includes the key simulation output, which is located in `sim_old$basic_output`, which is printed below.
+
+```{r, echo = FALSE}
+print(sim_old$basic_output);
+```
+
+We can then loop over 30 time steps to recreate the simulations from [Supporting Information 3]().
+In these simulations, we are specifically interested in the resource and observation outputs, as well as the manager policy and user actions for culling, which we record below in the data frame `sim_sum`.
+The inclusion of the argument `old_list` tells `gmse_apply` to use parameters and values from the list `sim_old` in the new time step.
+
+```{r}
+sim_sum_1 <- matrix(data = NA, nrow = 30, ncol = 5);
+for(time_step in 1:30){
+    sim_new                 <- gmse_apply(get_res = "Full", old_list = sim_old);
+    sim_sum_1[time_step, 1] <- time_step;
+    sim_sum_1[time_step, 2] <- sim_new$basic_output$resource_results[1];
+    sim_sum_1[time_step, 3] <- sim_new$basic_output$observation_results[1];
+    sim_sum_1[time_step, 4] <- sim_new$basic_output$manager_results[3];
+    sim_sum_1[time_step, 5] <- sum(sim_new$basic_output$user_results[,3]);
+    sim_old                 <- sim_new;
+}
+colnames(sim_sum_1) <- c("Time", "Pop_size", "Pop_est", "Cull_cost", 
+                         "Cull_count");
+print(sim_sum_1);
+```
+
+The above output from `sim_sum_1` shows the table that holds the information we were interested in pulling out of our simulation results.
+All of this information was available under the list element `sim_new$basic_output`, but other list elements of `sim_new` might also be useful to record.
+It is important to also remember that this example of `gmse_apply` is using the default resource, observation, manager, and user submodels.
+Custom submodels could produce different outputs in `sim_new` (see [Supporting Information 1]() for examples).
+For default options, there are some list elements that might be especially useful.
+All of these elements can be edited *within the above loop* to dynamically adjust simulations.
+
+- `sim_new$resource_array`: A table holding all information on resources.
+Rows correspond to a unique resource, and columns correspond to resource properties: (1) ID, (2-4) resource types (not currently in use), (5) x-location, (6) y-location, (7) movement parameter, (8) time, (9) density independent mortality parameter (`remove_pr`), (10) reproduction parameter (`lambda`), (11) offspring number, (12) age, (13-14) observation columns, (15) consumption rate (`res_consume`), and (16-20) recorded experiences of user actions.
+- `sim_new$AGENTS`: A table holding basic information on agents (manager and users).
+Rows correspond to a unique agent, and columns correspond to agent properties: (1) ID, (2) type (0 for the manager, 1 for users), (3-4) additional type options not currently in use, (5-6), x and y locations (usually ignored), (7) movement parameter (usually ignored), (8) time, (9) agent's viewing ability in cells (`agent_view`), (10) error parameter, (11-12) values for holding marks and tallies of resources, (13-15) values for holding observations, (16) yield from landscape cells, (17) budget (`manager_budget` and `user_budget`).
+- `sim_new$observation_vector`: Estimate of total resource number from the observation model (`observation_array` also holds this information in a different way depending on `observe_type`)
+- `sim_new$LAND`: The landscape on which interactions occur, which is stored as a 3D array with `land_dim_1` rows, `land_dim_2` columns, and 3 layers.
+Layer 1 (`sim_new$LAND[,,1]`) is not used, but could be used to store values that affect resources and agents.
+Layer 2 (`sim_new$LAND[,,2]`) stores crop yield from a cell, and layer 3 (`sim_new$LAND[,,3]`) stores the owner of the cell (value corresponds to the agent's ID).
+- `sim_new$manage_vector`: The cost of each action as set by the manager. For even more fine-tuning, individual costs for actions can be set for each user in `sim_new$manager_array`.
+- `sim_new$user_vector`: The total number of actions performed by each user. A more detailed breakdown of actions by individual users is held in `sim_new$user_array`.
+
+We now show how to adjust the landscape to manually adjust land ownership in `gmse_apply`.
+
+Customising the GMSE landscape
+================================================================================
+
+By default, all farmers in GMSE are allocated the same number of landscape cells, which are simply placed in order of the farmer's ID.
+Public landscape is produced by placing landscape cells that are technically owned by the manager, and therefore have landscape cell values of 0.
+The image below shows this landscape for the eight farmers from `sim_old`
+
+```{r}
+image(x = sim_old$LAND[,,3], colors = topo.colors(9));
+```

### 2018-04-24 notebook/ms/SI4.Rmd
New function to demonstrate

@@
 Here we demonstrate how simulations in GMSE can be more fine-tuned to specific empirical situations through the use of `gmse_apply`.
-To do this, we use the same scenario described in [Supporting Information 3](); we first recreate the basic scenario run in `gmse` using `gmse_apply`, and then build in additional modelling details including (1) custom placement of user land, (2) parameterisation of individual user budgets, and (3) density-dependent movement of resources.
+To do this, we use the same scenario described in [Supporting Information 3](); we first recreate the basic scenario run in `gmse` using `gmse_apply`, and then build in additional modelling details including (1) [custom placement of user land](#land), (2) [parameterisation of individual user budgets](#budgets), and (3) [density-dependent movement of resources](#movement).
 We emphasise that these simulations are provided only to demonstrate the use of GMSE, and specifically to show the flexibility of the `gmse_apply` function, not to accurately recreate the dynamics of a specific system or make management recommendations.
@@
 
-Additionally, we continue to use the following values for consistency, except in the case of `stakeholders`, where we reduce the number of farmers to `stakehoders = 8` and increase `user_budget` to 100000.
+Additionally, we continue to use the following values for consistency, except in the case of `stakeholders`, where we reduce the number of farmers to `stakeholders = 8` and increase `user_budget` to 100000.
 This is done to for two reasons.
@@
 
-Customising the GMSE landscape
+1. <a name="land">Custom placement of user land</a>
 ================================================================================
@@
 Public landscape is produced by placing landscape cells that are technically owned by the manager, and therefore have landscape cell values of 0.
-The image below shows this landscape for the eight farmers from `sim_old`
+The image below shows this landscape for the eight farmers from `sim_old`.
 
 ```{r}
-image(x = sim_old$LAND[,,3], colors = topo.colors(9));
+image(x = sim_old$LAND[,,3], col = topo.colors(9), xaxt = "n", yaxt = "n");
 ```
+
+We can change the ownership of cells by manipulating `sim_old$LAND[,,3]`. First we initialise a new `sim_old` below.
+
+```{r, echo = TRUE, eval = TRUE}
+sim_old  <- gmse_apply(get_res = "Full", remove_pr = 0.122, lambda = 0.275,
+                       res_death_K = 93870, RESOURCE_ini = 35000, 
+                       manage_target = 70000, res_death_type = 3,
+                       manager_budget = 10000, user_budget = 100000,
+                       public_land = 0.4, stakeholders = 8, 
+                       land_ownership = TRUE, res_consume = 0.02, 
+                       observe_type = 3, agent_view = 1);
+```
+
+Because we have not specified landscape dimensions in the above, the landscape reverts to the default size of 100 by 100 cells.
+We can then manually assign landscape cells to the eight farmers, whose IDs range from 2-9 (ID value 1 is the manager).
+Below we do this to make eight different sized farms.
+
+```{r}
+sim_old$LAND[1:20,   1:20,  3] <- 2; 
+sim_old$LAND[1:20,  21:40,  3] <- 3;
+sim_old$LAND[1:20,  41:60,  3] <- 4;
+sim_old$LAND[1:20,  61:80,  3] <- 5;
+sim_old$LAND[1:20,  81:100, 3] <- 6;
+sim_old$LAND[21:40,  1:50,  3] <- 7;
+sim_old$LAND[21:40, 51:100, 3] <- 8;
+sim_old$LAND[41:60,  1:100, 3] <- 9;
+sim_old$LAND[61:100, 1:100, 3] <- 1; # Public land
+image(x = sim_old$LAND[,,3], col = topo.colors(9), xaxt = "n", yaxt = "n");
+```
+
+The above image shows the modified landscape stored in `sim_old`, which can now be incorporated into simulations using `gmse_apply`.
+We can think of all the plots on the left side of the landscape as farms of various sizes, while the blue area of the landscape on the right is public land.
+
+2. <a name="budgets">Parameterisation of individual user budgets</a>
+================================================================================
+
+Perhaps we want to assume that farmers have different budgets, which are perhaps correlated in some way to the number of landscape cells that they own.
+Custom user budgets can be set by manipulating `sim_old$AGENTS`, the last column of which holds the budget for each user.
+Agent IDs (as stored on the landscape above) correspond to rows of `sim_old$AGENTS`, so individual budgets can be directly input as desired.
+We can do this manually (e.g., `sim_old$AGENTS[2, 17] <- 4000`), or, alternatively, if farmer budget positively correlates to landscape owned, we can use a loop to input values as below.
+
+```{r}
+for(ID in 2:9){
+    cells_owned            <- sum(sim_old$LAND[,,3] == ID);
+    sim_old$AGENTS[ID, 17] <- 100 * cells_owned;
+}
+```
+
+The number of cells owned by each farmer is therefore listed in the table below.
+
+```{r, echo = FALSE}
+budget_table <- rbind(sim_old$AGENTS[,1], sim_old$AGENTS[,17]);
+row.names(budget_table) <- c("ID", "Budget");
+kable(budget_table, caption = "Custom created budgets for manager and users");
+```
+
+As with `sim_old$LAND` values, changes to `sim_old$AGENTS` will be retained in simulations looped through `gmse_apply`.
+
+3. <a name="movement">Density-dependent movement of resources</a>
+================================================================================
+
+Lastly, we consider a more nuanced change to simulations, in which the rules for movement of resources are modified to account for density-dependence.
+Assume that geese tend to avoid aggregating, such that if a goose is located on the same cell as too many other geese, then it will move with a high probability at the end of a time step.
+Programming this movement rule can be accomplished by creating a new function to apply to the resource data array `sim_old$resource_array`.
+Below, a custom function is defined that causes a goose to move up to 5 cells in any direction if it finds itself on a cell with more than 10 other geese.
+As with default GMSE simulations, movement is based on a torus landscape.
+
+```{r}
+avoid_aggregation <- function(goose_table, land_dim_1 = 100, land_dim_2 = 100){
+    goose_number <- dim(goose_table)[1]    # How many geese are there?
+    for(goose in 1:goose_number){          # Loop through all rows of geese
+        x_loc  <- goose_table[goose, 5];
+        y_loc  <- goose_table[goose, 6];
+        shared <- sum(goose_table[,5] == x_loc & goose_table[,6] == y_loc);
+        if(shared > 10){
+            new_x <- x_loc <- sample(x = -5:5, size = 1);
+            new_y <- y_loc <- sample(x = -5:5, size = 1);
+            if(new_x < 0){ # The 'if' statements below apply the torus
+                new_x <- land_dim_1 + new_x;
+            }
+            if(new_x >= land_dim_1){
+                new_x <- new_x - land_dim_1;
+            }
+            if(new_y < 0){
+                new_y <- land_dim_2 + new_x;
+            }
+            if(new_y >= land_dim_2){
+                new_y <- new_y - land_dim_2;
+            }
+            goose_table[goose, 5] <- new_x;
+            goose_table[goose, 6] <- new_y;
+        }
+    }
+    return(goose_table);
+}
+```

### 2018-04-26 notebook/ms/SI4.Rmd
Small changes

@@
 | Parameter         | Value    | Description                                         |
-|-------------------|----------|--------------------------------------------------|
+|-------------------|----------|-----------------------------------------------------|
 | `remove_pr`       | 0.122    | Goose density-independent mortality probability     |
@@
 | `manage_target`   | 70000    | Manager's target goose abundance                    |
-| `res_death_type`  | 3        | Mortality (density and density-indepent sources) |
+| `res_death_type`  | 3        | Mortality (density and density-independent sources) |
 
@@
 
-Additionally, we continue to use the following values for consistency, except in the case of `stakeholders`, where we reduce the number of farmers to `stakeholders = 8` and increase `user_budget` to 100000.
+Additionally, we continue to use the following values for consistency, except in the case of `stakeholders`, where we reduce the number of farmers to `stakeholders = 8`.
 This is done to for two reasons.
@@
 | `manager_budget`  | 10000    | Manager's budget for setting policy options      |
-| `user_budget`     | 100000   | Users' budgets for actions                       |
+| `user_budget`     | 10000    | Users' budgets for actions                       |
 | `public_land`     | 0.4      | Proportion of the landscape that is public       |
@@
 We now recreate the simulations in [Supporting Information 3](), which were run using the `gmse` function, in `gmse_apply`.
-Doing so requires us to first initialise simulations using one run of `gmse_apply`, then loop through multiple time steps that again call `gmse_apply` and saving the results of interest.
-Following instructions in [Supporting Information 1](), we can call the initialising simulation `sim_old`, and use the code below to read in the relevant parameter values.
+Doing so requires us to first initialise simulations using one call of `gmse_apply`, then loop through multiple time steps that again call `gmse_apply`; results of interest are recorded in a data frame (`sim_sum_1`).
+Following the protocol introduced in [Supporting Information 2](), we can call the initialising simulation `sim_old`, and use the code below to read in the relevant parameter values.
 
@@
                        manager_budget = 10000, user_budget = 100000,
-                       public_land = 0.4, stakeholders = 8, 
-                       land_ownership = TRUE, res_consume = 0.02, 
-                       observe_type = 3, agent_view = 1);
+                       public_land = 0.4, stakeholders = 8, res_consume = 0.02, 
+                       res_birth_K = 200000, land_ownership = TRUE,
+                       observe_type = 3, agent_view = 1, converge_crit = 0.01,
+                       ga_mingen = 200);
 ```
@@
 
-The above output from `sim_sum_1` shows the table that holds the information we were interested in pulling out of our simulation results.
+The above output from `sim_sum_1` shows the data frame that holds the information we were interested in pulling out of our simulation results.
 All of this information was available under the list element `sim_new$basic_output`, but other list elements of `sim_new` might also be useful to record.
-It is important to also remember that this example of `gmse_apply` is using the default resource, observation, manager, and user submodels.
-Custom submodels could produce different outputs in `sim_new` (see [Supporting Information 1]() for examples).
-For default options, there are some list elements that might be especially useful.
-All of these elements can be edited *within the above loop* to dynamically adjust simulations.
+It is important to remember that this example of `gmse_apply` is using the default resource, observation, manager, and user submodels.
+Custom submodels could produce different outputs in `sim_new` (see [Supporting Information 2]() for examples).
+For default submodels, there are some list elements that might be especially useful.
+These elements can potentially be edited *within the above loop* to dynamically adjust simulations.
 
 - `sim_new$resource_array`: A table holding all information on resources.
-Rows correspond to a unique resource, and columns correspond to resource properties: (1) ID, (2-4) resource types (not currently in use), (5) x-location, (6) y-location, (7) movement parameter, (8) time, (9) density independent mortality parameter (`remove_pr`), (10) reproduction parameter (`lambda`), (11) offspring number, (12) age, (13-14) observation columns, (15) consumption rate (`res_consume`), and (16-20) recorded experiences of user actions.
+Rows correspond to discrete resources, and columns correspond to resource properties: (1) ID, (2-4) types (not currently in use), (5) x-location, (6) y-location, (7) movement parameter, (8) time, (9) density independent mortality parameter (`remove_pr`), (10) reproduction parameter (`lambda`), (11) offspring number, (12) age, (13-14) observation columns, (15) consumption rate (`res_consume`), and (16-20) recorded experiences of user actions (e.g., was the resource culled or scared?).
 - `sim_new$AGENTS`: A table holding basic information on agents (manager and users).
@@
 - `sim_new$LAND`: The landscape on which interactions occur, which is stored as a 3D array with `land_dim_1` rows, `land_dim_2` columns, and 3 layers.
-Layer 1 (`sim_new$LAND[,,1]`) is not used, but could be used to store values that affect resources and agents.
+Layer 1 (`sim_new$LAND[,,1]`) is not currently used in default submodels, but could be used to store values that affect resources and agents.
 Layer 2 (`sim_new$LAND[,,2]`) stores crop yield from a cell, and layer 3 (`sim_new$LAND[,,3]`) stores the owner of the cell (value corresponds to the agent's ID).
-- `sim_new$manage_vector`: The cost of each action as set by the manager. For even more fine-tuning, individual costs for actions can be set for each user in `sim_new$manager_array`.
+- `sim_new$manage_vector`: The cost of each action as set by the manager. For even more fine-tuning, individual costs for the actions of each agent can be set for each user in `sim_new$manager_array`.
 - `sim_new$user_vector`: The total number of actions performed by each user. A more detailed breakdown of actions by individual users is held in `sim_new$user_array`.
 
-We now show how to adjust the landscape to manually adjust land ownership in `gmse_apply`.
+Next, we show how to adjust the landscape to manually set land ownership in `gmse_apply`.
 
@@
 By default, all farmers in GMSE are allocated the same number of landscape cells, which are simply placed in order of the farmer's ID.
-Public landscape is produced by placing landscape cells that are technically owned by the manager, and therefore have landscape cell values of 0.
+Public land is produced by placing landscape cells that are technically owned by the manager, and therefore have landscape cell values of 1.
 The image below shows this landscape for the eight farmers from `sim_old`.
@@
 sim_old  <- gmse_apply(get_res = "Full", remove_pr = 0.122, lambda = 0.275,
-                       res_death_K = 93870, RESOURCE_ini = 35000, 
+                       res_d4eath_K = 93870, RESOURCE_ini = 35000, 
                        manage_target = 70000, res_death_type = 3,
-                       manager_budget = 10000, user_budget = 100000,
-                       public_land = 0.4, stakeholders = 8, 
-                       land_ownership = TRUE, res_consume = 0.02, 
-                       observe_type = 3, agent_view = 1);
+                       manager_budget = 10000, user_budget = 10000,
+                       public_land = 0.4, stakeholders = 8, res_consume = 0.02, 
+                       res_birth_K = 200000, land_ownership = TRUE,
+                       observe_type = 3, agent_view = 1, converge_crit = 0.01,
+                       ga_mingen = 200);
 ```
@@
 
-Perhaps we want to assume that farmers have different budgets, which are perhaps correlated in some way to the number of landscape cells that they own.
-Custom user budgets can be set by manipulating `sim_old$AGENTS`, the last column of which holds the budget for each user.
+Perhaps we want to assume that farmers have different budgets, which are correlated in some way to the number of landscape cells that they own.
+Custom user budgets can be set by manipulating `sim_old$AGENTS`, the last column of which (column 17) holds the budget for each user.
 Agent IDs (as stored on the landscape above) correspond to rows of `sim_old$AGENTS`, so individual budgets can be directly input as desired.
@@
 ```{r, echo = FALSE}
+library(pander);
 budget_table <- rbind(sim_old$AGENTS[,1], sim_old$AGENTS[,17]);
 row.names(budget_table) <- c("ID", "Budget");
-kable(budget_table, caption = "Custom created budgets for manager and users");
+pander(budget_table);
 ```
@@
 Lastly, we consider a more nuanced change to simulations, in which the rules for movement of resources are modified to account for density-dependence.
-Assume that geese tend to avoid aggregating, such that if a goose is located on the same cell as too many other geese, then it will move with a high probability at the end of a time step.
+Assume that geese tend to avoid aggregating, such that if a goose is located on the same cell as too many other geese, then it will move at the start of a time step.
 Programming this movement rule can be accomplished by creating a new function to apply to the resource data array `sim_old$resource_array`.
@@
         if(shared > 10){
-            new_x <- x_loc <- sample(x = -5:5, size = 1);
-            new_y <- y_loc <- sample(x = -5:5, size = 1);
+            new_x <- x_loc + sample(x = -5:5, size = 1);
+            new_y <- y_loc + sample(x = -5:5, size = 1);
             if(new_x < 0){ # The 'if' statements below apply the torus
@@
 ```
+
+With the above function written, we can apply the new movement rule along with our [custom farm placement](#land) and [custom farmer budgets](#budgets) to the simulation of goose population dynamics.
+
+Simulation with custom farms, budgets, and goose movement
+================================================================================
+
+Below shows an example of `gmse_apply` with custom landscapes, farmer budgets, and density-dependent goose movement rules.
+
+```{r, echo = TRUE, eval = TRUE}
+# First initialise a simulation
+sim_old  <- gmse_apply(get_res = "Full", remove_pr = 0.122, lambda = 0.275,
+                       res_death_K = 93870, RESOURCE_ini = 35000, 
+                       manage_target = 70000, res_death_type = 3,
+                       manager_budget = 10000, user_budget = 10000,
+                       public_land = 0.4, stakeholders = 8, res_consume = 0.02, 
+                       res_birth_K = 200000, land_ownership = TRUE,
+                       observe_type = 3, agent_view = 1, converge_crit = 0.01,
+                       ga_mingen = 200, res_move_type = 0);
+# By setting `res_move_type = 0`, no resource movement will occur in gmse_apply
+# Adjust the landscape ownership below
+sim_old$LAND[1:20,   1:20,  3] <- 2; 
+sim_old$LAND[1:20,  21:40,  3] <- 3;
+sim_old$LAND[1:20,  41:60,  3] <- 4;
+sim_old$LAND[1:20,  61:80,  3] <- 5;
+sim_old$LAND[1:20,  81:100, 3] <- 6;
+sim_old$LAND[21:40,  1:50,  3] <- 7;
+sim_old$LAND[21:40, 51:100, 3] <- 8;
+sim_old$LAND[41:60,  1:100, 3] <- 9;
+sim_old$LAND[61:100, 1:100, 3] <- 1;
+# Change the budgets of each farmer based on the land they own
+for(ID in 2:9){
+    cells_owned            <- sum(sim_old$LAND[,,3] == ID);
+    sim_old$AGENTS[ID, 17] <- 10 * cells_owned;
+}
+# Begin simulating time steps for the system
+sim_sum_2 <- matrix(data = NA, nrow = 30, ncol = 5);
+for(time_step in 1:30){
+    # Apply the new movement rules at the beginning of the loop
+    sim_old$resource_array  <- avoid_aggregation(sim_old$resource_array);
+    # Next, move on to simulate (old_list remembers that res_move_type = 0)
+    sim_new                 <- gmse_apply(get_res = "Full", old_list = sim_old);
+    sim_sum_2[time_step, 1] <- time_step;
+    sim_sum_2[time_step, 2] <- sim_new$basic_output$resource_results[1];
+    sim_sum_2[time_step, 3] <- sim_new$basic_output$observation_results[1];
+    sim_sum_2[time_step, 4] <- sim_new$basic_output$manager_results[3];
+    sim_sum_2[time_step, 5] <- sum(sim_new$basic_output$user_results[,3]);
+    sim_old                 <- sim_new;
+}
+colnames(sim_sum_2) <- c("Time", "Pop_size", "Pop_est", "Cull_cost", 
+                         "Cull_count");
+print(sim_sum_2);
+```
+
+Conclusions
+================================================================================
+
+In this example, we showed how the built-in resource, observation, manager, and user sub-models can be customised by manipulating the data within the data structures that they use.
+The goal was to show how software users can work with these existing sub-models and data structures to customise GMSE simulations.
+Software users seeking even greater flexibility (e.g., replacing an entire built-in submodel with a custom submodel) should refer to the [Supporting Information 2](https://bradduthie.github.io/gmse_SI2.pdf) that introduces `gmse_apply` more generally.
+Future versions of GMSE are likely to expand on the built-in options available for simulation; requests for such expansions, or contributions, can be submitted to [GitHub](https://github.com/bradduthie/gmse/issues).

### 2018-04-26 R/gmse.R
Adjust default K

@@ -78,7 +78,7 @@ gmse <- function( time_max       = 100,   # Max number of time steps in sim
                   lambda         = 0.30,  # Resource growth rate
                   agent_view     = 10,    # Number cells agent view around them
                   agent_move     = 50,    # Number cells agent can move
-                  res_birth_K    = 10000, # Carrying capacity applied to birth
+                  res_birth_K    = 100000,# Carrying capacity applied to birth
                   res_death_K    = 2000,  # Carrying capacity applied to death
                   edge_effect    = 1,     # What type of edge on the landscape
                   res_move_type  = 1,     # What type of movement for resources

### 2018-04-27 notebook/ms/SI5.Rmd
Initialise the fisheries SI example

---
title: 'GMSE: an R package for generalised management strategy evaluation'
author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
  Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;,
  Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes: null
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: gmse.bib
subtitle: Supporting Information 5
biblio-style: apalike
---

install.packages(c("ggplot2"));
install.packages(c("FLCore"), repos="http://flr-project.org/R");
install.packages(c("ggplotFL"), repos="http://flr-project.org/R");

library(FLCore);
library(ggplotFL);

data(ple4);
p4sr <-as.FLSR(ple4[-1])

model(p4sr) <- ricker()
p4sr2       <- fmle(p4sr)

a <- params(p4sr2)[[1]];
b <- params(p4sr2)[[2]];

rec(p4sr2);
ssb(p4sr2);

rec_year    <- as.numeric(attributes(rec(p4sr2))$dimnames$year);
ssb_year    <- as.numeric(attributes(ssb(p4sr2))$dimnames$year);
recruitment <- NULL;
for(year in 1:length(rec_year)){
    recruitment[year] <- rec(p4sr)[[year]];
}
spawning_stock_biomass <- NULL;
for(year in 1:length(ssb_year)){
    spawning_stock_biomass[year] <- ssb(p4sr)[[year]];
}

rec <- a * spawning_stock_biomass * exp(-b * spawning_stock_biomass);

rec(p4sr)[[51]] <- 4300

### 2018-04-27 notebook/ms/SI5.Rmd
Some more progress

@@
 
+Integration and simulation with fisheries
+================================================================================
+
+Early development of management strategy evaluation (MSE) models originated in fisheries [@Polacheck1999; @Smith1999; @Sainsbury2000].
+Consequently, fisheries-focused software for MSE has been extensively developed, including R libraries that focus on the management of species of exceptional interest, such as the Atlantic Bluefin Tuna (*Thunnus thynnus*) [[ABFTMSE](https://github.com/ICCAT/abft-mse/tree/master/R_package/ABTMSE); @Carruthers2018; @Carruthers2018a], and Indian Ocean Bigeye (*T.
+obesus*) and Yellowfin (*T.
+albacares*) Tuna [[MSE-IO-BET-YFT](https://github.com/pjumppanen/MSE-IO-BET-YFT); @Kolody2016].
+The largest of all such libraries is the Fisheries Library in R ([FLR](http://www.flr-project.org/)), which includes an extensive collection of tools targeted for fisheries science.
+The FLR library has been used in over a hundred [publications](http://www.flr-project.org/#publications) [recent publications include @Jardim2018; @Mackinson2018; @Utizi2018], and includes [an MSE framework](http://www.flr-project.org/doc/An_introduction_to_MSE_using_FLR.html) for evaluating different harvest control rules.
+
+As part of the [ConFooBio](https://sti-cs.org/confoobio/) project, a central focus of GMSE is on simulating the management of populations of conservation interest, with a particular emphasis on understanding conservation conflict; further development of GMSE is expected to continue with this as a priority, further building upon the decision-making algorithms of managers and users to better understand how conflict arises and can potentially be resolved.
+Hence, GMSE is not intended as a substitute for packages such as [FLR](http://www.flr-project.org/), but the integration of these packages with GMSE could make use of GSME's current and future simulation capabilities, and particularly the [genetic algorithm]().
+Such integration might be possible using the `gmse_apply` function, which allows for [custom defined submodels]() to be used within the GMSE framework, and with default GMSE submodels.
+Hence, GMSE might be especially useful for modelling the management of fisheries under conditions of increasing harvesting demands and stakeholder conflict.
+We do not attempt such an ambitious project here, but instead show how such a project could be developed through integration of [FLR](http://www.flr-project.org/) and `gmse_apply`.
+
+Here we follow a [Modelling Stock-Recruitment with FLSR](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html) example, then integrate this example with `gmse_apply` to explore the behaviour of simulated fishers who are goal-driven to maximise their own harvest.
+We emphasise that this example is provided only as demonstration of how GMSE can potentially be integrated with already developed fisheries models, and is not intended to make recommendations for management in any population.
+
+Integrating with the Fisheries Library in R ([FLR](http://www.flr-project.org/))
+================================================================================
+
+The FLR toolset includes [a series of pacakges](http://www.flr-project.org/#packages), with several [tutorials](http://www.flr-project.org/doc/index.html) for using them.
+For simplicity, we focus here on [a model of stock recruitment](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html) to be used as the population model in `gmse_apply`.
+This population model will use sample data and one of the many available stock-recruitment models available in FLR, and a custom function will be written to return a single value for stock recruitment.
+Currently, `gmse_apply` requires that submodels return subfunction results either as scalar values or data frames that are structured in the same way as GMSE submodels.
+But interpretation of scalar values is left up to the user (e.g., population model results could be interpreted as abundance or biomass; manager policy could be interpreted as cost of harvesting or as total allowable catch).
+For simplicity, the observation (i.e., estimation) model will simply be the stock reported from the population model with error, and the manager model will be a total allowable catch calculated from the stock-recruitment relationship that accounts for the number of fishers in the system.
+The user model, however, will employ the full power of the default GMSE user function to simulate user actions.
+We first show how a custom function can be made that applies the FLR toolset to a population model.
+
+Modelling stock-recruitment for the population model
+================================================================================
+
+<!---
+Just re-create the herring 'rec' and 'ssb' data, then read it into an FLquant data type. Then the whole analysis can be done as before, but with access to the data in GMSE. Add the model in too.
+--->
+
+Here we closely follow [a tutorial from the FLR project](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html).
+To build the stock-recruitment model, the following packages are needed.
+
+```{r, eval = FALSE}
 install.packages(c("ggplot2"));
@@
 install.packages(c("ggplotFL"), repos="http://flr-project.org/R");
+```
 
+To start, we need to read in the `FLCore`, `FLFleet` and `ggplotFL` libraries.
+
+```{r}
 library(FLCore);
+library(FLFleet)
 library(ggplotFL);
+```
 
-data(ple4);
-p4sr <-as.FLSR(ple4[-1])
+The data below are derived from the [FLR tutorial](http://www.flr-project.org/doc/Loading_your_data_into_FLR.html), and the code below downloads the example data files into a temporary folder.
 
-model(p4sr) <- ricker()
-p4sr2       <- fmle(p4sr)
+```{r}
+temp_dir <- tempdir()
+download.file("http://flr-project.org/doc/src/loading_data.zip", file.path(dir, "loading_data.zip"))
+unzip(file.path(temp_dir, "loading_data.zip"), exdir = temp_dir);
+```
 
-a <- params(p4sr2)[[1]];
-b <- params(p4sr2)[[2]];
+We can read in the example file below and convert it into a matrix to be used in an [FLQuant](http://www.flr-project.org/FLCore/reference/FLQuant.html) object. Rows in this matrix indicate ages.
+
+```{r}
+fish_data <- read.csv(file.path(temp_dir, "catch_numbers.csv"), row = 1);
+fish_data <- as.matrix(fish_data);
+print(fish_data[,1:8]);
+```
+
+For simplicity, we merge these ages down into a single vector.
+
+```{r}
+fish_data <- apply(X = fish_data, MARGIN = 2, FUN = sum);
+```
+
+We can now create a new [FLQuant](http://www.flr-project.org/FLCore/reference/FLQuant.html) object with the `fish_data` vector.
+This process is important; down below, it will be repeated in a function that gets used within the GMSE population model, as GMSE will need to incorporate newly simulated data in each time step.
+
+```{r}
+fish_data_flq <- FLQuant(fish_data, dimnames = list(year = 1957:2011));
+```
+
+The data below include sample results for recruitment and spawning-stock biomass (SSB) of North Sea herring.
+
+```{r}
+data(nsher);
+```
+
+These data are structured using `FLStock`, an [S4](https://stat.ethz.ch/R-manual/R-devel/library/methods/html/Classes_Details.html) class of R objects.
+The code below creates an `FLSR` object, needed to create a stock-recruitment model.
+
+```{r}
+summary(nsher);
+```
+
+Notice that under `Model`, a Ricker model of stock-recruitment has already been fitted (many stock-recruitment models are available in FLCore; see the [FLSR tutorial](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html) for more about fitting different models) and parameterised with Maximum Likelihood Estimation (MLE).
+The code below shows how to apply the Beverton-Holt model model as an alternative, and the  `fmle` function parameterises the Beverton-Holt model model using the data in `nsher` and MLE.
+
+```{r}
+model(nsher) <- bevholt();
+nsher        <- fmle(nsher);
+summary(nsher);
+```
+
+We will use these tools to update the population model in GMSE.
+
+a <- params(nsher)[[1]];
+b <- params(nsher)[[2]];
 

### 2018-04-30 notebook/ms/SI3.Rmd
Major add of vignettes and relevant data

@@
 ```{r, echo = FALSE, eval = TRUE}
-load("data/SI3-cull_sim.rda"); # Avoids long compiling time
+load("../data/SI3-cull_sim.rda"); # Avoids long compiling time
 sim_1 <- sim;
@@
 ```{r, echo = FALSE, eval = TRUE}
-load("data/SI3-scare_sim.rda"); # Avoids long compiling time
+load("../data/SI3-scare_sim.rda"); # Avoids long compiling time
 sim_2 <- sim;

### 2018-04-30 notebook/ms/SI5.Rmd
Major add of vignettes and relevant data

@@
 
-<!---
-Just re-create the herring 'rec' and 'ssb' data, then read it into an FLquant data type. Then the whole analysis can be done as before, but with access to the data in GMSE. Add the model in too.
---->
-
 Here we closely follow [a tutorial from the FLR project](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html).
@@
 
+<!---
+Just re-create the herring 'rec' and 'ssb' data, then read it into an FLquant data type. Then the whole analysis can be done as before, but with access to the data in GMSE. Add the model in too.
+--->
+
 The data below are derived from the [FLR tutorial](http://www.flr-project.org/doc/Loading_your_data_into_FLR.html), and the code below downloads the example data files into a temporary folder.
@@
 ```{r}
-temp_dir <- tempdir()
-download.file("http://flr-project.org/doc/src/loading_data.zip", file.path(dir, "loading_data.zip"))
-unzip(file.path(temp_dir, "loading_data.zip"), exdir = temp_dir);
+#temp_dir <- tempdir()
+#download.file("http://flr-project.org/doc/src/loading_data.zip", file.path(dir, "loading_data.zip"))
+#unzip(file.path(temp_dir, "loading_data.zip"), exdir = temp_dir);
 ```
@@
 ```{r}
-fish_data <- read.csv(file.path(temp_dir, "catch_numbers.csv"), row = 1);
-fish_data <- as.matrix(fish_data);
-print(fish_data[,1:8]);
+#fish_data <- read.csv(file.path(temp_dir, "catch_numbers.csv"), row = 1);
+#fish_data <- as.matrix(fish_data);
+#print(fish_data[,1:8]);
 ```

### 2018-04-30 notebook/ms/ms.Rmd
Minor edit
Major add of vignettes and relevant data

@@
 
-The R package 'GMSE' is a flexible modelling tool to simulate all key aspects of natural resource management.
+The R package 'GMSE' is a flexible modelling tool to simulate key aspects of natural resource management.
 GMSE offers a range of parameters to simulate resource dynamics and management policy options, and includes genetic algorithms to dynamically model stakeholder (manager and user) decision-making.

### 2018-04-30 R/SI4_data.R

@@ -1,0 +1,44 @@
+#' @name sim_init_basic
+#' @title R data needed for quick load basic sim in the S4 vignette
+#' @description This object contains SI4 vignette simulation output
+#' @docType data
+#' @usage sim_init_basic
+#' @format a set of \code{gmse_apply} output examples
+#' @author Brad Duthie
+NULL
+
+#' @name sim_old_1_LAND
+#' @title R data needed to show vignette example of LAND 1
+#' @description This object contains simulation output for SI4 for LAND 1
+#' @docType data
+#' @usage sim_old_1_LAND
+#' @format a set of \code{gmse_apply} output examples
+#' @author Brad Duthie
+NULL
+
+#' @name sim_old_2_LAND
+#' @title R data needed to show vignette example of LAND 2
+#' @description This object contains simulation output for SI4 for LAND 1
+#' @docType data
+#' @usage sim_old_2_LAND
+#' @format a set of \code{gmse_apply} output examples
+#' @author Brad Duthie
+NULL
+
+#' @name sim_sum_1
+#' @title R data needed to show a summary of example simulation 1
+#' @description This object contains simulation the simulation 1 example in S4
+#' @docType data
+#' @usage sim_sum_1
+#' @format a set of \code{gmse_apply} output examples
+#' @author Brad Duthie
+NULL
+
+#' @name sim_sum_2
+#' @title R data needed to show a summary of example simulation 2
+#' @description his object contains simulation the simulation 2 example in S4
+#' @docType data
+#' @usage sim_sum_2
+#' @format a set of \code{gmse_apply} output examples
+#' @author Brad Duthie
+NULL

### 2018-04-30 R/gmse.R

@@ -14,7 +14,7 @@
 #'@param lambda This value is the baseline population growth rate of resources. Each resource in the simulation produces Poisson(lambda) offspring in one time step within the resource model. The value of lambda might be increased or decreased by user actions, and juvenile survival can potentially be decreased by a carrying capacity placed on birth. The default value is 0.3, meaning that the average resource produces one offspring every three time steps.
 #'@param agent_view This value determines how far agents (managers and stakeholders) can see on the landscape. At the moment, this affects only the sampling ability of managers in the observation model for density-based and transect-based estimates of resource abundance. In these types of estimates, when managers have a higher agent_view, they are capable of observing a larger area of landscape and therefore of getting a larger (in the case of density-based estimation) or more efficient (in the case of transect-based estimation) sample of resources from which to estimate total resource abundance. The default value of agent_view is 10, so agents can see 10 cells away from their current cell in any direction.
 #'@param agent_move This value determines how far agents can move. At the moment, this does not affect much in the simulation because agent movement does not affect agent actions (interactions with resources can be limited to stakeholder's owned land, but do not currently depend on where an agent is on the landscape -- effectively assuming that agents are mobile enough to do what they want to do to resources). The one exception is for density-based estimation, which can be biased by low values of agent_move by causing the manager to sample the same (or nearby) landscape cells to estimate total resource abundance; if resources are spatially autocorrelated, then managers might over or under-estimate total abundance. Therefore, as a default, this value is set to 50 so that managers can move to any cell on a (torus) landscape in a time step, removing any bias for density sampling.
-#'@param res_birth_K This value is the carrying capacity on new resources added per time step (e.g., birth). If more offspring are born in a time step than res_birth_K, then offspring are randomly removed from the population until offspring born equals res_birth_K. By default, carrying capacity is effectively applied to death instead of birth, so the default value of res_birth_K is set to 10000 (and hence not enacted because the number of births is never this high).
+#'@param res_birth_K This value is the carrying capacity on new resources added per time step (e.g., birth). If more offspring are born in a time step than res_birth_K, then offspring are randomly removed from the population until offspring born equals res_birth_K. By default, carrying capacity is effectively applied to death instead of birth, so the default value of res_birth_K is set to 100000 (and hence not enacted because the number of births is never this high).
[…]

### 2018-04-30 notebook/ms/SI3.Rmd
Add correct links

--- 2018-04-24 notebook/ms/SI3.Rmd
@@
 Novice R users may prefer to run all of the simulations below using the browser-based GMSE GUI by calling `gmse_gui()` from the R command line.
-Alternatively, experienced R users may prefer to simulate by [looping time steps](https://bradduthie.github.io/gmse_SI2.pdf#page=5) through `gmse_apply`, which allows more flexibility for incorporating custom sub-models and dynamically adjusting parameter values.
+Alternatively, experienced R users may prefer to simulate by [looping time steps](SI2.pdf#page=5) through `gmse_apply`, which allows more flexibility for incorporating custom sub-models and dynamically adjusting parameter values.
 Where available, we use estimated demographic parameter values from @Johnson2018 and @AEWA2016.
@@
 load("data/SI3-cull_sim.rda"); # Avoids long compiling time
-sim_1 <- sim;
+sim_1 <- sim_cull;
 ```
@@
 load("data/SI3-scare_sim.rda"); # Avoids long compiling time
-sim_2 <- sim;
+sim_2 <- sim_scare;
 ```

### 2018-04-30 notebook/ms/SI4.Rmd
Add correct links

@@
 Here we demonstrate how simulations in GMSE can be more fine-tuned to specific empirical situations through the use of `gmse_apply`.
-To do this, we use the same scenario described in [Supporting Information 3](); we first recreate the basic scenario run in `gmse` using `gmse_apply`, and then build in additional modelling details including (1) [custom placement of user land](#land), (2) [parameterisation of individual user budgets](#budgets), and (3) [density-dependent movement of resources](#movement).
+To do this, we use the same scenario described in [Supporting Information 3](SI3.pdf); we first recreate the basic scenario run in `gmse` using `gmse_apply`, and then build in additional modelling details including (1) [custom placement of user land](#land), (2) [parameterisation of individual user budgets](#budgets), and (3) [density-dependent movement of resources](#movement).
 We emphasise that these simulations are provided only to demonstrate the use of GMSE, and specifically to show the flexibility of the `gmse_apply` function, not to accurately recreate the dynamics of a specific system or make management recommendations.
@@
 
-We now recreate the simulations in [Supporting Information 3](), which were run using the `gmse` function, in `gmse_apply`.
+We now recreate the simulations in [Supporting Information 3](SI3.pdf), which were run using the `gmse` function, in `gmse_apply`.
 Doing so requires us to first initialise simulations using one call of `gmse_apply`, then loop through multiple time steps that again call `gmse_apply`; results of interest are recorded in a data frame (`sim_sum_1`).
-Following the protocol introduced in [Supporting Information 2](), we can call the initialising simulation `sim_old`, and use the code below to read in the relevant parameter values.
+Following the protocol introduced in [Supporting Information 2](SI2.pdf), we can call the initialising simulation `sim_old`, and use the code below to read in the relevant parameter values.
 
@@
 
-We can then loop over 30 time steps to recreate the simulations from [Supporting Information 3]().
+We can then loop over 30 time steps to recreate the simulations from [Supporting Information 3](SI3.pdf).
 In these simulations, we are specifically interested in the resource and observation outputs, as well as the manager policy and user actions for culling, which we record below in the data frame `sim_sum`.
@@
 It is important to remember that this example of `gmse_apply` is using the default resource, observation, manager, and user submodels.
-Custom submodels could produce different outputs in `sim_new` (see [Supporting Information 2]() for examples).
+Custom submodels could produce different outputs in `sim_new` (see [Supporting Information 2](SI2.pdf) for examples).
 For default submodels, there are some list elements that might be especially useful.
@@
 The goal was to show how software users can work with these existing sub-models and data structures to customise GMSE simulations.
-Software users seeking even greater flexibility (e.g., replacing an entire built-in submodel with a custom submodel) should refer to the [Supporting Information 2](https://bradduthie.github.io/gmse_SI2.pdf) that introduces `gmse_apply` more generally.
+Software users seeking even greater flexibility (e.g., replacing an entire built-in submodel with a custom submodel) should refer to the [Supporting Information 2](SI2.pdf) that introduces `gmse_apply` more generally.
 Future versions of GMSE are likely to expand on the built-in options available for simulation; requests for such expansions, or contributions, can be submitted to [GitHub](https://github.com/bradduthie/gmse/issues).

### 2018-04-30 notebook/ms/SI5.Rmd
Fix the procedure for reading it all in

@@
 
+```{r}
+newFL <- FLSR()
+ssb.n <- read.csv("notebook/ms/data/SI5/nsher_ssb.csv")
+rec.n <- read.csv("notebook/ms/data/SI5/nsher_rec.csv")
+ssb.m <- as.matrix(ssb.n)
+rec.m <- as.matrix(rec.n);
+Frec.m <- FLQuant(ref.m, dimnames=list(age=1, year = 1960:2004))
+Fssb.m <- FLQuant(ssb.m, dimnames=list(age=1, year = 1960:2004))
+Frec.m@units <- "10^3"
+
+newFL@rec <- Frec.m
+newFL@ssb <- Fssb.m
+```
+
 The data below are derived from the [FLR tutorial](http://www.flr-project.org/doc/Loading_your_data_into_FLR.html), and the code below downloads the example data files into a temporary folder.

### 2018-05-01 notebook/ms/SI5.Rmd
Comment out all the old stuff

--- 2018-04-30 notebook/ms/SI5.Rmd
@@
 library(FLCore);
-library(FLFleet)
-library(ggplotFL);
+#library(FLFleet)
+#library(ggplotFL);
+```
+
+For a simplified example in GMSE, we will simulate the process of stock recruitment over multiple time steps using an example stock-recruitment model.
+The stock-recruitment model describes the relationship between stock-recruitment and spawning stock biomass.
+The sample that we will work from is a recreation of the North Sea Herring (`nsher`) dataset available in the `FLCore` package.
+This data set includes recruitment and spawning stock biomass data between 1960 and 2004.
+First, we initialise an empty `FLSR` object and read in the recreated CSV data files.
+
+```{r}
+newFL        <- FLSR(); # Initialises the empty FLSR object
+rec.n        <- read.csv("data/SI5/nsher_rec.csv");
+ssb.n        <- read.csv("data/SI5/nsher_ssb.csv");
+print(rec.n);
+```
+
+The recruitment (`rec.n`) and spawning stock biomass (`ssb.n`) data need to be in the form of a vector, array, matrix to use them with `FLQuant`. We will convert `rec.n` and `ssb.n` into matrices.
+
+```{r}
+rec.m        <- as.matrix(rec.n);
+ssb.m        <- as.matrix(ssb.n);
+```
+
+```{r}
+Frec.m       <- FLQuant(rec.m, dimnames=list(age=1, year = 1960:2004));
+Fssb.m       <- FLQuant(ssb.m, dimnames=list(age=1, year = 1960:2004));
+Frec.m@units <- "10^3";
+Fssb.m@units <- "t*10^3";
+
+newFL@rec    <- Frec.m;
+newFL@ssb    <- Fssb.m;
+newFL@range  <- c(0, 1960, 0, 2004);
+
+model(newFL) <- ricker();
 ```
@@
 <!---
-Just re-create the herring 'rec' and 'ssb' data, then read it into an FLquant data type. Then the whole analysis can be done as before, but with access to the data in GMSE. Add the model in too.
---->
 
@@
 rec(p4sr)[[51]] <- 4300
+
+--->

### 2018-05-01 notebook/ms/SI5.Rmd
Update author list

@@
 author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
-  Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;,
-  Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
+  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
 date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
@@
 Here we closely follow [a tutorial from the FLR project](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html).
-To build the stock-recruitment model, the following packages are needed.
+To build the stock-recruitment model, the `FLCore` package is needed [@Kell2007].
 
 ```{r, eval = FALSE}
-install.packages(c("ggplot2"));
 install.packages(c("FLCore"), repos="http://flr-project.org/R");
-install.packages(c("ggplotFL"), repos="http://flr-project.org/R");
 ```
 
-To start, we need to read in the `FLCore`, `FLFleet` and `ggplotFL` libraries.
+To start, we need to read in the `FLCore` library.
 
@@
 library(FLCore);
-#library(FLFleet)
-#library(ggplotFL);
 ```
@@
 The stock-recruitment model describes the relationship between stock-recruitment and spawning stock biomass.
-The sample that we will work from is a recreation of the North Sea Herring (`nsher`) dataset available in the `FLCore` package.
+The sample that we will work from is a recreation of the North Sea Herring (`nsher`) dataset available in the `FLCore` package [@Kell2007].
 This data set includes recruitment and spawning stock biomass data between 1960 and 2004.
@@
 ssb.n        <- read.csv("data/SI5/nsher_ssb.csv");
-print(rec.n);
 ```
@@
 
+We can then construct two `FLQuant` objects, specifying the relevant years and units.
+
 ```{r}
@@
 Fssb.m@units <- "t*10^3";
+```
 
-newFL@rec    <- Frec.m;
-newFL@ssb    <- Fssb.m;
-newFL@range  <- c(0, 1960, 0, 2004);
+We then place the recruitment and spawning stock biomass data into the `FLSR` object that we created.
 
+```{r}
+rec(newFL)    <- Frec.m;
+ssb(newFL)    <- Fssb.m;
+range(newFL)  <- c(0, 1960, 0, 2004);
+```
+
+The `FLCore` package offers several stock-recruitment models. Here we use a Ricker model of stock recruitment [@Ricker1954], and insert this model into the `FLSR` object below.
+
+```{r}
 model(newFL) <- ricker();
@@
 
-<!---
+Parameters for the Ricker stock-recruitment model can be estimated with maximum likelihood.
 
-The data below are derived from the [FLR tutorial](http://www.flr-project.org/doc/Loading_your_data_into_FLR.html), and the code below downloads the example data files into a temporary folder.
+```{r, echo = FALSE, messages = FALSE, cache = FALSE, comment = FALSE, results = "hide"}
+newFL <- fmle(newFL);
+```
+
+```{r, eval = FALSE}
+newFL <- fmle(newFL);
+```
+
+Diagnostic plots, identical to those of the [modelling stock-recruitment tutorial](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html) for the `nsher_ri` example, are shown below.
+
+```{r, fig.height = 7}
+plot(newFL);
+```
+
+We now have a working example of a stock-recruitment model, but for our integration with `gmse_apply`, we will want a function that automates the above to simulate the process of updating the stock-recruitment model.
+We do this using the custom function created below.
 
 ```{r}
-#temp_dir <- tempdir()
-#download.file("http://flr-project.org/doc/src/loading_data.zip", file.path(dir, "loading_data.zip"))
-#unzip(file.path(temp_dir, "loading_data.zip"), exdir = temp_dir);
+update_SR_model <- function(rec_m, ssb_m, years){
+    Frec_m       <- FLQuant(rec.m, dimnames=list(age=1, year = years));
+    Fssb_m       <- FLQuant(ssb.m, dimnames=list(age=1, year = years));
+    Frec_m@units <- "10^3";
+    Fssb_m@units <- "t*10^3";
+    rec(newFL)   <- Frec.m;
+    ssb(newFL)   <- Fssb.m;
+    range(newFL) <- c(0, years[1], 0, years[length(years)]);
+    model(newFL) <- ricker();
+    newFL        <- fmle(newFL);
+    return(newFL);
+}
 ```
 
-We can read in the example file below and convert it into a matrix to be used in an [FLQuant](http://www.flr-project.org/FLCore/reference/FLQuant.html) object. Rows in this matrix indicate ages.
+The above function will be used within another custom function to predict the next time step of recruitment.
 
 ```{r}
-#fish_data <- read.csv(file.path(temp_dir, "catch_numbers.csv"), row = 1);
-#fish_data <- as.matrix(fish_data);
-#print(fish_data[,1:8]);
+predict_recruitment <- function(rec_m, ssb_m, years, new_ssb){
+    newFL <- update_SR_model(rec_m, ssb_m, years);
+    a     <- params(newFL)[[1]] # Extract 'a' parameter of the Ricker model
+    b     <- params(newFL)[[2]] # Extract 'b' parameter of the Ricker model
+    rec   <- a * new_ssb * exp(-b * new_ssb); # Predict the new recruitment
+    return(recruitment)
+}
 ```
 
-For simplicity, we merge these ages down into a single vector.
+In `gmse_apply`, we will use the `predict_recruitment` function above as the resource (i.e., operational) model.
+The `new_ssb` reads in the new spawning stock biomass, which will be calculated from the built-in GMSE `user` model.
 
-```{r}
-fish_data <- apply(X = fish_data, MARGIN = 2, FUN = sum);
-```
-
-We can now create a new [FLQuant](http://www.flr-project.org/FLCore/reference/FLQuant.html) object with the `fish_data` vector.
-This process is important; down below, it will be repeated in a function that gets used within the GMSE population model, as GMSE will need to incorporate newly simulated data in each time step.
-
-```{r}
-fish_data_flq <- FLQuant(fish_data, dimnames = list(year = 1957:2011));
-```
-
-The data below include sample results for recruitment and spawning-stock biomass (SSB) of North Sea herring.
-
-```{r}
-data(nsher);
-```
-
-These data are structured using `FLStock`, an [S4](https://stat.ethz.ch/R-manual/R-devel/library/methods/html/Classes_Details.html) class of R objects.
-The code below creates an `FLSR` object, needed to create a stock-recruitment model.
-
-```{r}
-summary(nsher);
-```
-
-Notice that under `Model`, a Ricker model of stock-recruitment has already been fitted (many stock-recruitment models are available in FLCore; see the [FLSR tutorial](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html) for more about fitting different models) and parameterised with Maximum Likelihood Estimation (MLE).
-The code below shows how to apply the Beverton-Holt model model as an alternative, and the  `fmle` function parameterises the Beverton-Holt model model using the data in `nsher` and MLE.
-
-```{r}
-model(nsher) <- bevholt();
-nsher        <- fmle(nsher);
-summary(nsher);
-```
-
-We will use these tools to update the population model in GMSE.
-
-a <- params(nsher)[[1]];
-b <- params(nsher)[[2]];
-
-rec(p4sr2);
-ssb(p4sr2);
-
-rec_year    <- as.numeric(attributes(rec(p4sr2))$dimnames$year);
-ssb_year    <- as.numeric(attributes(ssb(p4sr2))$dimnames$year);
-recruitment <- NULL;
-for(year in 1:length(rec_year)){
-    recruitment[year] <- rec(p4sr)[[year]];
-}
-spawning_stock_biomass <- NULL;
-for(year in 1:length(ssb_year)){
-    spawning_stock_biomass[year] <- ssb(p4sr)[[year]];
-}
-
-rec <- a * spawning_stock_biomass * exp(-b * spawning_stock_biomass);
-
-rec(p4sr)[[51]] <- 4300
-
---->
+Integrating `predict_recruitment` with `gmse_apply`
+================================================================================

### 2018-05-01 R/plotting.R
Minor edit

@@ -188,7 +188,7 @@ case23plot <- function(res, obs, land1, land2, land3, agents, paras, COST,
         par(mar=c(4,4,1,4));
         plot(x=gens, y=abun, pch=20, type="l", lwd=2, ylim=c(0, ymaxi),
              xlim=c(0,time_max), xlab="Time Step", ylab="Abundance",
-             cex.lab=1.25);
+             cex.lab=1.25, yaxs="i");
         new_est   <- sum(obs_t[,13]);
         est       <- c(est, new_est);
         points(x=gens, y=est, pch=20, type="l", lwd=2, col="cyan4");
@@ -198,7 +198,7 @@ case23plot <- function(res, obs, land1, land2, land3, agents, paras, COST,
         par(new=TRUE);
         plot(x=gens, y=lnds, pch=20, type="l", lwd=3, col="orange", xlab = "",
              xlim=c(0, time_max), ylim = c(0, 100), xaxt="n", yaxt="n", 
-             ylab = "");
+             ylab = "", yaxs="i");
         axis(side=4, at=c(0, 25, 50, 75, 100));
         mtext("Mean % Yield", side = 4, line = 2.4);
         # ------------ Panel 4 (middle right);
@@ -206,7 +206,7 @@ case23plot <- function(res, obs, land1, land2, land3, agents, paras, COST,
         cell_number <- dim(land3)[1] * dim(land3)[2];
         plot(x=gens, y=gens, pch=20, type="n", lwd=2, ylim=c(0, 100),
              xlim=c(0,time_max), xlab="Time Step", ylab="Stake-holder % yield",
-             cex.lab=1.25);
+             cex.lab=1.25, yaxs="i");
         stake_colors <- topo.colors( dim(age_t)[1] );
         for(stakeholder in 1:dim(ages)[2]){
             max_yield   <- sum(land3 == stakeholder);
@@ -229,7 +229,7 @@ case23plot <- function(res, obs, land1, land2, land3, agents, paras, COST,
         y_upper_limit <- max_cost + (0.25 * max_cost);
         plot(x=gens, y=gens, pch=20, type="n", lwd=2, ylim=c(0, y_upper_limit),
              xlim=c(0,time_max), xlab="Time Step", ylab="Cost of actions",
-             cex.lab=1.25);
+             cex.lab=1.25, yaxs="i");
         if(paras[89] == 1){
[…]

### 2018-05-01 R/gmse.R
Tweak defaults again

@@ -106,8 +106,8 @@ gmse <- function( time_max       = 100,   # Max number of time steps in sim
                   move_agents    = TRUE,  # Move agents once per time step
                   max_ages       = 5,     # Maximum age of any resource(s)
                   minimum_cost   = 10,    # Minimum cost value
-                  user_budget    = 10000, # What is the budget of a user
-                  manager_budget = 10000, # The budget of a manager
+                  user_budget    = 1000,  # What is the budget of a user
+                  manager_budget = 1000,  # The budget of a manager
                   manage_target  = 1000,  # The target resource abundance
                   RESOURCE_ini   = 1000,  # Number of initial resources
                   scaring        = FALSE, # Scaring allowed in simulations
@@ -123,7 +123,7 @@ gmse <- function( time_max       = 100,   # Max number of time steps in sim
                   land_ownership = FALSE, # Do stake-holders act on their land?
                   manage_freq    = 1,     # Frequency that management enacted
                   converge_crit  = 0.1,   # Convergence criteria
-                  manager_sense  = 0.8,   # Manager sensitivity
+                  manager_sense  = 0.9,   # Manager sensitivity
                   public_land    = 0,     # Proportion of landscape public
                   group_think    = FALSE  # All users behave identically
 ){

### 2018-05-01 notebook/ms/SI5.Rmd
Update SI
Update to SI5

@@
 ---
+
+```{r, echo = FALSE}
+base_r_plot <- plot;
+```
 
@@
     rec   <- a * new_ssb * exp(-b * new_ssb); # Predict the new recruitment
-    return(recruitment)
+    return(rec)
 }
@@
 ================================================================================
+
+The [FLR project](http://www.flr-project.org/) includes libraries that can be used to [perform a management strategy evaluation](http://www.flr-project.org/doc/An_introduction_to_MSE_using_FLR.html) (MSE) under fisheries-focused observation, manager, and user models.
+We will not recreate [this approach](http://www.flr-project.org/doc/An_introduction_to_MSE_using_FLR.html), or integrate any other submodels into GMSE as was done for the population model above, although such integration of submodels should be possible using similar techniques.
+Our goal here is to instead show how the `predict_recruitment` model created above can be integrated with `gmse_apply`, which can then make use of the genetic algorithm to predict the behaviour fishers.
+
+We will use a custom observation model, which will simply estimate recruitment with some fixed error.
+
+```{r}
+obs_ssb <- function(resource_vector){
+    obs_err <- rnorm(n = 1, mean = 0, sd = 100);
+    the_obs <- resource_vector + obs_err;
+    return(the_obs);
+}
+```
+
+Hence, we can now feed the data from `rec.m` and `ssb.m` through `predict_recruitment`, which will return a value for new recruitment, and this new value can in turn be fed into `obs_ssb` to predict recruitment with some error.
+We also need a new spawning stock biomass `new_ssb`, which we can just initialise with the biomass from the last year in `ssb.m`
+
+```{r, echo = FALSE, messages = FALSE, cache = FALSE, comment = FALSE, results = "hide"}
+ssb_ini <- ssb.m[length(ssb.m)];
+new_rec <- predict_recruitment(rec_m = rec.m, ssb_m = ssb.m, years = 1960:2004,
+                               new_ssb = ssb_ini);
+obs_rec <- obs_ssb(new_rec);
+```
+
+```{r, eval = FALSE}
+ssb_ini <- ssb.m[length(ssb.m)];
+new_rec <- predict_recruitment(rec_m = rec.m, ssb_m = ssb.m, years = 1960:2004,
+                               new_ssb = ssb_ini);
+obs_rec <- obs_ssb(new_rec);
+```
+
+An initial run of these models gives values of `r round(new_rec, digits = 2)` for `new_rec` and `r round(obs_rec, digits = 2)` for `obs_rec`.
+We are now ready to use the built-in manager and user submodels in `gmse_apply`.
+We will assume that managers attempt to keep a recruitment of 5000, and that there are 4 independent fishers who attempt to maximise their catch.
+We assign a user budget of `manager_budget = 10000`, and all other values are set to GMSE defaults.
+In the built-in GMSE functions, the manager will use the estimate of recruitment based on `obs_rec` and use it to set the cost of harvesting (`culling` in GMSE).
+
+```{r, eval = FALSE}
+library(GMSE);
+rec.n        <- read.csv("data/SI5/nsher_rec.csv");
+ssb.n        <- read.csv("data/SI5/nsher_ssb.csv");
+yrspan       <- 1960:2004;
+rec.m        <- as.matrix(rec.n);
+ssb.m        <- as.matrix(ssb.n);
+
+sim <- gmse_apply(res_mod = predict_recruitment, obs_mod = obs_ssb,
+                  rec_m = rec.m, ssb_m = ssb.m, years = yrspan, 
+                  new_ssb = ssb_ini, manage_target = 5000, stakeholders = 10,
+                  manager_budget = 10000);
+print(sim);
+```
+
+```{r, echo = FALSE, messages = FALSE, cache = FALSE, comment = FALSE, results = "hide"}
+library(GMSE);
+rec.n        <- read.csv("data/SI5/nsher_rec.csv");
+ssb.n        <- read.csv("data/SI5/nsher_ssb.csv");
+yrspan       <- 1960:2004;
+rec.m        <- as.matrix(rec.n);
+ssb.m        <- as.matrix(ssb.n);
+
+sim <- gmse_apply(res_mod = predict_recruitment, obs_mod = obs_ssb,
+                  rec_m = rec.m, ssb_m = ssb.m, years = yrspan, 
+                  new_ssb = ssb_ini, manage_target = 5000, stakeholders = 10,
+                  manager_budget = 10000);
+```
+
+```{r, echo = FALSE}
+print(sim);
+```
+
+The resource and observation results above are interpreted in terms of recruitment, while the manager results are interpreted in terms of the cost of harvesting a unit of spawning stock biomass and the user results are interpreted in terms of how much biomass was harvested.
+Note in the run of `gmse_apply` that the arguments for our custom resource and observation models (`predict_recruitment` and `obs_ssb`, respectively) are read directly in as arguments of `gmse_apply` itself.
+The `gmse_apply` function will figure out which subfunctions custom arguments should go to, then update these arguments as needed over the course of a single run of `gmse_apply`.
+
+Simulation with `gmse_apply` over multiple time steps
+================================================================================
+
+We are now ready to loop the `gmse_apply` function over multiple time steps.
+To do this, we will update the `rec.m` and `ssb.m` matrices after each time step, simulating 20 years into the future.
+The population model `predict_recruitment` will use these data to dynamically update parameters of the Ricker model, as might occur in an empirical fishery that is being monitored.
+For simplicity, here we will use the results from the observation model to update recruiment for the new year in `rec.m`, and subtract the biomass culled by users

### 2018-05-02 notebook/ms/SI2.Rmd
Just remove some white space
Update author list
SI stuff, problem with gmse_apply found in development

@@
 subtitle: Supporting Information 2
-author: A.
-Bradley Duthie&#xb9;&#xb3;, Jeremy J.
-Cusack&#xb9;, Isabel L.
-Jones&#xb9;, Erlend B.
-Nilsen&#xb2;, Roc&#0237;o A.
-Pozo&#xb9;, O.
-Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
+  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
+  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
 date: "[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK
@@
 gmse_apply(res_mod = alt_res, obs_mod = observation, 
-           man_mod = alt_man, use_mod = alt_usr, X = 1000)
+           man_mod = alt_man, use_mod = alt_usr, X = 1000);
 ```
@@
 gmse_apply(res_mod = alt_res, obs_mod = alt_obs, 
-           man_mod = manager, use_mod = user, X = 1000)
+           man_mod = manager, use_mod = user, X = 1000);
 ```
@@
 This `old_list` produced when `get_res = Full` includes all data structures and parameter values necessary for a unique simulation of GMSE.
+Note that custom functions sent to `gmse_apply` still need to be specified (`res_mod`, `obs_mod`, `man_mod`, and `use_mod`).
 An exampe of using `get_res` and `old_list` in tandem to loop `gmse_apply` is shown below.

### 2018-05-02 notebook/ms/SI5.Rmd
SI stuff, problem with gmse_apply found in development

@@
 update_SR_model <- function(rec_m, ssb_m, years){
-    Frec_m       <- FLQuant(rec.m, dimnames=list(age=1, year = years));
-    Fssb_m       <- FLQuant(ssb.m, dimnames=list(age=1, year = years));
+    Frec_m       <- FLQuant(rec_m, dimnames=list(age = 1, year = years));
+    Fssb_m       <- FLQuant(ssb_m, dimnames=list(age = 1, year = years));
     Frec_m@units <- "10^3";
@@
 The population model `predict_recruitment` will use these data to dynamically update parameters of the Ricker model, as might occur in an empirical fishery that is being monitored.
-For simplicity, here we will use the results from the observation model to update recruiment for the new year in `rec.m`, and subtract the biomass culled by users
+We will use the results from the observation model to update recruiment for the new year in `rec.m`.
+For simplicity, spawning stock biomass prior to harvest will be randomly sampled from a value in last 10 years (i.e., from `ssb.m` between 1994 and 2004), but more realistic models could relate this spawning stock biomass to recruitment and environmental variables from a prevoius year; spawning stock biomass will be decreased after harvest based on user actions.
+The GMSE initialisation and simulation is below.
+
+```{r, echo = FALSE, messages = FALSE, cache = FALSE, comment = FALSE, results = "hide"}
+# This code initialises the simulation -----------------------------------------
+rec.n        <- read.csv("data/SI5/nsher_rec.csv");
+ssb.n        <- read.csv("data/SI5/nsher_ssb.csv");
+yrspan       <- 1960:2004;
+rec.m        <- as.matrix(rec.n);
+ssb.m        <- as.matrix(ssb.n);
+ssb_ini      <- ssb.m[length(ssb.m)];
+sim_old      <- gmse_apply(res_mod = predict_recruitment, obs_mod = obs_ssb,
+                           rec_m = rec.m, ssb_m = ssb.m, years = yrspan, 
+                           new_ssb = ssb_ini, manage_target = 3500, 
+                           stakeholders = 10, manager_budget = 10000, 
+                           get_res = "Full");
+# The code below simulates 20 time steps ---------------------------------------
+sim_sum  <- matrix(data = NA, nrow = 20, ncol = 6); # Hold results here
+for(time_step in 1:20){
+    # Update the relevant parameter values as necessary -----------------------
+    rand_ssb        <- sample(x = ssb.m[35:45], size = 1);
+    harvest         <- sum(sim_old$basic_output$user_results[,3]);
+    new_rec_m       <- c(sim_old$rec_m, sim_old$observation_vector);
+    new_ssb_m       <- c(sim_old$ssb_m, rand_ssb - harvest);
+    sim_old$rec_m   <- matrix(data = new_rec_m, nrow = 1);
+    sim_old$ssb_m   <- matrix(data = new_ssb_m, nrow = 1);
+    sim_old$years   <- c(sim_old$years, time_step + 2004);
+    sim_old$new_ssb <- sim_old$ssb_m[length(sim_old$ssb_m)];
+    # Run a new simulation in the loop: custom functions are always specified -
+    sim_new  <- gmse_apply(get_res = "Full", old_list = sim_old,
+                           res_mod = predict_recruitment, obs_mod = obs_ssb);
+    # Record the results in sim_sum -------------------------------------------
+    sim_sum[time_step, 1] <- time_step + 2004;
+    sim_sum[time_step, 2] <- sim_new$basic_output$resource_results[1];
+    sim_sum[time_step, 3] <- sim_new$basic_output$observation_results[1];
+    sim_sum[time_step, 4] <- sim_new$basic_output$manager_results[3];
+    sim_sum[time_step, 5] <- harvest;
+    sim_sum[time_step, 6] <- sim_new$new_ssb;
+    # Redefine the old list --------------------------------------------------
+    sim_old               <- sim_new;
+}
+colnames(sim_sum) <- c("Year", "Recruitment", "Recruit_estim", "Harvest_cost", 
+                         "Harvested", "SSB");
+print(sim_sum);
+```

### 2018-05-03 notebook/ms/SI5.Rmd
Minor edits, and a closing paragraph in SI5

@@
 
-To start, we need to read in the `FLCore` library.
+To start, we need to read in the `FLCore` and GMSE libraries.
 
@@
 library(FLCore);
+library(GMSE);
 ```
@@
 This data set includes recruitment and spawning stock biomass data between 1960 and 2004.
-First, we initialise an empty `FLSR` object and read in the recreated CSV data files.
+First, we initialise an empty `FLSR` object and read in the recreated herring data files from GMSE, which contains recruitment (`rec.n`) and spawning stock biomass (`ssb.n`)
 
@@
 newFL   <- FLSR(); # Initialises the empty FLSR object
-rec.n        <- read.csv("data/SI5/nsher_rec.csv");
-ssb.n        <- read.csv("data/SI5/nsher_ssb.csv");
+data(nsher_data);
 ```
@@
 ```{r, eval = FALSE}
-library(GMSE);
-rec.n        <- read.csv("data/SI5/nsher_rec.csv");
-ssb.n        <- read.csv("data/SI5/nsher_ssb.csv");
 yrspan       <- 1960:2004;
@@
 ```
+
+```{r, echo = TRUE, eval = FALSE}
+# This code initialises the simulation -----------------------------------------
+yrspan       <- 1960:2004;
+rec.m        <- as.matrix(rec.n);
+ssb.m        <- as.matrix(ssb.n);
+ssb_ini      <- ssb.m[length(ssb.m)];
+sim_old      <- gmse_apply(res_mod = predict_recruitment, obs_mod = obs_ssb,
+                           rec_m = rec.m, ssb_m = ssb.m, years = yrspan, 
+                           new_ssb = ssb_ini, manage_target = 3500, 
+                           stakeholders = 10, manager_budget = 10000, 
+                           get_res = "Full");
+# The code below simulates 20 time steps ---------------------------------------
+sim_sum  <- matrix(data = NA, nrow = 20, ncol = 6); # Hold results here
+for(time_step in 1:20){
+    # Update the relevant parameter values as necessary -----------------------
+    rand_ssb        <- sample(x = ssb.m[35:45], size = 1);
+    harvest         <- sum(sim_old$basic_output$user_results[,3]);
+    new_rec_m       <- c(sim_old$rec_m, sim_old$observation_vector);
+    new_ssb_m       <- c(sim_old$ssb_m, rand_ssb - harvest);
+    sim_old$rec_m   <- matrix(data = new_rec_m, nrow = 1);
+    sim_old$ssb_m   <- matrix(data = new_ssb_m, nrow = 1);
+    sim_old$years   <- c(sim_old$years, time_step + 2004);
+    sim_old$new_ssb <- sim_old$ssb_m[length(sim_old$ssb_m)];
+    # Run a new simulation in the loop: custom functions are always specified -
+    sim_new  <- gmse_apply(get_res = "Full", old_list = sim_old,
+                           res_mod = predict_recruitment, obs_mod = obs_ssb);
+    # Record the results in sim_sum -------------------------------------------
+    sim_sum[time_step, 1] <- time_step + 2004;
+    sim_sum[time_step, 2] <- sim_new$basic_output$resource_results[1];
+    sim_sum[time_step, 3] <- sim_new$basic_output$observation_results[1];
+    sim_sum[time_step, 4] <- sim_new$basic_output$manager_results[3];
+    sim_sum[time_step, 5] <- harvest;
+    sim_sum[time_step, 6] <- sim_new$new_ssb;
+    # Redefine the old list --------------------------------------------------
+    sim_old               <- sim_new;
+}
+colnames(sim_sum) <- c("Year", "Recruitment", "Recruit_estim", "Harvest_cost", 
+                         "Harvested", "SSB");
+print(sim_sum);
+```
+
+```{r, echo = FALSE}
+print(sim_sum);
+```
+
+The above output from `sim_sum` reports the recruitment (resource or operational model), recruitment estimate (observation error model), management (harvest control model), and user (implementation model) simulation results.
+This example simulation demonstrates the ability of GMSE to integrate with fisheries libraries such as [FLR](http://www.flr-project.org/) through `gmse_apply`.
+In addition to being a useful wrapping function for MSE submodels, `gmse_apply` can therefore be used to take advantage of the genetic algorithm in the GMSE default manager and user models.
+This flexibility will be retained in future versions of `gmse_apply`, allowing custom resource and observation models that are built for specific systems to be integrated with an increasingly complex genetic algorithm simulating various aspects of human decision-making.
+
+Conclusions
+================================================================================
+
+GMSE is a general, flexible, tool for simulating the management of resources under situations of uncertainty and conflict.
+Management Strategy Evaluation [@Bunnefeld2011; @Punt2016], the framework upon which GMSE is based, had its origin in fisheries management [@Polacheck1999; @Smith1999; @Sainsbury2000], and here we showed one example of how GMSE could be integrated with the core package of the [Fisheries Library in R](http://www.flr-project.org/).
+
+Future versions of GMSE will continue to be open-source and developed to avoid unecessary dependencies (GMSE v.0.4.0.3 requires only base R).
+Key goals including (1) providing highly general and useful default `resource`, `observation`, `manager`, and `user` submodels for a variety of MSE modelling tasks, (2) keeping these submodels highly modular so that they can be developed in isolation given standardised data structures, and (3) allowing these modular submodels to be integrated with custom defined submodels as flexibly as possible using `gmse_apply`.
+Contributions in line with these goals, and suggestions for new features, can be made on [GitHub](https://github.com/bradduthie/gmse).

### 2018-05-04 notebook/ms/SI5.Rmd
changed SI5 text changes

@@
 
-As part of the [ConFooBio](https://sti-cs.org/confoobio/) project, a central focus of GMSE is on simulating the management of populations of conservation interest, with a particular emphasis on understanding conservation conflict; further development of GMSE is expected to continue with this as a priority, further building upon the decision-making algorithms of managers and users to better understand how conflict arises and can potentially be resolved.
+D'sfl/,f';ldf As part of the [ConFooBio](https://sti-cs.org/confoobio/) project, a central focus of GMSE is on simulating the management of populations of conservation interest, with a particular emphasis on understanding conservation conflict; further development of GMSE is expected to continue with this as a priority, further building upon the decision-making algorithms of managers and users to better understand how conflict arises and can potentially be resolved.
 Hence, GMSE is not intended as a substitute for packages such as [FLR](http://www.flr-project.org/), but the integration of these packages with GMSE could make use of GSME's current and future simulation capabilities, and particularly the [genetic algorithm]().

### 2018-05-04 notebook/ms/SI5.Rmd
change text 2

--- 2018-05-03 notebook/ms/SI5.Rmd
@@
 
-As part of the [ConFooBio](https://sti-cs.org/confoobio/) project, a central focus of GMSE is on simulating the management of populations of conservation interest, with a particular emphasis on understanding conservation conflict; further development of GMSE is expected to continue with this as a priority, further building upon the decision-making algorithms of managers and users to better understand how conflict arises and can potentially be resolved.
+D'sfl;As part of the [ConFooBio](https://sti-cs.org/confoobio/) project, a central focus of GMSE is on simulating the management of populations of conservation interest, with a particular emphasis on understanding conservation conflict; further development of GMSE is expected to continue with this as a priority, further building upon the decision-making algorithms of managers and users to better understand how conflict arises and can potentially be resolved.
 Hence, GMSE is not intended as a substitute for packages such as [FLR](http://www.flr-project.org/), but the integration of these packages with GMSE could make use of GSME's current and future simulation capabilities, and particularly the [genetic algorithm]().

### 2018-05-04 notebook/ms/SI1.Rmd
Update author list
Updates

@@
 ---
-title: 'GMSE: an R package for generalised management strategy evaluation'
-subtitle: Supporting Information 1
-author: A.
-Bradley Duthie&#xb9;&#xb3;, Jeremy J.
-Cusack&#xb9;, Isabel L.
-Jones&#xb9;, Erlend B.
-Nilsen&#xb2;, Roc&#0237;o A.
-Pozo&#xb9;, O.
-Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+title: 'The Genetic Algorithm of GMSE'
+subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 1)'
+author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
+  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
+  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
 date: "[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK

### 2018-05-04 notebook/ms/SI4.Rmd
Update author list
Updates

@@
 ---
-title: 'GMSE: an R package for generalised management strategy evaluation'
+title: 'Advanced case study options'
+subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 4)'
 author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
-  Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;,
-  Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
+  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
 date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
@@
 bibliography: gmse.bib
-subtitle: Supporting Information 4
 biblio-style: apalike

### 2018-05-04 notebook/ms/SI5.Rmd

@@
 ---
-title: 'GMSE: an R package for generalised management strategy evaluation'
+title: "Fisheries example integrating FLR"
 author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
-  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
-  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy
+  Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
 date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
@@
   html_document: default
-  word_document:
-    fig_caption: yes
-    reference_docx: docx_template.docx
 header-includes: null
@@
 bibliography: gmse.bib
-subtitle: Supporting Information 5
+subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting
+  Information 5)'
 biblio-style: apalike

### 2018-05-04 notebook/ms/SI5.Rmd
notebook

@@
 
-D'sfl;As part of the [ConFooBio](https://sti-cs.org/confoobio/) project, a central focus of GMSE is on simulating the management of populations of conservation interest, with a particular emphasis on understanding conservation conflict; further development of GMSE is expected to continue with this as a priority, further building upon the decision-making algorithms of managers and users to better understand how conflict arises and can potentially be resolved.
+As part of the [ConFooBio](https://sti-cs.org/confoobio/) project, a central focus of GMSE is on simulating the management of populations of conservation interest, with a particular emphasis on understanding conservation conflict; further development of GMSE is expected to continue with this as a priority, further building upon the decision-making algorithms of managers and users to better understand how conflict arises and can potentially be resolved.
 Hence, GMSE is not intended as a substitute for packages such as [FLR](http://www.flr-project.org/), but the integration of these packages with GMSE could make use of GSME's current and future simulation capabilities, and particularly the [genetic algorithm]().

### 2018-05-04 notebook/ms/SI6.Rmd
New sentences

---
title: 'Management frequency and extinction risk'
subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 5)'
author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes: null
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: gmse.bib
biblio-style: apalike
---

Replicate simulations as a tool for model inference
================================================================================

The default submodels of GMSE (`resource`, `observation`, `manager`, `user`) are individual-based (also called 'agent-based'), meaning that they model discrete individuals (resources or agents), which GMSE are represented by individual table rows (`RESOURCES`, `AGENTS`, `OBSERVATION`) or layers of three-dimensional arrays (`COST`, `ACTION`). Individual-based modelling has been a useful approach in ecology for decades [@Uchmanski1996; @Grimm1999], providing both a pragmatic tool for the mechanistic modelling of complex populations and a powerful technique for theoretical investigation. A key advantage of the individual-based modelling approach is the discrete nature of individuals, which allows for detailed trait variation and complex interactions among individuals. In GMSE, some of the most important traits for resources include different types, ages, demographic paramater values, locations, etc., and for agents (manager and users), this can include different types, goals, budgets, etc.

### 2018-05-04 notebook/ms/SI5.Rmd
Text changes throughout
Changes relate mostly to the framing of the conflict situation. Also some minor changes throughout.

--- 2018-05-04 notebook/ms/SI5.Rmd
@@
 
-D'sfl;As part of the [ConFooBio](https://sti-cs.org/confoobio/) project, a central focus of GMSE is on simulating the management of populations of conservation interest, with a particular emphasis on understanding conservation conflict; further development of GMSE is expected to continue with this as a priority, further building upon the decision-making algorithms of managers and users to better understand how conflict arises and can potentially be resolved.
+As part of the [ConFooBio](https://sti-cs.org/confoobio/) project, a central focus of GMSE is on simulating the management of animal populations of conservation interest, with a particular emphasis on understanding conservation conflict; further development of GMSE is expected to continue with this as a priority, further building upon the decision-making algorithms of managers and users to better understand how conflict arises and can be managed and mitigated.
 Hence, GMSE is not intended as a substitute for packages such as [FLR](http://www.flr-project.org/), but the integration of these packages with GMSE could make use of GSME's current and future simulation capabilities, and particularly the [genetic algorithm]().
 Such integration might be possible using the `gmse_apply` function, which allows for [custom defined submodels]() to be used within the GMSE framework, and with default GMSE submodels.
-Hence, GMSE might be especially useful for modelling the management of fisheries under conditions of increasing harvesting demands and stakeholder conflict.
+Hence, GMSE might be especially useful for modelling the management of fisheries under conditions of increasing competing staleholder demands and conflicts.
 We do not attempt such an ambitious project here, but instead show how such a project could be developed through integration of [FLR](http://www.flr-project.org/) and `gmse_apply`.
 
-Here we follow a [Modelling Stock-Recruitment with FLSR](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html) example, then integrate this example with `gmse_apply` to explore the behaviour of simulated fishers who are goal-driven to maximise their own harvest.
+Here we follow a [Modelling Stock-Recruitment with FLSR](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html) example, then integrate this example with `gmse_apply` to explore the behaviour of a number of simulated fishers who are goal-driven to maximise their own harvest and a manager that aims to keep the fish stocks at a predefined target level.
+The core concept in GMSE is that manager can only incentivise fishers to harvest less or more by varying the cost of fishing (through e.g.
+taxes) given a set manager budget; please note that the manager cannot force the fisher to follow any policy.
+Based on the cost of fishing, the fisher can then given their own budget decide whether to invest in fishing or keep the budget.
+This concepts represents a nartural resource managmeent and conservation conflict, where one party aims to maximise their livelihood (fisher) and the other aims to keep a population at a sustainable level and prevent it from going extinct.
+Importantly, the manager does not have full control over fishers but can set policies to incentivise sustainable behaviour.
 We emphasise that this example is provided only as demonstration of how GMSE can potentially be integrated with already developed fisheries models, and is not intended to make recommendations for management in any population.
@@
 newFL   <- FLSR(); # Initialises the empty FLSR object
-data(nsher_data);
+data(nsher_data); #called from GMSE library rather than nsher from FLCore library
 ```
@@
 We are now ready to use the built-in manager and user submodels in `gmse_apply`.
-We will assume that managers attempt to keep a recruitment of 5000, and that there are 4 independent fishers who attempt to maximise their catch.
+We will assume that managers attempt to keep a recruitment of 5000, and that there are 4 independent fishers [stakehodlers in gmse_apply says 10, is that a difference?] who attempt to maximise their catch.
 We assign a user budget of `manager_budget = 10000`, and all other values are set to GMSE defaults.
@@
 
-The above output from `sim_sum` reports the recruitment (resource or operational model), recruitment estimate (observation error model), management (harvest control model), and user (implementation model) simulation results.
+The above output from `sim_sum` reports the recruitment (resource or operational model), recruitment estimate (observation error model), management set harvest cost (harvest control model), user harvested numbers (implementation model) and spawning stock biomass (SSB) simulation results.
 This example simulation demonstrates the ability of GMSE to integrate with fisheries libraries such as [FLR](http://www.flr-project.org/) through `gmse_apply`.

### 2018-05-04 notebook/ms/SI2.Rmd
Minor edits, and a closing paragraph in SI5

@@
 ---
-title: 'GMSE: an R package for generalised management strategy evaluation'
-subtitle: Supporting Information 2
+title: 'Use of the gmse_apply function'
+subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 2)'
 author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
@@
 Note that custom functions sent to `gmse_apply` still need to be specified (`res_mod`, `obs_mod`, `man_mod`, and `use_mod`).
-An exampe of using `get_res` and `old_list` in tandem to loop `gmse_apply` is shown below.
+An example of using `get_res` and `old_list` in tandem to loop `gmse_apply` is shown below.
 

### 2018-05-04 notebook/ms/SI3.Rmd
Update author list

@@
 ---
-title: 'GMSE: an R package for generalised management strategy evaluation'
+title: 'Example case study in GMSE'
+subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 3)'
 author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
-  Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;,
-  Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
+  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
 date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
@@
 bibliography: gmse.bib
-subtitle: Supporting Information 3
 biblio-style: apalike

### 2018-05-04 notebook/ms/SI6.Rmd

@@
 The default submodels of GMSE (`resource`, `observation`, `manager`, `user`) are individual-based (also called 'agent-based'), meaning that they model discrete individuals (resources or agents), which GMSE are represented by individual table rows (`RESOURCES`, `AGENTS`, `OBSERVATION`) or layers of three-dimensional arrays (`COST`, `ACTION`).
-Individual-based modelling has been a useful approach in ecology for decades [@Uchmanski1996; @Grimm1999], providing both a pragmatic tool for the mechanistic modelling of complex populations and a powerful technique for theoretical investigation.
-A key advantage of the individual-based modelling approach is the discrete nature of individuals, which allows for detailed trait variation and complex interactions among individuals.
-In GMSE, some of the most important traits for resources include different types, ages, demographic paramater values, locations, etc., and for agents (manager and users), this can include different types, goals, budgets, etc.

### 2018-05-04 notebook/ms/SI6.Rmd
More text

--- 2018-05-04 notebook/ms/SI6.Rmd
@@
 ---
-title: 'Management frequency and extinction risk'
-subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 5)'
+title: "Management frequency and extinction risk"
 author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
-  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
-  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy
+  Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
 date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
@@
 output:
+  html_document: default
   pdf_document:
@@
     fig_caption: yes
-  html_document: default
   word_document:
@@
 bibliography: gmse.bib
+subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting
+  Information 5)'
 biblio-style: apalike
@@
 A key advantage of the individual-based modelling approach is the discrete nature of individuals, which allows for detailed trait variation and complex interactions among individuals.
-In GMSE, some of the most important traits for resources include different types, ages, demographic paramater values, locations, etc., and for agents (manager and users), this can include different types, goals, budgets, etc.
+In GMSE, some of the most important traits for resources include types, ages, demographic paramater values, locations, etc., and for agents (manager and users), traits include different types, utilities, budgets, etc.
+The traits that resources and managers have can potentially affect their interactions, and default GMSE submodels take advantage of this by simulating interactions explicitly on a landscape (see [SI7](SI7.pdf) for an introduction to GMSE default data structures).

### 2018-05-04 notebook/ms/SI5.Rmd
Add Nils' corrections

@@
 As part of the [ConFooBio](https://sti-cs.org/confoobio/) project, a central focus of GMSE is on simulating the management of animal populations of conservation interest, with a particular emphasis on understanding conservation conflict; further development of GMSE is expected to continue with this as a priority, further building upon the decision-making algorithms of managers and users to better understand how conflict arises and can be managed and mitigated.
-Hence, GMSE is not intended as a substitute for packages such as [FLR](http://www.flr-project.org/), but the integration of these packages with GMSE could make use of GSME's current and future simulation capabilities, and particularly the [genetic algorithm]().
-Such integration might be possible using the `gmse_apply` function, which allows for [custom defined submodels]() to be used within the GMSE framework, and with default GMSE submodels.
-Hence, GMSE might be especially useful for modelling the management of fisheries under conditions of increasing competing staleholder demands and conflicts.
+Hence, GMSE is not intended as a substitute for packages such as [FLR](http://www.flr-project.org/), but the integration of these packages with GMSE could make use of GSME's current and future simulation capabilities, and particularly the [genetic algorithm](SI1.pdf).
+Such integration might be possible using the `gmse_apply` function, which allows for [custom defined submodels](SI2.pdf) to be used within the GMSE framework, and with default GMSE submodels.
+Hence, GMSE might be especially useful for modelling the management of fisheries under conditions of increasing competing stakeholder demands and conflicts.
 We do not attempt such an ambitious project here, but instead show how such a project could be developed through integration of [FLR](http://www.flr-project.org/) and `gmse_apply`.
@@
 newFL   <- FLSR(); # Initialises the empty FLSR object
-data(nsher_data); #called from GMSE library rather than nsher from FLCore library
+data(nsher_data);  # Called from GMSE library (not from FLCore)
 ```

### 2018-05-04 notebook/ms/SI6.Rmd
Add Nils' corrections

@@
 
-The default submodels of GMSE (`resource`, `observation`, `manager`, `user`) are individual-based (also called 'agent-based'), meaning that they model discrete individuals (resources or agents), which GMSE are represented by individual table rows (`RESOURCES`, `AGENTS`, `OBSERVATION`) or layers of three-dimensional arrays (`COST`, `ACTION`).
+The default submodels of GMSE (`resource`, `observation`, `manager`, `user`) are individual-based (also called 'agent-based'), meaning that they model discrete individuals (resources or agents), which in GMSE are represented by individual table rows (`RESOURCES`, `AGENTS`, `OBSERVATION`) or layers of three-dimensional arrays (`COST`, `ACTION`).
 Individual-based modelling has been a useful approach in ecology for decades [@Uchmanski1996; @Grimm1999], providing both a pragmatic tool for the mechanistic modelling of complex populations and a powerful technique for theoretical investigation.
 A key advantage of the individual-based modelling approach is the discrete nature of individuals, which allows for detailed trait variation and complex interactions among individuals.
-In GMSE, some of the most important traits for resources include types, ages, demographic paramater values, locations, etc., and for agents (manager and users), traits include different types, utilities, budgets, etc.
+In GMSE, some of the most important traits for resources include types, ages, demographic parameter values, locations, etc., and for agents (manager and users), traits include different types, utilities, budgets, etc.
 The traits that resources and managers have can potentially affect their interactions, and default GMSE submodels take advantage of this by simulating interactions explicitly on a landscape (see [SI7](SI7.pdf) for an introduction to GMSE default data structures).

### 2018-05-04 notebook/ms/SI6.Rmd
Updates

@@
 output:
-  html_document: default
   pdf_document:
@@
     fig_caption: yes
+  html_document: default
   word_document:
@@
 
+```{r, echo = FALSE}
+library(GMSE);
+```
+
+The individual-based approach of default GMSE submodels
+================================================================================
+
+The default submodels of GMSE (`resource`, `observation`, `manager`, `user`) are individual-based (also called 'agent-based'), meaning that they model discrete individuals (resources or agents), which in GMSE are represented by individual table rows (`RESOURCES`, `AGENTS`, `OBSERVATION`) or layers of three-dimensional arrays (`COST`, `ACTION`).
+Individual-based models (IBMs) have been a useful approach in ecology for decades [@Uchmanski1996; @Grimm1999], providing both a pragmatic tool for the mechanistic modelling of complex populations and a powerful technique for theoretical investigation.
+A key advantage of the individual-based modelling approach is the discrete nature of individuals, which allows for detailed trait variation and complex interactions among individuals.
+In GMSE, some of the most important traits for resources include types, ages, demographic parameter values, locations, etc., and for agents (manager and users), traits include different types, utilities, budgets, etc.
+The traits that resources and managers have can potentially affect their interactions, and default GMSE submodels take advantage of this by simulating interactions explicitly on a landscape (see [SI7](SI7.pdf) for an introduction to GMSE default data structures).
+
 Replicate simulations as a tool for model inference
@@
 
-The default submodels of GMSE (`resource`, `observation`, `manager`, `user`) are individual-based (also called 'agent-based'), meaning that they model discrete individuals (resources or agents), which in GMSE are represented by individual table rows (`RESOURCES`, `AGENTS`, `OBSERVATION`) or layers of three-dimensional arrays (`COST`, `ACTION`).
-Individual-based modelling has been a useful approach in ecology for decades [@Uchmanski1996; @Grimm1999], providing both a pragmatic tool for the mechanistic modelling of complex populations and a powerful technique for theoretical investigation.
-A key advantage of the individual-based modelling approach is the discrete nature of individuals, which allows for detailed trait variation and complex interactions among individuals.
-In GMSE, some of the most important traits for resources include types, ages, demographic parameter values, locations, etc., and for agents (manager and users), traits include different types, utilities, budgets, etc.
-The traits that resources and managers have can potentially affect their interactions, and default GMSE submodels take advantage of this by simulating interactions explicitly on a landscape (see [SI7](SI7.pdf) for an introduction to GMSE default data structures).
+Modelling complex interactions among discrete individuals mechanistically typically causes some degree of stochasticity in IBMs (in the code, this is caused by the sampling of random values, which determine probabilistically whether or not events such as birth or death occur for individuals), reflecting the uncertainty that is inherent to complex systems.
+We can see a simple example of this by calling `gmse_apply` under the same default conditions twice.
+
+```{r}
+rand_eg_1 <- gmse_apply();
+print(rand_eg_1);
+```
+
+Although the second call to `gmse_apply` has identical initial conditions, because resource demographics (e.g., birth and death) and agent decision making (e.g., policy generation and user actions) is not deterministic, a slightly different result is obtained.
+
+```{r}
+rand_eg_2 <- gmse_apply();
+print(rand_eg_2);
+```
+
+To make meaningful model inferences, it is often necessary to replicate simulations under the same initial conditions to understand the range of predicted outcomes for a particular set of parameter values.
+This can be computationally intense, but it can also lead to a more robust understanding of the range of dynamics that might be expected in a system.
+Additionally, when parameter values are unknown but believed to be important, replicate simulations can be applied across a range of values to understand how a particular parameter might affect system dynamics.
+Below, we show how to use the `gmse_replicates` function to simulate a simple example of a managed population that is hunted by users.
+This function calls `gmse` multiple times and aggregates the results from replicate simulations into a single table.

### 2018-05-04 R/gmse_table.R
New table maker

@@ -1,0 +1,23 @@
+#' GMSE table results
+#' 
+#' The gmse_table function takes results created from simulations of the gmse
+#' and concatenates key results from a large list into a more manageable data 
+#' table.
+#'
+#'@param gmse_sim The output of a `gmse` simulation.
+#'@param individual_actions Whether or not the actions of individual users will be included as columns of the resulting table. Aggregations of user actions will always be included. By default, this is set to `FALSE`.
+#'@param hide_unused_options Whether or not to hide results from policy options when creating the resulting table. If `TRUE` (default), then policy and user actions that are not allowed in a simulation will not be placed as columns. If `FALSE`, then these columns will be placed with values of `NA`.
+#'@param all_timesteps Whether or not results from each time step from the simulation should be individually placed as a row in the resulting table (`TRUE` by default). If `FALSE`, then only the last row will be placed.
+#'@return A table with one or more rows of results, each of which indicates a unique `gmse` simulation for a given time step. Columns represent key simulation including resource densities, observation estimates, policy, and user actions.
+#'@examples
+#'sim       <- gmse();
+#'sim_table <- gmse_table(gmse_sim = sim);
+#'@export
+gmse_table <- function(gmse_sim, individual_actions = FALSE, 
+                       hide_unused_options = TRUE, all_timesteps = TRUE){
+    
+    time_steps <- 1:length(gmse_sim$resource);
+    
+    
+    
+}

### 2018-05-04 notebook/ms/SI5.Rmd

@@
 output:
+  html_document: default
+  pdf_document:
+    citation_package: natbib
+    fig_caption: yes
   word_document:
@@
     reference_docx: docx_template.docx
-  pdf_document:
-    citation_package: natbib
-    fig_caption: yes
-  html_document: default
 header-includes: null

### 2018-05-04 notebook/ms/SI5.tex

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{setspace}
\setstretch{1}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdftitle={Fisheries example integrating FLR},
            pdfauthor={A. Bradley Duthie¹³, Jeremy J. Cusack¹, Isabel L. Jones¹, Jeroen Minderman¹, Erlend B. Nilsen², Rocío A. Pozo¹, O. Sarobidy Rakotonarivo¹, Bram Van Moorter², and Nils Bunnefeld¹},
            colorlinks=true,
            linkcolor=blue,
            citecolor=Blue,
            urlcolor=Blue,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{apalike}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\ImportTok}[1]{{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{{#1}}}}
\newcommand{\BuiltInTok}[1]{{#1}}
\newcommand{\ExtensionTok}[1]{{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\NormalTok}[1]{{#1}}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Fisheries example integrating FLR}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
\subtitle{GMSE: an R package for generalised management strategy evaluation
(Supporting Information 5)}
  \author{A. Bradley Duthie¹³, Jeremy J. Cusack¹, Isabel L. Jones¹, Jeroen
Minderman¹, Erlend B. Nilsen², Rocío A. Pozo¹, O. Sarobidy
Rakotonarivo¹, Bram Van Moorter², and Nils Bunnefeld¹}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{{[}1{]} Biological and Environmental Sciences, University of Stirling,
Stirling, UK {[}2{]} Norwegian Institute for Nature Research, Trondheim,
Norway {[}3{]}
\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}}

\begin{document}
\maketitle

\section{Integration and simulation with
fisheries}\label{integration-and-simulation-with-fisheries}

Early development of management strategy evaluation (MSE) models
originated in fisheries \citep{Polacheck1999, Smith1999, Sainsbury2000}.
Consequently, fisheries-focused software for MSE has been extensively
developed, including R libraries that focus on the management of species
of exceptional interest, such as the Atlantic Bluefin Tuna
(\emph{Thunnus thynnus})
\citep[\href{https://github.com/ICCAT/abft-mse/tree/master/R_package/ABTMSE}{ABFTMSE};][]{Carruthers2018, Carruthers2018a},
and Indian Ocean Bigeye (\emph{T. obesus}) and Yellowfin (\emph{T.
albacares}) Tuna
\citep[\href{https://github.com/pjumppanen/MSE-IO-BET-YFT}{MSE-IO-BET-YFT};][]{Kolody2016}.
The largest of all such libraries is the Fisheries Library in R
(\href{http://www.flr-project.org/}{FLR}), which includes an extensive
collection of tools targeted for fisheries science. The FLR library has
been used in over a hundred
\href{http://www.flr-project.org/\#publications}{publications}
\citep[recent publications
include][]{Jardim2018, Mackinson2018, Utizi2018}, and includes
\href{http://www.flr-project.org/doc/An_introduction_to_MSE_using_FLR.html}{an
MSE framework} for evaluating different harvest control rules.

\textless{}\textless{}\textless{}\textless{}\textless{}\textless{}\textless{}
HEAD As part of the \href{https://sti-cs.org/confoobio/}{ConFooBio}
project, a central focus of GMSE is on simulating the management of
animal populations of conservation interest, with a particular emphasis
on understanding conservation conflict; further development of GMSE is
expected to continue with this as a priority, further building upon the
decision-making algorithms of managers and users to better understand
how conflict arises and can be managed and mitigated. Hence, GMSE is not
intended as a substitute for packages such as
\href{http://www.flr-project.org/}{FLR}, but the integration of these
packages with GMSE could make use of GSME's current and future
simulation capabilities, and particularly the \href{}{genetic
algorithm}. Such integration might be possible using the
\texttt{gmse\_apply} function, which allows for \href{}{custom defined
submodels} to be used within the GMSE framework, and with default GMSE
submodels. Hence, GMSE might be especially useful for modelling the
management of fisheries under conditions of increasing competing
staleholder demands and conflicts. We do not attempt such an ambitious
project here, but instead show how such a project could be developed
through integration of \href{http://www.flr-project.org/}{FLR} and
\texttt{gmse\_apply}. ======= As part of the
\href{https://sti-cs.org/confoobio/}{ConFooBio} project, a central focus
of GMSE is on simulating the management of populations of conservation
interest, with a particular emphasis on understanding conservation
conflict; further development of GMSE is expected to continue with this
as a priority, further building upon the decision-making algorithms of
managers and users to better understand how conflict arises and can
potentially be resolved. Hence, GMSE is not intended as a substitute for
packages such as \href{http://www.flr-project.org/}{FLR}, but the
integration of these packages with GMSE could make use of GSME's current
and future simulation capabilities, and particularly the \href{}{genetic
algorithm}. Such integration might be possible using the
\texttt{gmse\_apply} function, which allows for \href{}{custom defined
submodels} to be used within the GMSE framework, and with default GMSE
submodels. Hence, GMSE might be especially useful for modelling the
management of fisheries under conditions of increasing harvesting
demands and stakeholder conflict. We do not attempt such an ambitious
project here, but instead show how such a project could be developed
through integration of \href{http://www.flr-project.org/}{FLR} and
\texttt{gmse\_apply}.
\textgreater{}\textgreater{}\textgreater{}\textgreater{}\textgreater{}\textgreater{}\textgreater{}
366b9295e25f74c310c7185eab57ad3e19bfa83c

Here we follow a
\href{http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html}{Modelling
Stock-Recruitment with FLSR} example, then integrate this example with
\texttt{gmse\_apply} to explore the behaviour of a number of simulated
fishers who are goal-driven to maximise their own harvest and a manager
that aims to keep the fish stocks at a predefined target level. The core
concept in GMSE is that manager can only incentivise fishers to harvest
less or more by varying the cost of fishing (through e.g.~taxes) given a
set manager budget; please note that the manager cannot force the fisher
to follow any policy. Based on the cost of fishing, the fisher can then
given their own budget decide whether to invest in fishing or keep the
budget. This concepts represents a nartural resource managmeent and
conservation conflict, where one party aims to maximise their livelihood
(fisher) and the other aims to keep a population at a sustainable level
and prevent it from going extinct. Importantly, the manager does not
have full control over fishers but can set policies to incentivise
sustainable behaviour. We emphasise that this example is provided only
as demonstration of how GMSE can potentially be integrated with already
developed fisheries models, and is not intended to make recommendations
for management in any population.

\section{\texorpdfstring{Integrating with the Fisheries Library in R
(\href{http://www.flr-project.org/}{FLR})}{Integrating with the Fisheries Library in R (FLR)}}\label{integrating-with-the-fisheries-library-in-r-flr}

The FLR toolset includes \href{http://www.flr-project.org/\#packages}{a
series of pacakges}, with several
\href{http://www.flr-project.org/doc/index.html}{tutorials} for using
them. For simplicity, we focus here on
\href{http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html}{a
model of stock recruitment} to be used as the population model in
\texttt{gmse\_apply}. This population model will use sample data and one
of the many available stock-recruitment models available in FLR, and a
custom function will be written to return a single value for stock
recruitment. Currently, \texttt{gmse\_apply} requires that submodels
return subfunction results either as scalar values or data frames that
are structured in the same way as GMSE submodels. But interpretation of
scalar values is left up to the user (e.g., population model results
could be interpreted as abundance or biomass; manager policy could be
interpreted as cost of harvesting or as total allowable catch). For
simplicity, the observation (i.e., estimation) model will simply be the
stock reported from the population model with error, and the manager
model will be a total allowable catch calculated from the
stock-recruitment relationship that accounts for the number of fishers
in the system. The user model, however, will employ the full power of
the default GMSE user function to simulate user actions. We first show
how a custom function can be made that applies the FLR toolset to a
population model.

\section{Modelling stock-recruitment for the population
model}\label{modelling-stock-recruitment-for-the-population-model}

Here we closely follow
\href{http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html}{a
tutorial from the FLR project}. To build the stock-recruitment model,
the \texttt{FLCore} package is needed \citep{Kell2007}.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{install.packages}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"FLCore"}\NormalTok{), }\DataTypeTok{repos=}\StringTok{"http://flr-project.org/R"}\NormalTok{);}
\end{Highlighting}
\end{Shaded}

To start, we need to read in the \texttt{FLCore} and GMSE libraries.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(FLCore);}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Loading required package: lattice
\end{verbatim}

\begin{verbatim}
## FLCore (Version 2.6.7, packaged: 2018-04-17 09:12:42 UTC)
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(GMSE);}
\end{Highlighting}
\end{Shaded}

For a simplified example in GMSE, we will simulate the process of stock
recruitment over multiple time steps using an example stock-recruitment
model. The stock-recruitment model describes the relationship between
stock-recruitment and spawning stock biomass. The sample that we will
work from is a recreation of the North Sea Herring (\texttt{nsher})
dataset available in the \texttt{FLCore} package \citep{Kell2007}. This
data set includes recruitment and spawning stock biomass data between
1960 and 2004. First, we initialise an empty \texttt{FLSR} object and
read in the recreated herring data files from GMSE, which contains
recruitment (\texttt{rec.n}) and spawning stock biomass (\texttt{ssb.n})

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{newFL   <-}\StringTok{ }\KeywordTok{FLSR}\NormalTok{(); }\CommentTok{# Initialises the empty FLSR object}
\KeywordTok{data}\NormalTok{(nsher_data); }\CommentTok{#called from GMSE library rather than nsher from FLCore library}
\end{Highlighting}
\end{Shaded}

The recruitment (\texttt{rec.n}) and spawning stock biomass
(\texttt{ssb.n}) data need to be in the form of a vector, array, matrix
to use them with \texttt{FLQuant}. We will convert \texttt{rec.n} and
\texttt{ssb.n} into matrices.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{rec.m        <-}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(rec.n);}
\NormalTok{ssb.m        <-}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(ssb.n);}
\end{Highlighting}
\end{Shaded}

We can then construct two \texttt{FLQuant} objects, specifying the
relevant years and units.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{Frec.m       <-}\StringTok{ }\KeywordTok{FLQuant}\NormalTok{(rec.m, }\DataTypeTok{dimnames=}\KeywordTok{list}\NormalTok{(}\DataTypeTok{age=}\DecValTok{1}\NormalTok{, }\DataTypeTok{year =} \DecValTok{1960}\NormalTok{:}\DecValTok{2004}\NormalTok{));}
\NormalTok{Fssb.m       <-}\StringTok{ }\KeywordTok{FLQuant}\NormalTok{(ssb.m, }\DataTypeTok{dimnames=}\KeywordTok{list}\NormalTok{(}\DataTypeTok{age=}\DecValTok{1}\NormalTok{, }\DataTypeTok{year =} \DecValTok{1960}\NormalTok{:}\DecValTok{2004}\NormalTok{));}
\NormalTok{Frec.m@units <-}\StringTok{ "10^3"}\NormalTok{;}
\NormalTok{Fssb.m@units <-}\StringTok{ "t*10^3"}\NormalTok{;}
\end{Highlighting}
\end{Shaded}

We then place the recruitment and spawning stock biomass data into the
\texttt{FLSR} object that we created.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{rec}\NormalTok{(newFL)    <-}\StringTok{ }\NormalTok{Frec.m;}
\KeywordTok{ssb}\NormalTok{(newFL)    <-}\StringTok{ }\NormalTok{Fssb.m;}
\KeywordTok{range}\NormalTok{(newFL)  <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1960}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{2004}\NormalTok{);}
\end{Highlighting}
\end{Shaded}

The \texttt{FLCore} package offers several stock-recruitment models.
Here we use a Ricker model of stock recruitment \citep{Ricker1954}, and
insert this model into the \texttt{FLSR} object below.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{model}\NormalTok{(newFL) <-}\StringTok{ }\KeywordTok{ricker}\NormalTok{();}
\end{Highlighting}
\end{Shaded}

Parameters for the Ricker stock-recruitment model can be estimated with
maximum likelihood.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{newFL <-}\StringTok{ }\KeywordTok{fmle}\NormalTok{(newFL);}
\end{Highlighting}
\end{Shaded}

Diagnostic plots, identical to those of the
\href{http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html}{modelling
stock-recruitment tutorial} for the \texttt{nsher\_ri} example, are
shown below.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot}\NormalTok{(newFL);}
\end{Highlighting}
\end{Shaded}

\includegraphics{SI5_files/figure-latex/unnamed-chunk-11-1.pdf}

We now have a working example of a stock-recruitment model, but for our
integration with \texttt{gmse\_apply}, we will want a function that
automates the above to simulate the process of updating the
stock-recruitment model. We do this using the custom function created
below.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{update_SR_model <-}\StringTok{ }\NormalTok{function(rec_m, ssb_m, years)\{}
    \NormalTok{Frec_m       <-}\StringTok{ }\KeywordTok{FLQuant}\NormalTok{(rec_m, }\DataTypeTok{dimnames=}\KeywordTok{list}\NormalTok{(}\DataTypeTok{age =} \DecValTok{1}\NormalTok{, }\DataTypeTok{year =} \NormalTok{years));}
    \NormalTok{Fssb_m       <-}\StringTok{ }\KeywordTok{FLQuant}\NormalTok{(ssb_m, }\DataTypeTok{dimnames=}\KeywordTok{list}\NormalTok{(}\DataTypeTok{age =} \DecValTok{1}\NormalTok{, }\DataTypeTok{year =} \NormalTok{years));}
    \NormalTok{Frec_m@units <-}\StringTok{ "10^3"}\NormalTok{;}
    \NormalTok{Fssb_m@units <-}\StringTok{ "t*10^3"}\NormalTok{;}
    \KeywordTok{rec}\NormalTok{(newFL)   <-}\StringTok{ }\NormalTok{Frec.m;}
    \KeywordTok{ssb}\NormalTok{(newFL)   <-}\StringTok{ }\NormalTok{Fssb.m;}
    \KeywordTok{range}\NormalTok{(newFL) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, years[}\DecValTok{1}\NormalTok{], }\DecValTok{0}\NormalTok{, years[}\KeywordTok{length}\NormalTok{(years)]);}
    \KeywordTok{model}\NormalTok{(newFL) <-}\StringTok{ }\KeywordTok{ricker}\NormalTok{();}
    \NormalTok{newFL        <-}\StringTok{ }\KeywordTok{fmle}\NormalTok{(newFL);}
    \KeywordTok{return}\NormalTok{(newFL);}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

The above function will be used within another custom function to
predict the next time step of recruitment.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{predict_recruitment <-}\StringTok{ }\NormalTok{function(rec_m, ssb_m, years, new_ssb)\{}
    \NormalTok{newFL <-}\StringTok{ }\KeywordTok{update_SR_model}\NormalTok{(rec_m, ssb_m, years);}
    \NormalTok{a     <-}\StringTok{ }\KeywordTok{params}\NormalTok{(newFL)[[}\DecValTok{1}\NormalTok{]] }\CommentTok{# Extract 'a' parameter of the Ricker model}
    \NormalTok{b     <-}\StringTok{ }\KeywordTok{params}\NormalTok{(newFL)[[}\DecValTok{2}\NormalTok{]] }\CommentTok{# Extract 'b' parameter of the Ricker model}
    \NormalTok{rec   <-}\StringTok{ }\NormalTok{a *}\StringTok{ }\NormalTok{new_ssb *}\StringTok{ }\KeywordTok{exp}\NormalTok{(-b *}\StringTok{ }\NormalTok{new_ssb); }\CommentTok{# Predict the new recruitment}
    \KeywordTok{return}\NormalTok{(rec)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

In \texttt{gmse\_apply}, we will use the \texttt{predict\_recruitment}
function above as the resource (i.e., operational) model. The
\texttt{new\_ssb} reads in the new spawning stock biomass, which will be
calculated from the built-in GMSE \texttt{user} model.

\section{\texorpdfstring{Integrating \texttt{predict\_recruitment} with
\texttt{gmse\_apply}}{Integrating predict\_recruitment with gmse\_apply}}\label{integrating-predict_recruitment-with-gmse_apply}

The \href{http://www.flr-project.org/}{FLR project} includes libraries
that can be used to
\href{http://www.flr-project.org/doc/An_introduction_to_MSE_using_FLR.html}{perform
a management strategy evaluation} (MSE) under fisheries-focused
observation, manager, and user models. We will not recreate
\href{http://www.flr-project.org/doc/An_introduction_to_MSE_using_FLR.html}{this
approach}, or integrate any other submodels into GMSE as was done for
the population model above, although such integration of submodels
should be possible using similar techniques. Our goal here is to instead
show how the \texttt{predict\_recruitment} model created above can be
integrated with \texttt{gmse\_apply}, which can then make use of the
genetic algorithm to predict the behaviour fishers.

We will use a custom observation model, which will simply estimate
recruitment with some fixed error.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{obs_ssb <-}\StringTok{ }\NormalTok{function(resource_vector)\{}
    \NormalTok{obs_err <-}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(}\DataTypeTok{n =} \DecValTok{1}\NormalTok{, }\DataTypeTok{mean =} \DecValTok{0}\NormalTok{, }\DataTypeTok{sd =} \DecValTok{100}\NormalTok{);}
    \NormalTok{the_obs <-}\StringTok{ }\NormalTok{resource_vector +}\StringTok{ }\NormalTok{obs_err;}
    \KeywordTok{return}\NormalTok{(the_obs);}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

Hence, we can now feed the data from \texttt{rec.m} and \texttt{ssb.m}
through \texttt{predict\_recruitment}, which will return a value for new
recruitment, and this new value can in turn be fed into
\texttt{obs\_ssb} to predict recruitment with some error. We also need a
new spawning stock biomass \texttt{new\_ssb}, which we can just
initialise with the biomass from the last year in \texttt{ssb.m}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ssb_ini <-}\StringTok{ }\NormalTok{ssb.m[}\KeywordTok{length}\NormalTok{(ssb.m)];}
\NormalTok{new_rec <-}\StringTok{ }\KeywordTok{predict_recruitment}\NormalTok{(}\DataTypeTok{rec_m =} \NormalTok{rec.m, }\DataTypeTok{ssb_m =} \NormalTok{ssb.m, }\DataTypeTok{years =} \DecValTok{1960}\NormalTok{:}\DecValTok{2004}\NormalTok{,}
                               \DataTypeTok{new_ssb =} \NormalTok{ssb_ini);}
\NormalTok{obs_rec <-}\StringTok{ }\KeywordTok{obs_ssb}\NormalTok{(new_rec);}
\end{Highlighting}
\end{Shaded}

An initial run of these models gives values of 3835.21 for
\texttt{new\_rec} and 4025.16 for \texttt{obs\_rec}. We are now ready to
use the built-in manager and user submodels in \texttt{gmse\_apply}. We
will assume that managers attempt to keep a recruitment of 5000, and
that there are 4 independent fishers {[}stakehodlers in gmse\_apply says
10, is that a difference?{]} who attempt to maximise their catch. We
assign a user budget of \texttt{manager\_budget\ =\ 10000}, and all
other values are set to GMSE defaults. In the built-in GMSE functions,
the manager will use the estimate of recruitment based on
\texttt{obs\_rec} and use it to set the cost of harvesting
(\texttt{culling} in GMSE).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{yrspan       <-}\StringTok{ }\DecValTok{1960}\NormalTok{:}\DecValTok{2004}\NormalTok{;}
\NormalTok{rec.m        <-}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(rec.n);}
\NormalTok{ssb.m        <-}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(ssb.n);}

\NormalTok{sim <-}\StringTok{ }\KeywordTok{gmse_apply}\NormalTok{(}\DataTypeTok{res_mod =} \NormalTok{predict_recruitment, }\DataTypeTok{obs_mod =} \NormalTok{obs_ssb,}
                  \DataTypeTok{rec_m =} \NormalTok{rec.m, }\DataTypeTok{ssb_m =} \NormalTok{ssb.m, }\DataTypeTok{years =} \NormalTok{yrspan, }
                  \DataTypeTok{new_ssb =} \NormalTok{ssb_ini, }\DataTypeTok{manage_target =} \DecValTok{5000}\NormalTok{, }\DataTypeTok{stakeholders =} \DecValTok{10}\NormalTok{,}
                  \DataTypeTok{manager_budget =} \DecValTok{10000}\NormalTok{);}
\KeywordTok{print}\NormalTok{(sim);}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## $resource_results
## [1] 3835
## 
## $observation_results
## [1] 3617.956
## 
## $manager_results
##          resource_type scaring culling castration feeding help_offspring
## policy_1             1      NA     456         NA      NA             NA
## 
## $user_results
##         resource_type scaring culling castration feeding help_offspring
## Manager             1      NA       0         NA      NA             NA
## user_1              1      NA       2         NA      NA             NA
## user_2              1      NA       2         NA      NA             NA
## user_3              1      NA       2         NA      NA             NA
## user_4              1      NA       2         NA      NA             NA
## user_5              1      NA       2         NA      NA             NA
## user_6              1      NA       2         NA      NA             NA
## user_7              1      NA       2         NA      NA             NA
## user_8              1      NA       2         NA      NA             NA
## user_9              1      NA       2         NA      NA             NA
## user_10             1      NA       2         NA      NA             NA
##         tend_crops kill_crops
## Manager         NA         NA
## user_1          NA         NA
## user_2          NA         NA
## user_3          NA         NA
## user_4          NA         NA
## user_5          NA         NA
## user_6          NA         NA
## user_7          NA         NA
## user_8          NA         NA
## user_9          NA         NA
## user_10         NA         NA
\end{verbatim}

The resource and observation results above are interpreted in terms of
recruitment, while the manager results are interpreted in terms of the
cost of harvesting a unit of spawning stock biomass and the user results
are interpreted in terms of how much biomass was harvested. Note in the
run of \texttt{gmse\_apply} that the arguments for our custom resource
and observation models (\texttt{predict\_recruitment} and
\texttt{obs\_ssb}, respectively) are read directly in as arguments of
\texttt{gmse\_apply} itself. The \texttt{gmse\_apply} function will
figure out which subfunctions custom arguments should go to, then update
these arguments as needed over the course of a single run of
\texttt{gmse\_apply}.

\section{\texorpdfstring{Simulation with \texttt{gmse\_apply} over
multiple time
steps}{Simulation with gmse\_apply over multiple time steps}}\label{simulation-with-gmse_apply-over-multiple-time-steps}

We are now ready to loop the \texttt{gmse\_apply} function over multiple
time steps. To do this, we will update the \texttt{rec.m} and
\texttt{ssb.m} matrices after each time step, simulating 20 years into
the future. The population model \texttt{predict\_recruitment} will use
these data to dynamically update parameters of the Ricker model, as
might occur in an empirical fishery that is being monitored. We will use
the results from the observation model to update recruiment for the new
year in \texttt{rec.m}. For simplicity, spawning stock biomass prior to
harvest will be randomly sampled from a value in last 10 years (i.e.,
from \texttt{ssb.m} between 1994 and 2004), but more realistic models
could relate this spawning stock biomass to recruitment and
environmental variables from a prevoius year; spawning stock biomass
will be decreased after harvest based on user actions. The GMSE
initialisation and simulation is below.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# This code initialises the simulation -----------------------------------------}
\NormalTok{yrspan       <-}\StringTok{ }\DecValTok{1960}\NormalTok{:}\DecValTok{2004}\NormalTok{;}
\NormalTok{rec.m        <-}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(rec.n);}
\NormalTok{ssb.m        <-}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(ssb.n);}
\NormalTok{ssb_ini      <-}\StringTok{ }\NormalTok{ssb.m[}\KeywordTok{length}\NormalTok{(ssb.m)];}
\NormalTok{sim_old      <-}\StringTok{ }\KeywordTok{gmse_apply}\NormalTok{(}\DataTypeTok{res_mod =} \NormalTok{predict_recruitment, }\DataTypeTok{obs_mod =} \NormalTok{obs_ssb,}
                           \DataTypeTok{rec_m =} \NormalTok{rec.m, }\DataTypeTok{ssb_m =} \NormalTok{ssb.m, }\DataTypeTok{years =} \NormalTok{yrspan, }
                           \DataTypeTok{new_ssb =} \NormalTok{ssb_ini, }\DataTypeTok{manage_target =} \DecValTok{3500}\NormalTok{, }
                           \DataTypeTok{stakeholders =} \DecValTok{10}\NormalTok{, }\DataTypeTok{manager_budget =} \DecValTok{10000}\NormalTok{, }
                           \DataTypeTok{get_res =} \StringTok{"Full"}\NormalTok{);}
\CommentTok{# The code below simulates 20 time steps ---------------------------------------}
\NormalTok{sim_sum  <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DataTypeTok{data =} \OtherTok{NA}\NormalTok{, }\DataTypeTok{nrow =} \DecValTok{20}\NormalTok{, }\DataTypeTok{ncol =} \DecValTok{6}\NormalTok{); }\CommentTok{# Hold results here}
\NormalTok{for(time_step in }\DecValTok{1}\NormalTok{:}\DecValTok{20}\NormalTok{)\{}
    \CommentTok{# Update the relevant parameter values as necessary -----------------------}
    \NormalTok{rand_ssb        <-}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DataTypeTok{x =} \NormalTok{ssb.m[}\DecValTok{35}\NormalTok{:}\DecValTok{45}\NormalTok{], }\DataTypeTok{size =} \DecValTok{1}\NormalTok{);}
    \NormalTok{harvest         <-}\StringTok{ }\KeywordTok{sum}\NormalTok{(sim_old$basic_output$user_results[,}\DecValTok{3}\NormalTok{]);}
    \NormalTok{new_rec_m       <-}\StringTok{ }\KeywordTok{c}\NormalTok{(sim_old$rec_m, sim_old$observation_vector);}
    \NormalTok{new_ssb_m       <-}\StringTok{ }\KeywordTok{c}\NormalTok{(sim_old$ssb_m, rand_ssb -}\StringTok{ }\NormalTok{harvest);}
    \NormalTok{sim_old$rec_m   <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DataTypeTok{data =} \NormalTok{new_rec_m, }\DataTypeTok{nrow =} \DecValTok{1}\NormalTok{);}
    \NormalTok{sim_old$ssb_m   <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DataTypeTok{data =} \NormalTok{new_ssb_m, }\DataTypeTok{nrow =} \DecValTok{1}\NormalTok{);}
    \NormalTok{sim_old$years   <-}\StringTok{ }\KeywordTok{c}\NormalTok{(sim_old$years, time_step +}\StringTok{ }\DecValTok{2004}\NormalTok{);}
    \NormalTok{sim_old$new_ssb <-}\StringTok{ }\NormalTok{sim_old$ssb_m[}\KeywordTok{length}\NormalTok{(sim_old$ssb_m)];}
    \CommentTok{# Run a new simulation in the loop: custom functions are always specified -}
    \NormalTok{sim_new  <-}\StringTok{ }\KeywordTok{gmse_apply}\NormalTok{(}\DataTypeTok{get_res =} \StringTok{"Full"}\NormalTok{, }\DataTypeTok{old_list =} \NormalTok{sim_old,}
                           \DataTypeTok{res_mod =} \NormalTok{predict_recruitment, }\DataTypeTok{obs_mod =} \NormalTok{obs_ssb);}
    \CommentTok{# Record the results in sim_sum -------------------------------------------}
    \NormalTok{sim_sum[time_step, }\DecValTok{1}\NormalTok{] <-}\StringTok{ }\NormalTok{time_step +}\StringTok{ }\DecValTok{2004}\NormalTok{;}
    \NormalTok{sim_sum[time_step, }\DecValTok{2}\NormalTok{] <-}\StringTok{ }\NormalTok{sim_new$basic_output$resource_results[}\DecValTok{1}\NormalTok{];}
    \NormalTok{sim_sum[time_step, }\DecValTok{3}\NormalTok{] <-}\StringTok{ }\NormalTok{sim_new$basic_output$observation_results[}\DecValTok{1}\NormalTok{];}
    \NormalTok{sim_sum[time_step, }\DecValTok{4}\NormalTok{] <-}\StringTok{ }\NormalTok{sim_new$basic_output$manager_results[}\DecValTok{3}\NormalTok{];}
    \NormalTok{sim_sum[time_step, }\DecValTok{5}\NormalTok{] <-}\StringTok{ }\NormalTok{harvest;}
    \NormalTok{sim_sum[time_step, }\DecValTok{6}\NormalTok{] <-}\StringTok{ }\NormalTok{sim_new$new_ssb;}
    \CommentTok{# Redefine the old list --------------------------------------------------}
    \NormalTok{sim_old               <-}\StringTok{ }\NormalTok{sim_new;}
\NormalTok{\}}
\KeywordTok{colnames}\NormalTok{(sim_sum) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"Year"}\NormalTok{, }\StringTok{"Recruitment"}\NormalTok{, }\StringTok{"Recruit_estim"}\NormalTok{, }\StringTok{"Harvest_cost"}\NormalTok{, }
                         \StringTok{"Harvested"}\NormalTok{, }\StringTok{"SSB"}\NormalTok{);}
\KeywordTok{print}\NormalTok{(sim_sum);}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##       Year Recruitment Recruit_estim Harvest_cost Harvested      SSB
##  [1,] 2005        4491      4535.352          478        20 135.9025
##  [2,] 2006        4647      4709.264          538        20 105.2627
##  [3,] 2007        4629      4821.850          583        10 115.2627
##  [4,] 2008        4387      4255.025          594        10 145.9025
##  [5,] 2009        3035      3078.288          636        10  35.5913
##  [6,] 2010        3035      3219.448          502        10  35.5913
##  [7,] 2011        4629      4685.483          613        10 115.2627
##  [8,] 2012        4629      4614.816          473        10 115.2627
##  [9,] 2013        2399      2365.097          530        20  25.5913
## [10,] 2014        2994      2930.048          486        10  34.8673
## [11,] 2015        2347      2336.880          530        20  24.8673
## [12,] 2016        3303      3357.941          492        10  40.6133
## [13,] 2017        4336      4227.214          523        20 150.1926
## [14,] 2018        4387      4401.982          513        10 145.9025
## [15,] 2019        4328      4356.673          544        10  70.7603
## [16,] 2020        3035      2979.304          546        10  35.5913
## [17,] 2021        4339      4457.722          536        10  71.3340
## [18,] 2022        4339      4319.507          557        10  71.3340
## [19,] 2023        4339      4490.953          523        10  71.3340
## [20,] 2024        3303      3274.942          473        10  40.6133
\end{verbatim}

The above output from \texttt{sim\_sum} reports the recruitment
(resource or operational model), recruitment estimate (observation error
model), management set harvest cost (harvest control model), user
harvested numbers (implementation model) and spawning stock biomass
(SSB) simulation results. This example simulation demonstrates the
ability of GMSE to integrate with fisheries libraries such as
\href{http://www.flr-project.org/}{FLR} through \texttt{gmse\_apply}. In
addition to being a useful wrapping function for MSE submodels,
\texttt{gmse\_apply} can therefore be used to take advantage of the
genetic algorithm in the GMSE default manager and user models. This
flexibility will be retained in future versions of \texttt{gmse\_apply},
allowing custom resource and observation models that are built for
specific systems to be integrated with an increasingly complex genetic
algorithm simulating various aspects of human decision-making.

\section{Conclusions}\label{conclusions}

GMSE is a general, flexible, tool for simulating the management of
resources under situations of uncertainty and conflict. Management
Strategy Evaluation \citep{Bunnefeld2011, Punt2016}, the framework upon
which GMSE is based, had its origin in fisheries management
\citep{Polacheck1999, Smith1999, Sainsbury2000}, and here we showed one
example of how GMSE could be integrated with the core package of the
\href{http://www.flr-project.org/}{Fisheries Library in R}.

Future versions of GMSE will continue to be open-source and developed to
avoid unecessary dependencies (GMSE v.0.4.0.3 requires only base R). Key
goals including (1) providing highly general and useful default
\texttt{resource}, \texttt{observation}, \texttt{manager}, and
\texttt{user} submodels for a variety of MSE modelling tasks, (2)
keeping these submodels highly modular so that they can be developed in
isolation given standardised data structures, and (3) allowing these
modular submodels to be integrated with custom defined submodels as
flexibly as possible using \texttt{gmse\_apply}. Contributions in line
with these goals, and suggestions for new features, can be made on
\href{https://github.com/bradduthie/gmse}{GitHub}.

\bibliography{gmse.bib}

\end{document}

### 2018-05-04 notebook/ms/SI3.Rmd

@@
 
-Agents in GMSE (managers and users) are goal-oriented, and their behaviour is therefore driven to maximise a particular utility of interest such as a target density of resources (managers), or a suitable resource or landscape harvest size (users).
-This model feature allows GMSE to evaluate the actions of agents in the context of their individual objectives, and to therefore quantify the degree to which those objectives are or are not achieved.
+Agents in GMSE (managers and users) are goal-oriented, and their behaviour is therefore driven to maximise a particular utility of interest such as a target density of resources, such as animals or trees of conservation interest (managers), or a suitable resource (e.g.
+animal population size, forest) or landscape harvest size, such as bag size or timber (users).
+This model feature allows GMSE to evaluate the actions of agents (e.g.
+managers and users) in the context of their individual objectives, and to therefore quantify the degree to which those objectives are or are not achieved.
 When the actions of one party clashes with the objectives of another party, the objectives of one might be expressed at the expense of the other, causing conservation conflict [@Redpath2013].
-Currently, there is no standard way to measure conservation conflict in a social-ecological system, and previous modelling approaches have not meaningfully separated agent objectives from agent actions.
+Currently, there is no standard way to measure conservation conflict in a social-ecological system where both the natural resource (e.g.
+animals, plants) and the people (agents, stakeholders, managers, users etc) are modeled in a single system, and previous modelling approaches have not meaningfully separated agent objectives from agent actions.
 We suggest that a starting point to developing a useful metric of conservation conflict is to quantify the deviation of an individual's actions from their objectives  (i.e., of actual actions from desired actions), the former of which is restricted by the actions of other individuals.
@@
 To demonstrate how GMSE can be used to understand conflict in social-ecological systems, we build upon the [example of resource management](https://www.biorxiv.org/content/biorxiv/early/2017/11/17/221432.full.pdf#page=5) in the main text.
-We consider a protected population of waterfowl that exploits agricultural land and is therefore a source of conservation conflict [e.g., @Fox2017a; @Mason2017; @Tulloch2017; @Cusack2018].
-As in the main text example, the objective of the manager is to keep waterfowl at a target abundance, while the objective of farmers is to maximise agricultural production on their landscape.
-Here we consider a more complex simulation, with a level of detail that more accurately reflects a scenario that might occur in a real social-ecological system.
+We consider a protected population of waterfowl that exploits and damages agricultural land and is therefore a source of conservation conflict between those that seek the conservation of waterfowl and those that are concerned with the loss of agriculture [e.g., @Fox2017a; @Mason2017; @Tulloch2017; @Cusack2018].
+As in the main text example, the objective of the manager is to keep waterfowl at a target abundance that minmises extinction risks, while the objective of farmers is to maximise agricultural production on their landscape.
+Here we consider a more complex simulation, with a level of detail that more accurately reflects a scenario that might occur in a real social-ecological system where the manager sets policies that incentivise the user to act in a way that ensures the persistence of the resource.
+The policies are backed up by a manager budget that can be allocated to set costs of actions and thus incentivise users to follw with the desired actions.
+The user also has a budget to carry out actions and the cost of the actions is affected by the manager policies and budget.
 Our objective is not to model the dynamics of a specific system, but to show how GMSE could be parameterised using demographic estimates from empirical studies.
@@
 We parameterise our model using demographic information from the Taiga Bean Goose (*Anser fabalis fabalis*), a managed population that is hunted for recreation in Fennoscandinavia [@Johnson2018].
-Taiga Bean Geese can cause agricultural damage [@Johnson2018], which could potentially lead to conflict between farmers and management objectives.
+Taiga Bean Geese can cause agricultural damage [@Johnson2018], which could potentially lead to conflict between farming and management/conservation objectives.
 

### 2018-05-04 notebook/ms/SI6.Rmd
More start to the sixth SI

@@
 This function calls `gmse` multiple times and aggregates the results from replicate simulations into a single table.
+
+For a single simulation, the `gmse_table` function prints out key information from a `gmse` simulation result. The example provided in the GMSE documentation is below.
+
+```{r}
+gmse_sim  <- gmse(time_max = 10, plotting = FALSE);
+sim_table <- gmse_table(gmse_sim = gmse_sim);
+print(sim_table)
+```
+
+We can also only record the last time step in `gmse_table`.
+
+```{r}
+sim_table_last <- gmse_table(gmse_sim = gmse_sim, all_time = FALSE);
+print(sim_table_last)
+```
+
+The `gmse_replicates` function replicates multiple simulations under the same initial conditions, then returns a table showing the values of all simulations.
+This can be useful, for example, for testing how often a population is expected to go to extinction or carrying capacity under a given set of parameter values.

### 2018-05-04 notebook/ms/SI3.Rmd
Nils' edits, minor changes

@@
 Currently, there is no standard way to measure conservation conflict in a social-ecological system where both the natural resource (e.g.
-animals, plants) and the people (agents, stakeholders, managers, users etc) are modeled in a single system, and previous modelling approaches have not meaningfully separated agent objectives from agent actions.
+animals, plants) and the people (agents, stakeholders, managers, users etc.) are modelled in a single system, and previous modelling approaches have not meaningfully separated agent objectives from agent actions.
 We suggest that a starting point to developing a useful metric of conservation conflict is to quantify the deviation of an individual's actions from their objectives  (i.e., of actual actions from desired actions), the former of which is restricted by the actions of other individuals.

### 2018-05-06 notebook/ms/SI6.Rmd
Different frequencies added

@@
 This can be useful, for example, for testing how often a population is expected to go to extinction or carrying capacity under a given set of parameter values.
+First, we demonstrate the `gmse_replicates` function for simulations of 20 time steps; other parameter values are set to default values, except `plotting`, which we set to `FALSE` to avoid plotting each simulation result.
+We run 10 replicates below.
+
+```{r, echo = FALSE, messages = FALSE, cache = FALSE, comment = FALSE, results = "hide"}
+gmse_reps1 <- gmse_replicates(replicates = 10, time_max = 20, plotting = FALSE);
+print(gmse_reps1);
+```
+
+```{r, eval = FALSE}
+gmse_reps1 <- gmse_replicates(replicates = 10, time_max = 20, plotting = FALSE);
+print(gmse_reps1);
+```
+
+```{r, echo = FALSE}
+print(gmse_reps1);
+```
+
+Note from the results above that resources in all simulations persisted for 20 time steps, which means that extinction never occurred.
+We can also see that the population in all simulations never terminated at a density near the default carrying capacity of `res_death_K = 2000`, and was instead consistently near the target population size of `manage_target = 1000`.
+If we wish to define management success as having a population density near target levels after 20 time steps (perhaps interpreted as 20 years), then we might assess this population as successfully managed under the conditions of the simulation.
+We can then see what happens if managers only respond to changes in the social-ecological system with a change in policy once every two years, perhaps as a consequence of reduced funding for management or increasing demands for management attention elsewhere.
+This can be done by changing the default `manage_freq = 1` to `manage_freq = 2`.
+
+```{r, echo = FALSE, messages = FALSE, cache = FALSE, comment = FALSE, results = "hide"}
+gmse_reps2 <- gmse_replicates(replicates = 10, time_max = 20, plotting = FALSE,
+                             manage_freq = 2);
+print(gmse_reps2);
+```
+
+```{r, eval = FALSE}
+gmse_reps2 <- gmse_replicates(replicates = 10, time_max = 20, plotting = FALSE, 
+                             manage_freq = 2);
+print(gmse_reps2);
+```
+
+```{r, echo = FALSE}
+print(gmse_reps2);
+```
+
+Note that while extinction is still non-existent in these simulations, when populations are managed less frequently, they tend to be less close to the target size of 1000 after 20 generations.
+The median population size of `gmse_reps1` (management in every time step) is `r median(gmse_reps1[,2])`, with a maximum of `r max(gmse_reps1[,2])` and minimum of `r min(gmse_reps1[,2])`.
+While the median population size of `gmse_reps2` (management every two time steps) is `r median(gmse_reps2[,2])`, with a maximum of `r max(gmse_reps2[,2])` and minimum of `r min(gmse_reps2[,2])`.
+We can then see what happens when management occurs only once in every three time steps.
+
+```{r, echo = FALSE, messages = FALSE, cache = FALSE, comment = FALSE, results = "hide"}
+gmse_reps3 <- gmse_replicates(replicates = 10, time_max = 20, plotting = FALSE,
+                             manage_freq = 3);
+print(gmse_reps3);
+```
+
+```{r, eval = FALSE}
+gmse_reps3 <- gmse_replicates(replicates = 10, time_max = 20, plotting = FALSE, 
+                             manage_freq = 3);
+print(gmse_reps3);
+```
+
+```{r, echo = FALSE}
+print(gmse_reps3);
+```
+
+Given a management frequency of once every three time steps, the median population size of `gmse_reps3` (management in every time step) is `r median(gmse_reps3[,2])`, with a maximum of `r max(gmse_reps3[,2])` and minimum of `r min(gmse_reps3[,2])`.
+There are still no extinctions in the population, so below we change the management frequency to once every four time steps.
+
+```{r, echo = FALSE, messages = FALSE, cache = FALSE, comment = FALSE, results = "hide"}
+gmse_reps4 <- gmse_replicates(replicates = 10, time_max = 20, plotting = FALSE,
+                             manage_freq = 4);
+print(gmse_reps4);
+```
+
+```{r, eval = FALSE}
+gmse_reps4 <- gmse_replicates(replicates = 10, time_max = 20, plotting = FALSE, 
+                             manage_freq = 4);
+print(gmse_reps4);
+```
+
+```{r, echo = FALSE}
+print(gmse_reps4);
+```

### 2018-05-07 notebook/ms/SI6.Rmd
Complete a draft of SI6

@@
 subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting
-  Information 5)'
+  Information 6)'
 biblio-style: apalike
@@
 
-The default submodels of GMSE (`resource`, `observation`, `manager`, `user`) are individual-based (also called 'agent-based'), meaning that they model discrete individuals (resources or agents), which in GMSE are represented by individual table rows (`RESOURCES`, `AGENTS`, `OBSERVATION`) or layers of three-dimensional arrays (`COST`, `ACTION`).
+The default submodels of GMSE (`resource`, `observation`, `manager`, `user`) are individual-based (also called 'agent-based'), meaning that they model discrete individuals (resources or agents), which in GMSE are represented by individual table rows (as in `RESOURCES`, `AGENTS`, and `OBSERVATION`) or layers of three-dimensional arrays (as in `COST` and `ACTION`).
 Individual-based models (IBMs) have been a useful approach in ecology for decades [@Uchmanski1996; @Grimm1999], providing both a pragmatic tool for the mechanistic modelling of complex populations and a powerful technique for theoretical investigation.
@@
 
-Modelling complex interactions among discrete individuals mechanistically typically causes some degree of stochasticity in IBMs (in the code, this is caused by the sampling of random values, which determine probabilistically whether or not events such as birth or death occur for individuals), reflecting the uncertainty that is inherent to complex systems.
+Mechanistically modelling complex interactions among discrete individuals typically causes some degree of stochasticity in IBMs (in the code, this is caused by the sampling of random values, which determine probabilistically whether or not events such as birth or death occur for individuals), reflecting the uncertainty that is inherent to complex systems.
 We can see a simple example of this by calling `gmse_apply` under the same default conditions twice.
@@
 
-Although the second call to `gmse_apply` has identical initial conditions, because resource demographics (e.g., birth and death) and agent decision making (e.g., policy generation and user actions) is not deterministic, a slightly different result is obtained.
+Although a second call of `gmse_apply` has identical initial conditions, because resource demographics (e.g., birth and death) and agent decision making (e.g., policy generation and user actions) is not deterministic, a slightly different result is obtained below.
 
@@
 To make meaningful model inferences, it is often necessary to replicate simulations under the same initial conditions to understand the range of predicted outcomes for a particular set of parameter values.
-This can be computationally intense, but it can also lead to a more robust understanding of the range of dynamics that might be expected in a system.
+This can be computationally intense, but it can also lead to a more robust understanding of the range of dynamics that might be expected within a system.
 Additionally, when parameter values are unknown but believed to be important, replicate simulations can be applied across a range of values to understand how a particular parameter might affect system dynamics.
@@
 
-We can also only record the last time step in `gmse_table`.
+The above table can be saved as a CSV file using the `write.csv` function.
+
+```{r, eval = FALSE}
+write.csv(x= sim_table, file = "file_path/gmse_table");
+```
+
+Instead of recording all time steps in the simulation, we can instead record only the last time step in `gmse_table` using the `all_time` argument.
 
@@
 
-The `gmse_replicates` function replicates multiple simulations under the same initial conditions, then returns a table showing the values of all simulations.
-This can be useful, for example, for testing how often a population is expected to go to extinction or carrying capacity under a given set of parameter values.
-First, we demonstrate the `gmse_replicates` function for simulations of 20 time steps; other parameter values are set to default values, except `plotting`, which we set to `FALSE` to avoid plotting each simulation result.
+The `gmse_replicates` function replicates multiple simulations `replicates` times under the same initial conditions, then returns a table showing the values of all simulations.
+This can be useful, for example, for testing how frequently a population is expected to go to extinction or carrying capacity under a given set of parameter values.
+First, we demonstrate the `gmse_replicates` function for simulations of up to 20 time steps.
+The `gmse_replicates` function accepts all arguments used in `gmse`, and also all arguments of `gmse_table` (`all_time` and `hide_unused_options`) to summarise multiple `gmse` results.
+Here we use default `gmse` values in replicate simulations, except `plotting`, which we set to `FALSE` to avoid plotting each simulation result.
 We run 10 replicates below.
@@
 
-Note that while extinction is still non-existent in these simulations, when populations are managed less frequently, they tend to be less close to the target size of 1000 after 20 generations.
-The median population size of `gmse_reps1` (management in every time step) is `r median(gmse_reps1[,2])`, with a maximum of `r max(gmse_reps1[,2])` and minimum of `r min(gmse_reps1[,2])`.
-While the median population size of `gmse_reps2` (management every two time steps) is `r median(gmse_reps2[,2])`, with a maximum of `r max(gmse_reps2[,2])` and minimum of `r min(gmse_reps2[,2])`.
-We can then see what happens when management occurs only once in every three time steps.
+Note that while extinction still does not occur in these simulations, when populations are managed less frequently, they tend to be less close to the target size of 1000 after 20 generations.
+The median population size of `gmse_reps1` (management in every time step) was `r median(gmse_reps1[,2])`, with a maximum of `r max(gmse_reps1[,2])` and minimum of `r min(gmse_reps1[,2])`.
+The median population size of the newly simulated `gmse_reps2` (management every two time steps) is `r median(gmse_reps2[,2])`, with a maximum of `r max(gmse_reps2[,2])` and minimum of `r min(gmse_reps2[,2])`.
+We can now see what happens when management occurs only once in every three time steps.
 
@@
 Given a management frequency of once every three time steps, the median population size of `gmse_reps3` (management in every time step) is `r median(gmse_reps3[,2])`, with a maximum of `r max(gmse_reps3[,2])` and minimum of `r min(gmse_reps3[,2])`.
-There are still no extinctions in the population, so below we change the management frequency to once every four time steps.
+The number of extinctions observed in these replicate populations was `r sum(gmse_reps3[,1] < 20)`.
+Below we change the management frequency to once every four time steps.
 
@@
 ```
+
+Now note from the first column of `gmse_reps4` above that `r sum(gmse_reps4[,1] < 20)` populations did not persist to the 20th time step; i.e., `r sum(gmse_reps4[,2] < 20)` populations went to extinction (note that GMSE has a minimum resource population size of 5).
+This has occured because managers cannot respond quickly enough to changes in the population density, and therefore cannot increase the cost of culling to maintain target resource levels if population size starts to decrease.
+We can see the extinction risk increase even further if management only occurs once every 5 time steps.
+
+```{r, echo = FALSE, messages = FALSE, cache = FALSE, comment = FALSE, results = "hide"}
+gmse_reps5 <- gmse_replicates(replicates = 10, time_max = 20, plotting = FALSE,
+                             manage_freq = 5);
+print(gmse_reps5);
+```
+
+```{r, eval = FALSE}
+gmse_reps5 <- gmse_replicates(replicates = 10, time_max = 20, plotting = FALSE, 
+                             manage_freq = 5);
+print(gmse_reps5);
+```
+
+```{r, echo = FALSE}
+print(gmse_reps5);
+```
+
+When manager can only make policy decisions once every five time steps, extinction occurs in `r sum(gmse_reps5[,2] < 20)` out of 10 simulated populations before year 20.
+If we wanted to summarise these results, we could plot how extinction risk changes with increasing `manage_freq`.
+
+```{r, fig.height = 5}
+ext_risk1 <- sum(gmse_reps1[,2] < 20);
+ext_risk2 <- sum(gmse_reps2[,2] < 20);
+ext_risk3 <- sum(gmse_reps3[,2] < 20);
+ext_risk4 <- sum(gmse_reps4[,2] < 20);
+ext_risk5 <- sum(gmse_reps5[,2] < 20);
+y_var     <- c(ext_risk1, ext_risk2, ext_risk3, ext_risk4, ext_risk5);
+x_var     <- 1:5;
+plot(x = x_var, y = y_var, type = "b", pch = 20, lwd = 2, cex = 1.5,
+     xlab = "Management every N time steps (manage_freq)",
+     ylab = "Freq. of population extinction", cex.lab = 1.25)
+```
+
+The above plot and the simulations from which it was derived illustrates a greatly simplified example of how GMSE might be used to assess the risk of extinction in a managed population.
+A comprehensive analysis would need more than 10 replicate simulations to accurately infer extinction risk, and would require careful pararmeterisation of all sub-models and a sensitivity analysis where such parameters are unknown.
+A benefit of this approach is that it allows for the simulation of multiple different scenarios under conditions of uncertainty and stochasticity, modelling the range of outcomes that might occur within and among scenarios and facilitating the development of social-ecological theory.
+Future expansion on the complexity of individual-based default submodels of GMSE will further increase the realism of targeted case studies.

### 2018-05-07 R/gmse_summary.R
Minor change

@@ -150,7 +150,6 @@ gmse_summary <- function(gmse_results){
 #' table.
 #'
 #'@param gmse_sim The output of a `gmse` simulation.
-#'@param print_CSV Should the table be printed to a CSV file? If so, specify the filename here.
 #'@param hide_unused_options Whether or not to hide results from policy options when creating the resulting table. If `TRUE` (default), then policy and user actions that are not allowed in a simulation will not be placed as columns. If `FALSE`, then these columns will be placed with values of `NA`.
 #'@param all_time Whether or not results from each time step from the simulation should be individually placed as a row in the resulting table (`TRUE` by default). If `FALSE`, then only the last row will be placed.
 #'@return A table with one or more rows of results, each of which indicates a unique `gmse` simulation for a given time step. Columns represent key simulation including resource densities, observation estimates, policy, and user actions.

### 2018-05-07 src/observation.c
Remove some white space

@@ -491,9 +491,6 @@ void transect(double **resource_array, double *paras, int start_x, int start_y,
     }
 }
 
-
-
-
 /* =============================================================================
  * MAIN OBSERVATION FUNCTION:
  * ===========================================================================*/

### 2018-05-07 notebook/ms/SI7.Rmd
Up to the manager (cost) array

---
title: "Default GMSE data structures"
author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy
  Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes: null
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: gmse.bib
subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting
  Information 7)'
biblio-style: apalike
---

```{r, echo = FALSE}
library(GMSE);
```

The most important (default) GMSE data structures
================================================================================

The default submodels of GMSE (`resource`, `observation`, `manager`, and `user`) use a small number of default data structures to hold the information needed in simulations. While these default submodels do not necessarily need to be used in every run in GMSE (see use of [use of](SI2.pdf) `gmse_apply`), they will be used in any run of `gmse`, and in any call of `gmse_apply` that does not run with entirely custom submodels. Simulation and model inference does not require an understanding of the default data structures, but such an understanding can be especially useful when running `gmse_apply` if there is a need to extract uncommonly used information, change key simulated values (e.g., landscape properties, agent budgets, or resource movement rules as in [Supporting Infromation 4](SI4.pdf)), or build custom individual-based submodels. Here we provide a brief explanation of the following key data structures (each name below is listed as named in the output `gmse_apply` when `get_res = "Full"`).

1. [`AGENTS`](#agents)
2. [`resource_array`](#resources) (or `RESOURCES`)
3. [`observation_array`](#observation) (or `OBSERVATION`)
4. [`manager_array`](#cost) (or `COST`)
5. `user_array` (or `ACTION`)
6. `LAND`

Note that these are not the only data structures used in GMSE, but they are the only ones that can potentially be usefully modified in GMSE v0.4.0.3 (see, e.g., [Supporting Infromation 4](SI4.pdf)), so they are the ones that we focus on here. Additionally, any custom subfunction that returns an array rather than a single value should adhere to the same structure as the these defaults if any default GMSE functions are to be used in `gmse_apply`. We can investigate each data structure by running a single simulation of `gmse_apply`.

```{r}
sim <- gmse_apply(get_res = "Full");
```

The full list output of `sim` holds each structure by name (in the case where two names are used, e.g., `resource_array` and `RESOURCES`, both are identical, but the lower case `resource_array` takes precedence in case of a change). Each data structure can be examined, changed, and incorporated into a new simulation (e.g., `new_sim <- gmse_apply(old_list = sim)`).

1. <a name = "agents">AGENTS</a>
================================================================================

The `AGENTS` data structure is a two dimensional array with a fixed number of `r dim(sim$AGENTS)[2]` columns and a number of rows that is always equal to the total number of manager and users (each row is an individual agent).

```{r}
print(sim$AGENTS);
```

In the default case above, there are five agents (one manager and four user), each represented by a unique row. Columns in the array represent the agent traits listed below.

1. ID (each agent gets a unique number)
2. Type 1 (0 indicates the manager; 1 indicates users)
3. Type 2 (currently unused)
4. Type 3 (currently unused)
5. x-location on the landscape (typically ignored)
6. y-location on the landscape (typically ignored)
7. Movement distance (typically ignored)
8. Time parameter (typically ignored)
9. Distance of vision (currently used only for managers)
10. Error parameter (currently unused)
11. Resource marking parameter (currently used only for managers)
12. Resource tally parameter (currently used only for managers)
13. Unused column 1
14. Unused column 2
15. Unused column 3
16. Yield from owned land (zero for users when default `land_ownership = FALSE`)
17. Budget

It is obvious from the above list that most columns represent traits that are either typically ignored or currently not in use. This is intended to allow for easier future development of default model options and potential customisation of submodels in `gmse_apply`. We anticipate that future versions of GMSE will contain multiple user types with unique traits and among-user interactions.

2. <a name = "resources">resource_array</a>
================================================================================

The `resource_array` (also accessible as `RESOURCES`) is a two dimensional array with a fixed number of `r dim(sim$resource_array)[2]` columns and a number of rows that is always equal to the total number of resources (each row is an individual resource). In the above simulation, `sim$resource_array` includes `r dim(sim$resource_array)[1]` rows, so we only print out the first eight for illustration.

```{r}
print(sim$resource_array[1:8,]);
```

Columns in the resource array represent the individual resource traits listed below.

1. ID (each resource gets a unique number)
2. Type 1 (currently all resources are of type 1)
3. Type 2 (currently unused)
4. Type 3 (currently unused)
5. x-location on the landscape
6. y-location on the landscape
7. Movement distance
8. Time parameter (typically ignored)
9. Removal (i.e., death) probability
10. Growth (i.e., birth) probability
11. Offspring produced
12. Age (initial resources are given a random age between 1 and the maximum age sampled from a uniform distribution; offspring always start at age zero in their time step of birth)
13. Marking indicator (used in the observation function)
14. Tallying indicator (used in the observation function)
15. Proportion of a landscape cell the resource consumes in a time step
16. Has the resource been scared by an agent?
17. Has the resource been culled by an agent?
18. Has the resource been castrated by an agent?
19. Has the resource's growth rath been increased by an agent?
20. Has the resource's offspring production been increased by an agent?

In the case of columns 16-20, the value is either zero (if no action has occurred), or some positive integer that matches the ID of the agent that has performed the act (e.g., if column 17 equals 3, then that means that the agent with ID 3 culled the resource in the corresponding row; where more than one agent's action is possible per time step -- as in scaring -- the integer reflects the most recently acting agent). We anticipate that future versions of gmse will contain multiple resource types, and might add rows to include additional resource traits.

3. <a name = "observation">observation_array</a>
================================================================================

The `observation_array` (also accessible as `OBSERVATION`) is a two dimensional array, the number of rows and columns of which depend on the type of observation being made (i.e., `observe_type`, which can take integer values from 0-3; see the [GMSE reference manual](https://cran.r-project.org/package=GMSE) for more information about built-in observation types that are available in GMSE). The first 20 columns of `observation_array` contain the same individual resource traits as in `resource_array`, while any additional columns provide information about how and when a resource was observed. The number of rows in `observation_array` is always equal to or less than that of `resource_array`; each resource that is observed at least once is placed into one unique row, while unobserved resources are not included as rows in the `observation_array`. In `sim`, there are `r dim(sim$observation_array)[1]` rows, meaning that `r dim(sim$resource_array)[1] - dim(sim$observation_array)[1]` were not observed at all in this time step. Below we print out the first eight rows of the observation array.

```{r}
print(sim$observation_array[1:8,]);
```

In the case of the default parameters, the observation array has only two additional columns; the first added column 21 is currently unused, and all values in this column are zero. The second added column 22 contains a value of 1 confirming that the resource was observed. Additional options will add different numbers of columns with different values. For example, when `observe_type = 0` (managers observe all resources on a random subset of the landscape, the size of which is determined by their distance of vision) but `times_observe > 1`, managers sample more than one random subset of the landscape. A new column is added for each sampled subset, and a 1 is placed in the relevant column if the resource is observed (these collected data are then used to estimate population size). This process simulates the data collection of resources (and potentially resource trait measurements) as might be performed by observers within the system. It therefore takes a virtual ecologist approach; this enables the integration of theory and empirical work and can improve the mechanistic understanding of social-ecological systems [@Zurell2010].

4. <a name = "cost">manager_array</a>
================================================================================

### 2018-05-07 notebook/ms/SI7.Rmd
More on cost array

@@
 4. [`manager_array`](#cost) (or `COST`)
-5. `user_array` (or `ACTION`)
+5. [`user_array`](#action) (or `ACTION`)
 6. `LAND`
@@
 A new column is added for each sampled subset, and a 1 is placed in the relevant column if the resource is observed (these collected data are then used to estimate population size).
+An example where `times_observe = 4` is shown below.
+
+```{r}
+sim_t0_4 <- gmse_apply(get_res = "Full", times_observe = 4);
+print(sim_t0_4$observation_array[1:8,]);
+```
+
 This process simulates the data collection of resources (and potentially resource trait measurements) as might be performed by observers within the system.
@@
 ================================================================================
+
+For context, it might be easier to understand `manager_array` after reading about `user_array` [below](#action).
+The `manager_array` (also accessible as `COST`) is a three dimensional array, each layer of which corresponds to a unique agent (rows of agents correpond to layers of `manager_array`).
+Hence, in the simulation output `sim$manager_array`, there are `r dim(sim$manager_array)[3]` layers.
+Each layer in `manager_array` has `r dim(sim$manager_array)[2]` columns, and a number of rows that varies depending on the number of agents and resource types.
+In GMSE v0.4.0.3, only the first four rows are used.
+Two layers of `sim$manager_array` are shown below, the first being that of the manager and the second being that of the first user.
+
+```{r}
+print(sim$manager_array[,,1:2]);
+```
+
+Each element in the array indicates the cost of performing a particular action.
+The minimum value in `sim$manager_array` is therefore `r min(sim$manager_array)`, reflecting the default `minimum_cost` value of `r sim$GMSE$minimum_cost`.
+
+5. <a name = "action">user_array</a>
+================================================================================

### 2018-05-08 notebook/ms/SI7.Rmd
More on actions

@@
 While these default submodels do not necessarily need to be used in every run in GMSE (see use of [use of](SI2.pdf) `gmse_apply`), they will be used in any run of `gmse`, and in any call of `gmse_apply` that does not run with entirely custom submodels.
-Simulation and model inference does not require an understanding of the default data structures, but such an understanding can be especially useful when running `gmse_apply` if there is a need to extract uncommonly used information, change key simulated values (e.g., landscape properties, agent budgets, or resource movement rules as in [Supporting Infromation 4](SI4.pdf)), or build custom individual-based submodels.
+Simulation and model inference does not require an understanding of the default data structures, but such an understanding can be especially useful when running `gmse_apply` if there is a need to extract uncommonly used information, change key simulated values (e.g., landscape properties, agent budgets, or resource movement rules as in [Supporting Information 4](SI4.pdf)), or build custom individual-based submodels.
 Here we provide a brief explanation of the following key data structures (each name below is listed as named in the output `gmse_apply` when `get_res = "Full"`).
@@
 
-The full list output of `sim` holds each structure by name (in the case where two names are used, e.g., `resource_array` and `RESOURCES`, both are identical, but the lower case `resource_array` takes precedence in case of a change).
+The full list output of `sim` holds each structure by name (in the case where two names are used, e.g., `resource_array` and `RESOURCES`, both are identical, but the lower case `resohttps://twitter.com/existentialcoms/status/993513186547548160urce_array` takes precedence in case of a change).
 Each data structure can be examined, changed, and incorporated into a new simulation (e.g., `new_sim <- gmse_apply(old_list = sim)`).
@@
 Each layer in `manager_array` has `r dim(sim$manager_array)[2]` columns, and a number of rows that varies depending on the number of agents and resource types.
-In GMSE v0.4.0.3, only the first four rows are used.
+As of GMSE v0.4.0.3, only the first three rows are used.
 Two layers of `sim$manager_array` are shown below, the first being that of the manager and the second being that of the first user.
@@
 Each element in the array indicates the cost of performing a particular action.
+In the code, this is the cost of changing an element in `user_array` (which has the same dimensions as `manager_array`).
 The minimum value in `sim$manager_array` is therefore `r min(sim$manager_array)`, reflecting the default `minimum_cost` value of `r sim$GMSE$minimum_cost`.
+The maximum value is 100001, which is one higher than the maximum possible manager or user budget.
+Where cost is 100001, actions can therefore never be performed.
+An explanation of the rows and columns of `manager_array` is provided [below](#action) in the description of `user_array`.
 
@@
 ================================================================================
+
+When considering the three dimensional `user_array` (also accessible as `ACTION`), it is helpful to keep in mind that each layer corresponds to the actions of a particular agent, that each column corresponds to a particular type of action, and that each row corresponds to a particular resource, agent, or group that the action will affect.
+The cost of performing any action in this array is held in `manager_array`, wherein an action's cost in `manager_array` is held in the same array element as the action itself in `user_array`.
+Recall from the [manager array](#cost) that the first layer of `user_array` corresponds to the manager actions, and that remaining layers correspond to user actions; there are therefore as many layers in `user_array` as there are agents in the model, and each row of [`AGENTS`](#agents) corresponds to equivalent layer of `user_array` (e.g., the manager agent, ID = 1, is in the first row of `AGENTS` and the first layer of `user_array`).
+The first two layers of `user_array` are shown below.
+
+```{r}
+print(sim$user_array[,,1:2]);
+```
+
+Note that there are more columns in this array than there are possible actions in GMSE.
+This is because there are several actions that are not actions per se, but properties of agents.
+As of GMSE v0.4.0.3, these properties cannot be changed by other agents.
+Column of `user_array` are as follows.
+
+1.
+The type of agent or resource being affected by an action.
+A value of -2 indicates that actions have a direct effect on a resource (e.g., scaring, culling, etc.).
+A value of -1 indicates that actions have a direct effect on a landscape layer.
+Positive integer values indicate actions that affect other agents, where each integer corresponds to the agents' IDs.
+Where the integer value is identical with the agent's own ID (e.g., row 3 in layer 1 where the element `sim$user_array[3, 1, 1] = 1`), actions affect all other agents in the model.
+As of GMSE v0.4.0.3, all rows except 1-3 are unused because agents do not affect one anothers actions individually; they either affect all other agents' actions indiscriminately (in the case of the manager setting policy) or do not affect other agents' actions at all (in the case of users).
+This data structure, however, is designed so that future versions of GMSE will allow users to affect one another directly (representing, e.g., different groups of agents lobbying for different interests, among-user conflict, etc.).
+2. Type 1 of the agent or resource of interest (in practice, this is currently unused).
+3. Type 2 of the agent or resource of interest (currently unused).
+4. Type 3 of the agent or resource of interest (currently unused).
+5.
+Utility associated with the recipient of the action.
+For example, in the case of the resource (row 1), positive values indicate that the agent wants more of these resources, while negative values indicate that the agent wants fewer.
+In the case of the manager (layer 1), the value in the first row is `manage_target`, while the value in the third row is the change in resource number needed to achieve target value (i.e., `manage_target = ` `r sim$GMSE$manage_target`, and the manager's estimate is `sim$observation_vector = ` `r sim$observation_vector`.
+The former minus the latter is `r round(sim$GMSE$manage_target - sim$observation_vector, digits = 5)`).
+6.
+Whether or not the utility associated with the recipient of the action is dependent upon that recipient being on land owned by the actor (e.g., if users only care about resources on landscape cells that they own, then this value is 1 instead of 0).
+7.
+Whether or not actions on the recipient are possible if the recipient is not on land owned by the actor (e.g., if users cannot cull resources that are not on their own land, then this value is 1 instead of 0).

### 2018-05-08 src/game.c
Remove unused variable

@@ -500,7 +500,7 @@ void policy_to_counts(double ***population, double **merged_acts, int agent,
                       int action_row, int manager_row, double *paras){
     
     int col, COLS;
-    double old_cost, new_cost, old_act, cost_change, new_action;
+    double old_cost, new_cost, old_act, new_action;
     
     COLS   = (int) paras[69];
     

### 2018-05-08 notebook/ms/SI7.Rmd
Draft complete

@@
 The default submodels of GMSE (`resource`, `observation`, `manager`, and `user`) use a small number of default data structures to hold the information needed in simulations.
-While these default submodels do not necessarily need to be used in every run in GMSE (see use of [use of](SI2.pdf) `gmse_apply`), they will be used in any run of `gmse`, and in any call of `gmse_apply` that does not run with entirely custom submodels.
-Simulation and model inference does not require an understanding of the default data structures, but such an understanding can be especially useful when running `gmse_apply` if there is a need to extract uncommonly used information, change key simulated values (e.g., landscape properties, agent budgets, or resource movement rules as in [Supporting Information 4](SI4.pdf)), or build custom individual-based submodels.
-Here we provide a brief explanation of the following key data structures (each name below is listed as named in the output `gmse_apply` when `get_res = "Full"`).
+While these default submodels do not necessarily need to be used in every run in GMSE (see [use of](SI2.pdf) `gmse_apply`), they will be used in any run of the `gmse` function, and in any call of the `gmse_apply` function that does not run with entirely custom submodels.
+Simulation and model inference do not require an understanding of the default data structures, but such an understanding can be especially useful when running `gmse_apply` if there is a need to extract uncommonly used information, change key simulated values (e.g., landscape properties, agent budgets, or resource movement rules as in [Supporting Information 4](SI4.pdf)), or build custom individual-based submodels.
+Here we provide a brief explanation of the following key data structures (each name below is listed as it is named in the output `gmse_apply` when `get_res = "Full"`).
 
@@
 5. [`user_array`](#action) (or `ACTION`)
-6. `LAND`
+6. [`LAND`](#land)
 
-Note that these are not the only data structures used in GMSE, but they are the only ones that can potentially be usefully modified in GMSE v0.4.0.3 (see, e.g., [Supporting Infromation 4](SI4.pdf)), so they are the ones that we focus on here.
+Note that these are not the only data structures used in GMSE, but they are the only ones that can be easily modified in GMSE v0.4.0.3 (see, e.g., [Supporting Infromation 4](SI4.pdf)), so they are the ones that we focus on here.
 Additionally, any custom subfunction that returns an array rather than a single value should adhere to the same structure as the these defaults if any default GMSE functions are to be used in `gmse_apply`.
@@
 
-The full list output of `sim` holds each structure by name (in the case where two names are used, e.g., `resource_array` and `RESOURCES`, both are identical, but the lower case `resohttps://twitter.com/existentialcoms/status/993513186547548160urce_array` takes precedence in case of a change).
+The full list output of `sim` holds each structure by name (in the case where two names are used, e.g., `resource_array` and `RESOURCES`, both are identical, but the lower case `resource_array` takes precedence in case of a change).
 Each data structure can be examined, changed, and incorporated into a new simulation (e.g., `new_sim <- gmse_apply(old_list = sim)`).
 
-1. <a name = "agents">AGENTS</a>
+1. AGENTS {#agents}
 ================================================================================
 
-The `AGENTS` data structure is a two dimensional array with a fixed number of `r dim(sim$AGENTS)[2]` columns and a number of rows that is always equal to the total number of manager and users (each row is an individual agent).
+The `AGENTS` data structure is a two dimensional array with a fixed number of `r dim(sim$AGENTS)[2]` columns and a number of rows that is always equal to the sum of the number of manager and users (i.e., each row is an individual agent).
 
@@
 
-In the default case above, there are five agents (one manager and four user), each represented by a unique row. Columns in the array represent the agent traits listed below.
+In the default case above, there are five agents (one manager and four users), each represented by a unique row. Columns in the array represent the agent traits listed below.
 
@@
 
-2. <a name = "resources">resource_array</a>
+2. resource_array {#resources}
 ================================================================================
@@
 8. Time parameter (typically ignored)
-9. Removal (i.e., death) probability
+9. Density-independent removal (i.e., death) probability
 10. Growth (i.e., birth) probability
@@
 18. Has the resource been castrated by an agent?
-19. Has the resource's growth rath been increased by an agent?
+19. Has the resource's growth rate been increased by an agent?
 20. Has the resource's offspring production been increased by an agent?
 
-In the case of columns 16-20, the value is either zero (if no action has occurred), or some positive integer that matches the ID of the agent that has performed the act (e.g., if column 17 equals 3, then that means that the agent with ID 3 culled the resource in the corresponding row; where more than one agent's action is possible per time step -- as in scaring -- the integer reflects the most recently acting agent).
+In the case of columns 16-20, the value is either zero (if no action has occurred), or some positive integer that matches the ID of the agent that has performed the act (e.g., if column 17 equals 3, then that means that the agent with ID = 3 culled the resource in the corresponding row; where more than one agent's action is possible per time step -- as in scaring -- the integer reflects the most recently acting agent).
 We anticipate that future versions of gmse will contain multiple resource types, and might add rows to include additional resource traits.
 
-3. <a name = "observation">observation_array</a>
+3. observation_array {#observation}
 ================================================================================
@@
 The number of rows in `observation_array` is always equal to or less than that of `resource_array`; each resource that is observed at least once is placed into one unique row, while unobserved resources are not included as rows in the `observation_array`.
-In `sim`, there are `r dim(sim$observation_array)[1]` rows, meaning that `r dim(sim$resource_array)[1] - dim(sim$observation_array)[1]` were not observed at all in this time step.
-Below we print out the first eight rows of the observation array.
+In `sim`, there are `r dim(sim$observation_array)[1]` rows, meaning that `r dim(sim$resource_array)[1] - dim(sim$observation_array)[1]` resources were not observed at all in this time step.
+Below, we print out the first eight rows of the observation array.
 
@@
 
-This process simulates the data collection of resources (and potentially resource trait measurements) as might be performed by observers within the system.
+This process simulates the data collection of resources (and potentially resource trait measurement) as might be performed by observers within the system.
 It therefore takes a virtual ecologist approach; this enables the integration of theory and empirical work and can improve the mechanistic understanding of social-ecological systems [@Zurell2010].
 
-4. <a name = "cost">manager_array</a>
+4. manager_array {#cost}
 ================================================================================
@@
 For context, it might be easier to understand `manager_array` after reading about `user_array` [below](#action).
-The `manager_array` (also accessible as `COST`) is a three dimensional array, each layer of which corresponds to a unique agent (rows of agents correpond to layers of `manager_array`).
+The `manager_array` (also accessible as `COST`) is a three dimensional array, each layer of which corresponds to a unique agent (rows in [`AGENT`](#agents) correpond to layers in `manager_array`).
 Hence, in the simulation output `sim$manager_array`, there are `r dim(sim$manager_array)[3]` layers.
@@
 Each element in the array indicates the cost of performing a particular action.
-In the code, this is the cost of changing an element in `user_array` (which has the same dimensions as `manager_array`).
+In the code, this is the cost of changing an element in [`user_array`](#action) (which has the same dimensions as `manager_array`).
 The minimum value in `sim$manager_array` is therefore `r min(sim$manager_array)`, reflecting the default `minimum_cost` value of `r sim$GMSE$minimum_cost`.
-The maximum value is 100001, which is one higher than the maximum possible manager or user budget.
-Where cost is 100001, actions can therefore never be performed.
+The maximum value is 100001, which is one higher than the maximum allowed manager or user budget.
+Where a cost is 100001, actions can therefore never be performed.
 An explanation of the rows and columns of `manager_array` is provided [below](#action) in the description of `user_array`.
 
-5. <a name = "action">user_array</a>
+5. user_array {#action}
 ================================================================================
 
-When considering the three dimensional `user_array` (also accessible as `ACTION`), it is helpful to keep in mind that each layer corresponds to the actions of a particular agent, that each column corresponds to a particular type of action, and that each row corresponds to a particular resource, agent, or group that the action will affect.
-The cost of performing any action in this array is held in `manager_array`, wherein an action's cost in `manager_array` is held in the same array element as the action itself in `user_array`.
+The `user_array` (also accessible as `ACTION`) is a three dimensional array, each layer of which corresponds to a unique agent.
+When considering the three dimensional `user_array`, it is helpful to keep in mind that each layer corresponds to the actions of a particular agent, that each column corresponds to a particular type of action, and that each row corresponds to a particular resource, agent, or group that the action will affect.
+The cost of performing any action in this array is held in [`manager_array`](#cost), wherein an action's cost in `manager_array` is held in the same array element as the action itself in `user_array`.
 Recall from the [manager array](#cost) that the first layer of `user_array` corresponds to the manager actions, and that remaining layers correspond to user actions; there are therefore as many layers in `user_array` as there are agents in the model, and each row of [`AGENTS`](#agents) corresponds to equivalent layer of `user_array` (e.g., the manager agent, ID = 1, is in the first row of `AGENTS` and the first layer of `user_array`).
@@
 Note that there are more columns in this array than there are possible actions in GMSE.
-This is because there are several actions that are not actions per se, but properties of agents.
+This is because there are several columns that do not map to actions per se, but properties of agents.
 As of GMSE v0.4.0.3, these properties cannot be changed by other agents.
@@
 Where the integer value is identical with the agent's own ID (e.g., row 3 in layer 1 where the element `sim$user_array[3, 1, 1] = 1`), actions affect all other agents in the model.
-As of GMSE v0.4.0.3, all rows except 1-3 are unused because agents do not affect one anothers actions individually; they either affect all other agents' actions indiscriminately (in the case of the manager setting policy) or do not affect other agents' actions at all (in the case of users).
+As of GMSE v0.4.0.3, all rows except 1-3 are unused because agents do not affect one anothers actions individually; they either affect all other agents' actions indiscriminately (in the case of the manager setting policy) or do not (directly) affect other agents' actions at all (in the case of users).
 This data structure, however, is designed so that future versions of GMSE will allow users to affect one another directly (representing, e.g., different groups of agents lobbying for different interests, among-user conflict, etc.).
@@
 For example, in the case of the resource (row 1), positive values indicate that the agent wants more of these resources, while negative values indicate that the agent wants fewer.
-In the case of the manager (layer 1), the value in the first row is `manage_target`, while the value in the third row is the change in resource number needed to achieve target value (i.e., `manage_target = ` `r sim$GMSE$manage_target`, and the manager's estimate is `sim$observation_vector = ` `r sim$observation_vector`.
+In the case of the manager (layer 1), the value in the first row equals `manage_target`, while the value in the third row is the change in resource number needed to achieve the target value (i.e., `manage_target = ` `r sim$GMSE$manage_target`, and the manager's estimate is `sim$observation_vector = ` `r sim$observation_vector`.
 The former minus the latter is `r round(sim$GMSE$manage_target - sim$observation_vector, digits = 5)`).
@@
 Whether or not actions on the recipient are possible if the recipient is not on land owned by the actor (e.g., if users cannot cull resources that are not on their own land, then this value is 1 instead of 0).
+8. The number of actions performed for scaring, which in row 3 of the manager layer 1 is interpreted as the scaring cost set by the manager for users.
+9. The number of actions performed for culling, which in row 3 of the manager layer 1 is interpreted as the culling cost set by the manager for users.
+10.
+The number of actions performed for castration, which in row 3 of the manager's layer 1 is interpreted as the castration cost set by the manager for users.
+Further, in row 2 for users (where column 1 equals -1), this value is instead the number of `tend_crop` actions (the number of cells on which crops are tended by users, which always is perfomed on users' own land, cannot be affected by the manager, and always equals `minimum_cost`).
+11.
+The number of actions performed for feeding resources (increasing their growth rate, `lambda`), which in row 3 of the manager's layer 1 is interpreted as the feeding cost set by the manager for users.
+Further, in row 2 for users (where column 1 equals -1), this value is instead the number of `kill_crop` actions (the number of cells on which crops are destroyed by users, which always is perfomed on users' own land, cannot be affected by the manager, and always equals `minimum_cost`)
+12.
+The number of actions performed for helping resource offspring (directly increasing offspring production), which in row 3 of the manager's layer 1 is interpreted as the helping offspring cost set by the manager for users.
+13.
+The number of actions unspent by the user or manager; any actions allocated to this row do nothing.
+These may be used when any action would lead the agent to a less than desirable outcome, such as if only culling exists as a policy option (default), but managers do not want to increase the cost of culling because resource density is above `manage_target`.
+
+In the [genetic algorithm](SI1.pdf), values in elements of a `user_array` layer are potentially modified according to each agent's objective, as constrained by costs in `manager_array`.
+
+6. LAND {#land}
+================================================================================
+
+Events in default GMSE submodels occur on a spatially-explicit landscape `LAND`, which is stored as a three dimensional array.
+The size of this landscape is specified with the `land_dim_1` and `land_dim_2` arguments of GMSE, which determine the length, in cells, of the y and x dimensions of the landscape, respectively (e.g., if `land_dim_1 = 10` and `land_dim_2 = 1000`, then the landscape will be one very long horizontal transect).
+The total number of landscape cells on which resources and agents can interact is therefore the product of `land_dim_1` and `land_dim_2`.
+In addition, all landscapes have three layers, which hold three separate values of information for each x-y location.
+The first layer is unused in GMSE v0.4.0.3; the second layer holds crop production on a cell, and the third layer holds the owner of the cell (corresponding to the ID of an agent, where the manager's ID = 0 defines public land).
+An $8 \times 8$ portion of the landscape from `sim` is shown below.
+
+```{r}
+print(sim$LAND[1:8,1:8,]);
+```
+
+In the case of the above, all of the cells in this square patch of landscape are owned by agent 1 (i.e., the manager; see `sim$LAND[,,3]`), and we can see that crop production on this patch of land has been decreased from 1 in several cells as a consequence of consumption by resources.
+In [Supporting Information 4](SI4.pdf), we show how landscape cell values can be manipulated to customise the placement of land ownership.
+
+Conclusions
+================================================================================
+
+We have focused on the data structures [`AGENTS`](#agents), [`resource_array`](#resources), [`observation_array`](#observation), [`manager_array`](#cost), [`user_array`](#action), and [`LAND`](#land) because these are the data structures that can be most readily manipulated to customise GMSE simulations.
+An example of how to do this within a loop using `gmse_apply` can be found in [Supporting Information 4](SI4.pdf).
+While other data structures exist within GMSE (e.g., see the output of `gmse_apply` when `get_res = Full`), we do not recommend manipulating these structures for custom simulations.
+
+Many data structures contain elements that are unused in GMSE v0.4.0.3, and in all cases this is designed for ease of ongoing development of new GMSE features.
+Requests for new features can be made on GitHub using the [GMSE Wiki](https://github.com/bradduthie/gmse/wiki) or the [GMSE Issues](https://github.com/bradduthie/gmse/issues) page.

### 2018-05-08 notebook/ms/SI5.Rmd
Spelling corrections pointed out by @ninbrm

@@
 The FLR toolset includes [a series of pacakges](http://www.flr-project.org/#packages), with several [tutorials](http://www.flr-project.org/doc/index.html) for using them.
-For simplicity, we focus here on [a model of stock recruitment](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html) to be used as the population model in `gmse_apply`.
+For simplicity, we focus on [a model of stock recruitment](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html) to be used as the population model in `gmse_apply`.
 This population model will use sample data and one of the many available stock-recruitment models available in FLR, and a custom function will be written to return a single value for stock recruitment.
-Currently, `gmse_apply` requires that submodels return subfunction results either as scalar values or data frames that are structured in the same way as GMSE submodels.
+Currently, `gmse_apply` requires that submodels return subfunction results either as scalar values or data frames that are structured [in the same way](SI7.pdf) as GMSE submodels.
 But interpretation of scalar values is left up to the user (e.g., population model results could be interpreted as abundance or biomass; manager policy could be interpreted as cost of harvesting or as total allowable catch).
-For simplicity, the observation (i.e., estimation) model will simply be the stock reported from the population model with error, and the manager model will be a total allowable catch calculated from the stock-recruitment relationship that accounts for the number of fishers in the system.
+For simplicity, the observation (i.e., estimation) model will be the stock reported from the population model with error, and the manager model will be a total allowable catch calculated from the stock-recruitment relationship that accounts for the number of fishers in the system.
 The user model, however, will employ the full power of the default GMSE user function to simulate user actions.
@@
 
-```{r, fig.height = 7}
+```{r, fig.height = 6.2}
 plot(newFL);
@@
 We will not recreate [this approach](http://www.flr-project.org/doc/An_introduction_to_MSE_using_FLR.html), or integrate any other submodels into GMSE as was done for the population model above, although such integration of submodels should be possible using similar techniques.
-Our goal here is to instead show how the `predict_recruitment` model created above can be integrated with `gmse_apply`, which can then make use of the genetic algorithm to predict the behaviour fishers.
+Our goal here is to instead show how the `predict_recruitment` model created above can be integrated with `gmse_apply`, which can then make use of the genetic algorithm to predict the fishers' behaviour.
 
@@
 We are now ready to use the built-in manager and user submodels in `gmse_apply`.
-We will assume that managers attempt to keep a recruitment of 5000, and that there are 4 independent fishers [stakehodlers in gmse_apply says 10, is that a difference?] who attempt to maximise their catch.
+We will assume that managers attempt to keep a recruitment of 5000, and that there are 4 independent fishers [stakeholders in gmse_apply says 10, is that a difference?] who attempt to maximise their catch.
 We assign a user budget of `manager_budget = 10000`, and all other values are set to GMSE defaults.
@@
 The population model `predict_recruitment` will use these data to dynamically update parameters of the Ricker model, as might occur in an empirical fishery that is being monitored.
-We will use the results from the observation model to update recruiment for the new year in `rec.m`.
+We will use the results from the observation model to update recruitment for the new year in `rec.m`.
 For simplicity, spawning stock biomass prior to harvest will be randomly sampled from a value in last 10 years (i.e., from `ssb.m` between 1994 and 2004), but more realistic models could relate this spawning stock biomass to recruitment and environmental variables from a prevoius year; spawning stock biomass will be decreased after harvest based on user actions.

### 2018-05-09 notebook/ms/ms.Rmd
Update author list
Small changes with new reference included

@@
 2.
-We present the R package GMSE, which generalises MSE using a game-theoretic approach, applying genetic algorithms to simulate adaptive decision-making management scenarios between stakeholders with competing objectives under complex social-ecological interactions and uncertainty.
+We present the R package GMSE, which applies genetic algorithms to provide a genearalised tool for simulating adaptive decision-making management scenarios between stakeholders with competing objectives under complex social-ecological interactions and uncertainty.
 3.
@@
 This approach is especially effective when considering multiple aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of stakeholders [@Bunnefeld2011; @Bunnefeld2014].
-Management strategy evaluation (MSE) is a modelling framework, first developed in fisheries, for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016].
+Management strategy evaluation (MSE) is a modelling framework, first developed in fisheries [e.g., @Sainsbury2000; @Polacheck1999; @Smith1999; @Moore2013], for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016].
 Nevertheless, MSE models developed hitherto have been limited in their ability to model human decision-making [@Fulton2011; @Dichmont2017]; manager decisions are typically based on fixed rules, and user behaviour likewise remains fixed over time instead of dynamically responding to changing resource availability and management decisions [@Schluter2012; @Melbourne-Thomas2017].
@@
 The R package 'GMSE' is a flexible modelling tool to simulate key aspects of natural resource management.
-GMSE offers a range of parameters to simulate resource dynamics and management policy options, and includes genetic algorithms to dynamically model stakeholder (manager and user) decision-making.
+GMSE offers a range of parameters to simulate resource dynamics (primarily, but not necessarily biological populations) and management policy options, and includes genetic algorithms to dynamically model stakeholder (manager and user) decision-making.
 Genetic algorithms find adaptive solutions to any simulated conditions given stakeholder-specific goals, allowing GMSE to model scenarios of conservation conflict.
@@
 In each time step, waterfowl are observed on a subset of cells, then managers extrapolate from density per cell to estimate total population size.
-Managers then use these estimates to set costs of culling and scaring (non-lethal) waterfowl for five farmers.
+Managers then use these estimates to set costs of culling and scaring waterfowl for five farmers (scaring is non-lethal, causing waterfowl to move to a random landscape cell).
 Farmers attempt to reduce the negative impact of waterfowl on the cropland that they own, working within the constraints of culling and scaring costs and their budget for performing these actions.

### 2018-05-09 R/gmse.R
Fix feeding

@@ -350,7 +350,7 @@ gmse <- function( time_max       = 100,   # Max number of time steps in sim
                0,       # 74. Manager's projected change if resource moved
                -1*mas,        # 75. Manager's proj change if res killed
                -1*mas*lambda, # 76. Manager's proj change if res castrated
-               1*lambda,      # 77. Manager's proj change if reso growth +
+               1*mas*lambda,  # 77. Manager's proj change if reso growth +
                1*mas,   # 78. Manager's projected change if resource offspring +
                tcy,     # 79. User's improvement of land (proportion)
                1,       # 80. Landscape layer on which crop yield is located

### 2018-05-09 notebook/ms/SI1.Rmd
Reviewer 1 concerns regarding SI1

@@
 For each action, no consequence is predicted if the policy is not allowed by a simulation of GMSE (e.g., `culling = FALSE`).
-For allowed actions, the parameter `manager_sense` ($\sigma$) modulates predicted consequences for abundance by some factor; this is useful because not all actions attempted by users will be realised, and a value of $\sigma = 1$ tends to greatly overestimate how much the actions attempted by users will actually translate to a change in resource abundance.
-In practice, the default $\sigma = 0.1$ performs well.
-Allowed actions are predicted by managers to have the following effects:
+For allowed actions, the parameter `manager_sense` ($\sigma$) modulates predicted consequences for abundance by some factor; this is useful because not all actions attempted by users will be realised, and a value of $\sigma = 1$ tends to slightly overestimate how much the actions attempted by users will actually translate to a change in resource abundance.
+In practice, the default $\sigma = 0.9$ performs well.
+Allowed actions are predicted by managers to have the following effects (again, we emphasise that whether or not these effects are realised will depend later on the user model, to which the manager does not have access):
 
-- `scaring` is assumed to be nonlethal and therefore have no effect on resource number.
-- `culling` decreases resource number by $\sigma(1 + \lambda)$, where $\lambda$ is the GMSE argument `lambda` that defines the baseline population growth rate of resources.
-- `castration` decreases resource number by $\sigma\lambda$.
-- `feeding` increases resource number by $\lambda$.
+- `scaring` is assumed to be nonlethal and therefore have no effect on resource number (resources are moved to a random cell on the landscape, as sampled from a uniform distribution such that movement to any given cell is equally probable).
+- `culling` decreases resource number by $\sigma$.
+- `castration` decreases resource number by $\sigma\lambda$, where $\lambda$ is the GMSE argument `lambda` that defines the baseline population growth rate of resources.
+- `feeding` increases resource number by $\sigma\lambda$.
 - `help_offspring` increases resource number by $\sigma$.
 
+Note that $\sigma$ is included in all of the predicted actions above as a modulator for how strongly the manager predicts users will respond to a change in manager policy (e.g., a value of 0 would predict no reaction on the part of users to a change in policy, while a value of 1 would predict that an action would increase in exact proportion to its decrease in cost).
 These effects cannot be altered directly in `gmse` or `gmse_apply` (though parameter values can of course be changed using `manager_sense` and `lambda` arguments), but future versions of GMSE might include different predicted effects to increase precision or allow for multiple resource types or different actions.

### 2018-05-09 notebook/ms/ms.Rmd
Reviewer 1 concerns regarding SI1

@@
 
-Overall, a single call of the genetic algorithm simulates the process of thinking and decision-making for one manager or user.
+A single call of the genetic algorithm simulates the process of thinking and decision-making for one manager or user.
 In each new call of the genetic algorithm, a unique population of temporary manager or user strategies is initialised.
@@
 |  RESOURCE_ini   |  1000   | The initial abundance of resources                                                               |
-|  scaring        |  FALSE  | Resource scaring is a policy option                                                              |
-|  culling        |  TRUE   | Resource culling is a policy option                                                              |
-|  castration     |  FALSE  | Resource castration is a policy option                                                           |
-|  feeding        |  FALSE  | Resource feeding (increases lambda) is a policy option                                           |
-|  help_offspring |  FALSE  | Resource helping (increases offspring number) is a policy option                                 |
+|  scaring        |  FALSE  | Resource scaring (moves a resource to a random landscape cell) is a policy option                |
+|  culling        |  TRUE   | Resource culling (removes a resource entirely) is a policy option                                |
+|  castration     |  FALSE  | Resource castration (sets a resource's lambda to zero) is a policy option                        |
+|  feeding        |  FALSE  | Resource feeding (increases a resource's lambda) is a policy option                              |
+|  help_offspring |  FALSE  | Resource helping (increases a resource's offspring number) is a policy option                    |
 |  tend_crops     |  FALSE  | Users can increase landscape cell values                                                         |

### 2018-05-09 notebook/ms/SI1.Rmd
Small edits, mostly as suggested by Izzy

@@
 
-The focal data structure used for tracking manager and user decisions is a three dimensional array, which we will call `ACTION` (also returned as `user_array` by `gmse_apply`).
+The focal data structure used for tracking manager and user decisions is a three dimensional array, which we will call `ACTION` (also returned as `user_array` by `gmse_apply`; see [SI7](SI7.pdf)).
 Rows of `ACTION` correspond to the entities affected by actions (resources, landscape properties, or potentially other agents), and columns correspond either to properties of the affected entities, or to the actions potentially allocated to them.
@@
 
-Constraints on the values that elements in the `ACTION` array can take are defined by a `COST` array (also returned as `manager_array` by `gmse_apply`) of dimensions identical to `ACTION`.
+Constraints on the values that elements in the `ACTION` array can take are defined by a `COST` array (also returned as `manager_array` by `gmse_apply`; see [SI7](SI7.pdf)) of dimensions identical to `ACTION`.
 Elements of `COST` define how many units from the `manager_budget` or `user_budget` are needed to perform a single action; a `minimum_cost` for actions is defined as an argument in GMSE (10 by default).

### 2018-05-09 notebook/ms/SI3.Rmd
Small edits, mostly as suggested by Izzy

@@
 
-@Johnson2018 recently estimated key demographic parameters of the Taiga Bean Geese from the Central Management Unit, which includes geese that breed in "Northern most Sweden, Northern Norway, Northern and Central Finland, and adjacent North-wester parts of Russia, wintering mostly in Southern Sweden and South-east Denmark" [@AEWA2016].
+@Johnson2018 recently estimated key demographic parameters of the Taiga Bean Geese from the Central Management Unit, which includes geese that breed in "Northern most Sweden, Northern Norway, Northern and Central Finland, and adjacent North-western parts of Russia, wintering mostly in Southern Sweden and South-east Denmark" [@AEWA2016].
 They estimated goose survival under ideal conditions to be ca 0.878; this can be interpreted in our model by setting mortality to `remove_pr = 1 - 0.878`.
@@
 Goose locations at the end of the simulation are shown in the upper left panel, while the upper right panel shows the same landscape broken down among the 80 farmers (upper 60% of the landscape in multiple colours), along with non-agricultural land (lower 40% of the landscape in blue).
-Actual goose abundance is shown in the middle left panel (black solid line), along with its estimate by the manager (blue solid line).
+Actual goose abundance is shown in the middle left panel (black solid line), along with its estimate by the manager (blue solid line, hidden underneath the black line).
 The horizontal red and blue dotted lines show goose carrying capacity and the manager's target for goose abundance, respectively.

### 2018-05-09 notebook/ms/SI4.Rmd
Small edits, mostly as suggested by Izzy

@@
 
-Note that the argument `get_res = "Full"` causes `sim_old` retain all of the relevant data structures for simulating a new time step and recording simulation results.
+Note that the argument `get_res = "Full"` causes `sim_old` to retain all of the relevant data structures for simulating a new time step and recording simulation results.
 This includes the key simulation output, which is located in `sim_old$basic_output`, which is printed below.
@@
 Below, a custom function is defined that causes a goose to move up to 5 cells in any direction if it finds itself on a cell with more than 10 other geese.
-As with default GMSE simulations, movement is based on a torus landscape.
+As with default GMSE simulations, movement is based on a torus landscape (where no landscape edge exists, so that if resources move off of one side of the landscape they appear on the opposite side).
 

### 2018-05-09 notebook/ms/SI5.Rmd
Small edits, mostly as suggested by Izzy

@@
 We will use the results from the observation model to update recruitment for the new year in `rec.m`.
-For simplicity, spawning stock biomass prior to harvest will be randomly sampled from a value in last 10 years (i.e., from `ssb.m` between 1994 and 2004), but more realistic models could relate this spawning stock biomass to recruitment and environmental variables from a prevoius year; spawning stock biomass will be decreased after harvest based on user actions.
+For simplicity, spawning stock biomass prior to harvest will be randomly sampled from a value in the last 10 years (i.e., from `ssb.m` between 1994 and 2004), but more realistic models could relate this spawning stock biomass to recruitment and environmental variables from a prevoius year; spawning stock biomass will be decreased after harvest based on user actions.
 The GMSE initialisation and simulation is below.

### 2018-05-09 notebook/ms/SI6.Rmd
Small edits, mostly as suggested by Izzy

@@
 
-When manager can only make policy decisions once every five time steps, extinction occurs in `r sum(gmse_reps5[,2] < 20)` out of 10 simulated populations before year 20.
+When a manager can only make policy decisions once every five time steps, extinction occurs in `r sum(gmse_reps5[,2] < 20)` out of 10 simulated populations before year 20.
 If we wanted to summarise these results, we could plot how extinction risk changes with increasing `manage_freq`.

### 2018-05-10 src/game.c
Remove redundant declared variables

@@ -458,12 +458,11 @@ void sum_array_layers(double ***array, double **out, int get_mean,
 int new_act(double old_cost, double new_cost, double old_act, double *paras){
     
     int total_acts;
-    double users, max_to_spend, acts_per_user, cost_per_user, total_cost;
+    double users, acts_per_user, total_cost;
     double budget_for_act, mgr_budget, min_cost;
     
     users        = paras[54] - 1; /* Minus one for the manager */
     min_cost     = paras[96];     /* Minimum cost of an action */
-    max_to_spend = paras[97];     /* Maximum per user budget   */
     mgr_budget   = paras[105];    /* Manager's total budget    */
     
     total_cost    = 0.0;

### 2018-05-10 notebook/ms/SI1.Rmd
Some work on the new SI1

@@
 
-```{r, echo=FALSE, fig.height=1.75, fig.width=6}
+```{r, echo=FALSE, fig.height=1.75, fig.width=6, fig.cap = "Conceptual overview of the GMSE genetic algorithm"}
 mbox <- function(x0, x1, y0, y1){
@@
 ```
-
-**Figure S1-1:** Conceptual overview of the GMSE genetic algorithm
 
@@
 Individual fitness as calculated for managers ($F_{i}^{m}$) is affected by a manager's utility for resources and the projected change in resource abundance caused by the individual's policy (i.e., the contents of their `POPULATION` layer, specifically row 3; here again we use 'individual' to refer to one of `ga_popsize` discrete strategies in `POPULATION`, which may be selected and reproduce within the genetic algorithm).
-Manager utility for a resource ($U^{m}_{res}$) is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model (see "[Key data structures used](#Key_data_structures_used)" above).
+Manager utility for a resource ($U^{m}_{res}$) is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model (see "[Key data structures used](#Key_data_structures_used)" above, and [SI7](SI7.pdf#action) for more information).
 Manager utility can therefore change in each GMSE time step as estimated resource abundance changes; when the estimated resource abundance is greater than `manage_target`, $U^{m}_{res}$ is negative, and when the estimated resource abundance is less than `manage_target`, $U^{m}_{res}$ is positive.
@@
 
-To predict change in resource abundance as a consequence of policy, an individual first needs to know the total number of actions of all types $j$ (e.g., scaring, culling, etc.) performed by users in the previous time step ($X_{\bullet, j}$), and the cost of performing each action ($C_{\bullet, j}$).
+To predict change in resource abundance as a consequence of policy, an individual first needs to know the total number of actions of all types $j$ (e.g., scaring, culling, etc.) performed by users in the previous time step ($X_{\bullet, j}$; a value of `manage_caution` is added to this value, with a default of `manage_caution = 1`, to ensure that managers do not na&#0239;vely assume that users will not perform an action just because they did not perform it in the previous time step), and the cost of performing each action ($C_{\bullet, j}$).
 This information is collected from `ACTION` and `COST` arrays.
 The individual then needs to predict how their policy (i.e., the costs that they set for users to perform an action) will affect the new total number of each action performed ($X_{i,j}$).
-To do this, the individual assumes that total user actions performed under their policy will change in proportion to that of the old policy.
-The predicted total number of a particular action $j$ performed under the policy of $i$ is thereby calculated as,
+To do this, the individual assumes that total user actions performed under their policy will change in proportion to that of the old policy, while also recognising that users have a maximum above which higher costs set by the manager will have no effect.
+Inclined readers might wish to examine the short [`new_act`]() function, which is summarised mathematically below; this function is called by the [`policy_to_counts`]() function in the [genetic algorithm source file]().
 
-$$ X_{i,j} =  (X_{\bullet, j} + 1) \frac{C_{\bullet, j}}{C_{i,j}}. $$
+The manager first calculates how much total budget of users' was devoted to an action by multiplying the old per action cost $C_{\bullet, j}$ by the total number of actions performed, $X_{\bullet, j}$.
+The manager then divides this by the new cost $C_{i, j}$ per action to calculate the new predicted number of actions,
 
-The variable $C_{i,j}$ defines the new cost set by the individual $i$ for action $j$.
-A value of 1 is added to ($X_{\bullet, j}$) to model some degree of caution by the manager (this can be changed from the default 1 using `manage_caution`), especially so that managers do not na&#0239;vely assume that users will not perform an action just because they did not perform it in the previous time step.
-Otherwise, if $X_{\bullet, j} = 0$, then the manager would always assume that a change in the cost of an action would have no effect on the number of times the action was performed by users; a value of 1 assumes that at least one user will perform the action in the new time step.
+$$X_{i, j} = \frac{X_{\bullet, j} \times C_{\bullet, j}}{C_{i, j}}.$$
+
+Note that if $C_{i, j} = C_{\bullet, j}$, then the total number of new predicted actions $j$ will remain unchanged.
+If $C_{i, j} > C_{\bullet, j}$, then the total number of new actions will decrease, and if $C_{i, j} < C_{\bullet, j}$, then the total number of new actions will increase.
 

### 2018-05-10 notebook/ms/SI4.Rmd
More notebook edits

@@
 
-```{r}
+```{r, fig.cap = "Default position of land ownership by farmers."}
 image(x = sim_old$LAND[,,3], col = topo.colors(9), xaxt = "n", yaxt = "n");
@@
 
-```{r}
+```{r, fig.cap = "Land ownership by farmers as customised in gmse_apply."}
 sim_old$LAND[1:20,   1:20,  3] <- 2; 

### 2018-05-10 notebook/ms/SI6.Rmd
More notebook edits

@@
 
-```{r, fig.height = 5}
+```{r, fig.height = 5, fig.cap = "Extinction risk given an increasing number of time steps between updating policy decisions for culling costs in a simulated population.
+Higher values on the x-axis correspond to more time passing before a new policy is set.
+For each point, a total of 10 replicate simulations were run."}
 ext_risk1 <- sum(gmse_reps1[,2] < 20);

### 2018-05-10 notebook/ms/SI1.Rmd
Large edits made

@@
 A genetic algorithm is called in the predefined GMSE manager and user models to simulate human decision making.
-As of GMSE version 0.3.1.9, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step.
-Therefore, one run of the genetic algorithm occurs to simulate the manager's policy-setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users copy).
+As of GMSE version 0.4.0.3, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step.
+Therefore, one run of the genetic algorithm occurs to simulate the manager's policy-setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1; e.g., see [SI6](SI6.pdf)), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users copy).
 Each run of the genetic algorithm mimics the evolution by natural selection of a population of potential manager or user strategies over multiple iterations, with the highest fitness strategy in the terminal iteration being selected as the one that the manager or user decides to implement.
-For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'iteration' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the main text).
-Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.3.1.9 (future versions of GMSE might expand upon this framework, and we highlight some of these potential avenues for expansion).
-We first explain the key data structures used, then provide an overview of how a population of strategies is initialised, and the  subsequent processes of crossover, mutation, cost constraint, fitness evaluation, tournament selection, and replacement.
-We then explain the fitness functions of managers and users in more detail.
+For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'iteration' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the [main text](ms.pdf)).
+Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.4.0.3 (future versions of GMSE might expand upon this framework, and we highlight some of these potential avenues for expansion).
+We first explain the [key data structures used](#Key_data_structures_used), then provide an overview of how a population of strategies is [initialised](#initialised), and the subsequent processes of [crossover](#crossover), [mutation](#mutation), [cost constraint](#constraint), [fitness](#eval) evaluation, [tournament selection](#tournament), and [replacement](#replacement).
+We then explain the fitness functions of [managers](#manfit) and [users](#ufit) in more detail.
 
-<a name="Key_data_structures_used">Key data structures used</a>
+Key data structures used {#Key_data_structures_used}
 ================================================================================
@@
 Positive values are currently only meaningful for `Manager_Actions`, where a value of 1 defines an action setting a uniform cost of users' direct actions on resources (i.e., costs where `Act = -2` for `User_1_Actions` and `User_2_Actions`).
-All other values for `Act` are meaningless in GMSE 0.3.1.9, but might be expanded upon in future versions to allow for modification of specific user costs enacted by managers (i.e., managers having different policies for different users) or other users (e.g., users increasing the costs of other users' actions due to conflict or cooperation).
-For the rest of this supporting information, we will therefore focus only on rows 1-3 of `ACTION`.
+All other values for `Act` are meaningless in GMSE 0.4.0.3, but might be expanded upon in future versions to allow for modification of specific user costs enacted by managers (i.e., managers having different policies for different users) or other users (e.g., users increasing the costs of other users' actions due to conflict or cooperation).
+We will therefore focus only on rows 1-3 of `ACTION`.
 
-Columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in predefined GMSE v0.3.1.9 manager and user sub-models (i.e., only one type of resource is permitted).
+Columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in predefined GMSE v0.4.0.3 manager and user sub-models (i.e., only one type of resource is permitted).
 Future versions might allow for different resource types (e.g., `Type_1` might be used to designate species, and `Type_2` and `Type_3` could designate stage or sex).
@@
 Elements of `COST` define how many units from the `manager_budget` or `user_budget` are needed to perform a single action; a `minimum_cost` for actions is defined as an argument in GMSE (10 by default).
-All values in `COST` columns 1-7 are set to 10001, one higher than the highest possible `manager_budget` or `user_budget`, so neither managers nor users can affect resource types or utilities.
+All values in `COST` columns 1-7 are set to 100001, one higher than the highest possible `manager_budget` or `user_budget`, so neither managers nor users can affect resource types or utilities.
 Columns 8-13 are also set to 10001, except where actions are allowed.
-Maximum values of 10000 are unrelated to landscape cells (the total number of which is also 10000 by default).
+Maximum values of 100000 are independent of any other parameter value specified in GMSE (e.g., landscape dimensions).
 Below shows the `COST` array that corresponds to the above `ACTION` array.
@@
 Note that in default GMSE parameters, `culling = TRUE`, but all other actions are set to `FALSE`.
-Hence the `Cull` column 9 is the only column besides column 13 `None` in which cost is less than 10001.
+Hence, the `Cull` column 9 is the only column besides column 13 `None` in which cost is less than 100001.
 Manager's actions in `ACTION` directly affect the cost of users performing one of the five possible actions on resources (columns 8-12).
@@
 In each iteration, these layers crossover and mutate, generating variation in potential agent decisions; costs constrain this variation from exceeding a maximum budget, then the fitness of each layer is evaluated based on how the layer is predicted to affect resources or landscape output to which the agent has assigned some utility.
-A tournament is used to select high fitness layers, and these selected layers become the new iteration of layers; iterations continue until a minimum number of iterations (`ga_mingen`) have passed and a convergence criteria is satisfied such that the increase in mean fitness from the previous iteration is below the threshold `converge_crit` (Figure S1-1).
+A tournament is used to select high fitness layers, and these selected layers become the new iteration of layers; iterations continue until a minimum number of iterations (`ga_mingen`) have passed and a convergence criteria is satisfied such that the increase in mean fitness from the previous iteration is below the threshold `converge_crit` (Figure 1 below).
 
@@
 
-Initialisation
+Initialisation {#initialisation}
 --------------------------------------------------------------------------------
@@
 The number of exact replicates initialised is set using `ga_seedrep` (if `ga_seedrep` $\geq$ `ga_popsize`, then all individuals are seeded as replicates).
-After the `POPULATION` of `ga_popsize` individuals is initialised, a loop simulating the adaptive evolution of `POPULATION` in non-overlapping iterations begins (see Figure S1-1 above).
+After the `POPULATION` of `ga_popsize` individuals is initialised, a loop simulating the adaptive evolution of `POPULATION` in non-overlapping iterations begins (see Figure 1 above).
 
-Crossover
+Crossover {#crossover}
 --------------------------------------------------------------------------------
@@
 
-Mutation
+Mutation {#mutation}
 --------------------------------------------------------------------------------
@@
 
-Cost constraint
+Cost constraint {#constraint}
 --------------------------------------------------------------------------------
@@
 
-Fitness evaluation
+Fitness evaluation {#eval}
 --------------------------------------------------------------------------------
@@
 
-Tournament selection
+Tournament selection {#tournament}
 --------------------------------------------------------------------------------
@@
 
-Replacement and termination
+Replacement and termination {#replacement}
 --------------------------------------------------------------------------------
@@
 The genetic algorithm terminates if and only if a minimum number of iterations has passed (`ga_mingen`) and a convergence criteria (`converge_crit`) is satisfied.
-The convergence criteria checks the difference between the mean fitness of individuals in the new iteration versus the previous iteration; if this difference is greater than `converge_crit`, then termination does not occur (this prevents termination from occuring while fitness is still increasing, though it is usually fine to use the default GMSE `converge_crit = 100` and `ga_mingen = 40`, which nearly always terminates the genetic algorithm after 40 iterations having identified adaptive manager or user strategies).
+The convergence criteria checks the difference between the mean fitness of individuals in the new iteration versus the previous iteration; if this difference is greater than `converge_crit`, then termination does not occur (this prevents termination from occuring while fitness is still increasing, though it is usually fine to use the default GMSE `converge_crit = 0.1` and `ga_mingen = 40`, which nearly always terminates the genetic algorithm after 40 iterations having identified adaptive manager or user strategies).
 Due to the way in which fitness is calculated (see below), in practice, `converge_crit` currently applies only to users.
-If termination conditions are not satisfied, then the `POPULATION` of individuals begins a new iteration of crossover, mutation, cost constraint, fitness evaluation, and tournament selection (Figure S1-1).
+If termination conditions are not satisfied, then the `POPULATION` of individuals begins a new iteration of crossover, mutation, cost constraint, fitness evaluation, and tournament selection (Figure 1).
 
@@
 Here we explain how the fitnesses of candidate manager and user strategies in a `POPULATION` array (see above) are calculated.
-We emphasise that the fitness functions used in GMSE v0.3.1.9 are intended to be heuristic tools for identifying reasonable manager and user behaviours.
+We emphasise that the fitness functions used in GMSE v0.4.0.3 are intended to be heuristic tools for identifying reasonable manager and user behaviours.
 In practice, our fitness functions identify behaviours that are well-aligned with manager and user interests for harvesting or crop yield, but they are not intended to identify *optimal* decisions.
@@
 
-Fitness function for managers
+Fitness function for managers {#manfit}
 --------------------------------------------------------------------------------
@@
 
-To predict change in resource abundance as a consequence of policy, an individual first needs to know the total number of actions of all types $j$ (e.g., scaring, culling, etc.) performed by users in the previous time step ($X_{\bullet, j}$; a value of `manage_caution` is added to this value, with a default of `manage_caution = 1`, to ensure that managers do not na&#0239;vely assume that users will not perform an action just because they did not perform it in the previous time step), and the cost of performing each action ($C_{\bullet, j}$).
+To predict change in resource abundance as a consequence of policy, an individual first needs to know the total number of actions of all types $j$ (e.g., scaring, culling, etc.) performed by users in the previous time step ($X_{\bullet, j}$; note that this value includes the increment `manage_caution`, with a default of `manage_caution = 1`, to ensure that managers do not na&#0239;vely assume that users will not perform an action just because they did not perform it in the previous time step), and the cost of performing each action ($C_{\bullet, j}$).
 This information is collected from `ACTION` and `COST` arrays.
-The individual then needs to predict how their policy (i.e., the costs that they set for users to perform an action) will affect the new total number of each action performed ($X_{i,j}$).
+The individual $i$ then needs to predict how their policy (i.e., the costs that they set for users to perform an action) will affect the new total number of each action $j$ performed ($X_{i,j}$).
 To do this, the individual assumes that total user actions performed under their policy will change in proportion to that of the old policy, while also recognising that users have a maximum above which higher costs set by the manager will have no effect.
-Inclined readers might wish to examine the short [`new_act`]() function, which is summarised mathematically below; this function is called by the [`policy_to_counts`]() function in the [genetic algorithm source file]().
+Interested readers might wish to examine the short [`new_act`](https://github.com/bradduthie/gmse/blob/master/src/game.c#L452) function, which is summarised mathematically below; this function is called by the [`policy_to_counts`](https://github.com/bradduthie/gmse/blob/master/src/game.c#L482) function in the [genetic algorithm source file](https://github.com/bradduthie/gmse/blob/dev/src/game.c).
 
-The manager first calculates how much total budget of users' was devoted to an action by multiplying the old per action cost $C_{\bullet, j}$ by the total number of actions performed, $X_{\bullet, j}$.
+The manager first calculates how much total budget, as summed over all users, was devoted to an action by multiplying the old per action cost $C_{\bullet, j}$ by the total number of actions performed, $X_{\bullet, j}$.
 The manager then divides this by the new cost $C_{i, j}$ per action to calculate the new predicted number of actions,
@@
 
-Note that if $C_{i, j} = C_{\bullet, j}$, then the total number of new predicted actions $j$ will remain unchanged.
+Note again that if $C_{i, j} = C_{\bullet, j}$, then the total number of new predicted actions $j$ will remain unchanged.
 If $C_{i, j} > C_{\bullet, j}$, then the total number of new actions will decrease, and if $C_{i, j} < C_{\bullet, j}$, then the total number of new actions will increase.
@@
 In practice, the default $\sigma = 0.9$ performs well.
-Allowed actions are predicted by managers to have the following effects (again, we emphasise that whether or not these effects are realised will depend later on the user model, to which the manager does not have access):
+Allowed actions are predicted by managers to have the following effects (again, we emphasise that whether or not these effects are realised will depend later on the user model, to which the manager -- by design -- does not have access):
 
@@
 Note that $\sigma$ is included in all of the predicted actions above as a modulator for how strongly the manager predicts users will respond to a change in manager policy (e.g., a value of 0 would predict no reaction on the part of users to a change in policy, while a value of 1 would predict that an action would increase in exact proportion to its decrease in cost).
-These effects cannot be altered directly in `gmse` or `gmse_apply` (though parameter values can of course be changed using `manager_sense` and `lambda` arguments), but future versions of GMSE might include different predicted effects to increase precision or allow for multiple resource types or different actions.
+
+The above effects cannot be altered directly in `gmse` or `gmse_apply` (though parameter values can of course be changed using `manager_sense` and `lambda` arguments), but future versions of GMSE might include different predicted effects to increase precision or allow for multiple resource types or different actions.
 The summation of $X_{i,j}$ for all actions defines the predicted change in resource abundance caused by the policy of an individual $i$, $\Delta A_{i}$.
 
-Fitness function for users
+Fitness function for users {#ufit}
 --------------------------------------------------------------------------------
@@
 Note that $F_{i}^{u}$ increases when $\Delta A_{i}$ and $\Delta L_{i}$ are of the same sign as $U^{u}_{res}$ and $U^{u}_{land}$, respectively.
-Further, in GMSE v0.3.1.9, only one term of the equation is nonzero.
+Further, in GMSE v0.4.0.3, only one term of the equation is nonzero.
 When `land_ownership = FALSE` (default, modelling users that harvest resources), $U^{u}_{res} = -1$ and $U^{u}_{land} = 0$, and when `land_ownership = TRUE`, $U^{u}_{res} = 0$ and $U^{u}_{land} = 100$ (modelling farmers trying to increase crop yield).
@@
 - `scaring` decreases resource number by $1$.
-- `culling` decreases resource number by $1 + \lambda$.
+- `culling` decreases resource number by $1$.
 - `castration` decreases resource number by $\lambda$.
 - `feeding` increases resource number by $\lambda$.
-- `help_offspring` increases resource number by $\sigma$.
+- `help_offspring` increases resource number by 1.
 
@@
 
-$$\Delta A_{i} = (\lambda) Feeds + Helps -Scares - (1 + \lambda) Culls - (\lambda) Castrations.$$
+$$\Delta A_{i} = (\lambda) Feeds + Helps -Scares - Culls - (\lambda) Castrations.$$
 
@@
 |---------------------|---------|----------------------------------------------|
-| `ga_popsize`        | 100     | The number of individuals in the population temporary simulated during a single run of the genetic algorithm. |
+| `ga_popsize`        | 100     | The number of individuals in the population temporarily simulated during a single run of the genetic algorithm. |
 | `ga_mingen`         | 40      | The minimum number of iterations that a genetic algorithm will run before settling on an agent's strategy. |
-| `ga_seedrep`        | 0       | The number of individuas in the population to be initiaised with the current agent's strategy (e.g., from a previous time step in the broader GMSE simulation), as opposed to being initialised with random strategies.
+| `ga_seedrep`        | 20      | The number of individuas in the population to be initiaised with the current agent's strategy (e.g., from a previous time step in the broader GMSE simulation), as opposed to being initialised with random strategies.
 |
@@
 | `ga_crossover`      | 0.1     | The crossover rate of any action in an agent's strategy; crossover events occur with a different randomly selected strategy in the population. |
-| `ga_converge_crit`  | 1       | The percent increase in strategy fitness from one iteration to the next below which the convergence criteria is satisfied.
+| `ga_converge_crit`  | 0.1     | The percent increase in strategy fitness from one iteration to the next below which the convergence criteria is satisfied.
 Iterations wil continue as long as fitness increase is above this convergence criteria.

### 2018-05-10 notebook/ms/SI2.Rmd
Consistent hyphens

@@
 The `gmse_apply` function is a flexible function that allows for user-defined sub-functions calling resource, observation, manager, and user models.
-Where such models are not specified, predefined GMSE submodels 'resource', 'observation', 'manager', and 'user' are run by default.
+Where such models are not specified, predefined GMSE sub-models 'resource', 'observation', 'manager', and 'user' are run by default.
 Any type of sub-model (e.g., numerical,  individual-based) is permitted as long as the input and output are appropriately specified.
@@
 
-A simple run of `gmse_apply()` returns one generation of GMSE using predefined submodels and default parameter values.
+A simple run of `gmse_apply()` returns one generation of GMSE using predefined sub-models and default parameter values.
 
@@
 
-For `sim_1`, the default 'basic' results are returned as below, which summarise key values for all submodels.
+For `sim_1`, the default 'basic' results are returned as below, which summarise key values for all sub-models.
 
@@
 
-How `gmse_apply` integrates across submodels
+How `gmse_apply` integrates across sub-models
 --------------------------------------------------------------------------------
 
-To integrate across different types of submodels, `gmse_apply` translates between vectors and arrays between each submodel.
+To integrate across different types of sub-models, `gmse_apply` translates between vectors and arrays between each sub-model.
 For example, because the default GMSE observation model requires a resource array with particular requirements for column identites, when a resource model subfunction returns a vector, or a list with a named element 'resource_vector', this vector is translated into an array that can be used by the observation model.
@@
 At each step, a translation between the two is made, with necessary adjustments that can be tweaked through arguments to `gmse_apply` when needed.
-Alternative observation, manager, and user, submodels, for example, are defined below; note that each requires a vector from the preceding model.
+Alternative observation, manager, and user, sub-models, for example, are defined below; note that each requires a vector from the preceding model.
 
 ```{r, eval = TRUE}
-# Alternative observation submodel
+# Alternative observation sub-model
 alt_obs <- function(resource_vector){ 
@@
 
-# Alternative manager submodel
+# Alternative manager sub-model
 alt_man <- function(observation_vector){
@@
 
-# Alternative user submodel
+# Alternative user sub-model
 alt_usr <- function(manager_vector){
@@
 
-All of these submodels are completely deterministic, so when run with the same parameter combinations, they produce replicable outputs.
+All of these sub-models are completely deterministic, so when run with the same parameter combinations, they produce replicable outputs.
 
@@
 Hence while manger output is set in terms of costs of performing each action, and user output is set in terms of action attempts, this need not be the case when using `gmse_apply` (though it should be recognised when using default GMSE manager and user functions).
-GMSE default submodels can be added in at any point.
+GMSE default sub-models can be added in at any point.
 

### 2018-05-10 notebook/ms/SI4.Rmd
Consistent hyphens

@@
 All of this information was available under the list element `sim_new$basic_output`, but other list elements of `sim_new` might also be useful to record.
-It is important to remember that this example of `gmse_apply` is using the default resource, observation, manager, and user submodels.
-Custom submodels could produce different outputs in `sim_new` (see [Supporting Information 2](SI2.pdf) for examples).
-For default submodels, there are some list elements that might be especially useful.
+It is important to remember that this example of `gmse_apply` is using the default resource, observation, manager, and user sub-models.
+Custom sub-models could produce different outputs in `sim_new` (see [Supporting Information 2](SI2.pdf) for examples).
+For default sub-models, there are some list elements that might be especially useful.
 These elements can potentially be edited *within the above loop* to dynamically adjust simulations.
@@
 - `sim_new$LAND`: The landscape on which interactions occur, which is stored as a 3D array with `land_dim_1` rows, `land_dim_2` columns, and 3 layers.
-Layer 1 (`sim_new$LAND[,,1]`) is not currently used in default submodels, but could be used to store values that affect resources and agents.
+Layer 1 (`sim_new$LAND[,,1]`) is not currently used in default sub-models, but could be used to store values that affect resources and agents.
 Layer 2 (`sim_new$LAND[,,2]`) stores crop yield from a cell, and layer 3 (`sim_new$LAND[,,3]`) stores the owner of the cell (value corresponds to the agent's ID).
@@
 The goal was to show how software users can work with these existing sub-models and data structures to customise GMSE simulations.
-Software users seeking even greater flexibility (e.g., replacing an entire built-in submodel with a custom submodel) should refer to the [Supporting Information 2](SI2.pdf) that introduces `gmse_apply` more generally.
+Software users seeking even greater flexibility (e.g., replacing an entire built-in sub-model with a custom sub-model) should refer to the [Supporting Information 2](SI2.pdf) that introduces `gmse_apply` more generally.
 Future versions of GMSE are likely to expand on the built-in options available for simulation; requests for such expansions, or contributions, can be submitted to [GitHub](https://github.com/bradduthie/gmse/issues).

### 2018-05-10 notebook/ms/SI5.Rmd
More notebook edits
Consistent hyphens

@@
 Hence, GMSE is not intended as a substitute for packages such as [FLR](http://www.flr-project.org/), but the integration of these packages with GMSE could make use of GSME's current and future simulation capabilities, and particularly the [genetic algorithm](SI1.pdf).
-Such integration might be possible using the `gmse_apply` function, which allows for [custom defined submodels](SI2.pdf) to be used within the GMSE framework, and with default GMSE submodels.
+Such integration might be possible using the `gmse_apply` function, which allows for [custom defined sub-models](SI2.pdf) to be used within the GMSE framework, and with default GMSE sub-models.
 Hence, GMSE might be especially useful for modelling the management of fisheries under conditions of increasing competing stakeholder demands and conflicts.
@@
 This population model will use sample data and one of the many available stock-recruitment models available in FLR, and a custom function will be written to return a single value for stock recruitment.
-Currently, `gmse_apply` requires that submodels return subfunction results either as scalar values or data frames that are structured [in the same way](SI7.pdf) as GMSE submodels.
+Currently, `gmse_apply` requires that sub-models return subfunction results either as scalar values or data frames that are structured [in the same way](SI7.pdf) as GMSE sub-models.
 But interpretation of scalar values is left up to the user (e.g., population model results could be interpreted as abundance or biomass; manager policy could be interpreted as cost of harvesting or as total allowable catch).
@@
 
-```{r, fig.height = 6.2}
+```{r, fig.height = 6.2, fig.cap = "Output of the FLR plot function for an example Ricker model of stock recruitment on North Sea Herring data."}
 plot(newFL);
@@
 The [FLR project](http://www.flr-project.org/) includes libraries that can be used to [perform a management strategy evaluation](http://www.flr-project.org/doc/An_introduction_to_MSE_using_FLR.html) (MSE) under fisheries-focused observation, manager, and user models.
-We will not recreate [this approach](http://www.flr-project.org/doc/An_introduction_to_MSE_using_FLR.html), or integrate any other submodels into GMSE as was done for the population model above, although such integration of submodels should be possible using similar techniques.
+We will not recreate [this approach](http://www.flr-project.org/doc/An_introduction_to_MSE_using_FLR.html), or integrate any other sub-models into GMSE as was done for the population model above, although such integration of sub-models should be possible using similar techniques.
 Our goal here is to instead show how the `predict_recruitment` model created above can be integrated with `gmse_apply`, which can then make use of the genetic algorithm to predict the fishers' behaviour.
@@
 An initial run of these models gives values of `r round(new_rec, digits = 2)` for `new_rec` and `r round(obs_rec, digits = 2)` for `obs_rec`.
-We are now ready to use the built-in manager and user submodels in `gmse_apply`.
+We are now ready to use the built-in manager and user sub-models in `gmse_apply`.
 We will assume that managers attempt to keep a recruitment of 5000, and that there are 4 independent fishers [stakeholders in gmse_apply says 10, is that a difference?] who attempt to maximise their catch.
@@
 This example simulation demonstrates the ability of GMSE to integrate with fisheries libraries such as [FLR](http://www.flr-project.org/) through `gmse_apply`.
-In addition to being a useful wrapping function for MSE submodels, `gmse_apply` can therefore be used to take advantage of the genetic algorithm in the GMSE default manager and user models.
+In addition to being a useful wrapping function for MSE sub-models, `gmse_apply` can therefore be used to take advantage of the genetic algorithm in the GMSE default manager and user models.
 This flexibility will be retained in future versions of `gmse_apply`, allowing custom resource and observation models that are built for specific systems to be integrated with an increasingly complex genetic algorithm simulating various aspects of human decision-making.
@@
 Future versions of GMSE will continue to be open-source and developed to avoid unecessary dependencies (GMSE v.0.4.0.3 requires only base R).
-Key goals including (1) providing highly general and useful default `resource`, `observation`, `manager`, and `user` submodels for a variety of MSE modelling tasks, (2) keeping these submodels highly modular so that they can be developed in isolation given standardised data structures, and (3) allowing these modular submodels to be integrated with custom defined submodels as flexibly as possible using `gmse_apply`.
+Key goals including (1) providing highly general and useful default `resource`, `observation`, `manager`, and `user` sub-models for a variety of MSE modelling tasks, (2) keeping these sub-models highly modular so that they can be developed in isolation given standardised data structures, and (3) allowing these modular sub-models to be integrated with custom defined sub-models as flexibly as possible using `gmse_apply`.
 Contributions in line with these goals, and suggestions for new features, can be made on [GitHub](https://github.com/bradduthie/gmse).

### 2018-05-10 notebook/ms/SI6.Rmd
Consistent hyphens

@@
 
-The individual-based approach of default GMSE submodels
+The individual-based approach of default GMSE sub-models
 ================================================================================
 
-The default submodels of GMSE (`resource`, `observation`, `manager`, `user`) are individual-based (also called 'agent-based'), meaning that they model discrete individuals (resources or agents), which in GMSE are represented by individual table rows (as in `RESOURCES`, `AGENTS`, and `OBSERVATION`) or layers of three-dimensional arrays (as in `COST` and `ACTION`).
+The default sub-models of GMSE (`resource`, `observation`, `manager`, `user`) are individual-based (also called 'agent-based'), meaning that they model discrete individuals (resources or agents), which in GMSE are represented by individual table rows (as in `RESOURCES`, `AGENTS`, and `OBSERVATION`) or layers of three-dimensional arrays (as in `COST` and `ACTION`).
 Individual-based models (IBMs) have been a useful approach in ecology for decades [@Uchmanski1996; @Grimm1999], providing both a pragmatic tool for the mechanistic modelling of complex populations and a powerful technique for theoretical investigation.
@@
 In GMSE, some of the most important traits for resources include types, ages, demographic parameter values, locations, etc., and for agents (manager and users), traits include different types, utilities, budgets, etc.
-The traits that resources and managers have can potentially affect their interactions, and default GMSE submodels take advantage of this by simulating interactions explicitly on a landscape (see [SI7](SI7.pdf) for an introduction to GMSE default data structures).
+The traits that resources and managers have can potentially affect their interactions, and default GMSE sub-models take advantage of this by simulating interactions explicitly on a landscape (see [SI7](SI7.pdf) for an introduction to GMSE default data structures).
 
@@
 A benefit of this approach is that it allows for the simulation of multiple different scenarios under conditions of uncertainty and stochasticity, modelling the range of outcomes that might occur within and among scenarios and facilitating the development of social-ecological theory.
-Future expansion on the complexity of individual-based default submodels of GMSE will further increase the realism of targeted case studies.
+Future expansion on the complexity of individual-based default sub-models of GMSE will further increase the realism of targeted case studies.

### 2018-05-10 notebook/ms/SI7.Rmd
Consistent hyphens

@@
 
-The default submodels of GMSE (`resource`, `observation`, `manager`, and `user`) use a small number of default data structures to hold the information needed in simulations.
-While these default submodels do not necessarily need to be used in every run in GMSE (see [use of](SI2.pdf) `gmse_apply`), they will be used in any run of the `gmse` function, and in any call of the `gmse_apply` function that does not run with entirely custom submodels.
-Simulation and model inference do not require an understanding of the default data structures, but such an understanding can be especially useful when running `gmse_apply` if there is a need to extract uncommonly used information, change key simulated values (e.g., landscape properties, agent budgets, or resource movement rules as in [Supporting Information 4](SI4.pdf)), or build custom individual-based submodels.
+The default sub-models of GMSE (`resource`, `observation`, `manager`, and `user`) use a small number of default data structures to hold the information needed in simulations.
+While these default sub-models do not necessarily need to be used in every run in GMSE (see [use of](SI2.pdf) `gmse_apply`), they will be used in any run of the `gmse` function, and in any call of the `gmse_apply` function that does not run with entirely custom sub-models.
+Simulation and model inference do not require an understanding of the default data structures, but such an understanding can be especially useful when running `gmse_apply` if there is a need to extract uncommonly used information, change key simulated values (e.g., landscape properties, agent budgets, or resource movement rules as in [Supporting Information 4](SI4.pdf)), or build custom individual-based sub-models.
 Here we provide a brief explanation of the following key data structures (each name below is listed as it is named in the output `gmse_apply` when `get_res = "Full"`).
@@
 It is obvious from the above list that most columns represent traits that are either typically ignored or currently not in use.
-This is intended to allow for easier future development of default model options and potential customisation of submodels in `gmse_apply`.
+This is intended to allow for easier future development of default model options and potential customisation of sub-models in `gmse_apply`.
 We anticipate that future versions of GMSE will contain multiple user types with unique traits and among-user interactions.
@@
 
-Events in default GMSE submodels occur on a spatially-explicit landscape `LAND`, which is stored as a three dimensional array.
+Events in default GMSE sub-models occur on a spatially-explicit landscape `LAND`, which is stored as a three dimensional array.
 The size of this landscape is specified with the `land_dim_1` and `land_dim_2` arguments of GMSE, which determine the length, in cells, of the y and x dimensions of the landscape, respectively (e.g., if `land_dim_1 = 10` and `land_dim_2 = 1000`, then the landscape will be one very long horizontal transect).

### 2018-05-10 notebook/ms/SI2.Rmd
More edits

@@
 Where model parameters are needed but not specified, defaults from GMSE are used.
-In this Supporting Information, we demonstrate some uses of `gmse_apply`, and how it might be used to simulate myriad management scenarios *in silico*.
+Here we demonstrate some uses of `gmse_apply`, and how it might be used to simulate myriad management scenarios *in silico*.
 
-A simple run of `gmse_apply()` returns one generation of GMSE using predefined sub-models and default parameter values.
+A simple run of `gmse_apply()` returns one time step of GMSE using predefined sub-models and default parameter values.
 
@@
 
-Note that in the case above we have the total abundance of resources returned (`sim_1$resource_results`), the estimate of resource abundance from the observation function (`sim_1$observation_results`, the costs the manager sets for the only available action of culling (`sim_1$manager_results`), and the number of culls attempted by each user (`sim_1$user_results`).
-By default, only one resource type is used, but custom subfunctions could potentially allow for models with multiple resource types.
-Any custom subfunctions can replace GMSE predefined functions, provided that they have appropriately defined inputs and outputs (see [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf)).
+Note that in the case above we have the total abundance of resources returned (`sim_1$resource_results`), the estimate of resource abundance from the observation function (`sim_1$observation_results`), the costs the manager sets for the only available action of culling (`sim_1$manager_results`), and the number of culls attempted by each user (`sim_1$user_results`).
+By default, only one resource type is used, but custom sub-functions could potentially allow for models with multiple resource types.
+Any custom sub-functions can replace GMSE predefined functions, provided that they have appropriately defined inputs and outputs (see [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf)).
 For example, we can define a very simple logistic growth function to send to `res_mod` instead.
@@
 The above function takes in a population size of `X` and returns a value `X_1` based on the population intrinsic growth rate `rate` and carrying capacity `K`.
-Iterating the logistic growth model by itself under default parameter values with a starting population of 100 will cause the population to increase to carrying capacity in ca seven generations.
-The function can be substituted into `gmse_apply` to use it instead of the predefined GMSE resource model.
+Iterating the logistic growth model by itself under default parameter values with a starting population of 100 will cause the population to increase to carrying capacity in ca seven time steps The function can be substituted into `gmse_apply` to use it instead of the predefined GMSE resource model.
 
@@
 To integrate across different types of sub-models, `gmse_apply` translates between vectors and arrays between each sub-model.
-For example, because the default GMSE observation model requires a resource array with particular requirements for column identites, when a resource model subfunction returns a vector, or a list with a named element 'resource_vector', this vector is translated into an array that can be used by the observation model.
+For example, because the default GMSE observation model requires a resource array with particular requirements for column identites, when a resource model sub-function returns a vector, or a list with a named element 'resource_vector', this vector is translated into an array that can be used by the observation model.
 Specifically, each element of the vector identifies the abundance of a resource type (and hence will usually be just a single value denoting abundance of the only focal population).
 If this is all the information provided, then a 'resource_array' will be made with default GMSE parameter values with an identical number of rows to the abundance value (floored if the value is a non-integer; non-default values can also be put into this transformation from vector to array if they are specified in `gmse_apply`, e.g., through an argument such as `lambda = 0.8`).
-Similarly, a `resource_array` is also translated into a vector after the default individual-based resource model is run, should the observation model require simple abundances instead of an array.
+Similarly, a `resource_array` is also translated into a vector after the default individual-based resource model is run, should a custom observation model require simple abundances instead of an array.
 The same is true of `observation_vector` and `observation_array` objects returned by observation models, of `manager_vector` and `manager_array` (i.e., `COST` in the `gmse` function) objects returned by manager models, and of `user_vector` and `user_array` (i.e., `ACTION` in the `gmse` function) objects returned by user models.
@@
 Note that the `manager_results` and `user_results` are ambiguous here, and can be interpreted as desired -- e.g., as total allowable catch and catches made, or as something like costs of catching set by the manager and effort to catching made by the user.
-Hence while manger output is set in terms of costs of performing each action, and user output is set in terms of action attempts, this need not be the case when using `gmse_apply` (though it should be recognised when using default GMSE manager and user functions).
+Hence, while manger output is set in terms of costs of performing each action, and user output is set in terms of action attempts, this need not be the case when using `gmse_apply` (though it should be recognised when using default GMSE manager and user functions).
 GMSE default sub-models can be added in at any point.
@@
 
-It is possible to, for example, specify a simple resource and observation model, but then take advantage of the genetic algorithm to predict policy decisions and user actions.
+It is possible to, e.g., specify a simple resource and observation model, but then take advantage of the genetic algorithm to predict policy decisions and user actions (see [SI5](SI5.pdf) for a fisheries example).
 This can be done by using the default GMSE manager and user functions (written below explicitly, though this is not necessary).
@@
 There is an important note to make about changing arguments to `gmse_apply` when `old_list` is being used: The function `gmse_apply` is trying to avoid a crash, so `gmse_apply` will accomodate parameter changes by rebuilding data structures if necessary.
-For example, if the number of stakeholders is changed (and by including an argument `stakeholders` to `gmse_apply`, it is assumed that stakeholders are changing even they are not), then a new array of agents will need to be built.
+For example, if the number of stakeholders is changed (and by including an argument such as `stakeholders` to `gmse_apply`, it is assumed that stakeholders are changing even they are not), then a new array of agents will need to be built.
 If landscape dimensions are changed (or just include the argument `land_dim_1` or `land_dim_2`), then a new landscape willl be built.

### 2018-05-10 notebook/ms/SI2.Rmd
Minor edits, new read

@@
 header-includes:
+- \usepackage{lineno}
+- \linenumbers
 linestretch: 1

### 2018-05-10 notebook/ms/SI3.Rmd
Minor edits, new read

@@
     reference_docx: docx_template.docx
-header-includes: null
+header-includes:
+- \usepackage{lineno}
+- \linenumbers
 linestretch: 1
@@
 
-Agents in GMSE (managers and users) are goal-oriented, and their behaviour is therefore driven to maximise a particular utility of interest such as a target density of resources, such as animals or trees of conservation interest (managers), or a suitable resource (e.g.
-animal population size, forest) or landscape harvest size, such as bag size or timber (users).
-This model feature allows GMSE to evaluate the actions of agents (e.g.
-managers and users) in the context of their individual objectives, and to therefore quantify the degree to which those objectives are or are not achieved.
+Agents in GMSE (managers and users) are goal-oriented, and their behaviour is therefore driven to maximise a particular utility of interest such as a target density of resources.
+For managers and users, this could include animals or trees of conservation interest.
+For users, it could additionally include a landscape harvest size such as bag size or timber.
+This model feature allows GMSE to evaluate the actions of agents in the context of their individual objectives, and to therefore quantify the degree to which those objectives are or are not achieved.
 When the actions of one party clashes with the objectives of another party, the objectives of one might be expressed at the expense of the other, causing conservation conflict [@Redpath2013].
-Currently, there is no standard way to measure conservation conflict in a social-ecological system where both the natural resource (e.g.
-animals, plants) and the people (agents, stakeholders, managers, users etc.) are modelled in a single system, and previous modelling approaches have not meaningfully separated agent objectives from agent actions.
+Currently, there is no standard way to measure conservation conflict in a social-ecological system where both the natural resource (e.g., animals, plants, or non-biological resources) and the people (e.g., stakeholders, managers, etc.) are modelled in a single system, and previous modelling approaches have not meaningfully separated agent objectives from agent actions.
 We suggest that a starting point to developing a useful metric of conservation conflict is to quantify the deviation of an individual's actions from their objectives  (i.e., of actual actions from desired actions), the former of which is restricted by the actions of other individuals.
@@
 We consider a protected population of waterfowl that exploits and damages agricultural land and is therefore a source of conservation conflict between those that seek the conservation of waterfowl and those that are concerned with the loss of agriculture [e.g., @Fox2017a; @Mason2017; @Tulloch2017; @Cusack2018].
-As in the main text example, the objective of the manager is to keep waterfowl at a target abundance that minmises extinction risks, while the objective of farmers is to maximise agricultural production on their landscape.
+As in the main text example, the objective of the manager is to keep waterfowl at a target abundance that minimises extinction risks, while the objective of farmers is to maximise agricultural production on their landscape.
 Here we consider a more complex simulation, with a level of detail that more accurately reflects a scenario that might occur in a real social-ecological system where the manager sets policies that incentivise the user to act in a way that ensures the persistence of the resource.
-The policies are backed up by a manager budget that can be allocated to set costs of actions and thus incentivise users to follw with the desired actions.
-The user also has a budget to carry out actions and the cost of the actions is affected by the manager policies and budget.
+The policies are backed up by a manager budget that can be allocated to set costs of actions and thus incentivise users to perform the desired actions.
+The user also has a budget to carry out actions, and the cost of the user actions is affected by the manager policies and budget.
 Our objective is not to model the dynamics of a specific system, but to show how GMSE could be parameterised using demographic estimates from empirical studies.
@@
 We parameterise our model using demographic information from the Taiga Bean Goose (*Anser fabalis fabalis*), a managed population that is hunted for recreation in Fennoscandinavia [@Johnson2018].
-Taiga Bean Geese can cause agricultural damage [@Johnson2018], which could potentially lead to conflict between farming and management/conservation objectives.
+Taiga Bean Geese can cause agricultural damage [@Johnson2018], which could potentially lead to conflict between farming and management or conservation objectives.
 
@@
 Where GMSE parameter values are not available, we use reasonable values or GMSE defaults.
-To make model inferences for real case studies, we strongly recommend simulating across a range of parameter values when empirical estimates are unavailable, as social-ecological dynamics might be sensitive to these unknown values.
+To make model inferences for real case studies, we strongly recommend [replicating simulations](SI6.pdf) and simulating across a range of parameter values when empirical estimates are unavailable, as social-ecological dynamics might be sensitive to these unknown values.
 
@@
 Similarly, @Johnson2018 estimated mean reproductive rate and carrying capacity to be 0.55 and 93870, respectively, so we set `lambda = 0.275` (for simplicity, we simply use half the mean reproductive rate; GMSE does not currently distinguish female and male individuals) and `res_death_K = 93870`.
-The global abundance of Taiga Bean Geese in 2009 was ca 63000 [@Fox2010], and ca 35000 in the Central Management Unit [@AEWA2016], which we can take as a starting abundance for our simulations (`RESOURCE_ini = 35000`).
-And the International Single Species Action Plan has a target population size of ca 70000 in the Central Management Unit [@AEWA2016], which we can use as a management target (`manage_target = 70000`).
+The global abundance of Taiga Bean Geese in 2009 was ca 63000 [@Fox2010], with ca 35000 in the Central Management Unit [@AEWA2016], which we can take as a starting abundance for our simulations (`RESOURCE_ini = 35000`).
+The International Single Species Action Plan has a target population size of ca 70000 in the Central Management Unit [@AEWA2016], which we can use as a management target (`manage_target = 70000`).
 We simulate social-ecological dynamics over 30 time steps, which could be interpreted as years.
@@
 Further, we consider the case of a region in which farmland makes up 60% of all land, with 40% of land being 'public' (`public_land = 0.4`; which might be interpreted as any land in which stakeholders are not, or cannot be, invested in goose presence), and divide the farmland amongst 80 individual farmers (`stakeholders = 80`; `land_ownership = TRUE`).
-Landscape size is set to default 100 by 100 cells, so each farmer effectively owns 75 cells, which might be interpreted as hectares of land (for instructions on how to more precisely control landscape ownership, see the advanced GMSE options using `gmse_apply`).
+Landscape size is set to default 100 by 100 cells, so each farmer effectively owns 75 cells, which might be interpreted as hectares of land (for instructions on how to more precisely control landscape ownership, see the [advanced GMSE options](SI4.pdf) using `gmse_apply`).
 Because we need both density-dependent (`res_death_K = 93870`) and density-independent (`remove_pr = 0.122`) sources of mortality, we set `res_death_type = 3`.
@@
 The horizontal red and blue dotted lines show goose carrying capacity and the manager's target for goose abundance, respectively.
-The organge line shows the total percent of landscape cell (including non-farmed cells) yield, as decreased by geese.
+The orange line shows the total percent of landscape cell (including non-farmed cells) yield, as decreased by geese.
 The middle right panel shows this yield for each farmer, and for the public land (lower line in blue).
@@
 In response, users begin to cull geese on their land, and the goose population begins to stabilise around the target of 7000 total geese.
-We can investigate the conflict between management policy and farmers more directly using the `plot_gmse_effort` function.
+We can investigate the conflict between management policy and farmers more directly using the `plot_gmse_effort` function (Figure 2).
 
@@
 
-Black lines indicate how permissive a manager is toward a particular action on a scale of 0 to 100, while coloured lines indicate how much effort farmers expend on a given action.
+Black lines in Figure 2 indicate how permissive a manager is toward a particular action on a scale of 0 to 100, while coloured lines indicate how much effort farmers expend on a given action.
 When black lines are far below coloured lines, we can (cautiously) interpret this as a conflict between management of the goose population and farmer's interest in agricultural production.
@@
 ```{r, eval = TRUE, fig.height = 8, fig.cap = "Results of a GMSE simulation using parameters estimated for Taiga Been Geese Central Management Unit in which scaring is permitted.
-Simulation output is interpreted as in Figure 1"}
+Simulation output is interpreted as in Figure 1."}
 plot_gmse_results(res = sim_2$resource, obs = sim_2$observation, 
@@
 Additionally, where entirely different types of sub-models are required, the `gmse_apply` function [can be used](https://bradduthie.github.io/gmse_SI2.pdf) to more flexibly simulate social-ecological systems.
+In [SI4](SI4.pdf), we show an example of this using the same Taiga Bean Geese case study that we did here.
+
+\clearpage

### 2018-05-10 notebook/ms/SI4.Rmd
Minor edits, new read

@@
 Here we demonstrate how simulations in GMSE can be more fine-tuned to specific empirical situations through the use of `gmse_apply`.
-To do this, we use the same scenario described in [Supporting Information 3](SI3.pdf); we first recreate the basic scenario run in `gmse` using `gmse_apply`, and then build in additional modelling details including (1) [custom placement of user land](#land), (2) [parameterisation of individual user budgets](#budgets), and (3) [density-dependent movement of resources](#movement).
+To do this, we use the same scenario described in [SI3](SI3.pdf); we first recreate the basic scenario run in `gmse` using `gmse_apply`, and then build in additional modelling details including (1) [custom placement of user land](#land), (2) [parameterisation of individual user budgets](#budgets), and (3) [density-dependent movement of resources](#movement).
 We emphasise that these simulations are provided only to demonstrate the use of GMSE, and specifically to show the flexibility of the `gmse_apply` function, not to accurately recreate the dynamics of a specific system or make management recommendations.
@@
 This is done to for two reasons.
-First, it speeds up simulations for the purpose of demonstration; second, it makes the presentation of landscape ownership easier (see below).
+First, it speeds up simulations for the purpose of demonstration; second, it makes the presentation of our custom landscape ownership easier to visualise (see below).
 
@@
 
-We now recreate the simulations in [Supporting Information 3](SI3.pdf), which were run using the `gmse` function, in `gmse_apply`.
+We now recreate the simulations in [SI3](SI3.pdf), which were run using the `gmse` function, in `gmse_apply`.
 Doing so requires us to first initialise simulations using one call of `gmse_apply`, then loop through multiple time steps that again call `gmse_apply`; results of interest are recorded in a data frame (`sim_sum_1`).
-Following the protocol introduced in [Supporting Information 2](SI2.pdf), we can call the initialising simulation `sim_old`, and use the code below to read in the relevant parameter values.
+Following the protocol introduced in [SI2](SI2.pdf), we can call the initialising simulation `sim_old`, and use the code below to read in the relevant parameter values.
 
@@
 
-We can then loop over 30 time steps to recreate the simulations from [Supporting Information 3](SI3.pdf).
-In these simulations, we are specifically interested in the resource and observation outputs, as well as the manager policy and user actions for culling, which we record below in the data frame `sim_sum`.
+We can then loop over 30 time steps to recreate the simulations from [SI3](SI3.pdf).
+In these simulations, we are specifically interested in the resource and observation outputs, as well as the manager policy and user actions for culling, which we record below in the data frame `sim_sum_1`.
 The inclusion of the argument `old_list` tells `gmse_apply` to use parameters and values from the list `sim_old` in the new time step.
@@
 It is important to remember that this example of `gmse_apply` is using the default resource, observation, manager, and user sub-models.
-Custom sub-models could produce different outputs in `sim_new` (see [Supporting Information 2](SI2.pdf) for examples).
+Custom sub-models could produce different outputs in `sim_new` (see [SI2](SI2.pdf) for examples).
 For default sub-models, there are some list elements that might be especially useful.
 These elements can potentially be edited *within the above loop* to dynamically adjust simulations.
+For more explanation of built-in GMSE data arrays, see [SI7](SI7.pdf).
 
-- `sim_new$resource_array`: A table holding all information on resources.
+- [`sim_new$resource_array`](SI7.pdf#resources): A table holding all information on resources.
 Rows correspond to discrete resources, and columns correspond to resource properties: (1) ID, (2-4) types (not currently in use), (5) x-location, (6) y-location, (7) movement parameter, (8) time, (9) density independent mortality parameter (`remove_pr`), (10) reproduction parameter (`lambda`), (11) offspring number, (12) age, (13-14) observation columns, (15) consumption rate (`res_consume`), and (16-20) recorded experiences of user actions (e.g., was the resource culled or scared?).
-- `sim_new$AGENTS`: A table holding basic information on agents (manager and users).
+- [`sim_new$AGENTS`](SI7.pdf#agents): A table holding basic information on agents (manager and users).
 Rows correspond to a unique agent, and columns correspond to agent properties: (1) ID, (2) type (0 for the manager, 1 for users), (3-4) additional type options not currently in use, (5-6), x and y locations (usually ignored), (7) movement parameter (usually ignored), (8) time, (9) agent's viewing ability in cells (`agent_view`), (10) error parameter, (11-12) values for holding marks and tallies of resources, (13-15) values for holding observations, (16) yield from landscape cells, (17) budget (`manager_budget` and `user_budget`).
-- `sim_new$observation_vector`: Estimate of total resource number from the observation model (`observation_array` also holds this information in a different way depending on `observe_type`)
-- `sim_new$LAND`: The landscape on which interactions occur, which is stored as a 3D array with `land_dim_1` rows, `land_dim_2` columns, and 3 layers.
+- `sim_new$observation_vector`: Estimate of total resource number from the observation model ([`observation_array`](SI7.pdf#observation) also holds this information in a different way depending on `observe_type`)
+- [`sim_new$LAND`](SI7.pdf#land): The landscape on which interactions occur, which is stored as a 3D array with `land_dim_1` rows, `land_dim_2` columns, and 3 layers.
 Layer 1 (`sim_new$LAND[,,1]`) is not currently used in default sub-models, but could be used to store values that affect resources and agents.
 Layer 2 (`sim_new$LAND[,,2]`) stores crop yield from a cell, and layer 3 (`sim_new$LAND[,,3]`) stores the owner of the cell (value corresponds to the agent's ID).
-- `sim_new$manage_vector`: The cost of each action as set by the manager. For even more fine-tuning, individual costs for the actions of each agent can be set for each user in `sim_new$manager_array`.
-- `sim_new$user_vector`: The total number of actions performed by each user. A more detailed breakdown of actions by individual users is held in `sim_new$user_array`.
+- `sim_new$manage_vector`: The cost of each action as set by the manager.
+For even more fine-tuning, individual costs for the actions of each agent can be set for each user in [`sim_new$manager_array`](SI7.pdf#cost).
+- `sim_new$user_vector`: The total number of actions performed by each user. A more detailed breakdown of actions by individual users is held in [`sim_new$user_array`](SI7.pdf#action).
 
@@
 
-The number of cells owned by each farmer is therefore listed in the table below.
+The number of cells owned by the manager (1) and each farmer (2-8) is therefore listed in the table below.
 
@@
     cells_owned            <- sum(sim_old$LAND[,,3] == ID);
-    sim_old$AGENTS[ID, 17] <- 10 * cells_owned;
+    sim_old$AGENTS[ID, 17] <- 100 * cells_owned;
 }
@@
 The goal was to show how software users can work with these existing sub-models and data structures to customise GMSE simulations.
-Software users seeking even greater flexibility (e.g., replacing an entire built-in sub-model with a custom sub-model) should refer to the [Supporting Information 2](SI2.pdf) that introduces `gmse_apply` more generally.
+Readers seeking even greater flexibility (e.g., replacing an entire built-in sub-model with a custom sub-model) should refer to [SI2](SI2.pdf) that introduces `gmse_apply` more generally.
 Future versions of GMSE are likely to expand on the built-in options available for simulation; requests for such expansions, or contributions, can be submitted to [GitHub](https://github.com/bradduthie/gmse/issues).

### 2018-05-10 notebook/ms/SI5.Rmd
Minor edits, new read

@@
 Early development of management strategy evaluation (MSE) models originated in fisheries [@Polacheck1999; @Smith1999; @Sainsbury2000].
-Consequently, fisheries-focused software for MSE has been extensively developed, including R libraries that focus on the management of species of exceptional interest, such as the Atlantic Bluefin Tuna (*Thunnus thynnus*) [[ABFTMSE](https://github.com/ICCAT/abft-mse/tree/master/R_package/ABTMSE); @Carruthers2018; @Carruthers2018a], and Indian Ocean Bigeye (*T.
+Consequently, fisheries-focused software for MSE has been extensively developed, including R libraries that focus on the management of species of exceptional interest, such as the Atlantic Bluefin Tuna (*Thunnus thynnus*) [[ABFTMSE](https://github.com/ICCAT/abft-mse/tree/master/R_package/ABTMSE); @Carruthers2018a; @Carruthers2018], and Indian Ocean Bigeye (*T.
 obesus*) and Yellowfin (*T.
@@
 
-The FLR toolset includes [a series of pacakges](http://www.flr-project.org/#packages), with several [tutorials](http://www.flr-project.org/doc/index.html) for using them.
+The FLR toolset includes [a series of packages](http://www.flr-project.org/#packages), with several [tutorials](http://www.flr-project.org/doc/index.html) for using them.
 For simplicity, we focus on [a model of stock recruitment](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html) to be used as the population model in `gmse_apply`.
@@
 But interpretation of scalar values is left up to the user (e.g., population model results could be interpreted as abundance or biomass; manager policy could be interpreted as cost of harvesting or as total allowable catch).
-For simplicity, the observation (i.e., estimation) model will be the stock reported from the population model with error, and the manager model will be a total allowable catch calculated from the stock-recruitment relationship that accounts for the number of fishers in the system.
-The user model, however, will employ the full power of the default GMSE user function to simulate user actions.
+For simplicity, the observation (i.e., estimation) model will be the stock reported from the population model with error.
+The manager and user models, however, will employ the full power of the default GMSE functions to simulate management and user actions.
 We first show how a custom function can be made that applies the FLR toolset to a population model.
@@
 ```{r, eval = FALSE}
-install.packages(c("FLCore"), repos="http://flr-project.org/R");
+install.packages("FLCore", repos="http://flr-project.org/R");
 ```
@@
 We are now ready to use the built-in manager and user sub-models in `gmse_apply`.
-We will assume that managers attempt to keep a recruitment of 5000, and that there are 4 independent fishers [stakeholders in gmse_apply says 10, is that a difference?] who attempt to maximise their catch.
+We will assume that managers attempt to keep a recruitment of 5000, and that there are 10 independent fishers [stakeholders in gmse_apply says 10, is that a difference?] who attempt to maximise their catch.
 We assign a user budget of `manager_budget = 10000`, and all other values are set to GMSE defaults.
@@
 We will use the results from the observation model to update recruitment for the new year in `rec.m`.
-For simplicity, spawning stock biomass prior to harvest will be randomly sampled from a value in the last 10 years (i.e., from `ssb.m` between 1994 and 2004), but more realistic models could relate this spawning stock biomass to recruitment and environmental variables from a prevoius year; spawning stock biomass will be decreased after harvest based on user actions.
+For simplicity, spawning stock biomass prior to harvest will be randomly sampled from a value in the last 10 years (i.e., from `ssb.m` between 1994 and 2004), but more realistic models could relate this spawning stock biomass to recruitment and environmental variables from a previous year; spawning stock biomass will be decreased after harvest based on user actions.
 The GMSE initialisation and simulation is below.

### 2018-05-10 notebook/ms/SI6.Rmd
New notes

@@
 
-For a single simulation, the `gmse_table` function prints out key information from a `gmse` simulation result. The example provided in the GMSE documentation is below.
+For a single simulation, the `gmse_table` function prints out key information from a `gmse` simulation result.
+The example provided in the [GMSE documentation](https://cran.r-project.org/package=GMSE) is below.
 
@@
 ```{r, eval = FALSE}
-write.csv(x= sim_table, file = "file_path/gmse_table");
+write.csv(x= sim_table, file = "file_path/gmse_table_name.csv");
 ```

### 2018-05-10 notebook/ms/ms.Rmd
Minor edits
Add acknowledgements

@@
 |  agent_view     |  10     | How far managers can see on the landscape for resource counting when observe_type = 0            |
-|  res_birth_K    |  10000  | Carrying capacity applied to the number of resources added during a time step                    |
+|  res_birth_K    |  100000 | Carrying capacity applied to the number of resources added during a time step                    |
 |  res_death_K    |  2000   | Carrying capacity applied to the number of resources removed during a time step                  |
@@
 
+Conclusions
+===================================================
+
+Acknowledgements
+===================================================
+
+This project was funded by the European Research Council under the European Union's H2020/ERC grant agreement no.
+679651 (ConFooBio) to NB.
+ABD is funded by a Leverhulme Trust Early Career Fellowship.
+We thank two anonymous reviewers for very helpful comments.
+
 References

### 2018-05-11 notebook/ms/ms.Rmd
Add new concluding section -- draft appears to be complete

@@
 GMSE offers a range of parameters to simulate resource dynamics (primarily, but not necessarily biological populations) and management policy options, and includes genetic algorithms to dynamically model stakeholder (manager and user) decision-making.
-Genetic algorithms find adaptive solutions to any simulated conditions given stakeholder-specific goals, allowing GMSE to model scenarios of conservation conflict.
+Genetic algorithms find adaptive solutions to any simulated conditions given stakeholder-specific goals (see [SI1](SI1.pdf)), allowing GMSE to model scenarios of conservation conflict.
 
@@
 1).
-The workhorse function `gmse` runs simulations using four predefined submodels, which can be parameterised to fit various case studies; more tailored submodels can also be defined using the `gmse_apply` function (see below).
+The workhorse function `gmse` runs simulations using four predefined submodels, which can be parameterised to fit various case studies; more tailored submodels can also be defined using the `gmse_apply` function (see [SI2](SI1.pdf)).
 (1) A population of discrete resources (e.g., a managed species) with individual traits (e.g., location, age) is modelled on a spatially-explicit landscape and can simulate resource birth, movement, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty.
@@
 Consistent with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for agents (stakeholders).
-Instead, genetic algorithms are used to heuristically find an adaptive strategy for each stakeholder in each time step (see Supporting Information 1 for details).
+Instead, genetic algorithms are used to heuristically find an adaptive strategy for each stakeholder in each time step (see [SI1](SI1.pdf) for details).
 Critically, all stakeholders involved in resource conservation are constrained in their decision-making; managing and using resources takes effort (e.g., time or money), and effort expended in developing or enforcing one policy (for managers) or performing one action (for users) will be effort not expendable elsewhere [@Milner-Gulland2011; @Muller-Hansen2017; @Schluter2017].
@@
 Allocations from a manager's budget can be used to increase the cost it takes a user to perform an action (i.e., 'policy'), and allocations from a user's budget can be used to perform the action at the cost set by the manager.
-Hence, stakeholders can have incomplete control over resource use and express competing management objectives.
+Hence, stakeholders can have incomplete control over resource use and express competing management objectives, potentially resulting in conflict.
 
@@
 Output of `gmse` is an exhaustive list that includes all resources and observations, all stakeholder decisions and actions, and all landscape properties in each time step of the simulation.
-Results are most easily interpreted visually, so a summary of simulation dynamics is plotted by default (the plot can also be called using the `plot_gmse_results` function, and a summary of results can be obtained using `gmse_summary`).
+Results are most easily interpreted visually, so a summary of simulation dynamics is plotted by default (the plot can also be called using the `plot_gmse_results` function, and summaries of results can be obtained using `gmse_summary` and `gmse_table`).
 An example below shows how simulations are set and interpreted.
@@
 Waterfowl consume and destroy all crop yield upon arrival to a landscape cell.
-In each time step, waterfowl are observed on a subset of cells, then managers extrapolate from density per cell to estimate total population size.
+In each time step, waterfowl are observed on a random subset of cells, then managers extrapolate from density per cell to estimate total population size.
 Managers then use these estimates to set costs of culling and scaring waterfowl for five farmers (scaring is non-lethal, causing waterfowl to move to a random landscape cell).
@@
 Parameters in `gmse` not listed are set to default values.
-By plotting the output with `plot_gmse_results`, simulation results can be interpreted visually (manager and user decisions can also be interpreted using the `plot_gmse_effort` function).
+By plotting the output with `plot_gmse_results`, simulation results can be interpreted visually (manager and user decisions can also be interpreted using the `plot_gmse_effort` function, see [SI3](SI3.pdf) and [SI4](SI4.pdf) for an expanded example).
 
@@
 
-The function `gmse_apply` allows custom resource, observation, manager, or user sub-models to be integrated into the GMSE framework.
+The function `gmse_apply` allows custom resource, observation, manager, or user sub-models to be integrated into the GMSE framework (see [SI2](SI2.pdf), [SI4](SI4.pdf), and [SI5](SI5.pdf)).
 Any type of sub-model (e.g., numerical, individual-based) is permitted by defining a function with appropriately specified inputs and outputs; where custom functions are not provided, `gmse_apply` runs default GMSE sub-models used in `gmse`.
@@
 The `gmse_apply` function simulates a single GMSE time step, and therefore must be looped for simulations over multiple time steps.
-Within loops, GMSE arguments can be redefined to simulate changing conditions (e.g., change in policy availability or stakeholder budgets, see Supporting Information 2), thereby allowing many management scenarios to be simulated *in silico*.
+Within loops, GMSE arguments can be redefined to simulate changing conditions (e.g., change in policy availability or stakeholder budgets, see [SI2](SI2.pdf) and [SI4](SI4.pdf)), thereby allowing many management scenarios to be simulated *in silico*.
 
@@
 
+Here we have introduced the R package GMSE v0.4.0.3, new software for modelling social-ecological dynamics under scenarios of potential conflict.
+GMSE provides a powerful tool for individual-based modelling simulations while also allowing for extensive model customisation.
+GMSE vignettes provide additional examples for getting started with `gmse` or `gmse_gui` for simulations using default resource, observation, manager, and user sub-models ([SI3](SI3.pdf) and [SI6](SI6.pdf)), and for using `gmse_apply` for advanced model customisation ([SI2](SI2.pdf), [SI4](SI4.pdf), and [SI7](SI7.pdf)) and integration with existing packages ([SI5](SI5.pdf)).
+Future versions of GMSE will include additional features and improve upon the realism of social and ecological modelling components, while also maintaining a high degree of flexibility and modulatarity for model customisation.
+
 Acknowledgements
@@
 ABD is funded by a Leverhulme Trust Early Career Fellowship.
-We thank two anonymous reviewers for very helpful comments.
+We thank two anonymous reviewers for helpful comments.
 

### 2018-05-14 notebook/ms/SI4.Rmd
Vignette made, need SI too

@@
     cells_owned            <- sum(sim_old$LAND[,,3] == ID);
-    sim_old$AGENTS[ID, 17] <- 100 * cells_owned;
+    sim_old$AGENTS[ID, 17] <- 10 * cells_owned;
 }

### 2018-05-14 R/gmse.R
Spacing deletion following valgrind check

@@ -10,7 +10,7 @@
 #'@param land_dim_1 This value sets the number of cells on the x dimension of the landscape (i.e., the number of columns in the landscape array; this can also be thought of as the x-axis when the landscape image is plotted). There is no maximum, but the minimum dimension of a landscape is 2 cells. The default is 100 cells.
 #'@param land_dim_2 This value sets the number of cells on the y dimension of the landscape (i.e., the number of columns in the landscape array; this can also be thought of as the y-axis when the landscape image is plotted). There is no maximum, but the minimum dimension of a landscape is 2 cells. The default is 100 cells.
 #'@param res_movement This value determines how far resources move during a time step. Exact movement is probabilistic and partly affected by `res_move_type` settings. Under default settings, during each time step, resources move from zero to res_movement cells away from their starting cell in any direction. Hence res_movement is the maximum distance away from a resources starting cell that it can move in a time step; other types of resource movement, however, interpret res_movement differently to get the raw distance moved (see res_move_type). The default value is 20.
-#'@param remove_pr This value is the density-independent and user-independent probability of a resource being removed (e.g., dying) during a time step in the resource model. Under default settings, this value is set to zero, with resource removal being determined entirely by carrying capcity on resource survival, and by user actions. 
+#'@param remove_pr This value is the density-independent and user-independent probability of a resource being removed (e.g., dying) during a time step in the resource model. Under default settings, this value is set to zero, with resource removal being determined entirely by carrying capacity on resource survival, and by user actions. 
 #'@param lambda This value is the baseline population growth rate of resources. Each resource in the simulation produces Poisson(lambda) offspring in one time step within the resource model. The value of lambda might be increased or decreased by user actions, and juvenile survival can potentially be decreased by a carrying capacity placed on birth. The default value is 0.3, meaning that the average resource produces one offspring every three time steps.
 #'@param agent_view This value determines how far agents (managers and stakeholders) can see on the landscape. At the moment, this affects only the sampling ability of managers in the observation model for density-based and transect-based estimates of resource abundance. In these types of estimates, when managers have a higher agent_view, they are capable of observing a larger area of landscape and therefore of getting a larger (in the case of density-based estimation) or more efficient (in the case of transect-based estimation) sample of resources from which to estimate total resource abundance. The default value of agent_view is 10, so agents can see 10 cells away from their current cell in any direction.
[…]

### 2018-05-14 R/gmse_apply.R

@@ -2255,4 +2255,3 @@ get_user_sum <- function(arg_list){
     return(act_mat);
 }
 ################################################################################
-

### 2018-05-15 notebook/ms/SI4.Rmd
MS ready SI

@@
     reference_docx: docx_template.docx
-header-includes: null
+header-includes: 
+- \usepackage{lineno}
+- \linenumbers
 linestretch: 1
@@
 
+\clearpage
+
 ```{r}

### 2018-05-15 notebook/ms/SI4.Rmd
New vignette without code page break

--- 2018-05-14 notebook/ms/SI4.Rmd
@@
     reference_docx: docx_template.docx
-header-includes: null
+header-includes: 
 linestretch: 1

### 2018-05-15 notebook/ms/SI5.Rmd
Remove the parenthetical comment

@@
 We are now ready to use the built-in manager and user sub-models in `gmse_apply`.
-We will assume that managers attempt to keep a recruitment of 5000, and that there are 10 independent fishers [stakeholders in gmse_apply says 10, is that a difference?] who attempt to maximise their catch.
+We will assume that managers attempt to keep a recruitment of 5000, and that there are 10 independent fishers who attempt to maximise their catch.
 We assign a user budget of `manager_budget = 10000`, and all other values are set to GMSE defaults.

### 2018-05-24 notebook/ms/SI1.Rmd

@@
 If $u$ is greater than `1 - (0.5 * ga_mutation)`, then the value of the array element is increased by 1.
-If $u$ is less than `0.5 * ga_mutation`, then the value of the array element is decreased by 1; when this decrease results in a negative value, the mutated value is multiplied by -1 to equal 1.
+If $u$ is less than `0.5 * ga_mutation`, then the value of the array element is decreased by 1; when this decrease results in a negative value, the mutated value is multiplied by -1 to be positive.
 

### 2018-05-25 notebook/ms/SI3.Rmd
Remove MS text to barely get over the hurdle of 3475 words

@@
 Alternatively, experienced R users may prefer to simulate by [looping time steps](SI2.pdf#page=5) through `gmse_apply`, which allows more flexibility for incorporating custom sub-models and dynamically adjusting parameter values.
+
+Simulations using the default GMSE sub-models described above are run using the `gmse` function, which offers a range of options for setting parameter values (see Table 1 for some select examples).
+Output of `gmse` is an exhaustive list that includes all resources and observations, all stakeholder decisions and actions, and all landscape properties in each time step of the simulation.
+Results are most easily interpreted visually, so a summary of simulation dynamics is plotted by default (the plot can also be called using the `plot_gmse_results` function, and summaries of results can be obtained using `gmse_summary` and `gmse_table`).
+An example below shows how simulations are set and interpreted.
+
+\small
+
+|  Argument       | Default | Description                                                                                      |
+|-----------------|---------|--------------------------------------------------------------------------------------------------|
+|  time_max       |  100    | Maximum time steps in simulation                                                                 |
+|  land_dim_1     |  100    | Width of the landscape (horizontal cells)                                                        |
+|  land_dim_2     |  100    | Height of the landscape (vertical cells)                                                         |
+|  res_movement   |  20     | Distance (cells) a resource can move in any direction (for movement rules, see res_move_type)    |
+|  remove_pr      |  0      | Density-independent probability of resource mortality during a time step                         |
+|  lambda         |  0.3    | Poisson rate parameter for resource offspring number produced during a time step                 |
+|  agent_view     |  10     | How far managers can see on the landscape for resource counting when observe_type = 0            |
+|  res_birth_K    |  100000 | Carrying capacity applied to the number of resources added during a time step                    |
+|  res_death_K    |  2000   | Carrying capacity applied to the number of resources removed during a time step                  |
+|  res_move_type  |  1      | Type of resource movement (default is up to res_movement cells in any direction)                 |
+|  observe_type   |  0      | Type of resource observation (default is density-based; i.e, counting a subset on the landscape) |
+|  fixed_mark     |  50     | For mark-recapture observation (observe_type = 1), number of marked resources                    |
+|  fixed_recapt   |  150    | For mark-recapture observation (observe_type = 1), number of recaptured resources                |
+|  times_observe  |  1      | For density-based observation (observe_type = 0), landscape subsets viewed during observation    |
+|  res_consume    |  0.5    | Pr. of a landscape cell's value reduced by the presence of a resource in a time step             |
+|  max_ages       |  5      | The maximum number of time steps a resource can persist before it is removed                     |
+|  minimum_cost   |  10     | The minimum cost of a user performing any action                                                 |
+|  user_budget    |  1000   | A user's budget per time step for performing any number of actions                               |
+|  manager_budget |  1000   | A manager's budget per time step for setting policy                                              |
+|  manage_target  |  1000   | The manager's target resource abundance                                                          |
+|  RESOURCE_ini   |  1000   | The initial abundance of resources                                                               |
+|  scaring        |  FALSE  | Resource scaring (moves a resource to a random landscape cell) is a policy option                |
+|  culling        |  TRUE   | Resource culling (removes a resource entirely) is a policy option                                |
+|  castration     |  FALSE  | Resource castration (sets a resource's lambda to zero) is a policy option                        |
+|  feeding        |  FALSE  | Resource feeding (increases a resource's lambda) is a policy option                              |
+|  help_offspring |  FALSE  | Resource helping (increases a resource's offspring number) is a policy option                    |
+|  tend_crops     |  FALSE  | Users can increase landscape cell values                                                         |
+|  tend_crop_yld  |  0.2    | Proportional increase per landscape cell from tend_crops action                                  |
+|  kill_crops     |  FALSE  | Users can decrease landscape cell values to zero                                                 |
+|  stakeholders   |  4      | Number of users in the simulation                                                                |
+|  land_ownership |  FALSE  | Users own land and increase utility indirectly from landscape instead of resource use            |
+|  manage_freq    |  1      | Frequency (in time steps) with which managers revise and enact policy                            |
+|  public_land    |  0      | Proportion of land that is public (un-owned by users) if land_ownership = TRUE                   |
+
+\normalsize
+
+Table 1: Select parameter values for initialising generalised management strategy evaluation simulations
+
 Where available, we use estimated demographic parameter values from @Johnson2018 and @AEWA2016.

### 2018-05-25 notebook/ms/ms.Rmd
Remove MS text to barely get over the hurdle of 3475 words

@@
 author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
-  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
-  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy
+  Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
 date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
@@
   html_document: default
+  word_document: default
 header-includes:
@@
 
-Game theory is the formal study of strategic interactions, and can therefore be applied to modelling stakeholder actions and addressing issues of cooperation and conflict in conservation [@Colyvan2011; @Lee2012; @Kark2015; @Adami2016; @Tilman2016].
-In game-theoretic models, agents adopt strategies to make decisions that maximise some type of payoff (e.g., utility, biological fitness).
-Agents are constrained in their decision-making, and realised pay-offs depend on decisions made by other agents.
-In simple models, it is often useful to assume that agents are perfectly rational decision-makers, then find optimal solutions for pay-off maximisation mathematically.
-But models that permit even moderately complex decision-making strategies or pay-off structures often include more possible strategies than are mathematically tractable [@Hamblin2013].
-In these models, genetic algorithms, which mimic the process of natural selection (mutation, recombination, selection, reproduction), can find adaptive (i.e., practical, but not necessarily optimal) solutions for game strategies [e.g., @Balmann2000; @Tu2000; @Hamblin2013].
-
 Consistent with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for agents (stakeholders).
@@
 The highest fitness strategy in the population then becomes the stakeholder's new strategy.
-
-GMSE arguments and output
----------------------------------------------------
-
-Simulations using the default GMSE sub-models described above are run using the `gmse` function, which offers a range of options for setting parameter values (see Table 1 for some select examples).
-Output of `gmse` is an exhaustive list that includes all resources and observations, all stakeholder decisions and actions, and all landscape properties in each time step of the simulation.
-Results are most easily interpreted visually, so a summary of simulation dynamics is plotted by default (the plot can also be called using the `plot_gmse_results` function, and summaries of results can be obtained using `gmse_summary` and `gmse_table`).
-An example below shows how simulations are set and interpreted.
-
-\small
-
-|  Argument       | Default | Description                                                                                      |
-|-----------------|---------|--------------------------------------------------------------------------------------------------|
-|  time_max       |  100    | Maximum time steps in simulation                                                                 |
-|  land_dim_1     |  100    | Width of the landscape (horizontal cells)                                                        |
-|  land_dim_2     |  100    | Height of the landscape (vertical cells)                                                         |
-|  res_movement   |  20     | Distance (cells) a resource can move in any direction (for movement rules, see res_move_type)    |
-|  remove_pr      |  0      | Density-independent probability of resource mortality during a time step                         |
-|  lambda         |  0.3    | Poisson rate parameter for resource offspring number produced during a time step                 |
-|  agent_view     |  10     | How far managers can see on the landscape for resource counting when observe_type = 0            |
-|  res_birth_K    |  100000 | Carrying capacity applied to the number of resources added during a time step                    |
-|  res_death_K    |  2000   | Carrying capacity applied to the number of resources removed during a time step                  |
-|  res_move_type  |  1      | Type of resource movement (default is up to res_movement cells in any direction)                 |
-|  observe_type   |  0      | Type of resource observation (default is density-based; i.e, counting a subset on the landscape) |
-|  fixed_mark     |  50     | For mark-recapture observation (observe_type = 1), number of marked resources                    |
-|  fixed_recapt   |  150    | For mark-recapture observation (observe_type = 1), number of recaptured resources                |
-|  times_observe  |  1      | For density-based observation (observe_type = 0), landscape subsets viewed during observation    |
-|  res_consume    |  0.5    | Pr. of a landscape cell's value reduced by the presence of a resource in a time step             |
-|  max_ages       |  5      | The maximum number of time steps a resource can persist before it is removed                     |
-|  minimum_cost   |  10     | The minimum cost of a user performing any action                                                 |
-|  user_budget    |  1000   | A user's budget per time step for performing any number of actions                               |
-|  manager_budget |  1000   | A manager's budget per time step for setting policy                                              |
-|  manage_target  |  1000   | The manager's target resource abundance                                                          |
-|  RESOURCE_ini   |  1000   | The initial abundance of resources                                                               |
-|  scaring        |  FALSE  | Resource scaring (moves a resource to a random landscape cell) is a policy option                |
-|  culling        |  TRUE   | Resource culling (removes a resource entirely) is a policy option                                |
-|  castration     |  FALSE  | Resource castration (sets a resource's lambda to zero) is a policy option                        |
-|  feeding        |  FALSE  | Resource feeding (increases a resource's lambda) is a policy option                              |
-|  help_offspring |  FALSE  | Resource helping (increases a resource's offspring number) is a policy option                    |
-|  tend_crops     |  FALSE  | Users can increase landscape cell values                                                         |
-|  tend_crop_yld  |  0.2    | Proportional increase per landscape cell from tend_crops action                                  |
-|  kill_crops     |  FALSE  | Users can decrease landscape cell values to zero                                                 |
-|  stakeholders   |  4      | Number of users in the simulation                                                                |
-|  land_ownership |  FALSE  | Users own land and increase utility indirectly from landscape instead of resource use            |
-|  manage_freq    |  1      | Frequency (in time steps) with which managers revise and enact policy                            |
-|  public_land    |  0      | Proportion of land that is public (un-owned by users) if land_ownership = TRUE                   |
-
-\normalsize
-
-Table 1: Select parameter values for initialising generalised management strategy evaluation simulations
 

### 2018-05-25 notebook/ms/SI1.Rmd

@@
 ================================================================================
+
+Game theory is the formal study of strategic interactions, and can therefore be applied to modelling stakeholder actions and addressing issues of cooperation and conflict in conservation [@Lee2012; @Kark2015; @Adami2016; @Tilman2016; @Redpath2018].
+In game-theoretic models, agents adopt strategies to make decisions that maximise some type of payoff (e.g., utility, biological fitness).
+Agents are constrained in their decision-making, and realised pay-offs depend on decisions made by other agents.
+In simple models, it is often useful to assume that agents are perfectly rational decision-makers, then find optimal solutions for pay-off maximisation mathematically.
+But models that permit even moderately complex decision-making strategies or pay-off structures often include more possible strategies than are mathematically tractable [@Hamblin2013].
+In these models, genetic algorithms, which mimic the process of natural selection (mutation, recombination, selection, reproduction), can find adaptive (i.e., practical, but not necessarily optimal) solutions for game strategies [e.g., @Balmann2000; @Tu2000; @Hamblin2013].
 

### 2018-07-08 notebook/opportunities/IAPETUS_AI_PhD.Rmd
Add PhD Advertisement

---
title: "PhD Studentship: Using artificial intelligence to improve decision-making in conservation conflicts"
author: ""
date: ""
output: html_document
---

********************************************************************************

**We are looking for a PhD student to work with us on conservation conflict at the [University of Stirling](http://stir.ac.uk), in partnership with [Newcastle University](https://www.ncl.ac.uk/) and [Scottish Natural Heritage](https://www.nature.scot/).**

********************************************************************************

Project Description
-----------------------------------------------

The conservation of biodiversity is increasingly coming into conflict with stakeholders who rely on natural resources for their livelihoods and well-being [1]. Long-term conservation of populations therefore requires an understanding of conflict among stakeholders, and a strategy for managing conflict effectively. Conflict and cooperation between rational decision-makers can be studied formally using game theory. But conservation conflicts are highly complex, meaning that simple game-theoretic models are unlikely to understand and predict human decision-making [2].

Meanwhile, artificial intelligence has shown rapid progress in recent years, particularly in the context of decision-making in complex games [3]. The use of flexible artificial intelligence provides a promising avenue for investigating situations of human conflict and finding solutions to major global problems. For conservation conflicts, the initial development of artificial intelligence in stakeholder decision-making is available in the recently published GMSE R package (Generalised Management Strategy Evaluation) [4], which uses genetic algorithms to model human decision-making and  its effect populations [5]. Nevertheless, modelling complex negotiations, interactions among stakeholders, and the biological impact on populations remains a challenge, and requires new conceptual ideas and modelling techniques.

This project will develop and apply artificial intelligence to simulate the decision-making of stakeholders under situations of conservation conflict and the biological response of populations. Stakeholders and organisms will be modelled discretely using an agent-based modelling framework, allowing for potentially complex interactions within simulations [6].

<br>

![](https://raw.githubusercontent.com/bradduthie/gmse/1727ea37f32f0f40df8ee6e8277d0d1723c88639/notebook/images/GMSE_logo_name.png)

<br>

The primary objective is to develop a flexible tool for the agent-based modelling of stakeholder decision-making and its  ecological consequences under conditions of conservation conflict, and to then apply this tool to one or more real-world case studies. Additional objectives might include the modelling and simulation of multi-species ecological systems under management, or the development of a graphical user interface for model users.

The student should have a background or very strong interest in programming (ideally in R, C, or C++), and an interest in applying social-ecological modelling to conservation biology. They will benefit from a team of experts in modelling and game-theory, and will interact more broadly with a network of researchers involved in case studies of real-world conservation conflicts.

Application Procedure
-----------------------------------------------

To apply, please email Dr. Duthie [alexander.duthie@stir.ac.uk](mailto:alexander.duthie@stir.ac.uk) (CC Dr. Mill [aileen.mill@ncl.ac.uk](mailto:aileen.mill@ncl.ac.uk)  and Prof. Bunnefeld [nils.bunnefeld@stir.ac.uk](nils.bunnefeld@stir.ac.uk)) with (1) a one page cover letter indicating why you are interested in this project, (2) your CV with contact information for two references, and (3) full transcripts of previous qualifications obtained. The deadline for application is 27 JUL 2018.

Funding Notes
-----------------------------------------------

Please note that this project is in competition with others for funding, and success will depend on the quality of applications received.

References
-----------------------------------------------

[1] Redpath, SM, et al. (2013). Understanding and managing conservation conflicts. Trends in Ecology & Evolution, 28(2), 100-109.

[2] Redpath, SM, et al. (2018). Games as tools to address conservation conflicts. Trends in Ecology and Evolution, 33(6), 415-426.

[3] Moravčík, M, et al. (2017). DeepStack: Expert-level artificial intelligence in heads-up no-limit poker. Science, 356(6337), 508-513.

[4] Duthie, AB, et al. (2018). GMSE: an R package for generalised management strategy evaluation. Methods in Ecology and Evolution. In review. (pre-print here)

[5] Hamblin, S (2013). On the practical usage of genetic algorithms in ecology and evolution. Methods in Ecology and Evolution, 4(2), 184-194.

[6] McLane, AJ, et al. (2011). The role of agent-based models in wildlife ecology and management. Ecological Modelling, 222(8), 1544–1556.

### 2018-07-08 notebook/opportunities/IAPETUS_AI_PhD.Rmd
Add PhD opp

@@
 The use of flexible artificial intelligence provides a promising avenue for investigating situations of human conflict and finding solutions to major global problems.
-For conservation conflicts, the initial development of artificial intelligence in stakeholder decision-making is available in the recently published GMSE R package (Generalised Management Strategy Evaluation) [4], which uses genetic algorithms to model human decision-making and  its effect populations [5].
+For conservation conflicts, the initial development of artificial intelligence in stakeholder decision-making is available in the recently published [GMSE R package](https://confoobio.github.io/gmse/) (Generalised Management Strategy Evaluation) [4], which uses genetic algorithms to model human decision-making and  its effect populations [5].
 Nevertheless, modelling complex negotiations, interactions among stakeholders, and the biological impact on populations remains a challenge, and requires new conceptual ideas and modelling techniques.

### 2018-07-24 notebook/opportunities/IAPETUS_AI_PhD.Rmd
add IAPETUS stuff

@@
 ---
-title: "PhD Studentship: Using artificial intelligence to improve decision-making in conservation conflicts"
-author: ""
-date: ""
-output: html_document
+title: 'PhD Studentship: Using artificial intelligence to improve decision-making
+  in conservation conflicts'
+author: ''
+date: ''
+output:
+  html_document: default
+  word_document: default
 ---
@@
 
-[4] Duthie, AB, et al. (2018). GMSE: an R package for generalised management strategy evaluation. Methods in Ecology and Evolution. In review. (pre-print here)
+[4] Duthie, AB, et al.
+(2018).
+GMSE: an R package for generalised management strategy evaluation.
+Methods in Ecology and Evolution.
+In review.
+(pre-print [here](https://www.biorxiv.org/content/early/2018/05/16/221432))
 

### 2018-07-25 R/gmse_summary.R
New website documents

@@ -184,6 +184,5 @@ gmse_table <- function(gmse_sim, hide_unused_options = TRUE, all_time = TRUE){
         results   <- results[last_step,];
     }
     
-    
     return(results);
 }

### 2018-07-25 tests/testthat/test-user.R

@@ -92,9 +92,9 @@ test_that("Users act according to costs", {
     expect_equal(max(usem[[5]][,1:7,2]), 100001);
     expect_equal(min(usem[[5]]), 10);
     expect_equal(max(usem[[5]]), 100001);
-    expect_equal(usem[[4]][1,8:13,2], c(0, 0, 0, 3, 0, 3));
-    expect_equal(usem[[4]][2,8:13,2], c(0, 1, 0, 0, 1, 0));
-    expect_equal(usem[[4]][3,8:13,2], c(0, 0, 0, 2, 0, 0));
+    expect_equal(usem[[4]][1,8:13,2], c(0, 0, 1, 0, 0, 0));
+    expect_equal(usem[[4]][2,8:13,2], c(0, 0, 0, 0, 0, 0));
+    expect_equal(usem[[4]][3,8:13,2], c(0, 0, 3, 6, 0, 0));
     expect_equal(max(usem[[4]][4:7,8:13,2]), 0);
 })
 

### 2018-08-17 notebook/ms/ms.Rmd
Address Reviewer 1 minor comment 1

@@
 1).
-The workhorse function `gmse` runs simulations using four predefined submodels, which can be parameterised to fit various case studies; more tailored submodels can also be defined using the `gmse_apply` function (see [SI2](SI1.pdf)).
-(1) A population of discrete resources (e.g., a managed species) with individual traits (e.g., location, age) is modelled on a spatially-explicit landscape and can simulate resource birth, movement, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty.
+The function `gmse` runs simulations using four predefined submodels, which can be parameterised to fit various case studies; more tailored submodels can also be defined using the `gmse_apply` function (see [SI2](SI1.pdf)).
+By default, GMSE models (1) a population of discrete resources (e.g., a managed species) with individual traits (e.g., location, age) on a spatially-explicit landscape and simulates resource birth, individual movement between landscape cells, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty.
 This sub-model is unique in not relying on other sub-models because ecological dynamics can be simulated in the absence of observation and management.
@@
 Policy is developed by calling the genetic algorithm (see below), which works within a manager's constraints to find costs for user actions on the resource (e.g., culling, scaring, etc.) that minimise deviation from the target abundance, as informed by the predicted consequences of each action on resource abundance and user action histories.
-After a suitable policy is found, (4) users perform actions that affect resources or landscape cells.
+After a suitable policy is found, (4) users perform actions that affect resources or landscape cells (e.g., culling, which causes immediate resource death).
 Users respond to policy individually, each calling the genetic algorithm to find actions that maximise their own utilities (e.g., maximise resource use or landscape yield) within their imposed constraints.

### 2018-08-17 notebook/ms/SI3.Rmd
Reviewer 1 minor comment 2

@@
 Simulations using the default GMSE sub-models described above are run using the `gmse` function, which offers a range of options for setting parameter values (see Table 1 for some select examples).
-Output of `gmse` is an exhaustive list that includes all resources and observations, all stakeholder decisions and actions, and all landscape properties in each time step of the simulation.
+Output of `gmse` is an exhaustive list that includes all resources and observations, all stakeholder decisions and actions, and all landscape properties in each time step of the simulation (see [SI7](SI7.pdf) for a description of key data structures).
 Results are most easily interpreted visually, so a summary of simulation dynamics is plotted by default (the plot can also be called using the `plot_gmse_results` function, and summaries of results can be obtained using `gmse_summary` and `gmse_table`).
@@
 |  land_dim_2     |  100    | Height of the landscape (vertical cells)                                                         |
-|  res_movement   |  20     | Distance (cells) a resource can move in any direction (for movement rules, see res_move_type)    |
+|  res_movement   |  20     | Distance (cells) a resource can move in any direction (see also res_move_type)                   |
 |  remove_pr      |  0      | Density-independent probability of resource mortality during a time step                         |
@@
 |  fixed_recapt   |  150    | For mark-recapture observation (observe_type = 1), number of recaptured resources                |
-|  times_observe  |  1      | For density-based observation (observe_type = 0), landscape subsets viewed during observation    |
+|  times_observe  |  1      | For density-based observation (observe_type = 0), landscape subsets observed                     |
 |  res_consume    |  0.5    | Pr. of a landscape cell's value reduced by the presence of a resource in a time step             |
@@
 In [SI4](SI4.pdf), we show an example of this using the same Taiga Bean Geese case study that we did here.
-
-\clearpage

### 2018-08-17 notebook/ms/SI5.Rmd
Attempt to addres Reviewer 1 minor comment 4 -- note multiple attempts make it clear that axis labels cannot be adjusted in FLR, so explain why this is the case

@@
 Diagnostic plots, identical to those of the [modelling stock-recruitment tutorial](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html) for the `nsher_ri` example, are shown below.
+We note that these plots are made using the `FLCore` package, and are not produced by, nor available in, the `GMSE` package.
 
-```{r, fig.height = 6.2, fig.cap = "Output of the FLR plot function for an example Ricker model of stock recruitment on North Sea Herring data."}
-plot(newFL);
+```{r, fig.height = 5.2, fig.cap = "Output of the FLR plot function for an example Ricker model of stock recruitment on North Sea Herring data."}
+plot(newFL, cex = 0.7);
 ```

### 2018-08-17 notebook/ms/SI3.Rmd
Add half of the response to Reviewer 1 minor comment 4. Also consider simply adding a new SI (probably not best because this will require yet more things to link to, possibly generating confusion)

@@
 
-Table 1: Select parameter values for initialising generalised management strategy evaluation simulations
+Table 1: Select parameter values for initialising generalised management strategy evaluation simulations. See [below](#nondefaults) for explanation of non-default values.
 
@@
 
+Non-default options in default GMSE sub-models {#nondefault}
+--------------------------------------------------------------------------------
+
+Within the default submodels of GMSE, there are several non-default options that can be set using arguments to `gmse` and `gmse_apply`.
+Brief explanations of these non-default options appear in the [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf) (also [listed](https://confoobio.github.io/gmse/reference/gmse.html) on the [GMSE website](https://confoobio.github.io/gmse/)).
+Here we further explain some of the less obvious or less intuitive options available in GMSE v0.4.0.6, for which additional explanation beyond the GMSE documentation may be warranted.
+Future versions of GMSE are expected to expand upon these options.
+
+**res_move_type**: Resources in default GMSE v0.4.0.7 sub-models can move in one of three ways.
+These different movement types are listed below, and are taken as arguments to `gmse` (e.g., `gmse(res_move_type = 2)`).
+
+  0. *No movement*. Given this option, resources do not move at all in the default resource sub-model, and instead remain on the location at which they are initialised.
+  1.
+*Uniform movement in any direction up to `res_movement` cells away from a resource's original location* (default).
+In uniform movement, the resource moves to a new cell within some distance `res_movement` of its original location in each time step.
+Distance therefore includes diagonal movement, such that a resource capable of moving `res_movement = 1` cell away can go to any of the eight adjacent cells around its current location; it can also remain on its current cell.
+Similarly, a resource capable of moving `res_movement = 2` cells away can go to any of the 24 cells surrounding its current location, or remain on its current cell.
+The distance (in cells) moved in the x-direction (i.e., left to right) and y-direction (i.e., up and down), is sampled from a uniform distribution, meaning that all reachable cells (including the resource's current cell) have equal probability of being selected as the resource's new cell location.
+  2.
+*Poisson selected movement distance and direction*.
+In this movement type, the distance (in cells) moved in the x-direction and the y-direction are sampled from a Poisson distribution using `res_movement` as the rate parameter (i.e., the mean movement distance in each direction).
+Using this movement type is not generally recommended because it is not likely to be biologically realistic (e.g., diagonal movement is rare because selection of both high x-direction and high y-direction values is unlikely).
+  3.
+*Uniform movement in any direction up to `res_movement` cells away from a resource's original location `res_movement` times.* Like option (1), when moving, resources can travel in any direction up to `res_movement` cells from their current location with equal probability.
+What is different about this option is that uniform movement in any direction is occurs $C =$ Poisson(`res_movement`) times.
+In other words, when a resource moves, it first decides how many different cells it will visit ($C$), as determined by sampling from a Poisson distribution with a rate parameter of `res_movement` (i.e., the mean number of visited cells will be `res_movement`).
+Upon leaving a cell, the resource will move in the same way as defined in option (1) above, by choosing a cell within `res_movement` distance (in cells) of its current cell.
+It continues moving in this way until $C$ cells have been visited.
+This type of movement has been used in some individual-based models of plant-pollinator interactions [e.g., @Duthie2013; @Wilson2003].
+
+**res_death_type**: Resources can be removed from the model in one of three ways within the default resource sub-model, thereby simulating resource death.
+It is important to note that `res_death_type` refers only to removal that occurs within the resource sub-model; removal from user actions (e.g., death from shooting) occurs within the user sub-model.
+
+  1.
+*Density-independent removal*.
+Given this option, there is no density-dependent removal of resources within the resource sub-model.
+Resources are simply removed with a probability of `removal_pr` each time the submodel is run.
+This option should be used with great care because in the absence of density-dependence, it is possible for the population of resources to increase without limit and thereby greatly slow down simulation times.
+  2.
+*Density-dependent removal*.
+Given this option, removal of resources within the resource sub-model is caused entirely by density effects.
+The probability that a resource is removed is a function of the resource carrying capacity set using the `res_death_K` parameter.
+If the number of resources ($N$) is greater than `res_death_K` ($K$), then the probability that an individual resource is removed is defined by $(N - K) / N$.
+Removal then occurs independently for each resource based on this probability.
+  3.
+*Density-independent and density-dependent removal*.
+This option allows a combination of both options (1) and (2), which affect removal independently.
+In other words, each resource is assigned a probability of removal `removal_pr` as in (1), and then independently assigned another probability of removal based on `res_death_K` as in (2).
+If removal occurs as a consequence of (1), (2), or both, then the resource is removed.
+
+**observe_type**: Observation of resources in the observation model can occur in one of four ways.
+
+**obs_move_type**:
+
 Conclusions and future development

### 2018-08-19 notebook/ms/SI5.Rmd
Definitively address Reviewer 1 minor comment 4 using the ggplotFL package

@@
 To build the stock-recruitment model, the `FLCore` package is needed [@Kell2007].
+We also include the `ggplotFL` package for plotting.
 
@@
 install.packages("FLCore", repos="http://flr-project.org/R");
+install.packages("ggplotFL", repos="http://flr-project.org/R")
 ```
 
-To start, we need to read in the `FLCore` and GMSE libraries.
+To start, we need to read in the `FLCore`, `ggplotFL` and `GMSE` libraries.
 
@@
 library(FLCore);
+library(ggplotFL);
 library(GMSE);
@@
 
-Diagnostic plots, identical to those of the [modelling stock-recruitment tutorial](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html) for the `nsher_ri` example, are shown below.
-We note that these plots are made using the `FLCore` package, and are not produced by, nor available in, the `GMSE` package.
+Diagnostic plots, identical to those of the [modelling stock-recruitment tutorial](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html) for the `nsher_ri` example, are shown below in Figure 1.
+We note that these plots are made using the `FLCore` and `ggplotFL` packages, and are not produced by, nor available in, the `GMSE` package.
 

### 2018-08-20 notebook/ms/SI3.Rmd
Address Reviewer 1 minor comment 5 with a new section of SI3 on non-default parameter values

@@
 Output of `gmse` is an exhaustive list that includes all resources and observations, all stakeholder decisions and actions, and all landscape properties in each time step of the simulation (see [SI7](SI7.pdf) for a description of key data structures).
-Results are most easily interpreted visually, so a summary of simulation dynamics is plotted by default (the plot can also be called using the `plot_gmse_results` function, and summaries of results can be obtained using `gmse_summary` and `gmse_table`).
-An example below shows how simulations are set and interpreted.
 
@@
 |  res_death_K    |  2000   | Carrying capacity applied to the number of resources removed during a time step                  |
-|  res_move_type  |  1      | Type of resource movement (default is up to res_movement cells in any direction)                 |
-|  observe_type   |  0      | Type of resource observation (default is density-based; i.e, counting a subset on the landscape) |
+|  res_death_type |  1      | Rules affecting resource death (default is density-dependent; [see below](#nondefault))          |
+|  res_move_type  |  1      | Type of resource movement (default moves uniformly in any direction; [see below](#nondefault))   |
+|  observe_type   |  0      | Type of resource observation (default is density-based; [see below](#nondefault))                |
+|  obs_move_type  |  1      | How agents (manager and stakeholders) move (typically ignored; [see below](#nondefault))         |
 |  fixed_mark     |  50     | For mark-recapture observation (observe_type = 1), number of marked resources                    |
@@
 
-Table 1: Select parameter values for initialising generalised management strategy evaluation simulations. See [below](#nondefaults) for explanation of non-default values.
+Table 1: Select parameter values for initialising generalised management strategy evaluation simulations.
+See [below](#nondefault) for explanation of non-default values of resource death and resource observation processes.
+
+Results are most easily interpreted visually, so a summary of simulation dynamics is plotted by default (the plot can also be called using the `plot_gmse_results` function, and summaries of results can be obtained using `gmse_summary` and `gmse_table`).
+An example below shows how simulations are set and interpreted.
 
@@
 
-Non-default options in default GMSE sub-models {#nondefault}
+Default and non-default options in default GMSE sub-models {#nondefault}
 --------------------------------------------------------------------------------
 
-Within the default submodels of GMSE, there are several non-default options that can be set using arguments to `gmse` and `gmse_apply`.
+Within the default sub-models of GMSE, there are several non-default options that can be set using arguments to `gmse` and `gmse_apply`.
 Brief explanations of these non-default options appear in the [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf) (also [listed](https://confoobio.github.io/gmse/reference/gmse.html) on the [GMSE website](https://confoobio.github.io/gmse/)).
-Here we further explain some of the less obvious or less intuitive options available in GMSE v0.4.0.6, for which additional explanation beyond the GMSE documentation may be warranted.
+Here we further explain some of the less obvious or less intuitive options available in GMSE v0.4.0.7, for which additional explanation beyond the GMSE documentation may be warranted for application to case studies such as the Taiga Bean Geese case study example demonstrated above.
 Future versions of GMSE are expected to expand upon these options.
 
-**res_move_type**: Resources in default GMSE v0.4.0.7 sub-models can move in one of three ways.
+**res_move_type**: Resources in default GMSE v0.4.0.7 sub-models can move on the landscape in one of three ways.
 These different movement types are listed below, and are taken as arguments to `gmse` (e.g., `gmse(res_move_type = 2)`).
 
-  0. *No movement*. Given this option, resources do not move at all in the default resource sub-model, and instead remain on the location at which they are initialised.
+  0. *No movement*. Given this option, resources do not move at all in the resource sub-model, and instead remain on the location at which they are initialised.
   1.
@@
 *Uniform movement in any direction up to `res_movement` cells away from a resource's original location `res_movement` times.* Like option (1), when moving, resources can travel in any direction up to `res_movement` cells from their current location with equal probability.
-What is different about this option is that uniform movement in any direction is occurs $C =$ Poisson(`res_movement`) times.
+What is different about this option is that uniform movement in any direction occurs $C =$ Poisson(`res_movement`) times.
 In other words, when a resource moves, it first decides how many different cells it will visit ($C$), as determined by sampling from a Poisson distribution with a rate parameter of `res_movement` (i.e., the mean number of visited cells will be `res_movement`).
@@
 
-**res_death_type**: Resources can be removed from the model in one of three ways within the default resource sub-model, thereby simulating resource death.
+**res_death_type**: Resources in GMSE v0.4.0.7 can be removed from the model in one of three ways within the default resource sub-model, thereby simulating resource death.
 It is important to note that `res_death_type` refers only to removal that occurs within the resource sub-model; removal from user actions (e.g., death from shooting) occurs within the user sub-model.
@@
 Given this option, there is no density-dependent removal of resources within the resource sub-model.
-Resources are simply removed with a probability of `removal_pr` each time the submodel is run.
+Resources are simply removed with a probability of `removal_pr` each time the sub-model is run.
 This option should be used with great care because in the absence of density-dependence, it is possible for the population of resources to increase without limit and thereby greatly slow down simulation times.
   2.
-*Density-dependent removal*.
+*Density-dependent removal* (default).
 Given this option, removal of resources within the resource sub-model is caused entirely by density effects.
@@
 
-**observe_type**: Observation of resources in the observation model can occur in one of four ways.
+**observe_type**: Observation of resources in default GMSE v0.4.0.7 sub-models can occur in one of four ways.
+Each observation type mimics some process of resource observation, with potential error affected by observation intensity.
 
-**obs_move_type**:
+  0.
+*Density-based observation*.
+In this option, managers observe resources on a subset of the landscape; subset size is determined by the manager's view as set using the `agent_view` parameter.
+In practice, this is done by having the manager pick a cell on the landscape at random, then count all of the resources within a distance (in any direction) of `agent_view` cells.
+Managers then extrapolate the density of resources within this subset to estimate the total number of resources on the landscape.
+For example, if the manager has an `agent_view` of 1, then they are capable of seeing all of the resources on nine landscape cells (the cell on which they are located, and each adjacent cell).
+If they count that there are 90 resource on these nine cells, then they calculate a resource density of 10 per cell.
+They then multiply the density of 10 by the total number of landscape cells to estimate the number of resources on the landscape.
+The error of this estimate naturally decreases with increasing `agent_view`, such that when the manager is capable of viewing the entire landscape, observation error is zero.
+If desired, the `times_observe` (default equals 1) option can also be used to allow managers to sample more than one landscape subset in their estimate (e.g., if `times_observe = 2`, then density is estimated from two different locations, with the mean of individual estimates used for a combined estimate of population density).
+  1.
+*Mark-recapture observation*.
+This option simulates the process of a mark-recapture analysis.
+In this process, managers randomly sample `fixed_mark` resources in the population; sampling occurs without replacement and without any spatial bias, and if `fixed_mark` is greater than the total number of resources on the landscape, then managers sample all resources.
+The manager then randomly samples `fixed_recapt` resources, again without replacement or spatial bias.
+The sampled resources from `fixed_mark` and `fixed_recapt` are interpreted as marked and recaptured resources, respectively, and a Chapman estimate [@Pollock2016] is then used in the manager model to calculate estimated population size form these observation data.
+Error in this estimate can be naturally minimised if both `fixed_mark` and `fixed_recapt` are very high.
+  2.
+*Sample linear transect*.
+This option simulates sampling from a linear transect of the landscape.
+The manager samples an entire set of cell rows on the landscape and counts the total resources on all cells; the manager then continues onto the next set of landscape rows until the entire landscape has been sampled.
+The number of rows observed in each sample is defined by `agent_view`.
+For example, if the landscape is of the dimensions $100 \times 100$ cells, and `agent_view = 2`, then the manager will first observe all resources in the upper 200 cells (first $2 \times 100$ cells in the upper two landscape cell rows), then move onto the next 200 cells, repeating the process 50 times until all rows have been sampled.
+This process of sampling will only generate observation error if `res_move_obs = TRUE`, which it is by default.
+In this case, resources can move on the landscape (according to the rules of `res_move_type` and distance of `res_movement`) in between transect observation, potentially causing some resources to escape observation and others to be observed more than once.
+  3.
+*Sample block transect*.
+This option is identical to the *sample linear transect* option above, except that instead of sampling an entire row of a landscape, the manager samples one square block of the landscape at a time.
+The cell length of each block side equals `agent_view`, meaning that the manager can observe all of the resources on `agent_view * agent_view` cells at a time.
+The manager proceeds to observe the entire landscape, block by block, until all cells on the landscape have been covered.
+For example, if `agent_view = 25`, and the landscape is of the dimensions $100 \times 100$, then the manager will sample four total blocks of the landscape, one at a time, and use the total number of resources observed as the estimate of resource density on the landscape.
+As with linear transect sampling, if `res_move_obs = TRUE` (which it is by default), then resources can move on the landscape between block samples, potentially causing observation error if some resources are missed due to movement, or counted multiple times.
+
+**obs_move_type**: Observation of agents (manager and users) in default GMSE v0.4.0.7 sub-models can occur in the same four ways as allowed in `res_move_type` (default `obs_move_type = 1`), with movement distance defined by the parameter `agent_move` (default 50).
+Currently, the only effect of this movement is to allow for spatially autocorrelated sampling of resources by the manager in the observation model.
+If the manager is using density-based sampling (`observe_type = 0`) and sampling more than one subset of the landscape (`times_observe > 1`), then the manager's movement between subsets is governed by `res_move_type` (movement rules) and `agent_move` (movement distance).
+Hence, to have the manager observe the population many times in one area of the landscape, the `times_observe` option should be increased above one and `agent_move` should be decreased to a low value.
 

### 2018-08-20 notebook/ms/SI1.Rmd
Update GMSE version number

@@
 For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'iteration' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the [main text](ms.pdf)).
-Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.4.0.3 (future versions of GMSE might expand upon this framework, and we highlight some of these potential avenues for expansion).
+Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.4.0.7 (future versions of GMSE might expand upon this framework, and we highlight some of these potential avenues for expansion).
 We first explain the [key data structures used](#Key_data_structures_used), then provide an overview of how a population of strategies is [initialised](#initialised), and the subsequent processes of [crossover](#crossover), [mutation](#mutation), [cost constraint](#constraint), [fitness](#eval) evaluation, [tournament selection](#tournament), and [replacement](#replacement).
@@
 
-Columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in predefined GMSE v0.4.0.3 manager and user sub-models (i.e., only one type of resource is permitted).
+Columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in predefined GMSE v0.4.0.7 manager and user sub-models (i.e., only one type of resource is permitted).
 Future versions might allow for different resource types (e.g., `Type_1` might be used to designate species, and `Type_2` and `Type_3` could designate stage or sex).
@@
 Here we explain how the fitnesses of candidate manager and user strategies in a `POPULATION` array (see above) are calculated.
-We emphasise that the fitness functions used in GMSE v0.4.0.3 are intended to be heuristic tools for identifying reasonable manager and user behaviours.
+We emphasise that the fitness functions used in GMSE v0.4.0.7 are intended to be heuristic tools for identifying reasonable manager and user behaviours.
 In practice, our fitness functions identify behaviours that are well-aligned with manager and user interests for harvesting or crop yield, but they are not intended to identify *optimal* decisions.
@@
 Note that $F_{i}^{u}$ increases when $\Delta A_{i}$ and $\Delta L_{i}$ are of the same sign as $U^{u}_{res}$ and $U^{u}_{land}$, respectively.
-Further, in GMSE v0.4.0.3, only one term of the equation is nonzero.
+Further, in GMSE v0.4.0.7, only one term of the equation is nonzero.
 When `land_ownership = FALSE` (default, modelling users that harvest resources), $U^{u}_{res} = -1$ and $U^{u}_{land} = 0$, and when `land_ownership = TRUE`, $U^{u}_{res} = 0$ and $U^{u}_{land} = 100$ (modelling farmers trying to increase crop yield).

### 2018-08-20 notebook/ms/SI7.Rmd
Update GMSE version number

@@
     reference_docx: docx_template.docx
-header-includes: null
-linestretch: 1
+header-includes: 
+- \usepackage{lineno}
+- \linenumbers
 link-citations: yes
@@
 While these default sub-models do not necessarily need to be used in every run in GMSE (see [use of](SI2.pdf) `gmse_apply`), they will be used in any run of the `gmse` function, and in any call of the `gmse_apply` function that does not run with entirely custom sub-models.
-Simulation and model inference do not require an understanding of the default data structures, but such an understanding can be especially useful when running `gmse_apply` if there is a need to extract uncommonly used information, change key simulated values (e.g., landscape properties, agent budgets, or resource movement rules as in [Supporting Information 4](SI4.pdf)), or build custom individual-based sub-models.
+Simulation and model inference do not require an understanding of the default data structures, but such an understanding can be especially useful when running `gmse_apply` if there is a need to extract uncommonly used information, change key simulated values (e.g., landscape properties, agent budgets, or resource movement rules, as in [SI4](SI4.pdf)), or build custom individual-based sub-models.
 Here we provide a brief explanation of the following key data structures (each name below is listed as it is named in the output `gmse_apply` when `get_res = "Full"`).
@@
 
-Note that these are not the only data structures used in GMSE, but they are the only ones that can be easily modified in GMSE v0.4.0.3 (see, e.g., [Supporting Infromation 4](SI4.pdf)), so they are the ones that we focus on here.
-Additionally, any custom subfunction that returns an array rather than a single value should adhere to the same structure as the these defaults if any default GMSE functions are to be used in `gmse_apply`.
+Note that these are not the only data structures used in GMSE, but they are the only ones that can be easily modified in GMSE v0.4.0.7 (see, e.g., [SI4](SI4.pdf)), so they are the ones that we focus on here.
+Additionally, any custom subfunction that returns an array rather than a single value should adhere to the same structure as these defaults if any default GMSE functions are to be used in `gmse_apply`.
 We can investigate each data structure by running a single simulation of `gmse_apply`.
@@
 In the case of columns 16-20, the value is either zero (if no action has occurred), or some positive integer that matches the ID of the agent that has performed the act (e.g., if column 17 equals 3, then that means that the agent with ID = 3 culled the resource in the corresponding row; where more than one agent's action is possible per time step -- as in scaring -- the integer reflects the most recently acting agent).
-We anticipate that future versions of gmse will contain multiple resource types, and might add rows to include additional resource traits.
+We anticipate that future versions of gmse will contain multiple resource types, and might add columns to include additional resource traits.
 
@@
 Each layer in `manager_array` has `r dim(sim$manager_array)[2]` columns, and a number of rows that varies depending on the number of agents and resource types.
-As of GMSE v0.4.0.3, only the first three rows are used.
+As of GMSE v0.4.0.7, only the first three rows are used.
 Two layers of `sim$manager_array` are shown below, the first being that of the manager and the second being that of the first user.
@@
 This is because there are several columns that do not map to actions per se, but properties of agents.
-As of GMSE v0.4.0.3, these properties cannot be changed by other agents.
+As of GMSE v0.4.0.7, these properties cannot be changed by other agents.
 Column of `user_array` are as follows.
@@
 Where the integer value is identical with the agent's own ID (e.g., row 3 in layer 1 where the element `sim$user_array[3, 1, 1] = 1`), actions affect all other agents in the model.
-As of GMSE v0.4.0.3, all rows except 1-3 are unused because agents do not affect one anothers actions individually; they either affect all other agents' actions indiscriminately (in the case of the manager setting policy) or do not (directly) affect other agents' actions at all (in the case of users).
+As of GMSE v0.4.0.7, all rows except 1-3 are unused because agents do not affect one anothers actions individually; they either affect all other agents' actions indiscriminately (in the case of the manager setting policy) or do not (directly) affect other agents' actions at all (in the case of users).
 This data structure, however, is designed so that future versions of GMSE will allow users to affect one another directly (representing, e.g., different groups of agents lobbying for different interests, among-user conflict, etc.).
@@
 In addition, all landscapes have three layers, which hold three separate values of information for each x-y location.
-The first layer is unused in GMSE v0.4.0.3; the second layer holds crop production on a cell, and the third layer holds the owner of the cell (corresponding to the ID of an agent, where the manager's ID = 0 defines public land).
+The first layer is unused in GMSE v0.4.0.7; the second layer holds crop production on a cell, and the third layer holds the owner of the cell (corresponding to the ID of an agent, where the manager's ID = 0 defines public land).
 An $8 \times 8$ portion of the landscape from `sim` is shown below.
@@
 
-In the case of the above, all of the cells in this square patch of landscape are owned by agent 1 (i.e., the manager; see `sim$LAND[,,3]`), and we can see that crop production on this patch of land has been decreased from 1 in several cells as a consequence of consumption by resources.
-In [Supporting Information 4](SI4.pdf), we show how landscape cell values can be manipulated to customise the placement of land ownership.
+In the case of the above, all of the cells in this square patch of landscape are owned by agent 1 (i.e., the manager; see `sim$LAND[,,3]`), and we can see that crop production on this patch of land has been decreased from 1 in several cells as a consequence of consumption by resources (see `sim$LAND[„2]`).
+In [SI4](SI4.pdf), we show how landscape cell values can be manipulated to customise the placement of land ownership.
 
@@
 We have focused on the data structures [`AGENTS`](#agents), [`resource_array`](#resources), [`observation_array`](#observation), [`manager_array`](#cost), [`user_array`](#action), and [`LAND`](#land) because these are the data structures that can be most readily manipulated to customise GMSE simulations.
-An example of how to do this within a loop using `gmse_apply` can be found in [Supporting Information 4](SI4.pdf).
+An example of how to do this within a loop using `gmse_apply` can be found in [SI4](SI4.pdf).
 While other data structures exist within GMSE (e.g., see the output of `gmse_apply` when `get_res = Full`), we do not recommend manipulating these structures for custom simulations.
 
-Many data structures contain elements that are unused in GMSE v0.4.0.3, and in all cases this is designed for ease of ongoing development of new GMSE features.
+Many data structures contain elements that are unused in GMSE v0.4.0.7, and in all cases this is designed for ease of ongoing development of new GMSE features.
 Requests for new features can be made on GitHub using the [GMSE Wiki](https://github.com/bradduthie/gmse/wiki) or the [GMSE Issues](https://github.com/bradduthie/gmse/issues) page.

### 2018-08-20 notebook/ms/SI3.Rmd
Rework SIs and MS with new plotting function

@@
 library(GMSE);
-plot_gmse_effort <- function(agents, paras, ACTION, COST){
+plot_gmse_effort <- function(sim_results){
+
+    agents <- sim_results$agents;
+    paras  <- sim_results$paras;
+    ACTION <- sim_results$action;
+    COST   <- sim_results$cost;
 
@@
 The lower left panel shows the cost of culling for farmers, as set by the manager, and the lower right panel shows the total number of culls attempted by farmers over time."}
-plot_gmse_results(res = sim_1$resource, obs = sim_1$observation, 
-                  land = sim_1$land, agents = sim_1$agents, paras = sim_1$paras, 
-                  ACTION = sim_1$action, COST = sim_1$cost);
+plot_gmse_results(sim_results = sim_1);
 ```
@@
 The right axis shows effort that farmers put into an action (coloured lines), which defined as the percentage of a farmer's budget put into a particular action (note, values might not add up to 100 because farmers are not forced to use their entire budget)."}
-plot_gmse_effort(sim_1$agents, sim_1$paras, ACTION = sim_1$action, 
-                 COST = sim_1$cost);
+plot_gmse_effort(sim_results = sim_1);
 ```
@@
 Simulation output is interpreted as in Figure 1."}
-plot_gmse_results(res = sim_2$resource, obs = sim_2$observation, 
-                  land = sim_2$land, agents = sim_2$agents, paras = sim_2$paras,
-                  ACTION = sim_2$action, COST = sim_2$cost);
+plot_gmse_results(sim_results = sim_2);
 ```
@@
 Simulation output is intepreted as in Figure 2."}
-plot_gmse_effort(sim_2$agents, sim_2$paras, ACTION = sim_2$action, 
-                 COST = sim_2$cost);
+plot_gmse_effort(sim_results = sim_2);
 ```

### 2018-08-20 notebook/ms/SI4.Rmd
Rework SIs and MS with new plotting function

--- 2018-05-15 notebook/ms/SI4.Rmd
@@
 As with default GMSE simulations, movement is based on a torus landscape (where no landscape edge exists, so that if resources move off of one side of the landscape they appear on the opposite side).
+We will use this custom function to modify `sim_old$resource_array` prior to running `gmse_apply`, thereby modelling a custom-built process affecting resource distribution that is integrated into GMSE.
 
@@
 ```{r}
-avoid_aggregation <- function(goose_table, land_dim_1 = 100, land_dim_2 = 100){
-    goose_number <- dim(goose_table)[1]    # How many geese are there?
+avoid_aggregation <- function(sim_resource_array, land_dim_1 = 100, 
+                              land_dim_2 = 100){
+    goose_number <- dim(sim_resource_array)[1]  # How many geese are there?
     for(goose in 1:goose_number){               # Loop through all rows of geese
-        x_loc  <- goose_table[goose, 5];
-        y_loc  <- goose_table[goose, 6];
-        shared <- sum(goose_table[,5] == x_loc & goose_table[,6] == y_loc);
+        x_loc  <- sim_resource_array[goose, 5];
+        y_loc  <- sim_resource_array[goose, 6];
+        shared <- sum( sim_resource_array[,5] == x_loc & 
+                       sim_resource_array[,6] == y_loc);
         if(shared > 10){
@@
             }
-            goose_table[goose, 5] <- new_x;
-            goose_table[goose, 6] <- new_y;
+            sim_resource_array[goose, 5] <- new_x;
+            sim_resource_array[goose, 6] <- new_y;
         }
     }
-    return(goose_table);
+    return(sim_resource_array);
 }
@@
     # Apply the new movement rules at the beginning of the loop
-    sim_old$resource_array  <- avoid_aggregation(sim_old$resource_array);
+    sim_old$resource_array  <- avoid_aggregation(sim_resource_array = 
+                                                     sim_old$resource_array);
     # Next, move on to simulate (old_list remembers that res_move_type = 0)

### 2018-08-20 notebook/ms/SI4.Rmd
Add notebook files and vinettes

@@
 header-includes: 
-- \usepackage{lineno}
-- \linenumbers
 linestretch: 1

### 2018-12-31 notebook/presentation/GMSE_presentation.Rmd
Update presentation

@@
 
-```{r, echo = FALSE}
-plot_gmse_effort <- function(agents, paras, ACTION, COST){
-
-    cols      <- c("green", "indianred1", "indianred3", "deepskyblue1",
-                   "deepskyblue2");
-
-    users    <- dim(agents[[1]])[1];
-    max_time <- dim(paras)[1];
-
-    para_vec <- paras[1,];
-
-    allowed  <- sum(para_vec[89:93]);
-
-    scar_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
-    cull_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
-    cast_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
-    feed_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
-    help_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
-
-    scar_act  <- matrix(data = 0, nrow = max_time, ncol = users);
-    cull_act  <- matrix(data = 0, nrow = max_time, ncol = users);
-    cast_act  <- matrix(data = 0, nrow = max_time, ncol = users);
-    feed_act  <- matrix(data = 0, nrow = max_time, ncol = users);
-    help_act  <- matrix(data = 0, nrow = max_time, ncol = users);
-
-    scar_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
-    cull_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
-    cast_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
-    feed_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
-    help_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
-
-    act_costs  <- matrix(data = 0, nrow = max_time, ncol = 5);
-    pol_effort <- matrix(data = 0, nrow = max_time, ncol = 5);
-    min_cost   <- para_vec[97];
-
-    for(time in 1:max_time){
-        #-- Scaring cost and actions
-        scar_cst[time, 1] <- COST[[time]][3, 8, 1];
-        for(user in 2:users){
-            scar_cst[time, user] <- COST[[time]][1, 8, user];
-        }
-        scar_act[time, 1] <- ACTION[[time]][3, 8, 1] - min_cost;
-        for(user in 2:users){
-            scar_act[time, user] <- ACTION[[time]][1, 8, user];
-        }
-        scar_eff[time,] <- scar_act[time,]*scar_cst[time,]/agents[[time]][,17];
-        #-- Culling cost and actions
-        cull_cst[time, 1] <- COST[[time]][3, 9, 1];
-        for(user in 2:users){
-            cull_cst[time, user] <- COST[[time]][1, 9, user];
-        }
-        cull_act[time, 1] <- ACTION[[time]][3, 9, 1] - min_cost;
-        for(user in 2:users){
-            cull_act[time, user] <- ACTION[[time]][1, 9, user];
-        }
-        cull_eff[time,] <- cull_act[time,]*cull_cst[time,]/agents[[time]][,17];
-        #-- Castration cost and actions
-        cast_cst[time, 1] <- COST[[time]][3, 10, 1];
-        for(user in 2:users){
-            cast_cst[time, user] <- COST[[time]][1, 10, user];
-        }
-        cast_act[time, 1] <- ACTION[[time]][3, 10, 1] - min_cost;
-        for(user in 2:users){
-            cast_act[time, user] <- ACTION[[time]][1, 10, user];
-        }
-        cast_eff[time,] <- cast_act[time,]*cast_cst[time,]/agents[[time]][,17];
-        #-- Feeding cost and actions
-        feed_cst[time, 1] <- COST[[time]][3, 11, 1];
-        for(user in 2:users){
-            feed_cst[time, user] <- COST[[time]][1, 11, user];
-        }
-        feed_act[time, 1] <- ACTION[[time]][3, 11, 1] - min_cost;
-        for(user in 2:users){
-            feed_act[time, user] <- ACTION[[time]][1, 11, user];
-        }
-        feed_eff[time,] <- feed_act[time,]*feed_cst[time,]/agents[[time]][,17];
-        #-- Helping cost and actions
-        help_cst[time, 1] <- COST[[time]][3, 12, 1];
-        for(user in 2:users){
-            help_cst[time, user] <- COST[[time]][1, 12, user];
-        }
-        help_act[time, 1] <- ACTION[[time]][3, 12, 1] - min_cost;
-        for(user in 2:users){
-            help_act[time, user] <- ACTION[[time]][1, 12, user];
-        }
-        help_eff[time,] <- help_act[time,]*help_cst[time,]/agents[[time]][,17];
-    }
-
-    # -- Turn these into percentages:
-    scar_eff <- scar_eff * 100;
-    cull_eff <- cull_eff * 100;
-    cast_eff <- cast_eff * 100;
-    feed_eff <- feed_eff * 100;
-    help_eff <- help_eff * 100;
-
-    par(mfrow = c(5, 1), mar = c(0, 0, 0, 0), oma = c(6, 6, 0, 6));
-    #---- Scaring
-    if(para_vec[89] == 1){
-        y1 <- 100;
-        y2 <- y1 * 1.2
-        plot(x = 1:max_time, y = 100 - scar_eff[,1], type = "l", lwd = 2, 
-             cex.axis = 1.5, xaxt = "n", ylim = c(0, y2), yaxt = "n");
-        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
-        par(new = TRUE);
-        plot(x = 1:max_time, y = scar_eff[,2], type = "n", lwd = 2, 
-             ylim = c(0, y2), xaxt="n", yaxt="n", 
-             cex.axis = 1.5);
-        for(stakeholder in 2:users){
-            points(x = 1:max_time, y = scar_eff[,stakeholder], type = "l", 
-                   lwd = 1, col = cols[1]);
-        }
-        axis(side = 4, at = c(50, 100),  labels = c(50, 100), cex.axis = 1.5);
-        legend(x = 1, y = y2 + y2*0.1, fill = cols[1:5], horiz = TRUE,
-               legend = c("scaring", "culling", "castration", "feeding", 
-                          "helping"), bty = "n", cex = 1);
-        abline(h = y1 + y1*0.1, lwd = 2);
-    }else{
-        plot(x = 1:max_time, y = scar_eff[,1], type = "n", xaxt= "n", 
-             yaxt = "n", ylim = c(0, 110));
-        text(x = 0, y = 90, cex = 2, labels = "No scaring allowed",
-             pos = 4);
-        abline(h = 100, lwd = 2);
-        legend(x = 1, y = 119, fill = cols[1:5], horiz = TRUE,
-               legend = c("scaring", "culling", "castration", "feeding", 
-                          "helping"), bty = "n", cex = 1);
-    }
-    #---- Culling
-    if(para_vec[90] == 1){
-        plot(x = 1:max_time, y = 100 - cull_eff[,1], type = "l", lwd = 2, 
-             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
-        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
-        par(new = TRUE);
-        plot(x = 1:max_time, y = cull_eff[,2], type = "n", lwd = 2, 
-             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
-        for(stakeholder in 2:users){
-            points(x = 1:max_time, y = cull_eff[,stakeholder], type = "l", 
-                   lwd = 1, col = cols[2]);
-        }
-        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
-    }else{
-        plot(x = 1:max_time, y = cull_act[,1], type = "n", xaxt= "n", 
-             yaxt = "n", ylim = c(0, 100));
-        text(x = 0, y = 90, cex = 2, labels = "No culling allowed",
-             pos = 4);
-    }
-    #---- Castrating
-    if(para_vec[91] == 1){
-        plot(x = 1:max_time, y = 100 - cast_eff[,1], type = "l", lwd = 2, 
-             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
-        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
-        par(new = TRUE);
-        plot(x = 1:max_time, y = cast_eff[,2], type = "n", lwd = 2, 
-             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
-        for(stakeholder in 2:users){
-            points(x = 1:max_time, y = cast_eff[,stakeholder], type = "l", 
-                   lwd = 1, col = cols[3]);
-        }
-        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
-    }else{
-        plot(x = 1:max_time, y = cast_act[,1], type = "n", xaxt= "n", 
-             yaxt = "n", ylim = c(0, 100));
-        text(x = 0, y = 90, cex = 2, labels = "No castration allowed",
-             pos = 4);
-    }
-    #---- Axes labels
-    mtext("Manager's permissiveness of user action", 
-          side = 2, line = 3.5, cex = 1.5, col = "black");
-    mtext("Percentage of user action effort expended", 
-          side = 4, line = 3.5, cex = 1.5, col = "black");
-    #---- Feeding
-    if(para_vec[92] == 1){
-        plot(x = 1:max_time, y = 100 - feed_eff[,1], type = "l", lwd = 2, 
-             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
-        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
-        par(new = TRUE);
-        plot(x = 1:max_time, y = feed_eff[,2], type = "n", lwd = 2, 
-             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
-        for(stakeholder in 2:users){
-            points(x = 1:max_time, y = feed_eff[,stakeholder], type = "l", 
-                   lwd = 1, col = cols[4]);
-        }
-        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
-    }else{
-        plot(x = 1:max_time, y = feed_act[,1], type = "n", xaxt= "n", 
-             yaxt = "n", ylim = c(0, 100));
-        text(x = 0, y = 90, cex = 2, labels = "No feeding allowed",
-             pos = 4);
-    }
-    #---- Helping
-    if(para_vec[93] == 1){
-        plot(x = 1:max_time, y = 100 - help_eff[,1], type = "l", lwd = 2, 
-             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
-        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
-        par(new = TRUE);
-        plot(x = 1:max_time, y = help_eff[,2], type = "n", lwd = 2, 
-             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
-        for(stakeholder in 2:users){
-            points(x = 1:max_time, y = help_eff[,stakeholder], type = "l", 
-                   lwd = 1, col = cols[5]);
-        }
-        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
-    }else{
-        plot(x = 1:max_time, y = help_act[,1], type = "n", cex.axis = 1.5,
-             yaxt = "n", ylim = c(0, 100));
-        text(x = 0, y = 90, cex = 2, labels = "No helping offspring allowed",
-             pos = 4);
-    }    
-    mtext("Time step", side = 1, line = 3.5, cex = 1.5, col = "black");
-}
-```
-
 <!---
@@
 ```{r, echo = TRUE, eval = FALSE}
-plot_gmse_results(res = sim$resource, obs = sim$observation, land = sim$land,
-                  agents = sim$agents, paras = sim$paras, ACTION = sim$action,
-                  COST = sim$cost);
+plot_gmse_results(sim_results = sim);
 ```
@@
 ```{r, echo = TRUE, eval = FALSE}
-plot_gmse_effort(agents = sim$agents, paras = sim$paras, ACTION = sim$action, 
-                 COST   = sim$cost);
+plot_gmse_effort(sim_results = sim);
 ```
@@
 ```{r, echo = FALSE, fig.height = 6, fig.width = 7}
-plot_gmse_results(res = sim$resource, obs = sim$observation, land = sim$land,
-                  agents = sim$agents, paras = sim$paras, ACTION = sim$action,
-                  COST = sim$cost);
+plot_gmse_results(sim_results = sim);
 ```
@@
 ```{r, echo = FALSE, fig.height = 6.5, fig.width = 6}
-plot_gmse_effort(agents = sim$agents, paras = sim$paras, ACTION = sim$action, 
-                 COST   = sim$cost);
+plot_gmse_effort(sim_results = sim);
 ```

### 2019-01-03 notebook/presentation/GMSE_presentation.Rmd
Remove unnecessary explanation of genetic algorithm

@@
 
+<style type="text/css">
+body p {
+  color: #000000;
+}
+ul {
+  color: black;
+}
+ol {
+  color: black;
+}
+a {
+    color: darkblue;
+    text-decoration: none
+}
+.fifty {
+   font-size: 50%;
+   }
+.seventy {
+   font-size: 70%;
+   }
+slides > slide.backdrop {
+  background: none !important;
+  background-color: white !important;
+}
+slides > slide.title-slide hgroup h1 {
+  color: black;
+}
+slides > slide.title-slide p {
+  color: black;
+}
+slides > slide.light > hgroup h2 {
+  color: black;
+}
+slides > slide.light > hgroup h1 {
+  color: black;
+}
+ul {
+  color: black;
+}
+ol {
+  color: black;
+}
+p {
+  color: black
+}
+h1, h2, h3, h4, h5, h6, h7, h8, p {
+    color: black;
+}
+.checkbox-inline, .radio-inline {
+    margin-right: 90px;
+}
+.forceBreak { -webkit-column-break-after: always; break-after: column; }
+</style>
+
 ```{r, echo=FALSE}
@@
 1. Know what to say about the Welikia project
-
 --->
@@
 https://welikia.org/explore/mannahatta-map/
-
 --->
@@
 
+<!---
+
 # Genetic algorithms, and an example using Prisoner's dilemma
-
-<!--- 
-    START OF THE GENETIC ALGORITHM SET OF SLIDES  ------------------------------
---->
 
@@
 
+--->
+
 <!---
-    END OF THE GENETIC ALGORITHM SET OF SLIDES  --------------------------------
---->
 
@@
 
+--->
+
 # Genetic algorithms in Management Strategy Evaluation

### 2019-01-03 notebook/presentation/GMSE_presentation.Rmd
Update presentation

--- 2018-12-31 notebook/presentation/GMSE_presentation.Rmd
@@
 ---
-title: "GMSE: a general tool for management strategy evaluation"
+title: "Generalised Management Strategy Evaluation: modelling biodiversity and food security"
 subtitle: "https://bradduthie.shinyapps.io/GMSE_presentation/#1"
-author: "Brad Duthie (@bradduthie)"
-date: "19 September 2017"
+author: "Brad Duthie (@bradduthie), Jeremy Cusack, Sarobidy Rakotonarivo, and Nils Bunnefeld"
+date: "8 January 2018"
 output: 
@@
 
+<style type="text/css">
+body p {
+  color: #000000;
+}
+ul {
+  color: black;
+}
+ol {
+  color: black;
+}
+a {
+    color: darkblue;
+    text-decoration: none
+}
+.fifty {
+   font-size: 50%;
+   }
+.seventy {
+   font-size: 70%;
+   }
+slides > slide.backdrop {
+  background: none !important;
+  background-color: white !important;
+}
+slides > slide.title-slide hgroup h1 {
+  color: black;
+}
+slides > slide.title-slide p {
+  color: black;
+}
+slides > slide.light > hgroup h2 {
+  color: black;
+}
+slides > slide.light > hgroup h1 {
+  color: black;
+}
+ul {
+  color: black;
+}
+ol {
+  color: black;
+}
+p {
+  color: black
+}
+h1, h2, h3, h4, h5, h6, h7, h8, p {
+    color: black;
+}
+.checkbox-inline, .radio-inline {
+    margin-right: 90px;
+}
+footer:after {
+  content: attr(label);
+  font-size: 12pt;
+  position: absolute;
+  bottom: 20px;
+  left: 60px;
+  line-height: 1.9;
+}
+div.footnotes {
+  position: absolute;
+  bottom: 0;
+  margin-bottom: 10px;
+  width: 80%;
+  font-size: 0.6em;
+  margin-left: 50px
+}
+.forceBreak { -webkit-column-break-after: always; break-after: column; }
+</style>
+
+<script src="https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script>
+<script>
+$(document).ready(function() {
+  $('slide:not(.backdrop):not(.title-slide)').append('<div class=\"footnotes\">');
+
+  $('footnote').each(function(index) {
+    var text  = $(this).html();
+    var fnNum = (index+1).toString();
+    $(this).html(fnNum.sup());
+
+    var footnote   = fnNum + '. ' + text + '<br/>';
+    var oldContent = $(this).parents('slide').children('div.footnotes').html();
+    var newContent = oldContent + footnote;
+    $(this).parents('slide').children('div.footnotes').html(newContent);
+  });
+});
+</script>
+
 ```{r, echo=FALSE}
@@
 
-</center>
-
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Computer Image: [Markley Boyer](https://www.facebook.com/markley.boyer), Photo Image: [Robert Clark](http://www.nationalgeographic.com/contributors/c/photographer-robert-clark/)
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [National Geographic](http://ngm.nationalgeographic.com/2009/09/manhattan/miller-text), [The Welikia Project](https://welikia.org/)
+Beyond Mannahatta<footnote>[Markley Boyer](https://www.facebook.com/markley.boyer), [Robert Clark](http://www.nationalgeographic.com/contributors/c/photographer-robert-clark/), [National Geographic](http://ngm.nationalgeographic.com/2009/09/manhattan/miller-text), [The Welikia Project](https://welikia.org/)</footnote>
+
+</center>
 
@@
 
-<br><br>
-
-</center>
-
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Image: [Janet Tench](https://commons.wikimedia.org/wiki/File:Norfolk_potato_field_-_geograph.org.uk_-_180521.jpg)
+Norfolk potato field<footnote>[Janet Tench](https://commons.wikimedia.org/wiki/File:Norfolk_potato_field_-_geograph.org.uk_-_180521.jpg)</footnote>
+
+</center>
 
@@
 
-</center>
-
-<br><br><br>
-
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Kehoe et al.
-(2017) *Nat.
-Ecol.
-Evol.* **1**:1129–1135](https://www.nature.com/articles/s41559-017-0234-3)
+Potential biodiversity loss due to increased agricultural development<footnote>[Kehoe et al. (2017) *Nat. Ecol. Evol.* **1**:1129–1135](https://www.nature.com/articles/s41559-017-0234-3)</footnote>
+
+</center>
 
@@
 
-# Uncertainty and conservation conflict
+# Modelling biodiversity dynamics and human decision-making
 
@@
 
-</center>
-
-<br>
-
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Bunnefeld et al.
+**Management Strategy Evaluation**
+
+</center>
+
+## {.smaller}
+
+<center>
+
+![](mse_overview_sm_un.png)
+
+**Management Strategy Evaluation**<footnote>[Bunnefeld et al.
 (2011) *Trends.
 Ecol.
-Evol.* **26**:441-447](https://doi.org/10.1016/j.tree.2011.05.003)
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Chadès et al. (2017) *Theor. Ecol.* **10**:1-10](https://doi.org/10.1007/s12080-016-0313-0)
-
-## {.smaller}
-
-<center>
-
-![](mse_overview_sm_un.png)
-
-</center>
-
-<br>
-
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Bunnefeld et al.
-(2011) *Trends.
+Evol.* **26**:441-447](https://doi.org/10.1016/j.tree.2011.05.003)</footnote><sup>,</sup><footnote>[Chadès et al.
+(2017) *Theor.
+Ecol.* **10**:1-10](https://doi.org/10.1007/s12080-016-0313-0)</footnote>
+
+</center>
+
+## The ConFooBio Project
+
+**Resolving conflicts between food security and biodiversity conservation under uncertainty**
+
+<div class = "columns-2">
+
+![](confoobio_sans.png)
+
+![](group.png)
+
+</div>
+
+## Generalised Management Strategy Evaluation (GMSE) software
+
+<div class = "columns-2">
+
+- **Model biodiversity dynamics & realistic human decision-making**
+- **Predict resource & land-use changes in social-ecological systems**
+- **Integrate flexibly with existing ecological models**<footnote>Duthie AB, et al.
+(2018) Method.
 Ecol.
-Evol.* **26**:441-447](https://doi.org/10.1016/j.tree.2011.05.003)
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Chadès et al. (2017) *Theor. Ecol.* **10**:1-10](https://doi.org/10.1007/s12080-016-0313-0)
+Evol.
+[**9**:2396-2401](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.13091) </footnote>
+- **Open-source on [CRAN](https://cran.r-project.org/package=GMSE) and [GitHub](https://github.com/ConFooBio/gmse)**<footnote>[http://confoobio.github.io/gmse](http://confoobio.github.io/gmse)</footnote>
+
+<p class="forceBreak"></p>
+
+<center>
+
+![](GMSE_Logo_Goose_300.png)
+
+[![](http://www.r-pkg.org/badges/version/GMSE?color=yellowgreen)](https://cran.r-project.org/package=GMSE) [![](http://cranlogs.r-pkg.org/badges/grand-total/GMSE?color=yellowgreen)](http://cranlogs.r-pkg.org/badges/grand-total/GMSE)
+
+</center>
+
+</div>
+
+## GMSE: population modelling (defaults)
+
+<div class = "columns-2"><br><br>
+
+- **Default models individual-based (IBM)**
+- **Spatially explicit landscape**
+- **Individual birth, movement, death**
+- **Can be over-ridden with custom models**
+
+<p class="forceBreak"></p>
+<br><br>
+<center>
+
+![](GMSE_Logo_Goose_300.png)
+
+[![](http://www.r-pkg.org/badges/version/GMSE?color=yellowgreen)](https://cran.r-project.org/package=GMSE) [![](http://cranlogs.r-pkg.org/badges/grand-total/GMSE?color=yellowgreen)](http://cranlogs.r-pkg.org/badges/grand-total/GMSE)
+
+</center>
+
+</div>
+
+## {.smaller}
+
+<center>
+
+![](case_studies.png)
+
+Modelling population dynamics<footnote>Cusack, JJ, et al.
+*Conserv.
+Lett.* (In press).
+DOI: <a ref = "https://onlinelibrary.wiley.com/doi/full/10.1111/conl.12450">10.1111/conl.12450</a>.</footnote><sup>,</sup><footnote>Cusack, JJ, et al.
+*In revision*</footnote>
+
+</center>
+
+##
+
+<center>
+
+![](mse_overview_sm.png)
+
+</center>
+
+##
+
+<center>
+
+![](gmse_overview_sm.png)
+
+</center>
 
@@
 
-## The ConFooBio Project
-
-**Resolving conflicts between food security and biodiversity conservation under uncertainty**
-
-<div class = "columns-2">
-
-![](confoobio_sans.png)
-
-![](group.png)
-
-</div>
+<!---
 
@@
 </div>
+
+--->
 
@@
 
+## {.smaller}
+
+<br>
+<center>
+
+![](funding.png)
+
+</center>
+
+<!---
+
 ##
@@
 </div>
+
+--->

### 2019-01-03 notebook/presentation/GMSE_presentation.Rmd
Add Magnus

@@
 
+## Game theory: conflict and cooperation
+
+<div class="columns-2">
+
+- Formalisation and study of conflict and cooperation between rational decision-makers
+- Compatibile with agent-based modelling<footnote>[Adami et al. (2016) *Phys. Life. Rev.* **19**:1-26](http://dx.doi.org/10.1016/j.plrev.2016.08.015)</footnote>
+- Can suggest management strategies<footnote>[Tilman et al.
+2016 *Theor.
+Ecol.*](http://link.springer.com/article/10.1007%2Fs12080-016-0318-8)</footnote> and be used to understand stake-holder actions<footnote>[Kark et al.
+2015 *Curr.
+Opin.
+Environ.
+Sustainability* **12**:12-24]()</footnote>
+- Conceptual unification
+
+<p class="forceBreak"></p>
+
+![](payoff_matrix_sm.png)
+<center>
+*Prisoner's dilemma*
+</center>
+
+</div>
+
+## {.smaller}
+
+<div class="columns-2">
+
+![Public license<footnote>test</footnote>](US_USSR_nukes_sm.png).
+
+![Wilkinson 1990](Wilkinson1990_sm.png).
+
+![Isaacdm](penalty_sm.png).
+
+![Dennig](brexit_sm.png).
+
+</div>
+
+<br>
+
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; **Images:** (1) [Public license](https://commons.wikimedia.org/wiki/File:US_and_USSR_nuclear_stockpiles.svg), (2) [Wilkinson 1990](https://projects.ncsu.edu/project/evoresources/Evolutionary%20medicine/Unitsofselection%20DM%20copy/wilkinson-foodsharing.pdf), (3) [Isaacdm](https://commons.wikimedia.org/wiki/File:PhysicsofPenalty.JPG?uselang=en-gb), (4) [Public license](http://www.publicdomainpictures.net/view-image.php?image=180560&picture=brexit-referendum-uk)
+
+## Game theory for wicked problems
+
+<div class="columns-2">
+
+<u>**Large strategy space:**</u>
+
+ - **Asymmetric payoffs** (payoffs affected by stake-holder identity)
+ - **Multiple choices**
+ - **Multiple players**
+ - **Iterated interactions** (strategies affected by interaction history)
+
+<p class="forceBreak"></p>
+
+![](payoff_matrix_sm.png)
+
+## Game theory for wicked problems
+
+<div class="columns-2">
+
+<u>**Large strategy space:**</u>
+
+ - **Asymmetric payoffs** (payoffs affected by stake-holder identity)
+ - **Multiple choices**
+ - **Multiple players**
+ - **Iterated interactions** (strategies affected by interaction history)
+
+<p class="forceBreak"></p>
+
+![](magnus.jpg)
+<center>
+*Magnus Carlsen*<footnote>[Andreas Kontokanis](https://commons.wikimedia.org/wiki/File:Carlsen_Magnus_(30238051906).jpg)</footnote>
+</center>
+
+##
+
+<center>
+
+![](gmse_overview_sm.png)
+
+</center>
+
 ## {.smaller}

### 2019-01-04 notebook/presentation/GMSE_presentation.Rmd
Really working example

--- 2019-01-03 notebook/presentation/GMSE_presentation.Rmd
@@
 }
-footer:after {
-  content: attr(label);
-  font-size: 12pt;
-  position: absolute;
-  bottom: 20px;
-  left: 60px;
-  line-height: 1.9;
-}
-div.footnotes {
-  position: absolute;
-  bottom: 0;
-  margin-bottom: 10px;
-  width: 80%;
-  font-size: 0.6em;
-  margin-left: 50px
-}
+
 .forceBreak { -webkit-column-break-after: always; break-after: column; }
@@
 
-<script src="https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script>
-<script>
-$(document).ready(function() {
-  $('slide:not(.backdrop):not(.title-slide)').append('<div class=\"footnotes\">');
-
-  $('footnote').each(function(index) {
-    var text  = $(this).html();
-    var fnNum = (index+1).toString();
-    $(this).html(fnNum.sup());
-
-    var footnote   = fnNum + '. ' + text + '<br/>';
-    var oldContent = $(this).parents('slide').children('div.footnotes').html();
-    var newContent = oldContent + footnote;
-    $(this).parents('slide').children('div.footnotes').html(newContent);
-  });
-});
-</script>
-
 ```{r, echo=FALSE}
-library(gamesGA);
 library(GMSE);
 library(knitr);
+library(shiny);
 ```
@@
 
-Beyond Mannahatta<footnote>[Markley Boyer](https://www.facebook.com/markley.boyer), [Robert Clark](http://www.nationalgeographic.com/contributors/c/photographer-robert-clark/), [National Geographic](http://ngm.nationalgeographic.com/2009/09/manhattan/miller-text), [The Welikia Project](https://welikia.org/)</footnote>
+[Markley Boyer](https://www.facebook.com/markley.boyer), [Robert Clark](http://www.nationalgeographic.com/contributors/c/photographer-robert-clark/), [National Geographic](http://ngm.nationalgeographic.com/2009/09/manhattan/miller-text), [Welikia Project](https://welikia.org/)
 
@@
 
-Norfolk potato field<footnote>[Janet Tench](https://commons.wikimedia.org/wiki/File:Norfolk_potato_field_-_geograph.org.uk_-_180521.jpg)</footnote>
+[Janet Tench](https://commons.wikimedia.org/wiki/File:Norfolk_potato_field_-_geograph.org.uk_-_180521.jpg)</footnote>
 
@@
 
-Potential biodiversity loss due to increased agricultural development<footnote>[Kehoe et al. (2017) *Nat. Ecol. Evol.* **1**:1129–1135](https://www.nature.com/articles/s41559-017-0234-3)</footnote>
+[Kehoe et al. (2017) *Nat. Ecol. Evol.* **1**:1129–1135](https://www.nature.com/articles/s41559-017-0234-3)</footnote>
 
@@
 
-**Management Strategy Evaluation**<footnote>[Bunnefeld et al.
-(2011) *Trends.
-Ecol.
-Evol.* **26**:441-447](https://doi.org/10.1016/j.tree.2011.05.003)</footnote><sup>,</sup><footnote>[Chadès et al.
-(2017) *Theor.
-Ecol.* **10**:1-10](https://doi.org/10.1007/s12080-016-0313-0)</footnote>
+</center>
+<br>
 
-</center>
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Bunnefeld et al. (2011) *Trends. Ecol. Evol.* [**26**:441-447](https://doi.org/10.1016/j.tree.2011.05.003) <br>
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Chadès et al. (2017) *Theor. Ecol.* [**10**:1-10](https://doi.org/10.1007/s12080-016-0313-0)
 
@@
 - **Predict resource & land-use changes in social-ecological systems**
-- **Integrate flexibly with existing ecological models**<footnote>Duthie AB, et al.
-(2018) Method.
-Ecol.
-Evol.
-[**9**:2396-2401](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.13091) </footnote>
-- **Open-source on [CRAN](https://cran.r-project.org/package=GMSE) and [GitHub](https://github.com/ConFooBio/gmse)**<footnote>[http://confoobio.github.io/gmse](http://confoobio.github.io/gmse)</footnote>
+- **Integrate flexibly with existing ecological models**
+- **Open-source on [CRAN](https://cran.r-project.org/package=GMSE) and [GitHub](https://github.com/ConFooBio/gmse)**
 
@@
 </div>
+<br><br>
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Duthie AB, et al. (2018) *Method. Ecol. Evol.* [**9**:2396-2401](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.13091)
 
@@
 
-Modelling population dynamics<footnote>Cusack, JJ, et al.
-*Conserv.
-Lett.* (In press).
-DOI: <a ref = "https://onlinelibrary.wiley.com/doi/full/10.1111/conl.12450">10.1111/conl.12450</a>.</footnote><sup>,</sup><footnote>Cusack, JJ, et al.
-*In revision*</footnote>
+<br>
+
+Cusack, JJ, et al. *Conserv. Lett.* (*In press*). DOI: [10.1111/conl.12450](https://onlinelibrary.wiley.com/doi/full/10.1111/conl.12450)
 
@@
 
-## {.smaller}
-
-<!--- Transition from global to local scale --->
-
-![](wildlife_conflict_sm.jpg)
-
-<center>
-**Image**: [Public domain](https://commons.wikimedia.org/wiki/File:Human-wildlife_conflict.jpg)
-</center>
-
-## {.smaller}
-
-<div class = "columns-2">
-
-<!--- ![](African_Bush_Elephant.jpg) --->
-
-![](grouse_shooting.jpg)
-
-**Conservation conflicts between biodiversity and food security result when the outcome of one stake-holder is affected by one or more other stake-holders** ([Colyvan et al.
-(2011) *Biol.
-Conserv.* **144**:1246-1253](http://www.sciencedirect.com/science/article/pii/S000632071000488X?via%3Dihub))
-
-<br>
-
-- People with strongly held views clash over objectives
-- Negative effect on sustainability, biodiversity, and human livelihood
-- MSE does not integrate the livelihood decisions of people
-
-</div>
-<br><br>
-<center>
-
-**Image:** [Public](https://commons.wikimedia.org/wiki/File:Hardy_Heywood_-_The_Grouse_Shoot_1898.jpg) [Domain](https://commons.wikimedia.org/wiki/File:Henharrier.jpg)
-<!--- **Image:** [Muhammad Mahdi Karim](https://en.wikipedia.org/wiki/User:Muhammad_Mahdi_Karim) --->
-
-</center>
-
-## {.smaller}
-
-<div class = "columns-2">
-
-<!--- ![](African_Bush_Elephant.jpg) --->
-
-![](grouse_shooting2.jpg)
-
-**Conservation conflicts between biodiversity and food security result when the outcome of one stake-holder is affected by one or more other stake-holders** ([Colyvan et al.
-(2011) *Biol.
-Conserv.* **144**:1246-1253](http://www.sciencedirect.com/science/article/pii/S000632071000488X?via%3Dihub))
-
-<br>
-
-- People with strongly held views clash over objectives
-- Negative effect on sustainability, biodiversity, and human livelihood
-- MSE does not integrate the livelihood decisions of people
-
-</div>
-
-<br><br>
-
-<center>
-
-**Image:** [Public](https://commons.wikimedia.org/wiki/File:Hardy_Heywood_-_The_Grouse_Shoot_1898.jpg) [Domain](https://commons.wikimedia.org/wiki/File:Henharrier.jpg)
-<!--- **Image:** [Muhammad Mahdi Karim](https://en.wikipedia.org/wiki/User:Muhammad_Mahdi_Karim) --->
-
-</center>
-
-<!---
-
-## The ConFooBio Project
-
-**Resolving conflicts between food security and biodiversity conservation under uncertainty**
-
-<div class = "columns-2">
-
-![](confoobio_sans.png)
-
-<br><br><br>
-
-- What leads to conservation conflict?
-- How to resolve conservation conflict?
-- Impact on sustainability of social-ecological systems?
-
-</div>
-
---->
-
-# Game theory in conservation conflict
-
-<!--- Get the ERC logo and EU flag in here? --->
-
 ## Game theory: conflict and cooperation
@@
 - Formalisation and study of conflict and cooperation between rational decision-makers
-- Broadly encompasses combinatorial games (e.g., tic-tac-toe, chess) and evolutionary invasion analysis (adaptive dynamics)
 - Compatibile with agent-based modelling
+- Can suggest management strategiesand be used to understand stake-holder actions
+- Conceptual unification
+
+<p class="forceBreak"></p>
 
 ![](payoff_matrix_sm.png)
+<center>
+*Prisoner's dilemma*
+</center>
 
-</div>
-<hr>
+</div><br>
 
-<br>
-
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Adami et al. (2016) *Phys. Life. Rev.* **19**:1-26](http://dx.doi.org/10.1016/j.plrev.2016.08.015)
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Adami et al. (2016) *Phys. Life. Rev.* [**19**:1-26](http://dx.doi.org/10.1016/j.plrev.2016.08.015) <br>
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Tilman et al. (2016) *Theor. Ecol.* [**10**:155-165](http://link.springer.com/article/10.1007%2Fs12080-016-0318-8)
 
@@
 
-![Public license](US_USSR_nukes_sm.png).
+![Public license<footnote>test</footnote>](US_USSR_nukes_sm.png).
 
@@
 
-## Conflict in conservation
-
-Game theory provides a natural framework for addressing issues of cooperation and conflict in conservation.
-
-- Game theory can be used to understand stake-holder decision making ([Kark et al. 2015 *Curr. Opin. Environ. Sustainability* **12**:12-24]()).
-
-- Game-theoretic models can suggest novel management strategies ([Colvan et al.
-2011 *Biol.
-Conserv.* **144**:1246-1253](http://dx.doi.org/10.1016/j.biocon.2010.10.028); [Tilman et al.
-2016 *Theor.
-Ecol.*](http://link.springer.com/article/10.1007%2Fs12080-016-0318-8)).
-
-- Game theory can consider multiple management objectives simultaneously ([Lee 2012 *Chemosphere* **87**:608-613](http://dx.doi.org/10.1016/j.chemosphere.2012.01.014)).
-
-## Stake-holders are complex
+## Game theory for wicked problems
 
@@
 
-**Simple models assume that stake-holders:**
+<u>**Large strategy space:**</u>
 
-- Know what pay-offs will be
-- Have clearly defined choices
-- Behave rationally
-- Understand the games
-- Prioritise outcomes clearly
+ - **Asymmetric payoffs** (payoffs affected by stake-holder identity)
+ - **Multiple choices**
+ - **Multiple players**
+ - **Iterated interactions** (strategies affected by interaction history)
+
+<p class="forceBreak"></p>
 
@@
 
-</div>
-
-## Conservation conflicts are complex {.smaller}
+## Game theory for wicked problems
 
@@
 
- - **Asymmetric payoffs**: Payoffs of choices may depend instead on stake-holder identity.
+<u>**Large strategy space:**</u>
 
-![](payoff_matrix_asym.png)
+ - **Asymmetric payoffs** (payoffs affected by stake-holder identity)
+ - **Multiple choices**
+ - **Multiple players**
+ - **Iterated interactions** (strategies affected by interaction history)
 
-</div>
+<p class="forceBreak"></p>
 
-## But most games are more complex {.smaller}
+![](magnus.jpg)
+<center>
+[Andreas Kontokanis](https://commons.wikimedia.org/wiki/File:Carlsen_Magnus_(30238051906).jpg)</footnote>
+</center>
 
-<div class="columns-2">
-
- - **Asymmetric payoffs**: Payoffs of choices may depend instead on stake-holder identity.
-  - **Multiple choices**: Players might have three or more choices.
-
-![](payoff_matrix_cho.png)
-
-</div>
-
-## But most games are more complex {.smaller}
-
-<div class="columns-2">
-
- - **Asymmetric payoffs**: Payoffs of choices may depend instead on stake-holder identity.
- - **Multiple choices**: Players might have three or more choices.
- - **Multiple players**: More than two stake-holders might interact at one time.
-
-![](payoff_matrix_3pl.png)
-
-</div>
-
-## But most games are more complex {.smaller}
-
-<div class="columns-2">
-
- - **Asymmetric payoffs**: Payoffs of choices may depend instead on stake-holder identity.
- - **Multiple choices**: Players might have three or more choices.
- - **Multiple players**: More than two stake-holders might interact at one time.
- - **Iterated interactions**: Strategies maybe affected by interaction history.
-
-```{r, echo=FALSE}
-rounds  <- as.character(1:9);
-players <- c("Player 1","Player 2");
-plays   <- sample(x=c(0,1), size=18, replace=TRUE);
-plays[plays == 0] <- "A";
-plays[plays == 1] <- "B";
-history <- matrix(data=plays, nrow=2);
-rownames(history) <- players;
-colnames(history) <- rounds;
-kable(noquote(history));
-```
-
-![](payoff_matrix_rd10_sm.png)
-
-</div>
-
-##{.smaller}
+##
 
@@
 
-![](mse_overview_sm.png)
+![](gmse_overview_sm.png)
 
 </center>
-
-## {.smaller}
-
-<center>
-
-![](gmse_overview_sm_ga.png)
-
-</center>
-
-<br><br><br>
-
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Bunnefeld et al.
-(2011) *Trends.
-Ecol.
-Evol.* **26**:441-447](https://doi.org/10.1016/j.tree.2011.05.003)
 
@@
 
-<!---
-
-# Genetic algorithms, and an example using Prisoner's dilemma
-
-## Structure of a genetic algorithm
-
-<div class="columns-2">
-
-```{r, echo=FALSE}
-mbox <- function(x0, x1, y0, y1){
-    xx <- seq(from=x0, to=x1, length.out = 100);
-    yy <- seq(from=y0, to=y1, length.out = 100);
-    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
-    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
-    return(list(x=xd, y=yd));
-}
-
-par(mar=c(0,0,0,0));
-# ===============================================================
-plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
-     xlab="",ylab="", bty="n");
-abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
-polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
-text(x=22, y=85, labels="Random", col="black");
-text(x=22, y=80, labels="seed of", col="black");
-text(x=22, y=75, labels="strategies", col="black");
-# --------------------------------------------------------------
-cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
-polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
-arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
-dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
-polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=84, labels="Begin new", col="black");
-text(x=105, y=79, labels="generation", col="black");
-# ===============================================================
-```
-
-<br>
-
-**Table of strategies**
-
-```{r, echo=FALSE}
-randvals <- sample(x=1:5, size=32, replace=TRUE);
-agents   <- matrix(data=randvals, nrow=8);
-rownames(agents) <- c("Strategy 1", "Strategy 2", "Strategy 3", "Strategy 4",
-                      "Strategy 5", "Strategy 6", "Strategy 7", "Strategy 8");
-colnames(agents) <- c("L1", "L2", "L3", "L4");
-kable(agents, format="html");
-```
-
-</div>
-
-## Structure of a genetic algorithm
-
-<div class="columns-2">
-
-```{r, echo=FALSE}
-mbox <- function(x0, x1, y0, y1){
-    xx <- seq(from=x0, to=x1, length.out = 100);
-    yy <- seq(from=y0, to=y1, length.out = 100);
-    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
-    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
-    return(list(x=xd, y=yd));
-}
-
-par(mar=c(0,0,0,0));
-# ===============================================================
-plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
-     xlab="",ylab="", bty="n");
-abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
-polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
-text(x=22, y=85, labels="Random", col="black");
-text(x=22, y=80, labels="seed of", col="black");
-text(x=22, y=75, labels="strategies", col="black");
-# --------------------------------------------------------------
-cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
-polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
-arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
-dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
-polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=84, labels="Begin new", col="black");
-text(x=105, y=79, labels="generation", col="black");
-# --------------------------------------------------------------
-ebox <- mbox(x0 = 75, x1 = 135, y0 = 63, y1 = 73);
-polygon(x=ebox$x, y=ebox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=68, labels="recombination", col="black");
-# ===============================================================
-```
-
-<br>
-
-**Table of strategies**
-
-```{r, echo=FALSE}
-for(i in 1:dim(agents)[1]){
-    for(j in 1:dim(agents)[2]){
-        check <- runif(n=1);
-        if(check < 0.2){
-            other            <- sample(x=1:dim(agents)[1], size=1);
-            temp             <- agents[i,j];
-            agents[i,j]      <- agents[other, j];
-            agents[other, j] <- temp;
-        }
-    }
-}
-kable(agents, format="html");
-```
-
-</div>
-
-## Structure of a genetic algorithm
-
-<div class="columns-2">
-
-```{r, echo=FALSE}
-mbox <- function(x0, x1, y0, y1){
-    xx <- seq(from=x0, to=x1, length.out = 100);
-    yy <- seq(from=y0, to=y1, length.out = 100);
-    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
-    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
-    return(list(x=xd, y=yd));
-}
-
-par(mar=c(0,0,0,0));
-# ===============================================================
-plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
-     xlab="",ylab="", bty="n");
-abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
-polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
-text(x=22, y=85, labels="Random", col="black");
-text(x=22, y=80, labels="seed of", col="black");
-text(x=22, y=75, labels="strategies", col="black");
-# --------------------------------------------------------------
-cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
-polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
-arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
-dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
-polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=84, labels="Begin new", col="black");
-text(x=105, y=79, labels="generation", col="black");
-# --------------------------------------------------------------
-ebox <- mbox(x0 = 75, x1 = 135, y0 = 63, y1 = 73);
-polygon(x=ebox$x, y=ebox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=68, labels="recombination", col="black");
-# --------------------------------------------------------------
-fbox <- mbox(x0 = 75, x1 = 135, y0 = 51, y1 = 61);
-polygon(x=fbox$x, y=fbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=56, labels="mutation", col="black");
-# ===============================================================
-```
-
-<br>
-
-**Table of strategies**
-
-```{r, echo=FALSE}
-for(i in 1:dim(agents)[1]){
-    for(j in 1:dim(agents)[2]){
-        check <- runif(n=1);
-        if(check < 0.2){
-            agents[i,j]  <- sample(x=1:5, size=1);
-        }
-    }
-}
-kable(agents, format="html");
-```
-
-</div>
-
-## Structure of a genetic algorithm
-
-<div class="columns-2">
-
-```{r, echo=FALSE}
-mbox <- function(x0, x1, y0, y1){
-    xx <- seq(from=x0, to=x1, length.out = 100);
-    yy <- seq(from=y0, to=y1, length.out = 100);
-    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
-    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
-    return(list(x=xd, y=yd));
-}
-
-par(mar=c(0,0,0,0));
-# ===============================================================
-plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
-     xlab="",ylab="", bty="n");
-abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
-polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
-text(x=22, y=85, labels="Random", col="black");
-text(x=22, y=80, labels="seed of", col="black");
-text(x=22, y=75, labels="strategies", col="black");
-# --------------------------------------------------------------
-cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
-polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
-arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
-dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
-polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=84, labels="Begin new", col="black");
-text(x=105, y=79, labels="generation", col="black");
-# --------------------------------------------------------------
-ebox <- mbox(x0 = 75, x1 = 135, y0 = 63, y1 = 73);
-polygon(x=ebox$x, y=ebox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=68, labels="recombination", col="black");
-# --------------------------------------------------------------
-fbox <- mbox(x0 = 75, x1 = 135, y0 = 51, y1 = 61);
-polygon(x=fbox$x, y=fbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=56, labels="mutation", col="black");
-# --------------------------------------------------------------
-gbox <- mbox(x0 = 75, x1 = 135, y0 = 39, y1 = 49);
-polygon(x=gbox$x, y=gbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=44, labels="check fitness", col="black");
-# ===============================================================
-```
-
-<br>
-
-**Table of strategies**
-
-```{r, echo=FALSE}
-fit <- apply(agents,1,sum);
-kable(cbind(agents,fit), format="html");
-```
-
-</div>
-
-## Structure of a genetic algorithm
-
-<div class="columns-2">
-
-```{r, echo=FALSE}
-mbox <- function(x0, x1, y0, y1){
-    xx <- seq(from=x0, to=x1, length.out = 100);
-    yy <- seq(from=y0, to=y1, length.out = 100);
-    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
-    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
-    return(list(x=xd, y=yd));
-}
-
-par(mar=c(0,0,0,0));
-# ===============================================================
-plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
-     xlab="",ylab="", bty="n");
-abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
-polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
-text(x=22, y=85, labels="Random", col="black");
-text(x=22, y=80, labels="seed of", col="black");
-text(x=22, y=75, labels="strategies", col="black");
-# --------------------------------------------------------------
-cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
-polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
-arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
-dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
-polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=84, labels="Begin new", col="black");
-text(x=105, y=79, labels="generation", col="black");
-# --------------------------------------------------------------
-ebox <- mbox(x0 = 75, x1 = 135, y0 = 63, y1 = 73);
-polygon(x=ebox$x, y=ebox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=68, labels="recombination", col="black");
-# --------------------------------------------------------------
-fbox <- mbox(x0 = 75, x1 = 135, y0 = 51, y1 = 61);
-polygon(x=fbox$x, y=fbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=56, labels="mutation", col="black");
-# --------------------------------------------------------------
-gbox <- mbox(x0 = 75, x1 = 135, y0 = 39, y1 = 49);
-polygon(x=gbox$x, y=gbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=44, labels="check fitness", col="black");
-# --------------------------------------------------------------
-hbox <- mbox(x0 = 75, x1 = 135, y0 = 27, y1 = 37);
-polygon(x=hbox$x, y=hbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=32, labels="selection", col="black");
-# ===============================================================
-```
-
-<br>
-
-**Table of strategies**
-
-```{r, echo=FALSE}
-ord <- order(fit, decreasing=TRUE);
-agents <- agents[ord,];
-fit    <- fit[ord];
-kable(cbind(agents,fit), format="html");
-```
-
-</div>
-
-## Structure of a genetic algorithm
-
-<div class="columns-2">
-
-```{r, echo=FALSE}
-mbox <- function(x0, x1, y0, y1){
-    xx <- seq(from=x0, to=x1, length.out = 100);
-    yy <- seq(from=y0, to=y1, length.out = 100);
-    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
-    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
-    return(list(x=xd, y=yd));
-}
-
-par(mar=c(0,0,0,0));
-# ===============================================================
-plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
-     xlab="",ylab="", bty="n");
-abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
-polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
-text(x=22, y=85, labels="Random", col="black");
-text(x=22, y=80, labels="seed of", col="black");
-text(x=22, y=75, labels="strategies", col="black");
-# --------------------------------------------------------------
-cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
-polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
-arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
-dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
-polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=84, labels="Begin new", col="black");
-text(x=105, y=79, labels="generation", col="black");
-# --------------------------------------------------------------
-ebox <- mbox(x0 = 75, x1 = 135, y0 = 63, y1 = 73);
-polygon(x=ebox$x, y=ebox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=68, labels="recombination", col="black");
-# --------------------------------------------------------------
-fbox <- mbox(x0 = 75, x1 = 135, y0 = 51, y1 = 61);
-polygon(x=fbox$x, y=fbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=56, labels="mutation", col="black");
-# --------------------------------------------------------------
-gbox <- mbox(x0 = 75, x1 = 135, y0 = 39, y1 = 49);
-polygon(x=gbox$x, y=gbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=44, labels="check fitness", col="black");
-# --------------------------------------------------------------
-hbox <- mbox(x0 = 75, x1 = 135, y0 = 27, y1 = 37);
-polygon(x=hbox$x, y=hbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=32, labels="selection", col="black");
-# --------------------------------------------------------------
-ibox <- mbox(x0 = 75, x1 = 135, y0 = 15, y1 = 25);
-polygon(x=ibox$x, y=ibox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=20, labels="replacement", col="black");
-# ===============================================================
-```
-
-<br>
-
-**Table of strategies**
-
-```{r, echo=FALSE}
-agents <- rbind(agents[1,],
-                agents[1,],
-                agents[2,],
-                agents[2,],
-                agents[3,],
-                agents[3,],
-                agents[4,],
-                agents[4,]
-                );
-fit    <- c(fit[1], fit[1], fit[2], fit[2], fit[3], fit[3], fit[4], fit[4]);
-kable(cbind(agents,fit), format="html");
-```
-
-</div>
-
-## Structure of a genetic algorithm
-
-<div class="columns-2">
-
-```{r, echo=FALSE}
-mbox <- function(x0, x1, y0, y1){
-    xx <- seq(from=x0, to=x1, length.out = 100);
-    yy <- seq(from=y0, to=y1, length.out = 100);
-    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
-    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
-    return(list(x=xd, y=yd));
-}
-
-par(mar=c(0,0,0,0));
-# ===============================================================
-plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
-     xlab="",ylab="", bty="n");
-abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
-polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
-text(x=22, y=85, labels="Random", col="black");
-text(x=22, y=80, labels="seed of", col="black");
-text(x=22, y=75, labels="strategies", col="black");
-# --------------------------------------------------------------
-cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
-polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
-arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
-dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
-polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=84, labels="Begin new", col="black");
-text(x=105, y=79, labels="generation", col="black");
-# --------------------------------------------------------------
-ebox <- mbox(x0 = 75, x1 = 135, y0 = 63, y1 = 73);
-polygon(x=ebox$x, y=ebox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=68, labels="recombination", col="black");
-# --------------------------------------------------------------
-fbox <- mbox(x0 = 75, x1 = 135, y0 = 51, y1 = 61);
-polygon(x=fbox$x, y=fbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=56, labels="mutation", col="black");
-# --------------------------------------------------------------
-gbox <- mbox(x0 = 75, x1 = 135, y0 = 39, y1 = 49);
-polygon(x=gbox$x, y=gbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=44, labels="check fitness", col="black");
-# --------------------------------------------------------------
-hbox <- mbox(x0 = 75, x1 = 135, y0 = 27, y1 = 37);
-polygon(x=hbox$x, y=hbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=32, labels="selection", col="black");
-# --------------------------------------------------------------
-ibox <- mbox(x0 = 75, x1 = 135, y0 = 15, y1 = 25);
-polygon(x=ibox$x, y=ibox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=20, labels="replacement", col="black");
-# --------------------------------------------------------------
-jbox <- mbox(x0 = 75, x1 = 135, y0 = 3, y1 = 13);
-polygon(x=jbox$x, y=jbox$y, lwd=3, border="goldenrod4", col="white");
-text(x=105, y=8, labels="converge?", col="black");
-# --------------------------------------------------------------
-bbox <- mbox(x0 = 0, x1 = 45, y0 = 20, y1 = 40);
-polygon(x=bbox$x, y=bbox$y, lwd=3, border="blue", col="lightsteelblue1");
-arrows(x0=70, x1=22.5, y0=5, y1=5, lwd=2, length=0.0);
-arrows(x0=22, x1=22, y0=5, y1=20, lwd=2, length=0.15);
-arrows(x0=55, x1=55, y0=5, y1=82, lwd=2, length=0.15);
-text(x=15, y=10, labels="Yes", col="black", srt=90);
-text(x=50, y=10, labels="No", col="black", srt=90);
-text(x=22, y=33, labels="Adaptive", col="black");
-text(x=22, y=28, labels="strategies", col="black");
-# ===============================================================
-```
-
-<br>
-
-**Table of strategies**
-
-```{r, echo=FALSE}
-kable(cbind(agents,fit), format="html");
-```
-
-</div>
-
---->
-
-<!---
-
-## Iterated Prisoner's dilemma rules {.smaller}
-
-<div class="columns-2">
-
-![](payoff_matrix_sm.png)
-**General structure of Prisoner's dilemma scenario**
-
-![](pd_pays.png)
-**Specific payoffs satisfying a Prisoner's dilemma scenario**
-
-</div>
-
-## Use of genetic algorithms {.smaller}
-
-**Each locus is an focal player's choice given an opponent's past three choices (plus one choice of first move)**
-
-<hr>
-
-|                  | **CCC** | **CCD** | **CDC** | **CDD** | **DCC** | **DCD** | **DDC** | **DDD** | **1st** |
-|------------------|---------|---------|---------|---------|---------|---------|---------|---------|---------|
-| **Strategy 1**   | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  |
-| **Strategy 2**   | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  |
-| **...**          |   ...   |   ...   |   ...   |   ...   |   ...   |   ...   |   ...   |   ...   |   ...   |
-| **Strategy N-1** | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  |
-| **Strategy N**   | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  |
-
-<hr>
-
-**Total of 512 possible strategies**
-
-## Use of genetic algorithms {.smaller}
-
-```{r, echo=FALSE}
-inputPanel(
-
-  numericInput("CC", label = "CC", 0, width="30%"),
-  numericInput("DC", label = "DC", 0, width="30%"),
-  numericInput("CD", label = "CD", 0, width="30%"),
-  numericInput("DD", label = "DD", 0, width="30%"),
-
-  style='width: 2000px; height: 60px'
-)
-```
-
-<br>
-
-```{r, echo=FALSE}
-gres <- reactive({ 
-     cc <- input$CC;
-     dc <- input$DC;
-     cd <- input$CD;
-     dd <- input$DD;
-     rd <- 100;
-     gn <- 100;
-     co <- 0.05;
-     mu <- 0.05;
-
-     game_res <- games_ga(CC = cc, DC = dc, CD = cd, DD = dd, rounds = rd,
-                          generations = gn, cross_prob = co, 
-                          mutation_prob = mu, num_opponents = 100);     
-     all <- c(game_res$genos, "break_1",
-              rownames(game_res$genos), "break_2",
-              game_res$fitness);
-    })
-```
-
-```{r, echo = FALSE}
-renderPlot({
-    break2    <- which(gres()=="break_2")[1]+1;
-    end       <- length(gres());
-    fitnesses <- gres()[break2:end];
-    fitnesses <- as.numeric(fitnesses);
-    mean_fit  <- fitnesses / (100 * 100);
-    maxpt     <- max(c(input$CC, input$CD, input$DC, input$DD));
-    par(mar=c(5,5,1,1));
-    plot(x = 1:length(mean_fit), y=mean_fit, type="l", ylim=c(0,maxpt), lwd = 3,
-        xlab="Generation", ylab = "Mean strategy fitness per round",
-         cex.axis=1.5, cex.lab=1.5);
-})
-```
-
-## Adaptive strategies simulated {.smaller}
-
-```{r, echo=FALSE}
-renderTable({
-      break1   <- which(gres()=="break_1")[1] - 1;
-      break2   <- which(gres()=="break_2")[1] - 1;
-      geno_vec <- gres()[1:break1];
-      geno_row <- gres()[(break1+2):break2];
-      genos    <- matrix(data =  geno_vec, ncol = 10, byrow = FALSE);
-      rownames(genos) <- geno_row;
-      colnames(genos) <- c("CCC", "CCD", "CDC", "CDD", "DCC", "DCD", "DDC",
-                           "DDD", "1st", "Final %");
-      final_table <- genos[1:10,];
-
-}, include.rownames=TRUE)
-```
-
---->
-
-# Genetic algorithms in Management Strategy Evaluation
-
-## Generalised MSE {.smaller}
+## GMSE: modelling decision-making
 
@@
 
-- Integrates game theory and ecological theory to construct social-ecological models for adaptive management
-- Simulates all aspects of management: population dynamics, manager observation of populations, manager decision making, and stake-holder responses to management decisions
-- Mimic the decision-making process of managers and stake-holders under conditions of change, uncertainty, and conflict
-- Provide a flexible modelling framework for future development of complex adaptive management models to be applied to conservation theory and specific case studies
+**Up to 5 possible actions**
 
+- Culling (&darr; survival)
+- Castration (&darr; reproduction)
+- Scaring (displacement)
+- Feeding (&uarr; survival)
+- Helping (&uarr; reproduction)
+
+**Manager sets policy**
+
+- Has a target population size
+- Has a budget to set costs
+
+<p class="forceBreak"></p>
 <br><br>
-
 <center>
 
-![](GMSE_Logo_Goose_375.png)
+![](GMSE_Logo_Goose_300.png)
+
+[![](http://www.r-pkg.org/badges/version/GMSE?color=yellowgreen)](https://cran.r-project.org/package=GMSE) [![](http://cranlogs.r-pkg.org/badges/grand-total/GMSE?color=yellowgreen)](http://cranlogs.r-pkg.org/badges/grand-total/GMSE)
 
@@
 
-## Generalised MSE ('GMSE' R package)
+## GMSE: modelling decision-making
 
@@
 
-- Individual-based
-- Spatially explicit
-- Written in R and C
-- Requires R (>= 3.3.3)
-- License GPL (>=2)
-- *v0.2.2.7* [available](https://cran.r-project.org/package=GMSE) on CRAN
-- Browser-based GUI [available](https://bradduthie.shinyapps.io/gmse_gui/) via shiny
-- Source code [available](https://github.com/bradduthie/gmse) on GitHub
-- Intended for a wide range of software users
+**Up to 5 possible actions**
 
-<br><br>
+- Culling (&darr; survival)
+- Castration (&darr; reproduction)
+- Scaring (displacement)
+- Feeding (&uarr; survival)
+- Helping (&uarr; reproduction)
+
+**Users respond to policy**
+
+- Maximise land yield
+- Maximise resource use
+
+<p class="forceBreak"></p>
 
@@
 
-![](GMSE_Logo_Goose_375.png)
+```{r, echo = FALSE}
+sim_old  <- gmse_apply(get_res = "Full", remove_pr = 0.122, lambda = 0.275,
+                       res_d4eath_K = 93870, RESOURCE_ini = 35000, 
+                       manage_target = 70000, res_death_type = 3,
+                       manager_budget = 10000, user_budget = 10000,
+                       public_land = 0.4, stakeholders = 8, res_consume = 0.02, 
+                       res_birth_K = 200000, land_ownership = TRUE,
+                       observe_type = 3, agent_view = 1, converge_crit = 0.01,
+                       ga_mingen = 200);
+sim_old$LAND[1:20,   1:20,  3] <- 2; 
+sim_old$LAND[1:20,  21:40,  3] <- 3;
+sim_old$LAND[1:20,  41:60,  3] <- 4;
+sim_old$LAND[1:20,  61:80,  3] <- 5;
+sim_old$LAND[1:20,  81:100, 3] <- 6;
+sim_old$LAND[21:40,  1:50,  3] <- 7;
+sim_old$LAND[21:40, 51:100, 3] <- 8;
+sim_old$LAND[41:60,  1:100, 3] <- 9;
+sim_old$LAND[61:100, 1:100, 3] <- 1; # Public land
+image(x = sim_old$LAND[,,3], col = topo.colors(9), xaxt = "n", yaxt = "n");
+```
 
@@
 
-##
+## GMSE: modelling decision-making {.smaller}
 
-<center>
+Default individual-based models can be run using `gmse` in R ([multiple options](https://confoobio.github.io/gmse/reference/gmse.html)).
 
-![](gmse_overview_sm.png)
-
-</center>
-
-## GMSE: Natural resources
-
-<div class = "columns-2">
-
-- Spatially explicit landscape
-- Discrete individual resources
-- Individual birth, movement, death
-
-<br>
-
-<center>
-
-![](sep_land.gif)
-
-</center>
-
-## GMSE: Resource observation {.smaller}
-
-<div class = "columns-2">
-
-- Four types of observation options
-    1. Counting on a subset of landscape
-    2. Mark and recapture
-    3. Linear transect sampling
-    4. Block transect sampling
-- Mechanistic sampling process
-- Generates simulated data frame used by the manager
-
-<br>
-
-<center>
-
-![](binos.png)
-
-</center>
-
-</div>
-
-<center>
-
-<br><br><br><br><br><br><br><br>
-
-**Image:** [Public Domain](https://commons.wikimedia.org/wiki/File:Fernglas(alt).JPG)
-
-</center>
-
-## GMSE: Management {.smaller}
-
-<div class = "columns-2">
-
-- One manager analyses observations and compares with a target abundance
-- Five possible policy options (scaring, culling, castration, feeding, helping)
-- Fixed budget and costs for setting policy
-- Genetic algorithm anticipates policy effects from previous actions to set costs of actions for stake-holders
-
-<br>
-
-<center>
-
-![](balance.png)
-
-</center>
-
-## GMSE: Stake-holder actions {.smaller}
-
-<div class = "columns-2">
-
-- Any number of stake-holders
-- Stake-holders want to maximise harvest of resources or minimise resource effects on their land
-- Seven possible actions (scaring, culling, castration, feeding, helping, tend crops, kill crops)
-- Fixed budget for performing actions of potentially differing costs
-- Genetic algorithm anticipates effects of actions on harvest or land yield
-
-<br>
-
-<center>
-
-![](harvesting.jpg)
-
-</center>
-
-</div>
-
-<center>
-
-<br>
-
-**Image:** [Public Domain](https://commons.wikimedia.org/wiki/File:Libiaz-zniwa1938.a.jpg)
-
-</center>
-
-# Example use of GMSE
-
-## Running the gmse() function {.smaller}
-
-```{r, echo = TRUE}
-sim <- gmse(land_ownership = TRUE, stakeholders = 10, RESOURCE_ini = 1000,
-            res_death_K = 2000, manage_target = 1000, res_consume = 1,
-            manager_budget = 5000, user_budget = 5000, plotting = FALSE,
-            scaring = TRUE, culling = TRUE, observe_type = 1, fixed_mark = 50, 
-            fixed_recapt = 300);
+```{r, eval = FALSE}
+sim <- gmse(time_max = 30, land_dim_1 = 100, land_dim_2 = 100,
+            observe_type = 1, res_death_K = 4000, 
+            manager_budget = 2000, user_budget = 500, 
+            scaring = TRUE, land_ownership = TRUE,
+            tend_crops = TRUE, stakeholders = 7,
+            res_consume = 1, RESOURCE_ini = 600);
 ```
 
-## Inferring GMSE from plots {.smaller}
-
-**General plot of simulation results.**
-
-```{r, echo = TRUE, eval = FALSE}
-plot_gmse_results(sim_results = sim);
+```{r, eval = FALSE}
+plot_gmse_results(sim);
 ```
 
-<br>
-
-**Plot of conflict among stake-holders**
-
-```{r, echo = TRUE, eval = FALSE}
-plot_gmse_effort(sim_results = sim);
+```{r, eval = FALSE}
+plot_gmse_effort(sim);
 ```
-
-##
-
-<center>
-
-```{r, echo = FALSE, fig.height = 6, fig.width = 7}
-plot_gmse_results(sim_results = sim);
-```
-
-</center>
 
@@
 
-<center>
+```{r, echo = FALSE}
+plot_gmse_effort <- function(sim){
 
-```{r, echo = FALSE, fig.height = 6.5, fig.width = 6}
-plot_gmse_effort(sim_results = sim);
+    agents <- sim$agents;
+    paras  <- sim$paras;
+    ACTION <- sim$action;
+    COST   <- sim$cost;
+
+    cols      <- c("green", "indianred1", "indianred3", "deepskyblue1",
+                   "deepskyblue2");
+
+    users    <- dim(agents[[1]])[1];
+    max_time <- dim(paras)[1];
+
+    para_vec <- paras[1,];
+
+    allowed  <- sum(para_vec[89:93]);
+    scar_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
+    cull_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
+    cast_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
+    feed_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
+    help_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
+
+    scar_act  <- matrix(data = 0, nrow = max_time, ncol = users);
+    cull_act  <- matrix(data = 0, nrow = max_time, ncol = users);
+    cast_act  <- matrix(data = 0, nrow = max_time, ncol = users);
+    feed_act  <- matrix(data = 0, nrow = max_time, ncol = users);
+    help_act  <- matrix(data = 0, nrow = max_time, ncol = users);
+    scar_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
+    cull_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
+    cast_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
+    feed_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
+    help_eff  <- matrix(data = 0, nrow = max_time, ncol = users);    
+    act_costs  <- matrix(data = 0, nrow = max_time, ncol = 5);
+    pol_effort <- matrix(data = 0, nrow = max_time, ncol = 5);
+    min_cost   <- para_vec[97];
+
+    for(time in 1:max_time){
+        #-- Scaring cost and actions
+        scar_cst[time, 1] <- COST[[time]][3, 8, 1];
+        for(user in 2:users){
+            scar_cst[time, user] <- COST[[time]][1, 8, user];
+        }
+        scar_act[time, 1] <- ACTION[[time]][3, 8, 1] - min_cost;
+        for(user in 2:users){
+            scar_act[time, user] <- ACTION[[time]][1, 8, user];
+        }
+        scar_eff[time,] <- scar_act[time,]*scar_cst[time,]/agents[[time]][,17];
+        #-- Culling cost and actions
+        cull_cst[time, 1] <- COST[[time]][3, 9, 1];
+        for(user in 2:users){
+            cull_cst[time, user] <- COST[[time]][1, 9, user];
+        }
+        cull_act[time, 1] <- ACTION[[time]][3, 9, 1] - min_cost;
+        for(user in 2:users){
+            cull_act[time, user] <- ACTION[[time]][1, 9, user];
+        }
+        cull_eff[time,] <- cull_act[time,]*cull_cst[time,]/agents[[time]][,17];
+        #-- Castration cost and actions
+        cast_cst[time, 1] <- COST[[time]][3, 10, 1];
+        for(user in 2:users){
+            cast_cst[time, user] <- COST[[time]][1, 10, user];
+        }
+        cast_act[time, 1] <- ACTION[[time]][3, 10, 1] - min_cost;
+        for(user in 2:users){
+            cast_act[time, user] <- ACTION[[time]][1, 10, user];
+        }
+        cast_eff[time,] <- cast_act[time,]*cast_cst[time,]/agents[[time]][,17];
+        #-- Feeding cost and actions
+        feed_cst[time, 1] <- COST[[time]][3, 11, 1];
+        for(user in 2:users){
+            feed_cst[time, user] <- COST[[time]][1, 11, user];
+        }
+        feed_act[time, 1] <- ACTION[[time]][3, 11, 1] - min_cost;
+        for(user in 2:users){
+            feed_act[time, user] <- ACTION[[time]][1, 11, user];
+        }
+        feed_eff[time,] <- feed_act[time,]*feed_cst[time,]/agents[[time]][,17];
+        #-- Helping cost and actions
+        help_cst[time, 1] <- COST[[time]][3, 12, 1];
+        for(user in 2:users){
+            help_cst[time, user] <- COST[[time]][1, 12, user];
+        }
+        help_act[time, 1] <- ACTION[[time]][3, 12, 1] - min_cost;
+        for(user in 2:users){
+            help_act[time, user] <- ACTION[[time]][1, 12, user];
+        }
+        help_eff[time,] <- help_act[time,]*help_cst[time,]/agents[[time]][,17];
+    }
+    # -- Turn these into percentages:
+    scar_eff <- scar_eff * 100;
+    cull_eff <- cull_eff * 100;
+    cast_eff <- cast_eff * 100;
+    feed_eff <- feed_eff * 100;
+    help_eff <- help_eff * 100;
+
+    par(mfrow = c(5, 1), mar = c(0, 0, 0, 0), oma = c(6, 6, 0, 6));
+    #---- Scaring
+    if(para_vec[89] == 1){
+        y1 <- 100;
+        y2 <- y1 * 1.2
+        plot(x = 1:max_time, y = 100 - scar_eff[,1], type = "l", lwd = 2, 
+             cex.axis = 1.5, xaxt = "n", ylim = c(0, y2), yaxt = "n");
+        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+        par(new = TRUE);
+        plot(x = 1:max_time, y = scar_eff[,2], type = "n", lwd = 2, 
+             ylim = c(0, y2), xaxt="n", yaxt="n", 
+             cex.axis = 1.5);
+        for(stakeholder in 2:users){
+            points(x = 1:max_time, y = scar_eff[,stakeholder], type = "l", 
+                   lwd = 1, col = cols[1]);
+        }
+        axis(side = 4, at = c(50, 100),  labels = c(50, 100), cex.axis = 1.5);
+        legend(x = 1, y = y2 + y2*0.1, fill = cols[1:5], horiz = TRUE,
+               legend = c("scaring", "culling", "castration", "feeding", 
+                          "helping"), bty = "n", cex = 1);
+        abline(h = y1 + y1*0.1, lwd = 2);
+    }else{
+        plot(x = 1:max_time, y = scar_eff[,1], type = "n", xaxt= "n", 
+             yaxt = "n", ylim = c(0, 110));
+        text(x = 0, y = 90, cex = 2, labels = "No scaring allowed",
+             pos = 4);
+        abline(h = 100, lwd = 2);
+        legend(x = 1, y = 119, fill = cols[1:5], horiz = TRUE,
+               legend = c("scaring", "culling", "castration", "feeding", 
+                          "helping"), bty = "n", cex = 1);
+    }
+    #---- Culling
+    if(para_vec[90] == 1){
+        plot(x = 1:max_time, y = 100 - cull_eff[,1], type = "l", lwd = 2, 
+             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
+        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+        par(new = TRUE);
+        plot(x = 1:max_time, y = cull_eff[,2], type = "n", lwd = 2, 
+             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
+        for(stakeholder in 2:users){
+            points(x = 1:max_time, y = cull_eff[,stakeholder], type = "l", 
+                   lwd = 1, col = cols[2]);
+        }
+        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+    }else{
+        plot(x = 1:max_time, y = cull_act[,1], type = "n", xaxt= "n", 
+             yaxt = "n", ylim = c(0, 100));
+        text(x = 0, y = 90, cex = 2, labels = "No culling allowed",
+             pos = 4);
+    }
+    #---- Castrating
+    if(para_vec[91] == 1){
+        plot(x = 1:max_time, y = 100 - cast_eff[,1], type = "l", lwd = 2, 
+             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
+        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+        par(new = TRUE);
+        plot(x = 1:max_time, y = cast_eff[,2], type = "n", lwd = 2, 
+             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
+        for(stakeholder in 2:users){
+            points(x = 1:max_time, y = cast_eff[,stakeholder], type = "l", 
+                   lwd = 1, col = cols[3]);
+        }
+        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+    }else{
+        plot(x = 1:max_time, y = cast_act[,1], type = "n", xaxt= "n", 
+             yaxt = "n", ylim = c(0, 100));
+        text(x = 0, y = 90, cex = 2, labels = "No castration allowed",
+             pos = 4);
+    }
+    #---- Axes labels
+    mtext("Manager's permissiveness of user action", 
+          side = 2, line = 3.5, cex = 1.5, col = "black");
+    mtext("Percentage of user action effort expended", 
+          side = 4, line = 3.5, cex = 1.5, col = "black");
+    #---- Feeding
+    if(para_vec[92] == 1){
+        plot(x = 1:max_time, y = 100 - feed_eff[,1], type = "l", lwd = 2, 
+             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
+        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+        par(new = TRUE);
+        plot(x = 1:max_time, y = feed_eff[,2], type = "n", lwd = 2, 
+             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
+        for(stakeholder in 2:users){
+            points(x = 1:max_time, y = feed_eff[,stakeholder], type = "l", 
+                   lwd = 1, col = cols[4]);
+        }
+        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+    }else{
+        plot(x = 1:max_time, y = feed_act[,1], type = "n", xaxt= "n", 
+             yaxt = "n", ylim = c(0, 100));
+        text(x = 0, y = 90, cex = 2, labels = "No feeding allowed",
+             pos = 4);
+    }
+    #---- Helping
+    if(para_vec[93] == 1){
+        plot(x = 1:max_time, y = 100 - help_eff[,1], type = "l", lwd = 2, 
+             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
+        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+        par(new = TRUE);
+        plot(x = 1:max_time, y = help_eff[,2], type = "n", lwd = 2, 
+             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
+        for(stakeholder in 2:users){
+            points(x = 1:max_time, y = help_eff[,stakeholder], type = "l", 
+                   lwd = 1, col = cols[5]);
+        }
+        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
+    }else{
+        plot(x = 1:max_time, y = help_act[,1], type = "n", cex.axis = 1.5,
+             yaxt = "n", ylim = c(0, 100));
+        text(x = 0, y = 90, cex = 2, labels = "No helping offspring allowed",
+             pos = 4);
+    }    
+    mtext("Time step", side = 1, line = 3.5, cex = 1.5, col = "black");
+}
 ```
 
-</center>
+```{r, echo = FALSE}
+inputPanel(
+  actionButton("run", "Run Simulation"),
+    actionButton("res", "Reset"),
+  radioButtons("typ", "",
+               c("General" = 1,
+                 "Effort"  = 2),
+               inline = TRUE, width = '800px')
+)
 
-# [GMSE GUI](https://bradduthie.shinyapps.io/gmse_gui/)
+sim      <- reactiveValues();
+okay     <- reactiveValues();
+sim$data <- NULL;
+okay$go  <- 0;
 
-## Use of GMSE: Model case studies
+sim_run <- observeEvent(input$run, {
+              sim$data <- gmse(time_max = 30, land_dim_1 = 100, land_dim_2 = 100, 
+                            observe_type = 1, res_death_K = 4000, plotting = FALSE,
+                            manager_budget = 2000, user_budget = 500, 
+                            scaring = TRUE, land_ownership = TRUE,
+                            tend_crops = TRUE, stakeholders = 7,
+                            res_consume = 1, RESOURCE_ini = 600)
+              showNotification("Simulation has finished")
+})
 
-<div class = "columns-2">
+observeEvent(input$res, {
+    sim$data <- NULL;
+})
 
-<center>
-
-![](African_Bush_Elephant.jpg)  
-**Image:** [M. M. Karim](https://en.wikipedia.org/wiki/User:Muhammad_Mahdi_Karim)
-
-![](barnacle_goose.JPG)  
-**Image:** [Raju Kasambe](https://commons.wikimedia.org/wiki/File:Barnacle_Geese_Branta_leucopsis_London_UK_(4).JPG)
-
-</center>
-
-</div>
-
-## Use of GMSE: Theoretical questions
-
-<center>
-
-![](man_freq.png)
-
-</center>
-
-## Future software development
-
-<div class = "columns-2">
-
-- Modular `gmse_apply()` function allowing user-supplied natural resource, observation, manager, and user models
-- Stuctured populations (e.g., sex, stage)
-- Multiple populations of interacting species
-- Improved strategy prediction for managers and users
-- Interactive decision-making with software users
-
-<center>
-
-![](GMSE_Logo_Goose_375.png)
-
-</center>
-
-</div>
+renderPlot(height = 500, {
+    if(is.null(sim$data) == FALSE){
+        if(input$typ == 1){
+            plot_gmse_results(sim$data)
+        }else{
+            plot_gmse_effort(sim$data)
+        }
+    }
+})
+```
 
@@
 </center>
-
-<!---
-
-##
-
-<br><br><br>
-
-<div class = "columns-2">
-
-<center>
-
-![](confoobio_sans.png)
-
-![](GMSE_Logo_Goose_340.png)
-
-</center>
-
-</div>
-
---->

### 2019-01-04 notebook/presentation/GMSE_presentation.Rmd
Up to gmse_apply in presentation

@@
 
+One function [shows](https://confoobio.github.io/gmse/reference/plot_gmse_results.html) the general results of simulation dynamics.
+
 ```{r, eval = FALSE}
@@
 ```
+
+Another function [shows](https://confoobio.github.io/gmse/reference/plot_gmse_effort.html) manager permissiveness, and stakeholder effort, for each action.
 
@@
 
+```{r, echo = FALSE}
+renderDataTable(
+    if(is.null(sim$data) == FALSE){
+        dat <- gmse_table(sim$data)
+        dat[,1:6];
+    }
+)
+```
+
+## {.smaller}
+
+<center>
+
+![](gmse_overview_sm_ga2.png)
+
+</center>
+
+## Custom sub-models with gmse_apply {.smaller}
+
+Custom models can be integrated into GMSE with `gmse_apply`, which [takes the arguments](https://confoobio.github.io/gmse/reference/gmse_apply.html) `res_mod`, `obs_mod`, `man_mod`, and `use_mod`.
+
+Consider a logistic growth model:
+
+$$N_{t+1} = N_{t} + rN_{t+1}\left(1 - \frac{N_{t}}{K}\right)$$
+
+Written as a function in R to take in values from $N_{t}$ and return $N_{t+1}$.
+
+```{r}
+logistic_res_mod <- function(N_0, K = 2000, gr = 1){
+    N_1 <- N_0 + gr * N_0 * ( 1 - N_0/K );
+    return(N_1);
+}
+
+mod_result <- logistic_res_mod(N_0 = 200, gr = 0.3);
+```
+
+```{r, echo = FALSE}
+print(mod_result);
+```
+
+The above function *and all of its arguments* can be integrated into `gmse_apply`.
+
+## Custom sub-models with gmse_apply {.smaller}
+
+```{r, echo = FALSE}
+mod <- gmse_apply(res_mod = function(N_0 = 200, K = 2000, gr = 0.3) N_1 <- N_0 + gr * N_0 * ( 1 - N_0/K ), stakeholders = 5);
+```
+
+```{r, eval = FALSE}
+mod <- gmse_apply(res_mod = logistic_res_mod, # Custom model
+                  N_0 = 200, gr = 0.3,        # Custom arguments
+                  stakeholders = 5);          # GMSE arguments
+```
+
+```{r, echo = FALSE}
+print(mod);
+```
+
+## {.smaller}
+
+<center>
+
+![](gmse_overview_sm_ga2.png)
+
+</center>
+
+## {.smaller}
+
 <br>

### 2019-01-04 notebook/presentation/GMSE_presentation.Rmd
Add near to end

@@
 ```{r, echo = FALSE}
-mod <- gmse_apply(res_mod = function(N_0 = 200, K = 2000, gr = 0.3) N_1 <- N_0 + gr * N_0 * ( 1 - N_0/K ), stakeholders = 5);
+mod <- gmse_apply(res_mod = function(N_0 = 200, K = 2000, gr = 0.3) N_1 <- N_0 + gr * N_0 * ( 1 - N_0/K ), stakeholders = 2);
 ```
@@
                   N_0 = 200, gr = 0.3,        # Custom arguments
-                  stakeholders = 5);          # GMSE arguments
+                  stakeholders = 2);          # GMSE arguments
 ```
@@
 </center>
+
+## Social conflict & management
+
+<div class = "columns-2">
+
+**General idea**
+
+- Detail 1
+- Detail 2
+- Detail 3
+
+<p class="forceBreak"></p>
+
+![](cusack.png)
+
+</div>
+
+##
+
+![](gmse_power.png)
+
+## Social conflict & management
+
+Results
+
+## {.smaller}
+
+<center>
+
+![](gmse_overview_sm.png)
+
+</center>
+
+## Future work PhD
+
+<div class = "columns-2">
+
+![](bach.jpeg)
+
+<p class="forceBreak"></p>
+
+**General idea of the paper**
+
+- Detail 1
+- Detail 2
+- Detail 3
+
+</div>
+
+## Future work ML
+
+Saro pictures
+
+## Future work ML
+
+Where next version will go
+
+## Summary and links
 

### 2019-01-04 notebook/presentation/GMSE_presentation.Rmd
More, closer to conclusion

@@
                             tend_crops = TRUE, stakeholders = 7,
-                            res_consume = 1, RESOURCE_ini = 600)
+                            res_consume = 1, RESOURCE_ini = 600,
+                            fixed_mark = 100, fixed_recapt = 200)
               showNotification("Simulation has finished")
@@
 </center>
+
+## Custom sub-models with gmse_apply {.smaller}
+
+Custom models can be integrated into GMSE with `gmse_apply`, which [takes the arguments](https://confoobio.github.io/gmse/reference/gmse_apply.html) `res_mod`, `obs_mod`, `man_mod`, and `use_mod`.
+
+```{r, eval = FALSE}
+new_sim <- gmse_apply(res_mod      = custom_resource_model,
+                      obs_mod      = custom_observation_model,
+                      man_mod      = manager, # GMSE default
+                      use_mod      = user,    # GMSE default
+                      stakeholders = 5,       # GMSE option
+                      custom_arg   = 200      # For custom function
+                      );
+```
+
+The `gmse_apply` function [needs to be looped](https://confoobio.github.io/gmse/articles/SI2.html#running-gmse-simulations-by-looping-gmse_apply) for modelling more than one time step (e.g., using the `old_list` argument).
 
@@
 
-## Social conflict & management
+## Social conflict and management {.smaller}
 
@@
 
-**General idea**
+**Investigate how disagreement over target population size affects natural resource management**
 
-- Detail 1
-- Detail 2
-- Detail 3
+<center>
+
+<br><br>
+
+![](cusack.png)
+
+</center>
+
+<br>
 
@@
 
-![](cusack.png)
+**General idea**
+
+- Model divergent goals expressed by conservationists and resource users (two stake-holders)
+- Modified `gmse_apply` to allow for policy lobbying and illegal harvesting
+- Varied manager 'impartiality' (susceptability to lobbying pressure) 
+- Compared population dynamics given conservationist and resource user disagreement over population target
+    - Manager ($N_{m}$)
+    - Conservationist ($N_{c}$)
+    - Resource user ($N_{u}$)
 
@@
 
-## Social conflict & management
+## Social conflict and management
 
@@
 
-## Future work ML
+## Behavioural games in conservation
 
-Saro pictures
+![](gabon3.jpg)
+
+## Behavioural games in conservation
+
+![](madagascar.jpg)
+
+## Behavioural games in conservation
+
+![](goosebump.png)
 

### 2019-01-05 notebook/presentation/GMSE_presentation.Rmd
Stable first draft

@@
 
+- Maximise resource use
 - Maximise land yield
-- Maximise resource use
 
@@
 
-## Social conflict and management
-
-Results
+##
+
+<center>
+
+![](gmse_power4.png)
+
+</center>
 
@@
 
-## Future work PhD
+## Using AI in population management
 
@@
 
+![](IAPETUS.png)
+
 <p class="forceBreak"></p>
 
-**General idea of the paper**
-
-- Detail 1
-- Detail 2
-- Detail 3
+**Using artificial intelligence to improve decision-making in conservation conflict**
+
+- Develop a framework for modelling AI for stake-holder decision-making
+- Develop AI software to simulate stake-holder interaction and negotiation
+- Apply agent-based decision-making to one or more case studies to predict long-term population dynamics
 
@@
 
-## Future work ML
-
-Where next version will go
-
-## Summary and links
+## AI of stake-holder decision-making
+
+<center>
+
+![](game_to_AI.png)
+
+</center>
+
+## {.smaller}
+
+<center>
+
+![](gmse_overview_sm.png)
+
+</center>
 
@@
 </center>
+
+# Supporting Information: Genetic algorithm
+
+<!--- 
+    START OF THE GENETIC ALGORITHM SET OF SLIDES  ------------------------------
+--->
+
+## Structure of a genetic algorithm
+
+<div class="columns-2">
+
+```{r, echo=FALSE}
+library(gamesGA);
+mbox <- function(x0, x1, y0, y1){
+    xx <- seq(from=x0, to=x1, length.out = 100);
+    yy <- seq(from=y0, to=y1, length.out = 100);
+    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
+    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
+    return(list(x=xd, y=yd));
+}
+par(mar=c(0,0,0,0));
+# ===============================================================
+plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
+     xlab="",ylab="", bty="n");
+abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
+polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
+text(x=22, y=85, labels="Random", col="black");
+text(x=22, y=80, labels="seed of", col="black");
+text(x=22, y=75, labels="strategies", col="black");
+# --------------------------------------------------------------
+cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
+polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
+arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
+dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
+polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=84, labels="Begin new", col="black");
+text(x=105, y=79, labels="generation", col="black");
+# ===============================================================
+```
+
+<br>
+
+**Table of strategies**
+
+```{r, echo=FALSE}
+randvals <- sample(x=1:5, size=32, replace=TRUE);
+agents   <- matrix(data=randvals, nrow=8);
+rownames(agents) <- c("Strategy 1", "Strategy 2", "Strategy 3", "Strategy 4",
+                      "Strategy 5", "Strategy 6", "Strategy 7", "Strategy 8");
+colnames(agents) <- c("L1", "L2", "L3", "L4");
+kable(agents, format="html");
+```
+
+</div>
+
+## Structure of a genetic algorithm
+
+<div class="columns-2">
+
+```{r, echo=FALSE}
+mbox <- function(x0, x1, y0, y1){
+    xx <- seq(from=x0, to=x1, length.out = 100);
+    yy <- seq(from=y0, to=y1, length.out = 100);
+    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
+    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
+    return(list(x=xd, y=yd));
+}
+par(mar=c(0,0,0,0));
+# ===============================================================
+plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
+     xlab="",ylab="", bty="n");
+abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
+polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
+text(x=22, y=85, labels="Random", col="black");
+text(x=22, y=80, labels="seed of", col="black");
+text(x=22, y=75, labels="strategies", col="black");
+# --------------------------------------------------------------
+cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
+polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
+arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
+dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
+polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=84, labels="Begin new", col="black");
+text(x=105, y=79, labels="generation", col="black");
+# --------------------------------------------------------------
+ebox <- mbox(x0 = 75, x1 = 135, y0 = 63, y1 = 73);
+polygon(x=ebox$x, y=ebox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=68, labels="recombination", col="black");
+# ===============================================================
+```
+
+<br>
+
+**Table of strategies**
+
+```{r, echo=FALSE}
+for(i in 1:dim(agents)[1]){
+    for(j in 1:dim(agents)[2]){
+        check <- runif(n=1);
+        if(check < 0.2){
+            other            <- sample(x=1:dim(agents)[1], size=1);
+            temp             <- agents[i,j];
+            agents[i,j]      <- agents[other, j];
+            agents[other, j] <- temp;
+        }
+    }
+}
+kable(agents, format="html");
+```
+
+</div>
+
+## Structure of a genetic algorithm
+
+<div class="columns-2">
+
+```{r, echo=FALSE}
+mbox <- function(x0, x1, y0, y1){
+    xx <- seq(from=x0, to=x1, length.out = 100);
+    yy <- seq(from=y0, to=y1, length.out = 100);
+    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
+    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
+    return(list(x=xd, y=yd));
+}
+par(mar=c(0,0,0,0));
+# ===============================================================
+plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
+     xlab="",ylab="", bty="n");
+abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
+polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
+text(x=22, y=85, labels="Random", col="black");
+text(x=22, y=80, labels="seed of", col="black");
+text(x=22, y=75, labels="strategies", col="black");
+# --------------------------------------------------------------
+cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
+polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
+arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
+dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
+polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=84, labels="Begin new", col="black");
+text(x=105, y=79, labels="generation", col="black");
+# --------------------------------------------------------------
+ebox <- mbox(x0 = 75, x1 = 135, y0 = 63, y1 = 73);
+polygon(x=ebox$x, y=ebox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=68, labels="recombination", col="black");
+# --------------------------------------------------------------
+fbox <- mbox(x0 = 75, x1 = 135, y0 = 51, y1 = 61);
+polygon(x=fbox$x, y=fbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=56, labels="mutation", col="black");
+# ===============================================================
+```
+
+<br>
+
+**Table of strategies**
+
+```{r, echo=FALSE}
+for(i in 1:dim(agents)[1]){
+    for(j in 1:dim(agents)[2]){
+        check <- runif(n=1);
+        if(check < 0.2){
+            agents[i,j]  <- sample(x=1:5, size=1);
+        }
+    }
+}
+kable(agents, format="html");
+```
+
+</div>
+
+## Structure of a genetic algorithm
+
+<div class="columns-2">
+
+```{r, echo=FALSE}
+mbox <- function(x0, x1, y0, y1){
+    xx <- seq(from=x0, to=x1, length.out = 100);
+    yy <- seq(from=y0, to=y1, length.out = 100);
+    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
+    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
+    return(list(x=xd, y=yd));
+}
+par(mar=c(0,0,0,0));
+# ===============================================================
+plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
+     xlab="",ylab="", bty="n");
+abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
+polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
+text(x=22, y=85, labels="Random", col="black");
+text(x=22, y=80, labels="seed of", col="black");
+text(x=22, y=75, labels="strategies", col="black");
+# --------------------------------------------------------------
+cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
+polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
+arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
+dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
+polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=84, labels="Begin new", col="black");
+text(x=105, y=79, labels="generation", col="black");
+# --------------------------------------------------------------
+ebox <- mbox(x0 = 75, x1 = 135, y0 = 63, y1 = 73);
+polygon(x=ebox$x, y=ebox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=68, labels="recombination", col="black");
+# --------------------------------------------------------------
+fbox <- mbox(x0 = 75, x1 = 135, y0 = 51, y1 = 61);
+polygon(x=fbox$x, y=fbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=56, labels="mutation", col="black");
+# --------------------------------------------------------------
+gbox <- mbox(x0 = 75, x1 = 135, y0 = 39, y1 = 49);
+polygon(x=gbox$x, y=gbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=44, labels="check fitness", col="black");
+# ===============================================================
+```
+
+<br>
+
+**Table of strategies**
+
+```{r, echo=FALSE}
+fit <- apply(agents,1,sum);
+kable(cbind(agents,fit), format="html");
+```
+
+</div>
+
+## Structure of a genetic algorithm
+
+<div class="columns-2">
+
+```{r, echo=FALSE}
+mbox <- function(x0, x1, y0, y1){
+    xx <- seq(from=x0, to=x1, length.out = 100);
+    yy <- seq(from=y0, to=y1, length.out = 100);
+    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
+    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
+    return(list(x=xd, y=yd));
+}
+par(mar=c(0,0,0,0));
+# ===============================================================
+plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
+     xlab="",ylab="", bty="n");
+abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
+polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
+text(x=22, y=85, labels="Random", col="black");
+text(x=22, y=80, labels="seed of", col="black");
+text(x=22, y=75, labels="strategies", col="black");
+# --------------------------------------------------------------
+cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
+polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
+arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
+dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
+polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=84, labels="Begin new", col="black");
+text(x=105, y=79, labels="generation", col="black");
+# --------------------------------------------------------------
+ebox <- mbox(x0 = 75, x1 = 135, y0 = 63, y1 = 73);
+polygon(x=ebox$x, y=ebox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=68, labels="recombination", col="black");
+# --------------------------------------------------------------
+fbox <- mbox(x0 = 75, x1 = 135, y0 = 51, y1 = 61);
+polygon(x=fbox$x, y=fbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=56, labels="mutation", col="black");
+# --------------------------------------------------------------
+gbox <- mbox(x0 = 75, x1 = 135, y0 = 39, y1 = 49);
+polygon(x=gbox$x, y=gbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=44, labels="check fitness", col="black");
+# --------------------------------------------------------------
+hbox <- mbox(x0 = 75, x1 = 135, y0 = 27, y1 = 37);
+polygon(x=hbox$x, y=hbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=32, labels="selection", col="black");
+# ===============================================================
+```
+
+<br>
+
+**Table of strategies**
+
+```{r, echo=FALSE}
+ord <- order(fit, decreasing=TRUE);
+agents <- agents[ord,];
+fit    <- fit[ord];
+kable(cbind(agents,fit), format="html");
+```
+
+</div>
+
+## Structure of a genetic algorithm
+
+<div class="columns-2">
+
+```{r, echo=FALSE}
+mbox <- function(x0, x1, y0, y1){
+    xx <- seq(from=x0, to=x1, length.out = 100);
+    yy <- seq(from=y0, to=y1, length.out = 100);
+    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
+    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
+    return(list(x=xd, y=yd));
+}
+par(mar=c(0,0,0,0));
+# ===============================================================
+plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
+     xlab="",ylab="", bty="n");
+abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
+polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
+text(x=22, y=85, labels="Random", col="black");
+text(x=22, y=80, labels="seed of", col="black");
+text(x=22, y=75, labels="strategies", col="black");
+# --------------------------------------------------------------
+cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
+polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
+arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
+dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
+polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=84, labels="Begin new", col="black");
+text(x=105, y=79, labels="generation", col="black");
+# --------------------------------------------------------------
+ebox <- mbox(x0 = 75, x1 = 135, y0 = 63, y1 = 73);
+polygon(x=ebox$x, y=ebox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=68, labels="recombination", col="black");
+# --------------------------------------------------------------
+fbox <- mbox(x0 = 75, x1 = 135, y0 = 51, y1 = 61);
+polygon(x=fbox$x, y=fbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=56, labels="mutation", col="black");
+# --------------------------------------------------------------
+gbox <- mbox(x0 = 75, x1 = 135, y0 = 39, y1 = 49);
+polygon(x=gbox$x, y=gbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=44, labels="check fitness", col="black");
+# --------------------------------------------------------------
+hbox <- mbox(x0 = 75, x1 = 135, y0 = 27, y1 = 37);
+polygon(x=hbox$x, y=hbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=32, labels="selection", col="black");
+# --------------------------------------------------------------
+ibox <- mbox(x0 = 75, x1 = 135, y0 = 15, y1 = 25);
+polygon(x=ibox$x, y=ibox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=20, labels="replacement", col="black");
+# ===============================================================
+```
+
+<br>
+
+**Table of strategies**
+
+```{r, echo=FALSE}
+agents <- rbind(agents[1,],
+                agents[1,],
+                agents[2,],
+                agents[2,],
+                agents[3,],
+                agents[3,],
+                agents[4,],
+                agents[4,]
+                );
+fit    <- c(fit[1], fit[1], fit[2], fit[2], fit[3], fit[3], fit[4], fit[4]);
+kable(cbind(agents,fit), format="html");
+```
+
+</div>
+
+## Structure of a genetic algorithm
+
+<div class="columns-2">
+
+```{r, echo=FALSE}
+mbox <- function(x0, x1, y0, y1){
+    xx <- seq(from=x0, to=x1, length.out = 100);
+    yy <- seq(from=y0, to=y1, length.out = 100);
+    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
+    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
+    return(list(x=xd, y=yd));
+}
+par(mar=c(0,0,0,0));
+# ===============================================================
+plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,100), xaxt="n", yaxt="n",
+     xlab="",ylab="", bty="n");
+abox <- mbox(x0 = 0, x1 = 45, y0 = 70, y1 = 90);
+polygon(x=abox$x, y=abox$y, lwd=3, border="red", col="rosybrown1");
+text(x=22, y=85, labels="Random", col="black");
+text(x=22, y=80, labels="seed of", col="black");
+text(x=22, y=75, labels="strategies", col="black");
+# --------------------------------------------------------------
+cbox <- mbox(x0 = 70, x1 = 140, y0 = 0, y1 = 90);
+polygon(x=cbox$x, y=cbox$y, lwd=3, border="goldenrod4", col="gold3");
+arrows(x0=46, x1=70, y0=82, y1=82, lwd=2, length=0.15);
+dbox <- mbox(x0 = 75, x1 = 135, y0 = 75, y1 = 88);
+polygon(x=dbox$x, y=dbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=84, labels="Begin new", col="black");
+text(x=105, y=79, labels="generation", col="black");
+# --------------------------------------------------------------
+ebox <- mbox(x0 = 75, x1 = 135, y0 = 63, y1 = 73);
+polygon(x=ebox$x, y=ebox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=68, labels="recombination", col="black");
+# --------------------------------------------------------------
+fbox <- mbox(x0 = 75, x1 = 135, y0 = 51, y1 = 61);
+polygon(x=fbox$x, y=fbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=56, labels="mutation", col="black");
+# --------------------------------------------------------------
+gbox <- mbox(x0 = 75, x1 = 135, y0 = 39, y1 = 49);
+polygon(x=gbox$x, y=gbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=44, labels="check fitness", col="black");
+# --------------------------------------------------------------
+hbox <- mbox(x0 = 75, x1 = 135, y0 = 27, y1 = 37);
+polygon(x=hbox$x, y=hbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=32, labels="selection", col="black");
+# --------------------------------------------------------------
+ibox <- mbox(x0 = 75, x1 = 135, y0 = 15, y1 = 25);
+polygon(x=ibox$x, y=ibox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=20, labels="replacement", col="black");
+# --------------------------------------------------------------
+jbox <- mbox(x0 = 75, x1 = 135, y0 = 3, y1 = 13);
+polygon(x=jbox$x, y=jbox$y, lwd=3, border="goldenrod4", col="white");
+text(x=105, y=8, labels="converge?", col="black");
+# --------------------------------------------------------------
+bbox <- mbox(x0 = 0, x1 = 45, y0 = 20, y1 = 40);
+polygon(x=bbox$x, y=bbox$y, lwd=3, border="blue", col="lightsteelblue1");
+arrows(x0=70, x1=22.5, y0=5, y1=5, lwd=2, length=0.0);
+arrows(x0=22, x1=22, y0=5, y1=20, lwd=2, length=0.15);
+arrows(x0=55, x1=55, y0=5, y1=82, lwd=2, length=0.15);
+text(x=15, y=10, labels="Yes", col="black", srt=90);
+text(x=50, y=10, labels="No", col="black", srt=90);
+text(x=22, y=33, labels="Adaptive", col="black");
+text(x=22, y=28, labels="strategies", col="black");
+# ===============================================================
+```
+
+<br>
+
+**Table of strategies**
+
+```{r, echo=FALSE}
+kable(cbind(agents,fit), format="html");
+```
+
+</div>
+
+<!--- 
+    END OF THE GENETIC ALGORITHM SET OF SLIDES  --------------------------------
+--->
+
+# Example of a genetic algorithm for Prisoner's dilemma
+
+## Iterated Prisoner's dilemma rules {.smaller}
+
+<div class="columns-2">
+
+![](payoff_matrix_sm.png)
+**General structure of Prisoner's dilemma scenario**
+
+![](pd_pays.png)
+**Specific payoffs satisfying a Prisoner's dilemma scenario**
+
+</div>
+
+## Use of genetic algorithms {.smaller}
+
+**Each locus is an focal player's choice given an opponent's past three choices (plus one choice of first move)**
+
+<hr>
+
+|                  | **CCC** | **CCD** | **CDC** | **CDD** | **DCC** | **DCD** | **DDC** | **DDD** | **1st** |
+|------------------|---------|---------|---------|---------|---------|---------|---------|---------|---------|
+| **Strategy 1**   | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  |
+| **Strategy 2**   | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  |
+| **...**          |   ...   |   ...   |   ...   |   ...   |   ...   |   ...   |   ...   |   ...   |   ...   |
+| **Strategy N-1** | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  |
+| **Strategy N**   | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  | Choice  |
+
+<hr>
+
+**Total of 512 possible strategies**
+
+## Use of genetic algorithms {.smaller}
+
+```{r, echo=FALSE}
+inputPanel(
+
+  numericInput("CC", label = "CC", 0, width="30%"),
+  numericInput("DC", label = "DC", 0, width="30%"),
+  numericInput("CD", label = "CD", 0, width="30%"),
+  numericInput("DD", label = "DD", 0, width="30%"),
+  style='width: 2000px; height: 60px'
+)
+```
+
+<br>
+
+```{r, echo=FALSE}
+gres <- reactive({ 
+     cc <- input$CC;
+     dc <- input$DC;
+     cd <- input$CD;
+     dd <- input$DD;
+     rd <- 100;
+     gn <- 100;
+     co <- 0.05;
+     mu <- 0.05;
+
+     game_res <- games_ga(CC = cc, DC = dc, CD = cd, DD = dd, rounds = rd,
+                          generations = gn, cross_prob = co, 
+                          mutation_prob = mu, num_opponents = 100);     
+     all <- c(game_res$genos, "break_1",
+              rownames(game_res$genos), "break_2",
+              game_res$fitness);
+    })
+```
+
+```{r, echo = FALSE}
+renderPlot({
+    break2    <- which(gres()=="break_2")[1]+1;
+    end       <- length(gres());
+    fitnesses <- gres()[break2:end];
+    fitnesses <- as.numeric(fitnesses);
+    mean_fit  <- fitnesses / (100 * 100);
+    maxpt     <- max(c(input$CC, input$CD, input$DC, input$DD));
+    par(mar=c(5,5,1,1));
+    plot(x = 1:length(mean_fit), y=mean_fit, type="l", ylim=c(0,maxpt), lwd = 3,
+        xlab="Generation", ylab = "Mean strategy fitness per round",
+         cex.axis=1.5, cex.lab=1.5);
+})
+```
+
+## Adaptive strategies simulated {.smaller}
+
+```{r, echo=FALSE}
+renderTable({
+      break1   <- which(gres()=="break_1")[1] - 1;
+      break2   <- which(gres()=="break_2")[1] - 1;
+      geno_vec <- gres()[1:break1];
+      geno_row <- gres()[(break1+2):break2];
+      genos    <- matrix(data =  geno_vec, ncol = 10, byrow = FALSE);
+      rownames(genos) <- geno_row;
+      colnames(genos) <- c("CCC", "CCD", "CDC", "CDD", "DCC", "DCD", "DDC",
+                           "DDD", "1st", "Final %");
+      final_table <- genos[1:10,];
+
+}, include.rownames=TRUE)
+```

### 2019-11-13 notebook/ms/SI7.Rmd
Committing local changes

@@
 header-includes: 
-- \usepackage{lineno}
-- \linenumbers
 link-citations: yes

### 2019-11-13 notebook/ms/ms.Rmd
Rework SIs and MS with new plotting function
Committing local changes

@@
 - \fancyhead[R]{GMSE, AN R PACKAGE}
-- \usepackage{lineno}
-- \linenumbers
-linestretch: 2
+linestretch: 1
 link-citations: yes
@@
 ******************************
+
+<!--- 
+Word count is 3497, including figure legends and references.
+--->
 
@@
 
-\newpage
-
 Abstract
@@
 2.
-We present the R package GMSE, which applies genetic algorithms to provide a genearalised tool for simulating adaptive decision-making management scenarios between stakeholders with competing objectives under complex social-ecological interactions and uncertainty.
+We present the R package GMSE, which applies genetic algorithms to provide a generalised tool for simulating adaptive decision-making management scenarios between stakeholders with competing objectives under complex social-ecological interactions and uncertainty.
 3.
@@
 1).
-The function `gmse` runs simulations using four predefined submodels, which can be parameterised to fit various case studies; more tailored submodels can also be defined using the `gmse_apply` function (see [SI2](SI1.pdf)).
+The function `gmse` runs simulations using four predefined individual-based submodels, which can be parameterised to fit various case studies; more tailored submodels can also be defined using the `gmse_apply` function (see [SI2](SI2.pdf)).
 By default, GMSE models (1) a population of discrete resources (e.g., a managed species) with individual traits (e.g., location, age) on a spatially-explicit landscape and simulates resource birth, individual movement between landscape cells, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty.
 This sub-model is unique in not relying on other sub-models because ecological dynamics can be simulated in the absence of observation and management.
-(2) Observation is modelled in one of four ways: resource counting on a subset of landscape cells [e.g., @Nuno2013], marking and recapturing a fixed number of resources, and resource counting across the whole landscape either one linear transect or one rectangular block at a time (during which resources might move).
+(2) Observation is modelled in one of three ways: resource counting on a subset of landscape cells [e.g., @Nuno2013], marking and recapturing a fixed number of resources, or counting resources across the whole landscape one set of landscape cells at a time (during which resources might move).
 Sampling error from all observation types generates a range of uncertainties that depend on monitoring effort.
@@
 Consistent with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for agents (stakeholders).
-Instead, genetic algorithms are used to heuristically find an adaptive strategy for each stakeholder in each time step (see [SI1](SI1.pdf) for details).
+Instead, a genetic algorithm is used to heuristically find strategies that reflect the individual objectives of each stakeholder in each time step (see [SI1](SI1.pdf) for details).
 Critically, all stakeholders involved in resource conservation are constrained in their decision-making; managing and using resources takes effort (e.g., time or money), and effort expended in developing or enforcing one policy (for managers) or performing one action (for users) will be effort not expendable elsewhere [@Milner-Gulland2011; @Muller-Hansen2017; @Schluter2017].
@@
 Using GMSE, we can simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy.
-We consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity is 2000.
+We consider a population of waterfowl with an initial abundance and manager target abundance of 200, but whose carrying capacity is 2000.
 Waterfowl consume and destroy all crop yield upon arrival to a landscape cell.
@@
 sim <- gmse(land_ownership = TRUE, stakeholders = 5, observe_type = 0, 
-            res_death_K = 2000, manage_target = 1000, RESOURCE_ini = 1000, 
+            res_death_K = 2000, manage_target = 1000, RESOURCE_ini = 200, 
             user_budget = 1000, manager_budget = 1000, res_consume = 1, 
@@
 ```{r, eval = FALSE}
-plot_gmse_results(res = sim$resource, obs = sim$observation, land = sim$land, 
-                  agents = sim$agents, paras = sim$paras, ACTION = sim$action, 
-                  COST = sim$cost);
+plot_gmse_results(sim_results = sim);
 ```
@@
 The upper right panel shows the same landscape broken down into five differently coloured regions, which correspond to areas of land owned by each of the five farmers.
-The middle left panel shows the actual abundance of resources (black solid line), and the abundance of resources as estimated by the manager (blue solid line; shading indicates 95 percent confidence intervals), over time.
+The middle left panel shows the actual abundance of resources (black solid line; i.e., `natural resources' or `operating' model), and the abundance of resources as estimated by the manager (blue solid line; i.e., `observation' or `assessment' model; shading indicates 95 percent confidence intervals), over time.
 The horizontal dotted red and blue lines show the landscape-level resource carrying capacity enacted on adult mortality and the manager's target for resource abundance, respectively.
@@
 The middle right panel shows total percent yield of landscape cells for each individual farmer, differentiated by colour, where line colours correspond to areas of the landscape in the upper right panel.
-The lower left panel shows the cost of farmers performing actions over time, as set by the manager.
-The lower right panel shows the total number of actions attempted to be performed by all farmers over time (some actions might be unsuccessful if resources are unavailable on a farmer's land to cull or scare, so, e.g., culling actions might be larger than resources actually culled)."}
+The lower left panel shows the cost of farmers performing actions over time, as set by the manager; the upper limit on cost of actions reflects the manager's limited budget for setting policy.
+The lower right panel shows the total number of actions attempted to be performed by all farmers over time (some actions might be unsuccessful if resources are not on a farmer's land to cull or scare, so, e.g., culling actions might be larger than resources actually culled)."}
 plot_gmse_results_vign(res = sim$resource, obs = sim$observation, 
@@
 }
-sim <- gmse_apply(res_mod = logistic_res_mod, X_0 = 1000, gr = 0.3, stakeholders = 5);
+sim <- gmse_apply(res_mod = logistic_res_mod, X_0 = 200, gr = 0.3, stakeholders = 5);
 ```
@@
 
-The GMSE package can be downloaded from CRAN (https://cran.r-project.org/package=GMSE) or GitHub (https://github.com/bradduthie/gmse). GMSE is open source under GNU Public License.
+The GMSE package can be downloaded from CRAN (https://cran.r-project.org/package=GMSE) or GitHub (https://confoobio.github.io/gmse/). GMSE is open source under GNU Public License.
 
@@
 GMSE provides a powerful tool for individual-based modelling simulations while also allowing for extensive model customisation.
-GMSE vignettes provide additional examples for getting started with `gmse` or `gmse_gui` for simulations using default resource, observation, manager, and user sub-models ([SI3](SI3.pdf) and [SI6](SI6.pdf)), and for using `gmse_apply` for advanced model customisation ([SI2](SI2.pdf), [SI4](SI4.pdf), and [SI7](SI7.pdf)) and integration with existing packages ([SI5](SI5.pdf)).
+GMSE vignettes provide additional examples for getting started with `gmse` or `gmse_gui` for simulations using default resource, observation, manager, and user sub-models ([SI3](SI3.pdf) and [SI6](SI6.pdf)), and for using `gmse_apply` for advanced model customisation ([SI2](SI2.pdf), [SI4](SI4.pdf), and [SI7](SI7.pdf)) and integration with existing packages (e.g., Fisheries Library in R, see [SI5](SI5.pdf)).
 Future versions of GMSE will include additional features and improve upon the realism of social and ecological modelling components, while also maintaining a high degree of flexibility and modulatarity for model customisation.

### 2020-02-04 notebook/ms/SI4.Rmd

@@
 sim_old  <- gmse_apply(get_res = "Full", remove_pr = 0.122, lambda = 0.275,
-                       res_d4eath_K = 93870, RESOURCE_ini = 35000, 
+                       res_death_K = 93870, RESOURCE_ini = 35000, 
                        manage_target = 70000, res_death_type = 3,

### 2020-05-12 notebook/ms/SI7.Rmd
Update SI7

@@
 12. Resource tally parameter (currently used only for managers)
-13. Unused column 1
-14. Unused column 2
-15. Unused column 3
+13. How many resources an agent has in view (unused)
+14. Total number of cells owned by an agent
+15. Unused column
 16. Yield from owned land (zero for users when default `land_ownership = FALSE`)
@@
 20. Has the resource's offspring production been increased by an agent?
+21. How much yield has the resource consumed?
+22. How many times can the resource consume yield in one time step?
 
@@
 In addition, all landscapes have three layers, which hold three separate values of information for each x-y location.
-The first layer is unused in GMSE v0.4.0.7; the second layer holds crop production on a cell, and the third layer holds the owner of the cell (corresponding to the ID of an agent, where the manager's ID = 0 defines public land).
+The first layer is unused as of GMSE v0.4.0.7; the second layer holds crop production on a cell, and the third layer holds the owner of the cell (corresponding to the ID of an agent, where the manager's ID = 0 defines public land).
 An $8 \times 8$ portion of the landscape from `sim` is shown below.
@@
 
-In the case of the above, all of the cells in this square patch of landscape are owned by agent 1 (i.e., the manager; see `sim$LAND[,,3]`), and we can see that crop production on this patch of land has been decreased from 1 in several cells as a consequence of consumption by resources (see `sim$LAND[„2]`).
+In the case of the above, all of the cells in this square patch of landscape are owned by agent 1 (i.e., the manager; see `sim$LAND[,,3]`).
+This occurs if `land_ownership = FALSE` and users are prohibited from owning land.
+We can also see that crop production on this patch of land has been decreased from its initialised value of 1 in several cells as a consequence of consumption by resources (see `sim$LAND[„2]`).
+By default, landscape yield starts each time step at a value of 1, and hence replenishes in each time even if consumed in the previous time step.
+Users that own landscape cells (when `land_ownership = TRUE`) can increase yield above 1 using the `tend_crop_yld` action.
+If resources consume landscape yield, the value of yield can drop to a minimum value of zero.
+
 In [SI4](SI4.pdf), we show how landscape cell values can be manipulated to customise the placement of land ownership.
+By default, GMSE v0.6.0.0+ places the ownership of landscape cells using a shortest-splitline algorithm (see the lab notebook for [a detailed explanation](https://confoobio.github.io/gmse/notebook/gmse_notes.html#u-10-MAY-2020)).
+The algorithm splits the landscape evenly into smaller and smaller rectangles until a sufficient number of rectangles are produced for all users to own a similar amount of land (note that how uniformly land is partitioned among users will naturally be influenced by how evenly the total number of landscape cells can be divided among them).
+Below is an example in which land of `land_dim_1 = 64` and `land_dim_2 = 64` is partitioned equally among eight users.
+
+```{r, echo = FALSE, fig.height = 5, fig.width = 5}
+par(mar = c(0.2, 0.2, 0.2, 0.2));
+land_1 <- owner_land_ssa(dim_x = 64, dim_y = 64, owners = 8, public_land = 0);
+image(land_1, xaxt = "n", yaxt = "n", col = topo.colors(max(land_1)));
+```
+
+Note that all users own exactly `r mean(as.vector(table(land_1)))` cells on the landscape.
+A bit more complexity can be generated by an uneven landscape of `land_dim_1 = 80` and `land_dim_2 = 100`, with land allocated among nine users.
+
+```{r, echo = FALSE, fig.height = 4, fig.width = 5}
+par(mar = c(0.2, 0.2, 0.2, 0.2));
+land_2 <- owner_land_ssa(dim_x = 100, dim_y = 80, owners = 9, public_land = 0);
+image(land_2, xaxt = "n", yaxt = "n", col = topo.colors(max(land_2)));
+```
+
+In the case of the above landscape, the number of cells allocated to each user ranges from `r min(as.vector(table(land_2)))` to `r max(as.vector(table(land_2)))`.
+
+Public land is placed on the landscape whenever `public_land > 0`, with `public_land` defining the proportion of landscape cells that are not be owned by an user.
+When the expected number of cells to be allocated to public land is calculated to be equal to or greater than that allocated to a single user, public land is allocated in a block as is with users.
+For example, the below shows a landscape in which `public_land = 0.5`, with seven other users owning land.
+
+```{r, echo = FALSE, fig.height = 5, fig.width = 5}
+par(mar = c(0.2, 0.2, 0.2, 0.2));
+land_3 <- owner_land_ssa(dim_x = 100, dim_y = 100, owners = 7, 
+                         public_land = 0.5);
+image(land_3, xaxt = "n", yaxt = "n", col = topo.colors(max(land_3)));
+```
+
+Public land is in dark blue above.
+If, however, the amount of public land is very small, the land is added to the centre of the landscape.
+Below illustrates a landscape that also includes seven land owning users, but with `public_land = 0.05`.
+
+```{r, echo = FALSE, fig.height = 5, fig.width = 5}
+par(mar = c(0.2, 0.2, 0.2, 0.2));
+land_4 <- owner_land_ssa(dim_x = 100, dim_y = 100, owners = 7, 
+                         public_land = 0.1);
+image(land_4, xaxt = "n", yaxt = "n", col = topo.colors(max(land_4)));
+```
+
+Where `public_land` is small, the default landscape ownership algorithm thereby attempts to be as precise as possible in allocating cell numbers.
+Of course, as previously mentioned, ownership can also be manually specified when using `gmse_apply`.
 
@@
 An example of how to do this within a loop using `gmse_apply` can be found in [SI4](SI4.pdf).
-While other data structures exist within GMSE (e.g., see the output of `gmse_apply` when `get_res = Full`), we do not recommend manipulating these structures for custom simulations.
+While other data structures exist within GMSE (e.g., see the output of `gmse_apply` when `get_res = "Full"`), we do not recommend manipulating these structures for custom simulations.
 
-Many data structures contain elements that are unused in GMSE v0.4.0.7, and in all cases this is designed for ease of ongoing development of new GMSE features.
+Many data structures contain elements that are unused in GMSE v0.6.0.0, and in all cases this is designed for ease of ongoing development of new GMSE features.
 Requests for new features can be made on GitHub using the [GMSE Wiki](https://github.com/bradduthie/gmse/wiki) or the [GMSE Issues](https://github.com/bradduthie/gmse/issues) page.
+
+Reference
+================================================================================

### 2020-05-12 notebook/presentation/GMSE_presentation.Rmd

@@
 - Compatibile with agent-based modelling
-- Can suggest management strategiesand be used to understand stake-holder actions
+- Can suggest management strategies and be used to understand stake-holder actions
 - Conceptual unification
@@
 
+<center>
+
+![](meadow.jpg)
+
+[Wing-Chi Poon](https://commons.wikimedia.org/wiki/File:Mist_Covering_a_Meadow_under_Forest_Encroachment.jpg)
+
+</center>
+
+## {.smaller}
+
 <br>

### 2020-05-21 notebook/ms/SI8.Rmd
Some minor edits to the vignette by @AdrianBach to ensure that the plots are produced just once when knitting

---
title: 'Adaptive timing of investment strategy'
subtitle: 'GMSE: an R package for generalised management strategy evaluation (New function vignette)'
author: Adrian Bach&#xb9; &#xb2;
date: "[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK [2] adrian.bach@stir.ac.uk"
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes:
- \usepackage{lineno}
- \linenumbers
linestretch: 1
bibliography: gmse.bib
biblio-style: apalike
link-citations: true
linkcolor: blue
---

```{r, echo = FALSE}
library(GMSE);
# sim <- gmse(action_thres = 0.1, budget_bonus = 0.5, plotting = FALSE);
```

Adaptive timing of investment strategy
================================================================================

In the context of adaptive management of conservation conflict, managers tend to invest their ressources in updating the conservation policy as soon as possible. But leaving the policy unchanged when the population is close enough to the target could actually save money, time, and energy, and enpower managers for the next iteration.

In this GMSE option, the notion of 'close enough to the target' is represented by the `action_thres` parameter (i.e., 'action threshold', which can take values between 0 and 1). The parameter is a fraction of the manager target (`manage_target`) added up and down the target value, defining a band around it. When managers observe the population size to be within this band, they assume that the current policy is satisfactory, so the genetic algorithm is not called and the policy is kept unchanged. For example, with `action_thres = 0.1` and `manage_target = 1000`, the manager will update the policy only if the population is under 900, or above 1100 individuals.

The notion of time, money, and energy saved when leaving the policy unchanged is represented by the `budget_bonus` parameter. If, in a previous time step, managers did not update the policy, they will receive an extra fraction of their initial budget in the following time step. For example, with `budget_bonus = 0.5` and `manage_budget = 1000`, this extra fraction will be 500 arbitrary budget units. This budget bonus is cumulative, so managers can stack budget bonuses if they do not update the policy for several consecutive time steps, but they lose it all as soon as they do.

Example of utilisation
================================================================================

Here is a default simulation where managers update every time step.

```{r, echo = TRUE, eval = FALSE}
sim.fli <- gmse();
```

```{r, echo = FALSE, eval = TRUE, fig.height = 8, fig.width = 6}
sim.fli <- gmse(plotting = FALSE);
plot_gmse_results(sim.fli);
```

And here the same simulation, exept managers update policy only when they observe the population to be plus or minus the target by 20%, and receive a bonus of 30% of their initial budget when they do not.

```{r, echo = TRUE, eval = FALSE}
sim.ati <- gmse(action_thres = 0.2, budget_bonus = 0.3);
```

```{r, echo = FALSE, eval = TRUE, fig.height = 8, fig.width = 6}
sim.ati <- gmse(action_thres = 0.2, budget_bonus = 0.3, plotting = FALSE);
plot_gmse_results(sim.ati);
```

Note how the costs plateau (lower left panel) when the population monitored (blue line in the upper left panel) is in between the two green dotted lines. Leaving the policy unchanged in these cases allowed managers to set the costs higher than the default maximum of 110 budget units. It also affects the number of cullings performed by users (lower right panel), which is more stable than when managers update policy every time step.

### 2020-05-21 notebook/ms/SI7.Rmd
Update the vignette here to account for the new columns in the agents array, just a heads up @jejoenje if you're editing the land parts

@@
 16. Yield from owned land (zero for users when default `land_ownership = FALSE`)
-17. Budget
+17. Agent's baseline budget
+18. Agent's perception of the efficacy of scaring one resource
+19. Agent's perception of the efficacy of culling one resource
+20. Agent's perception of the efficacy of castrating one resource
+21. Agent's perception of the efficacy of feeding one resource
+22. Agent's perception of the efficacy of helping offspring of one resource
+23. Agent's perception of the efficacy of tending the crop on one landscape cell
+24. Agent's perception of the efficacy of killing the crop on one landscape cell
+25. Increment added to the manager's (row 1) baseline budget caused by a budget bonus
+26. Increment added to an agent's budget caused by yield
+27. Unused column
 
-It is obvious from the above list that most columns represent traits that are either typically ignored or currently not in use.
+Note that for the perceptions of the efficacy of different actions (columns 18-24), this relates to whatever it is that the agents are acting upon.
+For example, for columns 18-22, actions are on the resources, so if an agent's perception of the efficacy of scaring is -1, this means that they believe that scaring a resource will decrease the number of resources affecting them by 1 (lower default values for landowners are caused based on the probability of a scared resource landing on a different cell owned by the same agent).
+For columns 23-24, actions are on the landscape, so if an agent tends the crop on one landscape cell, this value defines an agent's perception of how much yield on that cell will increase.
+
+It is obvious from the above list that several columns represent traits that are either typically ignored or currently not in use.
 This is intended to allow for easier future development of default model options and potential customisation of sub-models in `gmse_apply`.
@@
 par(mar = c(0.2, 0.2, 0.2, 0.2));
-land_1 <- owner_land_ssa(dim_x = 64, dim_y = 64, owners = 8, public_land = 0);
+land_1 <- owner_land_ssa(dim_x = 64, dim_y = 64, owners = 8, public_land = 0, 
+                         land_var = 0);
 image(land_1, xaxt = "n", yaxt = "n", col = topo.colors(max(land_1)));
@@
 par(mar = c(0.2, 0.2, 0.2, 0.2));
-land_2 <- owner_land_ssa(dim_x = 100, dim_y = 80, owners = 9, public_land = 0);
+land_2 <- owner_land_ssa(dim_x = 100, dim_y = 80, owners = 9, public_land = 0,
+                         land_var = 0);
 image(land_2, xaxt = "n", yaxt = "n", col = topo.colors(max(land_2)));
@@
 land_3 <- owner_land_ssa(dim_x = 100, dim_y = 100, owners = 7, 
-                         public_land = 0.5);
+                         public_land = 0.5, land_var = 0);
 image(land_3, xaxt = "n", yaxt = "n", col = topo.colors(max(land_3)));
@@
 land_4 <- owner_land_ssa(dim_x = 100, dim_y = 100, owners = 7, 
-                         public_land = 0.1);
+                         public_land = 0.1, land_var = 0);
 image(land_4, xaxt = "n", yaxt = "n", col = topo.colors(max(land_4)));

### 2020-05-21 notebook/ms/SI4.Rmd
Some small modifications to SI4 in progress

@@
 - [`sim_new$resource_array`](SI7.pdf#resources): A table holding all information on resources.
-Rows correspond to discrete resources, and columns correspond to resource properties: (1) ID, (2-4) types (not currently in use), (5) x-location, (6) y-location, (7) movement parameter, (8) time, (9) density independent mortality parameter (`remove_pr`), (10) reproduction parameter (`lambda`), (11) offspring number, (12) age, (13-14) observation columns, (15) consumption rate (`res_consume`), and (16-20) recorded experiences of user actions (e.g., was the resource culled or scared?).
+Rows correspond to discrete resources, and columns correspond to resource properties: (1) ID, (2-4) types (not currently in use), (5) x-location, (6) y-location, (7) movement parameter, (8) time, (9) density independent mortality parameter (`remove_pr`), (10) reproduction parameter (`lambda`), (11) offspring number, (12) age, (13-14) observation columns, (15) consumption rate (`res_consume`), (16-20) recorded experiences of user actions (e.g., was the resource culled or scared?), (21) how much yield has the resource consumed, and (22) how many times the resource can consume yield in one time step.
 - [`sim_new$AGENTS`](SI7.pdf#agents): A table holding basic information on agents (manager and users).
-Rows correspond to a unique agent, and columns correspond to agent properties: (1) ID, (2) type (0 for the manager, 1 for users), (3-4) additional type options not currently in use, (5-6), x and y locations (usually ignored), (7) movement parameter (usually ignored), (8) time, (9) agent's viewing ability in cells (`agent_view`), (10) error parameter, (11-12) values for holding marks and tallies of resources, (13-15) values for holding observations, (16) yield from landscape cells, (17) budget (`manager_budget` and `user_budget`).
+Rows correspond to a unique agent, and columns correspond to agent properties: (1) ID, (2) type (0 for the manager, 1 for users), (3-4) additional type options not currently in use, (5-6), x and y locations (usually ignored), (7) movement parameter (usually ignored), (8) time, (9) agent's viewing ability in cells (`agent_view`), (10) error parameter, (11-12) values for holding marks and tallies of resources, (13-15) values for holding observations, (16) yield from landscape cells, (17) baseline budget (`manager_budget` and `user_budget`), (18-24) agent's perception of the efficacy of scaring, culling, castrating, feeding, helping, tending crops, and killing crops, (25-26) increments to budget, (27) unused.
 - `sim_new$observation_vector`: Estimate of total resource number from the observation model ([`observation_array`](SI7.pdf#observation) also holds this information in a different way depending on `observe_type`)
@@
 
-By default, all farmers in GMSE are allocated the same number of landscape cells, which are simply placed in order of the farmer's ID.
+By default, all farmers in GMSE are allocated roughly the same number of landscape cells, which are placed on the landscape using a shortest-splitline algorithm that makes similar size rectangles.
+In the `LAND` array, ownership is designated by the agent's ID.
 Public land is produced by placing landscape cells that are technically owned by the manager, and therefore have landscape cell values of 1.
@@
 
-```{r, fig.cap = "Land ownership by farmers as customised in gmse_apply."}
+```{r, fig.cap = "Land ownership by farmers as customised in gmse_apply.", fig.height = 6, fig.width = 6}
 sim_old$LAND[1:20,   1:20,  3] <- 2; 
@@
 
-Perhaps we want to assume that farmers have different budgets, which are correlated in some way to the number of landscape cells that they own.
-Custom user budgets can be set by manipulating `sim_old$AGENTS`, the last column of which (column 17) holds the budget for each user.
-Agent IDs (as stored on the landscape above) correspond to rows of `sim_old$AGENTS`, so individual budgets can be directly input as desired.
+Perhaps we want to assume that farmers have different baseline budgets, which are correlated in some way to the number of landscape cells that they own.
+Custom user baseline budgets can be set by manipulating `sim_old$AGENTS`, column 17 of which holds the budget for each user.
+Agent IDs (as stored on the landscape above) correspond to rows of `sim_old$AGENTS`, so individual baseline budgets can be directly input as desired.
 We can do this manually (e.g., `sim_old$AGENTS[2, 17] <- 4000`), or, alternatively, if farmer budget positively correlates to landscape owned, we can use a loop to input values as below.
@@
 Future versions of GMSE are likely to expand on the built-in options available for simulation; requests for such expansions, or contributions, can be submitted to [GitHub](https://github.com/bradduthie/gmse/issues).
+
+References
+================================================================================

### 2020-05-22 notebook/ms/SI7.Rmd
Started working on some extra text in SI7 for ownership_var, examples to follow

@@
 In addition, all landscapes have three layers, which hold three separate values of information for each x-y location.
-The first layer is unused as of GMSE v0.4.0.7; the second layer holds crop production on a cell, and the third layer holds the owner of the cell (corresponding to the ID of an agent, where the manager's ID = 0 defines public land).
+The first layer is unused as of GMSE v0.6.0.0; the second layer holds crop production on a cell, and the third layer holds the owner of the cell (corresponding to the ID of an agent, where the manager's ID = 0 defines public land).
 An $8 \times 8$ portion of the landscape from `sim` is shown below.
@@
 In the case of the above landscape, the number of cells allocated to each user ranges from `r min(as.vector(table(land_2)))` to `r max(as.vector(table(land_2)))`.
+
+In addition, whereas in previous versions variable landscape ownership among users had to be customised using `gmse_apply`, GMSE v.
+0.6.0.0+ now includes an additional parameter `ownership_var`, which takes effect when `land_ownership = TRUE`.
+The value of this parameter is expected to be >=0 and <1.
+At its default value of 0, land ownership is equal among users as demonstrated in the code above.
+Values >0 increase the extent of variability of landownership among users.
+
+*** Working on examples here ***
 

### 2020-05-22 notebook/ms/SI7.Rmd
Added ownership_var example to SI7, feel free to change/add.

@@
 
-In addition, whereas in previous versions variable landscape ownership among users had to be customised using `gmse_apply`, GMSE v.
+In addition, [whereas in previous versions variable landscape ownership among users had to be customised using `gmse_apply`](SI4.pdf), GMSE v.
 0.6.0.0+ now includes an additional parameter `ownership_var`, which takes effect when `land_ownership = TRUE`.
@@
 At its default value of 0, land ownership is equal among users as demonstrated in the code above.
-Values >0 increase the extent of variability of landownership among users.
+Values >0 increase the extent of variability of landownership among users; in technical terms, `ownership_var` controls the extent to which each split in the shortest-splitline algorithm is proportional to the number of users the landscape is to be divided into, with larger numbers shifting the split away from even distribution.
+In practical terms, this means that increasing values of `ownership_var` produce more unevenly distributed landscapes among users.
+For example, a value of `ownership_var = 0.1` produces a relatively limited about of inequity in landscape distribution among users.
 
-*** Working on examples here ***
+```{r, echo = FALSE, fig.height = 5, fig.width = 10}
+par(mfrow=c(1,2))
+par(mar = c(2, 2, 2, 2))
+xd = 100
+yd = 100
+l1 = owner_land_ssa(dim_x = xd, dim_y = yd, owners = 48, public_land = 0, land_var = 0.2)
+tl1 = table(l1)/(xd*yd)
+image(l1, xaxt = "n", yaxt = "n", main = "ownership_var = 0.2", col = topo.colors(max(land_2)))
+hist(tl1, xlim = c(0,0.16), col = "grey", breaks = 12, main = "")
+```
+As can be seen from the frequency distribution, in this case, all users own between `r round((min(table(l1)/(xd*yd)))*100,1)`% and `r round((max(table(l1)/(xd*yd)))*100,1)`% of the landscape.
+
+By contrast, when increasing `ownership_var` to 0.8, overall inequity increases substantially:
+
+```{r, echo = FALSE, fig.height = 5, fig.width = 10}
+par(mfrow=c(1,2))
+par(mar = c(2, 2, 2, 2))
+xd = 100
+yd = 100
+l2 = owner_land_ssa(dim_x = xd, dim_y = yd, owners = 48, public_land = 0, land_var = 0.8)
+tl2 = table(l2)/(xd*yd)
+image(l2, xaxt = "n", yaxt = "n", main = "ownership_var = 0.8", col = topo.colors(max(land_2)))
+hist(tl2, xlim = c(0,0.16), col = "grey", breaks = 12, main = "")
+```
+
+In the latter case, user ownership varies between <0.001% and `r round((max(table(l2)/(xd*yd)))*100,1)`%.
+
+It should be noted that it is still entirely possible for the user to customise landscapes to whatever extent necessary, using custom `gmse_apply` loops [as described previously](SI4.pdf); however, the new `ownership_var` provides an option in GMSE to alter landscape ownership in a relatively straightforward way.
 

### 2020-05-23 notebook/ms/SI3.Rmd
Update the SI files

@@
 subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 3)'
-author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
-  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
-  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+author: |
+        | A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;, Jeroen Minderman&#xb9;, 
+        | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
+        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
 date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
@@
 
-```{r, echo = TRUE, eval = FALSE}
+```{r, echo = TRUE, eval = TRUE, results = 'hide'}
 sim_1 <- gmse(manager_budget = 10000, user_budget = 10000, res_death_K = 93870, 
@@
 ```{r, echo = FALSE, eval = TRUE}
-load("data/SI3-cull_sim.rda"); # Avoids long compiling time
-sim_1 <- sim_cull;
+# load("data/SI3-cull_sim.rda"); # Avoids long compiling time
+# sim_1 <- sim_cull;
 ```
@@
 
-```{r, echo = TRUE, fig.height = 6, eval = FALSE}
+```{r, echo = TRUE, fig.height = 6, eval = TRUE, results = 'hide'}
 sim_2 <- gmse(manager_budget = 10000, user_budget = 10000, res_death_K = 93870, 
@@
 ```{r, echo = FALSE, eval = TRUE}
-load("data/SI3-scare_sim.rda"); # Avoids long compiling time
-sim_2 <- sim_scare;
+# load("data/SI3-scare_sim.rda"); # Avoids long compiling time
+# sim_2 <- sim_scare;
 ```
@@
 Brief explanations of these non-default options appear in the [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf) (also [listed](https://confoobio.github.io/gmse/reference/gmse.html) on the [GMSE website](https://confoobio.github.io/gmse/)).
-Here we further explain some of the less obvious or less intuitive options available in GMSE v0.4.0.7, for which additional explanation beyond the GMSE documentation may be warranted for application to case studies such as the Taiga Bean Geese case study example demonstrated above.
+Here we further explain some of the less obvious or less intuitive options available in GMSE, for which additional explanation beyond the GMSE documentation may be warranted for application to case studies such as the Taiga Bean Geese case study example demonstrated above.
 Future versions of GMSE are expected to expand upon these options.
 
-**res_move_type**: Resources in default GMSE v0.4.0.7 sub-models can move on the landscape in one of three ways.
+**res_move_type**: Resources in default GMSE sub-models can move on the landscape in one of three ways.
 These different movement types are listed below, and are taken as arguments to `gmse` (e.g., `gmse(res_move_type = 2)`).
@@
 
-**res_death_type**: Resources in GMSE v0.4.0.7 can be removed from the model in one of three ways within the default resource sub-model, thereby simulating resource death.
+**res_birth_type**: Resources in GMSE can be added to the model in one of two ways within the default resource sub-model, thereby simulating resource birth.
+It is important to note that `res_birth_type` refers only to additions that occur within the resource sub-model; additions from user actions (e.g., the `help_offspring` option) occurs within the user sub-model.
+
+  0.
+*No explicit resource addition*.
+This option does not cause any baseline resource additions, but resource birth is still possible if `consume_repr > 0`.
+Under such conditions, landscape yield consumed by the resource will result in `floor(yield_consumed / consume_repr)` offspring produced.
+The amount of yield a resource consumes will be affected by both `times_feeding` (how many times the resource moves and feeds on a new landscape cell during a time step) and `res_consume` (the proportion of yield on a cell that is is consumed upon visit to the cell).
+Note that `consume_repr` can still be used when `res_birth_type > 0` (below), resulting in multiple sources of resource addition.
+  1. *Fixed number added per resource*. For this option, each resource produces `lambda` new resources in each time step.
+  2.
+*Poisson sampling*.
+For this option, each resource produces a number of offspring sampled independently from a Poisson distribution in which `lambda` is the rate parameter (i.e., each resource is expected to produce `lambda` offspring).
+
+**res_death_type**: Resources in GMSE can be removed from the model in one of four ways within the default resource sub-model, thereby simulating resource death.
 It is important to note that `res_death_type` refers only to removal that occurs within the resource sub-model; removal from user actions (e.g., death from shooting) occurs within the user sub-model.
 
+  0.
+*No explicit density independent or dependent removal*.
+This option does not cause removal to be *directly* caused in a density independent or dependent way.
+Nevertheless, removal may still be possible if the argument `consume_surv > 0` (its default value is 0).
+In this case, survival of a resource in a time step depends on that resource consuming at least `consume_surv` in yield on the landscape.
+In such cases, a natural carrying capacity will be generated as more resources result in higher total consumption.
+The amount yield that a resource consumes is further affected by both `times_feeding` (how many times the resource moves and feeds on a new landscape cell during a time step) and `res_consume` (the proportion of yield on a cell that is is consumed upon visit to the cell).
+Note that `consume_surv` is can still be used when `res_death_type > 0` (below), resulting in multiple potential sources of removal.
   1.
@@
 
-**obs_move_type**: Observation of agents (manager and users) in default GMSE v0.4.0.7 sub-models can occur in the same four ways as allowed in `res_move_type` (default `obs_move_type = 1`), with movement distance defined by the parameter `agent_move` (default 50).
+**obs_move_type**: Observation of agents (manager and users) in default GMSE sub-models can occur in the same four ways as allowed in `res_move_type` (default `obs_move_type = 1`), with movement distance defined by the parameter `agent_move` (default 50).
 Currently, the only effect of this movement is to allow for spatially autocorrelated sampling of resources by the manager in the observation model.

### 2020-05-23 notebook/ms/SI7.Rmd
Some proposed subheading additions in SI7
Update the SI files

--- 2020-05-21 notebook/ms/SI7.Rmd
@@
 title: "Default GMSE data structures"
-author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
-  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy
-  Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+author: |
+        | A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;, Jeroen Minderman&#xb9;, 
+        | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
+        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;'
 date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,

### 2020-05-23 notebook/ms/SI7.Rmd

@@
 If resources consume landscape yield, the value of yield can drop to a minimum value of zero.
+
+### 6.1. Default land distribution algorithm
 
@@
 In the case of the above landscape, the number of cells allocated to each user ranges from `r min(as.vector(table(land_2)))` to `r max(as.vector(table(land_2)))`.
+
+### 6.2. Varying land ownership: `ownership_var`
+
+In addition, [whereas in previous versions variable landscape ownership among users had to be customised using `gmse_apply`](SI4.pdf), GMSE v.
+0.6.0.0+ now includes an additional parameter `ownership_var`, which takes effect when `land_ownership = TRUE`.
+The value of this parameter is expected to be >=0 and <1.
+At its default value of 0, land ownership is equal among users as demonstrated in the code above.
+Values >0 increase the extent of variability of landownership among users; in technical terms, `ownership_var` controls the extent to which each split in the shortest-splitline algorithm is proportional to the number of users the landscape is to be divided into, with larger numbers shifting the split away from even distribution.
+In practical terms, this means that increasing values of `ownership_var` produce more unevenly distributed landscapes among users.
+For example, a value of `ownership_var = 0.1` produces a relatively limited about of inequity in landscape distribution among users.
+
+```{r, echo = FALSE, fig.height = 5, fig.width = 10}
+par(mfrow=c(1,2))
+par(mar = c(2, 2, 2, 2))
+xd = 100
+yd = 100
+l1 = owner_land_ssa(dim_x = xd, dim_y = yd, owners = 48, public_land = 0, land_var = 0.2)
+tl1 = table(l1)/(xd*yd)
+image(l1, xaxt = "n", yaxt = "n", main = "ownership_var = 0.2", col = topo.colors(max(land_2)))
+hist(tl1, xlim = c(0,0.16), col = "blue", breaks = 12, main = "", cex.axis = 1.25)
+```
+As can be seen from the frequency distribution, in this case, all users own between `r round((min(table(l1)/(xd*yd)))*100,1)`% and `r round((max(table(l1)/(xd*yd)))*100,1)`% of the landscape.
+
+By contrast, when increasing `ownership_var` to 0.8, overall inequity increases substantially:
+
+```{r, echo = FALSE, fig.height = 5, fig.width = 10}
+par(mfrow=c(1,2))
+par(mar = c(2, 2, 2, 2))
+xd = 100
+yd = 100
+l2 = owner_land_ssa(dim_x = xd, dim_y = yd, owners = 48, public_land = 0, land_var = 0.8)
+tl2 = table(l2)/(xd*yd)
+image(l2, xaxt = "n", yaxt = "n", main = "ownership_var = 0.8", col = topo.colors(max(land_2)))
+hist(tl2, xlim = c(0,0.16), col = "blue", breaks = 12, main = "", cex.axis = 1.25)
+```
+
+In the latter case, user ownership varies between <0.001% and `r round((max(table(l2)/(xd*yd)))*100,1)`%.
+
+It should be noted that it is still entirely possible for the user to customise landscapes to whatever extent necessary, using custom `gmse_apply` loops [as described previously](SI4.pdf); however, the new `ownership_var` provides an option in GMSE to alter landscape ownership in a relatively straightforward way.
+
+### 6.3. Public land
 

### 2020-05-23 notebook/ms/ms.Rmd
I'm not sure why the author list is now over-running the page on a PDF, but I've added a fix

@@
 title: 'GMSE: an R package for generalised management strategy evaluation'
-author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
-  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy
-  Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+author: |
+        | A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;, Jeroen Minderman&#xb9;, 
+        | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
+        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;'
 date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,

### 2020-05-23 notebook/ms/SI1.Rmd
Fix reference
Author name change, more to follow

@@
 subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 1)'
-author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
-  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
-  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
-date: "[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK
+author: |
+        | A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;, Jeroen Minderman&#xb9;, 
+        | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
+        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;'
 [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk"
@@
 This practical, metaheuristic approach is consistent with the objectives of management strategy evaluation [@Bunnefeld2011], and is well-suited for the use of genetic algorithms [@Hamblin2013].
-@Luke2013 describes the metaheuristic approach more generally (original emphasis retained):
+@Luke2009 describes the metaheuristic approach more generally (original emphasis retained):
 

### 2020-05-23 R/gmse_apply.R
I am changing the default genetic algorithm fitness function to take times_feeding into account, since this can lead to kind of unrealistic situations where times_feeding is very large and res_consume is very small

@@ -1902,9 +1902,13 @@ collect_agent_ini <- function(arg_list){
     if("tend_crop_yld" %in% arg_names){
         make_age_list[[26]] <- arg_list[["tend_crop_yld"]];
     }
-    make_age_list[[27]] <- NA;
+    make_age_list[[27]] <- arg_list[["GMSE"]][["times_feeding"]];
+    if("times_feeding" %in% arg_names){
+        make_age_list[[27]] <- arg_list[["times_feeding"]];
+    }
+    make_age_list[[28]] <- NA;
     if("LAND" %in% arg_names){
-        make_age_list[[27]] <- arg_list[["LAND"]];
+        make_age_list[[28]] <- arg_list[["LAND"]];
     }
     return(make_age_list);
 }

### 2020-05-23 R/initialise.R

@@ -106,6 +106,7 @@ make_resource <- function(model              = "IBM",
 #'@param res_consume The fraction of remaining biomass (e.g. crop production) that a resource consumes while occupying a landscape cell. The default value is 0.5, so if one resource occupies the cell, then landscape production is halved, if two resources occupy the cell, then landscape production drops to 0.25; if three, then production drops to 0.125, etc.
 #'@param consume_repr How much from a landscape does an individual resource need to produce one offspring (default 0)?
 #'@param tend_crop_yld The per landscape cell proportional increase in crop yield when stakeholders take one action to increase yield on their landscape. The default value is set to 0.5 (i.e., a 50 percent increase in yield on a cell).
+#'@param times_feeding Number of searches that resources are allowed per time step for feeding on the landscape. Resources will move between times feeding based on whatever `res_movement` and `res_move_type` parameters are specified.
 #'@param landscape The landscape on which some cells are owned. This needs to have been created with the make_landscape, or be a three dimensional array in which the third layer of the array corresponds to owned cells
 #'@return An initialised data frame of agents being modelled
 #'@examples
@@ -138,6 +139,7 @@ make_agents <- function(model              = "IBM",
                         res_consume        = 0,
                         consume_repr       = 0,
                         tend_crop_yld      = 0.2,
+                        times_feeding      = 1,
                         landscape          = NA
                         ){
     if(agent_number < 2 & length(type_counts) < 2){
@@ -190,7 +192,8 @@ make_agents <- function(model              = "IBM",
                                         perceive_feed, perceive_help, 
                                         perceive_tend, perceive_kill,
                                         manager_sense, lambda, res_consume,
-                                        consume_repr, tend_crop_yld, landscape);
+                                        consume_repr, tend_crop_yld, 
+                                        times_feeding ,landscape);
     }
     if( is.null(the_agents) ){
         stop("Invalid model selected (Must be 'IBM')");
@@ -218,6 +221,7 @@ agent_perceptions <- function(agents,
                               res_consume        = 0,
                               consume_repr       = 0,
                               tend_crop_yld      = 0.2,
+                              times_feeding      = 1,
                               landscape          = NA){
     
     if(is.array(agents) == FALSE){
@@ -232,7 +236,7 @@ agent_perceptions <- function(agents,
     }
     E_off <- lambda;
     if(consume_repr > 0){
[…]

### 2020-05-23 notebook/ms/SI1.Rmd
I have update SI1 for the new v0.6 GMSE

@@
         | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;'
-[2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk"
+date: '[2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
 output:
+  html_document: default
   pdf_document:
@@
 A genetic algorithm is called in the predefined GMSE manager and user models to simulate human decision making.
-As of GMSE version 0.4.0.3, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step.
+As of GMSE version 0.6, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step.
 Therefore, one run of the genetic algorithm occurs to simulate the manager's policy-setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1; e.g., see [SI6](SI6.pdf)), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users copy).
@@
 For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'iteration' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the [main text](ms.pdf)).
-Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.4.0.7 (future versions of GMSE might expand upon this framework, and we highlight some of these potential avenues for expansion).
+Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.6 (future versions of GMSE might expand upon this framework, and we highlight some of these potential avenues for expansion).
 We first explain the [key data structures used](#Key_data_structures_used), then provide an overview of how a population of strategies is [initialised](#initialised), and the subsequent processes of [crossover](#crossover), [mutation](#mutation), [cost constraint](#constraint), [fitness](#eval) evaluation, [tournament selection](#tournament), and [replacement](#replacement).
@@
 Positive values are currently only meaningful for `Manager_Actions`, where a value of 1 defines an action setting a uniform cost of users' direct actions on resources (i.e., costs where `Act = -2` for `User_1_Actions` and `User_2_Actions`).
-All other values for `Act` are meaningless in GMSE 0.4.0.3, but might be expanded upon in future versions to allow for modification of specific user costs enacted by managers (i.e., managers having different policies for different users) or other users (e.g., users increasing the costs of other users' actions due to conflict or cooperation).
+All other values for `Act` are meaningless in GMSE 0.6, but might be expanded upon in future versions to allow for modification of specific user costs enacted by managers (i.e., managers having different policies for different users) or other users (e.g., users increasing the costs of other users' actions due to conflict or cooperation).
 We will therefore focus only on rows 1-3 of `ACTION`.
 
-Columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in predefined GMSE v0.4.0.7 manager and user sub-models (i.e., only one type of resource is permitted).
+Columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in predefined GMSE v0.6 manager and user sub-models (i.e., only one type of resource is permitted).
 Future versions might allow for different resource types (e.g., `Type_1` might be used to designate species, and `Type_2` and `Type_3` could designate stage or sex).
@@
 All values in `COST` columns 1-7 are set to 100001, one higher than the highest possible `manager_budget` or `user_budget`, so neither managers nor users can affect resource types or utilities.
-Columns 8-13 are also set to 10001, except where actions are allowed.
+Columns 8-13 are also set to 100001, except where actions are allowed.
 Maximum values of 100000 are independent of any other parameter value specified in GMSE (e.g., landscape dimensions).
@@
 Here we explain how the fitnesses of candidate manager and user strategies in a `POPULATION` array (see above) are calculated.
-We emphasise that the fitness functions used in GMSE v0.4.0.7 are intended to be heuristic tools for identifying reasonable manager and user behaviours.
+We emphasise that the fitness functions used in GMSE v0.6 are intended to be heuristic tools for identifying reasonable manager and user behaviours.
 In practice, our fitness functions identify behaviours that are well-aligned with manager and user interests for harvesting or crop yield, but they are not intended to identify *optimal* decisions.
@@
 For allowed actions, the parameter `manager_sense` ($\sigma$) modulates predicted consequences for abundance by some factor; this is useful because not all actions attempted by users will be realised, and a value of $\sigma = 1$ tends to slightly overestimate how much the actions attempted by users will actually translate to a change in resource abundance.
-In practice, the default $\sigma = 0.9$ performs well.
+In practice, the default $\sigma = 0.9$ usually performs well.
+Managers also have access to an estimate of the number of offspring that resources produce ($E[offspring]$).
+For default settings in GMSE v0.6, $E[offspring] = \lambda$, where $\lambda$ is the GMSE argument `lambda` that defines the baseline population growth rate of resources.
+Given a non-default GMSE argument for `consume_repr` (where `consume_repr` is how much consumption of landscape yield is needed to increment offspring production by 1), $E[offspring]$ is further incremented by `( (times_feeding * res_consume) / consume_repr)` to account for reproduction enabled by feeding on the landscape (`res_consume` is the proportion of landscape yield that a resource consumes during feeding).
+
 Allowed actions are predicted by managers to have the following effects (again, we emphasise that whether or not these effects are realised will depend later on the user model, to which the manager -- by design -- does not have access):
@@
 - `scaring` is assumed to be nonlethal and therefore have no effect on resource number (resources are moved to a random cell on the landscape, as sampled from a uniform distribution such that movement to any given cell is equally probable).
-- `culling` decreases resource number by $\sigma$.
-- `castration` decreases resource number by $\sigma\lambda$, where $\lambda$ is the GMSE argument `lambda` that defines the baseline population growth rate of resources.
-- `feeding` increases resource number by $\sigma\lambda$.
+- `culling` decreases resource number by $\sigma (1 + E[offspring])$.
+- `castration` decreases resource number by $\sigma (E[offspring])$.
+- `feeding` increases resource number by $\sigma (E[offspring])$.
 - `help_offspring` increases resource number by $\sigma$.
@@
 
-The above effects cannot be altered directly in `gmse` or `gmse_apply` (though parameter values can of course be changed using `manager_sense` and `lambda` arguments), but future versions of GMSE might include different predicted effects to increase precision or allow for multiple resource types or different actions.
+The above effects cannot be altered directly in `gmse` or `gmse_apply` (though parameter values can of course be changed using `manager_sense` and arguments affecting $E[offspring]$), but future versions of GMSE might include different predicted effects to increase precision or allow for multiple resource types or different actions.
 The summation of $X_{i,j}$ for all actions defines the predicted change in resource abundance caused by the policy of an individual $i$, $\Delta A_{i}$.
@@
 Note that $F_{i}^{u}$ increases when $\Delta A_{i}$ and $\Delta L_{i}$ are of the same sign as $U^{u}_{res}$ and $U^{u}_{land}$, respectively.
-Further, in GMSE v0.4.0.7, only one term of the equation is nonzero.
+Further, in GMSE v0.6, only one term of the equation is nonzero.
 When `land_ownership = FALSE` (default, modelling users that harvest resources), $U^{u}_{res} = -1$ and $U^{u}_{land} = 0$, and when `land_ownership = TRUE`, $U^{u}_{res} = 0$ and $U^{u}_{land} = 100$ (modelling farmers trying to increase crop yield).
-Hence users only have a single objective of either decreasing resource abundance or increasing landscape output, though landscape output might be increased indirectly by decreasing resource abundance if `resource_consume` is greater than zero.
+Hence users only have a single objective of either decreasing resource abundance or increasing landscape output, though landscape output might be increased indirectly by decreasing resource abundance if `res_consume > 0`.
 
-User actions are predicted to affect resources in the following way:
+Unlike the [manager fitness function](#manfit), the predicted effects of user actions can be defined directly in `gmse` and `gmse_apply`.
+Hence, it is possible to specify how effective a user believes, e.g., culling or scaring to be with respect to their own utility.
+Arguments to do this include `perceive_scare`, `perceive_cull`, `perceive_cast`, `perceive_feed`, `perceive_help`, `perceive_tend` and `perceive_kill`.
+Each of these arguments define how much a user believes one action will affect either the user or the landscape.
+For example, if `perceive_scare = -1`, then a user will perceiving scaring as decreasing the number of resources feeding on their landscape by 1.
+A value of `perceive_feed = 0.5` would correspond to a user perceiving feeding to increase the number of resources by one half.
+Landscape effects (`perceive_tend` and `perceive_kill`) refer to effects of actions on the landscape, such that `perceive_tend = 0.2` would mean that a single action of tending crops (i.e., action taken on a single landscape cell) is perceived to increase crop yield by 0.2, while `perceive_kill = -1` would mean that a single action of killing crops will remove all yield on the landscape cell.
 
-- `scaring` decreases resource number by $1$.
-- `culling` decreases resource number by $1$.
-- `castration` decreases resource number by $\lambda$.
-- `feeding` increases resource number by $\lambda$.
-- `help_offspring` increases resource number by 1.
+By default, arguments affecting how users perceive the efficiacy of their own actions are set to `NA`.
+Default values are then calculated from other arguments specified in `gmse` or `gmse_apply`.
+Default user actions are predicted to affect resources in the following way (in the below $E[offspring]$ is calculated in the same way as in [the manager's fitness function](#manfit)):
+
+- `scaring` decreases resource number by $1$ times one minus the number of landscape cells that the user owns (i.e., `perceive_scare = -1 * (1 - pr_cells_owned)`).
+This accounts for the increasing probability that resource will be scared back to another landscape cell owned by the focal user.
+If a user owns no cells, then scaring is assumed to have no effect.
+- `culling` decreases resource number by $1 + E[offspring]$ (i.e., `perceive_cull = -1 * (1 + E_off)`).
+- `castration` decreases resource number by $E[offspring]$ (i.e., `perceive_cast = -1 * E_off`).
+- `feeding` increases resource number by $E[offspring]$ (i.e., `perceive_feed = E_off`).
+- `help_offspring` increases resource number by 1 (i.e., `perceive_help = 1`).
 
@@
 
-- `tend_crops` will increase landscape output by $\phi$.
-- `kill_crops` will decrease landscape output by 1 (since the output of a cell is 1 by default, this action removes all output on a landscape cell).
+- `tend_crops` will increase landscape output by $\phi$ (i.e., `perceive_tend = tend_crop_yld`).
+- `kill_crops` will decrease landscape output by 1 (since the output of a cell is 1 by default, this action removes all output on a landscape cell; i.e., `perceive_kill = -1`).
 

### 2020-05-24 notebook/ms/SI3.Rmd
Minor edit to the vignete and documentat

@@
         | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
-date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
-  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
+date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
 output:
@@
 |  manage_freq    |  1      | Frequency (in time steps) with which managers revise and enact policy                            |
-|  public_land    |  0      | Proportion of land that is public (un-owned by users) if land_ownership = TRUE                   |
+|  public_land    |  0      | Pr. of land that is public (un-owned by users) if land_ownership = TRUE                          |
+|  age_repr       |  1      | Age below which resources are incapable of reproducing                                           |
+|  action_thres   |  0      | Pr. Deviation of the estimated population from the manager target, above which policy is updated |
+|  budget_bonus   |  0      | Percentage of increase in budget a manager accrues by not updating policy in a time step         |
+|  consume_surv   |  0      | Amount of yield a resource need to consume in a timestep to survive                              |
+|  consume_repr   |  0      | Amount of yield a resource need to consume in a timestep to produce one offspring                |
+|  times_feeding  |  1      | Number of searches that resources are allowed per time step for feeding on the landscape         |
 
@@
 Further, we consider the case of a region in which farmland makes up 60% of all land, with 40% of land being 'public' (`public_land = 0.4`; which might be interpreted as any land in which stakeholders are not, or cannot be, invested in goose presence), and divide the farmland amongst 80 individual farmers (`stakeholders = 80`; `land_ownership = TRUE`).
-Landscape size is set to default 100 by 100 cells, so each farmer effectively owns 75 cells, which might be interpreted as hectares of land (for instructions on how to more precisely control landscape ownership, see the [advanced GMSE options](SI4.pdf) using `gmse_apply`).
+Landscape size is set to default 100 by 100 cells, so each farmer owns about 75 cells, which might be interpreted as hectares of land (for instructions on how to more precisely control landscape ownership, see the [advanced GMSE options](SI4.pdf) using `gmse_apply`).
 Because we need both density-dependent (`res_death_K = 93870`) and density-independent (`remove_pr = 0.122`) sources of mortality, we set `res_death_type = 3`.
@@
 **res_birth_type**: Resources in GMSE can be added to the model in one of two ways within the default resource sub-model, thereby simulating resource birth.
-It is important to note that `res_birth_type` refers only to additions that occur within the resource sub-model; additions from user actions (e.g., the `help_offspring` option) occurs within the user sub-model.
+It is important to note that `res_birth_type` refers only to additions that occur within the resource sub-model; additions from user actions (e.g., the `help_offspring` option) occur within the user sub-model.
 
@@
 *No explicit resource addition*.
-This option does not cause any baseline resource additions, but resource birth is still possible if `consume_repr > 0`.
+This option causes no baseline resource additions, but resource birth is still possible if `consume_repr > 0`.
 Under such conditions, landscape yield consumed by the resource will result in `floor(yield_consumed / consume_repr)` offspring produced.
-The amount of yield a resource consumes will be affected by both `times_feeding` (how many times the resource moves and feeds on a new landscape cell during a time step) and `res_consume` (the proportion of yield on a cell that is is consumed upon visit to the cell).
+The amount of yield a resource consumes will be affected by both `times_feeding` (how many times the resource moves and feeds on a new landscape cell during a time step) and `res_consume` (the proportion of yield on a cell that is consumed upon visit to the cell).
 Note that `consume_repr` can still be used when `res_birth_type > 0` (below), resulting in multiple sources of resource addition.
@@
 *Poisson sampling*.
-For this option, each resource produces a number of offspring sampled independently from a Poisson distribution in which `lambda` is the rate parameter (i.e., each resource is expected to produce `lambda` offspring).
+For this option, each resource produces a number of offspring as determined by sampling from a Poisson distribution in which `lambda` is the rate parameter (i.e., each resource is expected to produce `lambda` offspring).
 
@@
 Note that `consume_surv` is can still be used when `res_death_type > 0` (below), resulting in multiple potential sources of removal.
+Given `res_death_type = 0`, **user actions have no effect** on resource removal (to ensure that user actions do have an effect, set `res_death_type = 1` and `removal_pr = 0`; see below).
   1.
@@
 Given this option, there is no density-dependent removal of resources within the resource sub-model.
-Resources are simply removed with a probability of `removal_pr` each time the sub-model is run.
+Resources are simply removed with a fixed probability of `removal_pr` (its default value is 0) each time the sub-model is run.
 This option should be used with great care because in the absence of density-dependence, it is possible for the population of resources to increase without limit and thereby greatly slow down simulation times.
@@
 
-**observe_type**: Observation of resources in default GMSE v0.4.0.7 sub-models can occur in one of four ways.
+**observe_type**: Observation of resources in default GMSE sub-models can occur in one of four ways.
 Each observation type mimics some process of resource observation, with potential error affected by observation intensity.

### 2020-05-24 R/gmse.R

@@ -64,7 +64,7 @@
 #'@param group_think If TRUE, all users will have identical actions; the genetic algorithm will find actions for one user and copy them for all users. This is a useful option if a lot of users are required but variation among user decisions can be ignored. The default value is FALSE.
 #'@param age_repr The age below which resources are incapable of reproducing. The default value is 1.
 #'@param usr_budget_rng This specifies a range around the value of `user_budget`, such that the expected value of each user's budget will be `user_budget`, with a uniform distribution plus or minus `usr_budget_rng`. Note that the minimum `usr_budget_rng` allowed is 1 regardless of the range set, and the maximum is always 100000. The default value for this argument is 0.
-#'@param action_thres A value for the deviation of the estimated population from the manager target, above which manager will not update the policy. Recommended values are between 0 and 1, with the default value being 0.
+#'@param action_thres A value for the deviation of the estimated population from the manager target, below which manager will not update the policy. Recommended values are between 0 and 1, with the default value being 0.
 #'@param budget_bonus A percentage of the initial budget manager will receive if policy was not updated last time step. Corresponds to the time, energy and money saved by waiting for a better time to update the policy. Budget bonuses are cumulative, so many time steps of not updating policy can cause a compounding increase in the budget bonus. The default value is 0.
 #'@param consume_surv This value defines the amount of yield on a landscape that an individual resource need to consume in a timestep to survive. The default value is 0 (i.e., no consumption is required for survival).
 #'@param consume_repr This value defines the amount of yield on a landscape that an individual resource need to produce one offspring. Resources will produce as many offspring as is possible given their yield in take; e.g., if a resource consumes between three and four times the amount of yield required for reproduction, then they will produce three offspring. The default value is 0 (i.e., no consumption is required for reproduction). 

### 2020-05-24 notebook/ms/SI5.Rmd
Small edits to SI5

@@
 
-Future versions of GMSE will continue to be open-source and developed to avoid unecessary dependencies (GMSE v.0.4.0.3 requires only base R).
+Future versions of GMSE will continue to be open-source and developed to avoid unecessary dependencies (GMSE requires only base R, plus three shiny packages for running `gmse_gui`).
 Key goals including (1) providing highly general and useful default `resource`, `observation`, `manager`, and `user` sub-models for a variety of MSE modelling tasks, (2) keeping these sub-models highly modular so that they can be developed in isolation given standardised data structures, and (3) allowing these modular sub-models to be integrated with custom defined sub-models as flexibly as possible using `gmse_apply`.

### 2020-05-24 notebook/ms/SI5.Rmd
Get the authors listed in a way that won't run over the PDF

@@
 title: "Fisheries example integrating FLR"
-author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
-  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy
-  Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
-date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
-  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
+author: |
+        | A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;, Jeroen Minderman&#xb9;, 
+        | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
+        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
 output:

### 2020-05-24 notebook/ms/SI7.Rmd
Get the authors listed in a way that won't run over the PDF

@@
         | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
-        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;'
-date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
-  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
+        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
 output:

### 2020-05-24 notebook/ms/SI8.Rmd
Get the authors listed in a way that won't run over the PDF

@@
 subtitle: 'GMSE: an R package for generalised management strategy evaluation (New function vignette)'
-author: Adrian Bach&#xb9; &#xb2;
+author: | 
+        | Adrian Bach&#xb9; &#xb2;
 date: "[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK [2] adrian.bach@stir.ac.uk"

### 2020-05-26 notebook/cran_vignettes/SI2.Rmd
Recompile the website

--- 2018-05-10 notebook/ms/SI2.Rmd
@@
 subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 2)'
-author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
-  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
-  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
-date: "[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK
-[2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk"
+author: |
+        | A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;, Jeroen Minderman&#xb9;, 
+        | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
+        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
 output:
@@
 
-It is possible to, e.g., specify a simple resource and observation model, but then take advantage of the genetic algorithm to predict policy decisions and user actions (see [SI5](SI5.pdf) for a fisheries example).
+It is possible to, e.g., specify a simple resource and observation model, but then take advantage of the genetic algorithm to predict policy decisions and user actions (see [Fisheries example integrating FLR](SI5.pdf) for a fisheries example).
 This can be done by using the default GMSE manager and user functions (written below explicitly, though this is not necessary).
@@
 For most simulation purposes, this will not introduce any undesirable effect on simulation results, but it should be noted and understood when developing models.
+
+Special considerations for looping with custom sub-models
+================================================================================
+
+There are some special considerations that need to be made when using custom sub-models and the `old_list` argument within a loop as above. These considerations boil down to two key points.
+
+1. Custom sub-models *always* need to be read in explicitly as an argument in `gmse_apply` (i.e., they will not be remembered by `old_list`).
+2.
+Custom sub-model arguments also *always* need to be updated *outside* of `gmse_apply` before output is used as an argument in `old_list` (i.e., `gmse_apply` cannot know what custom function argument needs to be updated, so this needs to be done manually).
+
+An example below illustrates the above points more clearly.
+Assume that the custom resource sub-model defined above needs to be integrated with the default observation, manager, and user sub-models using `gmse_apply`.
+
+```{r, eval = TRUE}
+alt_res <- function(X, K = 2000, rate = 1){
+    X_1 <- X + rate*X*(1 - X/K);
+    return(X_1);
+}
+```
+
+The sub-model can be integrated once using `gmse_apply` as demonstrated above, but in the full `gmse_apply` output, the argument `X` will not change from its initial value (because sub-model functions can take any number of arbitrary arguments, `gmse_apply` has no way of knowing that `X` is meant to be the resource number and not some other parameter).
+
+```{r, eval = TRUE}
+sim_4 <- gmse_apply(res_mod = alt_res, X = 1000, get_res = "Full");
+print(sim_4$basic_output);
+```
+
+Note that in the above output, the resource abundance has increased and is now `sim_4$basic_output$resource_results`. But if we look at `sim_4$X`, the value is still `r sim_4$X`.
+
+```{r, eval = TRUE}
+print(sim_4$X);
+```
+
+To loop through multiple time steps with the custom function `alt_res`, it is therefore necessary to update `sim4$X` with the updated value from either `sim4$resource_vector` or `sim4$basic_output$resource_results` (the two values should be identical).
+The loop below shows a simple example.
+
+```{r, eval = TRUE}
+init_abun   <- 1000;
+sim_old     <- gmse_apply(get_res = "Full", res_mod = alt_res, X = init_abun);
+for(time_step in 1:20){
+    sim_new               <- gmse_apply(res_mod = alt_res, get_res = "Full", 
+                                        old_list = sim_old);
+    sim_old               <- sim_new;
+    sim_old$X             <- sim_new$resource_vector;
+}
+```
+
+Note again that the custom sub-model is read into to `gmse_apply` as an argument within the loop (`res_mod = alt_res`), and the output of `sim_new` is used to update the custom argument `X` in `alt_res` (`sim_old$X <- sim_new$resource_vector`).
+The population quickly increases to near carrying capacity, which can be summarised by using the same table structure explained above.
+
+```{r, eval = TRUE}
+init_abun   <- 1000;
+sim_old     <- gmse_apply(get_res = "Full", res_mod = alt_res, X = init_abun);
+sim_sum_4   <- matrix(data = NA, nrow = 5, ncol = 5);
+for(time_step in 1:5){
+    sim_new                 <- gmse_apply(res_mod = alt_res, get_res = "Full", 
+                                          old_list = sim_old);
+    sim_sum_4[time_step, 1] <- time_step;
+    sim_sum_4[time_step, 2] <- sim_new$basic_output$resource_results[1];
+    sim_sum_4[time_step, 3] <- sim_new$basic_output$observation_results[1];
+    sim_sum_4[time_step, 4] <- sim_new$basic_output$manager_results[3];
+    sim_sum_4[time_step, 5] <- sum(sim_new$basic_output$user_results[,3]);
+    sim_old                 <- sim_new;
+    sim_old$X               <- sim_new$resource_vector;
+}
+colnames(sim_sum_4) <- c("Time", "Pop_size", "Pop_est", "Cull_cost", 
+                         "Cull_count");
+print(sim_sum_4);
+```
+
+This is the recommended way to loop custom functions in `gmse_apply`.
+Note that elements of `old_list` will over-ride custom arguments to `gmse_apply` so **specifying custom arguments that are already present in `old_list` will not work**.

### 2020-05-26 notebook/cran_vignettes/SI4.Rmd
Recompile the website

--- 2020-05-23 notebook/ms/SI4.Rmd
@@
 subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 4)'
-author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
-  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
-  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
-date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
-  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
+author: |
+        | A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;, Jeroen Minderman&#xb9;, 
+        | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
+        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
 output:
-  html_document: default
   pdf_document:
@@
 Here we demonstrate how simulations in GMSE can be more fine-tuned to specific empirical situations through the use of `gmse_apply`.
-To do this, we use the same scenario described in [SI3](SI3.pdf); we first recreate the basic scenario run in `gmse` using `gmse_apply`, and then build in additional modelling details including (1) [custom placement of user land](#land), (2) [parameterisation of individual user budgets](#budgets), and (3) [density-dependent movement of resources](#movement).
+To do this, we use the same scenario described in [Example case study in GMSE](SI3.pdf); we first recreate the basic scenario run in `gmse` using `gmse_apply`, and then build in additional modelling details including (1) [custom placement of user land](#land), (2) [parameterisation of individual user budgets](#budgets), and (3) [density-dependent movement of resources](#movement).
 We emphasise that these simulations are provided only to demonstrate the use of GMSE, and specifically to show the flexibility of the `gmse_apply` function, not to accurately recreate the dynamics of a specific system or make management recommendations.
@@
 
-We now recreate the simulations in [SI3](SI3.pdf), which were run using the `gmse` function, in `gmse_apply`.
+We now recreate the simulations in [Example case study in GMSE](SI3.pdf), which were run using the `gmse` function, in `gmse_apply`.
 Doing so requires us to first initialise simulations using one call of `gmse_apply`, then loop through multiple time steps that again call `gmse_apply`; results of interest are recorded in a data frame (`sim_sum_1`).
-Following the protocol introduced in [SI2](SI2.pdf), we can call the initialising simulation `sim_old`, and use the code below to read in the relevant parameter values.
+Following the protocol introduced in [Use of the gmse_apply function](SI2.pdf), we can call the initialising simulation `sim_old`, and use the code below to read in the relevant parameter values.
 
@@
 
-We can then loop over 30 time steps to recreate the simulations from [SI3](SI3.pdf).
+We can then loop over 30 time steps to recreate the simulations from [Example case study in GMSE](SI3.pdf).
 In these simulations, we are specifically interested in the resource and observation outputs, as well as the manager policy and user actions for culling, which we record below in the data frame `sim_sum_1`.
@@
 It is important to remember that this example of `gmse_apply` is using the default resource, observation, manager, and user sub-models.
-Custom sub-models could produce different outputs in `sim_new` (see [SI2](SI2.pdf) for examples).
+Custom sub-models could produce different outputs in `sim_new` (see [Use of the gmse_apply function](SI2.pdf) for examples).
 For default sub-models, there are some list elements that might be especially useful.
 These elements can potentially be edited *within the above loop* to dynamically adjust simulations.
-For more explanation of built-in GMSE data arrays, see [SI7](SI7.pdf).
+For more explanation of built-in GMSE data arrays, see [Default GMSE data structures](SI7.pdf).
 
-- [`sim_new$resource_array`](SI7.pdf#resources): A table holding all information on resources.
+- [`sim_new$resource_array`](SI7.pdf): A table holding all information on resources.
 Rows correspond to discrete resources, and columns correspond to resource properties: (1) ID, (2-4) types (not currently in use), (5) x-location, (6) y-location, (7) movement parameter, (8) time, (9) density independent mortality parameter (`remove_pr`), (10) reproduction parameter (`lambda`), (11) offspring number, (12) age, (13-14) observation columns, (15) consumption rate (`res_consume`), (16-20) recorded experiences of user actions (e.g., was the resource culled or scared?), (21) how much yield has the resource consumed, and (22) how many times the resource can consume yield in one time step.
-- [`sim_new$AGENTS`](SI7.pdf#agents): A table holding basic information on agents (manager and users).
+- [`sim_new$AGENTS`](SI7.pdf): A table holding basic information on agents (manager and users).
 Rows correspond to a unique agent, and columns correspond to agent properties: (1) ID, (2) type (0 for the manager, 1 for users), (3-4) additional type options not currently in use, (5-6), x and y locations (usually ignored), (7) movement parameter (usually ignored), (8) time, (9) agent's viewing ability in cells (`agent_view`), (10) error parameter, (11-12) values for holding marks and tallies of resources, (13-15) values for holding observations, (16) yield from landscape cells, (17) baseline budget (`manager_budget` and `user_budget`), (18-24) agent's perception of the efficacy of scaring, culling, castrating, feeding, helping, tending crops, and killing crops, (25-26) increments to budget, (27) unused.
-- `sim_new$observation_vector`: Estimate of total resource number from the observation model ([`observation_array`](SI7.pdf#observation) also holds this information in a different way depending on `observe_type`)
-- [`sim_new$LAND`](SI7.pdf#land): The landscape on which interactions occur, which is stored as a 3D array with `land_dim_1` rows, `land_dim_2` columns, and 3 layers.
+- `sim_new$observation_vector`: Estimate of total resource number from the observation model ([`observation_array`](SI7.pdf) also holds this information in a different way depending on `observe_type`)
+- [`sim_new$LAND`](SI7.pdf): The landscape on which interactions occur, which is stored as a 3D array with `land_dim_1` rows, `land_dim_2` columns, and 3 layers.
 Layer 1 (`sim_new$LAND[,,1]`) is not currently used in default sub-models, but could be used to store values that affect resources and agents.
@@
 - `sim_new$manage_vector`: The cost of each action as set by the manager.
-For even more fine-tuning, individual costs for the actions of each agent can be set for each user in [`sim_new$manager_array`](SI7.pdf#cost).
-- `sim_new$user_vector`: The total number of actions performed by each user. A more detailed breakdown of actions by individual users is held in [`sim_new$user_array`](SI7.pdf#action).
+For even more fine-tuning, individual costs for the actions of each agent can be set for each user in [`sim_new$manager_array`](SI7.pdf).
+- `sim_new$user_vector`: The total number of actions performed by each user. A more detailed breakdown of actions by individual users is held in [`sim_new$user_array`](SI7.pdf).
 
@@
 
-```{r, fig.cap = "Default position of land ownership by farmers."}
+```{r, fig.cap = "Default position of land ownership by farmers.", fig.height = 6, fig.width = 6}
 image(x = sim_old$LAND[,,3], col = topo.colors(9), xaxt = "n", yaxt = "n");
@@
 We will use this custom function to modify `sim_old$resource_array` prior to running `gmse_apply`, thereby modelling a custom-built process affecting resource distribution that is integrated into GMSE.
-
-\clearpage
 
@@
 The goal was to show how software users can work with these existing sub-models and data structures to customise GMSE simulations.
-Readers seeking even greater flexibility (e.g., replacing an entire built-in sub-model with a custom sub-model) should refer to [SI2](SI2.pdf) that introduces `gmse_apply` more generally.
+Readers seeking even greater flexibility (e.g., replacing an entire built-in sub-model with a custom sub-model) should refer to [Use of the gmse_apply function](SI2.pdf) that introduces `gmse_apply` more generally.
 Future versions of GMSE are likely to expand on the built-in options available for simulation; requests for such expansions, or contributions, can be submitted to [GitHub](https://github.com/bradduthie/gmse/issues).

### 2020-05-26 notebook/cran_vignettes/SI5.Rmd
Recompile the website

--- 2020-05-24 notebook/ms/SI5.Rmd
@@
 
-```{r, fig.height = 5.2, fig.cap = "Output of the FLR plot function for an example Ricker model of stock recruitment on North Sea Herring data."}
+```{r, fig.height = 5.2, fig.cap = "Output of the FLR plot function for an example Ricker model of stock recruitment on North Sea Herring data.", message = FALSE}
 plot(newFL, cex = 0.7);

### 2020-05-26 notebook/ms/SI5.Rmd
Recompile the website

@@
 As part of the [ConFooBio](https://sti-cs.org/confoobio/) project, a central focus of GMSE is on simulating the management of animal populations of conservation interest, with a particular emphasis on understanding conservation conflict; further development of GMSE is expected to continue with this as a priority, further building upon the decision-making algorithms of managers and users to better understand how conflict arises and can be managed and mitigated.
-Hence, GMSE is not intended as a substitute for packages such as [FLR](http://www.flr-project.org/), but the integration of these packages with GMSE could make use of GSME's current and future simulation capabilities, and particularly the [genetic algorithm](SI1.pdf).
-Such integration might be possible using the `gmse_apply` function, which allows for [custom defined sub-models](SI2.pdf) to be used within the GMSE framework, and with default GMSE sub-models.
+Hence, GMSE is not intended as a substitute for packages such as [FLR](http://www.flr-project.org/), but the integration of these packages with GMSE could make use of GSME's current and future simulation capabilities, and particularly the [genetic algorithm](SI1.html).
+Such integration might be possible using the `gmse_apply` function, which allows for [custom defined sub-models](SI2.html) to be used within the GMSE framework, and with default GMSE sub-models.
 Hence, GMSE might be especially useful for modelling the management of fisheries under conditions of increasing competing stakeholder demands and conflicts.
@@
 This population model will use sample data and one of the many available stock-recruitment models available in FLR, and a custom function will be written to return a single value for stock recruitment.
-Currently, `gmse_apply` requires that sub-models return subfunction results either as scalar values or data frames that are structured [in the same way](SI7.pdf) as GMSE sub-models.
+Currently, `gmse_apply` requires that sub-models return subfunction results either as scalar values or data frames that are structured [in the same way](SI7.html) as GMSE sub-models.
 But interpretation of scalar values is left up to the user (e.g., population model results could be interpreted as abundance or biomass; manager policy could be interpreted as cost of harvesting or as total allowable catch).
@@
 Contributions in line with these goals, and suggestions for new features, can be made on [GitHub](https://github.com/bradduthie/gmse).
+
+References
+================================================================================

### 2020-05-27 notebook/ms/SI2.Rmd

--- 2018-05-10 notebook/ms/SI2.Rmd
@@
 subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 2)'
-author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
-  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
-  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
-date: "[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK
-[2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk"
+author: |
+        | A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;, Jeroen Minderman&#xb9;, 
+        | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
+        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;'
+date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
 output:

### 2020-05-27 notebook/ms/SI4.Rmd

@@
 subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 4)'
-author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
-  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
-  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
-date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
-  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
+author: |
+        | A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;, Jeroen Minderman&#xb9;, 
+        | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
+        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
 output:

### 2020-05-27 notebook/website_vignettes/SI1.Rmd

--- 2020-05-23 notebook/ms/SI1.Rmd
@@
         | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
-        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;'
+        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
 date: '[2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
@@
 As of GMSE version 0.6, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step.
-Therefore, one run of the genetic algorithm occurs to simulate the manager's policy-setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1; e.g., see [SI6](SI6.pdf)), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users copy).
+Therefore, one run of the genetic algorithm occurs to simulate the manager's policy-setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1; e.g., see [Management frequency and extinction risk](SI6.html)), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users copy).
 Each run of the genetic algorithm mimics the evolution by natural selection of a population of potential manager or user strategies over multiple iterations, with the highest fitness strategy in the terminal iteration being selected as the one that the manager or user decides to implement.
@@
 
-The focal data structure used for tracking manager and user decisions is a three dimensional array, which we will call `ACTION` (also returned as `user_array` by `gmse_apply`; see [SI7](SI7.pdf)).
+The focal data structure used for tracking manager and user decisions is a three dimensional array, which we will call `ACTION` (also returned as `user_array` by `gmse_apply`; see [Default GMSE data structure](SI7.html)).
 Rows of `ACTION` correspond to the entities affected by actions (resources, landscape properties, or potentially other agents), and columns correspond either to properties of the affected entities, or to the actions potentially allocated to them.
@@
 
-Constraints on the values that elements in the `ACTION` array can take are defined by a `COST` array (also returned as `manager_array` by `gmse_apply`; see [SI7](SI7.pdf)) of dimensions identical to `ACTION`.
+Constraints on the values that elements in the `ACTION` array can take are defined by a `COST` array (also returned as `manager_array` by `gmse_apply`; see [Default GMSE data structures](SI7.html)) of dimensions identical to `ACTION`.
 Elements of `COST` define how many units from the `manager_budget` or `user_budget` are needed to perform a single action; a `minimum_cost` for actions is defined as an argument in GMSE (10 by default).
@@
 Individual fitness as calculated for managers ($F_{i}^{m}$) is affected by a manager's utility for resources and the projected change in resource abundance caused by the individual's policy (i.e., the contents of their `POPULATION` layer, specifically row 3; here again we use 'individual' to refer to one of `ga_popsize` discrete strategies in `POPULATION`, which may be selected and reproduce within the genetic algorithm).
-Manager utility for a resource ($U^{m}_{res}$) is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model (see "[Key data structures used](#Key_data_structures_used)" above, and [SI7](SI7.pdf#action) for more information).
+Manager utility for a resource ($U^{m}_{res}$) is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model (see "[Key data structures used](#Key_data_structures_used)" above, and [Default GMSE data structures](SI7.html#action) for more information).
 Manager utility can therefore change in each GMSE time step as estimated resource abundance changes; when the estimated resource abundance is greater than `manage_target`, $U^{m}_{res}$ is negative, and when the estimated resource abundance is less than `manage_target`, $U^{m}_{res}$ is positive.

### 2020-05-29 notebook/cran_vignettes/SI6.Rmd

--- 2019-11-13 notebook/ms/SI6.Rmd
@@
 title: "Management frequency and extinction risk"
-author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
-  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy
-  Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
-date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
-  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
+author: |
+        | A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;, Jeroen Minderman&#xb9;, 
+        | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
+        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
 output:
@@
 In GMSE, some of the most important traits for resources include types, ages, demographic parameter values, locations, etc., and for agents (manager and users), traits include different types, utilities, budgets, etc.
-The traits that resources and managers have can potentially affect their interactions, and default GMSE sub-models take advantage of this by simulating interactions explicitly on a landscape (see [SI7](SI7.pdf) for an introduction to GMSE default data structures).
+The traits that resources and managers have can potentially affect their interactions, and default GMSE sub-models take advantage of this by simulating interactions explicitly on a landscape (see [Default GMSE data structures](SI7.pdf) for an introduction to GMSE default data structures).
 
@@
 Future expansion on the complexity of individual-based default sub-models of GMSE will further increase the realism of targeted case studies.
+
+References
+================================================================================

### 2020-05-29 notebook/website_vignettes/SI3.Rmd

--- 2020-05-24 notebook/ms/SI3.Rmd
@@
 
-To demonstrate how GMSE can be used to understand conflict in social-ecological systems, we build upon the [example of resource management](https://www.biorxiv.org/content/biorxiv/early/2017/11/17/221432.full.pdf#page=5) in the main text.
+To demonstrate how GMSE can be used to understand conflict in social-ecological systems, we build upon the [example of resource management](GMSE.html#an-example-of-resource-management) in the main text.
 We consider a protected population of waterfowl that exploits and damages agricultural land and is therefore a source of conservation conflict between those that seek the conservation of waterfowl and those that are concerned with the loss of agriculture [e.g., @Fox2017a; @Mason2017; @Tulloch2017; @Cusack2018].
@@
 Novice R users may prefer to run all of the simulations below using the browser-based GMSE GUI by calling `gmse_gui()` from the R command line.
-Alternatively, experienced R users may prefer to simulate by [looping time steps](SI2.pdf#page=5) through `gmse_apply`, which allows more flexibility for incorporating custom sub-models and dynamically adjusting parameter values.
+Alternatively, experienced R users may prefer to simulate by [looping time steps](SI2.html#running-gmse-simulations-by-looping-gmse_apply) through `gmse_apply`, which allows more flexibility for incorporating custom sub-models and dynamically adjusting parameter values.
 
 Simulations using the default GMSE sub-models described above are run using the `gmse` function, which offers a range of options for setting parameter values (see Table 1 for some select examples).
-Output of `gmse` is an exhaustive list that includes all resources and observations, all stakeholder decisions and actions, and all landscape properties in each time step of the simulation (see [SI7](SI7.pdf) for a description of key data structures).
+Output of `gmse` is an exhaustive list that includes all resources and observations, all stakeholder decisions and actions, and all landscape properties in each time step of the simulation (see [Default GMSE data structures](SI7.html) for a description of key data structures).
 
@@
 Where GMSE parameter values are not available, we use reasonable values or GMSE defaults.
-To make model inferences for real case studies, we strongly recommend [replicating simulations](SI6.pdf) and simulating across a range of parameter values when empirical estimates are unavailable, as social-ecological dynamics might be sensitive to these unknown values.
+To make model inferences for real case studies, we strongly recommend [replicating simulations](SI6.html) and simulating across a range of parameter values when empirical estimates are unavailable, as social-ecological dynamics might be sensitive to these unknown values.
 
@@
 Further, we consider the case of a region in which farmland makes up 60% of all land, with 40% of land being 'public' (`public_land = 0.4`; which might be interpreted as any land in which stakeholders are not, or cannot be, invested in goose presence), and divide the farmland amongst 80 individual farmers (`stakeholders = 80`; `land_ownership = TRUE`).
-Landscape size is set to default 100 by 100 cells, so each farmer owns about 75 cells, which might be interpreted as hectares of land (for instructions on how to more precisely control landscape ownership, see the [advanced GMSE options](SI4.pdf) using `gmse_apply`).
+Landscape size is set to default 100 by 100 cells, so each farmer owns about 75 cells, which might be interpreted as hectares of land (for instructions on how to more precisely control landscape ownership, see the [advanced GMSE options](SI4.html) using `gmse_apply`).
 Because we need both density-dependent (`res_death_K = 93870`) and density-independent (`remove_pr = 0.122`) sources of mortality, we set `res_death_type = 3`.
@@
 [Requests for new features](https://github.com/bradduthie/gmse/issues) and contributions to GMSE can be made through [GitHub](https://github.com/bradduthie/gmse).
-Additionally, where entirely different types of sub-models are required, the `gmse_apply` function [can be used](https://bradduthie.github.io/gmse_SI2.pdf) to more flexibly simulate social-ecological systems.
-In [SI4](SI4.pdf), we show an example of this using the same Taiga Bean Geese case study that we did here.
+Additionally, where entirely different types of sub-models are required, the `gmse_apply` function [can be used](SI2.html) to more flexibly simulate social-ecological systems.
+In [Advanced case study options](SI4.html), we show an example of this using the same Taiga Bean Geese case study that we did here.
+
+References
+================================================================================

### 2020-05-29 notebook/website_vignettes/SI7.Rmd

--- 2020-05-24 notebook/ms/SI7.Rmd
@@
 The default sub-models of GMSE (`resource`, `observation`, `manager`, and `user`) use a small number of default data structures to hold the information needed in simulations.
-While these default sub-models do not necessarily need to be used in every run in GMSE (see [use of](SI2.pdf) `gmse_apply`), they will be used in any run of the `gmse` function, and in any call of the `gmse_apply` function that does not run with entirely custom sub-models.
-Simulation and model inference do not require an understanding of the default data structures, but such an understanding can be especially useful when running `gmse_apply` if there is a need to extract uncommonly used information, change key simulated values (e.g., landscape properties, agent budgets, or resource movement rules, as in [SI4](SI4.pdf)), or build custom individual-based sub-models.
+While these default sub-models do not necessarily need to be used in every run in GMSE (see [use of](SI2.html) `gmse_apply`), they will be used in any run of the `gmse` function, and in any call of the `gmse_apply` function that does not run with entirely custom sub-models.
+Simulation and model inference do not require an understanding of the default data structures, but such an understanding can be especially useful when running `gmse_apply` if there is a need to extract uncommonly used information, change key simulated values (e.g., landscape properties, agent budgets, or resource movement rules, as in [Advanced case study options](SI4.html)), or build custom individual-based sub-models.
 Here we provide a brief explanation of the following key data structures (each name below is listed as it is named in the output `gmse_apply` when `get_res = "Full"`).
@@
 
-Note that these are not the only data structures used in GMSE, but they are the only ones that can be easily modified in GMSE v0.4.0.7 (see, e.g., [SI4](SI4.pdf)), so they are the ones that we focus on here.
+Note that these are not the only data structures used in GMSE, but they are the only ones that can be easily modified in GMSE (see, e.g., [Advanced case study options](SI4.html)), so they are the ones that we focus on here.
 Additionally, any custom subfunction that returns an array rather than a single value should adhere to the same structure as these defaults if any default GMSE functions are to be used in `gmse_apply`.
@@
 
-In the [genetic algorithm](SI1.pdf), values in elements of a `user_array` layer are potentially modified according to each agent's objective, as constrained by costs in `manager_array`.
+In the [genetic algorithm](SI1.html), values in elements of a `user_array` layer are potentially modified according to each agent's objective, as constrained by costs in `manager_array`.
 
@@
 
-In [SI4](SI4.pdf), we show how landscape cell values can be manipulated to customise the placement of land ownership.
-By default, GMSE v0.6.0.0+ places the ownership of landscape cells using a shortest-splitline algorithm (see the lab notebook for [a detailed explanation](https://confoobio.github.io/gmse/notebook/gmse_notes.html#u-10-MAY-2020)).
+In [Advanced case study options](SI4.html), we show how landscape cell values can be manipulated to customise the placement of land ownership.
+By default, GMSE v0.6+ places the ownership of landscape cells using a shortest-splitline algorithm (see the lab notebook for [a detailed explanation](https://confoobio.github.io/gmse/notebook/gmse_notes.html#u-10-MAY-2020)).
 The algorithm splits the landscape evenly into smaller and smaller rectangles until a sufficient number of rectangles are produced for all users to own a similar amount of land (note that how uniformly land is partitioned among users will naturally be influenced by how evenly the total number of landscape cells can be divided among them).
@@
 
-In addition, [whereas in previous versions variable landscape ownership among users had to be customised using `gmse_apply`](SI4.pdf), GMSE v.
-0.6.0.0+ now includes an additional parameter `ownership_var`, which takes effect when `land_ownership = TRUE`.
+In addition, [whereas in previous versions variable landscape ownership among users had to be customised using `gmse_apply`](SI4.html), GMSE v.
+0.6+ now includes an additional parameter `ownership_var`, which takes effect when `land_ownership = TRUE`.
 The value of this parameter is expected to be >=0 and <1.
@@
 
-It should be noted that it is still entirely possible for the user to customise landscapes to whatever extent necessary, using custom `gmse_apply` loops [as described previously](SI4.pdf); however, the new `ownership_var` provides an option in GMSE to alter landscape ownership in a relatively straightforward way.
+It should be noted that it is still entirely possible for the user to customise landscapes to whatever extent necessary, using custom `gmse_apply` loops [as described previously](SI4.html); however, the new `ownership_var` provides an option in GMSE to alter landscape ownership in a relatively straightforward way.
 
@@
 We have focused on the data structures [`AGENTS`](#agents), [`resource_array`](#resources), [`observation_array`](#observation), [`manager_array`](#cost), [`user_array`](#action), and [`LAND`](#land) because these are the data structures that can be most readily manipulated to customise GMSE simulations.
-An example of how to do this within a loop using `gmse_apply` can be found in [SI4](SI4.pdf).
+An example of how to do this within a loop using `gmse_apply` can be found in [Advanced case study options](SI4.html).
 While other data structures exist within GMSE (e.g., see the output of `gmse_apply` when `get_res = "Full"`), we do not recommend manipulating these structures for custom simulations.
 
-Many data structures contain elements that are unused in GMSE v0.6.0.0, and in all cases this is designed for ease of ongoing development of new GMSE features.
+Many data structures contain elements that are unused in GMSE v0.6, and in all cases this is designed for ease of ongoing development of new GMSE features.
 Requests for new features can be made on GitHub using the [GMSE Wiki](https://github.com/bradduthie/gmse/wiki) or the [GMSE Issues](https://github.com/bradduthie/gmse/issues) page.

### 2021-01-12 notebook/cran_vignettes/GMSE.Rmd

--- 2020-05-23 notebook/ms/ms.Rmd
@@
         | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
-        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;'
+        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
 date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
@@
 ---
-
-******************************
 
@@
 GMSE offers a range of parameters to simulate resource dynamics (primarily, but not necessarily biological populations) and management policy options, and includes genetic algorithms to dynamically model stakeholder (manager and user) decision-making.
-Genetic algorithms find adaptive solutions to any simulated conditions given stakeholder-specific goals (see [SI1](SI1.pdf)), allowing GMSE to model scenarios of conservation conflict.
+Genetic algorithms find adaptive solutions to any simulated conditions given stakeholder-specific goals (see [The Genetic Algorithm of GMSE](SI1.pdf)), allowing GMSE to model scenarios of conservation conflict.
 
@@
 1).
-The function `gmse` runs simulations using four predefined individual-based submodels, which can be parameterised to fit various case studies; more tailored submodels can also be defined using the `gmse_apply` function (see [SI2](SI2.pdf)).
+The function `gmse` runs simulations using four predefined individual-based submodels, which can be parameterised to fit various case studies; more tailored submodels can also be defined using the `gmse_apply` function (see [Use of the gmse_apply function](SI2.pdf)).
 By default, GMSE models (1) a population of discrete resources (e.g., a managed species) with individual traits (e.g., location, age) on a spatially-explicit landscape and simulates resource birth, individual movement between landscape cells, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty.
@@
 
-```{r, echo = FALSE, fig.height = 4, fig.width = 4, fig.align = 'left', fig.cap = "Description of one time step of the generalised management strategy evaluation framework, which is comprised of four separate sub-models."}
+```{r, echo = FALSE, fig.height = 4, fig.width = 4, fig.align = 'center', fig.cap = "Description of one time step of the generalised management strategy evaluation framework, which is comprised of four separate sub-models."}
 mbox <- function(x0, x1, y0, y1){
@@
 Consistent with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for agents (stakeholders).
-Instead, a genetic algorithm is used to heuristically find strategies that reflect the individual objectives of each stakeholder in each time step (see [SI1](SI1.pdf) for details).
+Instead, a genetic algorithm is used to heuristically find strategies that reflect the individual objectives of each stakeholder in each time step (see [The Genetic Algorithm of GMSE](SI1.pdf) for details).
 Critically, all stakeholders involved in resource conservation are constrained in their decision-making; managing and using resources takes effort (e.g., time or money), and effort expended in developing or enforcing one policy (for managers) or performing one action (for users) will be effort not expendable elsewhere [@Milner-Gulland2011; @Muller-Hansen2017; @Schluter2017].
@@
 Parameters in `gmse` not listed are set to default values.
-By plotting the output with `plot_gmse_results`, simulation results can be interpreted visually (manager and user decisions can also be interpreted using the `plot_gmse_effort` function, see [SI3](SI3.pdf) and [SI4](SI4.pdf) for an expanded example).
+By plotting the output with `plot_gmse_results`, simulation results can be interpreted visually (manager and user decisions can also be interpreted using the `plot_gmse_effort` function, see [Example case study in GMSE](SI3.pdf) and [Advanced case study options](SI4.pdf) for an expanded example).
 
@@
 
-```{r, echo = FALSE, fig.height = 6, fig.align = 'left', fig.cap = "Results of an example simulation illustrating the management of a protected resource that exploits the land of five farmers.
+```{r, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = "Results of an example simulation illustrating the management of a protected resource that exploits the land of five farmers.
 The upper left panel shows locations of resources (black dots) on the landscape in the final time step of the simulation (multiple resources can occur on the same landscape cell).
@@
 
-The function `gmse_apply` allows custom resource, observation, manager, or user sub-models to be integrated into the GMSE framework (see [SI2](SI2.pdf), [SI4](SI4.pdf), and [SI5](SI5.pdf)).
+The function `gmse_apply` allows custom resource, observation, manager, or user sub-models to be integrated into the GMSE framework (see [Use of the gmse_apply function](SI2.pdf), [Advanced case study options](SI4.pdf), and [Fisheries example integrating FLR](SI5.pdf)).
 Any type of sub-model (e.g., numerical, individual-based) is permitted by defining a function with appropriately specified inputs and outputs; where custom functions are not provided, `gmse_apply` runs default GMSE sub-models used in `gmse`.
@@
 The `gmse_apply` function simulates a single GMSE time step, and therefore must be looped for simulations over multiple time steps.
-Within loops, GMSE arguments can be redefined to simulate changing conditions (e.g., change in policy availability or stakeholder budgets, see [SI2](SI2.pdf) and [SI4](SI4.pdf)), thereby allowing many management scenarios to be simulated *in silico*.
+Within loops, GMSE arguments can be redefined to simulate changing conditions (e.g., change in policy availability or stakeholder budgets, see [Use of the gmse_apply function](SI2.pdf) and [Advanced case study options](SI4.pdf)), thereby allowing many management scenarios to be simulated *in silico*.
 
@@
 
-Here we have introduced the R package GMSE v0.4.0.3, new software for modelling social-ecological dynamics under scenarios of potential conflict.
+Here we have introduced the R package GMSE, software for modelling social-ecological dynamics under scenarios of potential conflict.
 GMSE provides a powerful tool for individual-based modelling simulations while also allowing for extensive model customisation.
-GMSE vignettes provide additional examples for getting started with `gmse` or `gmse_gui` for simulations using default resource, observation, manager, and user sub-models ([SI3](SI3.pdf) and [SI6](SI6.pdf)), and for using `gmse_apply` for advanced model customisation ([SI2](SI2.pdf), [SI4](SI4.pdf), and [SI7](SI7.pdf)) and integration with existing packages (e.g., Fisheries Library in R, see [SI5](SI5.pdf)).
+GMSE vignettes provide additional examples for getting started with `gmse` or `gmse_gui` for simulations using default resource, observation, manager, and user sub-models ([Example case study in GMSE](SI3.pdf), [Management frequency and extinction risk](SI6.pdf), and [Adaptive timing of investment strategy](SI8.pdf)), and for using `gmse_apply` for advanced model customisation ([Use of the gmse_apply function](SI2.pdf), [Advanced case study options](SI4.pdf), and [Default GMSE data structures](SI7.pdf)) and integration with existing packages (e.g., Fisheries Library in R, see [Fisheries example integrating FLR](SI5.pdf)).
 Future versions of GMSE will include additional features and improve upon the realism of social and ecological modelling components, while also maintaining a high degree of flexibility and modulatarity for model customisation.

### 2021-01-12 notebook/cran_vignettes/SI7.Rmd

--- 2020-05-24 notebook/ms/SI7.Rmd
@@
 While these default sub-models do not necessarily need to be used in every run in GMSE (see [use of](SI2.pdf) `gmse_apply`), they will be used in any run of the `gmse` function, and in any call of the `gmse_apply` function that does not run with entirely custom sub-models.
-Simulation and model inference do not require an understanding of the default data structures, but such an understanding can be especially useful when running `gmse_apply` if there is a need to extract uncommonly used information, change key simulated values (e.g., landscape properties, agent budgets, or resource movement rules, as in [SI4](SI4.pdf)), or build custom individual-based sub-models.
+Simulation and model inference do not require an understanding of the default data structures, but such an understanding can be especially useful when running `gmse_apply` if there is a need to extract uncommonly used information, change key simulated values (e.g., landscape properties, agent budgets, or resource movement rules, as in [Advanced case study options](SI4.pdf)), or build custom individual-based sub-models.
 Here we provide a brief explanation of the following key data structures (each name below is listed as it is named in the output `gmse_apply` when `get_res = "Full"`).
@@
 
-Note that these are not the only data structures used in GMSE, but they are the only ones that can be easily modified in GMSE v0.4.0.7 (see, e.g., [SI4](SI4.pdf)), so they are the ones that we focus on here.
+Note that these are not the only data structures used in GMSE, but they are the only ones that can be easily modified in GMSE (see, e.g., [Advanced case study options](SI4.pdf)), so they are the ones that we focus on here.
 Additionally, any custom subfunction that returns an array rather than a single value should adhere to the same structure as these defaults if any default GMSE functions are to be used in `gmse_apply`.
@@
 
-In [SI4](SI4.pdf), we show how landscape cell values can be manipulated to customise the placement of land ownership.
-By default, GMSE v0.6.0.0+ places the ownership of landscape cells using a shortest-splitline algorithm (see the lab notebook for [a detailed explanation](https://confoobio.github.io/gmse/notebook/gmse_notes.html#u-10-MAY-2020)).
+In [Advanced case study options](SI4.pdf), we show how landscape cell values can be manipulated to customise the placement of land ownership.
+By default, GMSE v0.6+ places the ownership of landscape cells using a shortest-splitline algorithm (see the lab notebook for [a detailed explanation](https://confoobio.github.io/gmse/notebook/gmse_notes.html#u-10-MAY-2020)).
 The algorithm splits the landscape evenly into smaller and smaller rectangles until a sufficient number of rectangles are produced for all users to own a similar amount of land (note that how uniformly land is partitioned among users will naturally be influenced by how evenly the total number of landscape cells can be divided among them).
@@
 In addition, [whereas in previous versions variable landscape ownership among users had to be customised using `gmse_apply`](SI4.pdf), GMSE v.
-0.6.0.0+ now includes an additional parameter `ownership_var`, which takes effect when `land_ownership = TRUE`.
+0.6+ now includes an additional parameter `ownership_var`, which takes effect when `land_ownership = TRUE`.
 The value of this parameter is expected to be >=0 and <1.
@@
 
-```{r, echo = FALSE, fig.height = 5, fig.width = 5}
+```{r, echo = FALSE, fig.height = 4.75, fig.width = 4.75}
 par(mar = c(0.2, 0.2, 0.2, 0.2));
@@
 We have focused on the data structures [`AGENTS`](#agents), [`resource_array`](#resources), [`observation_array`](#observation), [`manager_array`](#cost), [`user_array`](#action), and [`LAND`](#land) because these are the data structures that can be most readily manipulated to customise GMSE simulations.
-An example of how to do this within a loop using `gmse_apply` can be found in [SI4](SI4.pdf).
+An example of how to do this within a loop using `gmse_apply` can be found in [Advanced case study options](SI4.pdf).
 While other data structures exist within GMSE (e.g., see the output of `gmse_apply` when `get_res = "Full"`), we do not recommend manipulating these structures for custom simulations.
 
-Many data structures contain elements that are unused in GMSE v0.6.0.0, and in all cases this is designed for ease of ongoing development of new GMSE features.
+Many data structures contain elements that are unused in GMSE v0.6, and in all cases this is designed for ease of ongoing development of new GMSE features.
 Requests for new features can be made on GitHub using the [GMSE Wiki](https://github.com/bradduthie/gmse/wiki) or the [GMSE Issues](https://github.com/bradduthie/gmse/issues) page.

### 2021-01-12 notebook/ms/ms.Rmd

@@
     fig_caption: yes
-  html_document: default
   word_document: default

### 2021-01-12 notebook/website_vignettes/GMSE.Rmd

--- 2021-01-12 notebook/ms/ms.Rmd
@@
 GMSE offers a range of parameters to simulate resource dynamics (primarily, but not necessarily biological populations) and management policy options, and includes genetic algorithms to dynamically model stakeholder (manager and user) decision-making.
-Genetic algorithms find adaptive solutions to any simulated conditions given stakeholder-specific goals (see [SI1](SI1.pdf)), allowing GMSE to model scenarios of conservation conflict.
+Genetic algorithms find adaptive solutions to any simulated conditions given stakeholder-specific goals (see [The Genetic Algorithm of GMSE](SI1.html)), allowing GMSE to model scenarios of conservation conflict.
 
@@
 1).
-The function `gmse` runs simulations using four predefined individual-based submodels, which can be parameterised to fit various case studies; more tailored submodels can also be defined using the `gmse_apply` function (see [SI2](SI2.pdf)).
+The function [`gmse`](../reference/gmse.html) runs simulations using four predefined individual-based submodels, which can be parameterised to fit various case studies; more tailored submodels can also be defined using the [`gmse_apply`](../reference/gmse_apply.html) function (see [Use of the gmse_apply function](SI2.html)).
 By default, GMSE models (1) a population of discrete resources (e.g., a managed species) with individual traits (e.g., location, age) on a spatially-explicit landscape and simulates resource birth, individual movement between landscape cells, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty.
@@
 Once each user has found an adaptive strategy, user actions affect resources and landscape cells, feeding back into the resource sub-model.
+
+********************************************************************************
 
@@
 
+********************************************************************************
+
 Genetic Algorithm
@@
 Consistent with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for agents (stakeholders).
-Instead, a genetic algorithm is used to heuristically find strategies that reflect the individual objectives of each stakeholder in each time step (see [SI1](SI1.pdf) for details).
+Instead, a genetic algorithm is used to heuristically find strategies that reflect the individual objectives of each stakeholder in each time step (see [The Genetic Algorithm of GMSE](SI1.html) for details).
 Critically, all stakeholders involved in resource conservation are constrained in their decision-making; managing and using resources takes effort (e.g., time or money), and effort expended in developing or enforcing one policy (for managers) or performing one action (for users) will be effort not expendable elsewhere [@Milner-Gulland2011; @Muller-Hansen2017; @Schluter2017].
@@
 
-Parameters in `gmse` not listed are set to default values.
-By plotting the output with `plot_gmse_results`, simulation results can be interpreted visually (manager and user decisions can also be interpreted using the `plot_gmse_effort` function, see [SI3](SI3.pdf) and [SI4](SI4.pdf) for an expanded example).
+Parameters in [`gmse`](../reference/gmse.html) not listed are set to default values.
+By plotting the output with [`plot_gmse_results`](../reference/plot_gmse_results.html), simulation results can be interpreted visually (manager and user decisions can also be interpreted using the [`plot_gmse_effort`](../reference/plot_gmse_effort.html) function, see [Example case study in GMSE](SI3.html) and [Advanced case study options](SI4.html) for an expanded example).
+
+********************************************************************************
 
@@
 
-```{r, echo = FALSE, fig.height = 6, fig.align = 'left', fig.cap = "Results of an example simulation illustrating the management of a protected resource that exploits the land of five farmers.
+```{r, echo = FALSE, fig.width = 6, fig.height = 9, fig.align = 'left', fig.cap = "Results of an example simulation illustrating the management of a protected resource that exploits the land of five farmers.
 The upper left panel shows locations of resources (black dots) on the landscape in the final time step of the simulation (multiple resources can occur on the same landscape cell).
@@
 ```
+
+********************************************************************************
 
@@
 
-The function `gmse_gui` opens GMSE in a browser and allows simulations to be run for most `gmse` parameter options using the package 'shiny' [@Chang2017].
-Figures from `plot_gmse_results` and `plot_gmse_effort`, and tables from `gmse_summary` are provided as GUI output.
+The function [`gmse_gui`](../reference/gmse_gui.html) opens GMSE in a browser and allows simulations to be run for most [`gmse`](../reference/gmse.html) parameter options using the package 'shiny' [@Chang2017].
+Figures from [`plot_gmse_results`](../reference/plot_gmse_results) and [`plot_gmse_effort`](../reference/plot_gmse_effort), and tables from [`gmse_summary`](../reference/gmse_summary) are provided as GUI output.
 
@@
 
-The function `gmse_apply` allows custom resource, observation, manager, or user sub-models to be integrated into the GMSE framework (see [SI2](SI2.pdf), [SI4](SI4.pdf), and [SI5](SI5.pdf)).
-Any type of sub-model (e.g., numerical, individual-based) is permitted by defining a function with appropriately specified inputs and outputs; where custom functions are not provided, `gmse_apply` runs default GMSE sub-models used in `gmse`.
-Any parameter options available in `gmse` or in custom functions can be passed directly to `gmse_apply`, thereby allowing for high flexibility in model specification.
+The function [`gmse_apply`](../reference/gmse_apply.html) allows custom resource, observation, manager, or user sub-models to be integrated into the GMSE framework (see [Use of the gmse_apply function](SI2.html), [Advanced case study options](SI4.html), and [Fisheries example integrating FLR](SI5.html)).
+Any type of sub-model (e.g., numerical, individual-based) is permitted by defining a function with appropriately specified inputs and outputs; where custom functions are not provided, [`gmse_apply`](../reference/gmse_apply.html) runs default GMSE sub-models used in [`gmse`](../reference/gmse.html).
+Any parameter options available in [`gmse`](../reference/gmse.html) or in custom functions can be passed directly to [`gmse_apply`](../reference/gmse_apply.html), thereby allowing for high flexibility in model specification.
 For example, a simple logistic growth function can be integrated as a resource sub-model to replace the default `resource` function.
@@
 
-The `gmse_apply` function simulates a single GMSE time step, and therefore must be looped for simulations over multiple time steps.
-Within loops, GMSE arguments can be redefined to simulate changing conditions (e.g., change in policy availability or stakeholder budgets, see [SI2](SI2.pdf) and [SI4](SI4.pdf)), thereby allowing many management scenarios to be simulated *in silico*.
+The [`gmse_apply`](../reference/gmse_apply.html) function simulates a single GMSE time step, and therefore must be looped for simulations over multiple time steps.
+Within loops, GMSE arguments can be redefined to simulate changing conditions (e.g., change in policy availability or stakeholder budgets, see [Use of the gmse_apply function](SI2.html) and [Advanced case study options](SI4.html)), thereby allowing many management scenarios to be simulated *in silico*.
 
@@
 
-Here we have introduced the R package GMSE v0.4.0.3, new software for modelling social-ecological dynamics under scenarios of potential conflict.
+Here we have introduced the R package GMSE, software for modelling social-ecological dynamics under scenarios of potential conflict.
 GMSE provides a powerful tool for individual-based modelling simulations while also allowing for extensive model customisation.
-GMSE vignettes provide additional examples for getting started with `gmse` or `gmse_gui` for simulations using default resource, observation, manager, and user sub-models ([SI3](SI3.pdf) and [SI6](SI6.pdf)), and for using `gmse_apply` for advanced model customisation ([SI2](SI2.pdf), [SI4](SI4.pdf), and [SI7](SI7.pdf)) and integration with existing packages (e.g., Fisheries Library in R, see [SI5](SI5.pdf)).
+GMSE vignettes provide additional examples for getting started with [`gmse`](../reference/gmse.html) or [`gmse_gui`](../reference/gmse_gui.html) for simulations using default resource, observation, manager, and user sub-models ([Example case study in GMSE](SI3.html), [Management frequency and extinction risk](SI6.html), and [Adaptive timing of investment strategy](SI8.html)), and for using [`gmse_apply`](../reference/gmse_apply.html) for advanced model customisation ([Use of the gmse_apply function](SI2.html), [Advanced case study options](SI4.html), and [Default GMSE data structures](SI7.html)) and integration with existing packages (e.g., Fisheries Library in R, see [Fisheries example integrating FLR](SI5.html)).
 Future versions of GMSE will include additional features and improve upon the realism of social and ecological modelling components, while also maintaining a high degree of flexibility and modulatarity for model customisation.

### 2021-01-12 notebook/website_vignettes/SI2.Rmd

--- 2020-05-26 notebook/cran_vignettes/SI2.Rmd
@@
 
-It is possible to, e.g., specify a simple resource and observation model, but then take advantage of the genetic algorithm to predict policy decisions and user actions (see [Fisheries example integrating FLR](SI5.pdf) for a fisheries example).
+It is possible to, e.g., specify a simple resource and observation model, but then take advantage of the genetic algorithm to predict policy decisions and user actions (see [Fisheries example integrating FLR](SI5.html) for a fisheries example).
 This can be done by using the default GMSE manager and user functions (written below explicitly, though this is not necessary).

### 2021-01-12 notebook/website_vignettes/SI6.Rmd

--- 2020-05-29 notebook/cran_vignettes/SI6.Rmd
@@
 In GMSE, some of the most important traits for resources include types, ages, demographic parameter values, locations, etc., and for agents (manager and users), traits include different types, utilities, budgets, etc.
-The traits that resources and managers have can potentially affect their interactions, and default GMSE sub-models take advantage of this by simulating interactions explicitly on a landscape (see [Default GMSE data structures](SI7.pdf) for an introduction to GMSE default data structures).
+The traits that resources and managers have can potentially affect their interactions, and default GMSE sub-models take advantage of this by simulating interactions explicitly on a landscape (see [Default GMSE data structures](SI7.html) for an introduction to GMSE default data structures).
 

### 2021-07-30 notebook/cran_vignettes/SI2.Rmd
Update the vignette for Issue #64 to explain the need for a landscape reset in gmse_apply loops

@@
 Note that elements of `old_list` will over-ride custom arguments to `gmse_apply` so **specifying custom arguments that are already present in `old_list` will not work**.
+
+Replenishing crops after consumption
+================================================================================
+
+Unlike with the `gmse` function, `gmse_apply` does not automatically assume that crop production should be replenished after a single time step.
+The second layer of [the landscape](SI7.pdf) holds crop production on a cell.
+This will be depleted if resources consume crops on the landscape.
+
+```{r}
+sim_consume <- gmse_apply(land_dim_1 = 8, land_dim_2 = 8, 
+                          res_consume = 0.02, get_res = "Full");
+print(round(sim_consume$LAND[,,2], digits = 2));
+```
+
+If we run this for five time steps using a loop in `gmse_apply`, then resources will continue to deplete crops on the landscape.
+
+```{r, eval = TRUE}
+sim_old     <- gmse_apply(land_dim_1 = 8, land_dim_2 = 8, 
+                          res_consume = 0.02, get_res = "Full");
+for(time_step in 1:5){
+    sim_new                 <- gmse_apply(get_res = "Full", old_list = sim_old);
+    sim_old                 <- sim_new;
+}
+print(round(sim_old$LAND[,,2], digits = 2));
+```
+
+Notice that the amount of crops on each cell has decreased substantially after five time steps. To replenish the crops after every `repl` time step, we can use the following code.
+
+```{r, eval = TRUE}
+sim_old     <- gmse_apply(land_dim_1 = 8, land_dim_2 = 8, 
+                          res_consume = 0.02, get_res = "Full");
+repl        <- 1;
+for(time_step in 1:5){
+    sim_new                 <- gmse_apply(get_res = "Full", old_list = sim_old);
+    if(time_step %% repl == 0){ # If remainder of time_step / repl is zero
+        sim_new$LAND[,,2] <- 1;
+    }
+    sim_old                 <- sim_new;
+}
+print(round(sim_old$LAND[,,2], digits = 2));
+```
+
+Now with the crop on the landscape replenishing every time step, each new time step starts with the landscape crop values set to 1.
+This is likely to be critical if simulating resources that must consume a certain amount to survive (`consume_surv > 0`) or reproduce (`consume_repr > 0`), and `gmse_apply` thereby provides some flexibility in terms of how frequently (and how much) landscape values change from one time step to the next.

### 2021-07-30 notebook/website_vignettes/SI2.Rmd
Update the vignette for Issue #64 to explain the need for a landscape reset in gmse_apply loops

@@
 Note that elements of `old_list` will over-ride custom arguments to `gmse_apply` so **specifying custom arguments that are already present in `old_list` will not work**.
+
+Replenishing crops after consumption
+================================================================================
+
+Unlike with the `gmse` function, `gmse_apply` does not automatically assume that crop production should be replenished after a single time step.
+The second layer of [the landscape](https://confoobio.github.io/gmse/articles/SI7.html#6-land) holds crop production on a cell.
+This will be depleted if resources consume crops on the landscape.
+
+```{r}
+sim_consume <- gmse_apply(land_dim_1 = 8, land_dim_2 = 8, 
+                          res_consume = 0.02, get_res = "Full");
+print(round(sim$LAND[,,2], digits = 2));
+```
+
+If we run this for five time steps using a loop in `gmse_apply`, then resources will continue to deplete crops on the landscape.
+
+```{r, eval = TRUE}
+sim_old     <- gmse_apply(land_dim_1 = 8, land_dim_2 = 8, 
+                          res_consume = 0.02, get_res = "Full");
+for(time_step in 1:5){
+    sim_new                 <- gmse_apply(get_res = "Full", old_list = sim_old);
+    sim_sum_4[time_step, 1] <- time_step;
+    sim_old                 <- sim_new;
+}
+print(round(sim$LAND[,,2], digits = 2));
+```
+
+Notice that the amount of crops on each cell has decreased substantially after five time steps. To replenish the crops after every `repl` time step, we can use the following code.
+
+```{r, eval = TRUE}
+sim_old     <- gmse_apply(land_dim_1 = 8, land_dim_2 = 8, 
+                          res_consume = 0.02, get_res = "Full");
+repl        <- 1;
+for(time_step in 1:5){
+    sim_new                 <- gmse_apply(get_res = "Full", old_list = sim_old);
+    sim_sum_4[time_step, 1] <- time_step;
+    if(time_step %% repl == 0){ # If remainder of time_step / repl is zero
+        sim_old$LAND[,,2] <- 1;
+    }
+    sim_old                 <- sim_new;
+}
+print(round(sim$LAND[,,2], digits = 2));
+```
+
+Now with the crop on the landscape replenishing every time step, consumption after five time steps comparable to consumption after a single time step.
+This is likely to be critical if simulating resources that must consume a certain amount to survive (`consume_surv > 0`) or reproduce (`consume_repr > 0`), and `gmse_apply` thereby provides some flexibility in terms of how frequently (and how much) landscape values change from one time step to the next.

### 2021-07-30 notebook/cran_vignettes/SI1.Rmd
Update GMSE version and equation, checking off an item in Issue #64 for the documentation

--- 2020-05-23 notebook/ms/SI1.Rmd
@@
         | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
-        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;'
-date: '[2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
+        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
+date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
 output:
-  html_document: default
   pdf_document:
@@
 - \usepackage{lineno}
-- \linenumbers
 linestretch: 1
@@
 As of GMSE version 0.6, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step.
-Therefore, one run of the genetic algorithm occurs to simulate the manager's policy-setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1; e.g., see [SI6](SI6.pdf)), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users copy).
+Therefore, one run of the genetic algorithm occurs to simulate the manager's policy-setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1; e.g., see [Management frequency and extinction risk](SI6.pdf)), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users copy).
 Each run of the genetic algorithm mimics the evolution by natural selection of a population of potential manager or user strategies over multiple iterations, with the highest fitness strategy in the terminal iteration being selected as the one that the manager or user decides to implement.
 For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'iteration' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the [main text](ms.pdf)).
-Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.6 (future versions of GMSE might expand upon this framework, and we highlight some of these potential avenues for expansion).
+Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.6+ (future versions of GMSE might expand upon this framework, and we highlight some of these potential avenues for expansion).
 We first explain the [key data structures used](#Key_data_structures_used), then provide an overview of how a population of strategies is [initialised](#initialised), and the subsequent processes of [crossover](#crossover), [mutation](#mutation), [cost constraint](#constraint), [fitness](#eval) evaluation, [tournament selection](#tournament), and [replacement](#replacement).
@@
 
-The focal data structure used for tracking manager and user decisions is a three dimensional array, which we will call `ACTION` (also returned as `user_array` by `gmse_apply`; see [SI7](SI7.pdf)).
+The focal data structure used for tracking manager and user decisions is a three dimensional array, which we will call `ACTION` (also returned as `user_array` by `gmse_apply`; see [Default GMSE data structure](SI7.pdf)).
 Rows of `ACTION` correspond to the entities affected by actions (resources, landscape properties, or potentially other agents), and columns correspond either to properties of the affected entities, or to the actions potentially allocated to them.
@@
 
-Columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in predefined GMSE v0.6 manager and user sub-models (i.e., only one type of resource is permitted).
+Columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in predefined GMSE v0.6+ manager and user sub-models (i.e., only one type of resource is permitted).
 Future versions might allow for different resource types (e.g., `Type_1` might be used to designate species, and `Type_2` and `Type_3` could designate stage or sex).
@@
 
-Constraints on the values that elements in the `ACTION` array can take are defined by a `COST` array (also returned as `manager_array` by `gmse_apply`; see [SI7](SI7.pdf)) of dimensions identical to `ACTION`.
+Constraints on the values that elements in the `ACTION` array can take are defined by a `COST` array (also returned as `manager_array` by `gmse_apply`; see [Default GMSE data structures](SI7.pdf)) of dimensions identical to `ACTION`.
 Elements of `COST` define how many units from the `manager_budget` or `user_budget` are needed to perform a single action; a `minimum_cost` for actions is defined as an argument in GMSE (10 by default).
@@
 Here we explain how the fitnesses of candidate manager and user strategies in a `POPULATION` array (see above) are calculated.
-We emphasise that the fitness functions used in GMSE v0.6 are intended to be heuristic tools for identifying reasonable manager and user behaviours.
+We emphasise that the fitness functions used in GMSE v0.6+ are intended to be heuristic tools for identifying reasonable manager and user behaviours.
 In practice, our fitness functions identify behaviours that are well-aligned with manager and user interests for harvesting or crop yield, but they are not intended to identify *optimal* decisions.
@@
 Individual fitness as calculated for managers ($F_{i}^{m}$) is affected by a manager's utility for resources and the projected change in resource abundance caused by the individual's policy (i.e., the contents of their `POPULATION` layer, specifically row 3; here again we use 'individual' to refer to one of `ga_popsize` discrete strategies in `POPULATION`, which may be selected and reproduce within the genetic algorithm).
-Manager utility for a resource ($U^{m}_{res}$) is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model (see "[Key data structures used](#Key_data_structures_used)" above, and [SI7](SI7.pdf#action) for more information).
+Manager utility for a resource ($U^{m}_{res}$) is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model (see "[Key data structures used](#Key_data_structures_used)" above, and [Default GMSE data structures](SI7.pdf) for more information).
 Manager utility can therefore change in each GMSE time step as estimated resource abundance changes; when the estimated resource abundance is greater than `manage_target`, $U^{m}_{res}$ is negative, and when the estimated resource abundance is less than `manage_target`, $U^{m}_{res}$ is positive.
@@
 Managers also have access to an estimate of the number of offspring that resources produce ($E[offspring]$).
-For default settings in GMSE v0.6, $E[offspring] = \lambda$, where $\lambda$ is the GMSE argument `lambda` that defines the baseline population growth rate of resources.
+For default settings in GMSE v0.6+, $E[offspring] = \lambda$, where $\lambda$ is the GMSE argument `lambda` that defines the baseline population growth rate of resources.
 Given a non-default GMSE argument for `consume_repr` (where `consume_repr` is how much consumption of landscape yield is needed to increment offspring production by 1), $E[offspring]$ is further incremented by `( (times_feeding * res_consume) / consume_repr)` to account for reproduction enabled by feeding on the landscape (`res_consume` is the proportion of landscape yield that a resource consumes during feeding).
@@
 Note that $F_{i}^{u}$ increases when $\Delta A_{i}$ and $\Delta L_{i}$ are of the same sign as $U^{u}_{res}$ and $U^{u}_{land}$, respectively.
-Further, in GMSE v0.6, only one term of the equation is nonzero.
+Further, in GMSE v0.6+, only one term of the equation is nonzero.
 When `land_ownership = FALSE` (default, modelling users that harvest resources), $U^{u}_{res} = -1$ and $U^{u}_{land} = 0$, and when `land_ownership = TRUE`, $U^{u}_{res} = 0$ and $U^{u}_{land} = 100$ (modelling farmers trying to increase crop yield).
@@
 
-$$\Delta A_{i} = (\lambda) Feeds + Helps -Scares - Culls - (\lambda) Castrations.$$
+$$\Delta A_{i} = (\lambda) Feeds + Helps - Scares - (1 + \lambda)Culls - (\lambda) Castrations.$$
 

### 2021-07-30 notebook/cran_vignettes/SI7.Rmd
Update GMSE version and equation, checking off an item in Issue #64 for the documentation

@@
 
-Many data structures contain elements that are unused in GMSE v0.6, and in all cases this is designed for ease of ongoing development of new GMSE features.
+Many data structures contain elements that are unused in GMSE v0.6+, and in all cases this is designed for ease of ongoing development of new GMSE features.
 Requests for new features can be made on GitHub using the [GMSE Wiki](https://github.com/bradduthie/gmse/wiki) or the [GMSE Issues](https://github.com/bradduthie/gmse/issues) page.

### 2021-07-30 notebook/website_vignettes/SI1.Rmd
Update GMSE version and equation, checking off an item in Issue #64 for the documentation

@@
 For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'iteration' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the [main text](ms.pdf)).
-Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.6 (future versions of GMSE might expand upon this framework, and we highlight some of these potential avenues for expansion).
+Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.6+ (future versions of GMSE might expand upon this framework, and we highlight some of these potential avenues for expansion).
 We first explain the [key data structures used](#Key_data_structures_used), then provide an overview of how a population of strategies is [initialised](#initialised), and the subsequent processes of [crossover](#crossover), [mutation](#mutation), [cost constraint](#constraint), [fitness](#eval) evaluation, [tournament selection](#tournament), and [replacement](#replacement).

### 2021-07-31 notebook/cran_vignettes/SI3.Rmd
Update vignettes SI3 and SI4 for CRAN

--- 2020-05-24 notebook/ms/SI3.Rmd
@@
 
-To demonstrate how GMSE can be used to understand conflict in social-ecological systems, we build upon the [example of resource management](https://www.biorxiv.org/content/biorxiv/early/2017/11/17/221432.full.pdf#page=5) in the main text.
+To demonstrate how GMSE can be used to understand conflict in social-ecological systems, we build upon the [example of resource management](ms.pdf) in the main text.
 We consider a protected population of waterfowl that exploits and damages agricultural land and is therefore a source of conservation conflict between those that seek the conservation of waterfowl and those that are concerned with the loss of agriculture [e.g., @Fox2017a; @Mason2017; @Tulloch2017; @Cusack2018].
@@
 Novice R users may prefer to run all of the simulations below using the browser-based GMSE GUI by calling `gmse_gui()` from the R command line.
-Alternatively, experienced R users may prefer to simulate by [looping time steps](SI2.pdf#page=5) through `gmse_apply`, which allows more flexibility for incorporating custom sub-models and dynamically adjusting parameter values.
+Alternatively, experienced R users may prefer to simulate by [looping time steps](SI2.pdf) through `gmse_apply`, which allows more flexibility for incorporating custom sub-models and dynamically adjusting parameter values.
 
 Simulations using the default GMSE sub-models described above are run using the `gmse` function, which offers a range of options for setting parameter values (see Table 1 for some select examples).
-Output of `gmse` is an exhaustive list that includes all resources and observations, all stakeholder decisions and actions, and all landscape properties in each time step of the simulation (see [SI7](SI7.pdf) for a description of key data structures).
+Output of `gmse` is an exhaustive list that includes all resources and observations, all stakeholder decisions and actions, and all landscape properties in each time step of the simulation (see [Default GMSE data structures](SI7.pdf) for a description of key data structures).
 
@@
 Within the default sub-models of GMSE, there are several non-default options that can be set using arguments to `gmse` and `gmse_apply`.
-Brief explanations of these non-default options appear in the [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf) (also [listed](https://confoobio.github.io/gmse/reference/gmse.html) on the [GMSE website](https://confoobio.github.io/gmse/)).
+Brief explanations of these non-default options appear in the [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf) (also listed on the [GMSE website](https://confoobio.github.io/gmse/)).
 Here we further explain some of the less obvious or less intuitive options available in GMSE, for which additional explanation beyond the GMSE documentation may be warranted for application to case studies such as the Taiga Bean Geese case study example demonstrated above.
@@
 In this option, managers observe resources on a subset of the landscape; subset size is determined by the manager's view as set using the `agent_view` parameter.
-In practice, this is done by having the manager pick a cell on the landscape at random, then count all of the resources within a distance (in any direction) of `agent_view` cells.
+In practice, this is done by having the manager count all of the resources within a distance (in any direction) of `agent_view` cells of their location.
 Managers then extrapolate the density of resources within this subset to estimate the total number of resources on the landscape.
@@
 If desired, the `times_observe` (default equals 1) option can also be used to allow managers to sample more than one landscape subset in their estimate (e.g., if `times_observe = 2`, then density is estimated from two different locations, with the mean of individual estimates used for a combined estimate of population density).
+Note that if `times_observe = 1`, then manager estimates might be consistently biased if they are observing from a location with disproportionately many or few resources.
   1.
@@
 [Requests for new features](https://github.com/bradduthie/gmse/issues) and contributions to GMSE can be made through [GitHub](https://github.com/bradduthie/gmse).
-Additionally, where entirely different types of sub-models are required, the `gmse_apply` function [can be used](https://bradduthie.github.io/gmse_SI2.pdf) to more flexibly simulate social-ecological systems.
-In [SI4](SI4.pdf), we show an example of this using the same Taiga Bean Geese case study that we did here.
+Additionally, where entirely different types of sub-models are required, the `gmse_apply` function [can be used](SI2.pdf) to more flexibly simulate social-ecological systems.
+In [Advanced case study options](SI4.pdf), we show an example of this using the same Taiga Bean Geese case study that we did here.
+
+References
+================================================================================

### 2021-07-31 notebook/cran_vignettes/SI3.Rmd
Update the vignettes as necessary with changes mentioned in Issue #64 -- mostly typos, but also some small edits to the code that needed to be clarified

@@
   0.
-*No explicit density independent or dependent removal*.
-This option does not cause removal to be *directly* caused in a density independent or dependent way.
-Nevertheless, removal may still be possible if the argument `consume_surv > 0` (its default value is 0).
-In this case, survival of a resource in a time step depends on that resource consuming at least `consume_surv` in yield on the landscape.
-In such cases, a natural carrying capacity will be generated as more resources result in higher total consumption.
-The amount yield that a resource consumes is further affected by both `times_feeding` (how many times the resource moves and feeds on a new landscape cell during a time step) and `res_consume` (the proportion of yield on a cell that is is consumed upon visit to the cell).
-Note that `consume_surv` is can still be used when `res_death_type > 0` (below), resulting in multiple potential sources of removal.
-Given `res_death_type = 0`, **user actions have no effect** on resource removal (to ensure that user actions do have an effect, set `res_death_type = 1` and `removal_pr = 0`; see below).
+*No explicit density independent or dependent removal, and no old age*.
+Given this option, the only way in resources can be removed is if users remove them through their actions (i.e., culling).
+Resources cannot die due to lack of resource consumption, density independent or dependent effects, or old age.
   1.
 *Density-independent removal*.
-Given this option, there is no density-dependent removal of resources within the resource sub-model.
+Given this option, there is no explicit density-dependent removal of resources within the resource sub-model.
 Resources are simply removed with a fixed probability of `removal_pr` (its default value is 0) each time the sub-model is run.
 This option should be used with great care because in the absence of density-dependence, it is possible for the population of resources to increase without limit and thereby greatly slow down simulation times.
+Usually `res_death_type = 1` is most useful when combined with consumption limited survival (`consume_surv`) and reproduction (`consume_repr`); see below.
   2.
 *Density-dependent removal* (default).
-Given this option, removal of resources within the resource sub-model is caused entirely by density effects.
+Given this option, removal of resources within the resource sub-model is caused by density effects.
 The probability that a resource is removed is a function of the resource carrying capacity set using the `res_death_K` parameter.
@@
 If removal occurs as a consequence of (1), (2), or both, then the resource is removed.
+
+Note that for all `res_death_type > 0`, removal can occur when the argument `consume_surv > 0` (its default value is 0).
+In this case, survival of a resource in a time step depends on that resource consuming at least `consume_surv` in yield on the landscape.
+In such cases, a natural carrying capacity can be generated as more resources result in higher total consumption.
+The amount of yield that a resource consumes is further affected by both `times_feeding` (how many times the resource moves and feeds on a new landscape cell during a time step) and `res_consume` (the proportion of yield on a cell that is is consumed upon visit to the cell).
+To ensure that resource death is *only* limited by consumption and user actions, set `res_death_type = 1` and `removal_pr = 0`.
+Under this scenario, resources will only die if the do not consume enough, are culled by users, or are above the `max_age`.
 
@@
 
+**times_feeding**: By default, resources feed once at the start of each time step (`times_feeding = 1`), then move at the end the time step.
+Consequently, where they move to at the end of the time step will be where they feed at the start of the next time step, which gives users the opportunity to potentially cull or scare them prior to feeding.
+When `times_feeding > 1`, at the start of each time step, resource feeding on their cell, then immediately move to a new cell on which to feed.
+This feeding and moving continues until the resource has fed `times_feeding` times, at which point on the cell at which they last fed, potentially reproduce, then move again at the end of the time step.
+Feeding and moving actions happen in a random order for resources so that one resource does not do all of its feeding before another resource has any chance to feed.
+For example, if there are 10 resources and `times_feeding = 4`, then the 40 total feeding and moving actions happen in a random order.
+
 Conclusions and future development

### 2021-07-31 notebook/cran_vignettes/SI8.Rmd
Update the vignettes as necessary with changes mentioned in Issue #64 -- mostly typos, but also some small edits to the code that needed to be clarified

--- 2020-05-21 notebook/ms/SI8.Rmd
@@
 
-And here the same simulation, exept managers update policy only when they observe the population to be plus or minus the target by 20%, and receive a bonus of 30% of their initial budget when they do not.
+And here the same simulation, except managers update policy only when they observe the population to be plus or minus the target by 20%, and receive a bonus of 30% of their initial budget when they do not.
 
@@
 
-Note how the costs plateau (lower left panel) when the population monitored (blue line in the upper left panel) is in between the two green dotted lines.
+Note how the costs plateau (lower left panel) when the population monitored (blue line, upper left panel) is in between the two green dotted lines.
 Leaving the policy unchanged in these cases allowed managers to set the costs higher than the default maximum of 110 budget units.

### 2022-04-14 notebook/cran_vignettes/SI8.Rmd
Initialise a new vignette to explain the simulated annealing approach

@@
 
-In the context of adaptive management of conservation conflict, managers tend to invest their ressources in updating the conservation policy as soon as possible.
+In the context of adaptive management of conservation conflict, managers tend to invest their resources in updating the conservation policy as soon as possible.
 But leaving the policy unchanged when the population is close enough to the target could actually save money, time, and energy, and enpower managers for the next iteration.

### 2022-04-14 notebook/cran_vignettes/SI9.Rmd
Initialise a new vignette to explain the simulated annealing approach

---
title: 'The simulated annealing algorithm of GMSE'
subtitle: 'GMSE: an R package for generalised management strategy evaluation (New function vignette)'
author: | 
        | Brad Duthie&#xb9; &#xb2;, Gabriela Ochoa&#xb9;
date: "[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK [2] alexander.duthieh@stir.ac.uk"
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes:
- \usepackage{lineno}
linestretch: 1
bibliography: gmse.bib
biblio-style: apalike
link-citations: true
linkcolor: blue
---

```{r, echo = FALSE}
library(GMSE);
# sim <- gmse(action_thres = 0.1, budget_bonus = 0.5, plotting = FALSE);
```

Overview of simulated annealing
================================================================================

The primary algorithm used in GMSE to model manager and user decision-making is the [genetic algorithm](SI1.pdf). As of version 1.0, GMSE also allows for the use of a simulated annealing algorithm [@Kirkpatrick1983; @Cerny1985; @Rutenbar1989]. Simulated annealing searches a space of potential solutions (in GMSE, decisions made by agents to maximise their own utility) in an iterative way. Unlike the [genetic algorithm](SI1.pdf), it is not a population focused approach.

Example use of simulated annealing
================================================================================

Performance of simulated annealing
================================================================================

References
================================================================================

### 2022-04-15 src/game.c
Minor update

@@ -633,7 +633,7 @@ void manager_fitness(double *fitnesses, double ***population, double **jaco,
       if(sim_ann > 0){
         fitnesses[agent] = 10000000 - dev_from_util[agent];
       }else{
-        fitnesses[agent] = max_dev - dev_from_util[agent]; 
+        fitnesses[agent] = 10000000 - dev_from_util[agent]; 
       }
     }
     
@@ -1011,7 +1011,7 @@ void sa(double ***ACTION, double ***COST, double **AGENT, double **RESOURCES,
                        lookup, agentID); 
       paras[141]++;
     }
-
+    
     if(fitnesses_n[0] < fitnesses[0]){
       pr_jump = exp(-(fitnesses[0] - fitnesses_n[0]) / temp); 
       rand_pr = runif(0, 1);

### 2022-04-15 notebook/cran_vignettes/SI9.Rmd
Update notebook with a draft of a new vignette for simulated annealing option in GMSE

@@
 title: 'The simulated annealing algorithm of GMSE'
-subtitle: 'GMSE: an R package for generalised management strategy evaluation (New function vignette)'
+subtitle: 'GMSE: an R package for generalised management strategy evaluation'
 author: | 
@@
 Unlike the [genetic algorithm](SI1.pdf), it is not a population focused approach.
+The simulated annealing algorithm of GMSE instead iterates from an initial $k = 0$ to a maximum $k_{max}$.
+For each iteration, a single solution (agent decision) exists, then a new potential solution is considered; under some conditions, the new solution is adopted and the old discarded.
+
+More specifically, the simulated annealing algorithm of GMSE first initialises a random agent decision at $k = 0$ (i.e., either a set of user actions or manager policy choices, depending on the agent).
+The initialisation is the same as in the [genetic algorithm](SI1.pdf).
+In each iteration from $k = 0$ to $k_{max}$, the agent decision is first checked to ensure that it is not over the agent's budget using the same cost constraint procedure as in the [genetic algorithm](SI1.pdf).
+If over budget, then an actions are randomly removed until the decision is again within budget.
+Next, the fitness of the current agent decision is assessed (again using the same fitness function as in the [genetic algorithm](SI1.pdf)).
+
+After checking the fitness of the current agent decision, the mutation function of the [genetic algorithm](SI1.pdf) is used to explore a new potential agent decision.
+The potential agent decision is checked and constrained to be within budget if necessary, then the fitness of the potential agent decision is compared with the fitness of the current agent decision.
+If the fitness of the potential agent decision is higher, then it replaces the current agent decision.
+If the fitness of the current agent decision is higher, then the potential agent decision replaces it with a probability of,
+
+$$\Pr(replace) = exp\left(\frac{- \left(W_{current} - W_{potential} \right)}{1 - \frac{(1 + k)}{ k_{max}}}\right).$$
+In the above, $W_{current}$ is the fitness of the current strategy, and $W_{potential}$ is the fitness of the potential strategy.
+Note that as $k \to k_{max}$, the probability of the potential strategy replacing the current strategy decreases when the latter has higher fitness.
+
+Overall, the simulated annealing approach has the potential to be faster than the [genetic algorithm](SI1.pdf) approach because it does not require a full population of evolving agent decisions.
+Below, we show how to use the simulated annealing algorithm in place of the genetic algorithm for users and managers, then we compare the performance of the two approaches.
 
@@
 
+Here we use the same example of resource management as was used in the [main introduction](ms.pdf) to the GMSE package.
+This example considers a population of waterfowl that feed off of farmland, causing conservation conflict with farmers [e.g., @Fox2017a; @Mason2017; @Tulloch2017], while managers attempt to maintain waterfowl density at a target level.
+To replicate the example from the [main introduction](ms.pdf) using simulated annealing, we can set arguments in the `gmse` function (or `gmse_apply`) for `user_annealing` and `mana_annealing` (`TRUE` or `FALSE` whether simulated annealing is used in place of the genetic algorithm for users and managers, respectively).
+An additional parameter `kmax_annealing` sets the value for $k_{max}$.
+Additional relevant parameters (e.g., mutation rate) are set using the same arguments as used in the genetic algorithm.
+The code below runs the example using simulated annealing for agent decision-making
+
+```{r, echo = TRUE}
+sim <- gmse(land_ownership = TRUE, stakeholders = 5, observe_type = 0, 
+            res_death_K = 2000, manage_target = 1000, RESOURCE_ini = 1000, 
+            user_budget = 1000, manager_budget = 1000, res_consume = 1, 
+            scaring = TRUE, plotting = FALSE, user_annealing = TRUE,
+            mana_annealing = TRUE, kmax_annealing = 100);
+```
+
+All other `gmse` parameters are set to default values. Figure 1 shows the output, which can be compared with the [same output](ms.pdf) from using the genetic algorithm.
+
+```{r, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = "Example GMSE simulation in which a natural resource is managed on the land of five farmers.
+The upper left panel shows locations of the resource (black dots) on the landscape.
+The upper right panel shows each farmers land (different colours).
+The middle left panel shows the abundance of the resource (black line) and its estimate from the perspective of the manager (blue line with 95 per cent confidence itnervals).
+The middle right planel shows farmer agricultural yield by colour.
+The lower left panel shows manager policy (costs of farmer actions) over time, and the lower right panel shows farmer actions chosen over time."}
+plot_gmse_results(sim);
+```
+
+Note that the output is the same as is produced the the genetic algorithm, and any combination of simulated annealing and the genetic algorithm can be used between managers and users with the `user_annealing` and `mana_annealing` arguments (but all users must use the same algorithm for decision-making).
+Overall, simulated annealing can give faster simulation results, but usually tends to perform worse in terms of agent decision-making, except when used for users in very simple decision-making models (e.g., farmers when the only option is culling).
+Details of the performance are described in the next section.
+
 Performance of simulated annealing
@@
 
+Here we show the results of some simulations to compare the performance of the GMSE genetic algorithm and simulated annealing algorithm.
+The purpose here is to provide some guidance as to when the simulated annealing algorithm might be preferred.
+The best choice will likely be a trade-off between simulation time and the decision-making ability of GMSE agents.
+And because agents are, [by design](ms.pdf), not intended to be perfectly rational, what constitutes desirable 'decision-making ability' might also vary with simulation objectives.
+In other words, there might be situations in which it is desirable for agents to make decisions that are very closely aligned with their goals, and other situations in which it might desirable to have agents make fast but sub-optimal decisions.
+
+Performance in terms of real computation time
+--------------------------------------------------------------------------------
+
+First, we show the results of test runs from a highly simplified example in which a single user is modelled (i.e., a lone farmer).
+In these test runs, the resource, observation, and manager model are simplified to run as quickly possible using the alternative functions introduced for demonstrating [the gmse_apply function](SI2.pdf).
+
+```{r}
+# Alternative resource sub-model
+alt_res <- function(X, K = 2000, rate = 1){
+    X_1 <- X + rate*X*(1 - X/K);
+    return(X_1);
+}
+# Alternative observation sub-model
+alt_obs <- function(resource_vector){
+    X_obs <- resource_vector - 0.1 * resource_vector;
+    return(X_obs);
+}
+
+# Alternative manager sub-model
+alt_man <- function(observation_vector){
+    policy <- observation_vector - 1000;
+    if(policy < 0){
+        policy <- 0;
+    }
+    return(policy);
+}
+```
+
+The default user sub-model is then used with a single stakeholder.
+
+```{r, eval = FALSE}
+sim_old   <- gmse_apply(res_mod = alt_res, obs_mod = alt_obs, man_mod = alt_man,
+                        get_res = "Full", stakeholders = 1, X = 10000,
+                        ga_seedrep = 0, ga_mingen = ga_min, 
+                        user_annealing = FALSE, mana_annealing = FALSE, 
+                        kmax_annealing = kmx, user_budget = 10000);
+```
+
+The `gmse_apply` function is then looped for 40 time steps across different values of `ga_min` (number of generations for the genetic algorithm) and `kmx` (number of iterations for simulated annealing).
+Higher `ga_min` and `kmx` result in longer computation times, but also potentially better user decision-making.
+The decision-making scenario is highly simplified; resources are initialised to be super-abundant and managed to maintain this abundance (i.e., there are a lot of resources and managers consistently try to keep it that way).
+The best decision for the user is therefore clear.
+The user should simply apply their budget to culling as much as possible.
+Hence, the number of resources that the user culls can be used as a metric of decision-making performance.
+We can therefore investigate the trade-off between computation time and performance in for the genetic algorithm and simulated annealing.
+The results show simulations from a PC running Xubuntu 18.04 LTS with a Intel(R) Xeon(R) W-2295 CPU (3.00 GHz) processor, 18 dual threaded cores, and 32 GB RAM.
+
+```{r, echo = FALSE, fig.width = 6, fig.height = 6, fig.cap = "Comparison of stakeholder culls (a metric of decision-making performance) across different simulation times when running a genetic algorithm versus a simulated annealing algorithm in GMSE.
+Each point represents a unique simulation of 40 time steps run at a specific number of genetic algorithm generations (ga_min) or simulated annealing iterations (kmx)."}
+ga_repl <- read.csv("data/SI9/genetic_algorithm_reps.csv");
+sa_repl <- read.csv("data/SI9/sim_annealing_reps.csv");
+GA_summ <- read.csv("data/SI9/GA_summary.csv");
+SA_summ <- read.csv("data/SI9/SA_summary.csv");
+gafit   <- read.csv("data/SI9/gafit.csv");
+safit   <- read.csv("data/SI9/safit.csv");
+plot(x = SA_summ[,2], SA_summ[,4], xlim = c(7, 19), ylim = c(0, 1000),
+     xlab = "Simulation time (sec)", ylab = "Stakeholder culls", cex = 1.5,
+     pch = 20, col = "red");
+points(x = GA_summ[,2], GA_summ[,4], cex = 1.5, pch = 20, col = "black");
+legend(x = 12, y = 200, fill = c("black", "red"), cex = 1.25,
+       legend = c("Genetic Algorithm", "Simulated Annealing"));
+```
+
+Results from Figure 2 show that given sufficient time, agents will decide to use all of their actions to cull the highest possible number of resources (1000).
+But the simulated annealing algorithm finds this solution in less time than the genetic algorithm, suggesting that it can be an effective tool when simulation time is limited (especially if many users need to be modelled within simulations).
+The situation of only culling is highly simplified.
+Users clearly benefit from each additional cull, with no other option to maximise their utility.
+Note that while this case of decision-making performance is straightforward, performance cannot be as easily measured when multiple actions are available to the user; nor can it be measured easily for the manager.
+
+User fitness increase by iteration or generation
+--------------------------------------------------------------------------------
+
+We can compare the two algorithms in a slightly different way by looking at their 'fitness' in terms of utility within GMSE.
+This fitness is explained in detail in the overview of the [genetic algorithm](SI1.pdf).
+Since fitness is calculated the same way in both algorithms, we can compare how fitness increases in genetic algorithm generations versus how it increases with iterations of the simulated annealing algorithm (keeping in mind that generation and iteration times will vary in practice given other parameters specified in GMSE).
+For this, we now use the `gmse` function parameterised as below.
+
+```{r, eval = FALSE}
+sim <- gmse(time_max = 40, land_ownership = TRUE, scaring = FALSE, 
+            culling = TRUE, castration = FALSE, help_offspring = FALSE, 
+            tend_crops = FALSE, kill_crops = FALSE, plotting = FALSE,
+            ga_seedrep = 0, user_annealing = FALSE, mana_annealing = FALSE,
+            kmax_annealing = 40, ga_mingen = 40);
+```
+
+Note that land ownership has been turned on, but only culling is available (e.g., as modelling farmers attempting to cull waterfowl before they damage their agricultural yield).
+For simulations with simulated annealing, we instead set `user_annealing = TRUE` and `mana_annealing = TRUE` in the code above.
+Instead of recording an agent's decision after it has been made, we will now look at how each generation (genetic algorithm) or iteration (simulated annealing) increases user or manager fitness (see below) on average.
+Figure 3 shows how user fitness increases on average as the number of generations in the genetic algorithm or iterations of the simulated annealing algorithm increases (note that there are 5 users in these simulations, but Figure 3 only looks at the actions of a single user).
+
+```{r, echo = FALSE, eval = TRUE, fig.cap = "Comparison of user fitness across different genetic algorithm generation numbers or simulated annealing iteration numbers in GMSE when users only have culling available as an action."}
+# Only culling
+plot(x = gafit[,1], y = gafit[,2], type = "l", lwd = 2, cex = 0.5,
+     col = "black", pch = 20, ylim = c(1800, 3600), 
+     xlab = "Generation / Iteration", ylab = "Scaled user fitness", xlim = c(1, 39));
+points(x = safit[,1], y = safit[,2], type = "l", lwd = 2, cex = 0.5,
+     col = "red", pch = 20);
+legend(x = 22, y = 2300, fill = c("black", "red"),
+       legend = c("Genetic Algorithm", "Simulated Annealing"));
+text(x = 30, y = 3500, labels = "Only Culling", cex = 1.25);
+```
+
+From Figure 3, we can see that expected user fitness increases for each algorithm, but the genetic algorithm requires fewer generations than the simulated annealing requires iterations to maximise user fitness.
+This is still, however, a highly simplified case in which culling is the only action available to users.
+We can compare this with the opposite extreme, in which all seven possible actions are available to users to maximise their fitness.
+Figure 4 shows fitness increases for each algorithm under these more complex conditions.
+
+```{r, echo = FALSE, eval = TRUE, fig.cap = "Comparison of user fitness across different genetic algorithm generation numbers or simulated annealing iteration numbers in GMSE when users have all actions available to them."}
+plot(x = gafit[,1], y = gafit[,3], type = "l", lwd = 2, cex = 0.5,
+     col = "black", pch = 20, ylim = c(1200, 4800), 
+     xlab = "Generation / Iteration", ylab = "Scaled user fitness", xlim = c(1, 39));
+points(x = safit[,1], y = safit[,3], type = "l", lwd = 2, cex = 0.5,
+     col = "red", pch = 20);
+legend(x = 0, y = 4900, fill = c("black", "red"),
+       legend = c("Genetic Algorithm", "Simulated Anealing"));
+text(x = 25, y = 1600, labels = "All possible user actions", cex = 1.5);
+```
+
+Under the more complex conditions in which all actions are possible, the genetic algorithm more rapidly asymptotes toward a high fitness solution than the simulated annealing algorithm, which suggests that the algorithm might be preferable under complex decision-making situations.
+
+Manager fitness increase by iteration or generation
+--------------------------------------------------------------------------------
+
+We can now look at the same scenarios of only culling versus all actions from the perspective of the manager.
+Figure 5 shows the fitness increase for each algorithm when the manager is attempting to maintain natural resources at a target level and culling is the only policy option that they have to use.
+
+```{r, echo = FALSE, eval = TRUE, fig.cap = "Comparison of manager fitness across different genetic algorithm generation numbers or simulated annealing iteration numbers in GMSE when managers only have culling available as a policy option."}
+# Only culling
+plot(x = gafit[,1], y = gafit[,4], type = "l", lwd = 2, cex = 0.5,
+     col = "black", pch = 20, ylim = c(0, 74000), 
+     xlab = "Generation / Iteration", ylab = "Scaled user fitness", xlim = c(1, 39));
+points(x = safit[,1], y = safit[,4], type = "l", lwd = 2, cex = 0.5,
+     col = "red", pch = 20);
+legend(x = 22, y = 20000, fill = c("black", "red"),
+       legend = c("Genetic Algorithm", "Simulated Annealing"));
+text(x = 10, y = 3500, labels = "Only Culling", cex = 1.25);
+```
+
+Figure 5 shows that while the genetic algorithm can quickly identify the policy option that maximises fitness after about 15 generations, more iterations of the simulated annealing algorithm are needed to perform with equivalent fitness.
+The difference becomes even clearer when managers have all policy options at their disposal (Figure 6).
+
+```{r, echo = FALSE, eval = TRUE, fig.cap = "Comparison of manager fitness across different genetic algorithm generation numbers or simulated annealing iteration numbers in GMSE when managers have all policy options available to them."}
+plot(x = gafit[,1], y = gafit[,5], type = "l", lwd = 2, cex = 0.5,
+     col = "black", pch = 20, ylim = c(300000, 2000000), 
+     xlab = "Generation / Iteration", ylab = "Scaled user fitness", xlim = c(1, 39));
+points(x = safit[,1], y = safit[,5], type = "l", lwd = 2, cex = 0.5,
+     col = "red", pch = 20);
+legend(x = 0, y = 1300000, fill = c("black", "red"),
+       legend = c("Genetic Algorithm", "Simulated Anealing"));
+text(x = 28, y = 1000000, labels = "All possible user actions", cex = 1.5);
+```
+
+Given the highly complex situation in which managers must choose from a range of policy options, the genetic algorithm appears to perform more strongly in terms of maximising the utility of the manager.
+
+Conclusion
+================================================================================
+
+Simulated annealing can be useful for modelling simple user decision-making.
+When user decision-making includes many potential actions, the genetic algorithm is likely to perform better.
+For managers, simulated annealing should be used very cautiously, and only when the number of potential policy options that the manager needs to consider is low.
+Overall, simulated annealing is perhaps best restricted to users for simulations where the simple decisions of a lot of users need to be modelled simultaneously.
+
 References

### 2022-04-15 notebook/sim_annealing_vs_ga.R

@@ -1,11 +1,13 @@
 
 ################################################################################
-count <- 1;
+count      <- 1;
+AN_summary <- NULL;
+GA_summary <- NULL;
 while(count < 125){
 
-time_steps <- 200;
-ga_min     <- 25 + count;
-kmx        <- 250 + (count * 10);
+time_steps <- 40;
+ga_min     <- 10 * count;
+kmx        <- 10 * count;
 
 alt_res <- function(X, K = 10000, rate = 1){
     X_1 <- X + rate*X*(1 - X/K);
@@ -69,7 +71,7 @@ colnames(sim_sum_1) <- c("Time", "Pop_size", "Pop_est", "Cull_cost",
 end_time <- Sys.time();
 tot_time <- end_time - st_time;
 gal_fcal <- sum(sim_sum_1[,6]);
-GA_stats <- c(time_steps, tot_time, gal_fcal, mean(sim_sum_1[101:200,5]));
+GA_stats <- c(time_steps, tot_time, gal_fcal, mean(sim_sum_1[31:40,5]));
 ################################################################################
 
 ################################################################################
@@ -108,7 +110,7 @@ colnames(sim_sum_2) <- c("Time", "Pop_size", "Pop_est", "Cull_cost",
 end_time <- Sys.time();
 tot_time <- end_time - st_time;
 ann_fcal <- sum(sim_sum_2[,7]);
-AN_stats <- c(time_steps, tot_time, ann_fcal, mean(sim_sum_2[101:200,5]));
+AN_stats <- c(time_steps, tot_time, ann_fcal, mean(sim_sum_2[31:40,5]));
 ################################################################################
 
 
@@ -117,6 +119,7 @@ AN_summary <- rbind(AN_summary, AN_stats);
 GA_summary <- rbind(GA_summary, GA_stats);
 
 count <- count + 1;
+print(count);
 }
 
 
@@ -137,13 +140,154 @@ sim <- gmse(time_max = 40, land_ownership = TRUE, scaring = TRUE,
             culling = TRUE, castration = TRUE, help_offspring = TRUE, 
             tend_crops = TRUE, kill_crops = TRUE, plotting = FALSE,
             ga_seedrep = 0, user_annealing = TRUE, mana_annealing = TRUE,
[…]

### 2022-04-16 notebook/cran_vignettes/SI9.Rmd
Data on the simulated annealing versus genetic algorithm

@@
 safit   <- read.csv("data/SI9/safit.csv");
+ga_fitness_user_cull <- read.csv("data/SI9/ga_fitness_user_cull.csv");
+ga_fitness_user_all  <- read.csv("data/SI9/ga_fitness_user_all.csv");
+ga_fitness_mana_cull <- read.csv("data/SI9/ga_fitness_mana_cull.csv");
+ga_fitness_mana_all  <- read.csv("data/SI9/ga_fitness_mana_all.csv");
+sa_fitness_user_cull <- read.csv("data/SI9/sa_fitness_user_cull.csv");
+sa_fitness_user_all  <- read.csv("data/SI9/sa_fitness_user_all.csv");
+sa_fitness_mana_cull <- read.csv("data/SI9/sa_fitness_mana_cull.csv");
+sa_fitness_mana_all  <- read.csv("data/SI9/sa_fitness_mana_all.csv");
 plot(x = SA_summ[,2], SA_summ[,4], xlim = c(7, 19), ylim = c(0, 1000),
@@
 
-User fitness increase by iteration or generation
+User fitness increase fitness function call
 --------------------------------------------------------------------------------
@@
             culling = TRUE, castration = FALSE, help_offspring = FALSE, 
-            tend_crops = FALSE, kill_crops = FALSE, plotting = FALSE,
-            ga_seedrep = 0, user_annealing = FALSE, mana_annealing = FALSE,
-            kmax_annealing = 40, ga_mingen = 40);
+            feeding = FALSE, tend_crops = FALSE, kill_crops = FALSE, 
+            plotting = FALSE, ga_seedrep = 0, user_annealing = FALSE, 
+            mana_annealing = FALSE, kmax_annealing = 40, ga_mingen = 40);
 ```
@@
 # Only culling
-plot(x = gafit[,1], y = gafit[,2], type = "l", lwd = 2, cex = 0.5,
-     col = "black", pch = 20, ylim = c(1800, 3600), 
-     xlab = "Generation / Iteration", ylab = "Scaled user fitness", xlim = c(1, 39));
-points(x = safit[,1], y = safit[,2], type = "l", lwd = 2, cex = 0.5,
-     col = "red", pch = 20);
-legend(x = 22, y = 2300, fill = c("black", "red"),
+plot(x = ga_fitness_user_cull[,1], y = ga_fitness_user_cull[,2], type = "l", 
+     lwd = 2, cex = 0.5, col = "black", pch = 20, ylim = c(1000, 4000), 
+     xlab = "Calls to fitness function", ylab = "Scaled user fitness", 
+     xlim = c(0, 4000));
+points(x = sa_fitness_user_cull[,1], y = sa_fitness_user_cull[,2], 
+       type = "l", lwd = 2, cex = 0.5, col = "red", pch = 20);
+legend(x = 2500, y = 1700, fill = c("black", "red"),
        legend = c("Genetic Algorithm", "Simulated Annealing"));
-text(x = 30, y = 3500, labels = "Only Culling", cex = 1.25);
+text(x = 2000, y = 3900, labels = "Only Culling", cex = 1.25);
 ```
@@
 ```{r, echo = FALSE, eval = TRUE, fig.cap = "Comparison of user fitness across different genetic algorithm generation numbers or simulated annealing iteration numbers in GMSE when users have all actions available to them."}
-plot(x = gafit[,1], y = gafit[,3], type = "l", lwd = 2, cex = 0.5,
-     col = "black", pch = 20, ylim = c(1200, 4800), 
-     xlab = "Generation / Iteration", ylab = "Scaled user fitness", xlim = c(1, 39));
-points(x = safit[,1], y = safit[,3], type = "l", lwd = 2, cex = 0.5,
-     col = "red", pch = 20);
-legend(x = 0, y = 4900, fill = c("black", "red"),
-       legend = c("Genetic Algorithm", "Simulated Anealing"));
-text(x = 25, y = 1600, labels = "All possible user actions", cex = 1.5);
+# All options
+plot(x = ga_fitness_user_all[1:40,1], y = ga_fitness_user_all[1:40,2], type = "l", 
+     lwd = 2, cex = 0.5, col = "black", pch = 20, ylim = c(0, 5000), 
+     xlab = "Calls to fitness function", ylab = "Scaled user fitness", xlim = c(0, 4000));
+points(x = sa_fitness_user_all[,1], y = sa_fitness_user_cull[,2], 
+       type = "l", lwd = 2, cex = 0.5, col = "red", pch = 20);
+legend(x = 0, y = 5000, fill = c("black", "red"),
+       legend = c("Genetic Algorithm", "Simulated Annealing"));
+text(x = 2500, y = 5000, labels = "All actions", cex = 1.25);
 ```
@@
 # Only culling
-plot(x = gafit[,1], y = gafit[,4], type = "l", lwd = 2, cex = 0.5,
-     col = "black", pch = 20, ylim = c(0, 74000), 
-     xlab = "Generation / Iteration", ylab = "Scaled user fitness", xlim = c(1, 39));
-points(x = safit[,1], y = safit[,4], type = "l", lwd = 2, cex = 0.5,
-     col = "red", pch = 20);
-legend(x = 22, y = 20000, fill = c("black", "red"),
+ofst <- 9726594;
+plot(x = ga_fitness_mana_cull[,1], y = ga_fitness_mana_cull[,2] - ofst, 
+     type = "l",  lwd = 2, cex = 0.5, col = "black", pch = 20, 
+     ylim = c(50000, 170000), xlab = "Calls to fitness function", 
+     ylab = "Scaled manager fitness", xlim = c(0, 4000));
+points(x = sa_fitness_mana_cull[,1], y = sa_fitness_mana_cull[,2] - ofst,
+       type = "l", lwd = 2, cex = 0.5, col = "red", pch = 20);
+legend(x = 2500, y = 80000, fill = c("black", "red"),
        legend = c("Genetic Algorithm", "Simulated Annealing"));
-text(x = 10, y = 3500, labels = "Only Culling", cex = 1.25);
+text(x = 500, y = 165000, labels = "Only Culling", cex = 1.25);
 ```
@@
 ```{r, echo = FALSE, eval = TRUE, fig.cap = "Comparison of manager fitness across different genetic algorithm generation numbers or simulated annealing iteration numbers in GMSE when managers have all policy options available to them."}
-plot(x = gafit[,1], y = gafit[,5], type = "l", lwd = 2, cex = 0.5,
-     col = "black", pch = 20, ylim = c(300000, 2000000), 
-     xlab = "Generation / Iteration", ylab = "Scaled user fitness", xlim = c(1, 39));
-points(x = safit[,1], y = safit[,5], type = "l", lwd = 2, cex = 0.5,
-     col = "red", pch = 20);
-legend(x = 0, y = 1300000, fill = c("black", "red"),
-       legend = c("Genetic Algorithm", "Simulated Anealing"));
-text(x = 28, y = 1000000, labels = "All possible user actions", cex = 1.5);
+# All
+ofst <- 9035476;
+plot(x = ga_fitness_mana_all[,1], y = ga_fitness_mana_all[,2] - ofst, 
+     type = "l",  lwd = 2, cex = 0.5, col = "black", pch = 20, 
+     ylim = c(500000, 900000), xlab = "Calls to fitness function", 
+     ylab = "Scaled manager fitness", xlim = c(0, 4000));
+points(x = sa_fitness_mana_all[,1], y = sa_fitness_mana_all[,2] - ofst,
+       type = "l", lwd = 2, cex = 0.5, col = "red", pch = 20);
+legend(x = 0, y = 750000, fill = c("black", "red"),
+       legend = c("Genetic Algorithm", "Simulated Annealing"));
+text(x = 2000, y = 900000, labels = "All actions", cex = 1.25);
 ```

### 2022-04-16 notebook/sim_annealing_vs_ga.R

@@ -1,13 +1,16 @@
 
+
+## REDO WITH A GA POPULATION SIZE OF 100
+
 ################################################################################
 count      <- 1;
 AN_summary <- NULL;
 GA_summary <- NULL;
-while(count < 125){
+while(count < 200){
 
 time_steps <- 40;
-ga_min     <- 10 * count;
-kmx        <- 10 * count;
+ga_min     <- 1 * count;
+kmx        <- 100 * count;
 
 alt_res <- function(X, K = 10000, rate = 1){
     X_1 <- X + rate*X*(1 - X/K);
@@ -130,17 +133,17 @@ print(count);
 ################################################################################
 # Side-by-side fair comparison
 ################################################################################
-sim <- gmse(time_max = 40, land_ownership = TRUE, scaring = TRUE, 
-            culling = TRUE, castration = TRUE, help_offspring = TRUE, 
-            tend_crops = TRUE, kill_crops = TRUE, plotting = FALSE,
-            ga_seedrep = 0);
+sim <- gmse(time_max = 40, land_ownership = TRUE, scaring = FALSE, 
+            culling = TRUE, castration = FALSE, help_offspring = FALSE, 
+            tend_crops = FALSE, kill_crops = FALSE, feeding = FALSE, 
+            plotting = FALSE, ga_seedrep = 0);
 
 
-sim <- gmse(time_max = 40, land_ownership = TRUE, scaring = TRUE, 
-            culling = TRUE, castration = TRUE, help_offspring = TRUE, 
-            tend_crops = TRUE, kill_crops = TRUE, plotting = FALSE,
-            ga_seedrep = 0, user_annealing = TRUE, mana_annealing = TRUE,
-            kmax_annealing = 40);
+sim <- gmse(time_max = 40, land_ownership = TRUE, scaring = FALSE, 
+            culling = TRUE, castration = FALSE, help_offspring = FALSE, 
+            tend_crops = FALSE, kill_crops = FALSE, feeding = FALSE,
+            plotting = FALSE, ga_seedrep = 0, user_annealing = TRUE, 
+            mana_annealing = FALSE, kmax_annealing = 2000);
 
 
 end_time <- Sys.time();
@@ -289,9 +292,177 @@ legend(x = 14, y = 90, fill = c("black", "red"), cex = 1.25,
 
[…]

### 2022-06-15 tests/testthat/test-make_utilities.R
Skip the make utilities test on CRAN because it was causing problems with the Fedora libraries, for some reason

@@ -3,6 +3,7 @@ library("GMSE");
 context("Action array initialisation");
 
 test_that("Dimensions of action array are correct", {
+    skip_on_cran();
     agents  <-  make_agents(model        = "IBM",
                             agent_number = 2,
                             type_counts  = c(1,1),
@@ -34,6 +35,7 @@ test_that("Dimensions of action array are correct", {
 })
 
 test_that("Values on the action array are accurate", {
+    skip_on_cran();
     agents  <-  make_agents(model        = "IBM",
                             agent_number = 2,
                             type_counts  = c(1,1),
@@ -65,6 +67,7 @@ test_that("Values on the action array are accurate", {
 })
 
 test_that("IDs on the action array are accurate", {
+    skip_on_cran();
     ID_vec <- c(-2, -2, -1, 1, 1, 2, 2);
     agents  <-  make_agents(model        = "IBM",
                             agent_number = 2,
